### 2023-12-05 paper/paper-new.Rmd
init

---
title             : "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
shorttitle        : "Ordinal regression models made easy"

author: 
  - name          : "Filippo Gambarota"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "filippo.gambarota@unipd.it"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Gianmarco Altoè"
    affiliation   : "1"
    role:
      - "Writing - Review & Editing"
      - "Supervision"

affiliation:
  - id            : "1"
    institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |
  Abstract here

keywords          : "ordinal, likert, simulations, power"
wordcount         : "X"

bibliography      : "`r filor::fil()$bib`"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output: 
  papaja::apa6_pdf:
    latex_engine: xelatex
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      dev = "pdf",
                      fig.align = "center",
                      message = FALSE,
                      warning = FALSE)
```

```{r packages, include = FALSE}
library(tidyverse)
library(ordinal)
devtools::load_all()
```

# Introduction

Da citare in qualche modo @Kemp2021-dj

> “As a matter of fact, most of the scales used widely and effectively by psychologists are ordinal scales. In the strictest propriety the ordinary statistics involving means and standard deviations ought not to be used with these scales, for these statistics imply a knowledge of something more than the relative rank-order of data. On the other hand for this ‘illegal’ statisticizing there can be invoked a kind of pragmatic sanction: In numerous instances it leads to fruitful results.” (Stevens, 1946, p. 679.)

> It is straightforward to construct ordinal scales that do not involve rank ordering. For example, one can take the first element encountered and arbitrarily assign it the number 100. If the next element encountered is smaller it is given a smaller arbitrary number, 53 say. If the third element is between these two, it can be given the number 86, and so on. If this construction method is used, the difference between the elements assigned the numbers 70 and 80 will not in any important sense be equal to the difference between the elements assigned 90 and 100, and the intervals between the numbers are not really interpretable. Note, too, that monotonic transformations of the scale essentially leave the mea- sure unaffected.

interessante come @Cliff2016-ck descriva che la maggiorparte dellle domande di ricerca in psicologia siano in riferimento alla location di un costrutto e possano essere gestite in modo ordinale

The use of ordinal data is widespread in Psychology. Usually items from questionnaires are created using Likert scales where a certain psychological traits is measured using an intuitive scale with 3-5 or more anchor points. These measure cannot be considered *metric* measure on interval or ratio scales [stevens] but the categories are ordered.

Other examples here

Despite the usage of ordinal variables, statistical models made for these type of data are rarely used in Psychology. @Liddell2018-wu reported that the majority of published papers using likert-like measures used standard methods to analyze the data. In practical terms, they use *metric* models where the response variable cannot be considered fully numeric (vedi se c'è un termine tipo scala a rapporti).

altro sui modelli

Theoretically, using a *metric* for ordinal data is not appropriate but understanding the actual impact is not straightforward. @Liddell2018-wu did a comprehensive work about pitfalls of analzying ordinal data as metric. They showed that metric models produce higher type-1 and type-2 errors compared to the ordinal models. In particular, given the bounded nature of ordinal data, difference in the underlying latent distribution are not always captured by the metric model that simply estimate the mean of the ordinal variable. This is even more relevant when the underluing variance of the ordinal variables are not homogeneous. Furthermore they presented some situations where the metric model could be wrong in the opposite direction, finding an effect with the wrong sign (type-s error).

Finally, they also demonstrated that even in the best condition where e.g. comparing two groups the variance are equal, the ordinal model is more powerful than the metric model given the underestimation of the true effect size from the latter.

> Some authors have argued that, despite the ordinal character of individual Likert items, averaged ordinal items can have an emergent property of an interval scale and so it is appropriate to apply metric methods to the averaged values (e.g., Carifio & Perla, 2007, 2008).

vedi come gestire questo

anche questo è importante
> Ordered-probit models typically assume that the thresholds (θ k) are the same across all groups because the thresholds are theoretically linked to the response measure, not to the predictor value. For example, when asked, “How happy are you?” with response options ‘1’ = very unhappy, ‘2’ = mildly unhappy, ‘3’ = neutral, ‘4’ = mildly happy, ‘5’ = very happy,” the latent thresholds between ordinal levels are assumed to be implicit in the phrasing of the question, regardless of other aspects of the respondent or situation. In other words, the thresholds are assumed to be part of the measurement procedure, not dependent on the value of the predictor or covariate. This can be technically referred to as a type of measurement invariance.

Da citare come paper iniziale sui modelli [@McCullagh1980-cw].

questo per la tassonomia dei diversi modelli ordinali [@Tutz2022-dg]

da vedere magari per una critica tipo Gomila sul binary model [@Robitzsch2020-la]

## Demonstrating the proportional odds assumption

Let's simulate the effect of a binary predictor on ordinal scale 1-5:

```{r}
#| echo: true
b1 <- log(3) # log odds ratio
n <- 1e4
x <- rep(c("a", "b"), each = n/2)
dat <- data.frame(x = x)
probs <- rep(1/5, 5) # for the group "a", uniform probabilities
dat <- sim_ord_latent(~x, b1, probs = probs, data = dat, link = "logit")
fit <- clm(y ~ x, data = dat, link = "logit")
pr <- predict(fit, data.frame(x = unique(x)))$fit
pr
```
Basically the proportional odds suggest that:

$$
\text log(\frac{P(y \leq 1)}{P(y > 1)})
$$
Is the same regardless the level of the $x$ predictor. Thus:

$$
\text{log}\left(\frac{\frac{P(y \leq 1|x_0)}{P(y > 1|x_0)}}{\frac{P(y \leq 2|x_0)}{P(y > 2|x_0)}}\right) = \text{log}\left(\frac{\frac{P(y \leq 1|x_1)}{P(y > 1|x_1)}}{\frac{P(y \leq 2|x_1)}{P(y > 2|x_1)}}\right)
$$

```{r}
#| echo: true
a_or1vs2345 <- filor::odds(pr[1, 1]) # 1 vs 2 3 4 5
a_or12vs345 <- filor::odds(sum(pr[1, 1:2])) # 1 vs 2 3 4 5

b_or1vs2345 <- filor::odds(pr[2, 1]) # 1 vs 2 3 4 5
b_or12vs345 <- filor::odds(sum(pr[2, 1:2])) # 1 vs 2 3 4 5

c(xa = log(a_or1vs2345 / a_or12vs345), xb = log(b_or1vs2345 / b_or12vs345))
```

## Logit vs Probit model

When fitting an ordinal regression the two mostly used link functions are the *probit* and *logit*. From the distribution point of view the two functions are very similar. The *probit* model is based on a cumulative Gaussian distribution while the *logit* model is based on a logistic distribution. Figure \@ref(fig:logit-vs-probit) depict the two cumulative distributions.

```{r logit-vs-probit, fig.cap="Logit vs Probit"}
par(mfrow = c(1,2))

# density

curve(dnorm(x), -5, 5, lwd = 2, ylab = "Density", main = "Probability Density", cex.lab = 1.2, cex.axes = 1.2)
curve(dlogis(x), -5, 5, lwd = 2, col = "firebrick", add = TRUE)

# cdf

curve(pnorm(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", lwd = 2, main = "Cumulative Probability",
      cex.lab = 1.2, cex.axes = 1.2)
curve(plogis(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", col = "firebrick", lwd = 2, add = TRUE)
```

Given the different underlying distribution, the parameters have a different interpretation under the two models. The probit model assume a latent standard normal distribution with and $\sigma = 1$. The logit model assume a logistic distribution with $\sigma^2 = \pi^2/3$. Thus regression coefficients $\beta$ represents the increase in standard deviations units. The interpretation in terms of latent distribution is particularly useful for the probit model where $\beta$s can be interpreted in a Cohen's $d$ like manner. Furthermore, there is the possibility to directly map parameters from signal detection theory [@Stanislaw1999-jr; @Green1966-gy] into an ordinal probit regression. In practical terms, the thresholds are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].

## Simulating data

There are mainly two ways to simulate data. The first method concerns simulating starting from the latent formulation of the ordinal model. Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.

```{r}
library(distributional)
library(ggdist)
library(cowplot)

probs <- rep(1/5, 5)
ths <- probs_to_th(probs)

lat <- data.frame(x = c(0, 1),
                  m = c(0, 2),
                  s = c(1, 1))

lat$dist <- dist_normal(lat$m, lat$s)

latent_plot <- lat |> 
  ggplot(aes(x = x, y = m, dist = dist)) +
  stat_halfeye(aes(fill = after_stat(factor(findInterval(y, ths))))) +
  ylim(c(-4, 6)) +
  geom_hline(yintercept = probs_to_th(probs)) +
  theme(legend.position = "none")

probs_plot <- get_probs(~x, 2, probs, lat) |> 
  pivot_longer(starts_with("y")) |> 
  ggplot(aes(y = factor(x), x = value, fill = name)) +
  geom_col(position = position_dodge())

plot_grid(latent_plot, probs_plot)
```

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

### 2023-12-05 paper/paper-new.tex
init

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{ordinal, likert, simulations, power\newline\indent Word count: X}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.},
  pdfauthor={Filippo Gambarota1 \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={ordinal, likert, simulations, power},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.}
\author{Filippo Gambarota\textsuperscript{1} \& Gianmarco Altoè\textsuperscript{1}}
\date{}

\shorttitle{Ordinal regression models made easy}

\authornote{

Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

Enter author note here.

The authors made the following contributions. Filippo Gambarota: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Gianmarco Altoè: Writing - Review \& Editing, Supervision.

Correspondence concerning this article should be addressed to Filippo Gambarota, Postal address. E-mail: \href{mailto:filippo.gambarota@unipd.it}{\nolinkurl{filippo.gambarota@unipd.it}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Developmental Psychology and Socialization, University of Padova, Italy}

\abstract{%
Abstract here
}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

Da citare in qualche modo Kemp and Grace (2021)

\begin{quote}
``As a matter of fact, most of the scales used widely and effectively by psychologists are ordinal scales. In the strictest propriety the ordinary statistics involving means and standard deviations ought not to be used with these scales, for these statistics imply a knowledge of something more than the relative rank-order of data. On the other hand for this `illegal' statisticizing there can be invoked a kind of pragmatic sanction: In numerous instances it leads to fruitful results.'' (Stevens, 1946, p.~679.)
\end{quote}

\begin{quote}
It is straightforward to construct ordinal scales that do not involve rank ordering. For example, one can take the first element encountered and arbitrarily assign it the number 100. If the next element encountered is smaller it is given a smaller arbitrary number, 53 say. If the third element is between these two, it can be given the number 86, and so on. If this construction method is used, the difference between the elements assigned the numbers 70 and 80 will not in any important sense be equal to the difference between the elements assigned 90 and 100, and the intervals between the numbers are not really interpretable. Note, too, that monotonic transformations of the scale essentially leave the mea- sure unaffected.
\end{quote}

interessante come Cliff (2016) descriva che la maggiorparte dellle domande di ricerca in psicologia siano in riferimento alla location di un costrutto e possano essere gestite in modo ordinale

The use of ordinal data is widespread in Psychology. Usually items from questionnaires are created using Likert scales where a certain psychological traits is measured using an intuitive scale with 3-5 or more anchor points. These measure cannot be considered \emph{metric} measure on interval or ratio scales {[}stevens{]} but the categories are ordered.

Other examples here

Despite the usage of ordinal variables, statistical models made for these type of data are rarely used in Psychology. Liddell and Kruschke (2018) reported that the majority of published papers using likert-like measures used standard methods to analyze the data. In practical terms, they use \emph{metric} models where the response variable cannot be considered fully numeric (vedi se c'è un termine tipo scala a rapporti).

altro sui modelli

Theoretically, using a \emph{metric} for ordinal data is not appropriate but understanding the actual impact is not straightforward. Liddell and Kruschke (2018) did a comprehensive work about pitfalls of analzying ordinal data as metric. They showed that metric models produce higher type-1 and type-2 errors compared to the ordinal models. In particular, given the bounded nature of ordinal data, difference in the underlying latent distribution are not always captured by the metric model that simply estimate the mean of the ordinal variable. This is even more relevant when the underluing variance of the ordinal variables are not homogeneous. Furthermore they presented some situations where the metric model could be wrong in the opposite direction, finding an effect with the wrong sign (type-s error).

Finally, they also demonstrated that even in the best condition where e.g.~comparing two groups the variance are equal, the ordinal model is more powerful than the metric model given the underestimation of the true effect size from the latter.

\begin{quote}
Some authors have argued that, despite the ordinal character of individual Likert items, averaged ordinal items can have an emergent property of an interval scale and so it is appropriate to apply metric methods to the averaged values (e.g., Carifio \& Perla, 2007, 2008).
\end{quote}

vedi come gestire questo

anche questo è importante
\textgreater{} Ordered-probit models typically assume that the thresholds (θ k) are the same across all groups because the thresholds are theoretically linked to the response measure, not to the predictor value. For example, when asked, ``How happy are you?'' with response options `1' = very unhappy, `2' = mildly unhappy, `3' = neutral, `4' = mildly happy, `5' = very happy,'' the latent thresholds between ordinal levels are assumed to be implicit in the phrasing of the question, regardless of other aspects of the respondent or situation. In other words, the thresholds are assumed to be part of the measurement procedure, not dependent on the value of the predictor or covariate. This can be technically referred to as a type of measurement invariance.

Da citare come paper iniziale sui modelli (McCullagh, 1980).

questo per la tassonomia dei diversi modelli ordinali (Tutz, 2022)

da vedere magari per una critica tipo Gomila sul binary model (Robitzsch, 2020)

\subsection{Demonstrating the proportional odds assumption}\label{demonstrating-the-proportional-odds-assumption}

Let's simulate the effect of a binary predictor on ordinal scale 1-5:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{b1 }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(}\DecValTok{3}\NormalTok{) }\CommentTok{\# log odds ratio}
\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e4}
\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x)}
\NormalTok{probs }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{) }\CommentTok{\# for the group "a", uniform probabilities}
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, b1, }\AttributeTok{probs =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}
\NormalTok{pr}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##            1         2         3         4         5
## 1 0.18711805 0.2021825 0.2062092 0.2020689 0.2024214
## 2 0.07132155 0.1040579 0.1540144 0.2385607 0.4320454
\end{verbatim}

Basically the proportional odds suggest that:

\[
\text log(\frac{P(y \leq 1)}{P(y > 1)})
\]
Is the same regardless the level of the \(x\) predictor. Thus:

\[
\text{log}\left(\frac{\frac{P(y \leq 1|x_0)}{P(y > 1|x_0)}}{\frac{P(y \leq 2|x_0)}{P(y > 2|x_0)}}\right) = \text{log}\left(\frac{\frac{P(y \leq 1|x_1)}{P(y > 1|x_1)}}{\frac{P(y \leq 2|x_1)}{P(y > 2|x_1)}}\right)
\]

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{a\_or1vs2345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{a\_or12vs345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}

\NormalTok{b\_or1vs2345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
\NormalTok{b\_or12vs345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}

\FunctionTok{c}\NormalTok{(}\AttributeTok{xa =} \FunctionTok{log}\NormalTok{(a\_or1vs2345 }\SpecialCharTok{/}\NormalTok{ a\_or12vs345), }\AttributeTok{xb =} \FunctionTok{log}\NormalTok{(b\_or1vs2345 }\SpecialCharTok{/}\NormalTok{ b\_or12vs345))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##        xa        xb 
## -1.018593 -1.018593
\end{verbatim}

\subsection{Logit vs Probit model}\label{logit-vs-probit-model}

When fitting an ordinal regression the two mostly used link functions are the \emph{probit} and \emph{logit}. From the distribution point of view the two functions are very similar. The \emph{probit} model is based on a cumulative Gaussian distribution while the \emph{logit} model is based on a logistic distribution. Figure \ref{fig:logit-vs-probit} depict the two cumulative distributions.

\begin{figure}

{\centering \includegraphics{paper-new_files/figure-latex/logit-vs-probit-1}

}

\caption{Logit vs Probit}\label{fig:logit-vs-probit}
\end{figure}

Given the different underlying distribution, the parameters have a different interpretation under the two models. The probit model assume a latent standard normal distribution with and \(\sigma = 1\). The logit model assume a logistic distribution with \(\sigma^2 = \pi^2/3\). Thus regression coefficients \(\beta\) represents the increase in standard deviations units. The interpretation in terms of latent distribution is particularly useful for the probit model where \(\beta\)s can be interpreted in a Cohen's \(d\) like manner. Furthermore, there is the possibility to directly map parameters from signal detection theory (Green, Swets, \& Others, 1966; Stanislaw \& Todorov, 1999) into an ordinal probit regression. In practical terms, the thresholds are the criterion cut-offs and the \(\beta_1\) is the d\('\) parameter (DeCarlo, 1998; Knoblauch \& Maloney, 2012).

\subsection{Simulating data}\label{simulating-data}

There are mainly two ways to simulate data. The first method concerns simulating starting from the latent formulation of the ordinal model. Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.

\begin{figure}

{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-3-1}

}

\caption{ }\label{fig:unnamed-chunk-3}
\end{figure}

\newpage

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Cliff2016-ck}
Cliff, N. (2016). \emph{Ordinal methods for behavioral data analysis}. London, England: Psychology Press.

\bibitem[\citeproctext]{ref-DeCarlo1998-ay}
DeCarlo, L. T. (1998). Signal detection theory and generalized linear models. \emph{Psychological Methods}, \emph{3}, 186--205. \url{https://doi.org/10.1037/1082-989X.3.2.186}

\bibitem[\citeproctext]{ref-Green1966-gy}
Green, D. M., Swets, J. A., \& Others. (1966). \emph{Signal detection theory and psychophysics} (Vol. 1, pp. xi, 455--xi, 455). Oxford, England: Wiley New York.

\bibitem[\citeproctext]{ref-Kemp2021-dj}
Kemp, S., \& Grace, R. C. (2021). Using ordinal scales in psychology. \emph{Methods in Psychology (Online)}, \emph{5}, 100054. \url{https://doi.org/10.1016/j.metip.2021.100054}

\bibitem[\citeproctext]{ref-Knoblauch2012-to}
Knoblauch, K., \& Maloney, L. T. (2012). \emph{Modeling psychophysical data in r}. New York, NY: Springer New York. \url{https://doi.org/10.1007/978-1-4614-4475-6}

\bibitem[\citeproctext]{ref-Liddell2018-wu}
Liddell, T. M., \& Kruschke, J. K. (2018). Analyzing ordinal data with metric models: What could possibly go wrong? \emph{Journal of Experimental Social Psychology}, \emph{79}, 328--348. \url{https://doi.org/10.1016/j.jesp.2018.08.009}

\bibitem[\citeproctext]{ref-McCullagh1980-cw}
McCullagh, P. (1980). Regression models for ordinal data. \emph{Journal of the Royal Statistical Society}, \emph{42}, 109--127. \url{https://doi.org/10.1111/j.2517-6161.1980.tb01109.x}

\bibitem[\citeproctext]{ref-Robitzsch2020-la}
Robitzsch, A. (2020). Why ordinal variables can (almost) always be treated as continuous variables: Clarifying assumptions of robust continuous and ordinal factor analysis estimation methods. \emph{Frontiers in Education}, \emph{5}, 589965. \url{https://doi.org/10.3389/feduc.2020.589965}

\bibitem[\citeproctext]{ref-Stanislaw1999-jr}
Stanislaw, H., \& Todorov, N. (1999). Calculation of signal detection theory measures. \emph{Behavior Research Methods, Instruments, \& Computers: A Journal of the Psychonomic Society, Inc}, \emph{31}, 137--149. \url{https://doi.org/10.3758/bf03207704}

\bibitem[\citeproctext]{ref-Tutz2022-dg}
Tutz, G. (2022). Ordinal regression: A review and a taxonomy of models. \emph{Wiley Interdisciplinary Reviews. Computational Statistics}, \emph{14}. \url{https://doi.org/10.1002/wics.1545}

\end{CSLReferences}

\end{document}

### 2023-12-05 paper/paper.rmd
init

--- 2023-12-05 paper/paper-new.Rmd
@@
 ---
-title             : "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
-shorttitle        : "Ordinal regression models made easy"
+title             : "The title"
+shorttitle        : "Title"
 
 author: 
-  - name          : "Filippo Gambarota"
+  - name          : "First Author"
     affiliation   : "1"
@@
     address       : "Postal address"
-    email         : "filippo.gambarota@unipd.it"
+    email         : "my@email.com"
     role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
@@
       - "Writing - Review & Editing"
-  - name          : "Gianmarco Altoè"
-    affiliation   : "1"
+  - name          : "Ernst-August Doelle"
+    affiliation   : "1,2"
     role:
@@
   - id            : "1"
-    institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"
+    institution   : "Wilhelm-Wundt-University"
+  - id            : "2"
+    institution   : "Konstanz Business School"
 
@@
 abstract: |
-  Abstract here
+  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
 
-keywords          : "ordinal, likert, simulations, power"
+  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
+
+  One sentence clearly stating the **general problem** being addressed by  this particular study.
+
+  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
+
+  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
+
+  One or two sentences to put the results into a more **general context**.
+
+  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
+
+  <!-- https://tinyurl.com/ybremelq -->
+
+keywords          : "keywords"
 wordcount         : "X"
 
-bibliography      : "`r filor::fil()$bib`"
+bibliography      : "r-references.bib"
 
-floatsintext      : yes
-linenumbers       : no
+floatsintext      : no
+linenumbers       : yes
 draft             : no
@@
 classoption       : "man"
-output: 
-  papaja::apa6_pdf:
-    latex_engine: xelatex
+output            : papaja::apa6_pdf
 ---
@@
 ```{r setup, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE,
-                      dev = "pdf",
-                      fig.align = "center",
-                      message = FALSE,
-                      warning = FALSE)
+library("papaja")
+r_refs("r-references.bib")
 ```
 
-```{r packages, include = FALSE}
-library(tidyverse)
-library(ordinal)
-devtools::load_all()
+```{r analysis-preferences}
+# Seed for random number generation
+set.seed(42)
+knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
 ```
@@
 
-Da citare in qualche modo @Kemp2021-dj
-
-> “As a matter of fact, most of the scales used widely and effectively by psychologists are ordinal scales.
-In the strictest propriety the ordinary statistics involving means and standard deviations ought not to be used with these scales, for these statistics imply a knowledge of something more than the relative rank-order of data.
-On the other hand for this ‘illegal’ statisticizing there can be invoked a kind of pragmatic sanction: In numerous instances it leads to fruitful results.” (Stevens, 1946, p.
-679.)
-
-> It is straightforward to construct ordinal scales that do not involve rank ordering.
-For example, one can take the first element encountered and arbitrarily assign it the number 100.
-If the next element encountered is smaller it is given a smaller arbitrary number, 53 say.
-If the third element is between these two, it can be given the number 86, and so on.
-If this construction method is used, the difference between the elements assigned the numbers 70 and 80 will not in any important sense be equal to the difference between the elements assigned 90 and 100, and the intervals between the numbers are not really interpretable.
-Note, too, that monotonic transformations of the scale essentially leave the mea- sure unaffected.
-
-interessante come @Cliff2016-ck descriva che la maggiorparte dellle domande di ricerca in psicologia siano in riferimento alla location di un costrutto e possano essere gestite in modo ordinale
+Interessante come
 
@@
 
-## Demonstrating the proportional odds assumption
-
-Let's simulate the effect of a binary predictor on ordinal scale 1-5:
-
-```{r}
-#| echo: true
-b1 <- log(3) # log odds ratio
-n <- 1e4
-x <- rep(c("a", "b"), each = n/2)
-dat <- data.frame(x = x)
-probs <- rep(1/5, 5) # for the group "a", uniform probabilities
-dat <- sim_ord_latent(~x, b1, probs = probs, data = dat, link = "logit")
-fit <- clm(y ~ x, data = dat, link = "logit")
-pr <- predict(fit, data.frame(x = unique(x)))$fit
-pr
-```
-Basically the proportional odds suggest that:
-
-$$
-\text log(\frac{P(y \leq 1)}{P(y > 1)})
-$$
-Is the same regardless the level of the $x$ predictor. Thus:
-
-$$
-\text{log}\left(\frac{\frac{P(y \leq 1|x_0)}{P(y > 1|x_0)}}{\frac{P(y \leq 2|x_0)}{P(y > 2|x_0)}}\right) = \text{log}\left(\frac{\frac{P(y \leq 1|x_1)}{P(y > 1|x_1)}}{\frac{P(y \leq 2|x_1)}{P(y > 2|x_1)}}\right)
-$$
-
-```{r}
-#| echo: true
-a_or1vs2345 <- filor::odds(pr[1, 1]) # 1 vs 2 3 4 5
-a_or12vs345 <- filor::odds(sum(pr[1, 1:2])) # 1 vs 2 3 4 5
-
-b_or1vs2345 <- filor::odds(pr[2, 1]) # 1 vs 2 3 4 5
-b_or12vs345 <- filor::odds(sum(pr[2, 1:2])) # 1 vs 2 3 4 5
-
-c(xa = log(a_or1vs2345 / a_or12vs345), xb = log(b_or1vs2345 / b_or12vs345))
-```
-
-## Logit vs Probit model
-
-When fitting an ordinal regression the two mostly used link functions are the *probit* and *logit*.
-From the distribution point of view the two functions are very similar.
-The *probit* model is based on a cumulative Gaussian distribution while the *logit* model is based on a logistic distribution.
-Figure \@ref(fig:logit-vs-probit) depict the two cumulative distributions.
-
-```{r logit-vs-probit, fig.cap="Logit vs Probit"}
-par(mfrow = c(1,2))
-
-# density
-
-curve(dnorm(x), -5, 5, lwd = 2, ylab = "Density", main = "Probability Density", cex.lab = 1.2, cex.axes = 1.2)
-curve(dlogis(x), -5, 5, lwd = 2, col = "firebrick", add = TRUE)
-
-# cdf
-
-curve(pnorm(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", lwd = 2, main = "Cumulative Probability",
-      cex.lab = 1.2, cex.axes = 1.2)
-curve(plogis(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", col = "firebrick", lwd = 2, add = TRUE)
-```
-
-Given the different underlying distribution, the parameters have a different interpretation under the two models.
-The probit model assume a latent standard normal distribution with and $\sigma = 1$.
-The logit model assume a logistic distribution with $\sigma^2 = \pi^2/3$.
-Thus regression coefficients $\beta$ represents the increase in standard deviations units.
-The interpretation in terms of latent distribution is particularly useful for the probit model where $\beta$s can be interpreted in a Cohen's $d$ like manner.
-Furthermore, there is the possibility to directly map parameters from signal detection theory [@Stanislaw1999-jr; @Green1966-gy] into an ordinal probit regression.
-In practical terms, the thresholds are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].
-
-## Simulating data
-
-There are mainly two ways to simulate data.
-The first method concerns simulating starting from the latent formulation of the ordinal model.
-Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.
-
-```{r}
-library(distributional)
-library(ggdist)
-library(cowplot)
-
-probs <- rep(1/5, 5)
-ths <- probs_to_th(probs)
-
-lat <- data.frame(x = c(0, 1),
-                  m = c(0, 2),
-                  s = c(1, 1))
-
-lat$dist <- dist_normal(lat$m, lat$s)
-
-latent_plot <- lat |> 
-  ggplot(aes(x = x, y = m, dist = dist)) +
-  stat_halfeye(aes(fill = after_stat(factor(findInterval(y, ths))))) +
-  ylim(c(-4, 6)) +
-  geom_hline(yintercept = probs_to_th(probs)) +
-  theme(legend.position = "none")
-
-probs_plot <- get_probs(~x, 2, probs, lat) |> 
-  pivot_longer(starts_with("y")) |> 
-  ggplot(aes(y = factor(x), x = value, fill = name)) +
-  geom_col(position = position_dodge())
-
-plot_grid(latent_plot, probs_plot)
-```
-
 \newpage

### 2023-12-05 R/utils.R

@@ -1,0 +1,172 @@
+th_to_probs <- function(th, link = c("logit", "probit"), ...){
+  link <- match.arg(link)
+  if(link == "logit"){
+    lfun <- plogis
+  }else{
+    lfun <- pnorm
+  }
+  ths <- c(-Inf, unname(th), Inf)
+  cprobs <- lfun(ths, ...)
+  probs <- diff(cprobs)
+  return(probs)
+}
+
+probs_to_th <- function(probs, link = c("logit", "probit")){
+  link <- match.arg(link)
+  if(link == "logit"){
+    lfun <- qlogis
+  }else{
+    lfun <- qnorm
+  }
+  th <- lfun(Reduce(sum, probs, accumulate = TRUE))
+  th[-length(th)]
+}
+
+# get_probs <- function(formula, 
+#                       B, 
+#                       probs = NULL, 
+#                       th = NULL, 
+#                       ynames = NULL, 
+#                       data, 
+#                       get_lp = FALSE, 
+#                       link = c("logit", "probit")){
+#   link <- match.arg(link)
+#   X <- model.matrix(formula, data = data)
+#   if(is.null(th)){
+#     th <- probs_to_th(probs, link = "probit")
+#   }
+#   probs <- th_to_probs(th, "probit")
+#   
+#   pnorm(as.vector(B %*% X[, -1]), mean = as.vector(B %*% X[, -1]))
+#   
+#   
+#   
+#   
+#   
+#   
+#   lp <- lapply(ths, function(x) as.vector(pnorm(X %*% c(x, B))))
+#   lp <- data.frame(lp)
+#   probs <- data.frame(t(diff(t(lp))))
+#   
+#   
+#   
+#   
+#   return(probs)
+# }
+
+get_y <- function(data = NULL, P, ordered = TRUE){
+  y <- apply(P, 1, function(x) sample(names(P), 1, replace = TRUE, prob = x))
+  y <- factor(y, levels = names(P), ordered = TRUE)
+  if(!is.null(data)){
+    cbind(data, y)
+  }else{
+    y
+  }
+}
+
+show_th <- function(th = NULL, probs = NULL, link = c("logit", "probit"),
+                    cnames = NULL){
+  link <- match.arg(link)
+  if(!is.null(th) & !is.null(probs)){
+    stop("Only one between th and probs need to be supplied")
+  }
+  
+  fn <- get_link(link)
+  
[…]

### 2023-12-05 review.R

@@ -1,0 +1,95 @@
+# underestimating the true d ordinal vs metric
+
+library(ordinal)
+
+d <- 0.5
+n <- 50
+g <- c("a", "b")
+nsim <- 1e3
+b0 <- 0
+b1 <- d
+s <- 2
+
+d_clm <- rep(NA, nsim)
+d_lm <- rep(NA, nsim)
+p_clm <- rep(NA, nsim)
+p_lm <- rep(NA, nsim)
+s_lm <- rep(NA, nsim)
+
+for(i in 1:nsim){
+  group <- rep(g, each = n/2)
+  ystar <- b0 + b1 * ifelse(group == "a", 0, 1) + rnorm(n)
+  y <- cut(ystar, c(-Inf, probs_to_th(rep(1/4, 4)), Inf))
+  y <- ordered(as.integer(y))
+  yn <- as.integer(y)
+  fit <- clm(y ~ group, link = "probit")
+  fit_lm <- lm(yn ~ group)
+  d_clm[i] <- coef(fit)[4]
+  d_lm[i] <- coef(fit_lm)[2]
+  s_lm[i] <- sigma(fit_lm)
+  p_clm[i] <- summary(fit)$coefficients["groupb", 4]
+  p_lm[i] <- summary(fit_lm)$coefficients[2, 4]
+  filor::pb(1000, i)
+}
+
+d_lms <- d_lm / s_lm
+
+plot(d_lms, d_clm)
+
+mean(p_lm <= 0.05)
+mean(p_clm <= 0.05)
+
+
+
+n <- 1e4
+group <- rep(g, each = n/2)
+ystar <- b0 + b1 * ifelse(group == "a", 0, 1) + rnorm(n)
+y <- cut(ystar, c(-Inf, probs_to_th(rep(1/4, 4), link = "logit"), Inf))
+y <- ordered(as.integer(y))
+yn <- as.integer(y)
+fit <- clm(y ~ group, link = "logit")
+
+B <- c(b1)
+X <- model.matrix(~group)
+TH <- c(-Inf, probs_to_th(rep(1/4, 4), link = "logit"), Inf)
+
+lp <- lapply(TH, function(th) as.vector(plogis(X %*% c(th, B))))
+lp <- data.frame(lp)
+names(lp) <- 1:ncol(lp)
+
+dat <- data.frame(t(diff(t(lp))))
+names(dat) <- 1:ncol(dat)
[…]

### 2023-12-05 testing.R

@@ -1,0 +1,62 @@
+library(ordinal)
+library(effects)
+library(MASS)
+
+# difference response 1-4 between two groups
+
+ps1 <- c(0.25, 0.25, 0.25, 0.25)
+b1 <- log(2)
+b0 <- probs_to_th(ps1, link = "logit")
+n <- 1e5
+
+dat <- data.frame(x = rep(c("g1", "g2"), each = n/2))
+P <- get_probs(~x, b0, b1, "logit", cnames = 1:4, data = dat)
+dat$y <- get_y(P = P)
+
+fit <- polr(y ~ x, data = dat)
+
+pr <- predict(fit, 
+              newdata = data.frame(x = unique(dat$x)), 
+              type = "prob")
+
+pr <- data.frame(pr)
+
+apply(pr, 2, function(x) filor::odds_ratio(x[1], x[2]))
+
+pg1 <- predict(fit, newdata = data.frame(x = "g1"))[[1]]
+pg2 <- predict(fit, newdata = data.frame(x = "g2"))[[1]]
+
+filor::odds_ratio(pg1, pg2)
+
+filor::odds_ratio(plogis(fit$zeta - fit$coefficients),
+                  plogis(fit$zeta))
+
+
+pr <- predict(fit, 
+              newdata = data.frame(x = unique(dat$x)), 
+              type = "linear.predictor")
+
+
+
+
+
+
+curve(dnorm(x), -5, 5)
+curve(dnorm(x, b1), -5, 5, add = TRUE)
+
+xtabs(~x + y, data = dat) |> 
+  prop.table(margin = 1) |> 
+  apply(2, function(x) filor::odds_ratio(x[1], x[2]))
+
+
+get_probs(~x, b0, 1, "probit", cnames = 1:4, data = dat) |> 
+  distinct() |> 
+  mutate(group = c("g1", "g2")) |> 
+  pivot_longer(1:4) |> 
+  #apply(2, function(x) filor::odds_ratio(x[1], x[2])) |> 
+  ggplot(aes(x = name, y = value, fill = group)) +
+  geom_col(position = position_dodge())
+
+get_probs(~x, b0, 1, "probit", cnames = 1:4, data = dat) |> 
+  distinct()
+

### 2023-12-06 paper/paper-new.Rmd
update

@@
 library(cowplot)
+library(tidyverse)
 
@@
 latent_plot <- lat |> 
-  ggplot(aes(x = x, y = m, dist = dist)) +
+  ggplot(aes(x = factor(x), y = m, dist = dist)) +
   stat_halfeye(aes(fill = after_stat(factor(findInterval(y, ths))))) +
@@
   geom_hline(yintercept = probs_to_th(probs)) +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  xlab("X")
 
-probs_plot <- get_probs(~x, 2, probs, lat) |> 
+probs_plot <- get_probs(~x, 2, probs, lat, append = TRUE) |> 
   pivot_longer(starts_with("y")) |> 
-  ggplot(aes(y = factor(x), x = value, fill = name)) +
-  geom_col(position = position_dodge())
+  ggplot(aes(x = factor(x), y = value, fill = name)) +
+  geom_col(position = position_dodge()) +
+  xlab("X")
 

### 2023-12-06 R/utils.R

@@ -105,20 +105,23 @@ show_th <- function(th = NULL, probs = NULL, link = c("logit", "probit"),
 }
 
 get_link <- function(link = c("logit", "probit")){
+  link <- match.arg(link)
   if(link == "logit"){
     pfun <- plogis
     dfun <- dlogis
     qfun <- qlogis
+    rfun <- rlogis
     dfun_n <- "dlogis(x)"
     pfun_n <- "plogis(x)"
   }else{
     pfun <- pnorm
     dfun <- dnorm
     qfun <- qnorm
+    rfun <- rnorm
     dfun_n <- "dnorm(x)"
     pfun_n <- "pnorm(x)"
   }
-  list(pfun = pfun, dfun = dfun, qfun = qfun,
+  list(rfun = rfun, pfun = pfun, dfun = dfun, qfun = qfun,
        dfun_n = dfun_n, pfun_n = pfun_n)
 }
 
@@ -129,12 +132,7 @@ sim_ord_latent <- function(formula,
                            ynames = NULL, 
                            data, 
                            link = c("logit", "probit")){
-  link <- match.arg(link)
-  if(link == "logit"){
-    rdata <- rlogis
-  }else{
-    rdata <- rnorm
-  }
+  lf <- get_link(link)
   if(is.null(th)){
     th <- probs_to_th(probs, link = link)
   }
@@ -142,7 +140,7 @@ sim_ord_latent <- function(formula,
   k <- length(probs) # number of options
   n <- nrow(data)
   X <- model.matrix(formula, data = data)[, 2] # remove intercept
-  ystar <- rdata(n, B %*% X)
+  ystar <- lf$rfun(n, B %*% X)
   y <- findInterval(ystar, th)
   if(is.null(names(probs)) & is.null(ynames)){
     ynames <- 1:k
@@ -151,22 +149,26 @@ sim_ord_latent <- function(formula,
   return(data)
 }
 
-get_probs <- function(formula, B, probs0, data, append = FALSE, link = c("logit", "probit")){
-  link <- match.arg(link)
-  ths <- probs_to_th(probs0, link)
-  X <- model.matrix(formula, data = data)
-  lp <- c(B %*% X[, -1])
-  p <- lapply(lp, function(l) {
-    ps <- diff(pnorm(c(-Inf, ths, Inf), l))
-    names(ps) <- paste0("y", 1:length(ps))
-    t(data.frame(ps))
-  })
-  p <- do.call(rbind, p)
-  if(append){
-    out <- cbind(data, p)
[…]

### 2023-12-07 paper/paper-new.Rmd
update

--- 2023-12-05 paper/paper-new.Rmd
@@
 library(ordinal)
+library(here)
 devtools::load_all()
@@
 
+```{r funs, include = FALSE}
+funs <- filor::get_funs(here("R", "utils.R"))
+```
+
 # Introduction
+
+## Risolvere un pochino la nomenclatura
+
+https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf
+
+> The name, cumulative link models is adopted from Agresti (2002), but the model class has
+been referred to by several other names in the literature, such as ordered logit models and
+ordered probit models (Greene and Hensher 2010) for the logit and probit link functions. The
+cumulative link model with a logit link is widely known as the proportional odds model due
+to McCullagh (1980) and with a complementary log-log link, the model is sometimes referred
+to as the proportional hazards model for grouped survival times.
 
@@
 
+## Odds ratios and cumulative odds ratios
+
+- https://online.stat.psu.edu/stat504/lesson/4/4.1
+
+In this section we introduce the main concept to understand the effects of a logit model.
+
 ## Demonstrating the proportional odds assumption
+
+basically given that the effect of $\beta$ is constant the odds ratio is independent from the thresholds $\alpha_j$ [@Liu2023-bp].
+In the current paper we are simulating data using a single set of $\beta$s thus the proportional odds assumption is true.
+@Liu2023-bp reviewed the available methods to test this assumption.
+
+- https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione
+
+sarebbe da capire quali sono i rischi di assumere questo ed eventualmente se simulare non proportional odds è troppo complicato.
 
@@
 c(xa = log(a_or1vs2345 / a_or12vs345), xb = log(b_or1vs2345 / b_or12vs345))
+```
+
+Proportional odds can be also visualized by plotting the cumulative probabilities of $y$, in terms of $g(P(y \leq j))$ (where $g()$ is the logit link function) as a function of the predictor $x$.
+If the proportional odds assumptions holds the slopes are parallel (also known as parallel regression assumption).
+The Figure \@ref(fig:poa-plot-example) depicts the assumption of proportional odds in the probability and logit space.
+
+```{r poa-plot-example, fig.cap="POA assumption"}
+k <- 5 # number of levels for y
+grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
+probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
+cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
+cprobs <- data.frame(do.call(rbind, cprobs))
+names(cprobs) <- sprintf("y<=%s", 1:k)
+cprobs <- cprobs[, -ncol(cprobs)] # remove the last
+grid <- cbind(grid, cprobs)
+
+grid <- grid |> 
+  pivot_longer(starts_with("y")) |> 
+  mutate(lp = qlogis(value))
+
+poa_probs_plot <- grid |> 
+  ggplot(aes(x = x, y = value, color = name)) +
+  geom_line() +
+  theme_minimal(15) +
+  theme(legend.position = "bottom",
+        legend.title = element_blank(),
+        axis.title.y = element_blank()) +
+  ggtitle("Probability")
+
+poa_lp_plot <- grid |> 
+  ggplot(aes(x = x, y = lp, color = name)) +
+  geom_line() +
+  theme_minimal(15) +
+  theme(axis.title.y = element_blank()) +
+  ggtitle("Linear Predictor")
+
+bl <- cowplot::get_legend(
+  poa_probs_plot
+)
+
+poa_plot <- cowplot::plot_grid(
+  poa_probs_plot + theme(legend.position = "none"),
+  poa_lp_plot + theme(legend.position = "none")
+)
+
+cowplot::plot_grid(poa_plot, bl, ncol = 1, rel_heights = c(0.9, 0.1))
+```
+
+@Harrell-2015-no proposed a very intuitive plot to assess the proportional odds assumption and eventually the degree of deviation from the ideal case.
+Basically predictor is plotted against the logit of the cumulative probability.
+Distances between pairs of symbols should be similar across levels of the predictors.
+Numerical predictors can be binned before plotting the corresponding logit.
+Figure \@ref(fig:harrel-poa-plot) depicts an example with simulated data satisfying the proportional odds assumption.
+
+```{r harrel-poa-plot, fig.cap="Harrel Poa Plot"}
+k <- 5 # number of levels for y
+grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
+probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
+cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
+cprobs <- data.frame(do.call(rbind, cprobs))
+names(cprobs) <- sprintf("y<=%s", 1:k)
+cprobs <- cprobs[, -ncol(cprobs)] # remove the last
+grid <- cbind(grid, cprobs)
+
+grid <- grid |> 
+  pivot_longer(starts_with("y")) |> 
+  mutate(lp = qlogis(value))
+
+grid |> 
+  ggplot(aes(x = lp, y = x)) +
+  geom_point(aes(color = name, 
+                 shape = name),
+             size = 3) +
+  theme_minimal(20) +
+  xlab("Logit") +
+  theme(legend.position = "bottom",
+        legend.title = element_blank())
 ```
@@
 
-```{r}
-library(distributional)
-library(ggdist)
-library(cowplot)
+## Scale Effects
 
-probs <- rep(1/5, 5)
-ths <- probs_to_th(probs)
+The default ordinal regression model assume that the variance of th underlying latent distribution is the same across condition.
+This is similar to a standard linear regression assuming homogeneity of variance.
+For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy].
+In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution.
+This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
+Figure \@ref(fig:example-unequal-variance) depict an example of a comparison between two groups where the two underlyning latent distributions have unequal variance.
 
-lat <- data.frame(x = c(0, 1),
-                  m = c(0, 2),
-                  s = c(1, 1))
+## Visualizing effects
 
-lat$dist <- dist_normal(lat$m, lat$s)
+Before starting with the actual simulation it is useful to plot the predicted probabilities of the response variable $y$ as a function of the predictors.
+The `cat_latent_plot()` and the `num_latent_plot()` functions are able to visualize the effect of a single categorical or numerical predictor.
+Specifying the mean and standard deviations of the latent variable in each condition (for the categorical version) or the $\beta$ (for the numerical version) the function compute the expected probability for each level of the ordinal variable $y$ and visualize the result.
+In this way the $\beta$ can be choose in a meaningful way.
 
-latent_plot <- lat |> 
-  ggplot(aes(x = x, y = m, dist = dist)) +
-  stat_halfeye(aes(fill = after_stat(factor(findInterval(y, ths))))) +
-  ylim(c(-4, 6)) +
-  geom_hline(yintercept = probs_to_th(probs)) +
-  theme(legend.position = "none")
+```{r, echo=FALSE, results='asis'}
+filor::print_fun(c(funs$cat_latent_plot, funs$num_latent_plot))
+```
 
-probs_plot <- get_probs(~x, 2, probs, lat) |> 
-  pivot_longer(starts_with("y")) |> 
-  ggplot(aes(y = factor(x), x = value, fill = name)) +
-  geom_col(position = position_dodge())
+For example, the Figure \@ref(fig:ex-categorical) depicts the effect of a categorical predictor on a 5-level $y$ assuming uniform probabilities for the reference level and a mean difference on a standardized Normal distribution of 1.
+The Figure \@ref(fig:ex-numerical) depicts the effect of a continuous predictor $x$ sampled from a standard Normal distribution assuming a $\beta = 1$.
 
-plot_grid(latent_plot, probs_plot)
+```{r ex-categorical, fig.cap="Example categorical", fig.width=8, fig.height=8}
+cat_latent_plot(m = c(0, 1), s = 1, probs = rep(1/5, 5), link = "probit")
 ```
+
+```{r ex-numerical, fig.cap="Example numerical", fig.width=8, fig.height=8}
+num_latent_plot(x = rnorm(100), b1 = 1, probs = c(0.4, 0.2, 0.2, 0.05, 0.05), link = "probit")
+```
+
+## Disclaimer about the functions
+
+The current paper proposed a simplified way with some functions to generate ordinal data.
+For more complex simulations such as simulating correlated ordinal data the `simstudy` package [https://kgoldfeld.github.io/simstudy/articles/ordinal.html](https://kgoldfeld.github.io/simstudy/articles/ordinal.html) proposed a very comprehensive set of data simulation function also for ordinal data.
 

### 2023-12-07 paper/paper-new.tex
update

@@
 ##            1         2         3         4         5
-## 1 0.18711805 0.2021825 0.2062092 0.2020689 0.2024214
-## 2 0.07132155 0.1040579 0.1540144 0.2385607 0.4320454
+## 1 0.19466136 0.1980089 0.2047539 0.2030399 0.1995360
+## 2 0.07435021 0.1025037 0.1534161 0.2411113 0.4286187
 \end{verbatim}
@@
 ##         xa         xb 
-## -1.018593 -1.018593
+## -0.9838998 -0.9838998
 \end{verbatim}
@@
 
+\subsection{Scale Effects}\label{scale-effects}
+
+The default ordinal regression model assume that the variance of th underlying latent distribution is the same across condition.
+This is similar to a standard linear regression assuming homogeneity of variance.
+For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test (see Delacre, Lakens, \& Leys, 2017).
+In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution.
+This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
+
+\subsection{Visualizing effects}\label{visualizing-effects}
+
+Before starting with the actual simulation it is useful to plot the predicted probabilities of the response variable \(y\) as a function of the predictors.
+The \texttt{cat\_latent\_plot()} and the \texttt{num\_latent\_plot()} functions are able to visualize the effect of a single categorical or numerical predictor.
+Specifying the mean and standard deviations of the latent variable in each condition (for the categorical version) or the \(\beta\) (for the numerical version) the function compute the expected probability for each level of the ordinal variable \(y\) and visualize the result.
+In this way the \(\beta\) can be choose in a meaningful way.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{cat\_latent\_plot }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(}\AttributeTok{m =} \DecValTok{0}\NormalTok{, }
+                            \AttributeTok{s =} \DecValTok{1}\NormalTok{, }
+                            \AttributeTok{th =} \ConstantTok{NULL}\NormalTok{, }
+                            \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
+                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{))\{}
+\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
+\NormalTok{  x }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"g"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(m))}
+\NormalTok{  lat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
+    \AttributeTok{x =}\NormalTok{ x,}
+    \AttributeTok{m =}\NormalTok{ m,}
+    \AttributeTok{s =}\NormalTok{ s}
+\NormalTok{  )}
+
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(th))\{}
+\NormalTok{    th }\OtherTok{\textless{}{-}} \FunctionTok{probs\_to\_th}\NormalTok{(probs, link)}
+\NormalTok{  \}}
+
+\NormalTok{  thl }\OtherTok{\textless{}{-}}\NormalTok{ latex2exp}\SpecialCharTok{::}\FunctionTok{TeX}\NormalTok{(}\FunctionTok{sprintf}\NormalTok{(}\StringTok{"$}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{alpha\_\{\%s\}$"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(th)))}
+
+  \ControlFlowTok{if}\NormalTok{(link }\SpecialCharTok{==} \StringTok{"logit"}\NormalTok{)\{}
+\NormalTok{    dfun }\OtherTok{\textless{}{-}}\NormalTok{ distributional}\SpecialCharTok{::}\NormalTok{dist\_logistic}
+\NormalTok{    s }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(}\FunctionTok{vlogit}\NormalTok{(s))}
+\NormalTok{    title }\OtherTok{\textless{}{-}} \StringTok{"Logistic Distribution"}
+\NormalTok{  \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{    dfun }\OtherTok{\textless{}{-}}\NormalTok{ distributional}\SpecialCharTok{::}\NormalTok{dist\_normal}
+\NormalTok{    s }\OtherTok{\textless{}{-}}\NormalTok{ lat}\SpecialCharTok{$}\NormalTok{s}
+\NormalTok{    title }\OtherTok{\textless{}{-}} \StringTok{"Normal Distribution"}
+\NormalTok{  \}}
+\NormalTok{  lat}\SpecialCharTok{$}\NormalTok{dist }\OtherTok{\textless{}{-}} \FunctionTok{dfun}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m, lat}\SpecialCharTok{$}\NormalTok{s)}
+\NormalTok{  range\_y }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FunctionTok{min}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m) }\SpecialCharTok{{-}} \FunctionTok{max}\NormalTok{(s) }\SpecialCharTok{*} \DecValTok{5}\NormalTok{, }\FunctionTok{max}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m) }\SpecialCharTok{+} \FunctionTok{max}\NormalTok{(s) }\SpecialCharTok{*} \DecValTok{5}\NormalTok{)}
+  \FunctionTok{ggplot}\NormalTok{(lat,}
+         \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ m, }\AttributeTok{dist =}\NormalTok{ dist)) }\SpecialCharTok{+}
+\NormalTok{    ggdist}\SpecialCharTok{::}\FunctionTok{stat\_halfeye}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{fill =} \FunctionTok{after\_stat}\NormalTok{(}\FunctionTok{factor}\NormalTok{(}\FunctionTok{findInterval}\NormalTok{(y, th) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{))),}
+                 \AttributeTok{alpha =} \FloatTok{0.85}\NormalTok{) }\SpecialCharTok{+}
+    \FunctionTok{ylim}\NormalTok{(range\_y) }\SpecialCharTok{+}
+    \FunctionTok{geom\_hline}\NormalTok{(}\AttributeTok{yintercept =}\NormalTok{ th, }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{, }\AttributeTok{col =} \StringTok{"black"}\NormalTok{, }\AttributeTok{alpha =} \FloatTok{0.7}\NormalTok{) }\SpecialCharTok{+}
+    \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ m, }\AttributeTok{group =} \DecValTok{1}\NormalTok{)) }\SpecialCharTok{+}
+    \FunctionTok{annotate}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\AttributeTok{x =} \FloatTok{0.7}\NormalTok{, }\AttributeTok{y =}\NormalTok{ th, }\AttributeTok{label =}\NormalTok{ thl, }\AttributeTok{size =} \DecValTok{5}\NormalTok{) }\SpecialCharTok{+}
+    \FunctionTok{theme\_minimal}\NormalTok{(}\DecValTok{15}\NormalTok{) }\SpecialCharTok{+}
+    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{,}
+          \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{(),}
+          \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
+    \FunctionTok{ylab}\NormalTok{(latex2exp}\SpecialCharTok{::}\FunctionTok{TeX}\NormalTok{(}\StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{mu"}\NormalTok{)) }\SpecialCharTok{+}
+    \FunctionTok{ggtitle}\NormalTok{(title)}
+\NormalTok{\}}
+\NormalTok{num\_latent\_plot }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, }
+\NormalTok{                            b1, }
+                            \AttributeTok{th =} \ConstantTok{NULL}\NormalTok{, }
+                            \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{,}
+                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{), }
+                            \AttributeTok{nsample =} \FloatTok{1e3}\NormalTok{, }
+                            \AttributeTok{size =} \DecValTok{20}\NormalTok{)\{}
+\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(th))\{}
+\NormalTok{    th }\OtherTok{\textless{}{-}} \FunctionTok{probs\_to\_th}\NormalTok{(probs, link)}
+\NormalTok{  \}}
+\NormalTok{  data }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{seq}\NormalTok{(}\FunctionTok{min}\NormalTok{(x), }\FunctionTok{max}\NormalTok{(x), }\AttributeTok{length.out =}\NormalTok{ nsample))}
+\NormalTok{  data }\OtherTok{\textless{}{-}} \FunctionTok{get\_probs}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, b1, probs, data, link, }\AttributeTok{append =} \ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{  datal }\OtherTok{\textless{}{-}}\NormalTok{ tidyr}\SpecialCharTok{::}\FunctionTok{pivot\_longer}\NormalTok{(data, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}
+  \FunctionTok{ggplot}\NormalTok{(datal, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x, }\AttributeTok{y =}\NormalTok{ value, }\AttributeTok{color =}\NormalTok{ y)) }\SpecialCharTok{+}
+    \FunctionTok{geom\_line}\NormalTok{() }\SpecialCharTok{+}
+    \FunctionTok{theme\_minimal}\NormalTok{(size) }\SpecialCharTok{+}
+    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{,}
+          \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
+    \FunctionTok{ylab}\NormalTok{(}\StringTok{"Probability"}\NormalTok{) }\SpecialCharTok{+}
+    \FunctionTok{ggtitle}\NormalTok{(latex2exp}\SpecialCharTok{::}\FunctionTok{TeX}\NormalTok{(}\FunctionTok{sprintf}\NormalTok{(}\StringTok{"$}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{beta\_1 = \%s$"}\NormalTok{, b1)))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+For example, the Figure \ref{fig:ex-categorical} depicts the effect of a categorical predictor on a 5-level \(y\) assuming uniform probabilities for the reference level and a mean difference on a standardized Normal distribution of 1.
+The Figure \ref{fig:ex-numerical} depicts the effect of a continuous predictor \(x\) sampled from a standard Normal distribution assuming a \(\beta = 1\).
+
 \begin{figure}
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-3-1}
+{\centering \includegraphics{paper-new_files/figure-latex/ex-categorical-1}
 
@@
 
-\caption{ }\label{fig:unnamed-chunk-3}
+\caption{Example categorical}\label{fig:ex-categorical}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/ex-numerical-1}
+
+}
+
+\caption{Example numerical}\label{fig:ex-numerical}
 \end{figure}
@@
 DeCarlo, L. T. (1998). Signal detection theory and generalized linear models. \emph{Psychological Methods}, \emph{3}, 186--205. \url{https://doi.org/10.1037/1082-989X.3.2.186}
+
+\bibitem[\citeproctext]{ref-Delacre2017-qy}
+Delacre, M., Lakens, D., \& Leys, C.
+(2017).
+Why psychologists should by default use welch's \emph{t}-test instead of student's \emph{t}-test.
+\emph{International Review of Social Psychology}, \emph{30}, 92.
+\url{https://doi.org/10.5334/irsp.82}
 

### 2023-12-07 R/utils.R

@@ -125,27 +125,32 @@ get_link <- function(link = c("logit", "probit")){
        dfun_n = dfun_n, pfun_n = pfun_n)
 }
 
-sim_ord_latent <- function(formula, 
-                           B, 
+sim_ord_latent <- function(formula,
+                           sigma = NULL,
+                           By,
+                           Bsigma = NULL,
                            probs = NULL, 
                            th = NULL, 
-                           ynames = NULL, 
                            data, 
                            link = c("logit", "probit")){
   lf <- get_link(link)
   if(is.null(th)){
     th <- probs_to_th(probs, link = link)
   }
-  probs <- th_to_probs(th)
+  probs <- th_to_probs(th, link = link)
   k <- length(probs) # number of options
   n <- nrow(data)
-  X <- model.matrix(formula, data = data)[, 2] # remove intercept
-  ystar <- lf$rfun(n, B %*% X)
-  y <- findInterval(ystar, th)
-  if(is.null(names(probs)) & is.null(ynames)){
-    ynames <- 1:k
+  Xy <- model.matrix(formula, data = data)[, -1, drop = FALSE] # remove intercept
+  lpy <- c(Xy %*% By)
+  if(is.null(sigma)){
+    ystar <- lf$rfun(n, lpy)
+  }else{
+    Xsigma <- model.matrix(sigma, data = data)[, -1, drop = FALSE] # remove intercept
+    ystar <- lf$rfun(n, lpy, exp(c(Xsigma %*% Bsigma)))
   }
-  data$y <- factor(y, labels = ynames)
+  y <- findInterval(ystar, th) + 1 # to start from 1
+  data$y <- factor(y)
+  data$y <- ordered(data$y)
   return(data)
 }
 
@@ -163,7 +168,9 @@ get_probs <- function(formula,
   ths <- c(-Inf, probs_to_th(probs0, link = link), Inf)
   X <- model.matrix(formula, data = data)[, -1]
   X <- matrix(X)
-  P <- lapply(ths, function(t) c(plogis(t - X %*% B)))
+  # notice the minus sign t - X %*% B thus an increase in x
+  # is an increase in p
+  P <- lapply(ths, function(t) c(lf$pfun(t - X %*% B)))
   P <- data.frame(P)
   P <- data.frame(t(apply(P, 1, diff)))
   names(P) <- ynames
@@ -171,4 +178,118 @@ get_probs <- function(formula,
[…]

### 2023-12-07 power.R

@@ -1,0 +1,58 @@
+n <- 100
+dat <- data.frame(
+  g = rep(c(0, 1), each = n/2)
+)
+dat$g <- factor(dat$g)
+contrasts(dat$g) <- contr.sum(2)/2
+
+probs <- rep(1/6, 6)
+th <- probs_to_th(probs, "probit")
+
+nsim <- 1000
+p_clm <- rep(NA, nsim)
+p_lm <- rep(NA, nsim)
+b_clm <- rep(NA, nsim)
+b_lm <- rep(NA, nsim)
+
+for(i in 1:nsim){
+  dat <- sim_ord_latent(~g, ~g, B.y = 0.5, B.sigma = log(5), probs = probs, data = dat, link = "probit")
+  dat$yn <- as.integer(dat$y)
+  fit_clm <- clm(y ~ g, ~ g, data = dat, link = "probit")
+  fit_lm <- lm(yn ~ g, data = dat)
+  p_clm[i] <- summary(fit_clm)$coefficients[length(th) + 1, 4]
+  p_lm[i] <- summary(fit_lm)$coefficients[2, 4]
+  b_clm[i] <- summary(fit_clm)$coefficients[length(th) + 1, 1]
+  b_lm[i] <- summary(fit_lm)$coefficients[2, 1]/sigma(fit_lm)
+}
+
+plot(b_clm, b_lm, xlim = c(-1, 2), ylim = c(-0.5, 1.5))
+
+
+mean(p_lm <= 0.05)
+mean(p_clm <= 0.05, na.rm = TRUE)
+
+
+n <- 1e3
+dat <- data.frame(
+  x = rnorm(n)
+)
+
+dat <- sim_ord_latent(~x, ~x, B.y = 0.5, B.sigma = log(1.5),probs = probs, data = dat, link = "logit")
+dat$yn <- as.integer(dat$y)
+
+fit2 <- clm(y ~ x, ~ x, data = dat, link = "logit")
+
+summary(fit2)
+
+library(effects)
+
+summary(fit2)
+
+
+
+
+
+
+
+
+

### 2023-12-07 zzz.R

@@ -1,0 +1,100 @@
+# quick example with a random effect
+
+library(ordinal)
+
+n <- 1e5
+dat <- data.frame(
+  x = seq(0, 1, length.out = n)
+)
+
+dat <- sim_ord_latent(~x, By = 8, probs = rep(1/5, 5), data = dat, link = "logit")
+fit <- clm(y ~ 1, data = dat, link = "logit")
+pr <- predict(fit, type = "linear.predictor")
+dat <- cbind(dat, pr)
+
+dat$yn <- as.integer(dat$y)
+
+cum_prob <- function(y){
+  if(!is.numeric(y)){
+    y <- as.integer(y)
+  }
+  sy <- sort(unique(y))
+  cy <- sapply(sy, function(t) mean(y <= t))
+  cy <- cy[-length(cy)]
+  names(cy) <- sprintf("y<=%s", sy[-length(sy)])
+  return(cy)
+}
+
+dat$xc <- cut(dat$x, seq(0, 1, 0.1), include.lowest = TRUE)
+
+dd <- lapply(split(dat$y, dat$xc), function(x) qlogis(cum_prob(x))) |> 
+  data.frame() |> 
+  t() |> 
+  data.frame()
+
+names(dd) <- sprintf("y<=%s", 1:4)
+dd$x <- 1:nlevels(dat$xc)
+dd |> 
+  pivot_longer(starts_with("y")) |> 
+  ggplot(aes(x = x, y = value, color = name)) +
+  geom_line()
+
+dat <- data.frame(x = seq(0, 1, 0.1))
+dat <- get_probs(~x, 5, rep(1/4, 4), data = dat, link = "logit", append = TRUE)
+
+pp <- apply(dat[, -1], 1, cumsum, simplify = FALSE) |> 
+  bind_rows()
+pp <- pp[, -ncol(pp)]
+names(pp) <- sprintf("y<=%s", 1:3)
+
+cbind(dat, pp) |> 
+  select(x, 6:8) |> 
+  pivot_longer(2:4) |> 
+  ggplot(aes(x = x, y = qlogis(value), color = name)) +
+  geom_line()
+
+
+k <- 5 # number of levels for y
+grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
+probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
+cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
+cprobs <- data.frame(do.call(rbind, cprobs))
[…]

### 2023-12-10 paper/paper-new.Rmd
update

@@
 
+## Metriv vs ordinal models
+
+- the main difference is that the metric model assign a number to each label of the discrete variable assuming that the distance is the same. [@Liddell2018-wu]
+
+## lm on latent vs ordinal
+
+as suggested https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf, running a lm on the latent variables gives similar parameter as the clm.
+Of course, we are able to do this with real data given that the ordinal variable is the observed version of an unobserved latent variable.
+But in simulation this is useful to understand what the cumulative model is doing.
+
+```{r}
+dat <- data.frame(
+  x1 = rnorm(1e5),
+  x2 = rnorm(1e5)
+)
+
+dat <- sim_ord_latent(~x1 + x2, By = c(1, 0.5), probs = rep(1/5, 5), link = "probit", data = dat)
+
+fit <- lm(ys ~ x1 + x2, data = dat)
+summary(fit)
+
+clm(y ~ x1 + x2, data = dat, link = "probit")
+coef(fit)/sigma(fit)
+
+```
+
+## Kruscke parametrization
+
+@Liddell2018-wu and @Kruschke2015-re proposed an alternative parametrization to understand the model parameters.
+They used a probit model where thresholds and regression parameters are estimated on the scale of the ordinal variable compared to standard ordinal regression where they refers to the quantile of the latent variables for the threshold and z score or odds ratio for the regression coefficients.
+They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.
+
 ## Odds ratios and cumulative odds ratios
@@
 probs <- rep(1/5, 5) # for the group "a", uniform probabilities
-dat <- sim_ord_latent(~x, b1, probs = probs, data = dat, link = "logit")
+dat <- sim_ord_latent(~x, By = b1, probs = probs, data = dat, link = "logit")
 fit <- clm(y ~ x, data = dat, link = "logit")
@@
 
+From a statistical point of view, the proportional odds assumption can be assessed by fitting $k - 1$ binomial regressions and checking if the estimated $\beta$ is similar between regressions.
+The regressions are estimated by creating $k - 1$ dummy variables from the ordinal $y$.
+The code below show that simulating data with the proportional odds create $k - 1$ binomial regression with similar $\beta$s.
+
+```{r}
+#| echo: true
+k <- 4
+n <- 1e5
+dat <- data.frame(
+  x = runif(n)
+)
+
+dat <- sim_ord_latent(~x, By = 5, probs = rep(1/k, k), data = dat, link = "logit")
+
+# create dummy variables
+
+dat$y1vs234 <- ifelse(dat$y <= 1, 1, 0)
+dat$y12vs34 <- ifelse(dat$y <= 2, 1, 0)
+dat$y123vs4 <- ifelse(dat$y <= 3, 1, 0)
+
+fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "logit"))
+fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "logit"))
+fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "logit"))
+
+car::compareCoefs(fit123vs4, fit12vs34, fit1vs234)
+```
+
 ## Logit vs Probit model
@@
 In practical terms, the thresholds are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].
+
+The latent formulation of the ordinal regression model allow to think in standard linear regression terms for choosing and interpreting model parameters before converting into the probability space.
+Both the standard Normal and the logistic distribution are defined with a location ($\mu$) and a scale ($s$) parameter.
+For the Normal distribution, $\mu$ is the mean and scale is the variance ($\sigma^2$).
+For the logistic distribution the variance is $\sigma^2=\frac{s^2\pi^2}{3}$.
+Thus fixing $\mu$ and $\sigma^2 = 1$ (the default), the two distributions are similar with the logistic having more variance.
+For this reason, the latent formulation for parameter intepretation is particularly useful for the probit model because the $\beta$ is directly interpreted in standard deviaton units that by default are fixed to 1.
+With a binary predictor $x$, $\beta$ is the shift of the latent distribution increasing $x$ by one unit.
+This can be directly interpreted as a standardized mean difference effect size (e.g., Cohen's $d$).
+This is the same for the logistic distribution but a unit increase in $x$ shift the latent distribution by $\sigma = \sqrt{\frac{s^2 \pi^2}{3}} = \frac{s\pi}{\sqrt{3}}$.^[In the case of a standard logistic distribution ($s^2 = 1$), the standard deviation is $\frac{\pi}{\sqrt{3}}$].
 

### 2023-12-10 paper/paper-new.tex
update

@@
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break#1\hfill\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
 \newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
@@
 
+\subsection{Risolvere un pochino la nomenclatura}\label{risolvere-un-pochino-la-nomenclatura}
+
+\url{https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf}
+
+\begin{quote}
+The name, cumulative link models is adopted from Agresti (2002), but the model class has
+been referred to by several other names in the literature, such as ordered logit models and
+ordered probit models (Greene and Hensher 2010) for the logit and probit link functions. The
+cumulative link model with a logit link is widely known as the proportional odds model due
+to McCullagh (1980) and with a complementary log-log link, the model is sometimes referred
+to as the proportional hazards model for grouped survival times.
+\end{quote}
+
 Da citare in qualche modo Kemp and Grace (2021)
@@
 
+\subsection{Metriv vs ordinal models}\label{metriv-vs-ordinal-models}
+
+\begin{itemize}
+\tightlist
+\item
+  the main difference is that the metric model assign a number to each label of the discrete variable assuming that the distance is the same. (Liddell \& Kruschke, 2018)
+\end{itemize}
+
+\subsection{lm on latent vs ordinal}\label{lm-on-latent-vs-ordinal}
+
+as suggested \url{https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf}, running a lm on the latent variables gives similar parameter as the clm.
+Of course, we are able to do this with real data given that the ordinal variable is the observed version of an unobserved latent variable.
+But in simulation this is useful to understand what the cumulative model is doing.
+
+\begin{verbatim}
+## 
+## Call:
+## lm(formula = ys ~ x1 + x2, data = dat)
+## 
+## Residuals:
+##     Min      1Q  Median      3Q     Max 
+## -4.0126 -0.6777 -0.0003  0.6754  4.1015 
+## 
+## Coefficients:
+##              Estimate Std. Error t value Pr(>|t|)    
+## (Intercept) -0.003164   0.003161  -1.001    0.317    
+## x1           1.005849   0.003146 319.749   <2e-16 ***
+## x2           0.498501   0.003156 157.971   <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Residual standard error: 0.9996 on 99997 degrees of freedom
+## Multiple R-squared:  0.5589, Adjusted R-squared:  0.5589 
+## F-statistic: 6.335e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
+\end{verbatim}
+
+\begin{verbatim}
+## formula: y ~ x1 + x2
+## data:    dat
+## 
+##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
+##  probit flexible  1e+05 -122018.81 244049.62 5(0)  9.59e-08 9.1e+00
+## 
+## Coefficients:
+##     x1     x2 
+## 1.0064 0.4993 
+## 
+## Threshold coefficients:
+##     1|2     2|3     3|4     4|5 
+## -0.8347 -0.2504  0.2529  0.8444
+\end{verbatim}
+
+\begin{verbatim}
+##  (Intercept)           x1           x2 
+## -0.003165429  1.006273800  0.498711694
+\end{verbatim}
+
+\subsection{Kruscke parametrization}\label{kruscke-parametrization}
+
+Liddell and Kruschke (2018) and Kruschke (2015) proposed an alternative parametrization to understand the model parameters.
+They used a probit model where thresholds and regression parameters are estimated on the scale of the ordinal variable compared to standard ordinal regression where they refers to the quantile of the latent variables for the threshold and z score or odds ratio for the regression coefficients.
+They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.
+
+\subsection{Odds ratios and cumulative odds ratios}\label{odds-ratios-and-cumulative-odds-ratios}
+
+\begin{itemize}
+\tightlist
+\item
+  \url{https://online.stat.psu.edu/stat504/lesson/4/4.1}
+\end{itemize}
+
+In this section we introduce the main concept to understand the effects of a logit model.
+
 \subsection{Demonstrating the proportional odds assumption}\label{demonstrating-the-proportional-odds-assumption}
+
+basically given that the effect of \(\beta\) is constant the odds ratio is independent from the thresholds \(\alpha_j\) (Liu, He, Tu, \& Tang, 2023).
+In the current paper we are simulating data using a single set of \(\beta\)s thus the proportional odds assumption is true.
+Liu et al.
+(2023) reviewed the available methods to test this assumption.
+
+\begin{itemize}
+\tightlist
+\item
+  \url{https://hbiostat.org/ordinal/impactpo.pdf} blog su proportional odds, mi pare di capire che non sia così problematica come assunzione
+\end{itemize}
+
+sarebbe da capire quali sono i rischi di assumere questo ed eventualmente se simulare non proportional odds è troppo complicato.
 
@@
 \NormalTok{probs }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{) }\CommentTok{\# for the group "a", uniform probabilities}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, b1, }\AttributeTok{probs =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =}\NormalTok{ b1, }\AttributeTok{probs =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
 \NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
@@
 ##            1         2         3         4         5
-## 1 0.19466136 0.1980089 0.2047539 0.2030399 0.1995360
-## 2 0.07435021 0.1025037 0.1534161 0.2411113 0.4286187
+## 1 0.19720971 0.2039506 0.2009355 0.1973976 0.2005066
+## 2 0.07341871 0.1042636 0.1503074 0.2345890 0.4374213
 \end{verbatim}
@@
 ##        xa        xb 
-## -0.9838998 -0.9838998
+## -1.003193 -1.003193
+\end{verbatim}
+
+Proportional odds can be also visualized by plotting the cumulative probabilities of \(y\), in terms of \(g(P(y \leq j))\) (where \(g()\) is the logit link function) as a function of the predictor \(x\).
+If the proportional odds assumptions holds the slopes are parallel (also known as parallel regression assumption).
+The Figure \ref{fig:poa-plot-example} depicts the assumption of proportional odds in the probability and logit space.
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/poa-plot-example-1}
+
+}
+
+\caption{POA assumption}\label{fig:poa-plot-example}
+\end{figure}
+
+Harrell (2015) proposed a very intuitive plot to assess the proportional odds assumption and eventually the degree of deviation from the ideal case.
+Basically predictor is plotted against the logit of the cumulative probability.
+Distances between pairs of symbols should be similar across levels of the predictors.
+Numerical predictors can be binned before plotting the corresponding logit.
+Figure \ref{fig:harrel-poa-plot} depicts an example with simulated data satisfying the proportional odds assumption.
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/harrel-poa-plot-1}
+
+}
+
+\caption{Harrel Poa Plot}\label{fig:harrel-poa-plot}
+\end{figure}
+
+From a statistical point of view, the proportional odds assumption can be assessed by fitting \(k - 1\) binomial regressions and checking if the estimated \(\beta\) is similar between regressions.
+The regressions are estimated by creating \(k - 1\) dummy variables from the ordinal \(y\).
+The code below show that simulating data with the proportional odds create \(k - 1\) binomial regression with similar \(\beta\)s.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{4}
+\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e5}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
+  \AttributeTok{x =} \FunctionTok{runif}\NormalTok{(n)}
+\NormalTok{)}
+
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \DecValTok{5}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k), }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+
+\CommentTok{\# create dummy variables}
+
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y1vs234 }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y }\SpecialCharTok{\textless{}=} \DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y12vs34 }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y }\SpecialCharTok{\textless{}=} \DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y123vs4 }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y }\SpecialCharTok{\textless{}=} \DecValTok{3}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
+
+\NormalTok{fit1vs234 }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(y1vs234 }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link =} \StringTok{"logit"}\NormalTok{))}
+\NormalTok{fit12vs34 }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(y12vs34 }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link =} \StringTok{"logit"}\NormalTok{))}
+\NormalTok{fit123vs4 }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(y123vs4 }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link =} \StringTok{"logit"}\NormalTok{))}
+
+\NormalTok{car}\SpecialCharTok{::}\FunctionTok{compareCoefs}\NormalTok{(fit123vs4, fit12vs34, fit1vs234)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Calls:
+## 1: glm(formula = y123vs4 ~ x, family = binomial(link = "logit"), data = dat)
+## 2: glm(formula = y12vs34 ~ x, family = binomial(link = "logit"), data = dat)
+## 3: glm(formula = y1vs234 ~ x, family = binomial(link = "logit"), data = dat)
+## 
+##             Model 1 Model 2 Model 3
+## (Intercept)  1.1117  0.0148 -1.1014
+## SE           0.0151  0.0163  0.0216
+##                                    
+## x           -5.0263 -5.0329 -4.9763
+## SE           0.0365  0.0486  0.0732
+## 
 \end{verbatim}
@@
 
+The latent formulation of the ordinal regression model allow to think in standard linear regression terms for choosing and interpreting model parameters before converting into the probability space.
+Both the standard Normal and the logistic distribution are defined with a location (\(\mu\)) and a scale (\(s\)) parameter.
+For the Normal distribution, \(\mu\) is the mean and scale is the variance (\(\sigma^2\)).
+For the logistic distribution the variance is \(\sigma^2=\frac{s^2\pi^2}{3}\).
+Thus fixing \(\mu\) and \(\sigma^2 = 1\) (the default), the two distributions are similar with the logistic having more variance.
+For this reason, the latent formulation for parameter intepretation is particularly useful for the probit model because the \(\beta\) is directly interpreted in standard deviaton units that by default are fixed to 1.
+With a binary predictor \(x\), \(\beta\) is the shift of the latent distribution increasing \(x\) by one unit.
+This can be directly interpreted as a standardized mean difference effect size (e.g., Cohen's \(d\)).
+This is the same for the logistic distribution but a unit increase in \(x\) shift the latent distribution by \(\sigma = \sqrt{\frac{s^2 \pi^2}{3}} = \frac{s\pi}{\sqrt{3}}\).\footnote{In the case of a standard logistic distribution (\(s^2 = 1\)), the standard deviation is \(\frac{\pi}{\sqrt{3}}\)}.
+
 \subsection{Simulating data}\label{simulating-data}
@@
 This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
+Figure \ref{fig:example-unequal-variance} depict an example of a comparison between two groups where the two underlyning latent distributions have unequal variance.
 
@@
                             \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
-                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{))\{}
+                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{),}
+                            \AttributeTok{plot =} \FunctionTok{c}\NormalTok{(}\StringTok{"probs"}\NormalTok{, }\StringTok{"latent"}\NormalTok{, }\StringTok{"both"}\NormalTok{))\{}
+\NormalTok{  plot }\OtherTok{\textless{}{-}} \FunctionTok{match.arg}\NormalTok{(plot)}
 \NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
@@
 \NormalTok{  range\_y }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FunctionTok{min}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m) }\SpecialCharTok{{-}} \FunctionTok{max}\NormalTok{(s) }\SpecialCharTok{*} \DecValTok{5}\NormalTok{, }\FunctionTok{max}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m) }\SpecialCharTok{+} \FunctionTok{max}\NormalTok{(s) }\SpecialCharTok{*} \DecValTok{5}\NormalTok{)}
-  \FunctionTok{ggplot}\NormalTok{(lat,}
+
+\NormalTok{  lat\_plot }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(lat,}
          \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ m, }\AttributeTok{dist =}\NormalTok{ dist)) }\SpecialCharTok{+}
@@
     \FunctionTok{ggtitle}\NormalTok{(title)}
+
+\NormalTok{  th }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\SpecialCharTok{{-}}\ConstantTok{Inf}\NormalTok{, th, }\ConstantTok{Inf}\NormalTok{)}
+\NormalTok{  lat\_ms }\OtherTok{\textless{}{-}}\NormalTok{ lat[, }\FunctionTok{c}\NormalTok{(}\StringTok{"m"}\NormalTok{, }\StringTok{"s"}\NormalTok{)]}
+\NormalTok{  ps }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(lat\_ms, }\DecValTok{1}\NormalTok{, }\ControlFlowTok{function}\NormalTok{(x) }\FunctionTok{data.frame}\NormalTok{(}\FunctionTok{t}\NormalTok{(}\FunctionTok{diff}\NormalTok{(lf}\SpecialCharTok{$}\FunctionTok{pfun}\NormalTok{(th, x[}\DecValTok{1}\NormalTok{], x[}\DecValTok{2}\NormalTok{])))), }\AttributeTok{simplify =} \ConstantTok{FALSE}\NormalTok{)}
+\NormalTok{  ps }\OtherTok{\textless{}{-}} \FunctionTok{do.call}\NormalTok{(rbind, ps)}
+  \FunctionTok{names}\NormalTok{(ps) }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"y"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{ncol}\NormalTok{(ps))}
+\NormalTok{  latl }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(lat, ps)}
+\NormalTok{  latl }\OtherTok{\textless{}{-}} \FunctionTok{pivot\_longer}\NormalTok{(latl, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}
+\NormalTok{  probs\_plot }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(latl, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ value, }\AttributeTok{fill =}\NormalTok{ y)) }\SpecialCharTok{+}
+    \FunctionTok{geom\_col}\NormalTok{(}\AttributeTok{position =} \FunctionTok{position\_dodge}\NormalTok{()) }\SpecialCharTok{+}
+    \FunctionTok{theme\_minimal}\NormalTok{(}\DecValTok{15}\NormalTok{) }\SpecialCharTok{+}
+    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{,}
+          \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{(),}
+          \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
+    \FunctionTok{ylab}\NormalTok{(}\StringTok{"Probability"}\NormalTok{)}
+  \ControlFlowTok{if}\NormalTok{(plot }\SpecialCharTok{==} \StringTok{"probs"}\NormalTok{)\{}
+\NormalTok{    probs\_plot}
+\NormalTok{  \}}\ControlFlowTok{else} \ControlFlowTok{if}\NormalTok{(plot }\SpecialCharTok{==} \StringTok{"latent"}\NormalTok{)\{}
+\NormalTok{    lat\_plot}
+\NormalTok{  \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{    legend\_b }\OtherTok{\textless{}{-}} \FunctionTok{get\_legend}\NormalTok{(}
+\NormalTok{      probs\_plot}
+\NormalTok{    )}
+\NormalTok{    plts }\OtherTok{\textless{}{-}} \FunctionTok{plot\_grid}\NormalTok{(}
+\NormalTok{      lat\_plot }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{),}
+\NormalTok{      probs\_plot }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{)}
+\NormalTok{    )}
+    \FunctionTok{plot\_grid}\NormalTok{(plts, legend\_b, }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{, }\AttributeTok{rel\_heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{))}
+\NormalTok{  \}}
 \NormalTok{\}}
@@
 
+\subsection{Disclaimer about the functions}\label{disclaimer-about-the-functions}
+
+The current paper proposed a simplified way with some functions to generate ordinal data.
+For more complex simulations such as simulating correlated ordinal data the \texttt{simstudy} package \url{https://kgoldfeld.github.io/simstudy/articles/ordinal.html} proposed a very comprehensive set of data simulation function also for ordinal data.
+
 \newpage
@@
 
+\bibitem[\citeproctext]{ref-Harrell-2015-no}
+Harrell, F. E. (2015). \emph{Regression modeling strategies}. Cham: Springer International Publishing. \url{https://doi.org/10.1007/978-3-319-19425-7}
+
 \bibitem[\citeproctext]{ref-Kemp2021-dj}
@@
 Knoblauch, K., \& Maloney, L. T. (2012). \emph{Modeling psychophysical data in r}. New York, NY: Springer New York. \url{https://doi.org/10.1007/978-1-4614-4475-6}
+
+\bibitem[\citeproctext]{ref-Kruschke2015-re}
+Kruschke, J. K. (2015). \emph{Doing bayesian data analysis}. \url{https://doi.org/10.1016/c2012-0-00477-2}
 
@@
 \url{https://doi.org/10.1016/j.jesp.2018.08.009}
+
+\bibitem[\citeproctext]{ref-Liu2023-bp}
+Liu, A., He, H., Tu, X.
+M., \& Tang, W.
+(2023).
+On testing proportional odds assumptions for proportional odds models.
+\emph{General Psychiatry}, \emph{36}, e101048.
+\url{https://doi.org/10.1136/gpsych-2023-101048}
 

### 2023-12-10 R/utils.R

@@ -18,8 +18,8 @@ probs_to_th <- function(probs, link = c("logit", "probit")){
   }else{
     lfun <- qnorm
   }
-  th <- lfun(Reduce(sum, probs, accumulate = TRUE))
-  th[-length(th)]
+  cprobs <- cumsum(probs)
+  lfun(cprobs[-length(cprobs)])
 }
 
 # get_probs <- function(formula, 
@@ -151,6 +151,7 @@ sim_ord_latent <- function(formula,
   y <- findInterval(ystar, th) + 1 # to start from 1
   data$y <- factor(y)
   data$y <- ordered(data$y)
+  data$ys <- ystar
   return(data)
 }
 
@@ -292,4 +293,14 @@ codds <- function(p){
   nn <- Reduce(paste0, 1:(length(p) - 1), accumulate = TRUE)
   names(co) <- paste0("o", nn)
   co
+}
+
+dummy_ord <- function(y){
+  if(!is.numeric(y)) y <- as.integer(y)
+  yc <- sort(unique(y))
+  dummy <- lapply(yc, function(t) ifelse(y <= t, 1, 0))
+  nn1 <- Reduce(paste0, yc, accumulate = TRUE)
+  nn2 <- Reduce(paste0, yc, accumulate = TRUE, right = TRUE)
+  names(dummy) <- sprintf("y%svs%s", nn1[-length(nn1)], nn2[-1])
+  data.frame(dummy[-length(dummy)])
 }

### 2023-12-10 reviewq.R

@@ -1,0 +1,130 @@
+tab <- with(wine, table(contact, rating))
+dat <- data.frame(freq =c(tab),
+                  contact=rep(c("no", "yes"), 5),
+                  rating = factor(rep(1:5, each=2), ordered=TRUE))
+thresholds <- 1:4
+cum.rate <- as.vector(sapply(thresholds, function(x) dat$rating <= x))
+rating.factor <- gl(n=length(thresholds), k=nrow(dat),
+                    length=nrow(dat) * length(thresholds))
+thres.X <- model.matrix(~ rating.factor - 1)
+colnames(thres.X) <- paste("t", thresholds, sep="")
+old.X <- -model.matrix(~contact, dat)[, -1, drop=FALSE]
+new.X <- kronecker(matrix(rep(1, length(thresholds)), nc = 1), old.X)
+weights <- kronecker(matrix(rep(1, length(thresholds)), nc = 1), dat$freq)
+new.X <- cbind(thres.X, new.X)
+colnames(new.X)[-seq(length(thresholds))] <- colnames(old.X)
+p.df <- cbind(cum.rate = 1*cum.rate, as.data.frame(new.X), weights)
+p.df
+
+# John K. Kruschke, 2014-Nov-21.
+# This function is described at the blog post
+# http://doingbayesiandataanalysis.blogspot.com/2014/11/ordinal-probit-regression-transforming.html
+# A PDF copy of the blog post is available at
+# https://osf.io/fc6zd/
+# This function comes with no warranties, guarantees, etc. Use at your own risk.
+polrToOrdScale = function( polrObject ) {
+  polrThresh = polrObject$zeta
+  polrSlopes = polrObject$coefficients
+  polrInter = 0.0
+  polrSigma = 1.0
+  K = length(polrThresh) + 1  # K is number of ordinal levels
+  sigmaMult = unname( (K-2)/(polrThresh[K-1]-polrThresh[1]) )
+  inter = unname( 1.5 - ( sigmaMult*polrThresh[1] ) )
+  respThresh = sigmaMult*polrThresh + inter
+  respSigma = sigmaMult*polrSigma
+  respB0 = sigmaMult*polrInter + inter
+  respSlopes = sigmaMult*polrSlopes
+  return( list( sigma=respSigma ,
[…]

### 2023-12-11 paper/paper-new.Rmd
update

@@
 # Introduction
+
+## notation
+
+- idea di mettere $F$ o qualcosa di simile per la pdf ($\Phi$ e.g., per la normale)
+- $Y^\star$ per la latente
+- $k$ numero di livelli ordinali
+- $\alpha$ per le soglie
+
+per notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw
+
+per la parte di notazione e quella più formale anche @DeCarlo2010-lj
+
+## cumulative vs others
+
+- cumulative quando una variabile continua latente viene tagliata in modo osservato. Le likert possono essere gestite in questo modo
+
+from @Burkner2019-aw:
+
+> For  many  ordinal  variables,  the  assumption  of  a   single underlying continuous variable, as in cumulative  models, may not be appropriate.
+If the response can  be understood as being the result of a sequential pro- cess, such that a higher response category is possible  only after all lower categories are achieved, the sequen- tial model proposed by Tutz (1990) is usually appropri- ate.
+
+vedi se fare k-1 logistiche è più o meno la stessa cosa
+
+in realtà credo che l'adiaject category e sequential model comunque abbiamo la proportional odds assumption
+
+```{r}
+#| eval: false
+library(brms)
+library(ordinal)
+
+dat <- data.frame(x = rnorm(100))
+dat <- sim_ord_latent(~x, By = log(2), probs = c(0.5, 0.2, 0.2, 0.1), data = dat, link = "logit")
+
+fit_clm <- clm(y ~ x, data = dat, link = "logit")
+fit_cum <- brm(y ~ x, data = dat, family = cumulative(link = "logit"))
+fit_seq <- brm(y ~ x, data = dat, family = sratio(link = "logit"))
+fit_adj <- brm(y ~ x, data = dat, family = acat(link = "logit"))
+
+```
+
+come sottolinea @Burkner2019-aw e @Tutz2020-xq i sequential model sono adatti per un certo tipo di dati dove l'interesse è vedere l'impatto dei predittori sulla transizione ad un categoria $r$ considerando che la categoria $r - 1$ è stata raggiunta.
+
+@Burkner2019-aw suggest that while cumulative and sequentials models have a clear empirical distinction the adjiacent category model is usually used for computational advantages.
+The probability of passing from one category to the other is modeled as a function of predictors.
 
@@
 They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.
+
+The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation).
+Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion.
+The main improvement regards mapping the values (for thresholds $\alpha_i$ and slopes $\beta$) from latent standard distribution (gaussian or logistic) into the scale of the $y$ ordinal value.
+The scale of the variable depend on the numeric labels assigned to ordered categories.
+There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to $\alpha_1 + 0.5$ and $\alpha_k - 0.5$ where other thresholds are estimated.
+
+@Kurz2023-zv explain how to convert a model fitted with the `brms` package (citation) (an R package for regression modeling using *stan*) into the Kruschke [-@Kruschke2015-re] parametrization.
+
+Using the function proposed by @Kruschke2015-re (metti ref ad osf) and the equations presented in @Kurz2023-zv the `clm_to_ord()` function convert parameters fitted with the `clm` function into the corresponding parameters for the latent $Y^\star$ variable considering the actual range of values (from 1 to $k$).
+
+```{r, echo=FALSE, results='asis'}
+filor::print_fun(funs$clm_to_ord)
+```
+
+```{r}
+#| echo: true
+
+dat <- data.frame(
+  x = rnorm(1e5)
+)
+
+dat <- sim_ord_latent(~x, By = 1, probs = rep(1/5, 5), link = "probit", data = dat)
+fit <- clm(y ~ x, data = dat, link = "probit")
+clm_to_ord(fit)
+```
+
+## Gelman parametrization
+
+rivedi questa parametrizzazione in @Gelman2020-tg
+
+```{r}
+#| eval: false
+dat <- data.frame(x = runif(1e5, 0, 1))
+dat <- sim_ord_latent(~x, By = 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), data = dat, link = "logit")
+fit_clm <- clm(y ~ x, data = dat, link = "logit")
+
+num_latent_plot(dat$x, 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), link = "logit")
+
+summary(fit_clm)
+
+ths <- -(fit_clm$alpha/fit_clm$beta)
+betas <- 1/fit_clm$beta
+
+dat <- cbind(dat, dummy_ord(dat$y))
+
+fit <- glm(y1234vs5 ~ x, data = dat, family = binomial(link = "logit"))
+
+coefs <- coef(fit)
+
+x <- seq(0, 1, 0.01)
+
+plot(x, plogis(coefs[1] + coefs[2] * x), type = "l")
+
+-(fit_clm$alpha/fit_clm$beta)
+1/fit_clm$beta
+
+```
 
@@
 The code below show that simulating data with the proportional odds create $k - 1$ binomial regression with similar $\beta$s.
+In fact, fitting $k - 1$ binomial regressions can be considered also an alternative to fitting an ordinal regression [see @Gelman2020-tg] with more flexibility in parameters (there will be $k - 1$ regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).
 
@@
 ```
+
+spiegare qui la differenza tra proportional and non proportional odds @Tutz2020-xq.
 
@@
 
+### simulating using probabilities
+
+The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution.
+This is similar to the general method to simulate data for a generalized linear model where the linear predictors $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.
+The following code box simulate a $n$ binary trials with a continuous predictor $x$.
+
+```{r}
+xl <- seq(0, 1, 0.1)
+x <- rep(xl, 10)
+n <- length(x)
+b0 <- qlogis(0.01)
+b1 <- 8
+lp <- b0 + b1 * x # linear predictor
+y <- rbinom(n, 1, plogis(lp)) # sampling from a binomial distribution using the vector of probabilities
+p <- tapply(y, x, mean)
+plot(unique(x), p, pch = 19)
+curve(plogis(x, -(b0/b1), 1/b1), add = TRUE, lwd = 2, col = "firebrick")
+```
+
+We can apply the same idea to an ordinal outcome but we need $k - 1$ equations where $k$ is the number of ordinal levels.
+Let's simulate a similar design with a continuous $x$ predictor and $k = 4$ options.
+We fix the baseline probabilities where $x = 0$ as uniform thus $p(y_1) = p(y_2) = ...
+p(y_k) = 1/k$.
+The general workflow for the simulation is:
+
+1. Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
+2. Define the regression coefficients $\beta$s
+3. Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
+4. Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\mathbf{\beta}$.
+5. Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
+6.
+Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities.
+$p(y = 1) = 0 - p(y <= 1), p(y = 2) = p(y <= 2) - p(y <= 1)$ (sistema qua).
+This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities.
+7. Sample $n$ outcomes from a multinomial distribution, choosing between $k$ alternatives with the calculated probabilities.
+
+(rivedi se fare con for)
+This can be easily implemented in R. We are using a `for` loop to improve the clarity of the workflow but using the `*apply` family of functions can improve the efficiency of the computations.
+
+```{r}
+k <- 4  # number of options (1)
+n <- 1e5 # number of observations (1)
+b1 <- 0.5 # beta1, the shift in the latent distribution (2)
+probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
+alphas <- probs_to_th(probs0, link = "probit") # get true thresholds from probabilities (3)
+dat <- data.frame(x = rnorm(n)) # create dataframe
+lp <- lapply(alphas, function(a) a - b1 * dat$x) # k - 1 linear predictors (4)
+names(lp) <- sprintf("g[p(y <= %s)]", 1:(k - 1)) # giving appropriate names
+```
+
+The linear predictors can be seen in the Figure ref. The $g$ in this case is the cumulative *probit* function $\Phi$.
+
+```{r}
+#| echo: false
+lpt <- lp
+names(lpt) <- latex2exp::TeX(sprintf("$g[p(y \\leq %s)]$", 1:(k - 1)))
+datl <- cbind(dat, lpt)
+datl <- pivot_longer(datl, 2:ncol(datl), names_to = "cumy", values_to = "lp")
+
+ggplot(datl, aes(x = x, y = lp, color = cumy)) +
+  facet_wrap(~cumy, labeller = label_parsed) +
+  geom_hline(yintercept = alphas, lty = "dashed") +
+  geom_line(show.legend = FALSE) +
+  theme_minimal(20) +
+  ylab("Linear Predictor")
+```
+
+Now we can apply the inverse of the link function $g^{-1} = \Phi^{-1}$ to calculate the corresponding cumulative probabilities.
+
+```{r}
+cump <- lapply(lp, pnorm) # inverse of the link function (5)
+cump <- data.frame(cump)
+p <- apply(cbind(0, cump, 1), 1, diff, simplify = FALSE) # probability of each k outcome (6)
+p <- do.call(rbind, p)
+```
+
+```{r}
+p <- data.frame(p)
+names(p) <- sprintf("p_y%s", 1:k)
+filor::trim_df(data.frame(p))
+```
+
+We can plot the expected effect of $\beta_1$ on the $k$ probabilities using the `num_latent_plot()` function (see Figure \@ref(fig:sim-with-probs-lat-plot)).
+Finally we sample using the `sample()` function using the calculated probabilities.
+
+```{r sim-with-probs-lat-plot, fig.cap="sim-with-probs-lat-plot"}
+num_latent_plot(x = dat$x, b1 = b1, probs = probs0, link = "probit")
+```
+
+```{r}
+dat$y <- apply(p, 1, function(ps) sample(1:k, 1, prob = ps))
+```
+
+To check the simulation result we can increase the number of observations, fit the model using the `ordinal::clm()` function and assess the recovery of simulated parameters.
+
+```{r}
+dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
+fit <- clm(y ~ x, data = dat, link = "probit")
+```
+
+```{r}
+#| eval: false
+truth <- c(alphas, b1)
+names(truth) <- c(names(fit$alpha), names(fit$beta))
+clm_table(fit, truth)
+```
+
+### simulating using latent distribution
+
+An equivalent but more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
+This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and the cutting the latent values according to the thresholds $\alpha$.
+The workflow is slighlty different compared to the previous paragraph:
+
+1. Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
+2. Define the regression coefficients $\beta$
+<!-- controlla che le soglie non fanno riferimento a quando è 0 ma concettualmente si. essendo che non cambiano non è rilevante a quale x si faccia riferimento -->
+3. Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
+4.
+Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$, the regression coefficients $\mathbf{\beta}$ and the error $\mathbf{\epsilon}$ sampled from the appropriate probability distribution.
+5.
+Cut the latent variable into $k$ areas using the thresholds $\alpha$ and assign the corresponding ordinal value.
+This step simply checks if the latent value $Y^{\star}_i$ is between two threshold values and assign the corresponding value.
+This can be done using the `cut()` or the `findInterval()` functions.
+
+```{r}
+k <- 4  # number of options (1)
+n <- 1e5 # number of observations (1)
+b1 <- 0.5 # beta1, the shift in the latent distribution (2)
+probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
+alphas <- probs_to_th(probs0, link = "probit") # get true thresholds from probabilities (3)
+dat <- data.frame(x = rnorm(n)) # create dataframe
+dat$ystar <- b1 * dat$x + rnorm(n, 0, 1) # linear predictor (4)
+dat$y <- findInterval(dat$ystar, alphas) + 1 # cut the latent distribution (5). The + 1 because the first category is 0 by default.
+```
+
+The Figure (put crossref) depict the simulated $Y^{*}$ and the corresponding ordinal value. As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.
+
+```{r}
+alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))
+
+dat |> 
+  slice_sample(n = 1000) |> 
+  mutate(y = findInterval(ystar, alphas) + 1) |> 
+  ggplot(aes(x = x, y = ystar)) +
+  geom_point(alpha = 0.6, aes(color = factor(y)), size = 3) +
+  theme_minimal(15) +
+  theme(legend.position = "bottom",
+        legend.title = element_blank()) +
+  ylab(latex2exp::TeX("$Y^{*}$")) +
+  geom_hline(yintercept = alphas) +
+  annotate("label", x = min(dat$x), y = alphas, label = alpha_n)
+```
+
+```{r}
+dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
+fit <- clm(y ~ x, data = dat, link = "probit")
+```
+
+```{r}
+#| eval: false
+truth <- c(alphas, b1)
+names(truth) <- c(names(fit$alpha), names(fit$beta))
+clm_table(fit, truth)
+```
+
+The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function.
+
+```{r}
+#| results: asis
+
+filor::print_fun(funs$sim_ord_latent)
+```
+
 ## Scale Effects
@@
 Figure \@ref(fig:example-unequal-variance) depict an example of a comparison between two groups where the two underlyning latent distributions have unequal variance.
+
+spiegare qui che i location-scale models sono sostanzialmente un cambiamento delle soglie (e non dei beta) in funzione di predittori @Tutz2020-xq
+
+dire che gli scale effects possono essere visti come un'alterativa più parsimoniosa a modelli non-proportional odds @Christensen2019-cz
+
+se citi brms per modelli più complicati, ricordati di specificare che:
+
+> That is, s = 1/disc. Further, because disc must be strictly positive, it is by default modeled on the log scale.
+
+un modo intuitivo per spiegare gli scale effects è di far capire la relazione tra slope e scale.
+La slope della funzione cumulative corrisponde alla varianza.
+Se la slope è piuù steep, la varianza sarà minore e la discrimizione maggiore (in `brms` si chiama desc)
+
+```{r}
+#| eval: false
+library(distributional)
+library(ggdist)
+library(tidyverse)
+
+s <- c(0.5, 1, 2, 3)
+
+dat <- data.frame(
+  m = 0,
+  s = s
+)
+
+dat$dist <- dist_normal(dat$m, dat$s)
+
+par(mfrow = c(1, 2))
+curve(dnorm(x, m = dat$m[1], sd = dat$s[1]), -5, 5, lwd = 2)
+lapply(2:nrow(dat), function(i) curve(dnorm(x, dat$m[i], dat$s[i]), add = TRUE, col = i, lwd = 2))
+
+curve(pnorm(x, m = dat$m[1], sd = dat$s[1]), -5, 5, lwd = 2)
+lapply(2:nrow(dat), function(i) curve(pnorm(x, dat$m[i], dat$s[i]), add = TRUE, col = i, lwd = 2))
+
+# x <- seq(0, 1, 0.001)
+# p <- pnorm(qnorm(0.01) + 5 * x)
+# y <- rbinom(length(x), 1, p)
+# 
+# plot(x, p)
+# curve(pnorm(x, -(qnorm(0.01)/5), 1/5))
+```
+
+si può citare anche questo @Cox1995-ur come riferimento per i modelli
+
+una distinzione tra location-shift and location-scale models @Tutz2017-de dove i primi oltre ad un cambiamento di location prevedono un cambiamento di soglie mentre i secondi modellano la differenza di soglie come un cambiamento della scala (varianza) che quindi permette di aumetare o diminuire l'area associata ad un determinato intervallo.
+
+sempre in @Tutz2017-de propongono il location-scale/shift come una via di mezzo tra il proportional odds ed il non proportional odds
+
+### Simulating scale effects
+
+The location-scale model can be simulated using the `sim_ord_latent()` function and providing the predictors for the scale parameters.
+Given the `log` link function, predictors are provided on the `log` scale.
+For example, we simulate the effect of a binary variable $x$ representing two independent groups predicting the $k = 5$ response.
+We simulate a *location* effect of $\beta_1 = 0.5$ (in *probit* scale) and $\zeta_1 = \text{log}(2) = 0.70$.
+The first group has a $\sigma = 1$ and the second group has $\sigma = 2$.
+Again we simulate that the baseline probabilities are uniform for the first group.
+
+```{r}
+k <- 5  # number of options (1)
+n <- 1e5 # number of observations (1)
+b1 <- 0.5 # beta1, the shift in the latent distribution (2)
+z1 <- log(2) # zeta1, the change in the scale
+probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
+alphas <- probs_to_th(probs0, link = "probit") # get true thresholds from probabilities (3)
+dat <- data.frame(x = rep(c(0, 1), each = n/2))
+dat <- sim_ord_latent(~x, scale = ~x, By = b1, Bscale = z1, probs = probs0, data = dat, link = "probit")
+fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
+```
+
+The table (put ref) reports the simulation results.
+
+```{r}
+#| eval: false
+truth <- c(alphas, b1, z1)
+names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
+clm_table(fit, truth)
+```
 

### 2023-12-11 paper/paper-new.tex
update

@@
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLBlock}[1]{\hfill\break#1\hfill\break}
 \newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
@@
 
+\subsection{notation}\label{notation}
+
+\begin{itemize}
+\tightlist
+\item
+  idea di mettere \(F\) o qualcosa di simile per la pdf (\(\Phi\) e.g., per la normale)
+\item
+  \(Y^\star\) per la latente
+\item
+  \(k\) numero di livelli ordinali
+\item
+  \(\alpha\) per le soglie
+\end{itemize}
+
+per notazione \url{https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf} e Bürkner and Vuorre (2019)
+
+per la parte di notazione e quella più formale anche DeCarlo (2010)
+
+\subsection{cumulative vs others}\label{cumulative-vs-others}
+
+\begin{itemize}
+\tightlist
+\item
+  cumulative quando una variabile continua latente viene tagliata in modo osservato. Le likert possono essere gestite in questo modo
+\end{itemize}
+
+from Bürkner and Vuorre (2019):
+
+\begin{quote}
+For many ordinal variables, the assumption of a single underlying continuous variable, as in cumulative models, may not be appropriate.
+If the response can be understood as being the result of a sequential pro- cess, such that a higher response category is possible only after all lower categories are achieved, the sequen- tial model proposed by Tutz (1990) is usually appropri- ate.
+\end{quote}
+
+vedi se fare k-1 logistiche è più o meno la stessa cosa
+
+in realtà credo che l'adiaject category e sequential model comunque abbiamo la proportional odds assumption
+
+come sottolinea Bürkner and Vuorre (2019) e Gerhard Tutz and Berger (2020) i sequential model sono adatti per un certo tipo di dati dove l'interesse è vedere l'impatto dei predittori sulla transizione ad un categoria \(r\) considerando che la categoria \(r - 1\) è stata raggiunta.
+
+Bürkner and Vuorre (2019) suggest that while cumulative and sequentials models have a clear empirical distinction the adjiacent category model is usually used for computational advantages.
+The probability of passing from one category to the other is modeled as a function of predictors.
+
 \subsection{Risolvere un pochino la nomenclatura}\label{risolvere-un-pochino-la-nomenclatura}
@@
 
-questo per la tassonomia dei diversi modelli ordinali (Tutz, 2022)
+questo per la tassonomia dei diversi modelli ordinali (Gerhard Tutz, 2022)
 
@@
 ##     Min      1Q  Median      3Q     Max 
-## -4.0126 -0.6777 -0.0003  0.6754  4.1015 
+## -4.5222 -0.6765  0.0000  0.6741  4.3425 
 ## 
@@
 ##             Estimate Std. Error t value Pr(>|t|)    
-## (Intercept) -0.003164   0.003161  -1.001    0.317    
-## x1           1.005849   0.003146 319.749   <2e-16 ***
-## x2           0.498501   0.003156 157.971   <2e-16 ***
+## (Intercept) 0.001086   0.003159   0.344    0.731    
+## x1          0.998374   0.003167 315.209   <2e-16 ***
+## x2          0.502929   0.003163 159.004   <2e-16 ***
 ## ---
@@
 ## 
-## Residual standard error: 0.9996 on 99997 degrees of freedom
-## Multiple R-squared:  0.5589, Adjusted R-squared:  0.5589 
-## F-statistic: 6.335e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
+## Residual standard error: 0.9989 on 99997 degrees of freedom
+## Multiple R-squared:  0.5549, Adjusted R-squared:  0.5549 
+## F-statistic: 6.233e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
 \end{verbatim}
@@
 ##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
-##  probit flexible  1e+05 -122018.81 244049.62 5(0)  9.59e-08 9.1e+00
+##  probit flexible  1e+05 -122644.47 245300.95 5(0)  4.75e-08 9.0e+00
 ## 
@@
 ##     x1     x2 
-## 1.0064 0.4993 
+## 0.9974 0.5013 
 ## 
@@
 ##     1|2     2|3     3|4     4|5 
-## -0.8347 -0.2504  0.2529  0.8444
+## -0.8473 -0.2529  0.2551  0.8449
 \end{verbatim}
@@
 ## (Intercept)          x1          x2 
-## -0.003165429  1.006273800  0.498711694
+## 0.001086866 0.999502295 0.503497031
 \end{verbatim}
@@
 They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.
+
+The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation).
+Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion.
+The main improvement regards mapping the values (for thresholds \(\alpha_i\) and slopes \(\beta\)) from latent standard distribution (gaussian or logistic) into the scale of the \(y\) ordinal value.
+The scale of the variable depend on the numeric labels assigned to ordered categories.
+There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to \(\alpha_1 + 0.5\) and \(\alpha_k - 0.5\) where other thresholds are estimated.
+
+Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.
+
+Using the function proposed by Kruschke (2015) (metti ref ad osf) and the equations presented in Kurz (2023) the \texttt{clm\_to\_ord()} function convert parameters fitted with the \texttt{clm} function into the corresponding parameters for the latent \(Y^\star\) variable considering the actual range of values (from 1 to \(k\)).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{clm\_to\_ord }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(fit)\{}
+\NormalTok{  th }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(fit}\SpecialCharTok{$}\NormalTok{alpha) }\CommentTok{\# estimated thresholds}
+\NormalTok{  coefs }\OtherTok{\textless{}{-}}\NormalTok{ fit}\SpecialCharTok{$}\NormalTok{coefficients}
+\NormalTok{  coefs }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(coefs[(}\FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(coefs)])}
+\NormalTok{  k }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# number of levels for y}
+\NormalTok{  y }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k }\CommentTok{\# ordinal values}
+\NormalTok{  stdm }\OtherTok{\textless{}{-}} \DecValTok{0} \CommentTok{\# latent mean}
+\NormalTok{  stds }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# latent sigma}
+\NormalTok{  th\_y }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(th, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{+} \FloatTok{0.5}\NormalTok{, y[k] }\SpecialCharTok{{-}} \FloatTok{0.5}\NormalTok{))}
+\NormalTok{  lats }\OtherTok{\textless{}{-}}\NormalTok{ (th\_y[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th\_y[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\SpecialCharTok{/}\NormalTok{ (th[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\CommentTok{\# real latent sigma}
+\NormalTok{  b0 }\OtherTok{\textless{}{-}}\NormalTok{ th\_y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ lats}
+\NormalTok{  betas }\OtherTok{\textless{}{-}}\NormalTok{ lats }\SpecialCharTok{*}\NormalTok{ coefs}
+\NormalTok{  out }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}\AttributeTok{sigma =}\NormalTok{ lats, }\AttributeTok{b0 =}\NormalTok{ b0, }\AttributeTok{beta =}\NormalTok{ betas, }\AttributeTok{alpha =}\NormalTok{ th\_y)}
+  \FunctionTok{return}\NormalTok{(out)}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
+  \AttributeTok{x =} \FunctionTok{rnorm}\NormalTok{(}\FloatTok{1e5}\NormalTok{)}
+\NormalTok{)}
+
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \DecValTok{1}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{), }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+\FunctionTok{clm\_to\_ord}\NormalTok{(fit)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## $sigma
+## [1] 1.79423
+## 
+## $b0
+## [1] 2.999998
+## 
+## $beta
+## [1] 1.786346
+## 
+## $alpha
+## [1] 1.500000 2.543349 3.447639 4.500000
+\end{verbatim}
+
+\subsection{Gelman parametrization}\label{gelman-parametrization}
+
+rivedi questa parametrizzazione in Gelman, Hill, and Vehtari (2020)
 
@@
 ##            1         2         3         4         5
-## 1 0.19720971 0.2039506 0.2009355 0.1973976 0.2005066
-## 2 0.07341871 0.1042636 0.1503074 0.2345890 0.4374213
+## 1 0.19881313 0.1989910 0.1974955 0.1998745 0.2048259
+## 2 0.07912516 0.1070303 0.1513017 0.2359672 0.4265757
 \end{verbatim}
@@
 ##         xa         xb 
-## -1.003193 -1.003193
+## -0.9791059 -0.9791059
 \end{verbatim}
@@
 The code below show that simulating data with the proportional odds create \(k - 1\) binomial regression with similar \(\beta\)s.
+In fact, fitting \(k - 1\) binomial regressions can be considered also an alternative to fitting an ordinal regression (see Gelman et al., 2020) with more flexibility in parameters (there will be \(k - 1\) regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).
 
@@
 ##               Model 1   Model 2   Model 3
-## (Intercept)  1.1117  0.0148 -1.1014
-## SE           0.0151  0.0163  0.0216
+## (Intercept)  1.098911 -0.000607 -1.096559
+## SE           0.015143  0.016352  0.021668
 ##                                          
-## x           -5.0263 -5.0329 -4.9763
-## SE           0.0365  0.0486  0.0732
+## x             -5.0004   -4.9958   -5.0132
+## SE             0.0364    0.0484    0.0736
 ## 
 \end{verbatim}
+
+spiegare qui la differenza tra proportional and non proportional odds Gerhard Tutz and Berger (2020).
 
@@
 
+\subsubsection{simulating using probabilities}\label{simulating-using-probabilities}
+
+The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution.
+This is similar to the general method to simulate data for a generalized linear model where the linear predictors \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component.
+The following code box simulate a \(n\) binary trials with a continuous predictor \(x\).
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-9-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-9}
+\end{figure}
+
+We can apply the same idea to an ordinal outcome but we need \(k - 1\) equations where \(k\) is the number of ordinal levels.
+Let's simulate a similar design with a continuous \(x\) predictor and \(k = 4\) options.
+We fix the baseline probabilities where \(x = 0\) as uniform thus \(p(y_1) = p(y_2) = ...
+p(y_k) = 1/k\).
+The general workflow for the simulation is:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Define \(n\) (number of observations) and \(k\) (number of ordinal outcomes)
+\item
+  Define the regression coefficients \(\beta\)s
+\item
+  Define the baseline probabilities when \(x = 0\).
+These probabilities can be converted to the corresponding thresholds \(\alpha\) choosing the appropriate link function (\emph{logit} or \emph{probit}).
+\item
+  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\) and regression coefficients \(\mathbf{\beta}\).
+\item
+  Apply the inverse of the link function \(g^{-1}(\eta)\) on the linear predictor and calculate the cumulative probabilities \(p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)\).
+\item
+  Calculate for each observation the probability of the \(k\) outcome as the difference between cumulative probabilities.
+\(p(y = 1) = 0 - p(y <= 1), p(y = 2) = p(y <= 2) - p(y <= 1)\) (sistema qua).
+This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities.
+\item
+  Sample \(n\) outcomes from a multinomial distribution, choosing between \(k\) alternatives with the calculated probabilities.
+\end{enumerate}
+
+(rivedi se fare con for)
+This can be easily implemented in R.
+We are using a \texttt{for} loop to improve the clarity of the workflow but using the \texttt{*apply} family of functions can improve the efficiency of the computations.
+
+The linear predictors can be seen in the Figure ref. The \(g\) in this case is the cumulative \emph{probit} function \(\Phi\).
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-11-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-11}
+\end{figure}
+
+Now we can apply the inverse of the link function \(g^{-1} = \Phi^{-1}\) to calculate the corresponding cumulative probabilities.
+
+\begin{verbatim}
+##    p_y1  p_y2  p_y3  p_y4
+## 1  0.57 0.232 0.134 0.063
+## 2 0.128 0.194 0.262 0.415
+## 3   0.1 0.172 0.255 0.472
+## 4 0.043 0.105 0.207 0.645
+## 5   ...   ...   ...   ...
+## 6 0.381 0.264 0.207 0.147
+## 7 0.459 0.257 0.177 0.106
+## 8 0.332 0.263 0.225  0.18
+## 9 0.439  0.26 0.185 0.116
+\end{verbatim}
+
+We can plot the expected effect of \(\beta_1\) on the \(k\) probabilities using the \texttt{num\_latent\_plot()} function (see Figure \ref{fig:sim-with-probs-lat-plot}).
+Finally we sample using the \texttt{sample()} function using the calculated probabilities.
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/sim-with-probs-lat-plot-1}
+
+}
+
+\caption{sim-with-probs-lat-plot}\label{fig:sim-with-probs-lat-plot}
+\end{figure}
+
+To check the simulation result we can increase the number of observations, fit the model using the \texttt{ordinal::clm()} function and assess the recovery of simulated parameters.
+
+\subsubsection{simulating using latent distribution}\label{simulating-using-latent-distribution}
+
+An equivalent but more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
+This require simulating a standard linear regression using the appropriate data generation function (\emph{logistic} or \emph{normal}) and the cutting the latent values according to the thresholds \(\alpha\).
+The workflow is slighlty different compared to the previous paragraph:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Define \(n\) (number of observations) and \(k\) (number of ordinal outcomes)
+\item
+  Define the regression coefficients \(\beta\)
+\item
+  Define the baseline probabilities when \(x = 0\).
+These probabilities can be converted to the corresponding thresholds \(\alpha\) choosing the appropriate link function (\emph{logit} or \emph{probit}).
+\item
+  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\), the regression coefficients \(\mathbf{\beta}\) and the error \(\mathbf{\epsilon}\) sampled from the appropriate probability distribution.
+\item
+  Cut the latent variable into \(k\) areas using the thresholds \(\alpha\) and assign the corresponding ordinal value.
+This step simply checks if the latent value \(Y^{\star}_i\) is between two threshold values and assign the corresponding value.
+This can be done using the \texttt{cut()} or the \texttt{findInterval()} functions.
+\end{enumerate}
+
+The Figure (put crossref) depict the simulated \(Y^{*}\) and the corresponding ordinal value.
+As for the previous simulation we can fit the model using \texttt{clm()} and check the estimated parameters.
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-18-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-18}
+\end{figure}
+
+The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{sim\_ord\_latent }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(location,}
+                           \AttributeTok{scale =} \ConstantTok{NULL}\NormalTok{,}
+\NormalTok{                           By,}
+                           \AttributeTok{Bscale =} \ConstantTok{NULL}\NormalTok{,}
+                           \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
+                           \AttributeTok{alphas =} \ConstantTok{NULL}\NormalTok{, }
+\NormalTok{                           data, }
+                           \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{))\{}
+
+  \CommentTok{\# get all the correct functions according to the link}
+\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(alphas))\{}
+    \CommentTok{\# calculate thresholds if not provided}
+\NormalTok{    alphas }\OtherTok{\textless{}{-}} \FunctionTok{probs\_to\_th}\NormalTok{(probs, }\AttributeTok{link =}\NormalTok{ link)}
+\NormalTok{  \}}
+\NormalTok{  k }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(alphas) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# number of ordinal outcomes}
+\NormalTok{  n }\OtherTok{\textless{}{-}} \FunctionTok{nrow}\NormalTok{(data) }\CommentTok{\# number of observations}
+
+  \CommentTok{\# model matrix for the location effect}
+\NormalTok{  Xy }\OtherTok{\textless{}{-}} \FunctionTok{model.matrix}\NormalTok{(location, }\AttributeTok{data =}\NormalTok{ data)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, drop }\OtherTok{=} \ConstantTok{FALSE}\NormalTok{] }\CommentTok{\# remove intercept}
+\NormalTok{  lpy }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(Xy }\SpecialCharTok{\%*\%}\NormalTok{ By) }\CommentTok{\# linear predictor for the location}
+\NormalTok{  lps }\OtherTok{\textless{}{-}} \DecValTok{0} \CommentTok{\# default scale effect 0, exp(0) = 1 (the default scale)}
+
+  \CommentTok{\# check predictors on the scale parameter}
+  \ControlFlowTok{if}\NormalTok{(}\SpecialCharTok{!}\FunctionTok{is.null}\NormalTok{(scale))\{}
+    \CommentTok{\# model matrix for the scale effect}
+\NormalTok{    Xsigma }\OtherTok{\textless{}{-}} \FunctionTok{model.matrix}\NormalTok{(scale, }\AttributeTok{data =}\NormalTok{ data)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, drop }\OtherTok{=} \ConstantTok{FALSE}\NormalTok{] }\CommentTok{\# remove intercept}
+\NormalTok{    lps }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(Xsigma }\SpecialCharTok{\%*\%}\NormalTok{ Bscale) }\CommentTok{\# linear predictor for the scale}
+\NormalTok{  \}}
+
+  \CommentTok{\# latent variable with appropriate error function}
+\NormalTok{  ystar }\OtherTok{\textless{}{-}}\NormalTok{ lf}\SpecialCharTok{$}\FunctionTok{rfun}\NormalTok{(n, lpy, }\FunctionTok{exp}\NormalTok{(lps))}
+
+  \CommentTok{\# cut according to thresholds}
+\NormalTok{  y }\OtherTok{\textless{}{-}} \FunctionTok{findInterval}\NormalTok{(ystar, alphas) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# to start from 1}
+
+\NormalTok{  data}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{ordered}\NormalTok{(y) }\CommentTok{\# to ordered factor}
+\NormalTok{  data}\SpecialCharTok{$}\NormalTok{ys }\OtherTok{\textless{}{-}}\NormalTok{ ystar }\CommentTok{\# save also the latent}
+  \FunctionTok{return}\NormalTok{(data)}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
 \subsection{Scale Effects}\label{scale-effects}
@@
 Figure \ref{fig:example-unequal-variance} depict an example of a comparison between two groups where the two underlyning latent distributions have unequal variance.
+
+spiegare qui che i location-scale models sono sostanzialmente un cambiamento delle soglie (e non dei beta) in funzione di predittori Gerhard Tutz and Berger (2020)
+
+dire che gli scale effects possono essere visti come un'alterativa più parsimoniosa a modelli non-proportional odds Christensen (2019)
+
+se citi brms per modelli più complicati, ricordati di specificare che:
+
+\begin{quote}
+That is, s = 1/disc. Further, because disc must be strictly positive, it is by default modeled on the log scale.
+\end{quote}
+
+un modo intuitivo per spiegare gli scale effects è di far capire la relazione tra slope e scale.
+La slope della funzione cumulative corrisponde alla varianza.
+Se la slope è piuù steep, la varianza sarà minore e la discrimizione maggiore (in \texttt{brms} si chiama desc)
+
+si può citare anche questo Cox (1995) come riferimento per i modelli
+
+una distinzione tra location-shift and location-scale models G.
+Tutz and Berger (2017) dove i primi oltre ad un cambiamento di location prevedono un cambiamento di soglie mentre i secondi modellano la differenza di soglie come un cambiamento della scala (varianza) che quindi permette di aumetare o diminuire l'area associata ad un determinato intervallo.
+
+sempre in G. Tutz and Berger (2017) propongono il location-scale/shift come una via di mezzo tra il proportional odds ed il non proportional odds
+
+\subsubsection{Simulating scale effects}\label{simulating-scale-effects}
+
+The location-scale model can be simulated using the \texttt{sim\_ord\_latent()} function and providing the predictors for the scale parameters.
+Given the \texttt{log} link function, predictors are provided on the \texttt{log} scale.
+For example, we simulate the effect of a binary variable \(x\) representing two independent groups predicting the \(k = 5\) response.
+We simulate a \emph{location} effect of \(\beta_1 = 0.5\) (in \emph{probit} scale) and \(\zeta_1 = \text{log}(2) = 0.70\).
+The first group has a \(\sigma = 1\) and the second group has \(\sigma = 2\).
+Again we simulate that the baseline probabilities are uniform for the first group.
+
+The table (put ref) reports the simulation results.
 
@@
                             \AttributeTok{plot =} \FunctionTok{c}\NormalTok{(}\StringTok{"probs"}\NormalTok{, }\StringTok{"latent"}\NormalTok{, }\StringTok{"both"}\NormalTok{))\{}
+  \FunctionTok{require}\NormalTok{(ggplot2)}
 \NormalTok{  plot }\OtherTok{\textless{}{-}} \FunctionTok{match.arg}\NormalTok{(plot)}
@@
 \NormalTok{  latl }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(lat, ps)}
-\NormalTok{  latl }\OtherTok{\textless{}{-}} \FunctionTok{pivot\_longer}\NormalTok{(latl, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}
+\NormalTok{  latl }\OtherTok{\textless{}{-}}\NormalTok{ tidyr}\SpecialCharTok{::}\FunctionTok{pivot\_longer}\NormalTok{(latl, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}
 \NormalTok{  probs\_plot }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(latl, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ value, }\AttributeTok{fill =}\NormalTok{ y)) }\SpecialCharTok{+}
@@
 \NormalTok{  \}}\ControlFlowTok{else}\NormalTok{\{}
-\NormalTok{    legend\_b }\OtherTok{\textless{}{-}} \FunctionTok{get\_legend}\NormalTok{(}
+\NormalTok{    legend\_b }\OtherTok{\textless{}{-}}\NormalTok{ cowplot}\SpecialCharTok{::}\FunctionTok{get\_legend}\NormalTok{(}
 \NormalTok{      probs\_plot}
 \NormalTok{    )}
-\NormalTok{    plts }\OtherTok{\textless{}{-}} \FunctionTok{plot\_grid}\NormalTok{(}
+\NormalTok{    plts }\OtherTok{\textless{}{-}}\NormalTok{ cowplot}\SpecialCharTok{::}\FunctionTok{plot\_grid}\NormalTok{(}
 \NormalTok{      lat\_plot }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{),}
@@
 \NormalTok{    )}
-    \FunctionTok{plot\_grid}\NormalTok{(plts, legend\_b, }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{, }\AttributeTok{rel\_heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{))}
+\NormalTok{    cowplot}\SpecialCharTok{::}\FunctionTok{plot\_grid}\NormalTok{(plts, legend\_b, }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{, }\AttributeTok{rel\_heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{))}
 \NormalTok{  \}}
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-Burkner2019-aw}
+Bürkner, P.-C., \& Vuorre, M.
+(2019).
+Ordinal regression models in psychology: A tutorial.
+\emph{Advances in Methods and Practices in Psychological Science}, \emph{2}, 77--101.
+\url{https://doi.org/10.1177/2515245918823199}
+
+\bibitem[\citeproctext]{ref-Christensen2019-cz}
+Christensen, R. H. B. (2019). \emph{Ordinal---regression models for ordinal data}.
+
 \bibitem[\citeproctext]{ref-Cliff2016-ck}
@@
 
+\bibitem[\citeproctext]{ref-Cox1995-ur}
+Cox, C.
+(1995).
+Location-scale cumulative odds models for ordinal data: A generalized non-linear model approach.
+\emph{Statistics in Medicine}, \emph{14}, 1191--1203.
+\url{https://doi.org/10.1002/sim.4780141105}
+
 \bibitem[\citeproctext]{ref-DeCarlo1998-ay}
 DeCarlo, L. T. (1998). Signal detection theory and generalized linear models. \emph{Psychological Methods}, \emph{3}, 186--205. \url{https://doi.org/10.1037/1082-989X.3.2.186}
+
+\bibitem[\citeproctext]{ref-DeCarlo2010-lj}
+DeCarlo, L.
+T.
+(2010).
+On the statistical and theoretical basis of signal detection theory and extensions: Unequal variance, random coefficient, and mixture models.
+\emph{Journal of Mathematical Psychology}, \emph{54}, 304--313.
+\url{https://doi.org/10.1016/j.jmp.2010.01.001}
 
@@
 \url{https://doi.org/10.5334/irsp.82}
+
+\bibitem[\citeproctext]{ref-Gelman2020-tg}
+Gelman, A., Hill, J., \& Vehtari, A. (2020). \emph{Regression and other stories}. Cambridge University Press. \url{https://doi.org/10.1017/9781139161879}
 
@@
 Kruschke, J. K. (2015). \emph{Doing bayesian data analysis}. \url{https://doi.org/10.1016/c2012-0-00477-2}
+
+\bibitem[\citeproctext]{ref-Kurz2023-zv}
+Kurz, A. S. (2023). \emph{Doing bayesian data analysis in brms and the tidyverse} (Version 1.1.0). Retrieved from \url{https://bookdown.org/content/3686/}
 
@@
 \bibitem[\citeproctext]{ref-Tutz2022-dg}
-Tutz, G. (2022). Ordinal regression: A review and a taxonomy of models. \emph{Wiley Interdisciplinary Reviews. Computational Statistics}, \emph{14}. \url{https://doi.org/10.1002/wics.1545}
+Tutz, Gerhard. (2022). Ordinal regression: A review and a taxonomy of models. \emph{Wiley Interdisciplinary Reviews. Computational Statistics}, \emph{14}. \url{https://doi.org/10.1002/wics.1545}
+
+\bibitem[\citeproctext]{ref-Tutz2017-de}
+Tutz, G., \& Berger, M.
+(2017).
+Separating location and dispersion in ordinal regression models.
+\emph{Econometrics and Statistics}, \emph{2}, 131--148.
+\url{https://doi.org/10.1016/j.ecosta.2016.10.002}
+
+\bibitem[\citeproctext]{ref-Tutz2020-xq}
+Tutz, Gerhard, \& Berger, M.
+(2020).
+Non proportional odds models are widely dispensable -- sparser modeling based on parametric and additive location-shift approaches.
+Retrieved from \url{http://arxiv.org/abs/2006.03914}
 

### 2023-12-11 R/utils-paper.R

@@ -1,0 +1,45 @@
+tidy_clm <- function(fit, ...){
+  fits <- broom::tidy(fit, ...)
+  return(fits)
+}
+
+clm_table <- function(fit, truth = NULL){
+  require(flextable)
+  require(ftExtra)
+  fits <- tidy_clm(fit)
+  nalpha <- length(fit$alpha)
+  if(!is.null(truth)){
+    fits$estimate <- sprintf("%.3f (*%.3f*)", fits$estimate, truth)
+  }
+  fits <- dplyr::select(fits, coef.type, dplyr::everything())
+  fits$coef.type <- dplyr::case_when(
+    fits$coef.type == "intercept" ~ "$\\alpha$",
+    fits$coef.type == "location" ~ "$\\mu$",
+    fits$coef.type == "scale" ~ "$\\sigma$",
+    TRUE ~ NA
+  )
+  
+  flextable(fits) |> 
+    flextable::merge_v(1) |> 
+    colformat_double(digits = 3) |> 
+    autofit() |> 
+    theme_vanilla() |> 
+    set_header_labels(coef.type = "", 
+                      term = "Parameter", 
+                      estimate = "Value", 
+                      std.error = "SE",
+                      statistic = "z",
+                      p.value = "p") |> 
+    align(part = "all", align = "center") |> 
+    colformat_md(part = "all")
+}
+
+z_to_x <- function(z, mean, sd){
+  (z * sd) + mean
+}
+
+
+
+th_names <- function(k){
+  sprintf("%s|%s", 1:(k-1), 2:k)
+}

### 2023-12-11 R/utils.R

@@ -1,58 +1,16 @@
 th_to_probs <- function(th, link = c("logit", "probit"), ...){
-  link <- match.arg(link)
-  if(link == "logit"){
-    lfun <- plogis
-  }else{
-    lfun <- pnorm
-  }
+  lf <- get_link(link)
   ths <- c(-Inf, unname(th), Inf)
-  cprobs <- lfun(ths, ...)
+  cprobs <- lf$pfun(ths, ...)
   probs <- diff(cprobs)
   return(probs)
 }
 
-probs_to_th <- function(probs, link = c("logit", "probit")){
-  link <- match.arg(link)
-  if(link == "logit"){
-    lfun <- qlogis
-  }else{
-    lfun <- qnorm
-  }
+probs_to_th <- function(probs, link = c("logit", "probit"), ...){
+  lf <- get_link(link)
   cprobs <- cumsum(probs)
-  lfun(cprobs[-length(cprobs)])
+  lf$qfun(cprobs[-length(cprobs)], ...)
 }
-
-# get_probs <- function(formula, 
-#                       B, 
-#                       probs = NULL, 
-#                       th = NULL, 
-#                       ynames = NULL, 
-#                       data, 
-#                       get_lp = FALSE, 
-#                       link = c("logit", "probit")){
-#   link <- match.arg(link)
-#   X <- model.matrix(formula, data = data)
-#   if(is.null(th)){
-#     th <- probs_to_th(probs, link = "probit")
-#   }
-#   probs <- th_to_probs(th, "probit")
-#   
-#   pnorm(as.vector(B %*% X[, -1]), mean = as.vector(B %*% X[, -1]))
-#   
-#   
-#   
-#   
-#   
-#   
-#   lp <- lapply(ths, function(x) as.vector(pnorm(X %*% c(x, B))))
-#   lp <- data.frame(lp)
-#   probs <- data.frame(t(diff(t(lp))))
-#   
-#   
-#   
-#   
-#   return(probs)
-# }
 
 get_y <- function(data = NULL, P, ordered = TRUE){
   y <- apply(P, 1, function(x) sample(names(P), 1, replace = TRUE, prob = x))
@@ -125,33 +83,44 @@ get_link <- function(link = c("logit", "probit")){
        dfun_n = dfun_n, pfun_n = pfun_n)
 }
 
-sim_ord_latent <- function(formula,
-                           sigma = NULL,
+sim_ord_latent <- function(location,
+                           scale = NULL,
                            By,
[…]

### 2023-12-12 paper/paper-new.Rmd
update

@@
     latex_engine: xelatex
+    header_includes:
+      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline}
 ---
@@
 library(here)
+library(flextable)
 devtools::load_all()
@@
 probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- probs_to_th(probs0, link = "probit") # get true thresholds from probabilities (3)
+alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
 dat <- data.frame(x = rnorm(n)) # create dataframe
@@
 datl <- cbind(dat, lpt)
-datl <- pivot_longer(datl, 2:ncol(datl), names_to = "cumy", values_to = "lp")
+datl <- tidyr::pivot_longer(datl, 2:ncol(datl), names_to = "cumy", values_to = "lp")
 
@@
 
-```{r}
-#| eval: false
+```{r tab-sim-probs}
 truth <- c(alphas, b1)
 names(truth) <- c(names(fit$alpha), names(fit$beta))
-clm_table(fit, truth)
+clm_table(fit, truth = truth, caption = "caption") |> 
+  flextable_to_rmd()
 ```
@@
 
-```{r}
-#| eval: false
+```{r tab-sim-latent}
 truth <- c(alphas, b1)
@@
 probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- probs_to_th(probs0, link = "probit") # get true thresholds from probabilities (3)
+alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
 dat <- data.frame(x = rep(c(0, 1), each = n/2))

### 2023-12-12 R/utils.R

@@ -1,4 +1,100 @@
-th_to_probs <- function(th, link = c("logit", "probit"), ...){
+show_alpha <- function(prob = NULL, 
+                       alpha = NULL, 
+                       link = c("logit", "probit")){
+  require(ggplot2)
+  require(cowplot)
+  require(ggtext)
+  
+  if(is.null(prob) & is.null(alpha)){
+    stop("alpha or prob must be specified!")
+  }
+  
+  lf <- get_link(link)
+  
+  if(link == "logit"){
+    xlim <- c(-7, 7)
+    dist <- distributional::dist_logistic
+  }else{
+    xlim <- c(-5, 5)
+    dist <- distributional::dist_normal
+  }
+  
+  if(is.null(prob)){
+    prob <- alpha_to_prob(alpha, link)
+  }else{
+    alpha <- prob_to_alpha(prob, link)
+  }
+  
+  k <- length(prob)
+  D <- data.frame(l = 0, s = 1)
+  D$dist <- dist(D$l, D$s)
+  
+  alpha_n <- th_names(k)
+  
+  dplot <- ggplot(D) +
+    ggdist::stat_slab(aes(dist = dist,
+                          fill = factor(after_stat(findInterval(x, alpha) + 1))),
+                      orientation = "horizontal",
+                      alpha = 0.8) +
+    xlim(xlim) +
+    theme_minimal(15) +
+    theme(legend.title = element_blank(),
+          legend.position = "bottom",
+          axis.title.y = element_blank(),
+          axis.title.x = element_blank()) +
+    geom_vline(xintercept = alpha,
+               alpha = 0.5) +
+    ggtext::geom_richtext(data = data.frame(alpha, alpha_n),
+                          aes(x = alpha, 
+                              y = 1, 
+                              label = alpha_n), 
+                          angle = 90,
+                          hjust = 0.5)
+  
+  y_n <- sprintf("P(Y = %s)", 1:k)
+  y_cp <- lf$pfun(c(-Inf, alpha, Inf))
+  y_p <- 0.5 * (y_cp[1:(length(y_cp) - 1)] + y_cp[2:length(y_cp)])
+  
+  pplot <- ggplot() +
+    stat_function(geom = "line", fun = lf$pfun) +
+    xlim(xlim) +
[…]

### 2023-12-12 paper/paper-new.Rmd
updating

--- 2023-12-11 paper/paper-new.Rmd
@@
 # Introduction
+
+Psychological research make an extensive use of ordinal data.
+One of the main reason is probably the usage of Likert scales [@Likert1932-xw].
+Ordinal data refers to a specific type of measurement scale [@Stevens1946-te] where ordered numbers are assigned to a variable.
+Compared to nominal scale and as the name suggest the labels are ordered.
+Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
+(rivedi questo).
+An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
+Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer.
+@Stevens1946-te suggest that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation).
+This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data.
+In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable.
+@Liddell2018-wu reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models.
+In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric.
+They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui).
+(vedi se c'è qualche altro lavoro che fa vedere questa cosa).
+
+### Ordinal regression models
+
+Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without making metric assumptions.
+We can class this general class of models as *ordinal regression*.
+The nomenclature of these models can be confusing mainly because there are several subclasses of models with different assumptions and structures [@Tutz2022-dg].
+@Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
+We can identify three main classes: *cumulative models* [@McCullagh1980-cw], *sequential models* [@Tutz1990-fe] and *adjacent category models*.
+The cumulative is the mostly used model assuming the existence of a latent variable that when segmented using thresholds produces the observed ordinal variable.
+The psychological process underlying the response is clearly formalized in the signal detection theory framework where the respondent (vedi come sistemare).
+The sequential model as suggested by the name is appropriate when modelling sequential processes.
+Assuming to have 5 response options, the sequential model assume that responding 3 means already reaching the states 1 and 2.
+A clear example is proposed by @Burkner2019-aw where the marriage duration in years is predicted as a function of some explanatory variables.
+For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential regression.
+When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached.
+The adjacent category model compare the category $k$ with $k + 1$ still assuming a latent distribution for each $k$.
+As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
 

### 2023-12-12 paper/paper-new.tex

@@
 ##     Min      1Q  Median      3Q     Max 
-## -4.5222 -0.6765  0.0000  0.6741  4.3425 
+## -4.0357 -0.6735 -0.0024  0.6761  4.0977 
 ## 
@@
 ##              Estimate Std. Error t value Pr(>|t|)    
-## (Intercept) 0.001086   0.003159   0.344    0.731    
-## x1          0.998374   0.003167 315.209   <2e-16 ***
-## x2          0.502929   0.003163 159.004   <2e-16 ***
+## (Intercept) 0.0004437  0.0031671    0.14    0.889    
+## x1          0.9995687  0.0031605  316.27   <2e-16 ***
+## x2          0.5046426  0.0031633  159.53   <2e-16 ***
 ## ---
@@
 ## 
-## Residual standard error: 0.9989 on 99997 degrees of freedom
-## Multiple R-squared:  0.5549, Adjusted R-squared:  0.5549 
-## F-statistic: 6.233e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
+## Residual standard error: 1.002 on 99997 degrees of freedom
+## Multiple R-squared:  0.5577, Adjusted R-squared:  0.5576 
+## F-statistic: 6.303e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
 \end{verbatim}
@@
 ##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
-##  probit flexible  1e+05 -122644.47 245300.95 5(0)  4.75e-08 9.0e+00
+##  probit flexible  1e+05 -121998.29 244008.57 5(0)  6.33e-08 9.1e+00
 ## 
@@
 ##     x1     x2 
-## 0.9974 0.5013 
+## 1.0025 0.5039 
 ## 
@@
 ##     1|2     2|3     3|4     4|5 
-## -0.8473 -0.2529  0.2551  0.8449
+## -0.8418 -0.2514  0.2543  0.8413
 \end{verbatim}
@@
 ##  (Intercept)           x1           x2 
-## 0.001086866 0.999502295 0.503497031
+## 0.0004430131 0.9980388055 0.5038702203
 \end{verbatim}
@@
 ## $sigma
-## [1] 1.79423
+## [1] 1.776447
 ## 
 ## $b0
-## [1] 2.999998
+## [1] 2.999548
 ## 
 ## $beta
-## [1] 1.786346
+## [1] 1.777515
 ## 
 ## $alpha
-## [1] 1.500000 2.543349 3.447639 4.500000
+## [1] 1.500000 2.549889 3.449757 4.500000
 \end{verbatim}
@@
 ##            1         2         3         4         5
-## 1 0.19881313 0.1989910 0.1974955 0.1998745 0.2048259
-## 2 0.07912516 0.1070303 0.1513017 0.2359672 0.4265757
+## 1 0.20301120 0.1911931 0.1969380 0.2054224 0.2034352
+## 2 0.07967757 0.1014347 0.1483835 0.2414729 0.4290314
 \end{verbatim}
@@
 ##         xa         xb 
-## -0.9791059 -0.9791059
+## -0.9379058 -0.9379058
 \end{verbatim}
@@
 ##             Model 1 Model 2 Model 3
-## (Intercept)  1.098911 -0.000607 -1.096559
-## SE           0.015143  0.016352  0.021668
-##                                          
-## x             -5.0004   -4.9958   -5.0132
-## SE             0.0364    0.0484    0.0736
+## (Intercept)  1.0844 -0.0176 -1.1199
+## SE           0.0151  0.0163  0.0216
+##                                    
+## x           -4.9826 -4.9374 -4.9444
+## SE           0.0364  0.0481  0.0733
 ## 
@@
 ##    p_y1  p_y2  p_y3  p_y4
-## 1  0.57 0.232 0.134 0.063
-## 2 0.128 0.194 0.262 0.415
-## 3   0.1 0.172 0.255 0.472
-## 4 0.043 0.105 0.207 0.645
+## 1 0.481 0.254 0.169 0.096
+## 2 0.092 0.165 0.252 0.491
+## 3 0.098  0.17 0.254 0.478
+## 4 0.426 0.261  0.19 0.122
 ## 5   ...   ...   ...   ...
-## 6 0.381 0.264 0.207 0.147
-## 7 0.459 0.257 0.177 0.106
-## 8 0.332 0.263 0.225  0.18
-## 9 0.439  0.26 0.185 0.116
+## 6  0.11  0.18 0.258 0.452
+## 7 0.315 0.261 0.231 0.193
+## 8 0.108 0.179 0.258 0.456
+## 9 0.065 0.136 0.234 0.565
 \end{verbatim}
@@
 To check the simulation result we can increase the number of observations, fit the model using the \texttt{ordinal::clm()} function and assess the recovery of simulated parameters.
+
+\begin{longtable}{llrrrrr}
+\toprule
+coef.type & term & estimate & std.error & statistic & p.value & row \\ 
+\midrule
+\$\$\textbackslash{}alpha\$\$ & 1|2 & -0.680 (*-0.674*) & $0.004$ & $-152.618$ & $0.000$ & $1.000$ \\ 
+ & 2|3 & -0.005 (*0.000*) & $0.004$ & $-1.179$ & $0.238$ & $2.000$ \\ 
+ & 3|4 & 0.666 (*0.674*) & $0.004$ & $149.925$ & $0.000$ & $3.000$ \\ 
+\$\$\textbackslash{}mu\$\$ & x & 0.495 (*0.500*) & $0.004$ & $132.696$ & $0.000$ & $1.000$ \\ 
+\bottomrule
+\end{longtable}
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-18-1}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-17-1}
 
@@
 
-\caption{ }\label{fig:unnamed-chunk-18}
+\caption{ }\label{fig:unnamed-chunk-17}
 \end{figure}
+
+\begin{longtable}{llrrrrr}
+\toprule
+coef.type & term & estimate & std.error & statistic & p.value & row \\ 
+\midrule
+\$\$\textbackslash{}alpha\$\$ & 1|2 & -0.681 (*-0.674*) & $0.004$ & $-152.568$ & $0.000$ & $1.000$ \\ 
+ & 2|3 & -0.007 (*0.000*) & $0.004$ & $-1.708$ & $0.088$ & $2.000$ \\ 
+ & 3|4 & 0.664 (*0.674*) & $0.004$ & $149.541$ & $0.000$ & $3.000$ \\ 
+\$\$\textbackslash{}mu\$\$ & x & 0.498 (*0.500*) & $0.004$ & $133.483$ & $0.000$ & $1.000$ \\ 
+\bottomrule
+\end{longtable}
 
@@
     \CommentTok{\# calculate thresholds if not provided}
-\NormalTok{    alphas }\OtherTok{\textless{}{-}} \FunctionTok{probs\_to\_th}\NormalTok{(probs, }\AttributeTok{link =}\NormalTok{ link)}
+\NormalTok{    alphas }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs, }\AttributeTok{link =}\NormalTok{ link)}
 \NormalTok{  \}}
@@
   \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(th))\{}
-\NormalTok{    th }\OtherTok{\textless{}{-}} \FunctionTok{probs\_to\_th}\NormalTok{(probs, link)}
+\NormalTok{    th }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs, link)}
 \NormalTok{  \}}

### 2023-12-13 paper/paper-new.Rmd
update

@@
                       fig.align = "center",
+                      size = "scriptsize",
                       message = FALSE,
                       warning = FALSE)
+```
+
+```{r hooks, include = FALSE}
+def.chunk.hook  <- knitr::knit_hooks$get("chunk")
+knitr::knit_hooks$set(chunk = function(x, options) {
+  x <- def.chunk.hook(x, options)
+  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
+})
 ```
@@
 library(flextable)
+library(cowplot)
+library(viridis)
 devtools::load_all()
@@
 funs <- filor::get_funs(here("R", "utils.R"))
+```
+
+```{r ggplot, include = FALSE}
+mtheme <- function(){
+  ggthemes::theme_pander(15) +
+    theme(legend.position = "bottom",
+        plot.title = element_text(hjust = 0.5),
+        strip.text = element_text(face = "bold"),
+        panel.grid.minor = element_blank())
+}
+
+theme_set(mtheme())
+
+# palettes
+options(ggplot2.continuous.colour="viridis")
+options(ggplot2.continuous.fill = "viridis")
+scale_color_discrete <- scale_colour_viridis_d
+scale_fill_discrete <- scale_fill_viridis_d
+```
+
+```{r include = FALSE}
+# figures
+source(here("scripts", "figures.R"))
+rfigs <- readRDS(here("objects", "r-figs.rds"))
+```
+
+```{r tables, include=FALSE}
+mtab <- function(data){
+
+}
 ```
@@
 Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer.
-@Stevens1946-te suggest that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation).
+@Stevens1946-te and @Kemp2021-dj suggested that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation).
 This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data.
+Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels [@Liddell2018-wu].
+
 In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable.
 @Liddell2018-wu reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models.
-In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric.
+In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric [but see @Robitzsch2020-la for an alternative perspective].
 They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui).
 (vedi se c'è qualche altro lavoro che fa vedere questa cosa).
+Some authors suggested that despite individual items being ordinal, averaging multiple items could solve the problem of applying metric models (e.g., Carifio & Perla, 2007, 2008 metti queste citazioni).
+@Liddell2018-wu suggested that in some conditions even averaging multiple items and applying metric models could be problematic.
+@Cliff2016-ck suggest that most of the research questions in behavioral sciences can be considered as ordinal (*is the score $x$ higher than the score $y$*?) concerning variables where the most appropriate measurement scale is probably ordinal.
 
-### Ordinal regression models
+qualcosa ancora qui
+
+## Ordinal regression models
 
@@
 For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential regression.
-When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached.
+When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached [@Tutz2020-xq].
 The adjacent category model compare the category $k$ with $k + 1$ still assuming a latent distribution for each $k$.
 As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
+@Burkner2019-aw suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.
 
-## notation
+In the current paper we put the focus on the cumulative model for several reasons.
+The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation.
+The second reason is that several psychological variables can be formalized as a latent continuous construct that when measured is collected using ordinal items.
+Furthermore, also the signal detection theory framework models the decision process assuming a latent distribution for the signal and noise where thresholds are the response criteria [e.g., @DeCarlo2010-lj].
 
-- idea di mettere $F$ o qualcosa di simile per la pdf ($\Phi$ e.g., per la normale)
-- $Y^\star$ per la latente
-- $k$ numero di livelli ordinali
-- $\alpha$ per le soglie
-
-per notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw
-
-per la parte di notazione e quella più formale anche @DeCarlo2010-lj
-
-## cumulative vs others
-
-- cumulative quando una variabile continua latente viene tagliata in modo osservato. Le likert possono essere gestite in questo modo
-
-from @Burkner2019-aw:
-
-> For  many  ordinal  variables,  the  assumption  of  a   single underlying continuous variable, as in cumulative  models, may not be appropriate.
-If the response can  be understood as being the result of a sequential pro- cess, such that a higher response category is possible  only after all lower categories are achieved, the sequen- tial model proposed by Tutz (1990) is usually appropri- ate.
-
-vedi se fare k-1 logistiche è più o meno la stessa cosa
-
-in realtà credo che l'adiaject category e sequential model comunque abbiamo la proportional odds assumption
-
-```{r}
-#| eval: false
-library(brms)
-library(ordinal)
-
-dat <- data.frame(x = rnorm(100))
-dat <- sim_ord_latent(~x, By = log(2), probs = c(0.5, 0.2, 0.2, 0.1), data = dat, link = "logit")
-
-fit_clm <- clm(y ~ x, data = dat, link = "logit")
-fit_cum <- brm(y ~ x, data = dat, family = cumulative(link = "logit"))
-fit_seq <- brm(y ~ x, data = dat, family = sratio(link = "logit"))
-fit_adj <- brm(y ~ x, data = dat, family = acat(link = "logit"))
-
+```{r fig-ordinal-models, fig.cap=filor::get_caption(), out.width="100%"}
+knitr::include_graphics("img/fig-ordinal-models.png")
 ```
 
-come sottolinea @Burkner2019-aw e @Tutz2020-xq i sequential model sono adatti per un certo tipo di dati dove l'interesse è vedere l'impatto dei predittori sulla transizione ad un categoria $r$ considerando che la categoria $r - 1$ è stata raggiunta.
+## Model notation
 
-@Burkner2019-aw suggest that while cumulative and sequentials models have a clear empirical distinction the adjiacent category model is usually used for computational advantages.
-The probability of passing from one category to the other is modeled as a function of predictors.
-
-## Risolvere un pochino la nomenclatura
-
-https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf
+metti qualcosa sulla nomenclatura
 
@@
 
-Da citare in qualche modo @Kemp2021-dj
+In this section we introduce some notations for the cumulative model that is used through the paper and in the R code.
+The notation is as consistent as possible with the main textbooks and with the `ordinal` R package used in the tutorial.
+We define $Y_k$ as the observed ordinal response with $1, 2, \dots,k$ options and $Y^\star$ is the corresponding latent variable.
+The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \alpha_2, \dots, \alpha_{k - 1}$.
+With $g()$ we define the cumulative probability function assumed for the specific model.
+For example, when using a Gaussian distribution we are fitting a model with a `probit` link function thus $f() = \Phi()$.
+With $g^{-1}()$ we define the inverse of the cumulative probability function.
+The minus sign in $\mathbf{X}\mathbf{\beta}$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).
 
-> “As a matter of fact, most of the scales used widely and effectively by psychologists are ordinal scales.
-In the strictest propriety the ordinary statistics involving means and standard deviations ought not to be used with these scales, for these statistics imply a knowledge of something more than the relative rank-order of data.
-On the other hand for this ‘illegal’ statisticizing there can be invoked a kind of pragmatic sanction: In numerous instances it leads to fruitful results.” (Stevens, 1946, p.
-679.)
+\begin{equation} 
+P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, 2, \dots, k
+(\#eq:prob-cum-model1)
+\end{equation}
 
-> It is straightforward to construct ordinal scales that do not involve rank ordering.
-For example, one can take the first element encountered and arbitrarily assign it the number 100.
-If the next element encountered is smaller it is given a smaller arbitrary number, 53 say.
-If the third element is between these two, it can be given the number 86, and so on.
-If this construction method is used, the difference between the elements assigned the numbers 70 and 80 will not in any important sense be equal to the difference between the elements assigned 90 and 100, and the intervals between the numbers are not really interpretable.
-Note, too, that monotonic transformations of the scale essentially leave the mea- sure unaffected.
+We can call the $\mathbf{X}\mathbf{\beta}$ as the linear predictor $\eta$. Equation (num) describe how the probability of a single outcome $Y_k$.
 
-interessante come @Cliff2016-ck descriva che la maggiorparte dellle domande di ricerca in psicologia siano in riferimento alla location di un costrutto e possano essere gestite in modo ordinale
+\begin{equation} 
+P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta)
+(\#eq:prob-cum-model2)
+\end{equation}
 
-The use of ordinal data is widespread in Psychology.
-Usually items from questionnaires are created using Likert scales where a certain psychological traits is measured using an intuitive scale with 3-5 or more anchor points.
-These measure cannot be considered *metric* measure on interval or ratio scales [stevens] but the categories are ordered.
+<!-- notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw per la parte di notazione e quella più formale anche @DeCarlo2010-lj -->
 
-Other examples here
+The same model can be formalized using the latent formulation (see Equation). This is more similar to a standard linear regression.
 
-Despite the usage of ordinal variables, statistical models made for these type of data are rarely used in Psychology.
-@Liddell2018-wu reported that the majority of published papers using likert-like measures used standard methods to analyze the data.
-In practical terms, they use *metric* models where the response variable cannot be considered fully numeric (vedi se c'è un termine tipo scala a rapporti).
+\begin{equation} 
+Y^\star_i = \eta + \epsilon_i
+(\#eq:latent-model)
+\end{equation}
 
-altro sui modelli
+Where $\epsilon$ is the error part and comes from the distribution that is assumed from the latent model.
+For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution.
+Then the observed ordinal value $Y_i$ comes from $Y^\star_i$ being between the thresholds $\alpha_{k - 1} < Y^\star_i < \alpha_{k}$ with thresholds organized as $- \infty \equiv \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k \equiv \infty$.
+<!-- TODO rivedi questa -->
 
-Theoretically, using a *metric* for ordinal data is not appropriate but understanding the actual impact is not straightforward.
-@Liddell2018-wu did a comprehensive work about pitfalls of analzying ordinal data as metric.
-They showed that metric models produce higher type-1 and type-2 errors compared to the ordinal models.
-In particular, given the bounded nature of ordinal data, difference in the underlying latent distribution are not always captured by the metric model that simply estimate the mean of the ordinal variable.
-This is even more relevant when the underluing variance of the ordinal variables are not homogeneous.
-Furthermore they presented some situations where the metric model could be wrong in the opposite direction, finding an effect with the wrong sign (type-s error).
-
-Finally, they also demonstrated that even in the best condition where e.g.
-comparing two groups the variance are equal, the ordinal model is more powerful than the metric model given the underestimation of the true effect size from the latter.
-
-> Some authors have argued that, despite the ordinal character of individual Likert items, averaged ordinal items can have an emergent property of an interval scale and so it is appropriate to apply metric methods to the averaged values (e.g., Carifio & Perla, 2007, 2008).
-
-vedi come gestire questo
-
-anche questo è importante
-> Ordered-probit models typically assume that the thresholds (θ k) are the same across all groups because the thresholds are theoretically linked to the response measure, not to the predictor value.
-For example, when asked, “How happy are you?” with response options ‘1’ = very unhappy, ‘2’ = mildly unhappy, ‘3’ = neutral, ‘4’ = mildly happy, ‘5’ = very happy,” the latent thresholds between ordinal levels are assumed to be implicit in the phrasing of the question, regardless of other aspects of the respondent or situation.
-In other words, the thresholds are assumed to be part of the measurement procedure, not dependent on the value of the predictor or covariate.
-This can be technically referred to as a type of measurement invariance.
-
-Da citare come paper iniziale sui modelli [@McCullagh1980-cw].
-
-questo per la tassonomia dei diversi modelli ordinali [@Tutz2022-dg]
-
-da vedere magari per una critica tipo Gomila sul binary model [@Robitzsch2020-la]
-
-## Metriv vs ordinal models
-
-- the main difference is that the metric model assign a number to each label of the discrete variable assuming that the distance is the same. [@Liddell2018-wu]
+In the basic version of the model, the thresholds $\alpha$ are considered as fixed and do not vary as a function of predictors.
+The thresholds are part of the measurement procedure [@Liddell2018-wu].
+In a more sophisticated version of the model called location-shift [@Tutz2022-dg], both the location $\mu$ and the thresholds $\alpha$ can vary as a function of the predictors.
+In the following sections another version of the model called location-scale [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur] will be presented as a more flexible way to model latent distributions with heterogeneous variance (spiega meglio).
 
@@
 coef(fit)/sigma(fit)
-
 ```
@@
 
-## Odds ratios and cumulative odds ratios
+# Interpreting parameters
 
-- https://online.stat.psu.edu/stat504/lesson/4/4.1
+## Odds and odds ratio
 
-In this section we introduce the main concept to understand the effects of a logit model.
+una piccola spiegazione di odds e odds ratio
 
-## Demonstrating the proportional odds assumption
+## Proportional odds assumption
 
-basically given that the effect of $\beta$ is constant the odds ratio is independent from the thresholds $\alpha_j$ [@Liu2023-bp].
-In the current paper we are simulating data using a single set of $\beta$s thus the proportional odds assumption is true.
-@Liu2023-bp reviewed the available methods to test this assumption.
+Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the  proportional odds assumption.
+There are more advanced versions of the model relaxing this assumption completely (*non proportional odds* models) and partially (*partial proportional odds* models) (rivedi se la nomenclatura va bene).
+The proportional odds assumption is a crucial point when intepreting model parameters.
+Can be formalized as in Equation ().
+Basically if we use the *logit* link function $g()$ the $\beta$s are interpreted as standard odds ratio in the logistic regression.
+Given that we have $k > 2$ alternatives there is the need of $k - 1$ equations (as reported in Equation ()).
+The porportional odds is assuming that the regression coefficients $\beta$ are independent from the thresholds $\alpha$ thus the effect is the same regardless of the $Y$ level.
+Thus $\beta_1 = \beta_2 \dots = \beta_{k - 1}$.
+The Figure () depicts the proportional odds assumption for the $k - 1$ logistic curves both for probabilities and linear predictors $\eta$.
 
-- https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione
+```{r fig-prop-odds, fig.cap=filor::get_caption()}
+rfigs$fig_prop_odds
+```
 
-sarebbe da capire quali sono i rischi di assumere questo ed eventualmente se simulare non proportional odds è troppo complicato.
+The proportional odds assumption is convenient because regardless the number of $k$, the $\beta_j$ ($j$ being the number of regression coefficients) effect is assumed to be the same.
+The model is more parsimonious compared to estimating $k - 1$ coefficients for each $\beta_j$ as in the multinomial regression or the non-proportional version of the ordinal regression.
+Clearly this can be a strong assumption in some conditions.
+There are several methods to test the proportional odds assumption [see @Liu2023-bp for an overview].
+@Tutz2020-xq suggested that using alternative as the location-shift or location-scale models, the assumption can be made (improving model parsimony) but still having a more flexible modeling framework.
+In the next section, we provide more details about the location-scale model in terms of parameters interpretation and data simulation.
 
-Let's simulate the effect of a binary predictor on ordinal scale 1-5:
+To make a practical example, the dataset presented in Table contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
+The participants are divided into two groups (predictor $x$).
+We want to fit a cumulative ordinal model and check the proportional odds assumption.
+
+<!-- - https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione -->
 
 ```{r}
-#| echo: true
 b1 <- log(3) # log odds ratio
-n <- 1e4
+k <- 4
+n <- 100
 x <- rep(c("a", "b"), each = n/2)
 dat <- data.frame(x = x)
-probs <- rep(1/5, 5) # for the group "a", uniform probabilities
+probs <- rep(1/k, k) # for the group "a", uniform probabilities
 dat <- sim_ord_latent(~x, By = b1, probs = probs, data = dat, link = "logit")
@@
 pr <- predict(fit, data.frame(x = unique(x)))$fit
-pr
 ```
+
 Basically the proportional odds suggest that:
@@
 $$
+
 Is the same regardless the level of the $x$ predictor. Thus:
@@
 names(truth) <- c(names(fit$alpha), names(fit$beta))
-clm_table(fit, truth = truth, caption = "caption") |> 
-  flextable_to_rmd()
+#clm_table(fit, truth = truth)
 ```
@@
 names(truth) <- c(names(fit$alpha), names(fit$beta))
-clm_table(fit, truth)
+#clm_table(fit, truth)
 ```
@@
 #| results: asis
-
 filor::print_fun(funs$sim_ord_latent)
@@
 names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
-clm_table(fit, truth)
+#clm_table(fit, truth)
 ```
@@
 :::
+
+(ref:fig-ordinal-models) Theoretical figure about ordinal models. Voglio fare una figura come questa in @Burkner2019-aw ma con qualche modifica per fare vedere i tre modelli.
+
+(ref:fig-prop-odds) caption here

### 2023-12-13 paper/paper-new.tex
update

--- 2023-12-11 paper/paper-new.tex
@@
 
-\subsection{notation}\label{notation}
-
-\begin{itemize}
-\tightlist
-\item
-  idea di mettere \(F\) o qualcosa di simile per la pdf (\(\Phi\) e.g., per la normale)
-\item
-  \(Y^\star\) per la latente
-\item
-  \(k\) numero di livelli ordinali
-\item
-  \(\alpha\) per le soglie
-\end{itemize}
-
-per notazione \url{https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf} e Bürkner and Vuorre (2019)
-
-per la parte di notazione e quella più formale anche DeCarlo (2010)
-
-\subsection{cumulative vs others}\label{cumulative-vs-others}
-
-\begin{itemize}
-\tightlist
-\item
-  cumulative quando una variabile continua latente viene tagliata in modo osservato. Le likert possono essere gestite in questo modo
-\end{itemize}
-
-from Bürkner and Vuorre (2019):
-
-\begin{quote}
-For many ordinal variables, the assumption of a single underlying continuous variable, as in cumulative models, may not be appropriate.
-If the response can be understood as being the result of a sequential pro- cess, such that a higher response category is possible only after all lower categories are achieved, the sequen- tial model proposed by Tutz (1990) is usually appropri- ate.
-\end{quote}
-
-vedi se fare k-1 logistiche è più o meno la stessa cosa
-
-in realtà credo che l'adiaject category e sequential model comunque abbiamo la proportional odds assumption
-
-come sottolinea Bürkner and Vuorre (2019) e Gerhard Tutz and Berger (2020) i sequential model sono adatti per un certo tipo di dati dove l'interesse è vedere l'impatto dei predittori sulla transizione ad un categoria \(r\) considerando che la categoria \(r - 1\) è stata raggiunta.
-
-Bürkner and Vuorre (2019) suggest that while cumulative and sequentials models have a clear empirical distinction the adjiacent category model is usually used for computational advantages.
-The probability of passing from one category to the other is modeled as a function of predictors.
-
-\subsection{Risolvere un pochino la nomenclatura}\label{risolvere-un-pochino-la-nomenclatura}
-
-\url{https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf}
+Psychological research make an extensive use of ordinal data.
+One of the main reason is probably the usage of Likert scales (Likert, 1932).
+Ordinal data refers to a specific type of measurement scale (Stevens, 1946) where ordered numbers are assigned to a variable.
+Compared to nominal scale and as the name suggest the labels are ordered.
+Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
+(rivedi questo).
+An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
+Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer.
+Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation).
+This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data.
+Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018).
+
+In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable.
+Liddell and Kruschke (2018) reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models.
+In the same work, Liddell and Kruschke (2018) showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric (but see Robitzsch, 2020 for an alternative perspective).
+They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui).
+(vedi se c'è qualche altro lavoro che fa vedere questa cosa).
+Some authors suggested that despite individual items being ordinal, averaging multiple items could solve the problem of applying metric models (e.g., Carifio \& Perla, 2007, 2008 metti queste citazioni).
+Liddell and Kruschke (2018) suggested that in some conditions even averaging multiple items and applying metric models could be problematic.
+Cliff (2016) suggest that most of the research questions in behavioral sciences can be considered as ordinal (\emph{is the score \(x\) higher than the score \(y\)}?) concerning variables where the most appropriate measurement scale is probably ordinal.
+
+qualcosa ancora qui
+
+\subsection{Ordinal regression models}\label{ordinal-regression-models}
+
+Despite the actual modeling proposal by Liddell and Kruschke (2018), there is a class of regression models taking into account the ordinal nature of the response variable without making metric assumptions.
+We can class this general class of models as \emph{ordinal regression}.
+The nomenclature of these models can be confusing mainly because there are several subclasses of models with different assumptions and structures (Gerhard Tutz, 2022).
+Gerhard Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
+We can identify three main classes: \emph{cumulative models} (McCullagh, 1980), \emph{sequential models} (Gerhard Tutz, 1990) and \emph{adjacent category models}.
+The cumulative is the mostly used model assuming the existence of a latent variable that when segmented using thresholds produces the observed ordinal variable.
+The psychological process underlying the response is clearly formalized in the signal detection theory framework where the respondent (vedi come sistemare).
+The sequential model as suggested by the name is appropriate when modelling sequential processes.
+Assuming to have 5 response options, the sequential model assume that responding 3 means already reaching the states 1 and 2.
+A clear example is proposed by Bürkner and Vuorre (2019) where the marriage duration in years is predicted as a function of some explanatory variables.
+For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential regression.
+When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Gerhard Tutz \& Berger, 2020).
+The adjacent category model compare the category \(k\) with \(k + 1\) still assuming a latent distribution for each \(k\).
+As suggested by Gerhard Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
+Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.
+
+In the current paper we put the focus on the cumulative model for several reasons.
+The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation.
+The second reason is that several psychological variables can be formalized as a latent continuous construct that when measured is collected using ordinal items.
+Furthermore, also the signal detection theory framework models the decision process assuming a latent distribution for the signal and noise where thresholds are the response criteria (e.g., DeCarlo, 2010).
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{img/fig-ordinal-models}
+
+}
+
+\caption{Theoretical figure about ordinal models. Voglio fare una figura come questa in Bürkner and Vuorre (2019) ma con qualche modifica per fare vedere i tre modelli.}\label{fig:fig-ordinal-models}
+\end{figure}
+
+\normalsize
+
+\subsection{Model notation}\label{model-notation}
+
+metti qualcosa sulla nomenclatura
 
@@
 
-Da citare in qualche modo Kemp and Grace (2021)
-
-\begin{quote}
-``As a matter of fact, most of the scales used widely and effectively by psychologists are ordinal scales.
-In the strictest propriety the ordinary statistics involving means and standard deviations ought not to be used with these scales, for these statistics imply a knowledge of something more than the relative rank-order of data.
-On the other hand for this `illegal' statisticizing there can be invoked a kind of pragmatic sanction: In numerous instances it leads to fruitful results.'' (Stevens, 1946, p.~679.)
-\end{quote}
-
-\begin{quote}
-It is straightforward to construct ordinal scales that do not involve rank ordering.
-For example, one can take the first element encountered and arbitrarily assign it the number 100.
-If the next element encountered is smaller it is given a smaller arbitrary number, 53 say.
-If the third element is between these two, it can be given the number 86, and so on.
-If this construction method is used, the difference between the elements assigned the numbers 70 and 80 will not in any important sense be equal to the difference between the elements assigned 90 and 100, and the intervals between the numbers are not really interpretable.
-Note, too, that monotonic transformations of the scale essentially leave the mea- sure unaffected.
-\end{quote}
-
-interessante come Cliff (2016) descriva che la maggiorparte dellle domande di ricerca in psicologia siano in riferimento alla location di un costrutto e possano essere gestite in modo ordinale
-
-The use of ordinal data is widespread in Psychology.
-Usually items from questionnaires are created using Likert scales where a certain psychological traits is measured using an intuitive scale with 3-5 or more anchor points.
-These measure cannot be considered \emph{metric} measure on interval or ratio scales {[}stevens{]} but the categories are ordered.
-
-Other examples here
-
-Despite the usage of ordinal variables, statistical models made for these type of data are rarely used in Psychology.
-Liddell and Kruschke (2018) reported that the majority of published papers using likert-like measures used standard methods to analyze the data.
-In practical terms, they use \emph{metric} models where the response variable cannot be considered fully numeric (vedi se c'è un termine tipo scala a rapporti).
-
-altro sui modelli
-
-Theoretically, using a \emph{metric} for ordinal data is not appropriate but understanding the actual impact is not straightforward.
-Liddell and Kruschke (2018) did a comprehensive work about pitfalls of analzying ordinal data as metric.
-They showed that metric models produce higher type-1 and type-2 errors compared to the ordinal models.
-In particular, given the bounded nature of ordinal data, difference in the underlying latent distribution are not always captured by the metric model that simply estimate the mean of the ordinal variable.
-This is even more relevant when the underluing variance of the ordinal variables are not homogeneous.
-Furthermore they presented some situations where the metric model could be wrong in the opposite direction, finding an effect with the wrong sign (type-s error).
-
-Finally, they also demonstrated that even in the best condition where e.g.~comparing two groups the variance are equal, the ordinal model is more powerful than the metric model given the underestimation of the true effect size from the latter.
-
-\begin{quote}
-Some authors have argued that, despite the ordinal character of individual Likert items, averaged ordinal items can have an emergent property of an interval scale and so it is appropriate to apply metric methods to the averaged values (e.g., Carifio \& Perla, 2007, 2008).
-\end{quote}
-
-vedi come gestire questo
-
-anche questo è importante
-\textgreater{} Ordered-probit models typically assume that the thresholds (θ k) are the same across all groups because the thresholds are theoretically linked to the response measure, not to the predictor value.
-For example, when asked, ``How happy are you?'' with response options `1' = very unhappy, `2' = mildly unhappy, `3' = neutral, `4' = mildly happy, `5' = very happy,'' the latent thresholds between ordinal levels are assumed to be implicit in the phrasing of the question, regardless of other aspects of the respondent or situation.
-In other words, the thresholds are assumed to be part of the measurement procedure, not dependent on the value of the predictor or covariate.
-This can be technically referred to as a type of measurement invariance.
-
-Da citare come paper iniziale sui modelli (McCullagh, 1980).
-
-questo per la tassonomia dei diversi modelli ordinali (Gerhard Tutz, 2022)
-
-da vedere magari per una critica tipo Gomila sul binary model (Robitzsch, 2020)
-
-\subsection{Metriv vs ordinal models}\label{metriv-vs-ordinal-models}
-
-\begin{itemize}
-\tightlist
-\item
-  the main difference is that the metric model assign a number to each label of the discrete variable assuming that the distance is the same. (Liddell \& Kruschke, 2018)
-\end{itemize}
+In this section we introduce some notations for the cumulative model that is used through the paper and in the R code.
+The notation is as consistent as possible with the main textbooks and with the \texttt{ordinal} R package used in the tutorial.
+We define \(Y_k\) as the observed ordinal response with \(1, 2, \dots,k\) options and \(Y^\star\) is the corresponding latent variable.
+The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \alpha_2, \dots, \alpha_{k - 1}\).
+With \(g()\) we define the cumulative probability function assumed for the specific model.
+For example, when using a Gaussian distribution we are fitting a model with a \texttt{probit} link function thus \(f() = \Phi()\).
+With \(g^{-1}()\) we define the inverse of the cumulative probability function.
+The minus sign in \(\mathbf{X}\mathbf{\beta}\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).
+
+\begin{equation} 
+P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, 2, \dots, k
+\label{eq:prob-cum-model1}
+\end{equation}
+
+We can call the \(\mathbf{X}\mathbf{\beta}\) as the linear predictor \(\eta\). Equation (num) describe how the probability of a single outcome \(Y_k\).
+
+\begin{equation} 
+P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta)
+\label{eq:prob-cum-model2}
+\end{equation}
+
+The same model can be formalized using the latent formulation (see Equation). This is more similar to a standard linear regression.
+
+\begin{equation} 
+Y^\star_i = \eta + \epsilon_i
+\label{eq:latent-model}
+\end{equation}
+
+Where \(\epsilon\) is the error part and comes from the distribution that is assumed from the latent model.
+For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution.
+Then the observed ordinal value \(Y_i\) comes from \(Y^\star_i\) being between the thresholds \(\alpha_{k - 1} < Y^\star_i < \alpha_{k}\) with thresholds organized as \(- \infty \equiv \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k \equiv \infty\).
+
+In the basic version of the model, the thresholds \(\alpha\) are considered as fixed and do not vary as a function of predictors.
+The thresholds are part of the measurement procedure (Liddell \& Kruschke, 2018).
+In a more sophisticated version of the model called location-shift (Gerhard Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha\) can vary as a function of the predictors.
+In the following sections another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Gerhard Tutz, 2022) will be presented as a more flexible way to model latent distributions with heterogeneous variance (spiega meglio).
 
@@
 But in simulation this is useful to understand what the cumulative model is doing.
+
+\scriptsize
 
@@
 ##     Min      1Q  Median      3Q     Max 
-## -4.5222 -0.6765  0.0000  0.6741  4.3425 
+## -4.4772 -0.6765 -0.0005  0.6742  4.0796 
 ## 
@@
 ##              Estimate Std. Error t value Pr(>|t|)    
-## (Intercept) 0.001086   0.003159   0.344    0.731    
-## x1          0.998374   0.003167 315.209   <2e-16 ***
-## x2          0.502929   0.003163 159.004   <2e-16 ***
+## (Intercept) 0.0007926  0.0031780   0.249    0.803    
+## x1          1.0009911  0.0031859 314.192   <2e-16 ***
+## x2          0.4998991  0.0031800 157.203   <2e-16 ***
 ## ---
@@
 ## 
-## Residual standard error: 0.9989 on 99997 degrees of freedom
-## Multiple R-squared:  0.5549, Adjusted R-squared:  0.5549 
-## F-statistic: 6.233e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
+## Residual standard error: 1.005 on 99997 degrees of freedom
+## Multiple R-squared:  0.5514, Adjusted R-squared:  0.5513 
+## F-statistic: 6.144e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
 \end{verbatim}
@@
 ##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
-##  probit flexible  1e+05 -122644.47 245300.95 5(0)  4.75e-08 9.0e+00
+##  probit flexible  1e+05 -122909.88 245831.76 5(0)  2.04e-07 9.1e+00
 ## 
@@
 ##     x1     x2 
-## 0.9974 0.5013 
+## 0.9949 0.4965 
 ## 
@@
 ##     1|2     2|3     3|4     4|5 
-## -0.8473 -0.2529  0.2551  0.8449
+## -0.8360 -0.2568  0.2515  0.8421
 \end{verbatim}
@@
 ##  (Intercept)           x1           x2 
-## 0.001086866 0.999502295 0.503497031
+## 0.0007886378 0.9960398259 0.4974264662
 \end{verbatim}
+
+\normalsize
 
@@
 Using the function proposed by Kruschke (2015) (metti ref ad osf) and the equations presented in Kurz (2023) the \texttt{clm\_to\_ord()} function convert parameters fitted with the \texttt{clm} function into the corresponding parameters for the latent \(Y^\star\) variable considering the actual range of values (from 1 to \(k\)).
+
+\scriptsize
 
@@
 
+\normalsize
+
+\scriptsize
+
 \begin{Shaded}
@@
 ## $sigma
-## [1] 1.79423
+## [1] 1.78472
 ## 
 ## $b0
-## [1] 2.999998
+## [1] 2.993037
 ## 
 ## $beta
-## [1] 1.786346
+## [1] 1.797978
 ## 
 ## $alpha
-## [1] 1.500000 2.543349 3.447639 4.500000
+## [1] 1.500000 2.560652 3.446266 4.500000
 \end{verbatim}
 
+\normalsize
+
 \subsection{Gelman parametrization}\label{gelman-parametrization}
@@
 
-\subsection{Odds ratios and cumulative odds ratios}\label{odds-ratios-and-cumulative-odds-ratios}
-
-\begin{itemize}
-\tightlist
-\item
-  \url{https://online.stat.psu.edu/stat504/lesson/4/4.1}
-\end{itemize}
-
-In this section we introduce the main concept to understand the effects of a logit model.
+\scriptsize
+
+\normalsize
+
+\section{Interpreting parameters}\label{interpreting-parameters}
+
+\subsection{Odds and odds ratio}\label{odds-and-odds-ratio}
+
+una piccola spiegazione di odds e odds ratio
+
+\subsection{Proportional odds assumption}\label{proportional-odds-assumption}
+
+Following again the taxonomy by Gerhard Tutz (2022), each of the presented ordinal regression model has a basic version making the proportional odds assumption.
+There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds} models) and partially (\emph{partial proportional odds} models) (rivedi se la nomenclatura va bene).
+The proportional odds assumption is a crucial point when intepreting model parameters.
+Can be formalized as in Equation ().
+Basically if we use the \emph{logit} link function \(g()\) the \(\beta\)s are interpreted as standard odds ratio in the logistic regression.
+Given that we have \(k > 2\) alternatives there is the need of \(k - 1\) equations (as reported in Equation ()).
+The porportional odds is assuming that the regression coefficients \(\beta\) are independent from the thresholds \(\alpha\) thus the effect is the same regardless of the \(Y\) level.
+Thus \(\beta_1 = \beta_2 \dots = \beta_{k - 1}\).
+The Figure () depicts the proportional odds assumption for the \(k - 1\) logistic curves both for probabilities and linear predictors \(\eta\).
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-prop-odds-1}
+
+}
+
+\caption{caption here}\label{fig:fig-prop-odds}
+\end{figure}
+
+\normalsize
+
+The proportional odds assumption is convenient because regardless the number of \(k\), the \(\beta_j\) (\(j\) being the number of regression coefficients) effect is assumed to be the same.
+The model is more parsimonious compared to estimating \(k - 1\) coefficients for each \(\beta_j\) as in the multinomial regression or the non-proportional version of the ordinal regression.
+Clearly this can be a strong assumption in some conditions.
+There are several methods to test the proportional odds assumption (see Liu, He, Tu, \& Tang, 2023 for an overview).
+Gerhard Tutz and Berger (2020) suggested that using alternative as the location-shift or location-scale models, the assumption can be made (improving model parsimony) but still having a more flexible modeling framework.
+In the next section, we provide more details about the location-scale model in terms of parameters interpretation and data simulation.
 
@@
 
-basically given that the effect of \(\beta\) is constant the odds ratio is independent from the thresholds \(\alpha_j\) (Liu, He, Tu, \& Tang, 2023).
+basically given that the effect of \(\beta\) is constant the odds ratio is independent from the thresholds \(\alpha_j\) (Liu et al., 2023).
 In the current paper we are simulating data using a single set of \(\beta\)s thus the proportional odds assumption is true.
@@
 Let's simulate the effect of a binary predictor on ordinal scale 1-5:
+
+\scriptsize
 
@@
 ##            1         2         3         4         5
-## 1 0.19881313 0.1989910 0.1974955 0.1998745 0.2048259
-## 2 0.07912516 0.1070303 0.1513017 0.2359672 0.4265757
+## 1 0.20141151 0.2052163 0.2038036 0.1939345 0.1956341
+## 2 0.07031792 0.1001600 0.1492126 0.2324907 0.4478188
 \end{verbatim}
 
+\normalsize
 Basically the proportional odds suggest that:
@@
 \]
+
 Is the same regardless the level of the \(x\) predictor. Thus:
@@
 \]
+
+\scriptsize
 
@@
 ##         xa         xb 
-## -0.9791059 -0.9791059
+## -0.9995719 -0.9995719
 \end{verbatim}
+
+\normalsize
 
@@
 
+\scriptsize
+
 \begin{figure}
@@
 \end{figure}
+
+\normalsize
 
@@
 In fact, fitting \(k - 1\) binomial regressions can be considered also an alternative to fitting an ordinal regression (see Gelman et al., 2020) with more flexibility in parameters (there will be \(k - 1\) regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).
+
+\scriptsize
 
@@
 ##             Model 1 Model 2 Model 3
-## (Intercept)  1.098911 -0.000607 -1.096559
-## SE           0.015143  0.016352  0.021668
+## (Intercept)  1.1146  0.0128 -1.0869
+## SE           0.0151  0.0163  0.0216
 ##                                    
-## x             -5.0004   -4.9958   -5.0132
-## SE             0.0364    0.0484    0.0736
+## x           -5.0266 -5.0021 -5.0291
+## SE           0.0365  0.0482  0.0736
 ## 
 \end{verbatim}
+
+\normalsize
 
@@
 
+\scriptsize
+
+\normalsize
+
 The linear predictors can be seen in the Figure ref. The \(g\) in this case is the cumulative \emph{probit} function \(\Phi\).
+
+\scriptsize
 
@@
 
+\normalsize
+
 Now we can apply the inverse of the link function \(g^{-1} = \Phi^{-1}\) to calculate the corresponding cumulative probabilities.
+
+\scriptsize
+
+\normalsize
+
+\scriptsize
 
@@
 ##    p_y1  p_y2  p_y3  p_y4
-## 1  0.57 0.232 0.134 0.063
-## 2 0.128 0.194 0.262 0.415
-## 3   0.1 0.172 0.255 0.472
-## 4 0.043 0.105 0.207 0.645
+## 1 0.529 0.243  0.15 0.077
+## 2 0.242 0.248 0.252 0.258
+## 3 0.553 0.237 0.141 0.069
+## 4 0.166 0.218 0.264 0.352
 ## 5   ...   ...   ...   ...
-## 6 0.381 0.264 0.207 0.147
-## 7 0.459 0.257 0.177 0.106
-## 8 0.332 0.263 0.225  0.18
-## 9 0.439  0.26 0.185 0.116
+## 6  0.22 0.241 0.257 0.282
+## 7 0.357 0.264 0.216 0.163
+## 8 0.464 0.256 0.175 0.104
+## 9 0.173 0.221 0.263 0.342
 \end{verbatim}
+
+\normalsize
 
@@
 
+\normalsize
+
+\scriptsize
+
+\normalsize
+
 To check the simulation result we can increase the number of observations, fit the model using the \texttt{ordinal::clm()} function and assess the recovery of simulated parameters.
+
+\scriptsize
+
+\normalsize
+
+\scriptsize
+
+\normalsize
 
@@
 
+\scriptsize
+
+\normalsize
+
 The Figure (put crossref) depict the simulated \(Y^{*}\) and the corresponding ordinal value.
@@
 
+\scriptsize
+
 \begin{figure}
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-18-1}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-17-1}
 
@@
 
-\caption{ }\label{fig:unnamed-chunk-18}
+\caption{ }\label{fig:unnamed-chunk-17}
 \end{figure}
 
+\normalsize
+
+\scriptsize
+
+\normalsize
+
+\scriptsize
+
+\normalsize
+
 The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function.
+
+\scriptsize
 
@@
 
+\normalsize
+
 \subsection{Scale Effects}\label{scale-effects}
@@
 
+\scriptsize
+
+\normalsize
+
 si può citare anche questo Cox (1995) come riferimento per i modelli
@@
 
+\scriptsize
+
+\normalsize
+
 The table (put ref) reports the simulation results.
+
+\scriptsize
+
+\normalsize
 
@@
 In this way the \(\beta\) can be choose in a meaningful way.
+
+\scriptsize
 
@@
 
+\normalsize
+
 For example, the Figure \ref{fig:ex-categorical} depicts the effect of a categorical predictor on a 5-level \(y\) assuming uniform probabilities for the reference level and a mean difference on a standardized Normal distribution of 1.
@@
 
+\normalsize
+
+\scriptsize
+
 \begin{figure}
@@
 
+\bibitem[\citeproctext]{ref-Likert1932-xw}
+Likert, R. (1932). A technique for the measurement of attitudes. \emph{Archives of Psychology}, \emph{22}, 55. Retrieved from \url{https://psycnet.apa.org/record/1933-01885-001}
+
 \bibitem[\citeproctext]{ref-Liu2023-bp}
@@
 
+\bibitem[\citeproctext]{ref-Rigby2005-ko}
+Rigby, R.
+A., \& Stasinopoulos, D.
+M.
+(2005).
+Generalized additive models for location, scale and shape.
+\emph{Journal of the Royal Statistical Society.
+Series C, Applied Statistics}, \emph{54}, 507--554.
+\url{https://doi.org/10.1111/j.1467-9876.2005.00510.x}
+
 \bibitem[\citeproctext]{ref-Robitzsch2020-la}
@@
 
+\bibitem[\citeproctext]{ref-Stevens1946-te}
+Stevens, S. S. (1946). On the theory of scales of measurement. \emph{Science (New York, N.Y.)}, \emph{103}, 677--680. \url{https://doi.org/10.1126/science.103.2684.677}
+
+\bibitem[\citeproctext]{ref-Tutz1990-fe}
+Tutz, Gerhard.
+(1990).
+Sequential item response models with an ordered response.
+\emph{The British Journal of Mathematical and Statistical Psychology}, \emph{43}, 39--55.
+\url{https://doi.org/10.1111/j.2044-8317.1990.tb00925.x}
+
 \bibitem[\citeproctext]{ref-Tutz2022-dg}

### 2023-12-13 R/utils-paper.R

@@ -64,3 +64,11 @@ clm_table_old <- function(fit, truth = NULL, caption = NULL){
 th_names <- function(k){
   sprintf("%s|%s", 1:(k-1), 2:k)
 }
+
+plot_grid_common_legend <- function(...){
+  plts <- list(...)
+  legend <- cowplot::get_legend(plts[[1]] + theme(legend.position = "bottom", legend.title = element_blank()))
+  plts <- lapply(plts, function(p) p + theme(legend.position = "none"))
+  grid <- cowplot::plot_grid(plotlist = plts)
+  cowplot::plot_grid(grid, legend, nrow = 2, rel_heights = c(0.9, 0.1))
+}

### 2023-12-13 R/utils.R

@@ -337,3 +337,15 @@ clm_to_ord <- function(fit){
   out <- list(sigma = lats, b0 = b0, beta = betas, alpha = th_y)
   return(out)
 }
+
+get_th <- function(b0, b1){
+  -(b0/b1)
+}
+
+get_slope <- function(b1){
+  1/b1
+}
+
+get_b0_from_th <- function(th, b1){
+  -b1 * th
+}

### 2023-12-13 scripts/figures.R

@@ -1,0 +1,100 @@
+# Packages
+
+library(tidyverse)
+library(cowplot)
+library(distributional)
+library(ggdist)
+library(here)
+devtools::load_all()
+
+# Figure Ordinal Models from @Burkner2019-aw ------------------------------
+
+# k <- 4
+# dd <- data.frame(m = 0, s = 1)
+# dd$dist <- distributional::dist_normal(dd$m, dd$s)
+# alpha <- prob_to_alpha(c(0.1, 0.4, 0.4, 0.1), link = "probit")
+# 
+# 
+# cumulative <- ggplot(dd, aes(dist = dist, fill = after_stat(factor(findInterval(x, alpha) + 1)))) +
+#   stat_slab(orientation = "horizontal") +
+#   xlim(c(-4, 4)) +
+#   xlab(latex2exp::TeX("$Y^{*}$")) +
+#   theme(legend.position = "none") +
+#   geom_vline(xintercept = alpha) +
+#   annotate("label", x = alpha, y = 0, label = latex2exp::TeX(sprintf("$\\alpha_{%s}$", 1:(k - 1)))) +
+#   annotate("label", x = seq(-2, 2, length.out = 4), y = 1, label = sprintf("Y = %s", 1:k))
+# 
+# seq1 <- ggplot(dd, aes(dist = dist, fill = after_stat(factor(findInterval(x, alpha[1]) + 1)))) +
+#   stat_slab(orientation = "horizontal") +
+#   xlim(c(-4, 4)) +
+#   xlab(latex2exp::TeX("$Y^{*}_1$")) +
+#   cowplot::theme_nothing() +
+#   theme(axis.title.x = element_text(),
+#         legend.position = "none") +
+#   geom_vline(xintercept = alpha[1]) +
+#   annotate("label", x = alpha[1], y = 0, label = latex2exp::TeX(sprintf("$\\alpha_{%s}$", 1))) +
+#   annotate("label", x = c(-2, -0.5), y = 1, label = c("Y = 1", "Y > 1"))
+# 
+# seq2 <- ggplot(dd, aes(dist = dist, fill = after_stat(factor(findInterval(x, alpha[2]) + 1)))) +
+#   stat_slab(orientation = "horizontal") +
+#   xlim(c(-4, 4)) +
+#   xlab(latex2exp::TeX("$Y^{*}_2$")) +
+#   cowplot::theme_nothing() +
+#   theme(axis.title.x = element_text(),
[…]

### 2023-12-19 paper/paper-new.Rmd
updating

@@
     header_includes:
-      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline}
+      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline, bm}
 ---
@@
 library(viridis)
+library(ggeffects)
 devtools::load_all()
@@
 rfigs <- readRDS(here("objects", "r-figs.rds"))
-```
-
-```{r tables, include=FALSE}
-mtab <- function(data){
-
-}
 ```
@@
 With $g()$ we define the cumulative probability function assumed for the specific model.
-For example, when using a Gaussian distribution we are fitting a model with a `probit` link function thus $f() = \Phi()$.
+For example, when using a Gaussian distribution we are fitting a model with a `probit` link function thus $g() = \Phi()$.
 With $g^{-1}()$ we define the inverse of the cumulative probability function.
-The minus sign in $\mathbf{X}\mathbf{\beta}$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).
+The minus sign in $\mathbf{X} \beta$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).
 
 \begin{equation} 
-P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, 2, \dots, k
+P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, \dots, k - 1
 (\#eq:prob-cum-model1)
@@
 
-We can call the $\mathbf{X}\mathbf{\beta}$ as the linear predictor $\eta$. Equation (num) describe how the probability of a single outcome $Y_k$.
+We can call the $\mathbf{X} \beta$ as the linear predictor $\eta$. Equation (num) describe how the probability of a single outcome $Y_k$.
 
@@
 For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution.
-Then the observed ordinal value $Y_i$ comes from $Y^\star_i$ being between the thresholds $\alpha_{k - 1} < Y^\star_i < \alpha_{k}$ with thresholds organized as $- \infty \equiv \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k \equiv \infty$.
+Then the observed ordinal value $Y_i$ comes from $Y^\star_i$ being between the thresholds $\alpha_{k - 1} < Y^\star_i < \alpha_{k}$ with thresholds organized as $- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty$.
 <!-- TODO rivedi questa -->
@@
 
-## lm on latent vs ordinal
+## Alternative parametrizations
 
-as suggested https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf, running a lm on the latent variables gives similar parameter as the clm.
-Of course, we are able to do this with real data given that the ordinal variable is the observed version of an unobserved latent variable.
-But in simulation this is useful to understand what the cumulative model is doing.
+Several parametrizations of the ordinal models have been proposed by different authors.
+@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian probit model where the latent mean and standard deviations are estimated along with the threshold.
+The interesting part is that the thresholds are transformed to be on the scale of the variable (e.g, from 1 to 4 when $k = 4$) and not in terms of $z$ scores.
+@Kurz2023-zv explain how to convert a model fitted with the `brms` package (citation) (an R package for regression modeling using *stan*) into the Kruschke [-@Kruschke2015-re] parametrization.
 
-```{r}
-dat <- data.frame(
-  x1 = rnorm(1e5),
-  x2 = rnorm(1e5)
-)
+@Gelman2020-tg proposed three alternative parametrizations focusing on different definition of the thresholds and.
 
-dat <- sim_ord_latent(~x1 + x2, By = c(1, 0.5), probs = rep(1/5, 5), link = "probit", data = dat)
+vedi qualcosa in più qui
 
-fit <- lm(ys ~ x1 + x2, data = dat)
-summary(fit)
+# Model fitting
 
-clm(y ~ x1 + x2, data = dat, link = "probit")
-coef(fit)/sigma(fit)
+For fitting the cumulative models we used the `ordinal` package [@Christensen2019-cz].
+Despite the presence of other possibilities the `ordinal` package provide the most complete and intuitive way to implement the ordinal models.
+The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.^[For a very complete overview of the ordinal package see @Christensen2019-cz and the Package documentation https://cran.r-project.org/web/packages/ordinal/ordinal.pdf].
+
+The function to fit the model is `clm()` and the model equation is specified using the R formula syntax `y ~ x` where `y` is the ordinal dependent variable and `x` is one or more predictors eventually including also the interactions.
+The package also implements mixed-effects models (see the `clmm()` function) including random intercepts and slopes but the topic is beyond the scope of the current tutorial.
+
+When fitting the model the crucial arguments are the `formula`, the `link` function and the `data`. More advanced arguments are the `nominal`, `scale` and `threshold`.
+
+- `formula`: the formula `y ~ x` with the dependent variable and predictors.
+- `link`: is the link function. In this tutorial we consider only the *logit* and *probit* link but other link functions are available.
+- `data`: is the dataset where with the variables included in the `formula`
+- `nominal`: formula with predictors where the proportional odds assumption (See Section ) is relaxed (i.e., partial or non proportional odds)
+- `scale`: formula with predictors for the scale (standard deviation) parameter.
+This argument allow to fit a scale-location model (see Section ).
+The main `formula` argument refers to predictors on the location parameter (i.e., the mean $\mu$).
+- `threshold`: different structures for estimating the thresholds. The default is `threshold = "flexible"` where $k - 1$ threshold (where $k$ is the number of ordinal levels for $Y$) are estimated.
+
+We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
+Table contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
+The participants are divided into two groups (predictor $x$).
+We can fit a cumulative link model with `clm()` function and check the model summary.
+
+```{r tab-dataset-example1}
+b1 <- log(3) # log odds ratio
+k <- 4
+n <- 100
+x <- rep(c("a", "b"), each = n/2)
+dat <- data.frame(x = x)
+probs <- rep(1/k, k) # for the group "a", uniform probabilities
+dat <- sim_ord_latent(~x, By = b1, prob = probs, data = dat, link = "logit")
+dat |> 
+  mutate(id = 1:n(), .before = x) |> 
+  select(-ys) |> 
+  filor::trim_df() |> 
+  mtab(type = "kable", caption = filor::get_caption())
 ```
 
-## Kruscke parametrization
-
-@Liddell2018-wu and @Kruschke2015-re proposed an alternative parametrization to understand the model parameters.
-They used a probit model where thresholds and regression parameters are estimated on the scale of the ordinal variable compared to standard ordinal regression where they refers to the quantile of the latent variables for the threshold and z score or odds ratio for the regression coefficients.
-They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.
-
-The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation).
-Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion.
-The main improvement regards mapping the values (for thresholds $\alpha_i$ and slopes $\beta$) from latent standard distribution (gaussian or logistic) into the scale of the $y$ ordinal value.
-The scale of the variable depend on the numeric labels assigned to ordered categories.
-There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to $\alpha_1 + 0.5$ and $\alpha_k - 0.5$ where other thresholds are estimated.
-
-@Kurz2023-zv explain how to convert a model fitted with the `brms` package (citation) (an R package for regression modeling using *stan*) into the Kruschke [-@Kruschke2015-re] parametrization.
-
-Using the function proposed by @Kruschke2015-re (metti ref ad osf) and the equations presented in @Kurz2023-zv the `clm_to_ord()` function convert parameters fitted with the `clm` function into the corresponding parameters for the latent $Y^\star$ variable considering the actual range of values (from 1 to $k$).
-
-```{r, echo=FALSE, results='asis'}
-filor::print_fun(funs$clm_to_ord)
+```{r eda-example-1}
+funs <- list(mean = mean, median = median, sd = sd)
+lapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
+prop.table(table(dat$y, dat$x), margin = 2)
 ```
 
-```{r}
-#| echo: true
-
-dat <- data.frame(
-  x = rnorm(1e5)
-)
-
-dat <- sim_ord_latent(~x, By = 1, probs = rep(1/5, 5), link = "probit", data = dat)
-fit <- clm(y ~ x, data = dat, link = "probit")
-clm_to_ord(fit)
+```{r model-example-1}
+fit <- clm(y ~ x, data = dat, link = "logit")
+summary(fit)
 ```
 
-## Gelman parametrization
-
-rivedi questa parametrizzazione in @Gelman2020-tg
-
-```{r}
-#| eval: false
-dat <- data.frame(x = runif(1e5, 0, 1))
-dat <- sim_ord_latent(~x, By = 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), data = dat, link = "logit")
-fit_clm <- clm(y ~ x, data = dat, link = "logit")
-
-num_latent_plot(dat$x, 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), link = "logit")
-
-summary(fit_clm)
-
-ths <- -(fit_clm$alpha/fit_clm$beta)
-betas <- 1/fit_clm$beta
-
-dat <- cbind(dat, dummy_ord(dat$y))
-
-fit <- glm(y1234vs5 ~ x, data = dat, family = binomial(link = "logit"))
-
-coefs <- coef(fit)
-
-x <- seq(0, 1, 0.01)
-
-plot(x, plogis(coefs[1] + coefs[2] * x), type = "l")
-
--(fit_clm$alpha/fit_clm$beta)
-1/fit_clm$beta
-
-```
+The two main sections of the model summary are the *Coefficients* section reporting the regression coefficients $\beta$ and the *Threshold* section reporting the $\alpha$ estimation.
+Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect.
+As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the intepretation depends on the contrast coding [for an overview about contrasts coding schemes see @Schad2020-ht]^[In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
+By default, the first level of the factor is taken as the reference level and the $j - 1$ coefficients represent the comparison between the other levels and the reference.]
 
@@
 
-## Odds and odds ratio
+## *Logit Model*. Odds and odds ratio
 
-una piccola spiegazione di odds e odds ratio
+aggiungere cosa significano i parametri nella regressione
+
+The odds of a probability $p$ is defined as $\frac{p}{1 - p}$ thus the probability of success divided by the probability of failure.
+The odds takes value ranging from 0 to $\infty$.
+For example with a probability of $p = 0.8$ we have an odds of $4$, thus we have a 4 successes for each failure.
+The same as having $p = 0.2$ and an odds of $0.25$ means that for each $0.25$ successes we have a failure or that we have $4$ failures for each success.
+When comparing two groups or conditions we can compare the two odds calculating an odds ratio.
+The odds ratio is the mostly used statistics to compare groups or conditions on a binary outcome.
+An odds ratio of $4$ means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator.
+The Figure \@ref(fig:fig-odds-example) shows the relationship between probabilities and odds.
+The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from $-\infty$ to $\infty$ with $p = 0.5$ as the midpoint because $\text{log}(\frac{0.5}{1 - 0.5}) = 0$.
+The standard logistic regression with two outcome model the logit transformed probability.
+
+```{r fig-odds-example, fig.cap=filor::get_caption()}
+p <- seq(0, 1, 0.001)
+
+dd <- data.frame(
+  p,
+  odds = filor::odds(p)
+)
+
+dd$lodds <- log(dd$odds)
+
+dd |> 
+  tidyr::pivot_longer(c(odds, lodds)) |> 
+  dplyr::mutate(name = ifelse(name == "odds", "Odds", "Log(Odds)")) |> 
+  ggplot(aes(x = p, y = value)) +
+  geom_line() +
+  facet_wrap(~name, scales = "free") +
+  theme(axis.title.y = element_blank()) +
+  xlab("Probability")
+```
+
+Odds and odds ratios are clearly defined with $k = 2$ outcomes.
+With an ordinal variable and a cumulative model we can use the cumulative odds ratio.
+With e.g.
+$k = 4$ outcomes we have $k - 1$ models determined by the cumulative probability in terms of $P(Y \leq 1), \dots, P(Y \leq k - 1)$.
 
@@
 There are more advanced versions of the model relaxing this assumption completely (*non proportional odds* models) and partially (*partial proportional odds* models) (rivedi se la nomenclatura va bene).
-The proportional odds assumption is a crucial point when intepreting model parameters.
+The proportional odds assumption is a crucial point when interpreting model parameters.
 Can be formalized as in Equation ().
@@
 
-To make a practical example, the dataset presented in Table contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
-The participants are divided into two groups (predictor $x$).
-We want to fit a cumulative ordinal model and check the proportional odds assumption.
+To make a practical example, we can use again the the dataset presented in Table. We can check the proportional odds by extrating fitted probabilities and manually computing the odds ratios.
 
@@
 
+Equation \@ref(eq:prop-odds) formalize the proportional odds assumption.
+Basically the cumulative log odds ratio comparing $P(Y_k|x_0)$ with $P(Y_k|x_1)$ is the same regardless the specific $k$ or threshold $\alpha$.
+
+\begin{equation}
+\text{logit} (\frac{P(Y \leq 1 |x_1)}{P(Y \leq 1 |x_0)}) = \dots = \text{logit} (\frac{P(Y \leq k - 1 |x_1)}{P(Y \leq k -1 |x_0)})
+(\#eq:prop-odds)
+\end{equation}
+
+This can be shown by fitting the model with `clm()` that by default assume the proportional odds and using the predicted probabilities to compute the odds ratios.
+In the following code we computed the odds ratio comparing $x_a$ and $x_b$ when $k \leq 1$ and $k \leq 2$.
+The log odds ratio is the same.
+
 ```{r}
-b1 <- log(3) # log odds ratio
-k <- 4
-n <- 100
-x <- rep(c("a", "b"), each = n/2)
-dat <- data.frame(x = x)
-probs <- rep(1/k, k) # for the group "a", uniform probabilities
-dat <- sim_ord_latent(~x, By = b1, probs = probs, data = dat, link = "logit")
+#| echo: true
+
+# fitting the model
 fit <- clm(y ~ x, data = dat, link = "logit")
+
+# extracting the predicted probabilities for the two groups
 pr <- predict(fit, data.frame(x = unique(x)))$fit
+
+odds <- function(p) p / (1 - p) # function to compute odds
+
+# group a
+a_or1vs2345 <- odds(pr[1, 1]) # 1 vs 2 3 4 5
+a_or12vs345 <- odds(sum(pr[1, 1:2])) # 1 vs 2 3 4 5
+
+# group b
+b_or1vs2345 <- odds(pr[2, 1]) # 1 vs 2 3 4 5
+b_or12vs345 <- odds(sum(pr[2, 1:2])) # 1 vs 2 3 4 5
+
+c(xa = log(a_or1vs2345 / a_or12vs345), xb = log(b_or1vs2345 / b_or12vs345))
 ```
 
-Basically the proportional odds suggest that:
+Clearly, the odds ratio here is exactly the same because the model assume the proportional odds assumption putting this constraint on the predicted probabilities.
+The estimated $\beta$ for each $k - 1$ equation could differ and the amount of heterogeneity is actually a way to test the proportional odds with real data.
+In the supplementary materials we implemented an intuitive plot proposed by @Harrell-2015-no to visually assess the proportional odds assumption.
+
+### Proportional odds and *probit* model
+
+The proportional odds assumption is relevant only for the *logit* model because of the link function that put $\beta$ as log odds.
+However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model.
+Equation \@ref(eq:probit-prop-odds) depict the parallel slopes assumption for a probit model.
+Basically the difference in $z$ for a unit increase in $x$ (i.e., the slope) is the same regardless the $Y$ level.
+This can be directly demonstrated in the same way as for the logit model.
+We use the same dataset presented in Table \@ref(tab:tab-prop-ord-ex) but fitting a probit model.
+The difference between the $z$ scores comparing the two groups calculated on the cumulative probability is the same regardless the $k - 1$ level.
 
 $$
-\text log(\frac{P(y \leq 1)}{P(y > 1)})
-$$
-
-Is the same regardless the level of the $x$ predictor. Thus:
-
-$$
-\text{log}\left(\frac{\frac{P(y \leq 1|x_0)}{P(y > 1|x_0)}}{\frac{P(y \leq 2|x_0)}{P(y > 2|x_0)}}\right) = \text{log}\left(\frac{\frac{P(y \leq 1|x_1)}{P(y > 1|x_1)}}{\frac{P(y \leq 2|x_1)}{P(y > 2|x_1)}}\right)
+z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
 $$
@@
 #| echo: true
-a_or1vs2345 <- filor::odds(pr[1, 1]) # 1 vs 2 3 4 5
-a_or12vs345 <- filor::odds(sum(pr[1, 1:2])) # 1 vs 2 3 4 5
 
-b_or1vs2345 <- filor::odds(pr[2, 1]) # 1 vs 2 3 4 5
-b_or12vs345 <- filor::odds(sum(pr[2, 1:2])) # 1 vs 2 3 4 5
+# fitting the model
+fit <- clm(y ~ x, data = dat, link = "probit")
 
-c(xa = log(a_or1vs2345 / a_or12vs345), xb = log(b_or1vs2345 / b_or12vs345))
+# extracting the predicted probabilities for the two groups
+pr <- predict(fit, data.frame(x = unique(x)))$fit
+
+# group a
+a_1vs2345 <- qnorm(pr[1, 1]) # 1 vs 2 3 4 5
+a_12vs345 <- qnorm(sum(pr[1, 1:2])) # 1 2 vs 3 4 5
+
+# group b
+b_1vs2345 <- qnorm(pr[2, 1]) # 1 vs 2 3 4 5
+b_12vs345 <- qnorm(sum(pr[2, 1:2])) # 1 2 vs 3 4 5
+
+c(xa = a_1vs2345 - a_12vs345, xb = b_1vs2345 - b_12vs345)
 ```
 
-Proportional odds can be also visualized by plotting the cumulative probabilities of $y$, in terms of $g(P(y \leq j))$ (where $g()$ is the logit link function) as a function of the predictor $x$.
-If the proportional odds assumptions holds the slopes are parallel (also known as parallel regression assumption).
-The Figure \@ref(fig:poa-plot-example) depicts the assumption of proportional odds in the probability and logit space.
+### Beyond the proportional odds
 
-```{r poa-plot-example, fig.cap="POA assumption"}
-k <- 5 # number of levels for y
-grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
-probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
-cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
-cprobs <- data.frame(do.call(rbind, cprobs))
-names(cprobs) <- sprintf("y<=%s", 1:k)
-cprobs <- cprobs[, -ncol(cprobs)] # remove the last
-grid <- cbind(grid, cprobs)
+qualcosa che spieghi come andare oltre e qualche proposta per testare l'assunzione.
 
-grid <- grid |> 
-  pivot_longer(starts_with("y")) |> 
-  mutate(lp = qlogis(value))
+## *Probit Model* $z$ scores
 
-poa_probs_plot <- grid |> 
-  ggplot(aes(x = x, y = value, color = name)) +
-  geom_line() +
-  theme_minimal(15) +
-  theme(legend.position = "bottom",
-        legend.title = element_blank(),
-        axis.title.y = element_blank()) +
-  ggtitle("Probability")
+The alternative to fitting a cumulative logit model is using a standard normal distribution thus fitting a *probit* model.
+The main difference is that the latent distribution is now a Gaussian distribution with $\mu = 0$ and $\sigma^2 = 1$ and the link function is the cumulative gaussian distribution function $\Phi$ implemented in R with the `pnorm()` function.
+The inverse of the Gaussian cumulative distribution function is $\Phi^{-1}$ implemented in R with the `qnorm()` function.
+The logistic and normal distribution are very similar, with the logistic having more variance.
+Figure \@ref(fig:fig-logit-vs-probit) depicts the differences between the two distributions.
 
-poa_lp_plot <- grid |> 
-  ggplot(aes(x = x, y = lp, color = name)) +
-  geom_line() +
-  theme_minimal(15) +
-  theme(axis.title.y = element_blank()) +
-  ggtitle("Linear Predictor")
+```{r fig-logit-vs-probit, fig.cap=filor::get_caption()}
+dplot <- ggplot(data = NULL) +
+  stat_function(fun = dnorm, aes(color = "Normal"), lwd = 1) +
+  stat_function(fun = dlogis, aes(color = "Logistic"), lwd = 1) +
+  xlim(c(-7, 7))
 
-bl <- cowplot::get_legend(
-  poa_probs_plot
+pplot <- ggplot(data = NULL) +
+  stat_function(fun = pnorm, aes(color = "Normal"), lwd = 1) +
+  stat_function(fun = plogis, aes(color = "Logistic"), lwd = 1) +
+  xlim(c(-7, 7))
+
+plot_grid_common_legend(
+  dplot, 
+  pplot + theme(axis.title.y = element_blank())
+)
+```
+
+The main difference regards the parameters interpretation.
+In the logit model the $\beta$ are the log odds ratio.
+For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in $x$.
+For the probit model, the $\beta$ is the increase in terms of $z$ scores for a unit increase in $x$.
+This is very convenient especially for categorical variables because parameters can be intepreted as a Cohen's $d$ like measure.
+Thinking about the latent distributions, the $\beta$ is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric $x$.
+The intepretation in terms of shifting the latent mean holds also for the logistic model.
+However, the standard deviation of the standard logistic regression is $\frac{\pi}{\sqrt{3}} \approx 1.81$^[Actually the variance of the logistic distribution is $\frac{s^2\pi^2}{3}$ and the standard deviation $\frac{s\pi}{\sqrt{3}}$ where $s$ is the scale of the distribution.
+For the standard logistic distribution $s = 1$ (as for the standard normal distribution).
+For the standard normal distribution $s^2 = \sigma^2 = 1$].
+The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\frac{\pi}{\sqrt{3}}$ standard deviations.
+
+Another advantage of the probit model is the possibility to directly map parameters from signal detection theory [@Stanislaw1999-jr; @Green1966-gy] into the model coefficients.
+In practical terms, the thresholds $\alpha$ are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].
+
+qualcosa in più qui
+
+The Table \@ref(tab:tab-model-summary) summarise the presented cumulative models with the proposed link function and the corresponding R code.
+
+```{r tab-model-summary}
+#| eval: false
+models_table <- data.frame(
+       stringsAsFactors = FALSE,
+                  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
+          Link.Function = c("Equation",
+                            "$\\text{logit}(p) = \\text{log}(\\frac{p}{1-p})$",
+                            "$z = \\Phi(p)$"),
+          Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
+  Inverse.Link.Function = c("Equation",
+                            "$\\frac{e^{\\text{logit}(p)}}{1 + e^{\\text{logit}(p)}}$","$\\Phi^{-1}(z)$"),
+  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
 )
 
-poa_plot <- cowplot::plot_grid(
-  poa_probs_plot + theme(legend.position = "none"),
-  poa_lp_plot + theme(legend.position = "none")
-)
-
-cowplot::plot_grid(poa_plot, bl, ncol = 1, rel_heights = c(0.9, 0.1))
+models_table[-1, ] |> 
+  flextable() |> 
+  set_header_labels(values = list(Model = "Model", 
+                                  Link.Function = "Equation",
+                                  Link.Function.1 = "R Code",
+                                  Inverse.Link.Function = "Equation",
+                                  Inverse.Link.Function.1 = "R Code")) |> 
+  add_header_row(values = c("", "Link Function", "Link Function", "Inverse Link Function", "Inverse Link Function")) |> 
+  merge_h(1, part = "header") |> 
+  theme_booktabs(bold_header = TRUE) |> 
+  align(part = "all", align = "center") |> 
+  set_caption(caption = filor::get_caption())
 ```
-
-@Harrell-2015-no proposed a very intuitive plot to assess the proportional odds assumption and eventually the degree of deviation from the ideal case.
-Basically predictor is plotted against the logit of the cumulative probability.
-Distances between pairs of symbols should be similar across levels of the predictors.
-Numerical predictors can be binned before plotting the corresponding logit.
-Figure \@ref(fig:harrel-poa-plot) depicts an example with simulated data satisfying the proportional odds assumption.
-
-```{r harrel-poa-plot, fig.cap="Harrel Poa Plot"}
-k <- 5 # number of levels for y
-grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
-probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
-cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
-cprobs <- data.frame(do.call(rbind, cprobs))
-names(cprobs) <- sprintf("y<=%s", 1:k)
-cprobs <- cprobs[, -ncol(cprobs)] # remove the last
-grid <- cbind(grid, cprobs)
-
-grid <- grid |> 
-  pivot_longer(starts_with("y")) |> 
-  mutate(lp = qlogis(value))
-
-grid |> 
-  ggplot(aes(x = lp, y = x)) +
-  geom_point(aes(color = name, 
-                 shape = name),
-             size = 3) +
-  theme_minimal(20) +
-  xlab("Logit") +
-  theme(legend.position = "bottom",
-        legend.title = element_blank())
-```
-
-From a statistical point of view, the proportional odds assumption can be assessed by fitting $k - 1$ binomial regressions and checking if the estimated $\beta$ is similar between regressions.
-The regressions are estimated by creating $k - 1$ dummy variables from the ordinal $y$.
-The code below show that simulating data with the proportional odds create $k - 1$ binomial regression with similar $\beta$s.
-In fact, fitting $k - 1$ binomial regressions can be considered also an alternative to fitting an ordinal regression [see @Gelman2020-tg] with more flexibility in parameters (there will be $k - 1$ regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).
-
-```{r}
-#| echo: true
-k <- 4
-n <- 1e5
-dat <- data.frame(
-  x = runif(n)
-)
-
-dat <- sim_ord_latent(~x, By = 5, probs = rep(1/k, k), data = dat, link = "logit")
-
-# create dummy variables
-
-dat$y1vs234 <- ifelse(dat$y <= 1, 1, 0)
-dat$y12vs34 <- ifelse(dat$y <= 2, 1, 0)
-dat$y123vs4 <- ifelse(dat$y <= 3, 1, 0)
-
-fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "logit"))
-fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "logit"))
-fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "logit"))
-
-car::compareCoefs(fit123vs4, fit12vs34, fit1vs234)
-```
-
-spiegare qui la differenza tra proportional and non proportional odds @Tutz2020-xq.
-
-## Logit vs Probit model
-
-When fitting an ordinal regression the two mostly used link functions are the *probit* and *logit*.
-From the distribution point of view the two functions are very similar.
-The *probit* model is based on a cumulative Gaussian distribution while the *logit* model is based on a logistic distribution.
-Figure \@ref(fig:logit-vs-probit) depict the two cumulative distributions.
-
-```{r logit-vs-probit, fig.cap="Logit vs Probit"}
-par(mfrow = c(1,2))
-
-# density
-
-curve(dnorm(x), -5, 5, lwd = 2, ylab = "Density", main = "Probability Density", cex.lab = 1.2, cex.axes = 1.2)
-curve(dlogis(x), -5, 5, lwd = 2, col = "firebrick", add = TRUE)
-
-# cdf
-
-curve(pnorm(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", lwd = 2, main = "Cumulative Probability",
-      cex.lab = 1.2, cex.axes = 1.2)
-curve(plogis(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", col = "firebrick", lwd = 2, add = TRUE)
-```
-
-Given the different underlying distribution, the parameters have a different interpretation under the two models.
-The probit model assume a latent standard normal distribution with and $\sigma = 1$.
-The logit model assume a logistic distribution with $\sigma^2 = \pi^2/3$.
-Thus regression coefficients $\beta$ represents the increase in standard deviations units.
-The interpretation in terms of latent distribution is particularly useful for the probit model where $\beta$s can be interpreted in a Cohen's $d$ like manner.
-Furthermore, there is the possibility to directly map parameters from signal detection theory [@Stanislaw1999-jr; @Green1966-gy] into an ordinal probit regression.
-In practical terms, the thresholds are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].
-
-The latent formulation of the ordinal regression model allow to think in standard linear regression terms for choosing and interpreting model parameters before converting into the probability space.
-Both the standard Normal and the logistic distribution are defined with a location ($\mu$) and a scale ($s$) parameter.
-For the Normal distribution, $\mu$ is the mean and scale is the variance ($\sigma^2$).
-For the logistic distribution the variance is $\sigma^2=\frac{s^2\pi^2}{3}$.
-Thus fixing $\mu$ and $\sigma^2 = 1$ (the default), the two distributions are similar with the logistic having more variance.
-For this reason, the latent formulation for parameter intepretation is particularly useful for the probit model because the $\beta$ is directly interpreted in standard deviaton units that by default are fixed to 1.
-With a binary predictor $x$, $\beta$ is the shift of the latent distribution increasing $x$ by one unit.
-This can be directly interpreted as a standardized mean difference effect size (e.g., Cohen's $d$).
-This is the same for the logistic distribution but a unit increase in $x$ shift the latent distribution by $\sigma = \sqrt{\frac{s^2 \pi^2}{3}} = \frac{s\pi}{\sqrt{3}}$.^[In the case of a standard logistic distribution ($s^2 = 1$), the standard deviation is $\frac{\pi}{\sqrt{3}}$].
 
@@
 
-### simulating using probabilities
+### Simulating from a multinomial distribution
 
-The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution.
+The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution^[In R we are using the `sample()` function that simulate values from a *categorical* distribution.
+The *categorical* distribution is considered a special case of the multinomial distribution].
 This is similar to the general method to simulate data for a generalized linear model where the linear predictors $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.
@@
 
-We can apply the same idea to an ordinal outcome but we need $k - 1$ equations where $k$ is the number of ordinal levels.
+Essentially the `lp` ($\eta_i$) vector contains true probability of success ($p_i|x_i = g^{-1}(\eta_i)$) for the $x_i$ level of the predictor.
+Then to introduce the random part of the model, we can use $p_i$ to generate binary values from a Binomial (in this specific case a Bernoulli) distribution.
+This create `y` that is the vector of 0-1 values generated using the true probabilities of success.
+The supplementary materials contains a more extensive example for simulating data for generalized linear models.
+
+We can apply the same idea to an ordinal outcome but we need $k - 1$ equations (compared to the previous example) where $k$ is the number of ordinal levels.
 Let's simulate a similar design with a continuous $x$ predictor and $k = 4$ options.
@@
 p(y_k) = 1/k$.
-The general workflow for the simulation is:
+The proposed workflow for the simulation is:
 
 1. Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
-2. Define the regression coefficients $\beta$s
+2. Define the regression coefficients $\beta$s.
 3. Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
@@
 Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities.
-$p(y = 1) = 0 - p(y <= 1), p(y = 2) = p(y <= 2) - p(y <= 1)$ (sistema qua).
-This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities.
+For example, the probability of $P(y = 2) = P(y \leq 2) - P(y \leq 1)$.
+This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities because the probability of the first $Y = 1$ level is the cumulative probability from $- \infty$ to the first threshold and the probability of $Y = k$ is obtained removing $P(Y \leq k - 1)$ from one.
 7. Sample $n$ outcomes from a multinomial distribution, choosing between $k$ alternatives with the calculated probabilities.
-
-(rivedi se fare con for)
-This can be easily implemented in R. We are using a `for` loop to improve the clarity of the workflow but using the `*apply` family of functions can improve the efficiency of the computations.
 
@@
 
-The linear predictors can be seen in the Figure ref. The $g$ in this case is the cumulative *probit* function $\Phi$.
+The linear predictors ($\eta$) can be seen in the Figure ref. The $g$ in this case is the cumulative *probit* function $\Phi$.
 
@@
 
-### simulating using latent distribution
+### Simulating from the latent distribution
 
@@
 This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and the cutting the latent values according to the thresholds $\alpha$.
-The workflow is slighlty different compared to the previous paragraph:
+The workflow is slightly different compared to the previous approach.
 
@@
 
+### Choosing parameters values
+
+#### Thresholds $\alpha$
+
+The two proposed simulation workflows can be really flexible for simulating complex models with multiple predictors and interactions.
+The critical part of the simulation is setting appropriate and empirically meaningful parameters values.
+The thresholds are usually not the crucial part of the statistical inference and model interpretations.
+In fact, they can be considered as the intercept in standard linear models.
+In any case to set meaningful $\alpha$ values the best strategy is to convert them into the corresponding probabilities according to the link function.
+The thresholds are essentially quantiles of the latent distribution (logistic or normal) without a direct empirical interpretation.
+The distance between the thresholds determine the probabililty assigned to a specific $Y$ value.
+The function `show_alpha()` allow to produce a meaningful plot starting from the thresholds or the probabilities.
+For example, the Figure \@ref(fig:fig-show-th-example) depict the threshold/probabilities associated with a four-level ordinal variable.
+
+```{r}
+#| eval: false
+show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))
+```
+
+```{r fig-show-th-example}
+p1 <- show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))
+p2 <- show_alpha(prob = c(0.5, 0.3, 0.1, 0.1), link = "probit", plot = c("latent", "probs"))
+p3 <- show_alpha(prob = c(0.1, 0.2, 0.5, 0.2), link = "probit", plot = c("latent", "probs"))
+plot_grid(p1 + rm_legend(), p2 + rm_legend(), p3, ncol = 1)
+```
+
+Thus the thresholds can be fixed according to the desired probabilities distribution.
+For example, when simulating the difference between two groups on an ordinal variable, the threshold can be set thinking about the response distribution in the reference group.
+The `alpha_to_prob()` and `prob_to_alpha()` functions can be used to convert between probabilities and thresholds.
+
+```{r}
+#| collapse: true
+# k = 4, thus k probabilities and k - 1 thresholds
+prob_to_alpha(prob = c(0.4, 0.1, 0.3, 0.2), link = "probit")
+alpha_to_prob(alpha = c(-2, 0, 1), link = "probit")
+```
+
+#### Regression coefficients $\beta$
+
+As reported in the sections above, the regression coefficients have a different interpretation according to the link function.
+Fortunately, both parameters have an intuitive interpretation given that they are directly expressed in a interpretable scale.
+For probit models we can think about standardized mean differences (e.g., Cohen's $d$) between categorical variables or standardized regression coefficient for a numerical predictor.
+For logit models we can fix the parameters thinking about the odds ratios.
+Meaningful odds ratios can be derived from previous literature, meta-analyses or converting to other effect sizes.
+For example, @Sanchez-Meca2003-ji proposed some equations to convert between odds ratios and Cohen's $d$.
+Using their approach, a Cohen's $d$ of 0.5 usually considered a plausible medium effect size corresponds to and odds ratio of $\approx 2.47$.
+Another more informative strategy is about plotting the predicted probabilities $g^{-1}(\eta)$ associated with certain regression coefficients.
+The `cat_latent_plot()` and `num_latent_plot()` functions can be used to see the impact of choosing a specific regression coefficients for respectively a categorical (Figures \@ref(fig:fig-example-cat-latent)) and numerical predictor (Figures \@ref(fig:fig-example-num-latent)).
+
+```{r fig-example-cat-latent, fig.cap=filor::get_caption()}
+#| echo: true
+cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")
+```
+
+```{r fig-example-num-latent, fig.cap=filor::get_caption()}
+#| echo: true
+num_latent_plot(x = runif(100), b1 = 2, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit")
+```
+
 ## Scale Effects
 
-The default ordinal regression model assume that the variance of th underlying latent distribution is the same across condition.
-This is similar to a standard linear regression assuming homogeneity of variance.
+The default ordinal regression model assume that the variance of the underlying latent distribution is the same across condition.
+This is similar to a standard linear regression assuming the homogeneity of variance.
 For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy].
@@
 This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
-Figure \@ref(fig:example-unequal-variance) depict an example of a comparison between two groups where the two underlyning latent distributions have unequal variance.
+The Equation (number here) expand the previous model including the linear predictor on the scale of the latent distribution.
 
-spiegare qui che i location-scale models sono sostanzialmente un cambiamento delle soglie (e non dei beta) in funzione di predittori @Tutz2020-xq
+\begin{equation}
+g(P(Y \leq k)) = \frac{\alpha_k - X\beta}{e^{X\zeta}}
+\end{equation}
 
-dire che gli scale effects possono essere visti come un'alterativa più parsimoniosa a modelli non-proportional odds @Christensen2019-cz
+\begin{equation} 
+Y^\star_i = \eta + \epsilon_i
+\epsilon_i \sim \mathcal{N}(0, e^{X\zeta})
+\end{equation}
 
-se citi brms per modelli più complicati, ricordati di specificare che:
+Where $X\zeta$ is the linear predictor $\eta$ for the scale of the distribution.
+By dafult for both the logit and probit model the scale is fixed to 1.
+On scale-location models we put predictors on both parameters.
+Given that the scale cannot be negative we use a log link function $\eta = \text{log}(X\zeta)$.
+Figure \@ref(fig:example-unequal-variance) depict an example of a comparison between two groups where the two underlying latent distributions have unequal variance.
+Ignoring the heterogeneity of variances could lead to underestimation of the effect size, reduced power and inflated type-1 error rate (qualche ref qui).
+Furtermore, as suggested by @Tutz2017-de location-scale models can be considered as a more parsimonius approach compared to partially or completely relaxing the proportional odds assumption.
+Allowing the scale to be different as a function of the predictor create more modelling flexbility.
+Furthemore, two groups could be theoretically different only in the scale of the latent distribution with a similar location.
+In this example, the only way to capture group differences is by including a scale effect.
+The Figure \@ref(fig:fig-scale-effect-example) depict the impact of having different scales between two groups on the ordinal probabilities.
 
-> That is, s = 1/disc. Further, because disc must be strictly positive, it is by default modeled on the log scale.
-
-un modo intuitivo per spiegare gli scale effects è di far capire la relazione tra slope e scale.
-La slope della funzione cumulative corrisponde alla varianza.
-Se la slope è piuù steep, la varianza sarà minore e la discrimizione maggiore (in `brms` si chiama desc)
+@Tutz2022-dg provide a very clear and intuitive explanation of what happen when including a scale effect and how to interpret the result.
+As suggested before, the scale-location model allow to independently predict changes in the location and the scale.
+While location shifts are simply interpreted as increasing/decreasing the latent $\mu$ or the odds of responding a certain category scale effects are not straightforward.
+As the scale increase (e.g., the variance increase) there is an higher probability mass on extreme categories.
+On the other side as the scale decrease, responses are more concentrated on single categories.
+The categories are determined by the location parameter.
+For example, if one group have a certain latent mean $\mu_1$ and a small scale $\sigma^2 = 1/3$ (thus one third compared to the standard version of the distribution), all responses will be focused on categories around the latent mean.
+On the other side, increasing the scale will increase the cumulative probabilities for all categories and for values that tends to infinity extreme categories are preferred.
+Clearly, the scale parameter can somehow be interpreted as the response style (concentrated or variable).
+This is conceptually similar but implemented and interpreted differently to shift-location models [@Tutz2022-dg; Tutz2020-xq] where thresholds are allowed to vary as a function of predictors.
+Both models tries to increase the flexibility of modelling the probability structure of the responses by adding extra paprameters to predict the probabilities of the ordinal responses.
 
 ```{r}
-#| eval: false
-library(distributional)
-library(ggdist)
-library(tidyverse)
-
-s <- c(0.5, 1, 2, 3)
-
-dat <- data.frame(
-  m = 0,
-  s = s
-)
-
-dat$dist <- dist_normal(dat$m, dat$s)
-
-par(mfrow = c(1, 2))
-curve(dnorm(x, m = dat$m[1], sd = dat$s[1]), -5, 5, lwd = 2)
-lapply(2:nrow(dat), function(i) curve(dnorm(x, dat$m[i], dat$s[i]), add = TRUE, col = i, lwd = 2))
-
-curve(pnorm(x, m = dat$m[1], sd = dat$s[1]), -5, 5, lwd = 2)
-lapply(2:nrow(dat), function(i) curve(pnorm(x, dat$m[i], dat$s[i]), add = TRUE, col = i, lwd = 2))
-
-# x <- seq(0, 1, 0.001)
-# p <- pnorm(qnorm(0.01) + 5 * x)
-# y <- rbinom(length(x), 1, p)
-# 
-# plot(x, p)
-# curve(pnorm(x, -(qnorm(0.01)/5), 1/5))
+sigma_high <- cat_latent_plot(m = c(0, 0), 
+                s = c(1, 2), 
+                probs = rep(1/4, 4), 
+                link = "probit", 
+                plot = "both",
+                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$"))
+sigma_low <- cat_latent_plot(m = c(0, 0), 
+                s = c(1, 0.5), 
+                probs = rep(1/4, 4), 
+                link = "probit", 
+                plot = "both",
+                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$"))
+sigma_high_mu1 <- cat_latent_plot(m = c(0, 1), 
+                s = c(1, 2), 
+                probs = rep(1/4, 4), 
+                link = "probit", 
+                plot = "both",
+                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$, $\\mu_2 = 1$"))
+sigma_low_mu1 <- cat_latent_plot(m = c(0, 1), 
+                s = c(1, 0.5), 
+                probs = rep(1/4, 4), 
+                link = "probit", 
+                plot = "both",
+                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$, $\\mu_2 = 1$"))
+plot_grid(sigma_high, sigma_low, sigma_high_mu1, sigma_low_mu1)
 ```
-
-si può citare anche questo @Cox1995-ur come riferimento per i modelli
-
-una distinzione tra location-shift and location-scale models @Tutz2017-de dove i primi oltre ad un cambiamento di location prevedono un cambiamento di soglie mentre i secondi modellano la differenza di soglie come un cambiamento della scala (varianza) che quindi permette di aumetare o diminuire l'area associata ad un determinato intervallo.
-
-sempre in @Tutz2017-de propongono il location-scale/shift come una via di mezzo tra il proportional odds ed il non proportional odds
 
@@
 dat <- data.frame(x = rep(c(0, 1), each = n/2))
-dat <- sim_ord_latent(~x, scale = ~x, By = b1, Bscale = z1, probs = probs0, data = dat, link = "probit")
+dat <- sim_ord_latent(~x, scale = ~x, By = b1, Bscale = z1, prob = probs0, data = dat, link = "probit")
 fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
@@
 
-## Visualizing effects
+To better understand the impact of assuming (or simulating) a different latent scale we fit $k - 1$ binomial regressions and check the estimated coefficients.
+We are not simulating a specific beta for each outcome but simulating a scale effect is actually impacting the regression coefficients.
+When generating data for a binary outcome the linear predictor is composed by $\eta = \beta_0 + \beta_1x$.
+The threshold $\alpha$ and slope of the function can be estimated using $\alpha = -\frac{\beta_0}{\beta_1}$ and the slope is $\frac{1}{\beta_1}$ [@Knoblauch2012-to; @Faraggi2003-hm].
+Under the proportional odds assumption, there is only a change in thresholds $\alpha$ this a shift in the sigmoid along the $x$ axis.
+When including a scale effect a change in the sigmoid is combined with a change in the slope.
 
-Before starting with the actual simulation it is useful to plot the predicted probabilities of the response variable $y$ as a function of the predictors.
-The `cat_latent_plot()` and the `num_latent_plot()` functions are able to visualize the effect of a single categorical or numerical predictor.
-Specifying the mean and standard deviations of the latent variable in each condition (for the categorical version) or the $\beta$ (for the numerical version) the function compute the expected probability for each level of the ordinal variable $y$ and visualize the result.
-In this way the $\beta$ can be choose in a meaningful way.
+```{r}
+set.seed(2023)
+k <- 4
+n <- 1e5
+b1 <- 3
+d1 <- log(2)
+dat <- data.frame(x = runif(n))
+dat <- sim_ord_latent(~x, ~x, By = b1, Bscale = d1, prob = rep(1/k, k), data = dat, link = "probit")
 
-```{r, echo=FALSE, results='asis'}
-filor::print_fun(c(funs$cat_latent_plot, funs$num_latent_plot))
+dat$y1vs234 <- ifelse(dat$y <= 1, 1, 0)
+dat$y12vs34 <- ifelse(dat$y <= 2, 1, 0)
+dat$y123vs4 <- ifelse(dat$y <= 3, 1, 0)
+
+dat$y <- ordered(dat$y)
+fit <- clm(y ~ x, scale = ~x, data = dat, link = "probit")
+
+fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "probit"))
+fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "probit"))
+fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "probit"))
+
+fits <- list(y1vs234 = fit1vs234, fit12vs34 = fit12vs34, fit123vs4 = fit123vs4)
+
+lapply(fits, function(x) coef(x)) |> 
+  bind_rows(.id = "model") |> 
+  mutate(xp = list(seq(0, 5, 0.01))) |> 
+  unnest(xp) |> 
+  ggplot(aes(x = xp, y = `(Intercept)` + x * xp)) +
+  geom_line(aes(color = model))
 ```
 
-For example, the Figure \@ref(fig:ex-categorical) depicts the effect of a categorical predictor on a 5-level $y$ assuming uniform probabilities for the reference level and a mean difference on a standardized Normal distribution of 1.
-The Figure \@ref(fig:ex-numerical) depicts the effect of a continuous predictor $x$ sampled from a standard Normal distribution assuming a $\beta = 1$.
+## 2x2 interaction
 
-```{r ex-categorical, fig.cap="Example categorical", fig.width=8, fig.height=8}
-cat_latent_plot(m = c(0, 1), s = 1, probs = rep(1/5, 5), link = "probit")
+A common research design could be a 2x2 factorial design.
+In this example we have 2 main effects and the interaction.
+Regardless the link function, we can focus the model equation on the linear predictor.
+Equation depicts the linear predictor for a 2x2 interaction.
+With a 2x2 design, a good strategy is to use sum to zero contrasts for the two factors.
+By default R use dummy coding but here we set the contrasts for $X_1$ and $X_2$ as $0.5$ and $-0.5$.
+In this way $\beta_1$ will be the main effect of $X_1$, $\beta_2$ the main effect of $X_2$ and $\beta_3$ the interaction (thus the difference of differences).
+
+$$
+g(P(Y \leq k))^{-1} = \alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i}
+$$
+
+```{r}
+n <- 1e5
+dat <- expand.grid(x1 = c("a", "b"), x2 = c("c", "d"), n = 1:n)
+dat$x1 <- factor(dat$x1)
+dat$x2 <- factor(dat$x2)
+contrasts(dat$x1) <- c(0.5, -0.5)
+contrasts(dat$x2) <- c(0.5, -0.5)
+bprobs <- rep(1/5, 5)
+
+dat <- sim_ord_latent(~x1*x2, By = c(0, 1, 0.5), prob = bprobs, link = "probit", data = dat)
+fit <- clm(y ~ x1 * x2, data = dat, link = "probit")
+summary(fit)
 ```
 
-```{r ex-numerical, fig.cap="Example numerical", fig.width=8, fig.height=8}
-num_latent_plot(x = rnorm(100), b1 = 1, probs = c(0.4, 0.2, 0.2, 0.05, 0.05), link = "probit")
+The parameters intepretation is the same as introduced in the general case.
+The thresholds $\alpha$ are fixed the determines the probabilities $P(Y = k)$ when all predictors are zero.
+In this case by doing `alpha_to_prob(fit$alpha, link = "probit")` we should recover the `bprobs` vector.
+`x1` is the main effect thus the difference in $z$ scores between *a* and *b* averaging over `x2`.
+The same holds for `x2`.
+`x1:x2` is the difference of differences in $z$ scores thus $(ad - ac) - (bd - bc)$.
+
+To visualize the effects there is the `ggeffect::ggpredict()` function that compute the predictions according to the specified predictors combination.
+In this case we can plot the interaction using the estimated probability for each group and $Y$ response.
+
+```{r}
+# terms =  c("x1", "x2") means to plot the interaction
+plot(ggpredict(fit, terms = c("x1", "x2")))
+```
+
+## Numerical by factor interaction
+
+Another common scenario is to compute and interaction between a numerical variable $x$ and a factor $g$.
+For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution $x \sim \mathcal{U}(0, 1)$.
+Again, the thresholds $\alpha$ represents the baseline probabilities when $x = 0$ and $g = 0$.
+
+```{r}
+n <- 1e5
+dat <- data.frame(x = runif(n),
+                  g = rep(c("a", "b"), each = n/2))
+dat$g <- factor(dat$g)
+contrasts(dat$g) <- c(0.5, -0.5)
+bprobs <- rep(1/5, 5)
+
+dat <- sim_ord_latent(~ x * g, By = c(0.3, 0.5, 0.5), prob = bprobs, link = "probit", data = dat)
+fit <- clm(y ~ x * g, data = dat, link = "probit")
+summary(fit)
+```
+
+Again, `x` is the slope of the numerical predictor averaging over `g` (given that `g` has been coded with sum to zero contrasts) thus the increase in $z$ scores for a unit increase in `x`.
+`g1` is the main effect of the factor evaluated when $x = 0$.
+To change the interpretation of `g1` a possibility is to center `x` differently (e.g., mean centering).
+The `x:g1` is the difference in slopes between the two groups.
+
+Regardless of the model complexity, the `sim_ord_latent()` function can be expanded by including more predictors.
+Clearly, as the model complexity increase, choosing meaningful and empirically plausible parameters became more challenging.
+
+# Power Analysis
+
+In the previous sections we introduced how to simulate standard and scale-location cumulative models and how to fit the corresponding models using the `clm()` function from the `ordinal` package.
+In this section we introduce how to compute the power for some common scenarios.
+We present an example in a very general form that can be expanded for specific use cases.
+For example, by simulating a model fixing all effects to zero it is possibile to estimate the type-1 error rate or fixing the sample size and varying the effect size it is possible to do a sensitivity analysis focusing on the smallest effect size of interest (ref qui).
+The general workflow for a power simulation can be summarized as:
+
+1. Choose the research design, e.g., 2x2 factorial design
+2. Indentifify the specific effect to focus on. For example the interaction effect, a main effect or a specific contrast
+3. Write the model equation similarly to what presented in the previous sections and extract the model coefficients
+4. Choose one of the proposed strategy for simulating data and follows the corresponding steps
+5. Fit the model and extract the relevant statistics e.g., the p-value of the interaction
+6. Repeat the simulation a large number of times (e.g, 10000) and store the relevant values from each simulation
+7.
+Summarise the simulation results in the appropriate way.
+For example, the power of the interaction is the number of times the p-value for the interaction test is lower than the critical level divided by the number of simulations.
+
+To make a practical example, we simulate the power of a cumulative link model in detecting an effect of $d = 0.4$ (probit model) with $k = 6$ and $n = 40$ participants per group.
+We fix the baseline probabilities in the first group as uniform.
+
+descrizione migliore
+
+```{r}
+n <- 40
+k <- 6
+d <- 0.4
+nsim <- 1e3 # higher is better, just for example
+bprobs <- rep(1/k, k)
+alpha <- 0.05 # critical alpha
+
+p <- rep(NA, nsim) # preallocation for speed
+dat <- data.frame(group = rep(c("a", "b"), each = n))
+
+for(i in 1:nsim){
+  sim <- sim_ord_latent(~group, By = d, prob = bprobs, link = "probit", data = dat)
+  fit <- clm(y ~ group, data = sim, link = "probit")
+  p[i] <- summary(fit)$coefficients["groupb", "Pr(>|z|)"] # extract the pvalue
+}
+```
+
+Then to compute the power we can simply compute the proportion of p values lower than the critical level.
+This can be done in R as `mean(p <= alpha)` resulting in an estimated power of `r mean(p <= alpha)`.
+Clearly using only a single condition is not really informative thus we can vary the sample size across plausible values and check the corresponding power curve.
+
+```{r}
+#| cache: true
+n <- c(10, 40, 60, 100, 200)
+power <- rep(NA, length(n))
+
+for(i in 1:length(n)){
+  p <- rep(NA, 0) # preallocation for speed
+  dat <- data.frame(group = rep(c("a", "b"), each = n[i]))
+  for(j in 1:nsim){
+    sim <- sim_ord_latent(~group, By = d, prob = bprobs, link = "probit", data = dat)
+    fit <- clm(y ~ group, data = sim, link = "probit")
+    p[j] <- summary(fit)$coefficients["groupb", "Pr(>|z|)"] # extract the pvalue
+  }
+  power[i] <- mean(p <= alpha)
+}
+```
+
+```{r}
+power_df <- data.frame(power = power, n = n)
+
+ggplot(power_df, aes(x = n, y = power)) +
+  geom_point(size = 3) +
+  geom_line() +
+  ylim(c(0, 1)) +
+  ylab("Power") +
+  xlab("Sample Size (per group)") +
+  geom_hline(yintercept = 0.8, lty = "dashed", color = "firebrick")
 ```
@@
 (ref:fig-prop-odds) caption here
+
+(ref:tab-prop-ord-ex) caption here
+
+(ref:fig-logit-vs-probit) caption here
+
+(ref:tab-model-summary) caption here
+
+(ref:fig-example-cat-latent) caption here
+
+(ref:fig-example-num-latent) caption here

### 2023-12-19 paper/paper-new.tex
updating

@@
 With \(g()\) we define the cumulative probability function assumed for the specific model.
-For example, when using a Gaussian distribution we are fitting a model with a \texttt{probit} link function thus \(f() = \Phi()\).
+For example, when using a Gaussian distribution we are fitting a model with a \texttt{probit} link function thus \(g() = \Phi()\).
 With \(g^{-1}()\) we define the inverse of the cumulative probability function.
-The minus sign in \(\mathbf{X}\mathbf{\beta}\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).
+The minus sign in \(\mathbf{X} \beta\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).
 
 \begin{equation} 
-P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, 2, \dots, k
+P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, \dots, k - 1
 \label{eq:prob-cum-model1}
@@
 
-We can call the \(\mathbf{X}\mathbf{\beta}\) as the linear predictor \(\eta\). Equation (num) describe how the probability of a single outcome \(Y_k\).
+We can call the \(\mathbf{X} \beta\) as the linear predictor \(\eta\). Equation (num) describe how the probability of a single outcome \(Y_k\).
 
@@
 For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution.
-Then the observed ordinal value \(Y_i\) comes from \(Y^\star_i\) being between the thresholds \(\alpha_{k - 1} < Y^\star_i < \alpha_{k}\) with thresholds organized as \(- \infty \equiv \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k \equiv \infty\).
+Then the observed ordinal value \(Y_i\) comes from \(Y^\star_i\) being between the thresholds \(\alpha_{k - 1} < Y^\star_i < \alpha_{k}\) with thresholds organized as \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).
 
@@
 
-\subsection{lm on latent vs ordinal}\label{lm-on-latent-vs-ordinal}
-
-as suggested \url{https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf}, running a lm on the latent variables gives similar parameter as the clm.
-Of course, we are able to do this with real data given that the ordinal variable is the observed version of an unobserved latent variable.
-But in simulation this is useful to understand what the cumulative model is doing.
-
-\scriptsize
+\subsection{Alternative parametrizations}\label{alternative-parametrizations}
+
+Several parametrizations of the ordinal models have been proposed by different authors.
+Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian probit model where the latent mean and standard deviations are estimated along with the threshold.
+The interesting part is that the thresholds are transformed to be on the scale of the variable (e.g, from 1 to 4 when \(k = 4\)) and not in terms of \(z\) scores.
+Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.
+
+Gelman, Hill, and Vehtari (2020) proposed three alternative parametrizations focusing on different definition of the thresholds and.
+
+vedi qualcosa in più qui
+
+\section{Model fitting}\label{model-fitting}
+
+For fitting the cumulative models we used the \texttt{ordinal} package (Christensen, 2019).
+Despite the presence of other possibilities the \texttt{ordinal} package provide the most complete and intuitive way to implement the ordinal models.
+The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}.
+
+The function to fit the model is \texttt{clm()} and the model equation is specified using the R formula syntax \texttt{y\ \textasciitilde{}\ x} where \texttt{y} is the ordinal dependent variable and \texttt{x} is one or more predictors eventually including also the interactions.
+The package also implements mixed-effects models (see the \texttt{clmm()} function) including random intercepts and slopes but the topic is beyond the scope of the current tutorial.
+
+When fitting the model the crucial arguments are the \texttt{formula}, the \texttt{link} function and the \texttt{data}.
+More advanced arguments are the \texttt{nominal}, \texttt{scale} and \texttt{threshold}.
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{formula}: the formula \texttt{y\ \textasciitilde{}\ x} with the dependent variable and predictors.
+\item
+  \texttt{link}: is the link function. In this tutorial we consider only the \emph{logit} and \emph{probit} link but other link functions are available.
+\item
+  \texttt{data}: is the dataset where with the variables included in the \texttt{formula}
+\item
+  \texttt{nominal}: formula with predictors where the proportional odds assumption (See Section ) is relaxed (i.e., partial or non proportional odds)
+\item
+  \texttt{scale}: formula with predictors for the scale (standard deviation) parameter.
+This argument allow to fit a scale-location model (see Section ).
+The main \texttt{formula} argument refers to predictors on the location parameter (i.e., the mean \(\mu\)).
+\item
+  \texttt{threshold}: different structures for estimating the thresholds.
+The default is \texttt{threshold\ =\ "flexible"} where \(k - 1\) threshold (where \(k\) is the number of ordinal levels for \(Y\)) are estimated.
+\end{itemize}
+
+We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
+Table contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options.
+The participants are divided into two groups (predictor \(x\)).
+We can fit a cumulative link model with \texttt{clm()} function and check the model summary.
+
+\scriptsize
+
+\begin{table}
+
+\caption{\label{tab:tab-dataset-example1}(ref:tab-dataset-example1)}
+\centering
+\begin{tabular}[t]{ccc}
+\toprule
+id & x & y\\
+\midrule
+1 & a & 4\\
+2 & a & 2\\
+3 & a & 1\\
+4 & a & 3\\
+... & ... & ...\\
+\addlinespace
+97 & b & 1\\
+98 & b & 3\\
+99 & b & 4\\
+100 & b & 4\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\normalsize
+
+\scriptsize
+
+\begin{verbatim}
+## $mean
+##    a    b 
+## 2.32 3.18 
+## 
+## $median
+## a b 
+## 2 3 
+## 
+## $sd
+##         a         b 
+## 1.0961901 0.9833305
+\end{verbatim}
 
@@
 ##    
-## Call:
-## lm(formula = ys ~ x1 + x2, data = dat)
+##        a    b
+##   1 0.30 0.10
+##   2 0.26 0.10
+##   3 0.26 0.32
+##   4 0.18 0.48
+\end{verbatim}
+
+\normalsize
+
+\scriptsize
+
+\begin{verbatim}
+## formula: y ~ x
+## data:    dat
 ## 
-## Residuals:
-##     Min      1Q  Median      3Q     Max 
-## -4.4772 -0.6765 -0.0005  0.6742  4.0796 
+##  link  threshold nobs logLik  AIC    niter max.grad cond.H 
+##  logit flexible  100  -127.63 263.27 4(0)  1.25e-08 1.6e+01
 ## 
 ## Coefficients:
-##              Estimate Std. Error t value Pr(>|t|)    
-## (Intercept) 0.0007926  0.0031780   0.249    0.803    
-## x1          1.0009911  0.0031859 314.192   <2e-16 ***
-## x2          0.4998991  0.0031800 157.203   <2e-16 ***
+##    Estimate Std. Error z value Pr(>|z|)    
+## xb   1.4974     0.3885   3.855 0.000116 ***
 ## ---
@@
 ## 
-## Residual standard error: 1.005 on 99997 degrees of freedom
-## Multiple R-squared:  0.5514, Adjusted R-squared:  0.5513 
-## F-statistic: 6.144e+04 on 2 and 99997 DF,  p-value: < 2.2e-16
+## Threshold coefficients:
+##     Estimate Std. Error z value
+## 1|2  -0.8124     0.2891  -2.810
+## 2|3   0.1867     0.2724   0.685
+## 3|4   1.5612     0.3206   4.869
 \end{verbatim}
 
-\begin{verbatim}
-## formula: y ~ x1 + x2
-## data:    dat
-## 
-##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
-##  probit flexible  1e+05 -122909.88 245831.76 5(0)  2.04e-07 9.1e+00
-## 
-## Coefficients:
-##     x1     x2 
-## 0.9949 0.4965 
-## 
-## Threshold coefficients:
-##     1|2     2|3     3|4     4|5 
-## -0.8360 -0.2568  0.2515  0.8421
-\end{verbatim}
-
-\begin{verbatim}
-##  (Intercept)           x1           x2 
-## 0.0007886378 0.9960398259 0.4974264662
-\end{verbatim}
-
-\normalsize
-
-\subsection{Kruscke parametrization}\label{kruscke-parametrization}
-
-Liddell and Kruschke (2018) and Kruschke (2015) proposed an alternative parametrization to understand the model parameters.
-They used a probit model where thresholds and regression parameters are estimated on the scale of the ordinal variable compared to standard ordinal regression where they refers to the quantile of the latent variables for the threshold and z score or odds ratio for the regression coefficients.
-They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.
-
-The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation).
-Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion.
-The main improvement regards mapping the values (for thresholds \(\alpha_i\) and slopes \(\beta\)) from latent standard distribution (gaussian or logistic) into the scale of the \(y\) ordinal value.
-The scale of the variable depend on the numeric labels assigned to ordered categories.
-There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to \(\alpha_1 + 0.5\) and \(\alpha_k - 0.5\) where other thresholds are estimated.
-
-Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.
-
-Using the function proposed by Kruschke (2015) (metti ref ad osf) and the equations presented in Kurz (2023) the \texttt{clm\_to\_ord()} function convert parameters fitted with the \texttt{clm} function into the corresponding parameters for the latent \(Y^\star\) variable considering the actual range of values (from 1 to \(k\)).
-
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{clm\_to\_ord }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(fit)\{}
-\NormalTok{  th }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(fit}\SpecialCharTok{$}\NormalTok{alpha) }\CommentTok{\# estimated thresholds}
-\NormalTok{  coefs }\OtherTok{\textless{}{-}}\NormalTok{ fit}\SpecialCharTok{$}\NormalTok{coefficients}
-\NormalTok{  coefs }\OtherTok{\textless{}{-}} \FunctionTok{unname}\NormalTok{(coefs[(}\FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(coefs)])}
-\NormalTok{  k }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(th) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# number of levels for y}
-\NormalTok{  y }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k }\CommentTok{\# ordinal values}
-\NormalTok{  stdm }\OtherTok{\textless{}{-}} \DecValTok{0} \CommentTok{\# latent mean}
-\NormalTok{  stds }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# latent sigma}
-\NormalTok{  th\_y }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(th, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{+} \FloatTok{0.5}\NormalTok{, y[k] }\SpecialCharTok{{-}} \FloatTok{0.5}\NormalTok{))}
-\NormalTok{  lats }\OtherTok{\textless{}{-}}\NormalTok{ (th\_y[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th\_y[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\SpecialCharTok{/}\NormalTok{ (th[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[k }\SpecialCharTok{{-}} \DecValTok{2}\NormalTok{]) }\CommentTok{\# real latent sigma}
-\NormalTok{  b0 }\OtherTok{\textless{}{-}}\NormalTok{ th\_y[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{{-}}\NormalTok{ th[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ lats}
-\NormalTok{  betas }\OtherTok{\textless{}{-}}\NormalTok{ lats }\SpecialCharTok{*}\NormalTok{ coefs}
-\NormalTok{  out }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}\AttributeTok{sigma =}\NormalTok{ lats, }\AttributeTok{b0 =}\NormalTok{ b0, }\AttributeTok{beta =}\NormalTok{ betas, }\AttributeTok{alpha =}\NormalTok{ th\_y)}
-  \FunctionTok{return}\NormalTok{(out)}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
-  \AttributeTok{x =} \FunctionTok{rnorm}\NormalTok{(}\FloatTok{1e5}\NormalTok{)}
-\NormalTok{)}
-
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \DecValTok{1}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{), }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
-\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
-\FunctionTok{clm\_to\_ord}\NormalTok{(fit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## $sigma
-## [1] 1.78472
-## 
-## $b0
-## [1] 2.993037
-## 
-## $beta
-## [1] 1.797978
-## 
-## $alpha
-## [1] 1.500000 2.560652 3.446266 4.500000
-\end{verbatim}
-
-\normalsize
-
-\subsection{Gelman parametrization}\label{gelman-parametrization}
-
-rivedi questa parametrizzazione in Gelman, Hill, and Vehtari (2020)
-
-\scriptsize
-
-\normalsize
+\normalsize
+
+The two main sections of the model summary are the \emph{Coefficients} section reporting the regression coefficients \(\beta\) and the \emph{Threshold} section reporting the \(\alpha\) estimation.
+Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect.
+As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the intepretation depends on the contrast coding (for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, \& Kliegl, 2020)\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
+By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}
 
@@
 
-\subsection{Odds and odds ratio}\label{odds-and-odds-ratio}
-
-una piccola spiegazione di odds e odds ratio
+\subsection{\texorpdfstring{\emph{Logit Model}. Odds and odds ratio}{Logit Model. Odds and odds ratio}}\label{logit-model.-odds-and-odds-ratio}
+
+aggiungere cosa significano i parametri nella regressione
+
+The odds of a probability \(p\) is defined as \(\frac{p}{1 - p}\) thus the probability of success divided by the probability of failure.
+The odds takes value ranging from 0 to \(\infty\).
+For example with a probability of \(p = 0.8\) we have an odds of \(4\), thus we have a 4 successes for each failure.
+The same as having \(p = 0.2\) and an odds of \(0.25\) means that for each \(0.25\) successes we have a failure or that we have \(4\) failures for each success.
+When comparing two groups or conditions we can compare the two odds calculating an odds ratio.
+The odds ratio is the mostly used statistics to compare groups or conditions on a binary outcome.
+An odds ratio of \(4\) means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator.
+The Figure \ref{fig:fig-odds-example} shows the relationship between probabilities and odds.
+The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from \(-\infty\) to \(\infty\) with \(p = 0.5\) as the midpoint because \(\text{log}(\frac{0.5}{1 - 0.5}) = 0\).
+The standard logistic regression with two outcome model the logit transformed probability.
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-odds-example-1}
+
+}
+
+\caption{(ref:fig-odds-example)}\label{fig:fig-odds-example}
+\end{figure}
+
+\normalsize
+
+Odds and odds ratios are clearly defined with \(k = 2\) outcomes.
+With an ordinal variable and a cumulative model we can use the cumulative odds ratio.
+With e.g.~\(k = 4\) outcomes we have \(k - 1\) models determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\).
 
@@
 There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds} models) and partially (\emph{partial proportional odds} models) (rivedi se la nomenclatura va bene).
-The proportional odds assumption is a crucial point when intepreting model parameters.
+The proportional odds assumption is a crucial point when interpreting model parameters.
 Can be formalized as in Equation ().
@@
 
-\subsection{Demonstrating the proportional odds assumption}\label{demonstrating-the-proportional-odds-assumption}
-
-basically given that the effect of \(\beta\) is constant the odds ratio is independent from the thresholds \(\alpha_j\) (Liu et al., 2023).
-In the current paper we are simulating data using a single set of \(\beta\)s thus the proportional odds assumption is true.
-Liu et al.
-(2023) reviewed the available methods to test this assumption.
-
-\begin{itemize}
-\tightlist
-\item
-  \url{https://hbiostat.org/ordinal/impactpo.pdf} blog su proportional odds, mi pare di capire che non sia così problematica come assunzione
-\end{itemize}
-
-sarebbe da capire quali sono i rischi di assumere questo ed eventualmente se simulare non proportional odds è troppo complicato.
-
-Let's simulate the effect of a binary predictor on ordinal scale 1-5:
+To make a practical example, we can use again the the dataset presented in Table. We can check the proportional odds by extrating fitted probabilities and manually computing the odds ratios.
+
+Equation \eqref{eq:prop-odds} formalize the proportional odds assumption.
+Basically the cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) is the same regardless the specific \(k\) or threshold \(\alpha\).
+
+\begin{equation}
+\text{logit} (\frac{P(Y \leq 1 |x_1)}{P(Y \leq 1 |x_0)}) = \dots = \text{logit} (\frac{P(Y \leq k - 1 |x_1)}{P(Y \leq k -1 |x_0)})
+\label{eq:prop-odds}
+\end{equation}
+
+This can be shown by fitting the model with \texttt{clm()} that by default assume the proportional odds and using the predicted probabilities to compute the odds ratios.
+In the following code we computed the odds ratio comparing \(x_a\) and \(x_b\) when \(k \leq 1\) and \(k \leq 2\).
+The log odds ratio is the same.
 
@@
 \begin{Highlighting}[]
-\NormalTok{b1 }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(}\DecValTok{3}\NormalTok{) }\CommentTok{\# log odds ratio}
-\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e4}
-\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x)}
-\NormalTok{probs }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{) }\CommentTok{\# for the group "a", uniform probabilities}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =}\NormalTok{ b1, }\AttributeTok{probs =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\CommentTok{\# fitting the model}
 \NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+
+\CommentTok{\# extracting the predicted probabilities for the two groups}
 \NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}
-\NormalTok{pr}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##            1         2         3         4         5
-## 1 0.20141151 0.2052163 0.2038036 0.1939345 0.1956341
-## 2 0.07031792 0.1001600 0.1492126 0.2324907 0.4478188
-\end{verbatim}
-
-\normalsize
-Basically the proportional odds suggest that:
-
-\[
-\text log(\frac{P(y \leq 1)}{P(y > 1)})
-\]
-
-Is the same regardless the level of the \(x\) predictor. Thus:
-
-\[
-\text{log}\left(\frac{\frac{P(y \leq 1|x_0)}{P(y > 1|x_0)}}{\frac{P(y \leq 2|x_0)}{P(y > 2|x_0)}}\right) = \text{log}\left(\frac{\frac{P(y \leq 1|x_1)}{P(y > 1|x_1)}}{\frac{P(y \leq 2|x_1)}{P(y > 2|x_1)}}\right)
-\]
-
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{a\_or1vs2345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
-\NormalTok{a\_or12vs345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}
-
-\NormalTok{b\_or1vs2345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
-\NormalTok{b\_or12vs345 }\OtherTok{\textless{}{-}}\NormalTok{ filor}\SpecialCharTok{::}\FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}
+
+\NormalTok{odds }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(p) p }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ p) }\CommentTok{\# function to compute odds}
+
+\CommentTok{\# group a}
+\NormalTok{a\_or1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
+\NormalTok{a\_or12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}
+
+\CommentTok{\# group b}
+\NormalTok{b\_or1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
+\NormalTok{b\_or12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}
 
@@
 ##         xa         xb 
-## -0.9995719 -0.9995719
+## -0.9991102 -0.9991102
 \end{verbatim}
@@
 
-Proportional odds can be also visualized by plotting the cumulative probabilities of \(y\), in terms of \(g(P(y \leq j))\) (where \(g()\) is the logit link function) as a function of the predictor \(x\).
-If the proportional odds assumptions holds the slopes are parallel (also known as parallel regression assumption).
-The Figure \ref{fig:poa-plot-example} depicts the assumption of proportional odds in the probability and logit space.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/poa-plot-example-1}
-
-}
-
-\caption{POA assumption}\label{fig:poa-plot-example}
-\end{figure}
-
-\normalsize
-
-Harrell (2015) proposed a very intuitive plot to assess the proportional odds assumption and eventually the degree of deviation from the ideal case.
-Basically predictor is plotted against the logit of the cumulative probability.
-Distances between pairs of symbols should be similar across levels of the predictors.
-Numerical predictors can be binned before plotting the corresponding logit.
-Figure \ref{fig:harrel-poa-plot} depicts an example with simulated data satisfying the proportional odds assumption.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/harrel-poa-plot-1}
-
-}
-
-\caption{Harrel Poa Plot}\label{fig:harrel-poa-plot}
-\end{figure}
-
-\normalsize
-
-From a statistical point of view, the proportional odds assumption can be assessed by fitting \(k - 1\) binomial regressions and checking if the estimated \(\beta\) is similar between regressions.
-The regressions are estimated by creating \(k - 1\) dummy variables from the ordinal \(y\).
-The code below show that simulating data with the proportional odds create \(k - 1\) binomial regression with similar \(\beta\)s.
-In fact, fitting \(k - 1\) binomial regressions can be considered also an alternative to fitting an ordinal regression (see Gelman et al., 2020) with more flexibility in parameters (there will be \(k - 1\) regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).
+Clearly, the odds ratio here is exactly the same because the model assume the proportional odds assumption putting this constraint on the predicted probabilities.
+The estimated \(\beta\) for each \(k - 1\) equation could differ and the amount of heterogeneity is actually a way to test the proportional odds with real data.
+In the supplementary materials we implemented an intuitive plot proposed by Harrell (2015) to visually assess the proportional odds assumption.
+
+\subsubsection{\texorpdfstring{Proportional odds and \emph{probit} model}{Proportional odds and probit model}}\label{proportional-odds-and-probit-model}
+
+The proportional odds assumption is relevant only for the \emph{logit} model because of the link function that put \(\beta\) as log odds.
+However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model.
+Equation \eqref{eq:probit-prop-odds} depict the parallel slopes assumption for a probit model.
+Basically the difference in \(z\) for a unit increase in \(x\) (i.e., the slope) is the same regardless the \(Y\) level.
+This can be directly demonstrated in the same way as for the logit model.
+We use the same dataset presented in Table \ref{tab:tab-prop-ord-ex} but fitting a probit model.
+The difference between the \(z\) scores comparing the two groups calculated on the cumulative probability is the same regardless the \(k - 1\) level.
+
+\[
+z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
+\]
 
@@
 \begin{Highlighting}[]
-\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{4}
-\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e5}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
-  \AttributeTok{x =} \FunctionTok{runif}\NormalTok{(n)}
-\NormalTok{)}
-
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \DecValTok{5}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k), }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
-
-\CommentTok{\# create dummy variables}
-
-\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y1vs234 }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y }\SpecialCharTok{\textless{}=} \DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
-\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y12vs34 }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y }\SpecialCharTok{\textless{}=} \DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
-\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y123vs4 }\OtherTok{\textless{}{-}} \FunctionTok{ifelse}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y }\SpecialCharTok{\textless{}=} \DecValTok{3}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{)}
-
-\NormalTok{fit1vs234 }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(y1vs234 }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link =} \StringTok{"logit"}\NormalTok{))}
-\NormalTok{fit12vs34 }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(y12vs34 }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link =} \StringTok{"logit"}\NormalTok{))}
-\NormalTok{fit123vs4 }\OtherTok{\textless{}{-}} \FunctionTok{glm}\NormalTok{(y123vs4 }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link =} \StringTok{"logit"}\NormalTok{))}
-
-\NormalTok{car}\SpecialCharTok{::}\FunctionTok{compareCoefs}\NormalTok{(fit123vs4, fit12vs34, fit1vs234)}
+\CommentTok{\# fitting the model}
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+
+\CommentTok{\# extracting the predicted probabilities for the two groups}
+\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}
+
+\CommentTok{\# group a}
+\NormalTok{a\_1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
+\NormalTok{a\_12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 2 vs 3 4 5}
+
+\CommentTok{\# group b}
+\NormalTok{b\_1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
+\NormalTok{b\_12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 2 vs 3 4 5}
+
+\FunctionTok{c}\NormalTok{(}\AttributeTok{xa =}\NormalTok{ a\_1vs2345 }\SpecialCharTok{{-}}\NormalTok{ a\_12vs345, }\AttributeTok{xb =}\NormalTok{ b\_1vs2345 }\SpecialCharTok{{-}}\NormalTok{ b\_12vs345)}
 \end{Highlighting}
@@
 \begin{verbatim}
-## Calls:
-## 1: glm(formula = y123vs4 ~ x, family = binomial(link = "logit"), data = dat)
-## 2: glm(formula = y12vs34 ~ x, family = binomial(link = "logit"), data = dat)
-## 3: glm(formula = y1vs234 ~ x, family = binomial(link = "logit"), data = dat)
-## 
-##             Model 1 Model 2 Model 3
-## (Intercept)  1.1146  0.0128 -1.0869
-## SE           0.0151  0.0163  0.0216
-##                                    
-## x           -5.0266 -5.0021 -5.0291
-## SE           0.0365  0.0482  0.0736
-## 
+##         xa         xb 
+## -0.5888457 -0.5888457
 \end{verbatim}
@@
 
-spiegare qui la differenza tra proportional and non proportional odds Gerhard Tutz and Berger (2020).
-
-\subsection{Logit vs Probit model}\label{logit-vs-probit-model}
-
-When fitting an ordinal regression the two mostly used link functions are the \emph{probit} and \emph{logit}.
-From the distribution point of view the two functions are very similar.
-The \emph{probit} model is based on a cumulative Gaussian distribution while the \emph{logit} model is based on a logistic distribution.
-Figure \ref{fig:logit-vs-probit} depict the two cumulative distributions.
+\subsubsection{Beyond the proportional odds}\label{beyond-the-proportional-odds}
+
+qualcosa che spieghi come andare oltre e qualche proposta per testare l'assunzione.
+
+\subsection{\texorpdfstring{\emph{Probit Model} \(z\) scores}{Probit Model z scores}}\label{probit-model-z-scores}
+
+The alternative to fitting a cumulative logit model is using a standard normal distribution thus fitting a \emph{probit} model.
+The main difference is that the latent distribution is now a Gaussian distribution with \(\mu = 0\) and \(\sigma^2 = 1\) and the link function is the cumulative gaussian distribution function \(\Phi\) implemented in R with the \texttt{pnorm()} function.
+The inverse of the Gaussian cumulative distribution function is \(\Phi^{-1}\) implemented in R with the \texttt{qnorm()} function.
+The logistic and normal distribution are very similar, with the logistic having more variance.
+Figure \ref{fig:fig-logit-vs-probit} depicts the differences between the two distributions.
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/logit-vs-probit-1}
-
-}
-
-\caption{Logit vs Probit}\label{fig:logit-vs-probit}
+{\centering \includegraphics{paper-new_files/figure-latex/fig-logit-vs-probit-1}
+
+}
+
+\caption{caption here}\label{fig:fig-logit-vs-probit}
 \end{figure}
@@
 
-Given the different underlying distribution, the parameters have a different interpretation under the two models.
-The probit model assume a latent standard normal distribution with and \(\sigma = 1\).
-The logit model assume a logistic distribution with \(\sigma^2 = \pi^2/3\).
-Thus regression coefficients \(\beta\) represents the increase in standard deviations units.
-The interpretation in terms of latent distribution is particularly useful for the probit model where \(\beta\)s can be interpreted in a Cohen's \(d\) like manner.
-Furthermore, there is the possibility to directly map parameters from signal detection theory (Green, Swets, \& Others, 1966; Stanislaw \& Todorov, 1999) into an ordinal probit regression.
-In practical terms, the thresholds are the criterion cut-offs and the \(\beta_1\) is the d\('\) parameter (DeCarlo, 1998; Knoblauch \& Maloney, 2012).
-
-The latent formulation of the ordinal regression model allow to think in standard linear regression terms for choosing and interpreting model parameters before converting into the probability space.
-Both the standard Normal and the logistic distribution are defined with a location (\(\mu\)) and a scale (\(s\)) parameter.
-For the Normal distribution, \(\mu\) is the mean and scale is the variance (\(\sigma^2\)).
-For the logistic distribution the variance is \(\sigma^2=\frac{s^2\pi^2}{3}\).
-Thus fixing \(\mu\) and \(\sigma^2 = 1\) (the default), the two distributions are similar with the logistic having more variance.
-For this reason, the latent formulation for parameter intepretation is particularly useful for the probit model because the \(\beta\) is directly interpreted in standard deviaton units that by default are fixed to 1.
-With a binary predictor \(x\), \(\beta\) is the shift of the latent distribution increasing \(x\) by one unit.
-This can be directly interpreted as a standardized mean difference effect size (e.g., Cohen's \(d\)).
-This is the same for the logistic distribution but a unit increase in \(x\) shift the latent distribution by \(\sigma = \sqrt{\frac{s^2 \pi^2}{3}} = \frac{s\pi}{\sqrt{3}}\).\footnote{In the case of a standard logistic distribution (\(s^2 = 1\)), the standard deviation is \(\frac{\pi}{\sqrt{3}}\)}.
+The main difference regards the parameters interpretation.
+In the logit model the \(\beta\) are the log odds ratio.
+For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in \(x\).
+For the probit model, the \(\beta\) is the increase in terms of \(z\) scores for a unit increase in \(x\).
+This is very convenient especially for categorical variables because parameters can be intepreted as a Cohen's \(d\) like measure.
+Thinking about the latent distributions, the \(\beta\) is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric \(x\).
+The intepretation in terms of shifting the latent mean holds also for the logistic model.
+However, the standard deviation of the standard logistic regression is \(\frac{\pi}{\sqrt{3}} \approx 1.81\)\footnote{Actually the variance of the logistic distribution is \(\frac{s^2\pi^2}{3}\) and the standard deviation \(\frac{s\pi}{\sqrt{3}}\) where \(s\) is the scale of the distribution.
+For the standard logistic distribution \(s = 1\) (as for the standard normal distribution).
+For the standard normal distribution \(s^2 = \sigma^2 = 1\)}.
+The \(\beta\) for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by \(\frac{\pi}{\sqrt{3}}\) standard deviations.
+
+Another advantage of the probit model is the possibility to directly map parameters from signal detection theory (Green, Swets, \& Others, 1966; Stanislaw \& Todorov, 1999) into the model coefficients.
+In practical terms, the thresholds \(\alpha\) are the criterion cut-offs and the \(\beta_1\) is the d\('\) parameter (DeCarlo, 1998; Knoblauch \& Maloney, 2012).
+
+qualcosa in più qui
+
+The Table \ref{tab:tab-model-summary} summarise the presented cumulative models with the proposed link function and the corresponding R code.
+
+\scriptsize
+
+\normalsize
 
@@
 
-\subsubsection{simulating using probabilities}\label{simulating-using-probabilities}
-
-The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution.
+\subsubsection{Simulating from a multinomial distribution}\label{simulating-from-a-multinomial-distribution}
+
+The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution\footnote{In R we are using the \texttt{sample()} function that simulate values from a \emph{categorical} distribution.
+The \emph{categorical} distribution is considered a special case of the multinomial distribution}.
 This is similar to the general method to simulate data for a generalized linear model where the linear predictors \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component.
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-9-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-9}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-4-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-4}
 \end{figure}
@@
 
-We can apply the same idea to an ordinal outcome but we need \(k - 1\) equations where \(k\) is the number of ordinal levels.
+Essentially the \texttt{lp} (\(\eta_i\)) vector contains true probability of success (\(p_i|x_i = g^{-1}(\eta_i)\)) for the \(x_i\) level of the predictor.
+Then to introduce the random part of the model, we can use \(p_i\) to generate binary values from a Binomial (in this specific case a Bernoulli) distribution.
+This create \texttt{y} that is the vector of 0-1 values generated using the true probabilities of success.
+The supplementary materials contains a more extensive example for simulating data for generalized linear models.
+
+We can apply the same idea to an ordinal outcome but we need \(k - 1\) equations (compared to the previous example) where \(k\) is the number of ordinal levels.
 Let's simulate a similar design with a continuous \(x\) predictor and \(k = 4\) options.
@@
 p(y_k) = 1/k\).
-The general workflow for the simulation is:
+The proposed workflow for the simulation is:
 
@@
 \item
-  Define the regression coefficients \(\beta\)s
+  Define the regression coefficients \(\beta\)s.
 \item
@@
   Calculate for each observation the probability of the \(k\) outcome as the difference between cumulative probabilities.
-\(p(y = 1) = 0 - p(y <= 1), p(y = 2) = p(y <= 2) - p(y <= 1)\) (sistema qua).
-This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities.
+For example, the probability of \(P(y = 2) = P(y \leq 2) - P(y \leq 1)\).
+This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities because the probability of the first \(Y = 1\) level is the cumulative probability from \(- \infty\) to the first threshold and the probability of \(Y = k\) is obtained removing \(P(Y \leq k - 1)\) from one.
 \item
@@
 
-(rivedi se fare con for)
-This can be easily implemented in R.
-We are using a \texttt{for} loop to improve the clarity of the workflow but using the \texttt{*apply} family of functions can improve the efficiency of the computations.
-
-\scriptsize
-
-\normalsize
-
-The linear predictors can be seen in the Figure ref. The \(g\) in this case is the cumulative \emph{probit} function \(\Phi\).
+\scriptsize
+
+\normalsize
+
+The linear predictors (\(\eta\)) can be seen in the Figure ref. The \(g\) in this case is the cumulative \emph{probit} function \(\Phi\).
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-11-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-11}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-6-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-6}
 \end{figure}
@@
 ##    p_y1  p_y2  p_y3  p_y4
-## 1 0.529 0.243  0.15 0.077
-## 2 0.242 0.248 0.252 0.258
-## 3 0.553 0.237 0.141 0.069
-## 4 0.166 0.218 0.264 0.352
+## 1 0.578  0.23 0.131 0.061
+## 2  0.15 0.209 0.264 0.377
+## 3 0.048 0.112 0.215 0.625
+## 4 0.429 0.261 0.189 0.121
 ## 5   ...   ...   ...   ...
-## 6  0.22 0.241 0.257 0.282
-## 7 0.357 0.264 0.216 0.163
-## 8 0.464 0.256 0.175 0.104
-## 9 0.173 0.221 0.263 0.342
+## 6 0.144 0.205 0.264 0.387
+## 7 0.198 0.233 0.261 0.308
+## 8 0.207 0.236 0.259 0.298
+## 9 0.107 0.178 0.257 0.457
 \end{verbatim}
@@
 
-\subsubsection{simulating using latent distribution}\label{simulating-using-latent-distribution}
+\subsubsection{Simulating from the latent distribution}\label{simulating-from-the-latent-distribution}
 
@@
 This require simulating a standard linear regression using the appropriate data generation function (\emph{logistic} or \emph{normal}) and the cutting the latent values according to the thresholds \(\alpha\).
-The workflow is slighlty different compared to the previous paragraph:
+The workflow is slightly different compared to the previous approach.
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-17-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-17}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-12-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-12}
 \end{figure}
@@
 \begin{Highlighting}[]
-\NormalTok{sim\_ord\_latent }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(location,}
-                           \AttributeTok{scale =} \ConstantTok{NULL}\NormalTok{,}
-\NormalTok{                           By,}
-                           \AttributeTok{Bscale =} \ConstantTok{NULL}\NormalTok{,}
-                           \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
-                           \AttributeTok{alphas =} \ConstantTok{NULL}\NormalTok{, }
-\NormalTok{                           data, }
-                           \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{))\{}
-
-  \CommentTok{\# get all the correct functions according to the link}
-\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(alphas))\{}
-    \CommentTok{\# calculate thresholds if not provided}
-\NormalTok{    alphas }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs, }\AttributeTok{link =}\NormalTok{ link)}
-\NormalTok{  \}}
-\NormalTok{  k }\OtherTok{\textless{}{-}} \FunctionTok{length}\NormalTok{(alphas) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# number of ordinal outcomes}
-\NormalTok{  n }\OtherTok{\textless{}{-}} \FunctionTok{nrow}\NormalTok{(data) }\CommentTok{\# number of observations}
-
-  \CommentTok{\# model matrix for the location effect}
-\NormalTok{  Xy }\OtherTok{\textless{}{-}} \FunctionTok{model.matrix}\NormalTok{(location, }\AttributeTok{data =}\NormalTok{ data)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, drop }\OtherTok{=} \ConstantTok{FALSE}\NormalTok{] }\CommentTok{\# remove intercept}
-\NormalTok{  lpy }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(Xy }\SpecialCharTok{\%*\%}\NormalTok{ By) }\CommentTok{\# linear predictor for the location}
-\NormalTok{  lps }\OtherTok{\textless{}{-}} \DecValTok{0} \CommentTok{\# default scale effect 0, exp(0) = 1 (the default scale)}
-
-  \CommentTok{\# check predictors on the scale parameter}
-  \ControlFlowTok{if}\NormalTok{(}\SpecialCharTok{!}\FunctionTok{is.null}\NormalTok{(scale))\{}
-    \CommentTok{\# model matrix for the scale effect}
-\NormalTok{    Xsigma }\OtherTok{\textless{}{-}} \FunctionTok{model.matrix}\NormalTok{(scale, }\AttributeTok{data =}\NormalTok{ data)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{, drop }\OtherTok{=} \ConstantTok{FALSE}\NormalTok{] }\CommentTok{\# remove intercept}
-\NormalTok{    lps }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(Xsigma }\SpecialCharTok{\%*\%}\NormalTok{ Bscale) }\CommentTok{\# linear predictor for the scale}
-\NormalTok{  \}}
-
-  \CommentTok{\# latent variable with appropriate error function}
-\NormalTok{  ystar }\OtherTok{\textless{}{-}}\NormalTok{ lf}\SpecialCharTok{$}\FunctionTok{rfun}\NormalTok{(n, lpy, }\FunctionTok{exp}\NormalTok{(lps))}
-
-  \CommentTok{\# cut according to thresholds}
-\NormalTok{  y }\OtherTok{\textless{}{-}} \FunctionTok{findInterval}\NormalTok{(ystar, alphas) }\SpecialCharTok{+} \DecValTok{1} \CommentTok{\# to start from 1}
-
-\NormalTok{  data}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{ordered}\NormalTok{(y) }\CommentTok{\# to ordered factor}
-\NormalTok{  data}\SpecialCharTok{$}\NormalTok{ys }\OtherTok{\textless{}{-}}\NormalTok{ ystar }\CommentTok{\# save also the latent}
-  \FunctionTok{return}\NormalTok{(data)}
-\NormalTok{\}}
+
 \end{Highlighting}
@@
 
+\subsubsection{Choosing parameters values}\label{choosing-parameters-values}
+
+\paragraph{\texorpdfstring{Thresholds \(\alpha\)}{Thresholds \textbackslash alpha}}\label{thresholds-alpha}
+
+The two proposed simulation workflows can be really flexible for simulating complex models with multiple predictors and interactions.
+The critical part of the simulation is setting appropriate and empirically meaningful parameters values.
+The thresholds are usually not the crucial part of the statistical inference and model interpretations.
+In fact, they can be considered as the intercept in standard linear models.
+In any case to set meaningful \(\alpha\) values the best strategy is to convert them into the corresponding probabilities according to the link function.
+The thresholds are essentially quantiles of the latent distribution (logistic or normal) without a direct empirical interpretation.
+The distance between the thresholds determine the probabililty assigned to a specific \(Y\) value.
+The function \texttt{show\_alpha()} allow to produce a meaningful plot starting from the thresholds or the probabilities.
+For example, the Figure \ref{fig:fig-show-th-example} depict the threshold/probabilities associated with a four-level ordinal variable.
+
+\scriptsize
+
+\normalsize
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-show-th-example-1}
+
+}
+
+\caption{ }\label{fig:fig-show-th-example}
+\end{figure}
+
+\normalsize
+
+Thus the thresholds can be fixed according to the desired probabilities distribution.
+For example, when simulating the difference between two groups on an ordinal variable, the threshold can be set thinking about the response distribution in the reference group.
+The \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} functions can be used to convert between probabilities and thresholds.
+
+\scriptsize
+
+\begin{verbatim}
+## [1] -0.2533471  0.0000000  0.8416212
+## [1] 0.02275013 0.47724987 0.34134475 0.15865525
+\end{verbatim}
+
+\normalsize
+
+\paragraph{\texorpdfstring{Regression coefficients \(\beta\)}{Regression coefficients \textbackslash beta}}\label{regression-coefficients-beta}
+
+As reported in the sections above, the regression coefficients have a different interpretation according to the link function.
+Fortunately, both parameters have an intuitive interpretation given that they are directly expressed in a interpretable scale.
+For probit models we can think about standardized mean differences (e.g., Cohen's \(d\)) between categorical variables or standardized regression coefficient for a numerical predictor.
+For logit models we can fix the parameters thinking about the odds ratios.
+Meaningful odds ratios can be derived from previous literature, meta-analyses or converting to other effect sizes.
+For example, Sánchez-Meca, Marín-Martínez, and Chacón-Moscoso (2003) proposed some equations to convert between odds ratios and Cohen's \(d\).
+Using their approach, a Cohen's \(d\) of 0.5 usually considered a plausible medium effect size corresponds to and odds ratio of \(\approx 2.47\).
+Another more informative strategy is about plotting the predicted probabilities \(g^{-1}(\eta)\) associated with certain regression coefficients.
+The \texttt{cat\_latent\_plot()} and \texttt{num\_latent\_plot()} functions can be used to see the impact of choosing a specific regression coefficients for respectively a categorical (Figures \ref{fig:fig-example-cat-latent}) and numerical predictor (Figures \ref{fig:fig-example-num-latent}).
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{cat\_latent\_plot}\NormalTok{(}\AttributeTok{m =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\AttributeTok{s =} \DecValTok{1}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{4}\NormalTok{, }\DecValTok{4}\NormalTok{), }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{, }\AttributeTok{plot =} \StringTok{"both"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-example-cat-latent-1}
+
+}
+
+\caption{caption here}\label{fig:fig-example-cat-latent}
+\end{figure}
+
+\normalsize
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{num\_latent\_plot}\NormalTok{(}\AttributeTok{x =} \FunctionTok{runif}\NormalTok{(}\DecValTok{100}\NormalTok{), }\AttributeTok{b1 =} \DecValTok{2}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.6}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.1}\NormalTok{, }\FloatTok{0.1}\NormalTok{), }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-example-num-latent-1}
+
+}
+
+\caption{caption here}\label{fig:fig-example-num-latent}
+\end{figure}
+
+\normalsize
+
 \subsection{Scale Effects}\label{scale-effects}
 
-The default ordinal regression model assume that the variance of th underlying latent distribution is the same across condition.
-This is similar to a standard linear regression assuming homogeneity of variance.
+The default ordinal regression model assume that the variance of the underlying latent distribution is the same across condition.
+This is similar to a standard linear regression assuming the homogeneity of variance.
 For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test (see Delacre, Lakens, \& Leys, 2017).
@@
 This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
-Figure \ref{fig:example-unequal-variance} depict an example of a comparison between two groups where the two underlyning latent distributions have unequal variance.
-
-spiegare qui che i location-scale models sono sostanzialmente un cambiamento delle soglie (e non dei beta) in funzione di predittori Gerhard Tutz and Berger (2020)
-
-dire che gli scale effects possono essere visti come un'alterativa più parsimoniosa a modelli non-proportional odds Christensen (2019)
-
-se citi brms per modelli più complicati, ricordati di specificare che:
-
-\begin{quote}
-That is, s = 1/disc. Further, because disc must be strictly positive, it is by default modeled on the log scale.
-\end{quote}
-
-un modo intuitivo per spiegare gli scale effects è di far capire la relazione tra slope e scale.
-La slope della funzione cumulative corrisponde alla varianza.
-Se la slope è piuù steep, la varianza sarà minore e la discrimizione maggiore (in \texttt{brms} si chiama desc)
-
-\scriptsize
-
-\normalsize
-
-si può citare anche questo Cox (1995) come riferimento per i modelli
-
-una distinzione tra location-shift and location-scale models G.
-Tutz and Berger (2017) dove i primi oltre ad un cambiamento di location prevedono un cambiamento di soglie mentre i secondi modellano la differenza di soglie come un cambiamento della scala (varianza) che quindi permette di aumetare o diminuire l'area associata ad un determinato intervallo.
-
-sempre in G. Tutz and Berger (2017) propongono il location-scale/shift come una via di mezzo tra il proportional odds ed il non proportional odds
+The Equation (number here) expand the previous model including the linear predictor on the scale of the latent distribution.
+
+\begin{equation}
+g(P(Y \leq k)) = \frac{\alpha_k - X\beta}{e^{X\zeta}}
+\end{equation}
+
+\begin{equation} 
+Y^\star_i = \eta + \epsilon_i
+\epsilon_i \sim \mathcal{N}(0, e^{X\zeta})
+\end{equation}
+
+Where \(X\zeta\) is the linear predictor \(\eta\) for the scale of the distribution.
+By dafult for both the logit and probit model the scale is fixed to 1.
+On scale-location models we put predictors on both parameters.
+Given that the scale cannot be negative we use a log link function \(\eta = \text{log}(X\zeta)\).
+Figure \ref{fig:example-unequal-variance} depict an example of a comparison between two groups where the two underlying latent distributions have unequal variance.
+Ignoring the heterogeneity of variances could lead to underestimation of the effect size, reduced power and inflated type-1 error rate (qualche ref qui).
+Furtermore, as suggested by G.
+Tutz and Berger (2017) location-scale models can be considered as a more parsimonius approach compared to partially or completely relaxing the proportional odds assumption.
+Allowing the scale to be different as a function of the predictor create more modelling flexbility.
+Furthemore, two groups could be theoretically different only in the scale of the latent distribution with a similar location.
+In this example, the only way to capture group differences is by including a scale effect.
+The Figure \ref{fig:fig-scale-effect-example} depict the impact of having different scales between two groups on the ordinal probabilities.
+
+Gerhard Tutz (2022) provide a very clear and intuitive explanation of what happen when including a scale effect and how to interpret the result.
+As suggested before, the scale-location model allow to independently predict changes in the location and the scale.
+While location shifts are simply interpreted as increasing/decreasing the latent \(\mu\) or the odds of responding a certain category scale effects are not straightforward.
+As the scale increase (e.g., the variance increase) there is an higher probability mass on extreme categories.
+On the other side as the scale decrease, responses are more concentrated on single categories.
+The categories are determined by the location parameter.
+For example, if one group have a certain latent mean \(\mu_1\) and a small scale \(\sigma^2 = 1/3\) (thus one third compared to the standard version of the distribution), all responses will be focused on categories around the latent mean.
+On the other side, increasing the scale will increase the cumulative probabilities for all categories and for values that tends to infinity extreme categories are preferred.
+Clearly, the scale parameter can somehow be interpreted as the response style (concentrated or variable).
+This is conceptually similar but implemented and interpreted differently to shift-location models {[}Gerhard Tutz (2022); Tutz2020-xq{]} where thresholds are allowed to vary as a function of predictors.
+Both models tries to increase the flexibility of modelling the probability structure of the responses by adding extra paprameters to predict the probabilities of the ordinal responses.
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-17-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-17}
+\end{figure}
+
+\normalsize
 
@@
 
-\subsection{Visualizing effects}\label{visualizing-effects}
-
-Before starting with the actual simulation it is useful to plot the predicted probabilities of the response variable \(y\) as a function of the predictors.
-The \texttt{cat\_latent\_plot()} and the \texttt{num\_latent\_plot()} functions are able to visualize the effect of a single categorical or numerical predictor.
-Specifying the mean and standard deviations of the latent variable in each condition (for the categorical version) or the \(\beta\) (for the numerical version) the function compute the expected probability for each level of the ordinal variable \(y\) and visualize the result.
-In this way the \(\beta\) can be choose in a meaningful way.
-
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{cat\_latent\_plot }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(}\AttributeTok{m =} \DecValTok{0}\NormalTok{, }
-                            \AttributeTok{s =} \DecValTok{1}\NormalTok{, }
-                            \AttributeTok{th =} \ConstantTok{NULL}\NormalTok{, }
-                            \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{, }
-                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{),}
-                            \AttributeTok{plot =} \FunctionTok{c}\NormalTok{(}\StringTok{"probs"}\NormalTok{, }\StringTok{"latent"}\NormalTok{, }\StringTok{"both"}\NormalTok{))\{}
-  \FunctionTok{require}\NormalTok{(ggplot2)}
-\NormalTok{  plot }\OtherTok{\textless{}{-}} \FunctionTok{match.arg}\NormalTok{(plot)}
-\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
-\NormalTok{  x }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"g"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(m))}
-\NormalTok{  lat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}
-    \AttributeTok{x =}\NormalTok{ x,}
-    \AttributeTok{m =}\NormalTok{ m,}
-    \AttributeTok{s =}\NormalTok{ s}
-\NormalTok{  )}
-
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(th))\{}
-\NormalTok{    th }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs, link)}
-\NormalTok{  \}}
-
-\NormalTok{  thl }\OtherTok{\textless{}{-}}\NormalTok{ latex2exp}\SpecialCharTok{::}\FunctionTok{TeX}\NormalTok{(}\FunctionTok{sprintf}\NormalTok{(}\StringTok{"$}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{alpha\_\{\%s\}$"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(th)))}
-
-  \ControlFlowTok{if}\NormalTok{(link }\SpecialCharTok{==} \StringTok{"logit"}\NormalTok{)\{}
-\NormalTok{    dfun }\OtherTok{\textless{}{-}}\NormalTok{ distributional}\SpecialCharTok{::}\NormalTok{dist\_logistic}
-\NormalTok{    s }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(}\FunctionTok{vlogit}\NormalTok{(s))}
-\NormalTok{    title }\OtherTok{\textless{}{-}} \StringTok{"Logistic Distribution"}
-\NormalTok{  \}}\ControlFlowTok{else}\NormalTok{\{}
-\NormalTok{    dfun }\OtherTok{\textless{}{-}}\NormalTok{ distributional}\SpecialCharTok{::}\NormalTok{dist\_normal}
-\NormalTok{    s }\OtherTok{\textless{}{-}}\NormalTok{ lat}\SpecialCharTok{$}\NormalTok{s}
-\NormalTok{    title }\OtherTok{\textless{}{-}} \StringTok{"Normal Distribution"}
-\NormalTok{  \}}
-\NormalTok{  lat}\SpecialCharTok{$}\NormalTok{dist }\OtherTok{\textless{}{-}} \FunctionTok{dfun}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m, lat}\SpecialCharTok{$}\NormalTok{s)}
-\NormalTok{  range\_y }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FunctionTok{min}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m) }\SpecialCharTok{{-}} \FunctionTok{max}\NormalTok{(s) }\SpecialCharTok{*} \DecValTok{5}\NormalTok{, }\FunctionTok{max}\NormalTok{(lat}\SpecialCharTok{$}\NormalTok{m) }\SpecialCharTok{+} \FunctionTok{max}\NormalTok{(s) }\SpecialCharTok{*} \DecValTok{5}\NormalTok{)}
-
-\NormalTok{  lat\_plot }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(lat,}
-         \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ m, }\AttributeTok{dist =}\NormalTok{ dist)) }\SpecialCharTok{+}
-\NormalTok{    ggdist}\SpecialCharTok{::}\FunctionTok{stat\_halfeye}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{fill =} \FunctionTok{after\_stat}\NormalTok{(}\FunctionTok{factor}\NormalTok{(}\FunctionTok{findInterval}\NormalTok{(y, th) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{))),}
-                 \AttributeTok{alpha =} \FloatTok{0.85}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{ylim}\NormalTok{(range\_y) }\SpecialCharTok{+}
-    \FunctionTok{geom\_hline}\NormalTok{(}\AttributeTok{yintercept =}\NormalTok{ th, }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{, }\AttributeTok{col =} \StringTok{"black"}\NormalTok{, }\AttributeTok{alpha =} \FloatTok{0.7}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ m, }\AttributeTok{group =} \DecValTok{1}\NormalTok{)) }\SpecialCharTok{+}
-    \FunctionTok{annotate}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\AttributeTok{x =} \FloatTok{0.7}\NormalTok{, }\AttributeTok{y =}\NormalTok{ th, }\AttributeTok{label =}\NormalTok{ thl, }\AttributeTok{size =} \DecValTok{5}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{theme\_minimal}\NormalTok{(}\DecValTok{15}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{,}
-          \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{(),}
-          \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
-    \FunctionTok{ylab}\NormalTok{(latex2exp}\SpecialCharTok{::}\FunctionTok{TeX}\NormalTok{(}\StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{mu"}\NormalTok{)) }\SpecialCharTok{+}
-    \FunctionTok{ggtitle}\NormalTok{(title)}
-
-\NormalTok{  th }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\SpecialCharTok{{-}}\ConstantTok{Inf}\NormalTok{, th, }\ConstantTok{Inf}\NormalTok{)}
-\NormalTok{  lat\_ms }\OtherTok{\textless{}{-}}\NormalTok{ lat[, }\FunctionTok{c}\NormalTok{(}\StringTok{"m"}\NormalTok{, }\StringTok{"s"}\NormalTok{)]}
-\NormalTok{  ps }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(lat\_ms, }\DecValTok{1}\NormalTok{, }\ControlFlowTok{function}\NormalTok{(x) }\FunctionTok{data.frame}\NormalTok{(}\FunctionTok{t}\NormalTok{(}\FunctionTok{diff}\NormalTok{(lf}\SpecialCharTok{$}\FunctionTok{pfun}\NormalTok{(th, x[}\DecValTok{1}\NormalTok{], x[}\DecValTok{2}\NormalTok{])))), }\AttributeTok{simplify =} \ConstantTok{FALSE}\NormalTok{)}
-\NormalTok{  ps }\OtherTok{\textless{}{-}} \FunctionTok{do.call}\NormalTok{(rbind, ps)}
-  \FunctionTok{names}\NormalTok{(ps) }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"y"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\FunctionTok{ncol}\NormalTok{(ps))}
-\NormalTok{  latl }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(lat, ps)}
-\NormalTok{  latl }\OtherTok{\textless{}{-}}\NormalTok{ tidyr}\SpecialCharTok{::}\FunctionTok{pivot\_longer}\NormalTok{(latl, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}
-\NormalTok{  probs\_plot }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(latl, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =} \FunctionTok{factor}\NormalTok{(x), }\AttributeTok{y =}\NormalTok{ value, }\AttributeTok{fill =}\NormalTok{ y)) }\SpecialCharTok{+}
-    \FunctionTok{geom\_col}\NormalTok{(}\AttributeTok{position =} \FunctionTok{position\_dodge}\NormalTok{()) }\SpecialCharTok{+}
-    \FunctionTok{theme\_minimal}\NormalTok{(}\DecValTok{15}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{,}
-          \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{(),}
-          \AttributeTok{axis.title.x =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
-    \FunctionTok{ylab}\NormalTok{(}\StringTok{"Probability"}\NormalTok{)}
-  \ControlFlowTok{if}\NormalTok{(plot }\SpecialCharTok{==} \StringTok{"probs"}\NormalTok{)\{}
-\NormalTok{    probs\_plot}
-\NormalTok{  \}}\ControlFlowTok{else} \ControlFlowTok{if}\NormalTok{(plot }\SpecialCharTok{==} \StringTok{"latent"}\NormalTok{)\{}
-\NormalTok{    lat\_plot}
-\NormalTok{  \}}\ControlFlowTok{else}\NormalTok{\{}
-\NormalTok{    legend\_b }\OtherTok{\textless{}{-}}\NormalTok{ cowplot}\SpecialCharTok{::}\FunctionTok{get\_legend}\NormalTok{(}
-\NormalTok{      probs\_plot}
-\NormalTok{    )}
-\NormalTok{    plts }\OtherTok{\textless{}{-}}\NormalTok{ cowplot}\SpecialCharTok{::}\FunctionTok{plot\_grid}\NormalTok{(}
-\NormalTok{      lat\_plot }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{),}
-\NormalTok{      probs\_plot }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{)}
-\NormalTok{    )}
-\NormalTok{    cowplot}\SpecialCharTok{::}\FunctionTok{plot\_grid}\NormalTok{(plts, legend\_b, }\AttributeTok{ncol =} \DecValTok{1}\NormalTok{, }\AttributeTok{rel\_heights =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{))}
-\NormalTok{  \}}
-\NormalTok{\}}
-\NormalTok{num\_latent\_plot }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, }
-\NormalTok{                            b1, }
-                            \AttributeTok{th =} \ConstantTok{NULL}\NormalTok{, }
-                            \AttributeTok{probs =} \ConstantTok{NULL}\NormalTok{,}
-                            \AttributeTok{link =} \FunctionTok{c}\NormalTok{(}\StringTok{"logit"}\NormalTok{, }\StringTok{"probit"}\NormalTok{), }
-                            \AttributeTok{nsample =} \FloatTok{1e3}\NormalTok{, }
-                            \AttributeTok{size =} \DecValTok{20}\NormalTok{)\{}
-\NormalTok{  lf }\OtherTok{\textless{}{-}} \FunctionTok{get\_link}\NormalTok{(link)}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{is.null}\NormalTok{(th))\{}
-\NormalTok{    th }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs, link)}
-\NormalTok{  \}}
-\NormalTok{  data }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{seq}\NormalTok{(}\FunctionTok{min}\NormalTok{(x), }\FunctionTok{max}\NormalTok{(x), }\AttributeTok{length.out =}\NormalTok{ nsample))}
-\NormalTok{  data }\OtherTok{\textless{}{-}} \FunctionTok{get\_probs}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, b1, probs, data, link, }\AttributeTok{append =} \ConstantTok{TRUE}\NormalTok{)}
-\NormalTok{  datal }\OtherTok{\textless{}{-}}\NormalTok{ tidyr}\SpecialCharTok{::}\FunctionTok{pivot\_longer}\NormalTok{(data, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"y"}\NormalTok{), }\AttributeTok{names\_to =} \StringTok{"y"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"value"}\NormalTok{)}
-  \FunctionTok{ggplot}\NormalTok{(datal, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x, }\AttributeTok{y =}\NormalTok{ value, }\AttributeTok{color =}\NormalTok{ y)) }\SpecialCharTok{+}
-    \FunctionTok{geom\_line}\NormalTok{() }\SpecialCharTok{+}
-    \FunctionTok{theme\_minimal}\NormalTok{(size) }\SpecialCharTok{+}
-    \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{,}
-          \AttributeTok{legend.title =} \FunctionTok{element\_blank}\NormalTok{()) }\SpecialCharTok{+}
-    \FunctionTok{ylab}\NormalTok{(}\StringTok{"Probability"}\NormalTok{) }\SpecialCharTok{+}
-    \FunctionTok{ggtitle}\NormalTok{(latex2exp}\SpecialCharTok{::}\FunctionTok{TeX}\NormalTok{(}\FunctionTok{sprintf}\NormalTok{(}\StringTok{"$}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{beta\_1 = \%s$"}\NormalTok{, b1)))}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-For example, the Figure \ref{fig:ex-categorical} depicts the effect of a categorical predictor on a 5-level \(y\) assuming uniform probabilities for the reference level and a mean difference on a standardized Normal distribution of 1.
-The Figure \ref{fig:ex-numerical} depicts the effect of a continuous predictor \(x\) sampled from a standard Normal distribution assuming a \(\beta = 1\).
+To better understand the impact of assuming (or simulating) a different latent scale we fit \(k - 1\) binomial regressions and check the estimated coefficients.
+We are not simulating a specific beta for each outcome but simulating a scale effect is actually impacting the regression coefficients.
+When generating data for a binary outcome the linear predictor is composed by \(\eta = \beta_0 + \beta_1x\).
+The threshold \(\alpha\) and slope of the function can be estimated using \(\alpha = -\frac{\beta_0}{\beta_1}\) and the slope is \(\frac{1}{\beta_1}\) (Faraggi, Izikson, \& Reiser, 2003; Knoblauch \& Maloney, 2012).
+Under the proportional odds assumption, there is only a change in thresholds \(\alpha\) this a shift in the sigmoid along the \(x\) axis.
+When including a scale effect a change in the sigmoid is combined with a change in the slope.
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/ex-categorical-1}
-
-}
-
-\caption{Example categorical}\label{fig:ex-categorical}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-20-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-20}
 \end{figure}
@@
 
+\subsection{2x2 interaction}\label{x2-interaction}
+
+A common research design could be a 2x2 factorial design.
+In this example we have 2 main effects and the interaction.
+Regardless the link function, we can focus the model equation on the linear predictor.
+Equation depicts the linear predictor for a 2x2 interaction.
+With a 2x2 design, a good strategy is to use sum to zero contrasts for the two factors.
+By default R use dummy coding but here we set the contrasts for \(X_1\) and \(X_2\) as \(0.5\) and \(-0.5\).
+In this way \(\beta_1\) will be the main effect of \(X_1\), \(\beta_2\) the main effect of \(X_2\) and \(\beta_3\) the interaction (thus the difference of differences).
+
+\[
+g(P(Y \leq k))^{-1} = \alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i}
+\]
+
+\scriptsize
+
+\begin{verbatim}
+## formula: y ~ x1 * x2
+## data:    dat
+## 
+##  link   threshold nobs  logLik     AIC        niter max.grad cond.H 
+##  probit flexible  4e+05 -597389.03 1194792.06 5(0)  1.42e-07 4.0e+01
+## 
+## Coefficients:
+##         Estimate Std. Error z value Pr(>|z|)    
+## x11     0.004798   0.003410   1.407    0.159    
+## x21     1.004223   0.003519 285.383   <2e-16 ***
+## x11:x21 0.505333   0.006833  73.954   <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Threshold coefficients:
+##      Estimate Std. Error z value
+## 1|2 -0.841241   0.002332  -360.8
+## 2|3 -0.254088   0.002092  -121.4
+## 3|4  0.250954   0.002092   120.0
+## 4|5  0.839533   0.002331   360.2
+\end{verbatim}
+
+\normalsize
+
+The parameters intepretation is the same as introduced in the general case.
+The thresholds \(\alpha\) are fixed the determines the probabilities \(P(Y = k)\) when all predictors are zero.
+In this case by doing \texttt{alpha\_to\_prob(fit\$alpha,\ link\ =\ "probit")} we should recover the \texttt{bprobs} vector.
+\texttt{x1} is the main effect thus the difference in \(z\) scores between \emph{a} and \emph{b} averaging over \texttt{x2}.
+The same holds for \texttt{x2}.
+\texttt{x1:x2} is the difference of differences in \(z\) scores thus \((ad - ac) - (bd - bc)\).
+
+To visualize the effects there is the \texttt{ggeffect::ggpredict()} function that compute the predictions according to the specified predictors combination.
+In this case we can plot the interaction using the estimated probability for each group and \(Y\) response.
+
 \scriptsize
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/ex-numerical-1}
-
-}
-
-\caption{Example numerical}\label{fig:ex-numerical}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-22-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-22}
+\end{figure}
+
+\normalsize
+
+\subsection{Numerical by factor interaction}\label{numerical-by-factor-interaction}
+
+Another common scenario is to compute and interaction between a numerical variable \(x\) and a factor \(g\).
+For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution \(x \sim \mathcal{U}(0, 1)\).
+Again, the thresholds \(\alpha\) represents the baseline probabilities when \(x = 0\) and \(g = 0\).
+
+\scriptsize
+
+\begin{verbatim}
+## formula: y ~ x * g
+## data:    dat
+## 
+##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
+##  probit flexible  1e+05 -153256.51 306527.02 5(0)  7.97e-08 1.6e+02
+## 
+## Coefficients:
+##      Estimate Std. Error z value Pr(>|z|)    
+## x     0.29419    0.01172   25.11   <2e-16 ***
+## g1    0.49679    0.01351   36.78   <2e-16 ***
+## x:g1  0.51690    0.02343   22.06   <2e-16 ***
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Threshold coefficients:
+##      Estimate Std. Error z value
+## 1|2 -0.841442   0.007474 -112.59
+## 2|3 -0.258062   0.007133  -36.18
+## 3|4  0.250592   0.007124   35.18
+## 4|5  0.838293   0.007404  113.23
+\end{verbatim}
+
+\normalsize
+
+Again, \texttt{x} is the slope of the numerical predictor averaging over \texttt{g} (given that \texttt{g} has been coded with sum to zero contrasts) thus the increase in \(z\) scores for a unit increase in \texttt{x}.
+\texttt{g1} is the main effect of the factor evaluated when \(x = 0\).
+To change the interpretation of \texttt{g1} a possibility is to center \texttt{x} differently (e.g., mean centering).
+The \texttt{x:g1} is the difference in slopes between the two groups.
+
+Regardless of the model complexity, the \texttt{sim\_ord\_latent()} function can be expanded by including more predictors.
+Clearly, as the model complexity increase, choosing meaningful and empirically plausible parameters became more challenging.
+
+\section{Power Analysis}\label{power-analysis}
+
+In the previous sections we introduced how to simulate standard and scale-location cumulative models and how to fit the corresponding models using the \texttt{clm()} function from the \texttt{ordinal} package.
+In this section we introduce how to compute the power for some common scenarios.
+We present an example in a very general form that can be expanded for specific use cases.
+For example, by simulating a model fixing all effects to zero it is possibile to estimate the type-1 error rate or fixing the sample size and varying the effect size it is possible to do a sensitivity analysis focusing on the smallest effect size of interest (ref qui).
+The general workflow for a power simulation can be summarized as:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Choose the research design, e.g., 2x2 factorial design
+\item
+  Indentifify the specific effect to focus on. For example the interaction effect, a main effect or a specific contrast
+\item
+  Write the model equation similarly to what presented in the previous sections and extract the model coefficients
+\item
+  Choose one of the proposed strategy for simulating data and follows the corresponding steps
+\item
+  Fit the model and extract the relevant statistics e.g., the p-value of the interaction
+\item
+  Repeat the simulation a large number of times (e.g, 10000) and store the relevant values from each simulation
+\item
+  Summarise the simulation results in the appropriate way.
+For example, the power of the interaction is the number of times the p-value for the interaction test is lower than the critical level divided by the number of simulations.
+\end{enumerate}
+
+To make a practical example, we simulate the power of a cumulative link model in detecting an effect of \(d = 0.4\) (probit model) with \(k = 6\) and \(n = 40\) participants per group.
+We fix the baseline probabilities in the first group as uniform.
+
+descrizione migliore
+
+\scriptsize
+
+\normalsize
+
+Then to compute the power we can simply compute the proportion of p values lower than the critical level.
+This can be done in R as \texttt{mean(p\ \textless{}=\ alpha)} resulting in an estimated power of 0.39.
+Clearly using only a single condition is not really informative thus we can vary the sample size across plausible values and check the corresponding power curve.
+
+\scriptsize
+
+\normalsize
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-26-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-26}
 \end{figure}
@@
 \url{https://doi.org/10.5334/irsp.82}
+
+\bibitem[\citeproctext]{ref-Faraggi2003-hm}
+Faraggi, D., Izikson, P., \& Reiser, B. (2003). Confidence intervals for the 50 per cent response dose. \emph{Statistics in Medicine}, \emph{22}, 1977--1988. \url{https://doi.org/10.1002/sim.1368}
 
@@
 
+\bibitem[\citeproctext]{ref-Sanchez-Meca2003-ji}
+Sánchez-Meca, J., Marín-Martínez, F., \& Chacón-Moscoso, S.
+(2003).
+Effect-size indices for dichotomized outcomes in meta-analysis.
+\emph{Psychological Methods}, \emph{8}, 448--467.
+\url{https://doi.org/10.1037/1082-989X.8.4.448}
+
+\bibitem[\citeproctext]{ref-Schad2020-ht}
+Schad, D.
+J., Vasishth, S., Hohenstein, S., \& Kliegl, R.
+(2020).
+How to capitalize on a priori contrasts in linear (mixed) models: A tutorial.
+\emph{Journal of Memory and Language}, \emph{110}, 104038.
+\url{https://doi.org/10.1016/j.jml.2019.104038}
+
 \bibitem[\citeproctext]{ref-Stanislaw1999-jr}
@@
 Non proportional odds models are widely dispensable -- sparser modeling based on parametric and additive location-shift approaches.
+\emph{arXiv {[}Stat.ME{]}}.
 Retrieved from \url{http://arxiv.org/abs/2006.03914}

### 2023-12-19 scripts-review/review.Rmd
updating

--- 2023-12-12 paper/paper-new.Rmd
@@
 ---
-title             : "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
-shorttitle        : "Ordinal regression models made easy"
-
-author: 
-  - name          : "Filippo Gambarota"
-    affiliation   : "1"
-    corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
-    email         : "filippo.gambarota@unipd.it"
-    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
-      - "Conceptualization"
-      - "Writing - Original Draft Preparation"
-      - "Writing - Review & Editing"
-  - name          : "Gianmarco Altoè"
-    affiliation   : "1"
-    role:
-      - "Writing - Review & Editing"
-      - "Supervision"
-
-affiliation:
-  - id            : "1"
-    institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"
-
-authornote: |
-  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-
-  Enter author note here.
-
-abstract: |
-  Abstract here
-
-keywords          : "ordinal, likert, simulations, power"
-wordcount         : "X"
-
-bibliography      : "`r filor::fil()$bib`"
-
-floatsintext      : yes
-linenumbers       : no
-draft             : no
-mask              : no
-
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-
-classoption       : "man"
-output: 
-  papaja::apa6_pdf:
-    latex_engine: xelatex
-    header_includes:
-      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline}
+title: "Review"
+author: "Filippo Gambarota"
+date: "`r Sys.Date()`"
+output: html_document
 ---
 
-```{r setup, include = FALSE}
-knitr::opts_chunk$set(echo = FALSE,
-                      dev = "pdf",
-                      fig.align = "center",
-                      message = FALSE,
-                      warning = FALSE)
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+devtools::load_all()
+library(tidyverse)
+library(ordinal)
 ```
 
-```{r packages, include = FALSE}
-library(tidyverse)
-library(ordinal)
-library(here)
-library(flextable)
-devtools::load_all()
-```
+# Scale effects and proportional odds
 
-```{r funs, include = FALSE}
-funs <- filor::get_funs(here("R", "utils.R"))
-```
-
-# Introduction
-
-Psychological research make an extensive use of ordinal data.
-One of the main reason is probably the usage of Likert scales [@Likert1932-xw].
-Ordinal data refers to a specific type of measurement scale [@Stevens1946-te] where ordered numbers are assigned to a variable.
-Compared to nominal scale and as the name suggest the labels are ordered.
-Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
-(rivedi questo).
-An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
-Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer.
-@Stevens1946-te suggest that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation).
-This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data.
-In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable.
-@Liddell2018-wu reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models.
-In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric.
-They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui).
-(vedi se c'è qualche altro lavoro che fa vedere questa cosa).
-
-### Ordinal regression models
-
-Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without making metric assumptions.
-We can class this general class of models as *ordinal regression*.
-The nomenclature of these models can be confusing mainly because there are several subclasses of models with different assumptions and structures [@Tutz2022-dg].
-@Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
-We can identify three main classes: *cumulative models* [@McCullagh1980-cw], *sequential models* [@Tutz1990-fe] and *adjacent category models*.
-The cumulative is the mostly used model assuming the existence of a latent variable that when segmented using thresholds produces the observed ordinal variable.
-The psychological process underlying the response is clearly formalized in the signal detection theory framework where the respondent (vedi come sistemare).
-The sequential model as suggested by the name is appropriate when modelling sequential processes.
-Assuming to have 5 response options, the sequential model assume that responding 3 means already reaching the states 1 and 2.
-A clear example is proposed by @Burkner2019-aw where the marriage duration in years is predicted as a function of some explanatory variables.
-For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential regression.
-When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached.
-The adjacent category model compare the category $k$ with $k + 1$ still assuming a latent distribution for each $k$.
-As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
-
-## notation
-
-- idea di mettere $F$ o qualcosa di simile per la pdf ($\Phi$ e.g., per la normale)
-- $Y^\star$ per la latente
-- $k$ numero di livelli ordinali
-- $\alpha$ per le soglie
-
-per notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw
-
-per la parte di notazione e quella più formale anche @DeCarlo2010-lj
-
-## cumulative vs others
-
-- cumulative quando una variabile continua latente viene tagliata in modo osservato. Le likert possono essere gestite in questo modo
-
-from @Burkner2019-aw:
-
-> For  many  ordinal  variables,  the  assumption  of  a   single underlying continuous variable, as in cumulative  models, may not be appropriate.
-If the response can  be understood as being the result of a sequential pro- cess, such that a higher response category is possible  only after all lower categories are achieved, the sequen- tial model proposed by Tutz (1990) is usually appropri- ate.
-
-vedi se fare k-1 logistiche è più o meno la stessa cosa
-
-in realtà credo che l'adiaject category e sequential model comunque abbiamo la proportional odds assumption
+To better understand the impact of assuming (or simulating) a different latent scale we fit $k - 1$ binomial regressions and check the estimated coefficients.
+We are not simulating a specific beta for each outcome but simulating a scale effect is actually impacting the regression coefficients.
+When generating data for a binary outcome the linear predictor is composed by $\eta = \beta_0 + \beta_1x$.
+The threshold $\alpha$ and slope of the function can be estimated using $\alpha = -\frac{\beta_0}{\beta_1}$ and the slope is $\frac{1}{\beta_1}$ [@Knoblauch2012-to; @Faraggi2003-hm].
+Under the proportional odds assumption, there is only a change in thresholds $\alpha$ this a shift in the sigmoid along the $x$ axis.
+When including a scale effect a change in the sigmoid is combined with a change in the slope.
 
 ```{r}
-#| eval: false
-library(brms)
-library(ordinal)
-
-dat <- data.frame(x = rnorm(100))
-dat <- sim_ord_latent(~x, By = log(2), probs = c(0.5, 0.2, 0.2, 0.1), data = dat, link = "logit")
-
-fit_clm <- clm(y ~ x, data = dat, link = "logit")
-fit_cum <- brm(y ~ x, data = dat, family = cumulative(link = "logit"))
-fit_seq <- brm(y ~ x, data = dat, family = sratio(link = "logit"))
-fit_adj <- brm(y ~ x, data = dat, family = acat(link = "logit"))
-
-```
-
-come sottolinea @Burkner2019-aw e @Tutz2020-xq i sequential model sono adatti per un certo tipo di dati dove l'interesse è vedere l'impatto dei predittori sulla transizione ad un categoria $r$ considerando che la categoria $r - 1$ è stata raggiunta.
-
-@Burkner2019-aw suggest that while cumulative and sequentials models have a clear empirical distinction the adjiacent category model is usually used for computational advantages.
-The probability of passing from one category to the other is modeled as a function of predictors.
-
-## Risolvere un pochino la nomenclatura
-
-https://cran.r-project.org/web/packages/ordinal/vignettes/clm_article.pdf
-
-> The name, cumulative link models is adopted from Agresti (2002), but the model class has
-been referred to by several other names in the literature, such as ordered logit models and
-ordered probit models (Greene and Hensher 2010) for the logit and probit link functions. The
-cumulative link model with a logit link is widely known as the proportional odds model due
-to McCullagh (1980) and with a complementary log-log link, the model is sometimes referred
-to as the proportional hazards model for grouped survival times.
-
-Da citare in qualche modo @Kemp2021-dj
-
-> “As a matter of fact, most of the scales used widely and effectively by psychologists are ordinal scales.
-In the strictest propriety the ordinary statistics involving means and standard deviations ought not to be used with these scales, for these statistics imply a knowledge of something more than the relative rank-order of data.
-On the other hand for this ‘illegal’ statisticizing there can be invoked a kind of pragmatic sanction: In numerous instances it leads to fruitful results.” (Stevens, 1946, p.
-679.)
-
-> It is straightforward to construct ordinal scales that do not involve rank ordering.
-For example, one can take the first element encountered and arbitrarily assign it the number 100.
-If the next element encountered is smaller it is given a smaller arbitrary number, 53 say.
-If the third element is between these two, it can be given the number 86, and so on.
-If this construction method is used, the difference between the elements assigned the numbers 70 and 80 will not in any important sense be equal to the difference between the elements assigned 90 and 100, and the intervals between the numbers are not really interpretable.
-Note, too, that monotonic transformations of the scale essentially leave the mea- sure unaffected.
-
-interessante come @Cliff2016-ck descriva che la maggiorparte dellle domande di ricerca in psicologia siano in riferimento alla location di un costrutto e possano essere gestite in modo ordinale
-
-The use of ordinal data is widespread in Psychology.
-Usually items from questionnaires are created using Likert scales where a certain psychological traits is measured using an intuitive scale with 3-5 or more anchor points.
-These measure cannot be considered *metric* measure on interval or ratio scales [stevens] but the categories are ordered.
-
-Other examples here
-
-Despite the usage of ordinal variables, statistical models made for these type of data are rarely used in Psychology.
-@Liddell2018-wu reported that the majority of published papers using likert-like measures used standard methods to analyze the data.
-In practical terms, they use *metric* models where the response variable cannot be considered fully numeric (vedi se c'è un termine tipo scala a rapporti).
-
-altro sui modelli
-
-Theoretically, using a *metric* for ordinal data is not appropriate but understanding the actual impact is not straightforward.
-@Liddell2018-wu did a comprehensive work about pitfalls of analzying ordinal data as metric.
-They showed that metric models produce higher type-1 and type-2 errors compared to the ordinal models.
-In particular, given the bounded nature of ordinal data, difference in the underlying latent distribution are not always captured by the metric model that simply estimate the mean of the ordinal variable.
-This is even more relevant when the underluing variance of the ordinal variables are not homogeneous.
-Furthermore they presented some situations where the metric model could be wrong in the opposite direction, finding an effect with the wrong sign (type-s error).
-
-Finally, they also demonstrated that even in the best condition where e.g.
-comparing two groups the variance are equal, the ordinal model is more powerful than the metric model given the underestimation of the true effect size from the latter.
-
-> Some authors have argued that, despite the ordinal character of individual Likert items, averaged ordinal items can have an emergent property of an interval scale and so it is appropriate to apply metric methods to the averaged values (e.g., Carifio & Perla, 2007, 2008).
-
-vedi come gestire questo
-
-anche questo è importante
-> Ordered-probit models typically assume that the thresholds (θ k) are the same across all groups because the thresholds are theoretically linked to the response measure, not to the predictor value.
-For example, when asked, “How happy are you?” with response options ‘1’ = very unhappy, ‘2’ = mildly unhappy, ‘3’ = neutral, ‘4’ = mildly happy, ‘5’ = very happy,” the latent thresholds between ordinal levels are assumed to be implicit in the phrasing of the question, regardless of other aspects of the respondent or situation.
-In other words, the thresholds are assumed to be part of the measurement procedure, not dependent on the value of the predictor or covariate.
-This can be technically referred to as a type of measurement invariance.
-
-Da citare come paper iniziale sui modelli [@McCullagh1980-cw].
-
-questo per la tassonomia dei diversi modelli ordinali [@Tutz2022-dg]
-
-da vedere magari per una critica tipo Gomila sul binary model [@Robitzsch2020-la]
-
-## Metriv vs ordinal models
-
-- the main difference is that the metric model assign a number to each label of the discrete variable assuming that the distance is the same. [@Liddell2018-wu]
-
-## lm on latent vs ordinal
-
-as suggested https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf, running a lm on the latent variables gives similar parameter as the clm.
-Of course, we are able to do this with real data given that the ordinal variable is the observed version of an unobserved latent variable.
-But in simulation this is useful to understand what the cumulative model is doing.
-
-```{r}
-dat <- data.frame(
-  x1 = rnorm(1e5),
-  x2 = rnorm(1e5)
-)
-
-dat <- sim_ord_latent(~x1 + x2, By = c(1, 0.5), probs = rep(1/5, 5), link = "probit", data = dat)
-
-fit <- lm(ys ~ x1 + x2, data = dat)
-summary(fit)
-
-clm(y ~ x1 + x2, data = dat, link = "probit")
-coef(fit)/sigma(fit)
-
-```
-
-## Kruscke parametrization
-
-@Liddell2018-wu and @Kruschke2015-re proposed an alternative parametrization to understand the model parameters.
-They used a probit model where thresholds and regression parameters are estimated on the scale of the ordinal variable compared to standard ordinal regression where they refers to the quantile of the latent variables for the threshold and z score or odds ratio for the regression coefficients.
-They implemented the model in Jags and provided some equations and R functions to convert from the standard parametrization to the proposed one.
-
-The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation).
-Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion.
-The main improvement regards mapping the values (for thresholds $\alpha_i$ and slopes $\beta$) from latent standard distribution (gaussian or logistic) into the scale of the $y$ ordinal value.
-The scale of the variable depend on the numeric labels assigned to ordered categories.
-There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to $\alpha_1 + 0.5$ and $\alpha_k - 0.5$ where other thresholds are estimated.
-
-@Kurz2023-zv explain how to convert a model fitted with the `brms` package (citation) (an R package for regression modeling using *stan*) into the Kruschke [-@Kruschke2015-re] parametrization.
-
-Using the function proposed by @Kruschke2015-re (metti ref ad osf) and the equations presented in @Kurz2023-zv the `clm_to_ord()` function convert parameters fitted with the `clm` function into the corresponding parameters for the latent $Y^\star$ variable considering the actual range of values (from 1 to $k$).
-
-```{r, echo=FALSE, results='asis'}
-filor::print_fun(funs$clm_to_ord)
-```
-
-```{r}
-#| echo: true
-
-dat <- data.frame(
-  x = rnorm(1e5)
-)
-
-dat <- sim_ord_latent(~x, By = 1, probs = rep(1/5, 5), link = "probit", data = dat)
-fit <- clm(y ~ x, data = dat, link = "probit")
-clm_to_ord(fit)
-```
-
-## Gelman parametrization
-
-rivedi questa parametrizzazione in @Gelman2020-tg
-
-```{r}
-#| eval: false
-dat <- data.frame(x = runif(1e5, 0, 1))
-dat <- sim_ord_latent(~x, By = 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), data = dat, link = "logit")
-fit_clm <- clm(y ~ x, data = dat, link = "logit")
-
-num_latent_plot(dat$x, 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), link = "logit")
-
-summary(fit_clm)
-
-ths <- -(fit_clm$alpha/fit_clm$beta)
-betas <- 1/fit_clm$beta
-
-dat <- cbind(dat, dummy_ord(dat$y))
-
-fit <- glm(y1234vs5 ~ x, data = dat, family = binomial(link = "logit"))
-
-coefs <- coef(fit)
-
-x <- seq(0, 1, 0.01)
-
-plot(x, plogis(coefs[1] + coefs[2] * x), type = "l")
-
--(fit_clm$alpha/fit_clm$beta)
-1/fit_clm$beta
-
-```
-
-## Odds ratios and cumulative odds ratios
-
-- https://online.stat.psu.edu/stat504/lesson/4/4.1
-
-In this section we introduce the main concept to understand the effects of a logit model.
-
-## Demonstrating the proportional odds assumption
-
-basically given that the effect of $\beta$ is constant the odds ratio is independent from the thresholds $\alpha_j$ [@Liu2023-bp].
-In the current paper we are simulating data using a single set of $\beta$s thus the proportional odds assumption is true.
-@Liu2023-bp reviewed the available methods to test this assumption.
-
-- https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione
-
-sarebbe da capire quali sono i rischi di assumere questo ed eventualmente se simulare non proportional odds è troppo complicato.
-
-Let's simulate the effect of a binary predictor on ordinal scale 1-5:
-
-```{r}
-#| echo: true
-b1 <- log(3) # log odds ratio
-n <- 1e4
-x <- rep(c("a", "b"), each = n/2)
-dat <- data.frame(x = x)
-probs <- rep(1/5, 5) # for the group "a", uniform probabilities
-dat <- sim_ord_latent(~x, By = b1, probs = probs, data = dat, link = "logit")
-fit <- clm(y ~ x, data = dat, link = "logit")
-pr <- predict(fit, data.frame(x = unique(x)))$fit
-pr
-```
-Basically the proportional odds suggest that:
-
-$$
-\text log(\frac{P(y \leq 1)}{P(y > 1)})
-$$
-Is the same regardless the level of the $x$ predictor. Thus:
-
-$$
-\text{log}\left(\frac{\frac{P(y \leq 1|x_0)}{P(y > 1|x_0)}}{\frac{P(y \leq 2|x_0)}{P(y > 2|x_0)}}\right) = \text{log}\left(\frac{\frac{P(y \leq 1|x_1)}{P(y > 1|x_1)}}{\frac{P(y \leq 2|x_1)}{P(y > 2|x_1)}}\right)
-$$
-
-```{r}
-#| echo: true
-a_or1vs2345 <- filor::odds(pr[1, 1]) # 1 vs 2 3 4 5
-a_or12vs345 <- filor::odds(sum(pr[1, 1:2])) # 1 vs 2 3 4 5
-
-b_or1vs2345 <- filor::odds(pr[2, 1]) # 1 vs 2 3 4 5
-b_or12vs345 <- filor::odds(sum(pr[2, 1:2])) # 1 vs 2 3 4 5
-
-c(xa = log(a_or1vs2345 / a_or12vs345), xb = log(b_or1vs2345 / b_or12vs345))
-```
-
-Proportional odds can be also visualized by plotting the cumulative probabilities of $y$, in terms of $g(P(y \leq j))$ (where $g()$ is the logit link function) as a function of the predictor $x$.
-If the proportional odds assumptions holds the slopes are parallel (also known as parallel regression assumption).
-The Figure \@ref(fig:poa-plot-example) depicts the assumption of proportional odds in the probability and logit space.
-
-```{r poa-plot-example, fig.cap="POA assumption"}
-k <- 5 # number of levels for y
-grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
-probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
-cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
-cprobs <- data.frame(do.call(rbind, cprobs))
-names(cprobs) <- sprintf("y<=%s", 1:k)
-cprobs <- cprobs[, -ncol(cprobs)] # remove the last
-grid <- cbind(grid, cprobs)
-
-grid <- grid |> 
-  pivot_longer(starts_with("y")) |> 
-  mutate(lp = qlogis(value))
-
-poa_probs_plot <- grid |> 
-  ggplot(aes(x = x, y = value, color = name)) +
-  geom_line() +
-  theme_minimal(15) +
-  theme(legend.position = "bottom",
-        legend.title = element_blank(),
-        axis.title.y = element_blank()) +
-  ggtitle("Probability")
-
-poa_lp_plot <- grid |> 
-  ggplot(aes(x = x, y = lp, color = name)) +
-  geom_line() +
-  theme_minimal(15) +
-  theme(axis.title.y = element_blank()) +
-  ggtitle("Linear Predictor")
-
-bl <- cowplot::get_legend(
-  poa_probs_plot
-)
-
-poa_plot <- cowplot::plot_grid(
-  poa_probs_plot + theme(legend.position = "none"),
-  poa_lp_plot + theme(legend.position = "none")
-)
-
-cowplot::plot_grid(poa_plot, bl, ncol = 1, rel_heights = c(0.9, 0.1))
-```
-
-@Harrell-2015-no proposed a very intuitive plot to assess the proportional odds assumption and eventually the degree of deviation from the ideal case.
-Basically predictor is plotted against the logit of the cumulative probability.
-Distances between pairs of symbols should be similar across levels of the predictors.
-Numerical predictors can be binned before plotting the corresponding logit.
-Figure \@ref(fig:harrel-poa-plot) depicts an example with simulated data satisfying the proportional odds assumption.
-
-```{r harrel-poa-plot, fig.cap="Harrel Poa Plot"}
-k <- 5 # number of levels for y
-grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
-probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
-cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
-cprobs <- data.frame(do.call(rbind, cprobs))
-names(cprobs) <- sprintf("y<=%s", 1:k)
-cprobs <- cprobs[, -ncol(cprobs)] # remove the last
-grid <- cbind(grid, cprobs)
-
-grid <- grid |> 
-  pivot_longer(starts_with("y")) |> 
-  mutate(lp = qlogis(value))
-
-grid |> 
-  ggplot(aes(x = lp, y = x)) +
-  geom_point(aes(color = name, 
-                 shape = name),
-             size = 3) +
-  theme_minimal(20) +
-  xlab("Logit") +
-  theme(legend.position = "bottom",
-        legend.title = element_blank())
-```
-
-From a statistical point of view, the proportional odds assumption can be assessed by fitting $k - 1$ binomial regressions and checking if the estimated $\beta$ is similar between regressions.
-The regressions are estimated by creating $k - 1$ dummy variables from the ordinal $y$.
-The code below show that simulating data with the proportional odds create $k - 1$ binomial regression with similar $\beta$s.
-In fact, fitting $k - 1$ binomial regressions can be considered also an alternative to fitting an ordinal regression [see @Gelman2020-tg] with more flexibility in parameters (there will be $k - 1$ regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).
-
-```{r}
-#| echo: true
 k <- 4
 n <- 1e5
-dat <- data.frame(
-  x = runif(n)
-)
-
-dat <- sim_ord_latent(~x, By = 5, probs = rep(1/k, k), data = dat, link = "logit")
-
-# create dummy variables
+b1 <- 3
+d1 <- log(2)
+dat <- data.frame(x = runif(n))
+dat <- sim_ord_latent(~x, ~x, By = b1, Bscale = d1, prob = rep(1/k, k), data = dat, link = "probit")
 
@@
 
-fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "logit"))
-fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "logit"))
-fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "logit"))
+dat$y <- ordered(dat$y)
+fit <- clm(y ~ x, scale = ~x, data = dat, link = "probit")
 
-car::compareCoefs(fit123vs4, fit12vs34, fit1vs234)
+fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "probit"))
+fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "probit"))
+fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "probit"))
+
+fits <- list(y1vs234 = fit1vs234, fit12vs34 = fit12vs34, fit123vs4 = fit123vs4)
+
+lapply(fits, function(x) coef(x)) |> 
+  bind_rows(.id = "model") |> 
+  mutate(xp = list(seq(0, 5, 0.01))) |> 
+  unnest(xp) |> 
+  ggplot(aes(x = xp, y = `(Intercept)` + x * xp)) +
+  geom_line(aes(color = model))
 ```
 
-spiegare qui la differenza tra proportional and non proportional odds @Tutz2020-xq.
+When including a scale effect, the proportional odds or parallel slopes are no longer assumed.
+However, compared to fitting a non-proportional odds model, there is still a single parameter for each predictor regardless the number of $k$ levels.
+This is clear when fitting a ordinal model including nominal effects i.e.
+allowing $\beta_j$ to differs for different $k$.
 
-## Logit vs Probit model
-
-When fitting an ordinal regression the two mostly used link functions are the *probit* and *logit*.
-From the distribution point of view the two functions are very similar.
-The *probit* model is based on a cumulative Gaussian distribution while the *logit* model is based on a logistic distribution.
-Figure \@ref(fig:logit-vs-probit) depict the two cumulative distributions.
-
-```{r logit-vs-probit, fig.cap="Logit vs Probit"}
-par(mfrow = c(1,2))
-
-# density
-
-curve(dnorm(x), -5, 5, lwd = 2, ylab = "Density", main = "Probability Density", cex.lab = 1.2, cex.axes = 1.2)
-curve(dlogis(x), -5, 5, lwd = 2, col = "firebrick", add = TRUE)
-
-# cdf
-
-curve(pnorm(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", lwd = 2, main = "Cumulative Probability",
-      cex.lab = 1.2, cex.axes = 1.2)
-curve(plogis(x), ylim = c(0, 1), xlim = c(-5, 5), y = "Probability", col = "firebrick", lwd = 2, add = TRUE)
+```{r}
+fit_nominal <- clm(y ~ 1, nominal = ~x, data = dat, link = "probit")
+summary(fit_nominal)
+sapply(fits, coef)["x", ]
 ```
 
-Given the different underlying distribution, the parameters have a different interpretation under the two models.
-The probit model assume a latent standard normal distribution with and $\sigma = 1$.
-The logit model assume a logistic distribution with $\sigma^2 = \pi^2/3$.
-Thus regression coefficients $\beta$ represents the increase in standard deviations units.
-The interpretation in terms of latent distribution is particularly useful for the probit model where $\beta$s can be interpreted in a Cohen's $d$ like manner.
-Furthermore, there is the possibility to directly map parameters from signal detection theory [@Stanislaw1999-jr; @Green1966-gy] into an ordinal probit regression.
-In practical terms, the thresholds are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].
-
-The latent formulation of the ordinal regression model allow to think in standard linear regression terms for choosing and interpreting model parameters before converting into the probability space.
-Both the standard Normal and the logistic distribution are defined with a location ($\mu$) and a scale ($s$) parameter.
-For the Normal distribution, $\mu$ is the mean and scale is the variance ($\sigma^2$).
-For the logistic distribution the variance is $\sigma^2=\frac{s^2\pi^2}{3}$.
-Thus fixing $\mu$ and $\sigma^2 = 1$ (the default), the two distributions are similar with the logistic having more variance.
-For this reason, the latent formulation for parameter intepretation is particularly useful for the probit model because the $\beta$ is directly interpreted in standard deviaton units that by default are fixed to 1.
-With a binary predictor $x$, $\beta$ is the shift of the latent distribution increasing $x$ by one unit.
-This can be directly interpreted as a standardized mean difference effect size (e.g., Cohen's $d$).
-This is the same for the logistic distribution but a unit increase in $x$ shift the latent distribution by $\sigma = \sqrt{\frac{s^2 \pi^2}{3}} = \frac{s\pi}{\sqrt{3}}$.^[In the case of a standard logistic distribution ($s^2 = 1$), the standard deviation is $\frac{\pi}{\sqrt{3}}$].
-
-## Simulating data
-
-There are mainly two ways to simulate data.
-The first method concerns simulating starting from the latent formulation of the ordinal model.
-Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.
-
-### simulating using probabilities
-
-The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution.
-This is similar to the general method to simulate data for a generalized linear model where the linear predictors $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.
-The following code box simulate a $n$ binary trials with a continuous predictor $x$.
-
-```{r}
-xl <- seq(0, 1, 0.1)
-x <- rep(xl, 10)
-n <- length(x)
-b0 <- qlogis(0.01)
-b1 <- 8
-lp <- b0 + b1 * x # linear predictor
-y <- rbinom(n, 1, plogis(lp)) # sampling from a binomial distribution using the vector of probabilities
-p <- tapply(y, x, mean)
-plot(unique(x), p, pch = 19)
-curve(plogis(x, -(b0/b1), 1/b1), add = TRUE, lwd = 2, col = "firebrick")
-```
-
-We can apply the same idea to an ordinal outcome but we need $k - 1$ equations where $k$ is the number of ordinal levels.
-Let's simulate a similar design with a continuous $x$ predictor and $k = 4$ options.
-We fix the baseline probabilities where $x = 0$ as uniform thus $p(y_1) = p(y_2) = ...
-p(y_k) = 1/k$.
-The general workflow for the simulation is:
-
-1. Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
-2. Define the regression coefficients $\beta$s
-3. Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
-4. Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\mathbf{\beta}$.
-5. Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
-6.
-Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities.
-$p(y = 1) = 0 - p(y <= 1), p(y = 2) = p(y <= 2) - p(y <= 1)$ (sistema qua).
-This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities.
-7. Sample $n$ outcomes from a multinomial distribution, choosing between $k$ alternatives with the calculated probabilities.
-
-(rivedi se fare con for)
-This can be easily implemented in R. We are using a `for` loop to improve the clarity of the workflow but using the `*apply` family of functions can improve the efficiency of the computations.
-
-```{r}
-k <- 4  # number of options (1)
-n <- 1e5 # number of observations (1)
-b1 <- 0.5 # beta1, the shift in the latent distribution (2)
-probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
-dat <- data.frame(x = rnorm(n)) # create dataframe
-lp <- lapply(alphas, function(a) a - b1 * dat$x) # k - 1 linear predictors (4)
-names(lp) <- sprintf("g[p(y <= %s)]", 1:(k - 1)) # giving appropriate names
-```
-
-The linear predictors can be seen in the Figure ref. The $g$ in this case is the cumulative *probit* function $\Phi$.
-
-```{r}
-#| echo: false
-lpt <- lp
-names(lpt) <- latex2exp::TeX(sprintf("$g[p(y \\leq %s)]$", 1:(k - 1)))
-datl <- cbind(dat, lpt)
-datl <- tidyr::pivot_longer(datl, 2:ncol(datl), names_to = "cumy", values_to = "lp")
-
-ggplot(datl, aes(x = x, y = lp, color = cumy)) +
-  facet_wrap(~cumy, labeller = label_parsed) +
-  geom_hline(yintercept = alphas, lty = "dashed") +
-  geom_line(show.legend = FALSE) +
-  theme_minimal(20) +
-  ylab("Linear Predictor")
-```
-
-Now we can apply the inverse of the link function $g^{-1} = \Phi^{-1}$ to calculate the corresponding cumulative probabilities.
-
-```{r}
-cump <- lapply(lp, pnorm) # inverse of the link function (5)
-cump <- data.frame(cump)
-p <- apply(cbind(0, cump, 1), 1, diff, simplify = FALSE) # probability of each k outcome (6)
-p <- do.call(rbind, p)
-```
-
-```{r}
-p <- data.frame(p)
-names(p) <- sprintf("p_y%s", 1:k)
-filor::trim_df(data.frame(p))
-```
-
-We can plot the expected effect of $\beta_1$ on the $k$ probabilities using the `num_latent_plot()` function (see Figure \@ref(fig:sim-with-probs-lat-plot)).
-Finally we sample using the `sample()` function using the calculated probabilities.
-
-```{r sim-with-probs-lat-plot, fig.cap="sim-with-probs-lat-plot"}
-num_latent_plot(x = dat$x, b1 = b1, probs = probs0, link = "probit")
-```
-
-```{r}
-dat$y <- apply(p, 1, function(ps) sample(1:k, 1, prob = ps))
-```
-
-To check the simulation result we can increase the number of observations, fit the model using the `ordinal::clm()` function and assess the recovery of simulated parameters.
-
-```{r}
-dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
-fit <- clm(y ~ x, data = dat, link = "probit")
-```
-
-```{r tab-sim-probs}
-truth <- c(alphas, b1)
-names(truth) <- c(names(fit$alpha), names(fit$beta))
-clm_table(fit, truth = truth, caption = "caption") |> 
-  flextable_to_rmd()
-```
-
-### simulating using latent distribution
-
-An equivalent but more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
-This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and the cutting the latent values according to the thresholds $\alpha$.
-The workflow is slighlty different compared to the previous paragraph:
-
-1. Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
-2. Define the regression coefficients $\beta$
-<!-- controlla che le soglie non fanno riferimento a quando è 0 ma concettualmente si. essendo che non cambiano non è rilevante a quale x si faccia riferimento -->
-3. Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
-4.
-Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$, the regression coefficients $\mathbf{\beta}$ and the error $\mathbf{\epsilon}$ sampled from the appropriate probability distribution.
-5.
-Cut the latent variable into $k$ areas using the thresholds $\alpha$ and assign the corresponding ordinal value.
-This step simply checks if the latent value $Y^{\star}_i$ is between two threshold values and assign the corresponding value.
-This can be done using the `cut()` or the `findInterval()` functions.
-
-```{r}
-k <- 4  # number of options (1)
-n <- 1e5 # number of observations (1)
-b1 <- 0.5 # beta1, the shift in the latent distribution (2)
-probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
-dat <- data.frame(x = rnorm(n)) # create dataframe
-dat$ystar <- b1 * dat$x + rnorm(n, 0, 1) # linear predictor (4)
-dat$y <- findInterval(dat$ystar, alphas) + 1 # cut the latent distribution (5). The + 1 because the first category is 0 by default.
-```
-
-The Figure (put crossref) depict the simulated $Y^{*}$ and the corresponding ordinal value. As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.
-
-```{r}
-alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))
-
-dat |> 
-  slice_sample(n = 1000) |> 
-  mutate(y = findInterval(ystar, alphas) + 1) |> 
-  ggplot(aes(x = x, y = ystar)) +
-  geom_point(alpha = 0.6, aes(color = factor(y)), size = 3) +
-  theme_minimal(15) +
-  theme(legend.position = "bottom",
-        legend.title = element_blank()) +
-  ylab(latex2exp::TeX("$Y^{*}$")) +
-  geom_hline(yintercept = alphas) +
-  annotate("label", x = min(dat$x), y = alphas, label = alpha_n)
-```
-
-```{r}
-dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
-fit <- clm(y ~ x, data = dat, link = "probit")
-```
-
-```{r tab-sim-latent}
-truth <- c(alphas, b1)
-names(truth) <- c(names(fit$alpha), names(fit$beta))
-clm_table(fit, truth)
-```
-
-The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function.
-
-```{r}
-#| results: asis
-
-filor::print_fun(funs$sim_ord_latent)
-```
-
-## Scale Effects
-
-The default ordinal regression model assume that the variance of th underlying latent distribution is the same across condition.
-This is similar to a standard linear regression assuming homogeneity of variance.
-For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy].
-In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution.
-This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
-Figure \@ref(fig:example-unequal-variance) depict an example of a comparison between two groups where the two underlyning latent distributions have unequal variance.
-
-spiegare qui che i location-scale models sono sostanzialmente un cambiamento delle soglie (e non dei beta) in funzione di predittori @Tutz2020-xq
-
-dire che gli scale effects possono essere visti come un'alterativa più parsimoniosa a modelli non-proportional odds @Christensen2019-cz
-
-se citi brms per modelli più complicati, ricordati di specificare che:
-
-> That is, s = 1/disc. Further, because disc must be strictly positive, it is by default modeled on the log scale.
-
-un modo intuitivo per spiegare gli scale effects è di far capire la relazione tra slope e scale.
-La slope della funzione cumulative corrisponde alla varianza.
-Se la slope è piuù steep, la varianza sarà minore e la discrimizione maggiore (in `brms` si chiama desc)
-
-```{r}
-#| eval: false
-library(distributional)
-library(ggdist)
-library(tidyverse)
-
-s <- c(0.5, 1, 2, 3)
-
-dat <- data.frame(
-  m = 0,
-  s = s
-)
-
-dat$dist <- dist_normal(dat$m, dat$s)
-
-par(mfrow = c(1, 2))
-curve(dnorm(x, m = dat$m[1], sd = dat$s[1]), -5, 5, lwd = 2)
-lapply(2:nrow(dat), function(i) curve(dnorm(x, dat$m[i], dat$s[i]), add = TRUE, col = i, lwd = 2))
-
-curve(pnorm(x, m = dat$m[1], sd = dat$s[1]), -5, 5, lwd = 2)
-lapply(2:nrow(dat), function(i) curve(pnorm(x, dat$m[i], dat$s[i]), add = TRUE, col = i, lwd = 2))
-
-# x <- seq(0, 1, 0.001)
-# p <- pnorm(qnorm(0.01) + 5 * x)
-# y <- rbinom(length(x), 1, p)
-# 
-# plot(x, p)
-# curve(pnorm(x, -(qnorm(0.01)/5), 1/5))
-```
-
-si può citare anche questo @Cox1995-ur come riferimento per i modelli
-
-una distinzione tra location-shift and location-scale models @Tutz2017-de dove i primi oltre ad un cambiamento di location prevedono un cambiamento di soglie mentre i secondi modellano la differenza di soglie come un cambiamento della scala (varianza) che quindi permette di aumetare o diminuire l'area associata ad un determinato intervallo.
-
-sempre in @Tutz2017-de propongono il location-scale/shift come una via di mezzo tra il proportional odds ed il non proportional odds
-
-### Simulating scale effects
-
-The location-scale model can be simulated using the `sim_ord_latent()` function and providing the predictors for the scale parameters.
-Given the `log` link function, predictors are provided on the `log` scale.
-For example, we simulate the effect of a binary variable $x$ representing two independent groups predicting the $k = 5$ response.
-We simulate a *location* effect of $\beta_1 = 0.5$ (in *probit* scale) and $\zeta_1 = \text{log}(2) = 0.70$.
-The first group has a $\sigma = 1$ and the second group has $\sigma = 2$.
-Again we simulate that the baseline probabilities are uniform for the first group.
-
-```{r}
-k <- 5  # number of options (1)
-n <- 1e5 # number of observations (1)
-b1 <- 0.5 # beta1, the shift in the latent distribution (2)
-z1 <- log(2) # zeta1, the change in the scale
-probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
-dat <- data.frame(x = rep(c(0, 1), each = n/2))
-dat <- sim_ord_latent(~x, scale = ~x, By = b1, Bscale = z1, probs = probs0, data = dat, link = "probit")
-fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
-```
-
-The table (put ref) reports the simulation results.
-
-```{r}
-#| eval: false
-truth <- c(alphas, b1, z1)
-names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
-clm_table(fit, truth)
-```
-
-## Visualizing effects
-
-Before starting with the actual simulation it is useful to plot the predicted probabilities of the response variable $y$ as a function of the predictors.
-The `cat_latent_plot()` and the `num_latent_plot()` functions are able to visualize the effect of a single categorical or numerical predictor.
-Specifying the mean and standard deviations of the latent variable in each condition (for the categorical version) or the $\beta$ (for the numerical version) the function compute the expected probability for each level of the ordinal variable $y$ and visualize the result.
-In this way the $\beta$ can be choose in a meaningful way.
-
-```{r, echo=FALSE, results='asis'}
-filor::print_fun(c(funs$cat_latent_plot, funs$num_latent_plot))
-```
-
-For example, the Figure \@ref(fig:ex-categorical) depicts the effect of a categorical predictor on a 5-level $y$ assuming uniform probabilities for the reference level and a mean difference on a standardized Normal distribution of 1.
-The Figure \@ref(fig:ex-numerical) depicts the effect of a continuous predictor $x$ sampled from a standard Normal distribution assuming a $\beta = 1$.
-
-```{r ex-categorical, fig.cap="Example categorical", fig.width=8, fig.height=8}
-cat_latent_plot(m = c(0, 1), s = 1, probs = rep(1/5, 5), link = "probit")
-```
-
-```{r ex-numerical, fig.cap="Example numerical", fig.width=8, fig.height=8}
-num_latent_plot(x = rnorm(100), b1 = 1, probs = c(0.4, 0.2, 0.2, 0.05, 0.05), link = "probit")
-```
-
-## Disclaimer about the functions
-
-The current paper proposed a simplified way with some functions to generate ordinal data.
-For more complex simulations such as simulating correlated ordinal data the `simstudy` package [https://kgoldfeld.github.io/simstudy/articles/ordinal.html](https://kgoldfeld.github.io/simstudy/articles/ordinal.html) proposed a very comprehensive set of data simulation function also for ordinal data.
-
-\newpage
-
-# References
-
-::: {#refs custom-style="Bibliography"}
-:::
+The effect estimated by the $k - 1$ binomial regression are the same.
+Crucially, we did not simulated 3 $\beta$s but estimating these values is a consequence of allowing the scale to vary as a function of $x$.
+Clearly, as the number of $x$ and $k$ increase, the nominal model is less parsimonious estimating more parameters.
+From that point of view, the same dataset can be modelled using a scale-location model.

### 2023-12-19 scripts-review/supplementary.Rmd
updating

---
title: "Untitled"
author: "Filippo Gambarota"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## lm on latent vs ordinal

as suggested https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf, running a lm on the latent variables gives similar parameter as the clm. Of course, we are able to do this with real data given that the ordinal variable is the observed version of an unobserved latent variable. But in simulation this is useful to understand what the cumulative model is doing.

```{r}
dat <- data.frame(
  x1 = rnorm(1e5),
  x2 = rnorm(1e5)
)

dat <- sim_ord_latent(~x1 + x2, By = c(1, 0.5), probs = rep(1/5, 5), link = "probit", data = dat)

fit <- lm(ys ~ x1 + x2, data = dat)
summary(fit)

clm(y ~ x1 + x2, data = dat, link = "probit")
coef(fit)/sigma(fit)
```

The proposed model is fitted within a Bayesian framework using either Jags (citation) or Stan (citation). Kruskche proposed a simple method to convert the parameters fitted with a standard model within the proposed parametriazion. The main improvement regards mapping the values (for thresholds $\alpha_i$ and slopes $\beta$) from latent standard distribution (gaussian or logistic) into the scale of the $y$ ordinal value. The scale of the variable depend on the numeric labels assigned to ordered categories. There is an additional feature of the proposed parametrization where the first and the last thresholds are fixed respectively to $\alpha_1 + 0.5$ and $\alpha_k - 0.5$ where other thresholds are estimated.

Using the function proposed by @Kruschke2015-re (metti ref ad osf) and the equations presented in @Kurz2023-zv the `clm_to_ord()` function convert parameters fitted with the `clm` function into the corresponding parameters for the latent $Y^\star$ variable considering the actual range of values (from 1 to $k$).

```{r, echo=FALSE, results='asis'}
filor::print_fun(funs$clm_to_ord)
```

```{r}
#| echo: true

dat <- data.frame(
  x = rnorm(1e5)
)

dat <- sim_ord_latent(~x, By = 1, probs = rep(1/5, 5), link = "probit", data = dat)
fit <- clm(y ~ x, data = dat, link = "probit")
clm_to_ord(fit)
```

## Gelman parametrization

rivedi questa parametrizzazione in @Gelman2020-tg

```{r}
#| eval: false
dat <- data.frame(x = runif(1e5, 0, 1))
dat <- sim_ord_latent(~x, By = 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), data = dat, link = "logit")
fit_clm <- clm(y ~ x, data = dat, link = "logit")

num_latent_plot(dat$x, 2, probs = c(0.5, 0.2, 0.1, 0.1, 0.1), link = "logit")

summary(fit_clm)

ths <- -(fit_clm$alpha/fit_clm$beta)
betas <- 1/fit_clm$beta

dat <- cbind(dat, dummy_ord(dat$y))

fit <- glm(y1234vs5 ~ x, data = dat, family = binomial(link = "logit"))

coefs <- coef(fit)

x <- seq(0, 1, 0.01)

plot(x, plogis(coefs[1] + coefs[2] * x), type = "l")

-(fit_clm$alpha/fit_clm$beta)
1/fit_clm$beta

```

Proportional odds can be also visualized by plotting the cumulative probabilities of $y$, in terms of $g(P(y \leq j))$ (where $g()$ is the logit link function) as a function of the predictor $x$. If the proportional odds assumptions holds the slopes are parallel (also known as parallel regression assumption). The Figure \@ref(fig:poa-plot-example) depicts the assumption of proportional odds in the probability and logit space.

```{r poa-plot-example, fig.cap="POA assumption"}
k <- 5 # number of levels for y
grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
cprobs <- data.frame(do.call(rbind, cprobs))
names(cprobs) <- sprintf("y<=%s", 1:k)
cprobs <- cprobs[, -ncol(cprobs)] # remove the last
grid <- cbind(grid, cprobs)

grid <- grid |> 
  pivot_longer(starts_with("y")) |> 
  mutate(lp = qlogis(value))

poa_probs_plot <- grid |> 
  ggplot(aes(x = x, y = value, color = name)) +
  geom_line() +
  theme_minimal(15) +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Probability")

poa_lp_plot <- grid |> 
  ggplot(aes(x = x, y = lp, color = name)) +
  geom_line() +
  theme_minimal(15) +
  theme(axis.title.y = element_blank()) +
  ggtitle("Linear Predictor")

bl <- cowplot::get_legend(
  poa_probs_plot
)

poa_plot <- cowplot::plot_grid(
  poa_probs_plot + theme(legend.position = "none"),
  poa_lp_plot + theme(legend.position = "none")
)

cowplot::plot_grid(poa_plot, bl, ncol = 1, rel_heights = c(0.9, 0.1))
```

@Harrell-2015-no proposed a very intuitive plot to assess the proportional odds assumption and eventually the degree of deviation from the ideal case. Basically predictor is plotted against the logit of the cumulative probability. Distances between pairs of symbols should be similar across levels of the predictors. Numerical predictors can be binned before plotting the corresponding logit. Figure \@ref(fig:harrel-poa-plot) depicts an example with simulated data satisfying the proportional odds assumption.

```{r harrel-poa-plot, fig.cap="Harrel Poa Plot"}
k <- 5 # number of levels for y
grid <- data.frame(x = seq(0, 1, 0.1)) # numeric predictor
probs <- get_probs(~x, 5, rep(1/k, k), data = grid, link = "logit") # probabilities
cprobs <- apply(probs, 1, cumsum, simplify = FALSE)
cprobs <- data.frame(do.call(rbind, cprobs))
names(cprobs) <- sprintf("y<=%s", 1:k)
cprobs <- cprobs[, -ncol(cprobs)] # remove the last
grid <- cbind(grid, cprobs)

grid <- grid |> 
  pivot_longer(starts_with("y")) |> 
  mutate(lp = qlogis(value))

grid |> 
  ggplot(aes(x = lp, y = x)) +
  geom_point(aes(color = name, 
                 shape = name),
             size = 3) +
  theme_minimal(20) +
  xlab("Logit") +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

From a statistical point of view, the proportional odds assumption can be assessed by fitting $k - 1$ binomial regressions and checking if the estimated $\beta$ is similar between regressions. The regressions are estimated by creating $k - 1$ dummy variables from the ordinal $y$. The code below show that simulating data with the proportional odds create $k - 1$ binomial regression with similar $\beta$s. In fact, fitting $k - 1$ binomial regressions can be considered also an alternative to fitting an ordinal regression [see @Gelman2020-tg] with more flexibility in parameters (there will be $k - 1$ regression coefficients instead of a single one) but losing the latent intepretation (magari qualcosa di più qui).

```{r}
#| echo: true
k <- 4
n <- 1e5
dat <- data.frame(
  x = runif(n)
)

dat <- sim_ord_latent(~x, By = 5, probs = rep(1/k, k), data = dat, link = "logit")

# create dummy variables

dat$y1vs234 <- ifelse(dat$y <= 1, 1, 0)
dat$y12vs34 <- ifelse(dat$y <= 2, 1, 0)
dat$y123vs4 <- ifelse(dat$y <= 3, 1, 0)

fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "logit"))
fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "logit"))
fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "logit"))

car::compareCoefs(fit123vs4, fit12vs34, fit1vs234)
```

### 2023-12-19 R/deprecated.R

@@ -8,6 +8,15 @@ get_y <- function(data = NULL, P, ordered = TRUE){
   }
 }
 
+codds <- function(p){
+  co <- cumsum(p)/(1 - cumsum(p))
+  co <- co[-length(co)]
+  nn <- Reduce(paste0, 1:(length(p) - 1), accumulate = TRUE)
+  names(co) <- paste0("o", nn)
+  co
+}
+
+
 show_th <- function(th = NULL, probs = NULL, link = c("logit", "probit"),
                     cnames = NULL){
   link <- match.arg(link)
@@ -46,4 +55,42 @@ show_th <- function(th = NULL, probs = NULL, link = c("logit", "probit"),
           names = cnames)
   
   par(mfrow = c(1,1))
+}
+
+dummy_ord <- function(y){
+  if(!is.numeric(y)) y <- as.integer(y)
+  yc <- sort(unique(y))
+  dummy <- lapply(yc, function(t) ifelse(y <= t, 1, 0))
+  nn1 <- Reduce(paste0, yc, accumulate = TRUE)
+  nn2 <- Reduce(paste0, yc, accumulate = TRUE, right = TRUE)
+  names(dummy) <- sprintf("y%svs%s", nn1[-length(nn1)], nn2[-1])
+  data.frame(dummy[-length(dummy)])
+}
+
+clm_to_ord <- function(fit){
+  th <- unname(fit$alpha) # estimated thresholds
+  coefs <- fit$coefficients
+  coefs <- unname(coefs[(length(th) + 1):length(coefs)])
+  k <- length(th) + 1 # number of levels for y
+  y <- 1:k # ordinal values
+  stdm <- 0 # latent mean
+  stds <- 1 # latent sigma
+  th_y <- scales::rescale(th, to = c(y[1] + 0.5, y[k] - 0.5))
+  lats <- (th_y[2] - th_y[k - 2]) / (th[2] - th[k - 2]) # real latent sigma
+  b0 <- th_y[1] - th[1] * lats
+  betas <- lats * coefs
+  out <- list(sigma = lats, b0 = b0, beta = betas, alpha = th_y)
+  return(out)
+}
+
+get_th <- function(b0, b1){
+  -(b0/b1)
+}
+
+get_slope <- function(b1){
+  1/b1
+}
+
+get_b0_from_th <- function(th, b1){
+  -b1 * th
 }

### 2023-12-19 R/utils-paper.R

@@ -65,10 +65,54 @@ th_names <- function(k){
   sprintf("%s|%s", 1:(k-1), 2:k)
 }
 
-plot_grid_common_legend <- function(...){
+plot_grid_common_legend <- function(..., nrow = NULL, ncol = NULL){
   plts <- list(...)
   legend <- cowplot::get_legend(plts[[1]] + theme(legend.position = "bottom", legend.title = element_blank()))
   plts <- lapply(plts, function(p) p + theme(legend.position = "none"))
   grid <- cowplot::plot_grid(plotlist = plts)
-  cowplot::plot_grid(grid, legend, nrow = 2, rel_heights = c(0.9, 0.1))
+  cowplot::plot_grid(grid, legend, nrow = nrow, ncol = ncol, rel_heights = c(0.9, 0.1))
+}
+
+mtab <- function(data, type = c("flextable", "kable", "apa"), caption = NULL, digits = 3){
+  type <- match.arg(type)
+  if(type == "flextable"){
+    qflex(data, caption, digits)
+  }else if(type == "kable"){
+    qkable(data, caption, digits)
+  }else{
+    qapa(data, caption)
+  }
+}
+
+qkable <- function(data, caption = NULL, digits = 3){
+  require(kableExtra)
+  data |> 
+    kable(booktabs = TRUE,
+          format = "latex",
+          align = "c",  
+          digits = digits, 
+          caption = caption) |> 
+    kable_styling(full_width = FALSE,
+                  position = "center")
+}
+
+qflex <- function(data, caption = NULL, digits = 3){
+  require(flextable)
+  data |> 
+    flextable() |> 
+    theme_booktabs(bold_header = TRUE) |> 
+    autofit() |> 
+    align(part = "all", align = "center") |> 
+    set_caption(caption = caption) |> 
+    colformat_double(digits = digits)
+}
+
+qapa <- function(data, caption){
+  require(papaja)
+  data |> 
+    apa_table(caption = caption)
+}
+
+rm_legend <- function(){
+  theme(legend.position = "none")
 }

### 2023-12-19 R/utils-plotting.R

@@ -1,0 +1,199 @@
+show_alpha <- function(prob = NULL, 
+                       alpha = NULL, 
+                       link = c("logit", "probit"),
+                       plot = c("latent", "cumulative", "probs")){
+  require(ggplot2)
+  require(cowplot)
+  require(ggtext)
+  
+  if(is.null(prob) & is.null(alpha)){
+    stop("alpha or prob must be specified!")
+  }
+  
+  lf <- get_link(link)
+  
+  if(link == "logit"){
+    xlim <- c(-7, 7)
+    dist <- distributional::dist_logistic
+  }else{
+    xlim <- c(-5, 5)
+    dist <- distributional::dist_normal
+  }
+  
+  if(is.null(prob)){
+    prob <- alpha_to_prob(alpha, link)
+  }else{
+    if(sum(prob) != 1){
+      stop("the prob vector must sum to 1!")
+    }
+    alpha <- prob_to_alpha(prob, link)
+  }
+  
+  k <- length(prob)
+  D <- data.frame(l = 0, s = 1)
+  D$dist <- dist(D$l, D$s)
+  
+  alpha_n <- th_names(k)
+  
+  dplot <- ggplot(D) +
+    ggdist::stat_slab(aes(dist = dist,
+                          fill = factor(after_stat(findInterval(x, alpha) + 1))),
+                      orientation = "horizontal",
+                      alpha = 0.8) +
+    xlim(xlim) +
+    theme_minimal(15) +
+    theme(legend.title = element_blank(),
+          legend.position = "bottom",
+          axis.title.y = element_blank(),
+          axis.title.x = element_blank()) +
+    geom_vline(xintercept = alpha,
+               alpha = 0.5) +
+    ggtext::geom_richtext(data = data.frame(alpha, alpha_n),
+                          aes(x = alpha, 
+                              y = 1, 
+                              label = alpha_n), 
+                          angle = 90,
+                          hjust = 0.5)
+  
+  y_n <- sprintf("P(Y = %s)", 1:k)
+  y_cp <- lf$pfun(c(-Inf, alpha, Inf))
+  y_p <- 0.5 * (y_cp[1:(length(y_cp) - 1)] + y_cp[2:length(y_cp)])
+  
+  pplot <- ggplot() +
+    stat_function(geom = "line", fun = lf$pfun) +
+    xlim(xlim) +
[…]

### 2023-12-19 R/utils.R

@@ -1,176 +1,3 @@
-show_alpha <- function(prob = NULL, 
-                       alpha = NULL, 
-                       link = c("logit", "probit")){
-  require(ggplot2)
-  require(cowplot)
-  require(ggtext)
-  
-  if(is.null(prob) & is.null(alpha)){
-    stop("alpha or prob must be specified!")
-  }
-  
-  lf <- get_link(link)
-  
-  if(link == "logit"){
-    xlim <- c(-7, 7)
-    dist <- distributional::dist_logistic
-  }else{
-    xlim <- c(-5, 5)
-    dist <- distributional::dist_normal
-  }
-  
-  if(is.null(prob)){
-    prob <- alpha_to_prob(alpha, link)
-  }else{
-    alpha <- prob_to_alpha(prob, link)
-  }
-  
-  k <- length(prob)
-  D <- data.frame(l = 0, s = 1)
-  D$dist <- dist(D$l, D$s)
-  
-  alpha_n <- th_names(k)
-  
-  dplot <- ggplot(D) +
-    ggdist::stat_slab(aes(dist = dist,
-                          fill = factor(after_stat(findInterval(x, alpha) + 1))),
-                      orientation = "horizontal",
-                      alpha = 0.8) +
-    xlim(xlim) +
-    theme_minimal(15) +
-    theme(legend.title = element_blank(),
-          legend.position = "bottom",
-          axis.title.y = element_blank(),
-          axis.title.x = element_blank()) +
-    geom_vline(xintercept = alpha,
-               alpha = 0.5) +
-    ggtext::geom_richtext(data = data.frame(alpha, alpha_n),
-                          aes(x = alpha, 
-                              y = 1, 
-                              label = alpha_n), 
-                          angle = 90,
-                          hjust = 0.5)
-  
-  y_n <- sprintf("P(Y = %s)", 1:k)
-  y_cp <- lf$pfun(c(-Inf, alpha, Inf))
-  y_p <- 0.5 * (y_cp[1:(length(y_cp) - 1)] + y_cp[2:length(y_cp)])
-  
-  pplot <- ggplot() +
-    stat_function(geom = "line", fun = lf$pfun) +
-    xlim(xlim) +
-    geom_segment(aes(x = alpha, y = 0, xend = alpha, yend = lf$pfun(alpha))) +
[…]

### 2023-12-19 scripts-review/aaa.R

@@ -1,0 +1,8 @@
+library(tidyverse)
+
+
+
+mtab(iris[1:5, ], "apa")
+
+
+

### 2023-12-21 paper/paper-new.Rmd
update

@@
 We can class this general class of models as *ordinal regression*.
-The nomenclature of these models can be confusing mainly because there are several subclasses of models with different assumptions and structures [@Tutz2022-dg].
+The nomenclature of these models can be confusing mainly because there are several types of models with different assumptions and structures [@Tutz2022-dg].
 @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
@@
 In this section we introduce some notations for the cumulative model that is used through the paper and in the R code.
-The notation is as consistent as possible with the main textbooks and with the `ordinal` R package used in the tutorial.
-We define $Y_k$ as the observed ordinal response with $1, 2, \dots,k$ options and $Y^\star$ is the corresponding latent variable.
-The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \alpha_2, \dots, \alpha_{k - 1}$.
-With $g()$ we define the cumulative probability function assumed for the specific model.
-For example, when using a Gaussian distribution we are fitting a model with a `probit` link function thus $g() = \Phi()$.
-With $g^{-1}()$ we define the inverse of the cumulative probability function.
-The minus sign in $\mathbf{X} \beta$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).
+We proposed a notation as consistent as possible with the main textbooks and seminal papers and with the `ordinal` R package used in the tutorial.
+We define $Y_k$ as the observed ordinal response with $1, \dots,k$ levels and $Y^\star$ is the underlying latent variable.
+The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \dots, \alpha_{k - 1}$.
+Similarly to the generalized linear models framework, we define $g(x) = \eta$ as the link function that maps that transforms probabilities into the linear predictor $\eta$.
+To transform $\eta$ into probabilities we use the inverse of the link function $x = g^{-1}(\eta)$.
+The specific link function define the type of model and require a different R function.
+For example, when assuming a Gaussian distribution we are fitting a model with a `probit` link function is the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
+When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y < k), k = 1, \dots, k - 1$.
+Equation \@ref(eq:prob-cum-model1) depict the general cumulative model including predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
+The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).
 
 \begin{equation} 
-P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, \dots, k - 1
+P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}), \;\;k = 1, \dots, k - 1
 (\#eq:prob-cum-model1)
@@
 
-We can call the $\mathbf{X} \beta$ as the linear predictor $\eta$. Equation (num) describe how the probability of a single outcome $Y_k$.
+The $\mathbf{X} \boldsymbol{\beta}$ is the linear predictor $\eta$ that is the cumulative probability $P(Y \leq k)$ transformed using the link function $g()$.
+To obtain the probability of a single outcome $P(Y = k)$ we can compute the difference between cumulative probabilities as shown in Equation \@ref(eq:eq:prob-cum-model2).
 
 \begin{equation} 
-P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta)
+P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta), \;\;k = 1, \dots, k - 1
 (\#eq:prob-cum-model2)
@@
 
+There are always two special cases when computing the probability of a single outcome $Y$ that is when $Y = 1$ and $Y = k$.
+In the first case the cumulative probability is calculated from $-\infty$ that is the same as temporary assuming an $\alpha_0 = -\infty$.
+In the second case ($Y = 1$) the probability is calculated as $P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)$ that is the same as assuming a temporary threshold $\alpha_k = +\infty$.
+The Figure \@ref(fig:fig-explain-cumulative) shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.
+
+```{r fig-explain-cumulative, fig.cap=filor::get_caption()}
+probs <- c(0.4, 0.2, 0.3, 0.1)
+th <- prob_to_alpha(probs, link = "probit")
+lat <- data.frame(m = 0, s = 1)
+lat$norm <- distributional::dist_normal(lat$m, lat$s)
+
+thl <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", 0:length(probs)))
+
+ggplot(lat, aes(dist = norm)) +
+  ggdist::stat_slab(orientation = "horizontal",
+                    aes(fill = after_stat(sprintf("P(Y = %s)", findInterval(x, th) + 1)))) +
+  geom_vline(xintercept = c(-4, th, 4)) +
+  theme_minimal(15) +
+  theme(legend.position = "bottom",
+        legend.title = element_blank(),
+        axis.text.x = element_text(size = 17),
+        axis.title.x = element_blank(),
+        axis.title.y = element_blank()) +
+  annotate("label", x = th, y = 0, label = thl[2:(length(thl) - 1)], size = 6) +
+  annotate("label", x = c(-4, 4), y = 0, label = thl[c(1, length(thl))], size = 6, fill = "grey") +
+  scale_x_continuous(breaks = seq(-4, 4, 2), 
+                     labels = c(latex2exp::TeX("$-\\infty$"), "-2", "0", "2", latex2exp::TeX("$\\infty$")))
+```
+
 <!-- notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw per la parte di notazione e quella più formale anche @DeCarlo2010-lj -->
 
-The same model can be formalized using the latent formulation (see Equation). This is more similar to a standard linear regression.
+The model presented in Equation \@ref(eq:prob-cum-model1) can be considered similar to a standard binomial generalized linear model.
+In this case we have $k - 1$ equations, one for each threshold $\alpha$ and we model the cumulative probability.
+The same model can be written as in Equation \@ref(eq:latent-model) assuming a continuous latent variable $Y^\star$ explained by the linear combination of predictors $\eta = \mathbf{X} \boldsymbol{\beta}$ as in a standard linear regression model.
 
@@
 
-Where $\epsilon$ is the error part and comes from the distribution that is assumed from the latent model.
+The crucial part is $\epsilon_i$ that is the random part of the model coming from a certain probability distribution.
 For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution.
-Then the observed ordinal value $Y_i$ comes from $Y^\star_i$ being between the thresholds $\alpha_{k - 1} < Y^\star_i < \alpha_{k}$ with thresholds organized as $- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty$.
-<!-- TODO rivedi questa -->
-
-In the basic version of the model, the thresholds $\alpha$ are considered as fixed and do not vary as a function of predictors.
-The thresholds are part of the measurement procedure [@Liddell2018-wu].
-In a more sophisticated version of the model called location-shift [@Tutz2022-dg], both the location $\mu$ and the thresholds $\alpha$ can vary as a function of the predictors.
-In the following sections another version of the model called location-scale [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur] will be presented as a more flexible way to model latent distributions with heterogeneous variance (spiega meglio).
+Following the notation by @Tutz2022-dg, the observed ordinal value $Y_i = k$ comes from $Y^\star_i$ belonging to the interval defined by the thresholds $Y_i = k  \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}$ where $- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty$.
+
+In the basic version of the model, the thresholds $\alpha_k$ are considered as fixed and being part of the measurement procedure [@Liddell2018-wu] and do not vary as a function of the predictors.
+In a more sophisticated version of the model called location-shift [@Tutz2022-dg], both the location $\mu$ and the thresholds $\alpha_k$ can vary as a function of the predictors.
+@Tutz2022-dg described also another version of the model called location-scale [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur] where the location $\mu$ and the scale $\sigma^2$ of the distribution can vary as a function of the predictors.
 
@@
 Several parametrizations of the ordinal models have been proposed by different authors.
-@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian probit model where the latent mean and standard deviations are estimated along with the threshold.
-The interesting part is that the thresholds are transformed to be on the scale of the variable (e.g, from 1 to 4 when $k = 4$) and not in terms of $z$ scores.
+@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian *probit* model where the latent mean and standard deviations are estimated along with the threshold.
+The interesting part is that the thresholds are transformed to be on the scale of the ordinal variable (e.g, from 1 to 4 when $k = 4$) and not in terms of $z$ scores.
 @Kurz2023-zv explain how to convert a model fitted with the `brms` package (citation) (an R package for regression modeling using *stan*) into the Kruschke [-@Kruschke2015-re] parametrization.
-
-@Gelman2020-tg proposed three alternative parametrizations focusing on different definition of the thresholds and.
-
-vedi qualcosa in più qui
+@Gelman2020-tg proposed three alternative parametrizations focusing on different definition of the thresholds.
+The current tutorial is focused on the standard parametrization of the cumulative link model as described in the equations above.
+
+## Link function
+
+The cumulative link model implemented in Equations \@ref(eq:prob-cum-model1) and \@ref(eq:latent-model) can be considered the general formulation that requires specifying the link function $g(\cdot)$ or the latent distribution $\epsilon_i \sim D(\mu, \sigma^2)$.
+Among several available functions the *logit* and *probit* models are the most common.
+The *logit* model assume a *logit* link function and a logistic distribution as latent variable.
+On the other side, the *probit* model assume a Gaussian distribution.
+The two models provide similar results but the parameters intepretation and the way data can be simulated is different.
+In the next sections we will illustrate the differences and simulation strategies.
+Figure \@ref(fig:fig-logit-vs-probit) depicts the two distributions while Table \@ref(tab:tab-model-summary) summarise the presented cumulative models with the proposed link function and the corresponding R code.
+In terms of parameters, both distributions can be defined with a location $\mu$ and a scale $s$ parameter.
+The standard normal distribution has $\mu = 0$ and $s = 1$.
+Furthermore the variance corresponds to the scale $s^2 = \sigma^2 = 1$.
+The variance of the logistic distribution is $\sigma^2 = \frac{s^2\pi^2}{3}$.
+The standard logistic distribution has $\mu = 0$ and $s^2 = 1$ thus the standard deviation simplified to  $\frac{\pi}{\sqrt{3}} \approx 1.81$.
+In practical terms, the standard version of the two distributions are very similar with the logistic having more variance.
+The R implementation of these functions consist in the `q` and ``
+
+```{r fig-logit-vs-probit, fig.cap=filor::get_caption()}
+dplot <- ggplot(data = NULL) +
+  stat_function(fun = dnorm, aes(color = "Normal"), lwd = 1) +
+  stat_function(fun = dlogis, aes(color = "Logistic"), lwd = 1) +
+  xlim(c(-7, 7)) +
+  theme(legend.position = "bottom")
+
+pplot <- ggplot(data = NULL) +
+  stat_function(fun = pnorm, aes(color = "Normal"), lwd = 1) +
+  stat_function(fun = plogis, aes(color = "Logistic"), lwd = 1) +
+  xlim(c(-7, 7))
+
+plot_grid_common_legend(
+  dplot, 
+  pplot + theme(axis.title.y = element_blank()), nrow = 2
+)
+```
+
+```{r tab-model-summary-old}
+#| eval: false
+models_table <- data.frame(
+       stringsAsFactors = FALSE,
+                  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
+          Link.Function = c("Equation",
+                            "$\\text{logit}(p) = \\text{log}(\\frac{p}{1-p})$",
+                            "$z = \\Phi(p)$"),
+          Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
+  Inverse.Link.Function = c("Equation",
+                            "$\\frac{e^{\\text{logit}(p)}}{1 + e^{\\text{logit}(p)}}$","$\\Phi^{-1}(z)$"),
+  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
+)
+
+models_table[-1, ] |> 
+  flextable() |> 
+  set_header_labels(values = list(Model = "Model", 
+                                  Link.Function = "Equation",
+                                  Link.Function.1 = "R Code",
+                                  Inverse.Link.Function = "Equation",
+                                  Inverse.Link.Function.1 = "R Code")) |> 
+  add_header_row(values = c("", "Link Function", "Link Function", "Inverse Link Function", "Inverse Link Function")) |> 
+  merge_h(1, part = "header") |> 
+  theme_booktabs(bold_header = TRUE) |> 
+  align(part = "all", align = "center") |> 
+  set_caption(caption = filor::get_caption())
+```
+
+```{r tab-model-summary}
+library(kableExtra)
+models_table <- data.frame(
+       stringsAsFactors = FALSE,
+                  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
+          Link.Function = c("Equation",
+                            "$\\text{logit}(p) = \\text{log}(\\frac{p}{1-p})$",
+                            "$z = \\Phi(p)$"),
+          Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
+  Inverse.Link.Function = c("Equation",
+                            "$\\frac{e^{\\text{logit}(p)}}{1 + e^{\\text{logit}(p)}}$","$\\Phi^{-1}(z)$"),
+  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
+)
+
+models_table[-1, ] |> 
+  kable(col.names = c("Model", "Equation", "R Code", "Equation", "R Code"), row.names = FALSE,
+        booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE) |> 
+  kable_styling(full_width = FALSE) |> 
+  add_header_above(header = c("", "Link Function" = 2, "Inverse Link Function" = 2))
+```
 
@@
 Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect.
-As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the intepretation depends on the contrast coding [for an overview about contrasts coding schemes see @Schad2020-ht]^[In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
+As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [for an overview about contrasts coding schemes see @Schad2020-ht]^[In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
 By default, the first level of the factor is taken as the reference level and the $j - 1$ coefficients represent the comparison between the other levels and the reference.]
@@
 The logistic and normal distribution are very similar, with the logistic having more variance.
-Figure \@ref(fig:fig-logit-vs-probit) depicts the differences between the two distributions.
-
-```{r fig-logit-vs-probit, fig.cap=filor::get_caption()}
-dplot <- ggplot(data = NULL) +
-  stat_function(fun = dnorm, aes(color = "Normal"), lwd = 1) +
-  stat_function(fun = dlogis, aes(color = "Logistic"), lwd = 1) +
-  xlim(c(-7, 7))
-
-pplot <- ggplot(data = NULL) +
-  stat_function(fun = pnorm, aes(color = "Normal"), lwd = 1) +
-  stat_function(fun = plogis, aes(color = "Logistic"), lwd = 1) +
-  xlim(c(-7, 7))
-
-plot_grid_common_legend(
-  dplot, 
-  pplot + theme(axis.title.y = element_blank())
-)
-```
 
@@
 
-The Table \@ref(tab:tab-model-summary) summarise the presented cumulative models with the proposed link function and the corresponding R code.
-
-```{r tab-model-summary}
-#| eval: false
-models_table <- data.frame(
-       stringsAsFactors = FALSE,
-                  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
-          Link.Function = c("Equation",
-                            "$\\text{logit}(p) = \\text{log}(\\frac{p}{1-p})$",
-                            "$z = \\Phi(p)$"),
-          Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
-  Inverse.Link.Function = c("Equation",
-                            "$\\frac{e^{\\text{logit}(p)}}{1 + e^{\\text{logit}(p)}}$","$\\Phi^{-1}(z)$"),
-  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
-)
-
-models_table[-1, ] |> 
-  flextable() |> 
-  set_header_labels(values = list(Model = "Model", 
-                                  Link.Function = "Equation",
-                                  Link.Function.1 = "R Code",
-                                  Inverse.Link.Function = "Equation",
-                                  Inverse.Link.Function.1 = "R Code")) |> 
-  add_header_row(values = c("", "Link Function", "Link Function", "Inverse Link Function", "Inverse Link Function")) |> 
-  merge_h(1, part = "header") |> 
-  theme_booktabs(bold_header = TRUE) |> 
-  align(part = "all", align = "center") |> 
-  set_caption(caption = filor::get_caption())
-```
-
 ## Simulating data
@@
 3. Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
-4. Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\mathbf{\beta}$.
+4. Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
 5. Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
@@
 4.
-Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$, the regression coefficients $\mathbf{\beta}$ and the error $\mathbf{\epsilon}$ sampled from the appropriate probability distribution.
+Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$, the regression coefficients $\boldsymbol{\beta}$ and the error $\mathbf{\epsilon}$ sampled from the appropriate probability distribution.
 5.
@@
 (ref:fig-example-num-latent) caption here
+
+(ref:fig-explain-cumulative) caption here

### 2023-12-21 paper/paper-new.tex
update

@@
 We can class this general class of models as \emph{ordinal regression}.
-The nomenclature of these models can be confusing mainly because there are several subclasses of models with different assumptions and structures (Gerhard Tutz, 2022).
+The nomenclature of these models can be confusing mainly because there are several types of models with different assumptions and structures (Gerhard Tutz, 2022).
 Gerhard Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
@@
 In this section we introduce some notations for the cumulative model that is used through the paper and in the R code.
-The notation is as consistent as possible with the main textbooks and with the \texttt{ordinal} R package used in the tutorial.
-We define \(Y_k\) as the observed ordinal response with \(1, 2, \dots,k\) options and \(Y^\star\) is the corresponding latent variable.
-The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \alpha_2, \dots, \alpha_{k - 1}\).
-With \(g()\) we define the cumulative probability function assumed for the specific model.
-For example, when using a Gaussian distribution we are fitting a model with a \texttt{probit} link function thus \(g() = \Phi()\).
-With \(g^{-1}()\) we define the inverse of the cumulative probability function.
-The minus sign in \(\mathbf{X} \beta\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).
+We proposed a notation as consistent as possible with the main textbooks and seminal papers and with the \texttt{ordinal} R package used in the tutorial.
+We define \(Y_k\) as the observed ordinal response with \(1, \dots,k\) levels and \(Y^\star\) is the underlying latent variable.
+The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\).
+Similarly to the generalized linear models framework, we define \(g(x) = \eta\) as the link function that maps that transforms probabilities into the linear predictor \(\eta\).
+To transform \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\).
+The specific link function define the type of model and require a different R function.
+For example, when assuming a Gaussian distribution we are fitting a model with a \texttt{probit} link function is the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
+When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y < k), k = 1, \dots, k - 1\).
+Equation \eqref{eq:prob-cum-model1} depict the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
+The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).
 
 \begin{equation} 
-P(Y \leq k) = g(\alpha_k - \mathbf{X} \mathbf{\beta}), \;\;k = 1, \dots, k - 1
+P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}), \;\;k = 1, \dots, k - 1
 \label{eq:prob-cum-model1}
@@
 
-We can call the \(\mathbf{X} \beta\) as the linear predictor \(\eta\). Equation (num) describe how the probability of a single outcome \(Y_k\).
+The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g()\).
+To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:eq:prob-cum-model2}.
 
 \begin{equation} 
-P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta)
+P(Y = k) = g^{-1}(\alpha_k - \eta) -  g^{-1}(\alpha_{k - 1} - \eta), \;\;k = 1, \dots, k - 1
 \label{eq:prob-cum-model2}
@@
 
-The same model can be formalized using the latent formulation (see Equation). This is more similar to a standard linear regression.
+There are always two special cases when computing the probability of a single outcome \(Y\) that is when \(Y = 1\) and \(Y = k\).
+In the first case the cumulative probability is calculated from \(-\infty\) that is the same as temporary assuming an \(\alpha_0 = -\infty\).
+In the second case (\(Y = 1\)) the probability is calculated as \(P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)\) that is the same as assuming a temporary threshold \(\alpha_k = +\infty\).
+The Figure \ref{fig:fig-explain-cumulative} shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-explain-cumulative-1}
+
+}
+
+\caption{caption here}\label{fig:fig-explain-cumulative}
+\end{figure}
+
+\normalsize
+
+The model presented in Equation \eqref{eq:prob-cum-model1} can be considered similar to a standard binomial generalized linear model.
+In this case we have \(k - 1\) equations, one for each threshold \(\alpha\) and we model the cumulative probability.
+The same model can be written as in Equation \eqref{eq:latent-model} assuming a continuous latent variable \(Y^\star\) explained by the linear combination of predictors \(\eta = \mathbf{X} \boldsymbol{\beta}\) as in a standard linear regression model.
 
@@
 
-Where \(\epsilon\) is the error part and comes from the distribution that is assumed from the latent model.
+The crucial part is \(\epsilon_i\) that is the random part of the model coming from a certain probability distribution.
 For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution.
-Then the observed ordinal value \(Y_i\) comes from \(Y^\star_i\) being between the thresholds \(\alpha_{k - 1} < Y^\star_i < \alpha_{k}\) with thresholds organized as \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).
-
-In the basic version of the model, the thresholds \(\alpha\) are considered as fixed and do not vary as a function of predictors.
-The thresholds are part of the measurement procedure (Liddell \& Kruschke, 2018).
-In a more sophisticated version of the model called location-shift (Gerhard Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha\) can vary as a function of the predictors.
-In the following sections another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Gerhard Tutz, 2022) will be presented as a more flexible way to model latent distributions with heterogeneous variance (spiega meglio).
+Following the notation by Gerhard Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).
+
+In the basic version of the model, the thresholds \(\alpha_k\) are considered as fixed and being part of the measurement procedure (Liddell \& Kruschke, 2018) and do not vary as a function of the predictors.
+In a more sophisticated version of the model called location-shift (Gerhard Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha_k\) can vary as a function of the predictors.
+Gerhard Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Gerhard Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.
 
@@
 Several parametrizations of the ordinal models have been proposed by different authors.
-Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian probit model where the latent mean and standard deviations are estimated along with the threshold.
-The interesting part is that the thresholds are transformed to be on the scale of the variable (e.g, from 1 to 4 when \(k = 4\)) and not in terms of \(z\) scores.
+Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian \emph{probit} model where the latent mean and standard deviations are estimated along with the threshold.
+The interesting part is that the thresholds are transformed to be on the scale of the ordinal variable (e.g, from 1 to 4 when \(k = 4\)) and not in terms of \(z\) scores.
 Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.
-
-Gelman, Hill, and Vehtari (2020) proposed three alternative parametrizations focusing on different definition of the thresholds and.
-
-vedi qualcosa in più qui
+Gelman, Hill, and Vehtari (2020) proposed three alternative parametrizations focusing on different definition of the thresholds.
+The current tutorial is focused on the standard parametrization of the cumulative link model as described in the equations above.
+
+\subsection{Link function}\label{link-function}
+
+The cumulative link model implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} can be considered the general formulation that requires specifying the link function \(g(\cdot)\) or the latent distribution \(\epsilon_i \sim D(\mu, \sigma^2)\).
+Among several available functions the \emph{logit} and \emph{probit} models are the most common.
+The \emph{logit} model assume a \emph{logit} link function and a logistic distribution as latent variable.
+On the other side, the \emph{probit} model assume a Gaussian distribution.
+The two models provide similar results but the parameters intepretation and the way data can be simulated is different.
+In the next sections we will illustrate the differences and simulation strategies.
+Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table \ref{tab:tab-model-summary} summarise the presented cumulative models with the proposed link function and the corresponding R code.
+In terms of parameters, both distributions can be defined with a location \(\mu\) and a scale \(s\) parameter.
+The standard normal distribution has \(\mu = 0\) and \(s = 1\).
+Furthermore the variance corresponds to the scale \(s^2 = \sigma^2 = 1\).
+The variance of the logistic distribution is \(\sigma^2 = \frac{s^2\pi^2}{3}\).
+The standard logistic distribution has \(\mu = 0\) and \(s^2 = 1\) thus the standard deviation simplified to \(\frac{\pi}{\sqrt{3}} \approx 1.81\).
+In practical terms, the standard version of the two distributions are very similar with the logistic having more variance.
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-logit-vs-probit-1}
+
+}
+
+\caption{caption here}\label{fig:fig-logit-vs-probit}
+\end{figure}
+
+\normalsize
+
+\scriptsize
+
+\normalsize
+
+\scriptsize
+
+\begin{table}
+
+\caption{\label{tab:tab-model-summary}caption here}
+\centering
+\begin{tabular}[t]{lllll}
+\toprule
+\multicolumn{1}{c}{} & \multicolumn{2}{c}{Link Function} & \multicolumn{2}{c}{Inverse Link Function} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Model & Equation & R Code & Equation & R Code\\
+\midrule
+Cumulative Logit & $\text{logit}(p) = \text{log}(\frac{p}{1-p})$ & `qlogis()` & $\frac{e^{\text{logit}(p)}}{1 + e^{\text{logit}(p)}}$ & `plogis()`\\
+Cumulative Probit & $z = \Phi(p)$ & `qnorm()` & $\Phi^{-1}(z)$ & `pnorm()`\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\normalsize
 
@@
 1 & a & 4\\
-2 & a & 2\\
-3 & a & 1\\
-4 & a & 3\\
+2 & a & 3\\
+3 & a & 2\\
+4 & a & 2\\
 ... & ... & ...\\
 \addlinespace
-97 & b & 1\\
+97 & b & 3\\
 98 & b & 3\\
-99 & b & 4\\
+99 & b & 2\\
 100 & b & 4\\
@@
 ##    a    b 
-## 2.32 3.18 
+## 2.52 3.04 
 ## 
@@
 ##   a   b 
-## 2 3 
+## 2.0 3.5 
 ## 
@@
 ##        a        b 
-## 1.0961901 0.9833305
+## 1.092460 1.124132
 \end{verbatim}
@@
 ##        a    b
-##   1 0.30 0.10
-##   2 0.26 0.10
-##   3 0.26 0.32
-##   4 0.18 0.48
+##   1 0.20 0.14
+##   2 0.34 0.18
+##   3 0.20 0.18
+##   4 0.26 0.50
 \end{verbatim}
@@
 ##  link  threshold nobs logLik  AIC    niter max.grad cond.H 
-##  logit flexible  100  -127.63 263.27 4(0)  1.25e-08 1.6e+01
+##  logit flexible  100  -130.61 269.23 5(0)  2.61e-13 2.0e+01
 ## 
@@
 ##    Estimate Std. Error z value Pr(>|z|)  
-## xb   1.4974     0.3885   3.855 0.000116 ***
+## xb   0.8795     0.3719   2.365    0.018 *
 ## ---
@@
 ##     Estimate Std. Error z value
-## 1|2  -0.8124     0.2891  -2.810
-## 2|3   0.1867     0.2724   0.685
-## 3|4   1.5612     0.3206   4.869
+## 1|2  -1.2246     0.3040  -4.029
+## 2|3   0.1231     0.2653   0.464
+## 3|4   0.9396     0.2841   3.308
 \end{verbatim}
@@
 Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect.
-As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the intepretation depends on the contrast coding (for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, \& Kliegl, 2020)\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
+As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding (for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, \& Kliegl, 2020)\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
 By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}
@@
 ##        xa        xb 
-## -0.9991102 -0.9991102
+## -1.347649 -1.347649
 \end{verbatim}
@@
 ##         xa         xb 
-## -0.5888457 -0.5888457
+## -0.7963604 -0.7963604
 \end{verbatim}
@@
 The logistic and normal distribution are very similar, with the logistic having more variance.
-Figure \ref{fig:fig-logit-vs-probit} depicts the differences between the two distributions.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/fig-logit-vs-probit-1}
-
-}
-
-\caption{caption here}\label{fig:fig-logit-vs-probit}
-\end{figure}
-
-\normalsize
 
@@
 
-The Table \ref{tab:tab-model-summary} summarise the presented cumulative models with the proposed link function and the corresponding R code.
-
-\scriptsize
-
-\normalsize
-
 \subsection{Simulating data}\label{simulating-data}
@@
 \item
-  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\) and regression coefficients \(\mathbf{\beta}\).
+  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
 \item
@@
 ##    p_y1  p_y2  p_y3  p_y4
-## 1 0.578  0.23 0.131 0.061
-## 2  0.15 0.209 0.264 0.377
-## 3 0.048 0.112 0.215 0.625
-## 4 0.429 0.261 0.189 0.121
+## 1 0.088 0.161  0.25 0.502
+## 2 0.296 0.259 0.237 0.208
+## 3 0.265 0.254 0.246 0.235
+## 4 0.239 0.247 0.253 0.261
 ## 5   ...   ...   ...   ...
-## 6 0.144 0.205 0.264 0.387
-## 7 0.198 0.233 0.261 0.308
-## 8 0.207 0.236 0.259 0.298
-## 9 0.107 0.178 0.257 0.457
+## 6 0.143 0.205 0.264 0.389
+## 7 0.167 0.218 0.264  0.35
+## 8 0.338 0.263 0.223 0.176
+## 9 0.265 0.253 0.246 0.236
 \end{verbatim}
@@
 \item
-  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\), the regression coefficients \(\mathbf{\beta}\) and the error \(\mathbf{\epsilon}\) sampled from the appropriate probability distribution.
+  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\), the regression coefficients \(\boldsymbol{\beta}\) and the error \(\mathbf{\epsilon}\) sampled from the appropriate probability distribution.
 \item

### 2023-12-22 paper/paper-new.Rmd
updating

@@
 Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
-(rivedi questo).
 An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
@@
 qualcosa ancora qui
+
+aggiungi @Cliff1996-qo per considerazioni ancora riguardo dati ordinali in psicologia e misure appropriate di correlazione e differenza
+
+> We usually do not want our conclusions to be confined to the current, somewhat arbitrary version of the variables; moreover, these are often assumed to be manifest versions of latent variables that are nonlinearly related to them.
+
+@Agresti2010-rz
+
+This approach assigns numerical scores to the ordered categories and then
+uses ordinary least squares (OLS) methods such as linear regression and analysis
+of variance (ANOVA). The second way restricts analyses solely to methods that
+use only the ordering information about the categories. Examples of this approach
+are nonparametric methods based on ranks and models for cumulative response
+probabilities.
+
+That being said, we recommend against the simplistic approach of posing linear
+regression models for ordinal response scores and fitting them using OLS methods.
+Although that approach can be useful for identifying variables that clearly affect
+a response variable, and for simple descriptions, limitations occur. First, there is
+usually not a clear-cut choice for the scores. Second, a particular response outcome
+is likely to be consistent with a range of values for some underlying latent variable,
+and an ordinary regression analysis does not allow for the measurement error that
+results from replacing such a range by a single numerical value. Third, unlike the
+methods presented in this book, that approach does not yield estimated probabilities
+for the response categories at fixed settings of the explanatory variables. Fourth, that
+approach can yield predicted values above the highest category score or below the
+lowest. Fifth, that approach ignores the fact that the variability of the responses is
+naturally nonconstant for categorical data: For an ordinal response variable, there is
+little variability at predictor values for which observations fall mainly in the highest
+category (or mainly in the lowest category), but there is considerable variability at
+predictor values for which observations tend to be spread among the categories.
 
@@
 The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).
+
+@Agresti2010-rz
+For simplicity of notation, unless we need to refer to
+particular subjects or to particular values of the explanatory variables, we replace
+P (Yi ≤ j | xi) in such equations by P (Y ≤ j ), keeping in mind that in the model
+this is actually a conditional probability at each fixed value for the explanatory
+variables.
 

### 2024-01-03 paper/paper-new.Rmd
update

--- 2023-12-21 paper/paper-new.Rmd
@@
 library(here)
-library(flextable)
+library(kableExtra)
 library(cowplot)
@@
 # figures
-source(here("scripts", "figures.R"))
+source(here("scripts", "figures.R"), local = knitr::knit_global())
 rfigs <- readRDS(here("objects", "r-figs.rds"))
+
+# tables
+source(here("scripts", "figures.R"), local = knitr::knit_global())
+rtabs <- readRDS(here("objects", "r-tabs.rds"))
 ```
@@
 Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
-(rivedi questo).
 An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
 Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer.
-@Stevens1946-te and @Kemp2021-dj suggested that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation).
+@Stevens1946-te and @Kemp2021-dj suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools [e.g., @Cliff1996-qo; @Agresti2010-rz].
 This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data.
 Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels [@Liddell2018-wu].
+More generally, @Cliff2016-ck suggest that most of the research questions in behavioral sciences can be considered as ordinal (*is the score* $x$ higher than the score $y$?) concerning variables where the most appropriate measurement scale is probably ordinal.
 
@@
 In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric [but see @Robitzsch2020-la for an alternative perspective].
-They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui).
-(vedi se c'è qualche altro lavoro che fa vedere questa cosa).
-Some authors suggested that despite individual items being ordinal, averaging multiple items could solve the problem of applying metric models (e.g., Carifio & Perla, 2007, 2008 metti queste citazioni).
-@Liddell2018-wu suggested that in some conditions even averaging multiple items and applying metric models could be problematic.
-@Cliff2016-ck suggest that most of the research questions in behavioral sciences can be considered as ordinal (*is the score $x$ higher than the score $y$*?) concerning variables where the most appropriate measurement scale is probably ordinal.
-
-qualcosa ancora qui
+They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and distorted effect size estimates.
+Some authors suggested that individual ordinal items and a collection of ordinal (averaged or summed) items can safely be considered and analyzed as metric [@Carifio2008-qd; @Carifio2007-kk; but see @Jamieson2004-xf].
+Despite @Liddell2018-wu provide some example that averaging ordinal items and applying metric models is not appropriate, in the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.).
 
@@
 
-Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without making metric assumptions.
-We can class this general class of models as *ordinal regression*.
-The nomenclature of these models can be confusing mainly because there are several types of models with different assumptions and structures [@Tutz2022-dg].
+Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions.
+We can name this general class of models as *ordinal regression*.
+The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures [@Tutz2022-dg].
 @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
-We can identify three main classes: *cumulative models* [@McCullagh1980-cw], *sequential models* [@Tutz1990-fe] and *adjacent category models*.
-The cumulative is the mostly used model assuming the existence of a latent variable that when segmented using thresholds produces the observed ordinal variable.
-The psychological process underlying the response is clearly formalized in the signal detection theory framework where the respondent (vedi come sistemare).
-The sequential model as suggested by the name is appropriate when modelling sequential processes.
-Assuming to have 5 response options, the sequential model assume that responding 3 means already reaching the states 1 and 2.
+We can identify three main classes: *cumulative models* [CM; @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe] and *adjacent category models*.
+The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.
+
+The *sequential model* as suggested by the name is appropriate when modelling sequential processes.
+Assuming to have five response options, the sequential model assume that responding "3" assume a sequential process where steps "1" and "2" are already reached.
 A clear example is proposed by @Burkner2019-aw where the marriage duration in years is predicted as a function of some explanatory variables.
-For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential regression.
+For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential model.
 When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached [@Tutz2020-xq].
@@
 
-In the current paper we put the focus on the cumulative model for several reasons.
+In the current paper we put the focus on the CM for several reasons.
 The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation.
-The second reason is that several psychological variables can be formalized as a latent continuous construct that when measured is collected using ordinal items.
-Furthermore, also the signal detection theory framework models the decision process assuming a latent distribution for the signal and noise where thresholds are the response criteria [e.g., @DeCarlo2010-lj].
+The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item.
+Furthermore, CM are also used to model data under a signal detection theory framework [e.g., @DeCarlo2010-lj].
+The Figure \@ref(fig:fig-ordinal-models) depict the overall structure of the *cumulative* model.
 
@@
 
-metti qualcosa sulla nomenclatura
-
-> The name, cumulative link models is adopted from Agresti (2002), but the model class has
-been referred to by several other names in the literature, such as ordered logit models and
-ordered probit models (Greene and Hensher 2010) for the logit and probit link functions. The
-cumulative link model with a logit link is widely known as the proportional odds model due
-to McCullagh (1980) and with a complementary log-log link, the model is sometimes referred
-to as the proportional hazards model for grouped survival times.
-
-In this section we introduce some notations for the cumulative model that is used through the paper and in the R code.
-We proposed a notation as consistent as possible with the main textbooks and seminal papers and with the `ordinal` R package used in the tutorial.
-We define $Y_k$ as the observed ordinal response with $1, \dots,k$ levels and $Y^\star$ is the underlying latent variable.
+In this section we introduce some notation for the CM that is used through the paper and in the R code.
+We proposed a notation as consistent as possible with the literature and the `ordinal` R package used in the tutorial.
+We define $Y_k$ as the observed ordinal response with $1, \dots, k$ levels and $Y^\star$ is the underlying latent variable.
 The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \dots, \alpha_{k - 1}$.
-Similarly to the generalized linear models framework, we define $g(x) = \eta$ as the link function that maps that transforms probabilities into the linear predictor $\eta$.
-To transform $\eta$ into probabilities we use the inverse of the link function $x = g^{-1}(\eta)$.
+Similarly to the generalized linear models framework, we define $g(x) = \eta$ as the link function that maps probabilities into the linear predictor $\eta$.
+To transform back $\eta$ into probabilities we use the inverse of the link function $x = g^{-1}(\eta)$.
 The specific link function define the type of model and require a different R function.
 For example, when assuming a Gaussian distribution we are fitting a model with a `probit` link function is the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
-When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y < k), k = 1, \dots, k - 1$.
+When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$[^1].
 Equation \@ref(eq:prob-cum-model1) depict the general cumulative model including predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
-The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to give the usual interpretation of standard regression models where an increase in $\beta$ corresponds to an increase in the probability of higher categories of $Y$ (metti il libro di agresti sui dati ordinali).
-
+The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to interpret the $\beta$ as in the standard regression model [@Agresti2010-rz] where higher $\beta$ values corresponds to increased probability of responding higher $k$ categories.
+
+[^1]: As done by @Agresti2010-rz, when referring to $P(Y \leq k)$ we are implicitly conditioning on a particular $x$ value $P(Y \leq k | x_i)$
+
+```{=tex}
 \begin{equation} 
@@
 \end{equation}
+```
 
@@
 
+```{=tex}
 \begin{equation} 
@@
 \end{equation}
-
+```
 There are always two special cases when computing the probability of a single outcome $Y$ that is when $Y = 1$ and $Y = k$.
@@
 
-The model presented in Equation \@ref(eq:prob-cum-model1) can be considered similar to a standard binomial generalized linear model.
-In this case we have $k - 1$ equations, one for each threshold $\alpha$ and we model the cumulative probability.
-The same model can be written as in Equation \@ref(eq:latent-model) assuming a continuous latent variable $Y^\star$ explained by the linear combination of predictors $\eta = \mathbf{X} \boldsymbol{\beta}$ as in a standard linear regression model.
-
+The same model can be written in the so-called latent formulation as reported in Equation \@ref(eq:latent-model).
+The model is no longer directly about the cumulative probabilities but focused on the continuous latent variable $Y^\star$ as a function of the linear predictor $\eta = \mathbf{X} \boldsymbol{\beta}$ similar to a standard linear regression.
+
+```{=tex}
 \begin{equation} 
@@
 \end{equation}
-
-The crucial part is $\epsilon_i$ that is the random part of the model coming from a certain probability distribution.
+```
+The crucial part is $\epsilon_i$ that is the random component of the model coming from a certain probability distribution.
 For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution.
@@
 
-## Alternative parametrizations
-
-Several parametrizations of the ordinal models have been proposed by different authors.
-@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian *probit* model where the latent mean and standard deviations are estimated along with the threshold.
-The interesting part is that the thresholds are transformed to be on the scale of the ordinal variable (e.g, from 1 to 4 when $k = 4$) and not in terms of $z$ scores.
-@Kurz2023-zv explain how to convert a model fitted with the `brms` package (citation) (an R package for regression modeling using *stan*) into the Kruschke [-@Kruschke2015-re] parametrization.
+The model can be also formalized in alternative ways.
+@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian version of the model with a different threshold parametrization.
 @Gelman2020-tg proposed three alternative parametrizations focusing on different definition of the thresholds.
-The current tutorial is focused on the standard parametrization of the cumulative link model as described in the equations above.
 
@@
 
-The cumulative link model implemented in Equations \@ref(eq:prob-cum-model1) and \@ref(eq:latent-model) can be considered the general formulation that requires specifying the link function $g(\cdot)$ or the latent distribution $\epsilon_i \sim D(\mu, \sigma^2)$.
+The cumulative link model implemented in Equations \@ref(eq:prob-cum-model1) and \@ref(eq:latent-model) can be considered the general formulation that requires specifying the link function $g(\cdot)$ or the errors distribution $\epsilon_i \sim D(\mu, \sigma^2)$.
 Among several available functions the *logit* and *probit* models are the most common.
@@
 On the other side, the *probit* model assume a Gaussian distribution.
-The two models provide similar results but the parameters intepretation and the way data can be simulated is different.
+
+The two models provide similar results with a different parameters interpretation.
 In the next sections we will illustrate the differences and simulation strategies.
 Figure \@ref(fig:fig-logit-vs-probit) depicts the two distributions while Table \@ref(tab:tab-model-summary) summarise the presented cumulative models with the proposed link function and the corresponding R code.
+
 In terms of parameters, both distributions can be defined with a location $\mu$ and a scale $s$ parameter.
@@
 The standard logistic distribution has $\mu = 0$ and $s^2 = 1$ thus the standard deviation simplified to $\frac{\pi}{\sqrt{3}} \approx 1.81$.
-In practical terms, the standard version of the two distributions are very similar with the logistic having more variance.
-The R implementation of these functions consist in the `q` and ``
+In practical terms, fixing $\mu$ and $s$ lead to an higher standard deviation for the logistic distribution.
 
@@
 
-```{r tab-model-summary-old}
-#| eval: false
-models_table <- data.frame(
-       stringsAsFactors = FALSE,
-                  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
-          Link.Function = c("Equation",
-                            "$\\text{logit}(p) = \\text{log}(\\frac{p}{1-p})$",
-                            "$z = \\Phi(p)$"),
-          Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
-  Inverse.Link.Function = c("Equation",
-                            "$\\frac{e^{\\text{logit}(p)}}{1 + e^{\\text{logit}(p)}}$","$\\Phi^{-1}(z)$"),
-  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
-)
-
-models_table[-1, ] |> 
-  flextable() |> 
-  set_header_labels(values = list(Model = "Model", 
-                                  Link.Function = "Equation",
-                                  Link.Function.1 = "R Code",
-                                  Inverse.Link.Function = "Equation",
-                                  Inverse.Link.Function.1 = "R Code")) |> 
-  add_header_row(values = c("", "Link Function", "Link Function", "Inverse Link Function", "Inverse Link Function")) |> 
-  merge_h(1, part = "header") |> 
-  theme_booktabs(bold_header = TRUE) |> 
-  align(part = "all", align = "center") |> 
-  set_caption(caption = filor::get_caption())
-```
-
 ```{r tab-model-summary}
-library(kableExtra)
-models_table <- data.frame(
-       stringsAsFactors = FALSE,
-                  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
-          Link.Function = c("Equation",
-                            "$\\text{logit}(p) = \\text{log}(\\frac{p}{1-p})$",
-                            "$z = \\Phi(p)$"),
-          Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
-  Inverse.Link.Function = c("Equation",
-                            "$\\frac{e^{\\text{logit}(p)}}{1 + e^{\\text{logit}(p)}}$","$\\Phi^{-1}(z)$"),
-  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
-)
-
-models_table[-1, ] |> 
+rtabs$models_table[-1, ] |> 
   kable(col.names = c("Model", "Equation", "R Code", "Equation", "R Code"), row.names = FALSE,
@@
 Despite the presence of other possibilities the `ordinal` package provide the most complete and intuitive way to implement the ordinal models.
-The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.^[For a very complete overview of the ordinal package see @Christensen2019-cz and the Package documentation https://cran.r-project.org/web/packages/ordinal/ordinal.pdf].
+The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.[^2].
+
+[^2]: For a very complete overview of the ordinal package see @Christensen2019-cz and the Package documentation <https://cran.r-project.org/web/packages/ordinal/ordinal.pdf>
 
@@
 
-```{r tab-dataset-example1}
+```{r dataset-example1}
+
 b1 <- log(3) # log odds ratio
@@
 dat <- sim_ord_latent(~x, By = b1, prob = probs, data = dat, link = "logit")
-dat |> 
-  mutate(id = 1:n(), .before = x) |> 
-  select(-ys) |> 
-  filor::trim_df() |> 
-  mtab(type = "kable", caption = filor::get_caption())
-```
-
-```{r eda-example-1}
+
 funs <- list(mean = mean, median = median, sd = sd)
-lapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
-prop.table(table(dat$y, dat$x), margin = 2)
+ss <- sapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
+ff <- t(table(dat$y, dat$x))
+
+ss <- data.frame(cbind(ss, ff))
+names(ss) <- c("mean", "median", "sd", paste0("Y", 1:4))
+
+ss$tot <- apply(ss[, 4:ncol(ss)], 1, sum)
+pp <- ss[, 4:7]/50
+ss <- round(ss, 2)
+ss$Y1 <- sprintf("**%s** (*p = %.2f*)", ss$Y1, pp$Y1)
+ss$Y2 <- sprintf("**%s** (*p = %.2f*)", ss$Y2, pp$Y2)
+ss$Y3 <- sprintf("**%s** (*p = %.2f*)", ss$Y3, pp$Y3)
+ss$Y4 <- sprintf("**%s** (*p = %.2f*)", ss$Y4, pp$Y4)
+
+ss$group <- rownames(ss)
+ss <- select(ss, group, everything(), -tot)
+ss
+```
+
+```{r tab-dataset-example1}
+#| eval: false
+
+flextable(ss) |> 
+  autofit() |> 
+  set_caption(caption = filor::get_caption()) |> 
+  align(part = "all", align = "center") |> 
+  ftExtra::colformat_md()
 ```
@@
 Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect.
-As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [for an overview about contrasts coding schemes see @Schad2020-ht]^[In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
-By default, the first level of the factor is taken as the reference level and the $j - 1$ coefficients represent the comparison between the other levels and the reference.]
+As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [for an overview about contrasts coding schemes see @Schad2020-ht][^3]
+
+[^3]: In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
+By default, the first level of the factor is taken as the reference level and the $j - 1$ coefficients represent the comparison between the other levels and the reference.
 
@@
 
-aggiungere cosa significano i parametri nella regressione
-
+To understand the logit model we need to introduce odds and odds ratio.
 The odds of a probability $p$ is defined as $\frac{p}{1 - p}$ thus the probability of success divided by the probability of failure.
@@
 $k = 4$ outcomes we have $k - 1$ models determined by the cumulative probability in terms of $P(Y \leq 1), \dots, P(Y \leq k - 1)$.
+We can manually calculate the odds ratio starting from the previous dataset.
+
+```{r}
+#| echo: true
+#| collapse: true
+
+# function to calculate odds
+odds <- function(p) p / (1 - p)
+
+# probability of Y = 1 for the group a
+dat$yn <- as.integer(dat$y) # better as number here
+(p1a <- mean(dat$yn[dat$x == "a"] == 1))
+
+# thus P(Y > 1) = 1 - p1a
+odds(p1a)
+
+# odds ratio Y = 1 a vs b
+(p1b <- mean(dat$yn[dat$x == "b"] == 1))
+odds(p1a) / odds(p1b)
+```
+
+For example the probability of responding $Y = 1$ in the group "a" is `r p1a` corresponding to an odds of `r odds(p1a)`.
+When calculating the odds ratio comparing "a" vs "b" for $Y = 1$ we obtain that the group "a" has `r odds(p1a) / odds(p1b)` times the odds of responding $Y = 1$ compared to the group "b".
+
+For an ordinal variable we have $k$ levels and we can calculate $k - 1$ cumulative probabilities (excluding the last associated with $p = 1$).
+Thus we have $k - 1$ odds ratio or what is called a cumulative odds ratio.
+Basically we repeat the previous steps but on the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$.
+The three odds ratios are very similar correspond to the model estimate $e^\beta_1$, in R `exp(fit$beta)` `r exp(fit$beta)`.
+In fact, the logit model essentially estimate the cumulative odds ratio.
+
+```{r}
+#| echo: true
+#| collapse: true
+# the dummy_ord() is a custom function creating k - 1 dummy variable for the cumulative probabilities
+
+# data.frame with the group (x) and the k - 1 dummy variables
+cum_p <- cbind(x = dat$x, dummy_ord(dat$y))
+
+# calculating the cumulative probability for k - 1 variables. 
+# Taking the average of a series of 0-1 is the same as computing the proportion of 1s.
+(group_a <- apply(cum_p[cum_p$x == "a", -1], 2, mean))
+(group_b <- apply(cum_p[cum_p$x == "b", -1], 2, mean))
+
+# calculating k - 1 odds ratios on the cumulative probabilities as a/b
+odds(group_a) / odds(group_b)
+```
 
@@
 
-Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the  proportional odds assumption.
-There are more advanced versions of the model relaxing this assumption completely (*non proportional odds* models) and partially (*partial proportional odds* models) (rivedi se la nomenclatura va bene).
+From the previous example, we noticed that the $k - 1$ odds ratios are very similar (the similarity increase if $n$ increase).
+This is not a coincidence but an implicit (until now) assumption made by the CM called *proportional odds* (POA).
+Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the (POA).
+There are more advanced versions of the model relaxing this assumption completely (*non proportional odds*) and partially (*partial proportional odds*).
 The proportional odds assumption is a crucial point when interpreting model parameters.
 Can be formalized as in Equation ().
+
 Basically if we use the *logit* link function $g()$ the $\beta$s are interpreted as standard odds ratio in the logistic regression.
-Given that we have $k > 2$ alternatives there is the need of $k - 1$ equations (as reported in Equation ()).
-The porportional odds is assuming that the regression coefficients $\beta$ are independent from the thresholds $\alpha$ thus the effect is the same regardless of the $Y$ level.
+Given that we have $k > 2$ alternatives there is the need of $k - 1$ equations.
+The POA is formalized in Equation \@ref(eq:prop-odds).
+Basically the cumulative log odds ratio comparing $P(Y_k|x_0)$ with $P(Y_k|x_1)$ is the same regardless the specific $k$ or threshold $\alpha$.
+
+The proportional odds is assuming that the regression coefficients $\beta$ are independent from the thresholds $\alpha$ thus the effect is the same regardless of the $Y$ level.
 Thus $\beta_1 = \beta_2 \dots = \beta_{k - 1}$.
-The Figure () depicts the proportional odds assumption for the $k - 1$ logistic curves both for probabilities and linear predictors $\eta$.
-
-```{r fig-prop-odds, fig.cap=filor::get_caption()}
-rfigs$fig_prop_odds
-```
-
-The proportional odds assumption is convenient because regardless the number of $k$, the $\beta_j$ ($j$ being the number of regression coefficients) effect is assumed to be the same.
-The model is more parsimonious compared to estimating $k - 1$ coefficients for each $\beta_j$ as in the multinomial regression or the non-proportional version of the ordinal regression.
-Clearly this can be a strong assumption in some conditions.
-There are several methods to test the proportional odds assumption [see @Liu2023-bp for an overview].
-@Tutz2020-xq suggested that using alternative as the location-shift or location-scale models, the assumption can be made (improving model parsimony) but still having a more flexible modeling framework.
-In the next section, we provide more details about the location-scale model in terms of parameters interpretation and data simulation.
-
-To make a practical example, we can use again the the dataset presented in Table. We can check the proportional odds by extrating fitted probabilities and manually computing the odds ratios.
+The Figure (\@ref(fig:fig-prop-odds)) depicts the proportional odds assumption for the $k - 1$ logistic curves both for probabilities and linear predictors $\eta$.
 
 <!-- - https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione -->
-
-Equation \@ref(eq:prop-odds) formalize the proportional odds assumption.
-Basically the cumulative log odds ratio comparing $P(Y_k|x_0)$ with $P(Y_k|x_1)$ is the same regardless the specific $k$ or threshold $\alpha$.
 
@@
 
-This can be shown by fitting the model with `clm()` that by default assume the proportional odds and using the predicted probabilities to compute the odds ratios.
+```{r fig-prop-odds, fig.cap=filor::get_caption()}
+rfigs$fig_prop_odds
+```
+
+The proportional odds assumption is convenient because regardless of $k$, the $\beta_j$ ($j$ being the number of regression coefficients) effect is assumed to be the same.
+The model is more parsimonious compared to estimating $k - 1$ coefficients for each $\beta_j$ as in the multinomial regression or the non-proportional model.
+
+Given the relatively strong assumption, there are several methods for testing if data are supporting the POA [see @Liu2023-bp for an overview].
+@Tutz2020-xq suggested a trade-off between assuming/relaxing the POA in terms of implementing *location-shift* or *location-scale* models.
+Basically these methods should guarantee more flexibility in modelling the observed probabilities reducing the number of parameters.
+However, these methods and models are outside the scope of the tutorial.
+In the next section, we provide more details only about the *location-scale* model in terms of parameters interpretation and data simulation.
+
+This can be shown by fitting the model with `clm()` that by default assume the POA and using the predicted probabilities to compute the odds ratios.
 In the following code we computed the odds ratio comparing $x_a$ and $x_b$ when $k \leq 1$ and $k \leq 2$.
-The log odds ratio is the same.
 
@@
 #| echo: true
+#| collapse: true
 
@@
 
-odds <- function(p) p / (1 - p) # function to compute odds
-
-# group a
-a_or1vs2345 <- odds(pr[1, 1]) # 1 vs 2 3 4 5
-a_or12vs345 <- odds(sum(pr[1, 1:2])) # 1 vs 2 3 4 5
-
-# group b
-b_or1vs2345 <- odds(pr[2, 1]) # 1 vs 2 3 4 5
-b_or12vs345 <- odds(sum(pr[2, 1:2])) # 1 vs 2 3 4 5
-
-c(xa = log(a_or1vs2345 / a_or12vs345), xb = log(b_or1vs2345 / b_or12vs345))
-```
-
-Clearly, the odds ratio here is exactly the same because the model assume the proportional odds assumption putting this constraint on the predicted probabilities.
-The estimated $\beta$ for each $k - 1$ equation could differ and the amount of heterogeneity is actually a way to test the proportional odds with real data.
-In the supplementary materials we implemented an intuitive plot proposed by @Harrell-2015-no to visually assess the proportional odds assumption.
+# y <= 1
+y1a <- pr[1, 1]
+y1b <- pr[2, 1]
+
+# y <= 2
+y12a <- sum(pr[1, 1:2])
+y12b <- sum(pr[2, 1:2])
+
+# odds ratio y <= 1 (a vs b) VS y <= 2 (a vs b)
+odds(y1a) / odds(y1b)
+odds(y12a) / odds(y12b)
+```
+
+In this case the odds ratio is exactly the same because the model (thus the predicted probabilities) assume the POA.
+When estimating the $k - 1$ odds ratio from the data results could be different.
+In fact, the discrepancy between the odds ratio is an index of POA violation.
+metti qualche riferimento, magari a supplementary.
 
@@
 
-# group a
-a_1vs2345 <- qnorm(pr[1, 1]) # 1 vs 2 3 4 5
-a_12vs345 <- qnorm(sum(pr[1, 1:2])) # 1 2 vs 3 4 5
-
-# group b
-b_1vs2345 <- qnorm(pr[2, 1]) # 1 vs 2 3 4 5
-b_12vs345 <- qnorm(sum(pr[2, 1:2])) # 1 2 vs 3 4 5
-
-c(xa = a_1vs2345 - a_12vs345, xb = b_1vs2345 - b_12vs345)
-```
-
-### Beyond the proportional odds
-
-qualcosa che spieghi come andare oltre e qualche proposta per testare l'assunzione.
+# y <= 1
+y1a <- pr[1, 1]
+y1b <- pr[2, 1]
+
+# y <= 2
+y12a <- sum(pr[1, 1:2])
+y12b <- sum(pr[2, 1:2])
+
+# z score difference y <= 1 (a vs b) VS y <= 2 (a vs b)
+qnorm(y1a) - qnorm(y1b)
+qnorm(y12a) - qnorm(y12b)
+```
 
@@
 The alternative to fitting a cumulative logit model is using a standard normal distribution thus fitting a *probit* model.
-The main difference is that the latent distribution is now a Gaussian distribution with $\mu = 0$ and $\sigma^2 = 1$ and the link function is the cumulative gaussian distribution function $\Phi$ implemented in R with the `pnorm()` function.
+The main difference is that the latent distribution is now a Gaussian distribution with $\mu = 0$ and $\sigma^2 = 1$ and the link function is the cumulative Gaussian distribution function $\Phi$ implemented in R with the `pnorm()` function.
 The inverse of the Gaussian cumulative distribution function is $\Phi^{-1}$ implemented in R with the `qnorm()` function.
@@
 The intepretation in terms of shifting the latent mean holds also for the logistic model.
-However, the standard deviation of the standard logistic regression is $\frac{\pi}{\sqrt{3}} \approx 1.81$^[Actually the variance of the logistic distribution is $\frac{s^2\pi^2}{3}$ and the standard deviation $\frac{s\pi}{\sqrt{3}}$ where $s$ is the scale of the distribution.
+However, the standard deviation of the standard logistic regression is $\frac{\pi}{\sqrt{3}} \approx 1.81$[^4].
+The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\frac{\pi}{\sqrt{3}}$ standard deviations.
+
+[^4]: Actually the variance of the logistic distribution is $\frac{s^2\pi^2}{3}$ and the standard deviation $\frac{s\pi}{\sqrt{3}}$ where $s$ is the scale of the distribution.
 For the standard logistic distribution $s = 1$ (as for the standard normal distribution).
-For the standard normal distribution $s^2 = \sigma^2 = 1$].
-The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\frac{\pi}{\sqrt{3}}$ standard deviations.
+For the standard normal distribution $s^2 = \sigma^2 = 1$
 
@@
 
-The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution^[In R we are using the `sample()` function that simulate values from a *categorical* distribution.
-The *categorical* distribution is considered a special case of the multinomial distribution].
+The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution[^5].
 This is similar to the general method to simulate data for a generalized linear model where the linear predictors $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.
 The following code box simulate a $n$ binary trials with a continuous predictor $x$.
+
+[^5]: In R we are using the `sample()` function that simulate values from a *categorical* distribution. The *categorical* distribution is considered a special case of the multinomial distribution
 
@@
 1.  Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
-2. Define the regression coefficients $\beta$
-<!-- controlla che le soglie non fanno riferimento a quando è 0 ma concettualmente si. essendo che non cambiano non è rilevante a quale x si faccia riferimento -->
+2.
+Define the regression coefficients $\beta$ <!-- controlla che le soglie non fanno riferimento a quando è 0 ma concettualmente si.
+essendo che non cambiano non è rilevante a quale x si faccia riferimento -->
 3.  Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
@@
 
+```{=tex}
 \begin{equation}
@@
 \end{equation}
-
+```
+```{=tex}
 \begin{equation} 
@@
 \end{equation}
-
+```
 Where $X\zeta$ is the linear predictor $\eta$ for the scale of the distribution.
@@
 The current paper proposed a simplified way with some functions to generate ordinal data.
-For more complex simulations such as simulating correlated ordinal data the `simstudy` package [https://kgoldfeld.github.io/simstudy/articles/ordinal.html](https://kgoldfeld.github.io/simstudy/articles/ordinal.html) proposed a very comprehensive set of data simulation function also for ordinal data.
+For more complex simulations such as simulating correlated ordinal data the `simstudy` package <https://kgoldfeld.github.io/simstudy/articles/ordinal.html> proposed a very comprehensive set of data simulation function also for ordinal data.
 
@@
 
-(ref:fig-ordinal-models) Theoretical figure about ordinal models. Voglio fare una figura come questa in @Burkner2019-aw ma con qualche modifica per fare vedere i tre modelli.
+(ref:fig-ordinal-models) Cumulative model. Adapted from @Burkner2019-aw.
 

### 2024-01-03 paper/paper-new.tex
update

@@
 Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
-(rivedi questo).
 An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
 Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer.
-Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation).
+Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools (Agresti, 2010; e.g., N.
+Cliff, 1996).
 This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data.
 Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018).
+More generally, Norman Cliff (2016) suggest that most of the research questions in behavioral sciences can be considered as ordinal (\emph{is the score} \(x\) higher than the score \(y\)?) concerning variables where the most appropriate measurement scale is probably ordinal.
 
@@
 In the same work, Liddell and Kruschke (2018) showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric (but see Robitzsch, 2020 for an alternative perspective).
-They reported problems in terms of lack of power, inversion of the effects and distortion in estimating the effect size (rivedi meglio qui).
-(vedi se c'è qualche altro lavoro che fa vedere questa cosa).
-Some authors suggested that despite individual items being ordinal, averaging multiple items could solve the problem of applying metric models (e.g., Carifio \& Perla, 2007, 2008 metti queste citazioni).
-Liddell and Kruschke (2018) suggested that in some conditions even averaging multiple items and applying metric models could be problematic.
-Cliff (2016) suggest that most of the research questions in behavioral sciences can be considered as ordinal (\emph{is the score \(x\) higher than the score \(y\)}?) concerning variables where the most appropriate measurement scale is probably ordinal.
-
-qualcosa ancora qui
+They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and distorted effect size estimates.
+Some authors suggested that individual ordinal items and a collection of ordinal (averaged or summed) items can safely be considered and analyzed as metric (Carifio \& Perla, 2008; Carifio \& Perla, 2007; but see Jamieson, 2004).
+Despite Liddell and Kruschke (2018) provide some example that averaging ordinal items and applying metric models is not appropriate, in the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.).
 
@@
 
-Despite the actual modeling proposal by Liddell and Kruschke (2018), there is a class of regression models taking into account the ordinal nature of the response variable without making metric assumptions.
-We can class this general class of models as \emph{ordinal regression}.
-The nomenclature of these models can be confusing mainly because there are several types of models with different assumptions and structures (Gerhard Tutz, 2022).
+Despite the actual modeling proposal by Liddell and Kruschke (2018), there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions.
+We can name this general class of models as \emph{ordinal regression}.
+The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures (Gerhard Tutz, 2022).
 Gerhard Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
-We can identify three main classes: \emph{cumulative models} (McCullagh, 1980), \emph{sequential models} (Gerhard Tutz, 1990) and \emph{adjacent category models}.
-The cumulative is the mostly used model assuming the existence of a latent variable that when segmented using thresholds produces the observed ordinal variable.
-The psychological process underlying the response is clearly formalized in the signal detection theory framework where the respondent (vedi come sistemare).
-The sequential model as suggested by the name is appropriate when modelling sequential processes.
-Assuming to have 5 response options, the sequential model assume that responding 3 means already reaching the states 1 and 2.
+We can identify three main classes: \emph{cumulative models} {[}CM; McCullagh (1980); Agresti (2010){]}, \emph{sequential models} (Gerhard Tutz, 1990) and \emph{adjacent category models}.
+The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.
+
+The \emph{sequential model} as suggested by the name is appropriate when modelling sequential processes.
+Assuming to have five response options, the sequential model assume that responding ``3'' assume a sequential process where steps ``1'' and ``2'' are already reached.
 A clear example is proposed by Bürkner and Vuorre (2019) where the marriage duration in years is predicted as a function of some explanatory variables.
-For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential regression.
+For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential model.
 When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Gerhard Tutz \& Berger, 2020).
@@
 
-In the current paper we put the focus on the cumulative model for several reasons.
+In the current paper we put the focus on the CM for several reasons.
 The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation.
-The second reason is that several psychological variables can be formalized as a latent continuous construct that when measured is collected using ordinal items.
-Furthermore, also the signal detection theory framework models the decision process assuming a latent distribution for the signal and noise where thresholds are the response criteria (e.g., DeCarlo, 2010).
+The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item.
+Furthermore, CM are also used to model data under a signal detection theory framework (e.g., DeCarlo, 2010).
+The Figure \ref{fig:fig-ordinal-models} depict the overall structure of the \emph{cumulative} model.
 
@@
 
-\caption{Theoretical figure about ordinal models. Voglio fare una figura come questa in Bürkner and Vuorre (2019) ma con qualche modifica per fare vedere i tre modelli.}\label{fig:fig-ordinal-models}
+\caption{Cumulative model. Adapted from Bürkner and Vuorre (2019).}\label{fig:fig-ordinal-models}
 \end{figure}
@@
 
-metti qualcosa sulla nomenclatura
-
-\begin{quote}
-The name, cumulative link models is adopted from Agresti (2002), but the model class has
-been referred to by several other names in the literature, such as ordered logit models and
-ordered probit models (Greene and Hensher 2010) for the logit and probit link functions. The
-cumulative link model with a logit link is widely known as the proportional odds model due
-to McCullagh (1980) and with a complementary log-log link, the model is sometimes referred
-to as the proportional hazards model for grouped survival times.
-\end{quote}
-
-In this section we introduce some notations for the cumulative model that is used through the paper and in the R code.
-We proposed a notation as consistent as possible with the main textbooks and seminal papers and with the \texttt{ordinal} R package used in the tutorial.
-We define \(Y_k\) as the observed ordinal response with \(1, \dots,k\) levels and \(Y^\star\) is the underlying latent variable.
+In this section we introduce some notation for the CM that is used through the paper and in the R code.
+We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} R package used in the tutorial.
+We define \(Y_k\) as the observed ordinal response with \(1, \dots, k\) levels and \(Y^\star\) is the underlying latent variable.
 The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\).
-Similarly to the generalized linear models framework, we define \(g(x) = \eta\) as the link function that maps that transforms probabilities into the linear predictor \(\eta\).
-To transform \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\).
+Similarly to the generalized linear models framework, we define \(g(x) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\).
+To transform back \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\).
 The specific link function define the type of model and require a different R function.
 For example, when assuming a Gaussian distribution we are fitting a model with a \texttt{probit} link function is the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
-When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y < k), k = 1, \dots, k - 1\).
+When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(x\) value \(P(Y \leq k | x_i)\)}.
 Equation \eqref{eq:prob-cum-model1} depict the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
-The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to give the usual interpretation of standard regression models where an increase in \(\beta\) corresponds to an increase in the probability of higher categories of \(Y\) (metti il libro di agresti sui dati ordinali).
+The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta\) as in the standard regression model (Agresti, 2010) where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories.
 
@@
 
-The model presented in Equation \eqref{eq:prob-cum-model1} can be considered similar to a standard binomial generalized linear model.
-In this case we have \(k - 1\) equations, one for each threshold \(\alpha\) and we model the cumulative probability.
-The same model can be written as in Equation \eqref{eq:latent-model} assuming a continuous latent variable \(Y^\star\) explained by the linear combination of predictors \(\eta = \mathbf{X} \boldsymbol{\beta}\) as in a standard linear regression model.
+The same model can be written in the so-called latent formulation as reported in Equation \eqref{eq:latent-model}.
+The model is no longer directly about the cumulative probabilities but focused on the continuous latent variable \(Y^\star\) as a function of the linear predictor \(\eta = \mathbf{X} \boldsymbol{\beta}\) similar to a standard linear regression.
 
@@
 
-The crucial part is \(\epsilon_i\) that is the random part of the model coming from a certain probability distribution.
+The crucial part is \(\epsilon_i\) that is the random component of the model coming from a certain probability distribution.
 For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution.
@@
 
-\subsection{Alternative parametrizations}\label{alternative-parametrizations}
-
-Several parametrizations of the ordinal models have been proposed by different authors.
-Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian \emph{probit} model where the latent mean and standard deviations are estimated along with the threshold.
-The interesting part is that the thresholds are transformed to be on the scale of the ordinal variable (e.g, from 1 to 4 when \(k = 4\)) and not in terms of \(z\) scores.
-Kurz (2023) explain how to convert a model fitted with the \texttt{brms} package (citation) (an R package for regression modeling using \emph{stan}) into the Kruschke (2015) parametrization.
+The model can be also formalized in alternative ways.
+Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model with a different threshold parametrization.
 Gelman, Hill, and Vehtari (2020) proposed three alternative parametrizations focusing on different definition of the thresholds.
-The current tutorial is focused on the standard parametrization of the cumulative link model as described in the equations above.
 
@@
 
-The cumulative link model implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} can be considered the general formulation that requires specifying the link function \(g(\cdot)\) or the latent distribution \(\epsilon_i \sim D(\mu, \sigma^2)\).
+The cumulative link model implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} can be considered the general formulation that requires specifying the link function \(g(\cdot)\) or the errors distribution \(\epsilon_i \sim D(\mu, \sigma^2)\).
 Among several available functions the \emph{logit} and \emph{probit} models are the most common.
@@
 On the other side, the \emph{probit} model assume a Gaussian distribution.
-The two models provide similar results but the parameters intepretation and the way data can be simulated is different.
+
+The two models provide similar results with a different parameters interpretation.
 In the next sections we will illustrate the differences and simulation strategies.
 Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table \ref{tab:tab-model-summary} summarise the presented cumulative models with the proposed link function and the corresponding R code.
+
 In terms of parameters, both distributions can be defined with a location \(\mu\) and a scale \(s\) parameter.
@@
 The standard logistic distribution has \(\mu = 0\) and \(s^2 = 1\) thus the standard deviation simplified to \(\frac{\pi}{\sqrt{3}} \approx 1.81\).
-In practical terms, the standard version of the two distributions are very similar with the logistic having more variance.
+In practical terms, fixing \(\mu\) and \(s\) lead to an higher standard deviation for the logistic distribution.
 
@@
 \end{figure}
-
-\normalsize
-
-\scriptsize
 
@@
 \midrule
-Cumulative Logit & $\text{logit}(p) = \text{log}(\frac{p}{1-p})$ & `qlogis()` & $\frac{e^{\text{logit}(p)}}{1 + e^{\text{logit}(p)}}$ & `plogis()`\\
+Cumulative Logit & $\text{logit}(p) = \text{log}(p / (1-p))$ & `qlogis()` & $e^{\text{logit}(p)} / (1 + e^{\text{logit}(p)})$ & `plogis()`\\
 Cumulative Probit & $z = \Phi(p)$ & `qnorm()` & $\Phi^{-1}(z)$ & `pnorm()`\\
@@
 
-\begin{table}
-
-\caption{\label{tab:tab-dataset-example1}(ref:tab-dataset-example1)}
-\centering
-\begin{tabular}[t]{ccc}
-\toprule
-id & x & y\\
-\midrule
-1 & a & 4\\
-2 & a & 3\\
-3 & a & 2\\
-4 & a & 2\\
-... & ... & ...\\
-\addlinespace
-97 & b & 3\\
-98 & b & 3\\
-99 & b & 2\\
-100 & b & 4\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\normalsize
-
-\scriptsize
-
 \begin{verbatim}
-## $mean
-##    a    b 
-## 2.52 3.04 
-## 
-## $median
-##   a   b 
-## 2.0 3.5 
-## 
-## $sd
-##        a        b 
-## 1.092460 1.124132
+##   group mean median   sd                  Y1                  Y2
+## a     a 2.46      2 1.20 **15** (*p = 0.30*) **11** (*p = 0.22*)
+## b     b 3.26      4 0.94  **2** (*p = 0.04*) **11** (*p = 0.22*)
+##                    Y3                  Y4
+## a **10** (*p = 0.20*) **14** (*p = 0.28*)
+## b  **9** (*p = 0.18*) **28** (*p = 0.56*)
 \end{verbatim}
 
-\begin{verbatim}
-##    
-##        a    b
-##   1 0.20 0.14
-##   2 0.34 0.18
-##   3 0.20 0.18
-##   4 0.26 0.50
-\end{verbatim}
+\normalsize
+
+\scriptsize
 
@@
 ##  link  threshold nobs logLik  AIC    niter max.grad cond.H 
-##  logit flexible  100  -130.61 269.23 5(0)  2.61e-13 2.0e+01
+##  logit flexible  100  -125.31 258.63 5(0)  3.87e-12 2.1e+01
 ## 
@@
 ##    Estimate Std. Error z value Pr(>|z|)    
-## xb   0.8795     0.3719   2.365    0.018 *
+## xb   1.3088     0.3828   3.419 0.000629 ***
 ## ---
@@
 ##     Estimate Std. Error z value
-## 1|2  -1.2246     0.3040  -4.029
-## 2|3   0.1231     0.2653   0.464
-## 3|4   0.9396     0.2841   3.308
+## 1|2  -1.0373     0.3074  -3.374
+## 2|3   0.1947     0.2791   0.698
+## 3|4   1.0282     0.2947   3.489
 \end{verbatim}
@@
 Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect.
-As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding (for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, \& Kliegl, 2020)\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
+As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding {[}for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, and Kliegl (2020){]}\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
 By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}
@@
 
-aggiungere cosa significano i parametri nella regressione
-
+To understand the logit model we need to introduce odds and odds ratio.
 The odds of a probability \(p\) is defined as \(\frac{p}{1 - p}\) thus the probability of success divided by the probability of failure.
@@
 With e.g.~\(k = 4\) outcomes we have \(k - 1\) models determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\).
+We can manually calculate the odds ratio starting from the previous dataset.
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# function to calculate odds}
+\NormalTok{odds }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(p) p }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ p)}
+
+\CommentTok{\# probability of Y = 1 for the group a}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{yn }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# better as number here}
+\NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
+\DocumentationTok{\#\# [1] 0.3}
+
+\CommentTok{\# thus P(Y \textgreater{} 1) = 1 {-} p1a}
+\FunctionTok{odds}\NormalTok{(p1a)}
+\DocumentationTok{\#\# [1] 0.4285714}
+
+\CommentTok{\# odds ratio Y = 1 a vs b}
+\NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
+\DocumentationTok{\#\# [1] 0.04}
+\FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
+\DocumentationTok{\#\# [1] 10.28571}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+For example the probability of responding \(Y = 1\) in the group ``a'' is 0.30 corresponding to an odds of 0.43.
+When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 10.29 times the odds of responding \(Y = 1\) compared to the group ``b''.
+
+For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(p = 1\)).
+Thus we have \(k - 1\) odds ratio or what is called a cumulative odds ratio.
+Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\).
+The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 3.70.
+In fact, the logit model essentially estimate the cumulative odds ratio.
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# the dummy\_ord() is a custom function creating k {-} 1 dummy variable for the cumulative probabilities}
+
+\CommentTok{\# data.frame with the group (x) and the k {-} 1 dummy variables}
+\NormalTok{cum\_p }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(}\AttributeTok{x =}\NormalTok{ dat}\SpecialCharTok{$}\NormalTok{x, }\FunctionTok{dummy\_ord}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y))}
+
+\CommentTok{\# calculating the cumulative probability for k {-} 1 variables. }
+\CommentTok{\# Taking the average of a series of 0{-}1 is the same as computing the proportion of 1s.}
+\NormalTok{(group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
+\DocumentationTok{\#\# y1vs234 y12vs34 y123vs4 }
+\DocumentationTok{\#\#    0.30    0.52    0.72}
+\NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
+\DocumentationTok{\#\# y1vs234 y12vs34 y123vs4 }
+\DocumentationTok{\#\#    0.04    0.26    0.44}
+
+\CommentTok{\# calculating k {-} 1 odds ratios on the cumulative probabilities as a/b}
+\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b)}
+\DocumentationTok{\#\#   y1vs234   y12vs34   y123vs4 }
+\DocumentationTok{\#\# 10.285714  3.083333  3.272727}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
 
@@
 
-Following again the taxonomy by Gerhard Tutz (2022), each of the presented ordinal regression model has a basic version making the proportional odds assumption.
-There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds} models) and partially (\emph{partial proportional odds} models) (rivedi se la nomenclatura va bene).
+From the previous example, we noticed that the \(k - 1\) odds ratios are very similar (the similarity increase if \(n\) increase).
+This is not a coincidence but an implicit (until now) assumption made by the CM called \emph{proportional odds} (POA).
+Following again the taxonomy by Gerhard Tutz (2022), each of the presented ordinal regression model has a basic version making the (POA).
+There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}).
 The proportional odds assumption is a crucial point when interpreting model parameters.
 Can be formalized as in Equation ().
+
 Basically if we use the \emph{logit} link function \(g()\) the \(\beta\)s are interpreted as standard odds ratio in the logistic regression.
-Given that we have \(k > 2\) alternatives there is the need of \(k - 1\) equations (as reported in Equation ()).
-The porportional odds is assuming that the regression coefficients \(\beta\) are independent from the thresholds \(\alpha\) thus the effect is the same regardless of the \(Y\) level.
+Given that we have \(k > 2\) alternatives there is the need of \(k - 1\) equations.
+The POA is formalized in Equation \eqref{eq:prop-odds}.
+Basically the cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) is the same regardless the specific \(k\) or threshold \(\alpha\).
+
+The proportional odds is assuming that the regression coefficients \(\beta\) are independent from the thresholds \(\alpha\) thus the effect is the same regardless of the \(Y\) level.
 Thus \(\beta_1 = \beta_2 \dots = \beta_{k - 1}\).
-The Figure () depicts the proportional odds assumption for the \(k - 1\) logistic curves both for probabilities and linear predictors \(\eta\).
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/fig-prop-odds-1}
-
-}
-
-\caption{caption here}\label{fig:fig-prop-odds}
-\end{figure}
-
-\normalsize
-
-The proportional odds assumption is convenient because regardless the number of \(k\), the \(\beta_j\) (\(j\) being the number of regression coefficients) effect is assumed to be the same.
-The model is more parsimonious compared to estimating \(k - 1\) coefficients for each \(\beta_j\) as in the multinomial regression or the non-proportional version of the ordinal regression.
-Clearly this can be a strong assumption in some conditions.
-There are several methods to test the proportional odds assumption (see Liu, He, Tu, \& Tang, 2023 for an overview).
-Gerhard Tutz and Berger (2020) suggested that using alternative as the location-shift or location-scale models, the assumption can be made (improving model parsimony) but still having a more flexible modeling framework.
-In the next section, we provide more details about the location-scale model in terms of parameters interpretation and data simulation.
-
-To make a practical example, we can use again the the dataset presented in Table. We can check the proportional odds by extrating fitted probabilities and manually computing the odds ratios.
-
-Equation \eqref{eq:prop-odds} formalize the proportional odds assumption.
-Basically the cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) is the same regardless the specific \(k\) or threshold \(\alpha\).
+The Figure (\ref{fig:fig-prop-odds}) depicts the proportional odds assumption for the \(k - 1\) logistic curves both for probabilities and linear predictors \(\eta\).
 
@@
 
-This can be shown by fitting the model with \texttt{clm()} that by default assume the proportional odds and using the predicted probabilities to compute the odds ratios.
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-prop-odds-1}
+
+}
+
+\caption{caption here}\label{fig:fig-prop-odds}
+\end{figure}
+
+\normalsize
+
+The proportional odds assumption is convenient because regardless of \(k\), the \(\beta_j\) (\(j\) being the number of regression coefficients) effect is assumed to be the same.
+The model is more parsimonious compared to estimating \(k - 1\) coefficients for each \(\beta_j\) as in the multinomial regression or the non-proportional model.
+
+Given the relatively strong assumption, there are several methods for testing if data are supporting the POA (see Liu, He, Tu, \& Tang, 2023 for an overview).
+Gerhard Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the POA in terms of implementing \emph{location-shift} or \emph{location-scale} models.
+Basically these methods should guarantee more flexibility in modelling the observed probabilities reducing the number of parameters.
+However, these methods and models are outside the scope of the tutorial.
+In the next section, we provide more details only about the \emph{location-scale} model in terms of parameters interpretation and data simulation.
+
+This can be shown by fitting the model with \texttt{clm()} that by default assume the POA and using the predicted probabilities to compute the odds ratios.
 In the following code we computed the odds ratio comparing \(x_a\) and \(x_b\) when \(k \leq 1\) and \(k \leq 2\).
-The log odds ratio is the same.
 
@@
 
-\NormalTok{odds }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(p) p }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ p) }\CommentTok{\# function to compute odds}
-
-\CommentTok{\# group a}
-\NormalTok{a\_or1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
-\NormalTok{a\_or12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}
-
-\CommentTok{\# group b}
-\NormalTok{b\_or1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
-\NormalTok{b\_or12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 vs 2 3 4 5}
-
-\FunctionTok{c}\NormalTok{(}\AttributeTok{xa =} \FunctionTok{log}\NormalTok{(a\_or1vs2345 }\SpecialCharTok{/}\NormalTok{ a\_or12vs345), }\AttributeTok{xb =} \FunctionTok{log}\NormalTok{(b\_or1vs2345 }\SpecialCharTok{/}\NormalTok{ b\_or12vs345))}
+\CommentTok{\# y \textless{}= 1}
+\NormalTok{y1a }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]}
+\NormalTok{y1b }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]}
+
+\CommentTok{\# y \textless{}= 2}
+\NormalTok{y12a }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
+\NormalTok{y12b }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
+
+\CommentTok{\# odds ratio y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
+\FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
+\DocumentationTok{\#\# [1] 3.701591}
+\FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
+\DocumentationTok{\#\# [1] 3.701591}
 \end{Highlighting}
@@
 
-\begin{verbatim}
-##        xa        xb 
-## -1.347649 -1.347649
-\end{verbatim}
-
-\normalsize
-
-Clearly, the odds ratio here is exactly the same because the model assume the proportional odds assumption putting this constraint on the predicted probabilities.
-The estimated \(\beta\) for each \(k - 1\) equation could differ and the amount of heterogeneity is actually a way to test the proportional odds with real data.
-In the supplementary materials we implemented an intuitive plot proposed by Harrell (2015) to visually assess the proportional odds assumption.
+\normalsize
+
+In this case the odds ratio is exactly the same because the model (thus the predicted probabilities) assume the POA.
+When estimating the \(k - 1\) odds ratio from the data results could be different.
+In fact, the discrepancy between the odds ratio is an index of POA violation.
+metti qualche riferimento, magari a supplementary.
 
@@
 
-\CommentTok{\# group a}
-\NormalTok{a\_1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
-\NormalTok{a\_12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 2 vs 3 4 5}
-
-\CommentTok{\# group b}
-\NormalTok{b\_1vs2345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]) }\CommentTok{\# 1 vs 2 3 4 5}
-\NormalTok{b\_12vs345 }\OtherTok{\textless{}{-}} \FunctionTok{qnorm}\NormalTok{(}\FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])) }\CommentTok{\# 1 2 vs 3 4 5}
-
-\FunctionTok{c}\NormalTok{(}\AttributeTok{xa =}\NormalTok{ a\_1vs2345 }\SpecialCharTok{{-}}\NormalTok{ a\_12vs345, }\AttributeTok{xb =}\NormalTok{ b\_1vs2345 }\SpecialCharTok{{-}}\NormalTok{ b\_12vs345)}
+\CommentTok{\# y \textless{}= 1}
+\NormalTok{y1a }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]}
+\NormalTok{y1b }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]}
+
+\CommentTok{\# y \textless{}= 2}
+\NormalTok{y12a }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
+\NormalTok{y12b }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
+
+\CommentTok{\# z score difference y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
+\FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
+\DocumentationTok{\#\# [1] 0.8261077}
+\FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
+\DocumentationTok{\#\# [1] 0.8261077}
 \end{Highlighting}
@@
 
-\begin{verbatim}
-##         xa         xb 
-## -0.7963604 -0.7963604
-\end{verbatim}
-
-\normalsize
-
-\subsubsection{Beyond the proportional odds}\label{beyond-the-proportional-odds}
-
-qualcosa che spieghi come andare oltre e qualche proposta per testare l'assunzione.
+\normalsize
 
@@
 The alternative to fitting a cumulative logit model is using a standard normal distribution thus fitting a \emph{probit} model.
-The main difference is that the latent distribution is now a Gaussian distribution with \(\mu = 0\) and \(\sigma^2 = 1\) and the link function is the cumulative gaussian distribution function \(\Phi\) implemented in R with the \texttt{pnorm()} function.
+The main difference is that the latent distribution is now a Gaussian distribution with \(\mu = 0\) and \(\sigma^2 = 1\) and the link function is the cumulative Gaussian distribution function \(\Phi\) implemented in R with the \texttt{pnorm()} function.
 The inverse of the Gaussian cumulative distribution function is \(\Phi^{-1}\) implemented in R with the \texttt{qnorm()} function.
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-4-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-4}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-6-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-6}
 \end{figure}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-6-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-6}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-8-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-8}
 \end{figure}
@@
 ##    p_y1  p_y2  p_y3  p_y4
-## 1 0.088 0.161  0.25 0.502
-## 2 0.296 0.259 0.237 0.208
-## 3 0.265 0.254 0.246 0.235
-## 4 0.239 0.247 0.253 0.261
+## 1 0.409 0.263 0.197 0.131
+## 2 0.389 0.264 0.204 0.143
+## 3 0.429 0.261 0.189 0.121
+## 4 0.119 0.188 0.261 0.433
 ## 5   ...   ...   ...   ...
-## 6 0.143 0.205 0.264 0.389
-## 7 0.167 0.218 0.264  0.35
-## 8 0.338 0.263 0.223 0.176
-## 9 0.265 0.253 0.246 0.236
+## 6 0.147 0.207 0.264 0.383
+## 7 0.328 0.263 0.226 0.183
+## 8 0.389 0.264 0.204 0.143
+## 9 0.488 0.252 0.166 0.094
 \end{verbatim}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-12-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-12}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-14-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-14}
 \end{figure}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-17-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-17}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-19-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-19}
 \end{figure}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-20-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-20}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-22-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-22}
 \end{figure}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-22-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-22}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-24-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-24}
 \end{figure}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-26-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-26}
+{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-28-1}
+
+}
+
+\caption{ }\label{fig:unnamed-chunk-28}
 \end{figure}
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-Agresti2010-rz}
+Agresti, A. (2010). \emph{Analysis of ordinal categorical data}. Hoboken, NJ, USA: John Wiley \& Sons, Inc. \url{https://doi.org/10.1002/9780470594001}
+
 \bibitem[\citeproctext]{ref-Burkner2019-aw}
@@
 
+\bibitem[\citeproctext]{ref-Carifio2008-qd}
+Carifio, J., \& Perla, R.
+(2008).
+Resolving the 50-year debate around using and misusing likert scales.
+\emph{Medical Education}, \emph{42}, 1150--1152.
+\url{https://doi.org/10.1111/j.1365-2923.2008.03172.x}
+
+\bibitem[\citeproctext]{ref-Carifio2007-kk}
+Carifio, J., \& Perla, R.
+J.
+(2007).
+Ten common misunderstandings, misconceptions, persistent myths and urban legends about likert scales and likert response formats and their antidotes.
+\emph{Journal of Social Sciences}, \emph{3}, 106--116.
+\url{https://doi.org/10.3844/jssp.2007.106.116}
+
 \bibitem[\citeproctext]{ref-Christensen2019-cz}
@@
 
+\bibitem[\citeproctext]{ref-Cliff1996-qo}
+Cliff, N. (1996). Answering ordinal questions with ordinal data using ordinal statistics. \emph{Multivariate Behavioral Research}, \emph{31}, 331--350. \url{https://doi.org/10.1207/s15327906mbr3103_4}
+
 \bibitem[\citeproctext]{ref-Cliff2016-ck}
-Cliff, N. (2016). \emph{Ordinal methods for behavioral data analysis}. London, England: Psychology Press.
+Cliff, Norman. (2016). \emph{Ordinal methods for behavioral data analysis}. London, England: Psychology Press.
 
@@
 
-\bibitem[\citeproctext]{ref-Harrell-2015-no}
-Harrell, F. E. (2015). \emph{Regression modeling strategies}. Cham: Springer International Publishing. \url{https://doi.org/10.1007/978-3-319-19425-7}
+\bibitem[\citeproctext]{ref-Jamieson2004-xf}
+Jamieson, S. (2004). Likert scales: How to (ab)use them. \emph{Medical Education}, \emph{38}, 1217--1218. \url{https://doi.org/10.1111/j.1365-2929.2004.02012.x}
 
@@
 Kruschke, J. K. (2015). \emph{Doing bayesian data analysis}. \url{https://doi.org/10.1016/c2012-0-00477-2}
-
-\bibitem[\citeproctext]{ref-Kurz2023-zv}
-Kurz, A. S. (2023). \emph{Doing bayesian data analysis in brms and the tidyverse} (Version 1.1.0). Retrieved from \url{https://bookdown.org/content/3686/}
 

### 2024-01-03 R/utils.R

@@ -155,3 +155,30 @@ get_link <- function(link = c("logit", "probit")){
   }
   list(rfun = rfun, pfun = pfun, dfun = dfun, qfun = qfun)
 }
+
+#' Create k - 1 dummy variables from an ordinal variable
+#' @description
+#' Given an ordinal variable with `k` levels the function return a data.frame with k - 1 dummy variables.
+#'
+#' @param y a vector. Can be numeric, integer or (ordered) factor. If `y` is a factor, the function extract the underlying integer representation.
+#'
+#' @return a data.frame with k - 1 dummy variables.
+#' @export
+#'
+#' @examples
+#' k <- 4
+#' y <- sample(1:k, size = 10, replace = TRUE)
+#' dummy_ord(y)
+dummy_ord <- function(y){
+  if(!all(is.numeric(y) | is.integer(y) | is.factor(y))){
+    stop("y need to be numeric, integer or (ordered) factor")
+  }
+  # nothing for numeric and integers, safe for (ordered) factors
+  y <- as.integer(y)
+  yc <- sort(unique(y))
+  dummy <- lapply(yc, function(t) ifelse(y <= t, 1, 0))
+  nn1 <- Reduce(paste0, yc, accumulate = TRUE)
+  nn2 <- Reduce(paste0, yc, accumulate = TRUE, right = TRUE)
+  names(dummy) <- sprintf("y%svs%s", nn1[-length(nn1)], nn2[-1])
+  data.frame(dummy[-length(dummy)])
+}

### 2024-01-03 scripts/tables.R

@@ -1,0 +1,34 @@
+library(flextable)
+
+models_table <- data.frame(
+  stringsAsFactors = FALSE,
+  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
+  Link.Function = c("Equation",
+                    "$\\text{logit}(p) = \\text{log}(p / (1-p))$",
+                    "$z = \\Phi(p)$"),
+  Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
+  Inverse.Link.Function = c("Equation",
+                            "$e^{\\text{logit}(p)} / (1 + e^{\\text{logit}(p)})$","$\\Phi^{-1}(z)$"),
+  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
+)
+
+models_table[-1, ] |> 
+  flextable() |> 
+  autofit() |> 
+  set_header_labels(values = list(Model = "Model",
+                                  Link.Function = "Equation",
+                                  Link.Function.1 = "R Code",
+                                  Inverse.Link.Function = "Equation",
+                                  Inverse.Link.Function.1 = "R Code")) |>
+  add_header_row(values = c("", "Link Function", "Link Function", "Inverse Link Function", "Inverse Link Function")) |>
+  merge_h(1, part = "header") |>
+  theme_booktabs(bold_header = TRUE) |>
+  align(part = "all", align = "center") |> 
+  ftExtra::colformat_md(part = "all") |> 
+  save_as_docx(path = sprintf(here::here("paper", "tables", "%s.docx"), "notation-table"))
+
+tables <- filor::named_list(
+  models_table
+)
+
+saveRDS(tables, here::here("objects", "r-tabs.rds"))

### 2024-01-04 paper/paper-new.Rmd
update

@@
                       size = "scriptsize",
+                      collapse = TRUE,
                       message = FALSE,
-                      warning = FALSE)
+                      warning = FALSE,
+                      comment = "#>")
 ```
@@
 
-```{r include = FALSE}
+```{r scripts, include = FALSE}
+temp <- new.env()
 # figures
-source(here("scripts", "figures.R"), local = knitr::knit_global())
+# source(here("scripts", "figures.R"), local = temp)
 rfigs <- readRDS(here("objects", "r-figs.rds"))
@@
 # tables
-source(here("scripts", "figures.R"), local = knitr::knit_global())
+# source(here("scripts", "tables.R"), local = temp)
 rtabs <- readRDS(here("objects", "r-tabs.rds"))
+```
+
+```{r paper-functions}
+head_tail <- function(x, n = 5){
+  rbind(
+    head(x, n = n),
+    tail(x, n = n)
+  )
+}
 ```
@@
 @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
-We can identify three main classes: *cumulative models* [CM; @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe] and *adjacent category models*.
+We can identify three main classes: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe] and *adjacent category models*.
 The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.
@@
 Furthermore, CM are also used to model data under a signal detection theory framework [e.g., @DeCarlo2010-lj].
-The Figure \@ref(fig:fig-ordinal-models) depict the overall structure of the *cumulative* model.
-
-```{r fig-ordinal-models, fig.cap=filor::get_caption(), out.width="100%"}
-knitr::include_graphics("img/fig-ordinal-models.png")
-```
+The Figure \@ref(fig:fig-explain-cumulative) depict the overall structure of the *cumulative* model.
 
@@
 
-```{=tex}
 \begin{equation} 
@@
 \end{equation}
-```
 
 The $\mathbf{X} \boldsymbol{\beta}$ is the linear predictor $\eta$ that is the cumulative probability $P(Y \leq k)$ transformed using the link function $g()$.
-To obtain the probability of a single outcome $P(Y = k)$ we can compute the difference between cumulative probabilities as shown in Equation \@ref(eq:eq:prob-cum-model2).
-
-```{=tex}
+To obtain the probability of a single outcome $P(Y = k)$ we can compute the difference between cumulative probabilities as shown in Equation \@ref(eq:prob-cum-model2).
+
 \begin{equation} 
@@
 \end{equation}
-```
+
 There are always two special cases when computing the probability of a single outcome $Y$ that is when $Y = 1$ and $Y = k$.
@@
 \end{equation}
-```
+
 The crucial part is $\epsilon_i$ that is the random component of the model coming from a certain probability distribution.
@@
   xlim(c(-7, 7)) +
-  theme(legend.position = "bottom")
+  theme(legend.position = "bottom") +
+  viridis::scale_color_viridis(discrete = TRUE)
 
@@
   stat_function(fun = plogis, aes(color = "Logistic"), lwd = 1) +
-  xlim(c(-7, 7))
-
-plot_grid_common_legend(
-  dplot, 
-  pplot + theme(axis.title.y = element_blank()), nrow = 2
-)
+  xlim(c(-7, 7)) +
+  viridis::scale_color_viridis(discrete = TRUE)
+
+legend <- cowplot::get_legend(dplot + theme(legend.position = "bottom", legend.title = element_blank()))
+grid <- cowplot::plot_grid(dplot + rm_legend(), pplot + rm_legend())
+cowplot::plot_grid(grid, legend, nrow = 2, rel_heights = c(0.9, 0.1))
 ```
@@
 
-```{r dataset-example1}
-
-b1 <- log(3) # log odds ratio
-k <- 4
-n <- 100
-x <- rep(c("a", "b"), each = n/2)
-dat <- data.frame(x = x)
-probs <- rep(1/k, k) # for the group "a", uniform probabilities
-dat <- sim_ord_latent(~x, By = b1, prob = probs, data = dat, link = "logit")
-
-funs <- list(mean = mean, median = median, sd = sd)
-ss <- sapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
-ff <- t(table(dat$y, dat$x))
-
-ss <- data.frame(cbind(ss, ff))
-names(ss) <- c("mean", "median", "sd", paste0("Y", 1:4))
-
-ss$tot <- apply(ss[, 4:ncol(ss)], 1, sum)
-pp <- ss[, 4:7]/50
-ss <- round(ss, 2)
-ss$Y1 <- sprintf("**%s** (*p = %.2f*)", ss$Y1, pp$Y1)
-ss$Y2 <- sprintf("**%s** (*p = %.2f*)", ss$Y2, pp$Y2)
-ss$Y3 <- sprintf("**%s** (*p = %.2f*)", ss$Y3, pp$Y3)
-ss$Y4 <- sprintf("**%s** (*p = %.2f*)", ss$Y4, pp$Y4)
-
-ss$group <- rownames(ss)
-ss <- select(ss, group, everything(), -tot)
-ss
-```
-
-```{r tab-dataset-example1}
-#| eval: false
-
-flextable(ss) |> 
-  autofit() |> 
-  set_caption(caption = filor::get_caption()) |> 
-  align(part = "all", align = "center") |> 
-  ftExtra::colformat_md()
+```{r tab-dataset-example}
+dat <- rtabs$data_example_table$dat
+map(rtabs$data_example_table$table, ~gsub("\\**", "", .x)) |> 
+  bind_rows() |> 
+  kable(row.names = FALSE,
+        booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE) |> 
+  kable_styling(full_width = FALSE)
 ```
@@
 ```{r model-example-1}
+#| echo: true
 fit <- clm(y ~ x, data = dat, link = "logit")
@@
 Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect.
-As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [for an overview about contrasts coding schemes see @Schad2020-ht][^3]
+As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [see @Schad2020-ht][^3]
 
@@
 
-```{r}
+```{r odds-example}
 #| echo: true
@@
 
-```{r}
+```{r cumulative-odds-example}
 #| echo: true
@@
 
-```{r}
+```{r poa-logit-example}
 #| echo: true
@@
 # extracting the predicted probabilities for the two groups
-pr <- predict(fit, data.frame(x = unique(x)))$fit
+pr <- predict(fit, data.frame(x = unique(dat$x)))$fit
 
@@
 This can be directly demonstrated in the same way as for the logit model.
-We use the same dataset presented in Table \@ref(tab:tab-prop-ord-ex) but fitting a probit model.
+We use the same dataset of the previous simulation.
 The difference between the $z$ scores comparing the two groups calculated on the cumulative probability is the same regardless the $k - 1$ level.
 
-$$
+\begin{equation}
 z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
-$$
-
-```{r}
+(\#eq:probit-prop-odds)
+\end{equation}
+
+```{r poa-probit-example}
 #| echo: true
@@
 For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in $x$.
-For the probit model, the $\beta$ is the increase in terms of $z$ scores for a unit increase in $x$.
-This is very convenient especially for categorical variables because parameters can be intepreted as a Cohen's $d$ like measure.
+
+For the *probit* model, the $\beta$ is the increase in terms of $z$ scores for a unit increase in $x$.
+This is very convenient especially for categorical variables because parameters can be interpreted as a Cohen's $d$ like measure.
 Thinking about the latent distributions, the $\beta$ is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric $x$.
-The intepretation in terms of shifting the latent mean holds also for the logistic model.
+More formally the shift in the latent distribution is $\frac{\beta}{\sigma}$ (for the *probit* model $\sigma = 1$).
+The interpretation in terms of shifting the latent mean holds also for the logistic model.
 However, the standard deviation of the standard logistic regression is $\frac{\pi}{\sqrt{3}} \approx 1.81$[^4].
-The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\frac{\pi}{\sqrt{3}}$ standard deviations.
+The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\beta/(\frac{\pi}{\sqrt{3}})$ standard deviations.
 
@@
 For the standard logistic distribution $s = 1$ (as for the standard normal distribution).
-For the standard normal distribution $s^2 = \sigma^2 = 1$
-
-Another advantage of the probit model is the possibility to directly map parameters from signal detection theory [@Stanislaw1999-jr; @Green1966-gy] into the model coefficients.
-In practical terms, the thresholds $\alpha$ are the criterion cut-offs and the $\beta_1$ is the d$'$ parameter [@DeCarlo1998-ay; @Knoblauch2012-to].
-
-qualcosa in più qui
 
@@
 
-There are mainly two ways to simulate data.
-The first method concerns simulating starting from the latent formulation of the ordinal model.
-Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.
+In this tutorial we present two method for simulating ordinal data.
+The first method concerns calculating the probabilities of each $k$ of $Y$ as a function of predictors and then sampling from a multinomial distribution.
+The second method simulate the latent variable $Y^\star$ as a function of predictors.
 
@@
 
-The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution[^5].
-This is similar to the general method to simulate data for a generalized linear model where the linear predictors $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.
-The following code box simulate a $n$ binary trials with a continuous predictor $x$.
-
-[^5]: In R we are using the `sample()` function that simulate values from a *categorical* distribution. The *categorical* distribution is considered a special case of the multinomial distribution
-
-```{r}
-xl <- seq(0, 1, 0.1)
-x <- rep(xl, 10)
-n <- length(x)
-b0 <- qlogis(0.01)
-b1 <- 8
-lp <- b0 + b1 * x # linear predictor
-y <- rbinom(n, 1, plogis(lp)) # sampling from a binomial distribution using the vector of probabilities
-p <- tapply(y, x, mean)
-plot(unique(x), p, pch = 19)
-curve(plogis(x, -(b0/b1), 1/b1), add = TRUE, lwd = 2, col = "firebrick")
-```
-
-Essentially the `lp` ($\eta_i$) vector contains true probability of success ($p_i|x_i = g^{-1}(\eta_i)$) for the $x_i$ level of the predictor.
-Then to introduce the random part of the model, we can use $p_i$ to generate binary values from a Binomial (in this specific case a Bernoulli) distribution.
-This create `y` that is the vector of 0-1 values generated using the true probabilities of success.
-The supplementary materials contains a more extensive example for simulating data for generalized linear models.
-
-We can apply the same idea to an ordinal outcome but we need $k - 1$ equations (compared to the previous example) where $k$ is the number of ordinal levels.
-Let's simulate a similar design with a continuous $x$ predictor and $k = 4$ options.
-We fix the baseline probabilities where $x = 0$ as uniform thus $p(y_1) = p(y_2) = ...
-p(y_k) = 1/k$.
-The proposed workflow for the simulation is:
-
-1.  Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
-2.  Define the regression coefficients $\beta$s.
-3.  Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
+For the first method we need to calculate $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution^[In R we are using the `sample()` function that simulate values from a *categorical* distribution.
+The *categorical* distribution is considered a special case of the multinomial distribution].
+This is similar to the general method to simulate data for a generalized linear model where the linear predictor $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.
+
+As a simple example, let's simulate two groups or condition with $n = 100$ participants responding to an item with $k = 4$ ordered options.
+We simulate that the second group has more probability of responding higher categories of $Y$.
+As explained in the previous sections, we can summarise the effect size of a CM (assuming POA) using a single $\beta = \log(OR)$.
+We can assume that the odds ratio is $OR = 2$.
+For simplicity, the probabilities of the first group $x = 0$ are uniform thus $P(Y = 1|x_0) = ...
+P(Y = k|x_0) = 1/k$.
+The following code summarise the first steps of the simulation.
+Basically we define the simulation parameters, we calculate the $k - 1$ thresholds $\alpha$ and we apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2).
+In this way we calculated the true probability of each $Y$ level for the two groups.
+
+An important step is converting from $\alpha$ to probability and the opposite.
+This steps are implemented in the `alpha_to_prob()` and `prob_to_alpha()` functions.
+Basically given the input and the link function, these functions return the thresholds associated with $k$ probabilities or the $k - 1$ thresholds.
+
+```{r alpha-prob-example}
+#| echo: true
+#| collapse: true
+k <- 5
+(p <- rep(1/k, k)) # uniform
+names(p) <- paste0("y", 1:k)
+
+(alpha <- prob_to_alpha(p, link = "logit")) # or prob_to_alpha(p, "probit")
+alpha_to_prob(alpha, link = "logit")
+```
+
+```{r simulating-from-multinomial1}
+#| echo: true
+#| collapse: true
+
+## SIMULATION PARAMETERS
+
+N <- 100 # sample size
+or <- 2 # odds ratio
+k <- 4 # number of ordinal alternatives
+probs0 <- rep(1/k, k) # probabilities for the first group
+alpha <- prob_to_alpha(probs0, link = "logit")
+dat <- data.frame(x = rep(c(0, 1), each = N/2))
+
+## LINEAR PREDICTOR
+
+# calculate linear predictor using equation TODO put equation
+# obtaining k - 1 equations
+
+lp <- lapply(alpha, function(a) a - log(or) * dat$x)
+names(lp) <- sprintf("lp_leq%s", 1:(k - 1)) # giving appropriate names
+lp <- data.frame(lp)
+head(lp)
+
+## CUMULATIVE PROBABILITIES
+
+# apply the inverse of the link function (invlogit) to calculate cumulative probabilities
+cump <- lapply(lp, plogis)
+cump <- data.frame(cump)
+names(cump) <- sprintf("cump_leq%s", 1:(k - 1)) # giving appropriate names
+head(cump)
+
+## PROBABILITIES OF Y
+
+# for each row, we can calculate P(Y = k) using equation TODO put equation
+# P(Y = 1) = P(Y <= 1)
+# P(Y = 2) = P(Y <= 2) - P(Y <= 1)
+# P(Y = 3) = P(Y <= 3) - P(Y <= 2)
+# P(Y = 4) = 1 - P(Y <= 3)
+
+# adding a columns of 0 and 1, then diff() for adjacent differences
+cump <- cbind(0, cump, 1)
+p <- apply(cump, 1, diff, simplify = FALSE)
+
+p <- data.frame(do.call(rbind, p)) # collapse list of rows into a dataframe
+names(p) <- sprintf("p%s", 1:k) # giving appropriate names
+head(p)
+
+# probabilities for id = 1 (x = 0) and id = 51 (x = 1)
+p[c(1, 51), ]
+
+## CUMULATIVE ODDS RATIO
+
+# calculate the (cumulative) odds ratio
+
+x0 <- cump[1, 2:k]
+x1 <- cump[51, 2:k]
+
+# this is the same as the or (the b1) and we are assuming POA
+odds(x0) / odds(x1)
+```
+
+```{r simulating-from-multinomial2}
+#| echo: true
+#| collapse: true
+
+## SAMPLING FROM THE MULTINOMIAL DISTRIBUTION
+
+# example of a random Y outcome based on the probabilities
+sample(x = 1:k, size = 1, prob = p[1, ])
+
+# we can apply it to the full dataset, this step lead to different results
+# each time we run the code because we are sampling from a distribution
+
+dat$y <- apply(p, 1, function(ps) sample(1:k, size = 1, prob = ps))
+head(dat)
+
+# let's compute the observed probabilities, to be compared to the true
+# probabilities
+
+# observed
+(op <- prop.table(table(dat$x, dat$y), margin = 1))
+
+# true (just selecting a row from x = 0 and x = 1)
+p[c(1, 51), ]
+
+# similarly we can compute the observed cumulative odds ratios
+
+(cum_op <- apply(op, 1, cumsum))
+odds(cum_op[, 1]) / odds(cum_op[, 2])
+
+# the odds ratios are not the same as the parameter. as we increase N
+# the parameter will converge to the true value
+```
+
+```{r simulating-from-multinomial3}
+#| echo: true
+#| collapse: true
+dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
+fit <- clm(y ~ x, data = dat, link = "probit")
+summary(fit)
+```
+
+We can generalize the previous workflow as:
+
+1.  Define simulation parameters, baseline probabilities, etc. as in the previous workflow
 4.  Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
 5.  Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
-6.
-Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities.
-For example, the probability of $P(y = 2) = P(y \leq 2) - P(y \leq 1)$.
-This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities because the probability of the first $Y = 1$ level is the cumulative probability from $- \infty$ to the first threshold and the probability of $Y = k$ is obtained removing $P(Y \leq k - 1)$ from one.
+6.  Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities.
 7.  Sample $n$ outcomes from a multinomial distribution, choosing between $k$ alternatives with the calculated probabilities.
-
-```{r}
-k <- 4  # number of options (1)
-n <- 1e5 # number of observations (1)
-b1 <- 0.5 # beta1, the shift in the latent distribution (2)
-probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
-dat <- data.frame(x = rnorm(n)) # create dataframe
-lp <- lapply(alphas, function(a) a - b1 * dat$x) # k - 1 linear predictors (4)
-names(lp) <- sprintf("g[p(y <= %s)]", 1:(k - 1)) # giving appropriate names
-```
-
-The linear predictors ($\eta$) can be seen in the Figure ref. The $g$ in this case is the cumulative *probit* function $\Phi$.
-
-```{r}
-#| echo: false
-lpt <- lp
-names(lpt) <- latex2exp::TeX(sprintf("$g[p(y \\leq %s)]$", 1:(k - 1)))
-datl <- cbind(dat, lpt)
-datl <- tidyr::pivot_longer(datl, 2:ncol(datl), names_to = "cumy", values_to = "lp")
-
-ggplot(datl, aes(x = x, y = lp, color = cumy)) +
-  facet_wrap(~cumy, labeller = label_parsed) +
-  geom_hline(yintercept = alphas, lty = "dashed") +
-  geom_line(show.legend = FALSE) +
-  theme_minimal(20) +
-  ylab("Linear Predictor")
-```
-
-Now we can apply the inverse of the link function $g^{-1} = \Phi^{-1}$ to calculate the corresponding cumulative probabilities.
-
-```{r}
-cump <- lapply(lp, pnorm) # inverse of the link function (5)
-cump <- data.frame(cump)
-p <- apply(cbind(0, cump, 1), 1, diff, simplify = FALSE) # probability of each k outcome (6)
-p <- do.call(rbind, p)
-```
-
-```{r}
-p <- data.frame(p)
-names(p) <- sprintf("p_y%s", 1:k)
-filor::trim_df(data.frame(p))
-```
-
-We can plot the expected effect of $\beta_1$ on the $k$ probabilities using the `num_latent_plot()` function (see Figure \@ref(fig:sim-with-probs-lat-plot)).
-Finally we sample using the `sample()` function using the calculated probabilities.
-
-```{r sim-with-probs-lat-plot, fig.cap="sim-with-probs-lat-plot"}
-num_latent_plot(x = dat$x, b1 = b1, probs = probs0, link = "probit")
-```
-
-```{r}
-dat$y <- apply(p, 1, function(ps) sample(1:k, 1, prob = ps))
-```
-
-To check the simulation result we can increase the number of observations, fit the model using the `ordinal::clm()` function and assess the recovery of simulated parameters.
-
-```{r}
-dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
-fit <- clm(y ~ x, data = dat, link = "probit")
-```
-
-```{r tab-sim-probs}
-truth <- c(alphas, b1)
-names(truth) <- c(names(fit$alpha), names(fit$beta))
-#clm_table(fit, truth = truth)
-```
+8.  Fit the appropriate model using `ordinal::clm()`
+
+A general suggestion is to increase the sample size $n$ to a very large value reducing the random variability to check if simulated parameters are recovered.
 
@@
 
+```{r simulating-from-latent1}
+#| echo: true
+#| collapse: true
+
+## SIMULATION PARAMETERS
+
+N <- 1e3 # sample size
+or <- 4 # odds ratio, higher here just for a more clear plot
+k <- 4 # number of ordinal alternatives
+probs0 <- rep(1/k, k) # probabilities for the first group
+alpha <- prob_to_alpha(probs0, link = "logit")
+dat <- data.frame(x = rep(c(0, 1), each = N/2))
+
+## LINEAR PREDICTOR
+
+# calculate the linear predictor using equation TODO put equation
+dat$lp <- log(or) * dat$x
+
+# add the random part by sampling errors from a standard logistic (or normal) distribution
+dat$ystar <- dat$lp + rlogis(N, location = 0, scale = 1)
+
+# cut the latent distribution. The + 1 because the first category is 0 by default.
+dat$y <- findInterval(dat$ystar, alpha) + 1
+
+head_tail(dat, n = 3)
+```
+
+The Figure \@ref(fig:fig-sim-from-latent) depict the simulated $Y^{*}$ and the corresponding ordinal value.
+As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.
+
+```{r fig-sim-from-latent}
+alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))
+
+dat |> 
+  ggplot(aes(x = factor(x), y = ystar)) +
+  geom_point(aes(color = factor(y)),
+             position = position_jitter(width = 0.1),
+             alpha = 0.7) +
+  geom_hline(yintercept = alpha) +
+  annotate("label", x = 0.5, y = alpha, label = alpha_n) +
+  ylab(latex2exp::TeX("$Y^{*}$")) +
+  xlab("Group (x)") +
+  mtheme() +
+  viridis::scale_color_viridis(discrete = TRUE, direction = -1) +
+  theme(legend.position = "bottom",
+        legend.title = element_blank())
+```
+
+```{r simulating-from-latent2}
+#| echo: true
+#| collapse: true
+dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
+fit <- clm(y ~ x, data = dat, link = "logit")
+summary(fit)
+```
+
+The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function. Details about the function can be found on the Github repository LINK.
+
 ```{r}
-k <- 4  # number of options (1)
-n <- 1e5 # number of observations (1)
-b1 <- 0.5 # beta1, the shift in the latent distribution (2)
-probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
-dat <- data.frame(x = rnorm(n)) # create dataframe
-dat$ystar <- b1 * dat$x + rnorm(n, 0, 1) # linear predictor (4)
-dat$y <- findInterval(dat$ystar, alphas) + 1 # cut the latent distribution (5). The + 1 because the first category is 0 by default.
-```
-
-The Figure (put crossref) depict the simulated $Y^{*}$ and the corresponding ordinal value. As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.
-
-```{r}
-alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))
-
-dat |> 
-  slice_sample(n = 1000) |> 
-  mutate(y = findInterval(ystar, alphas) + 1) |> 
-  ggplot(aes(x = x, y = ystar)) +
-  geom_point(alpha = 0.6, aes(color = factor(y)), size = 3) +
-  theme_minimal(15) +
-  theme(legend.position = "bottom",
-        legend.title = element_blank()) +
-  ylab(latex2exp::TeX("$Y^{*}$")) +
-  geom_hline(yintercept = alphas) +
-  annotate("label", x = min(dat$x), y = alphas, label = alpha_n)
-```
-
-```{r}
-dat$y <- ordered(dat$y) # make an ordered factor in R where 1 < 2 < 3 < 4
-fit <- clm(y ~ x, data = dat, link = "probit")
-```
-
-```{r tab-sim-latent}
-truth <- c(alphas, b1)
-names(truth) <- c(names(fit$alpha), names(fit$beta))
-#clm_table(fit, truth)
-```
-
-The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function.
-
-```{r}
-#| results: asis
-filor::print_fun(funs$sim_ord_latent)
+#| echo: true
+#| collapse: true
+N <- 100 # sample size
+or <- 4 # odds ratio, higher here just for a more clear plot
+k <- 4 # number of ordinal alternatives
+probs0 <- rep(1/k, k) # probabilities for the first group
+alpha <- prob_to_alpha(probs0, link = "logit")
+dat <- data.frame(x = rep(c(0, 1), each = N/2))
+
+# same as the previous simulation
+dat <- sim_ord_latent(~x, By = log(or), prob = probs0, data = dat, link = "logit")
+
+head_tail(dat, n = 3)
 ```
@@
 
-The two proposed simulation workflows can be really flexible for simulating complex models with multiple predictors and interactions.
-The critical part of the simulation is setting appropriate and empirically meaningful parameters values.
-The thresholds are usually not the crucial part of the statistical inference and model interpretations.
-In fact, they can be considered as the intercept in standard linear models.
-In any case to set meaningful $\alpha$ values the best strategy is to convert them into the corresponding probabilities according to the link function.
-The thresholds are essentially quantiles of the latent distribution (logistic or normal) without a direct empirical interpretation.
-The distance between the thresholds determine the probabililty assigned to a specific $Y$ value.
-The function `show_alpha()` allow to produce a meaningful plot starting from the thresholds or the probabilities.
-For example, the Figure \@ref(fig:fig-show-th-example) depict the threshold/probabilities associated with a four-level ordinal variable.
-
-```{r}
-#| eval: false
-show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))
-```
+The previous simulation can be easily extended by adding predictors and their interactions.
+The crucial part is setting appropriate and empirically meaningful parameters.
+Thresholds $\alpha$ are usually not of main interest [but see the location-shift models @Tutz2022-dg] and can be considered as intercepts in standard linear regression.
+The thresholds are quantiles of the latent distribution that produced a certain probability distribution of the ordinal variable.
+To set meaningful $\alpha$ values we can convert probabilities into thresholds (`alpha_to_prob()`).
+The function `show_alpha()` create a meaningful visual representation of using a specific set of thresholds (see Figure \@ref(fig:fig-show-th-example)).
+The function can be used as `show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))`.
+In regression terms, the thresholds determine the probabilities when $x = 0$ (as the intercept).
+Thus with two groups for example, the thresholds are the $k$ probabilities of the ordinal variable $Y$ for the reference group.
 
@@
 p1 <- show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))
-p2 <- show_alpha(prob = c(0.5, 0.3, 0.1, 0.1), link = "probit", plot = c("latent", "probs"))
-p3 <- show_alpha(prob = c(0.1, 0.2, 0.5, 0.2), link = "probit", plot = c("latent", "probs"))
-plot_grid(p1 + rm_legend(), p2 + rm_legend(), p3, ncol = 1)
-```
-
-Thus the thresholds can be fixed according to the desired probabilities distribution.
-For example, when simulating the difference between two groups on an ordinal variable, the threshold can be set thinking about the response distribution in the reference group.
-The `alpha_to_prob()` and `prob_to_alpha()` functions can be used to convert between probabilities and thresholds.
-
-```{r}
-#| collapse: true
-# k = 4, thus k probabilities and k - 1 thresholds
-prob_to_alpha(prob = c(0.4, 0.1, 0.3, 0.2), link = "probit")
-alpha_to_prob(alpha = c(-2, 0, 1), link = "probit")
+p2 <- show_alpha(prob = c(0.1, 0.2, 0.5, 0.2), link = "probit", plot = c("latent", "probs"))
+plot_grid(p1, p2, ncol = 1)
 ```
@@
 
-As reported in the sections above, the regression coefficients have a different interpretation according to the link function.
-Fortunately, both parameters have an intuitive interpretation given that they are directly expressed in a interpretable scale.
-For probit models we can think about standardized mean differences (e.g., Cohen's $d$) between categorical variables or standardized regression coefficient for a numerical predictor.
-For logit models we can fix the parameters thinking about the odds ratios.
+Regression coefficients are the effect size parameters.
+For *probit* models we can set the $\beta_j$ to be in standardized (i.e., Cohen's $d$-like) units.
+For a categorical variable as the previous example with the group, $\beta$ is the degree of separation in standard deviation unit between the two latent distributions.
+For *logit* models we can set the odds ratio.
 Meaningful odds ratios can be derived from previous literature, meta-analyses or converting to other effect sizes.
@@
 Using their approach, a Cohen's $d$ of 0.5 usually considered a plausible medium effect size corresponds to and odds ratio of $\approx 2.47$.
-Another more informative strategy is about plotting the predicted probabilities $g^{-1}(\eta)$ associated with certain regression coefficients.
-The `cat_latent_plot()` and `num_latent_plot()` functions can be used to see the impact of choosing a specific regression coefficients for respectively a categorical (Figures \@ref(fig:fig-example-cat-latent)) and numerical predictor (Figures \@ref(fig:fig-example-num-latent)).
+
+We can also calculate and plot the predicted probabilities (i.e., $g^{-1}(\eta)$, without actually sampling) given the predictors and the chosen regression coefficients.
+In this way we can try different values and see if predicted probabilities are plausible or not.
+The `cat_latent_plot()` and `num_latent_plot()` functions can be for respectively a categorical (Figures \@ref(fig:fig-example-cat-latent)) and numerical predictor (Figures \@ref(fig:fig-example-num-latent)).
 
 ```{r fig-example-cat-latent, fig.cap=filor::get_caption()}
-#| echo: true
 cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")
@@
 ```{r fig-example-num-latent, fig.cap=filor::get_caption()}
-#| echo: true
-num_latent_plot(x = runif(100), b1 = 2, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit")
-```
-
-## Scale Effects
-
-The default ordinal regression model assume that the variance of the underlying latent distribution is the same across condition.
-This is similar to a standard linear regression assuming the homogeneity of variance.
-For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy].
-In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution.
-This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
-The Equation (number here) expand the previous model including the linear predictor on the scale of the latent distribution.
-
-```{=tex}
-\begin{equation}
-g(P(Y \leq k)) = \frac{\alpha_k - X\beta}{e^{X\zeta}}
-\end{equation}
-```
-```{=tex}
-\begin{equation} 
-Y^\star_i = \eta + \epsilon_i
-\epsilon_i \sim \mathcal{N}(0, e^{X\zeta})
-\end{equation}
-```
-Where $X\zeta$ is the linear predictor $\eta$ for the scale of the distribution.
-By dafult for both the logit and probit model the scale is fixed to 1.
-On scale-location models we put predictors on both parameters.
-Given that the scale cannot be negative we use a log link function $\eta = \text{log}(X\zeta)$.
-Figure \@ref(fig:example-unequal-variance) depict an example of a comparison between two groups where the two underlying latent distributions have unequal variance.
-Ignoring the heterogeneity of variances could lead to underestimation of the effect size, reduced power and inflated type-1 error rate (qualche ref qui).
-Furtermore, as suggested by @Tutz2017-de location-scale models can be considered as a more parsimonius approach compared to partially or completely relaxing the proportional odds assumption.
-Allowing the scale to be different as a function of the predictor create more modelling flexbility.
-Furthemore, two groups could be theoretically different only in the scale of the latent distribution with a similar location.
-In this example, the only way to capture group differences is by including a scale effect.
-The Figure \@ref(fig:fig-scale-effect-example) depict the impact of having different scales between two groups on the ordinal probabilities.
-
-@Tutz2022-dg provide a very clear and intuitive explanation of what happen when including a scale effect and how to interpret the result.
-As suggested before, the scale-location model allow to independently predict changes in the location and the scale.
-While location shifts are simply interpreted as increasing/decreasing the latent $\mu$ or the odds of responding a certain category scale effects are not straightforward.
-As the scale increase (e.g., the variance increase) there is an higher probability mass on extreme categories.
-On the other side as the scale decrease, responses are more concentrated on single categories.
-The categories are determined by the location parameter.
-For example, if one group have a certain latent mean $\mu_1$ and a small scale $\sigma^2 = 1/3$ (thus one third compared to the standard version of the distribution), all responses will be focused on categories around the latent mean.
-On the other side, increasing the scale will increase the cumulative probabilities for all categories and for values that tends to infinity extreme categories are preferred.
-Clearly, the scale parameter can somehow be interpreted as the response style (concentrated or variable).
-This is conceptually similar but implemented and interpreted differently to shift-location models [@Tutz2022-dg; Tutz2020-xq] where thresholds are allowed to vary as a function of predictors.
-Both models tries to increase the flexibility of modelling the probability structure of the responses by adding extra paprameters to predict the probabilities of the ordinal responses.
-
-```{r}
-sigma_high <- cat_latent_plot(m = c(0, 0), 
-                s = c(1, 2), 
-                probs = rep(1/4, 4), 
-                link = "probit", 
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$"))
-sigma_low <- cat_latent_plot(m = c(0, 0), 
-                s = c(1, 0.5), 
-                probs = rep(1/4, 4), 
-                link = "probit", 
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$"))
-sigma_high_mu1 <- cat_latent_plot(m = c(0, 1), 
-                s = c(1, 2), 
-                probs = rep(1/4, 4), 
-                link = "probit", 
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$, $\\mu_2 = 1$"))
-sigma_low_mu1 <- cat_latent_plot(m = c(0, 1), 
-                s = c(1, 0.5), 
-                probs = rep(1/4, 4), 
-                link = "probit", 
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$, $\\mu_2 = 1$"))
-plot_grid(sigma_high, sigma_low, sigma_high_mu1, sigma_low_mu1)
-```
-
-### Simulating scale effects
-
-The location-scale model can be simulated using the `sim_ord_latent()` function and providing the predictors for the scale parameters.
-Given the `log` link function, predictors are provided on the `log` scale.
-For example, we simulate the effect of a binary variable $x$ representing two independent groups predicting the $k = 5$ response.
-We simulate a *location* effect of $\beta_1 = 0.5$ (in *probit* scale) and $\zeta_1 = \text{log}(2) = 0.70$.
-The first group has a $\sigma = 1$ and the second group has $\sigma = 2$.
-Again we simulate that the baseline probabilities are uniform for the first group.
-
-```{r}
-k <- 5  # number of options (1)
-n <- 1e5 # number of observations (1)
-b1 <- 0.5 # beta1, the shift in the latent distribution (2)
-z1 <- log(2) # zeta1, the change in the scale
-probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
-dat <- data.frame(x = rep(c(0, 1), each = n/2))
-dat <- sim_ord_latent(~x, scale = ~x, By = b1, Bscale = z1, prob = probs0, data = dat, link = "probit")
-fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
-```
-
-The table (put ref) reports the simulation results.
-
-```{r}
-#| eval: false
-truth <- c(alphas, b1, z1)
-names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
-#clm_table(fit, truth)
-```
-
-To better understand the impact of assuming (or simulating) a different latent scale we fit $k - 1$ binomial regressions and check the estimated coefficients.
-We are not simulating a specific beta for each outcome but simulating a scale effect is actually impacting the regression coefficients.
-When generating data for a binary outcome the linear predictor is composed by $\eta = \beta_0 + \beta_1x$.
-The threshold $\alpha$ and slope of the function can be estimated using $\alpha = -\frac{\beta_0}{\beta_1}$ and the slope is $\frac{1}{\beta_1}$ [@Knoblauch2012-to; @Faraggi2003-hm].
-Under the proportional odds assumption, there is only a change in thresholds $\alpha$ this a shift in the sigmoid along the $x$ axis.
-When including a scale effect a change in the sigmoid is combined with a change in the slope.
-
-```{r}
-set.seed(2023)
-k <- 4
-n <- 1e5
-b1 <- 3
-d1 <- log(2)
-dat <- data.frame(x = runif(n))
-dat <- sim_ord_latent(~x, ~x, By = b1, Bscale = d1, prob = rep(1/k, k), data = dat, link = "probit")
-
-dat$y1vs234 <- ifelse(dat$y <= 1, 1, 0)
-dat$y12vs34 <- ifelse(dat$y <= 2, 1, 0)
-dat$y123vs4 <- ifelse(dat$y <= 3, 1, 0)
-
-dat$y <- ordered(dat$y)
-fit <- clm(y ~ x, scale = ~x, data = dat, link = "probit")
-
-fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "probit"))
-fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "probit"))
-fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "probit"))
-
-fits <- list(y1vs234 = fit1vs234, fit12vs34 = fit12vs34, fit123vs4 = fit123vs4)
-
-lapply(fits, function(x) coef(x)) |> 
-  bind_rows(.id = "model") |> 
-  mutate(xp = list(seq(0, 5, 0.01))) |> 
-  unnest(xp) |> 
-  ggplot(aes(x = xp, y = `(Intercept)` + x * xp)) +
-  geom_line(aes(color = model))
+num_latent_plot(x = runif(100), b1 = 2, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit") +
+  mtheme() +
+  viridis::scale_color_viridis(discrete = TRUE)
 ```
@@
 In this way $\beta_1$ will be the main effect of $X_1$, $\beta_2$ the main effect of $X_2$ and $\beta_3$ the interaction (thus the difference of differences).
-
-$$
-g(P(Y \leq k))^{-1} = \alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i}
-$$
-
-```{r}
-n <- 1e5
+Equation \@ref(eq:interaction-equation) depict the model formula.
+
+\begin{equation}
+P(Y \leq k) = g^{-1}(\alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i})
+(\#eq:interaction-equation)
+\end{equation}
+
+```{r cat-cat-simulation-example}
+#| echo: true
+#| collapse: true
+n <- 100
+k <- 3
+betas <- c(b1 = 0, b2 = 1, b3 = 0.5) # b1 = main effect X1, b2 = main effect X2, b3 = interaction
+
 dat <- expand.grid(x1 = c("a", "b"), x2 = c("c", "d"), n = 1:n)
@@
 dat$x2 <- factor(dat$x2)
+
+# sum to 0 coding
 contrasts(dat$x1) <- c(0.5, -0.5)
 contrasts(dat$x2) <- c(0.5, -0.5)
-bprobs <- rep(1/5, 5)
-
-dat <- sim_ord_latent(~x1*x2, By = c(0, 1, 0.5), prob = bprobs, link = "probit", data = dat)
+probs0 <- rep(1/k, k)
+
+dat <- sim_ord_latent(~ x1 * x2, By = c(0, 1, 0.5), prob = probs0, link = "probit", data = dat)
 fit <- clm(y ~ x1 * x2, data = dat, link = "probit")
@@
 
-The parameters intepretation is the same as introduced in the general case.
-The thresholds $\alpha$ are fixed the determines the probabilities $P(Y = k)$ when all predictors are zero.
-In this case by doing `alpha_to_prob(fit$alpha, link = "probit")` we should recover the `bprobs` vector.
+The parameters interpretation is the same as introduced in the general case.
+The thresholds $\alpha$ are fixed representing the probabilities $P(Y = k)$ when all predictors are zero.
+In this case by doing `alpha_to_prob(fit$alpha, link = "probit")` we should recover the `probs0` vector.
 `x1` is the main effect thus the difference in $z$ scores between *a* and *b* averaging over `x2`.
 The same holds for `x2`.
-`x1:x2` is the difference of differences in $z$ scores thus $(ad - ac) - (bd - bc)$.
-
-To visualize the effects there is the `ggeffect::ggpredict()` function that compute the predictions according to the specified predictors combination.
-In this case we can plot the interaction using the estimated probability for each group and $Y$ response.
-
-```{r}
-# terms =  c("x1", "x2") means to plot the interaction
-plot(ggpredict(fit, terms = c("x1", "x2")))
-```
-
-## Numerical by factor interaction
-
-Another common scenario is to compute and interaction between a numerical variable $x$ and a factor $g$.
-For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution $x \sim \mathcal{U}(0, 1)$.
-Again, the thresholds $\alpha$ represents the baseline probabilities when $x = 0$ and $g = 0$.
-
-```{r}
-n <- 1e5
+`x1:x2` is the difference of differences in $z$ scores (i.e., the interaction).
+Figure \@ref(fig:fig-effects-2-by-2-interaction) depicts the model results.
+
+```{r fig-effects-2-by-2-interaction}
+# plot(ggpredict(fit, terms = c("x1", "x2")))
+ggpredict(fit, terms = c("x1", "x2")) |> 
+  data.frame() |>
+  mutate(response.level = paste("Y = ", response.level)) |> 
+  ggplot(aes(x = x, y = predicted, color = group)) +
+  geom_point(size = 4) +
+  geom_line(aes(group = group), linewidth = 1) +
+  facet_grid(~response.level) +
+  ylab("Probability") +
+  mtheme() +
+  viridis::scale_color_viridis(discrete = TRUE)
+```
+
+## Numerical by categorical interaction
+
+Another common scenario is to compute and interaction between a numerical variable $x$ and a categorical variable $g$.
+For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution between 0 and 1.
+
+```{r cat-num-simulation-example}
+#| echo: true
+#| collapse: true
+n <- 100
+k <- 3
 dat <- data.frame(x = runif(n),
@@
 contrasts(dat$g) <- c(0.5, -0.5)
-bprobs <- rep(1/5, 5)
-
-dat <- sim_ord_latent(~ x * g, By = c(0.3, 0.5, 0.5), prob = bprobs, link = "probit", data = dat)
+probs0 <- rep(1/k, k)
+
+dat <- sim_ord_latent(~ x * g, By = c(0.3, 0.5, 0.5), prob = probs0, link = "probit", data = dat)
 fit <- clm(y ~ x * g, data = dat, link = "probit")
@@
 The `x:g1` is the difference in slopes between the two groups.
-
-Regardless of the model complexity, the `sim_ord_latent()` function can be expanded by including more predictors.
-Clearly, as the model complexity increase, choosing meaningful and empirically plausible parameters became more challenging.
+Figure \@ref(fig:fig-effects-num-by-cat-interaction) depicts the model results.
+
+```{r fig-effects-num-by-cat-interaction}
+# plot(ggpredict(fit, terms = c("x[all]", "g")))
+ggpredict(fit, terms = c("x[all]", "g")) |> 
+  data.frame() |>
+  mutate(response.level = paste("Y = ", response.level)) |> 
+  ggplot(aes(x = x, y = predicted)) +
+  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group),
+              alpha = 0.3) +
+  geom_line(aes(group = group), linewidth = 1) +
+  facet_grid(~response.level) +
+  ylab("Probability") +
+  mtheme() +
+  viridis::scale_fill_viridis(discrete = TRUE)
+```
 
@@
 
-In the previous sections we introduced how to simulate standard and scale-location cumulative models and how to fit the corresponding models using the `clm()` function from the `ordinal` package.
-In this section we introduce how to compute the power for some common scenarios.
-We present an example in a very general form that can be expanded for specific use cases.
-For example, by simulating a model fixing all effects to zero it is possibile to estimate the type-1 error rate or fixing the sample size and varying the effect size it is possible to do a sensitivity analysis focusing on the smallest effect size of interest (ref qui).
+In this section we introduce how to estimate the statistical power.
+Formally the power is defined as the probability of correctly rejecting the null hypothesis $H_0$.
+In some simple case, the power can be computed analytically however using data simulation we can handle every scenario.
+We present very general example that, as for the previous simulations, can be extended and modified.
+
+For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting $H_0$) or assessing an eventual bias in model estimates.
 The general workflow for a power simulation can be summarized as:
 
-1.  Choose the research design, e.g., 2x2 factorial design
-2.  Indentifify the specific effect to focus on. For example the interaction effect, a main effect or a specific contrast
-3.  Write the model equation similarly to what presented in the previous sections and extract the model coefficients
-4.  Choose one of the proposed strategy for simulating data and follows the corresponding steps
-5.  Fit the model and extract the relevant statistics e.g., the p-value of the interaction
-6.  Repeat the simulation a large number of times (e.g, 10000) and store the relevant values from each simulation
-7.
-Summarise the simulation results in the appropriate way.
-For example, the power of the interaction is the number of times the p-value for the interaction test is lower than the critical level divided by the number of simulations.
-
-To make a practical example, we simulate the power of a cumulative link model in detecting an effect of $d = 0.4$ (probit model) with $k = 6$ and $n = 40$ participants per group.
-We fix the baseline probabilities in the first group as uniform.
-
-descrizione migliore
-
-```{r}
-n <- 40
-k <- 6
-d <- 0.4
-nsim <- 1e3 # higher is better, just for example
-bprobs <- rep(1/k, k)
+1.  Specify the research design, e.g., 2x2 factorial design
+2.  Specify the effect/parameter of interest, e.g., the interaction effect
+3.  Define the simulation conditions, e.g., a range of sample size, effect size, etc.
+4.  Implement one of the simulation workflow described in the previous sections
+5.  Repeat the simulation a large number of times (e.g, 10000) and store the relevant values from each simulation
+6.  Summarise the simulation results
+
+To make a practical example, we simulate the power of detecting a group difference of $d = 0.4$ (assuming a *probit* model).
+Participants respond to an ordinal variable with $k = 5$.
+The iteration is performed using a `for` loop that repeat 1000 times the data simulation, model fitting and extract the p-value for the $\beta_1$.
+The power is then estimated as the number of p-values lower than the critical level over the number of simulations.
+
+```{r power-example}
+#| echo: true
+#| collapse: true
+n <- 40 # sample size
+k <- 5  # number of ordinal variables
+d <- 0.4 # effect size (i.e., our regression coefficients)
+nsim <- 1e3 # higher is better, here using 1000 for an example
+probs0 <- rep(1/k, k)
 alpha <- 0.05 # critical alpha
 
-p <- rep(NA, nsim) # preallocation for speed
-dat <- data.frame(group = rep(c("a", "b"), each = n))
+p <- rep(NA, nsim) # preallocation to improve loop computational efficiency
+dat <- data.frame(group = rep(c("a", "b"), each = n)) # data frame
+
+head_tail(dat, n = 3)
 
 for(i in 1:nsim){
-  sim <- sim_ord_latent(~group, By = d, prob = bprobs, link = "probit", data = dat)
+  sim <- sim_ord_latent(~group, By = d, prob = probs0, link = "probit", data = dat)
   fit <- clm(y ~ group, data = sim, link = "probit")
@@
 }
-```
-
-Then to compute the power we can simply compute the proportion of p values lower than the critical level.
-This can be done in R as `mean(p <= alpha)` resulting in an estimated power of `r mean(p <= alpha)`.
-Clearly using only a single condition is not really informative thus we can vary the sample size across plausible values and check the corresponding power curve.
-
-```{r}
+
+# estimate the power
+mean(p <= alpha, na.rm = TRUE)
+```
+
+Despite useful, calculating power curves instead a single value is more informative.
+For this reason we repeat the previous simulation but for different sample sizes.
+Figure \@ref(fig:fig-power-curve) depict the power curve resulting from the simulation.
+
+```{r power-curve-example}
 #| cache: true
-n <- c(10, 40, 60, 100, 200)
+#| echo: true
+#| collapse: true
+
+n <- c(20, 40, 60, 100, 200)
 power <- rep(NA, length(n))
@@
 for(i in 1:length(n)){
-  p <- rep(NA, 0) # preallocation for speed
+  p <- rep(NA, nsim) # preallocation for speed
   dat <- data.frame(group = rep(c("a", "b"), each = n[i]))
   for(j in 1:nsim){
-    sim <- sim_ord_latent(~group, By = d, prob = bprobs, link = "probit", data = dat)
+    sim <- sim_ord_latent(~group, By = d, prob = probs0, link = "probit", data = dat)
     fit <- clm(y ~ group, data = sim, link = "probit")
-    p[j] <- summary(fit)$coefficients["groupb", "Pr(>|z|)"] # extract the pvalue
+    p[j] <- summary(fit)$coefficients["groupb", "Pr(>|z|)"]
   }
@@
 }
-```
-
-```{r}
+
+power
+```
+
+```{r fig-power-curve, fig.cap=filor::get_caption()}
 power_df <- data.frame(power = power, n = n)
@@
 ggplot(power_df, aes(x = n, y = power)) +
-  geom_point(size = 3) +
-  geom_line() +
+  geom_point(size = 4) +
+  geom_line(linewidth = 1) +
   ylim(c(0, 1)) +
@@
 
+<!-- FIGURES -->
+
 (ref:fig-ordinal-models) Cumulative model. Adapted from @Burkner2019-aw.
@@
 
-(ref:tab-prop-ord-ex) caption here
-
 (ref:fig-logit-vs-probit) caption here
 
+(ref:fig-example-cat-latent) `cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")`
+
+(ref:fig-example-num-latent) `num_latent_plot(x = runif(100), b1 = 2, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit")`
+
+(ref:fig-explain-cumulative) caption here
+
+(ref:fig-odds-example) caption here
+
+(ref:fig-effects-2-by-2-interaction) `plot(ggpredict(fit, terms = c("x1", "x2")))`
+
+(ref:fig-effects-num-by-cat-interaction) `plot(ggpredict(fit, terms = c("x[all]", "g")))`
+
+(ref:fig-power-curve) caption
+
+<!-- TABLES -->
+
 (ref:tab-model-summary) caption here
 
-(ref:fig-example-cat-latent) caption here
-
-(ref:fig-example-num-latent) caption here
-
-(ref:fig-explain-cumulative) caption here
+(ref:tab-dataset-example) caption here

### 2024-01-04 paper/paper-new.tex
update

@@
 
+\scriptsize
+
+\normalsize
+
 \section{Introduction}\label{introduction}
@@
 We can name this general class of models as \emph{ordinal regression}.
-The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures (Gerhard Tutz, 2022).
-Gerhard Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
-We can identify three main classes: \emph{cumulative models} {[}CM; McCullagh (1980); Agresti (2010){]}, \emph{sequential models} (Gerhard Tutz, 1990) and \emph{adjacent category models}.
+The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures (Tutz, 2022).
+Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
+We can identify three main classes: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990) and \emph{adjacent category models}.
 The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.
@@
 For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential model.
-When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Gerhard Tutz \& Berger, 2020).
+When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Tutz \& Berger, 2020).
 The adjacent category model compare the category \(k\) with \(k + 1\) still assuming a latent distribution for each \(k\).
-As suggested by Gerhard Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
+As suggested by Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
 Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.
@@
 Furthermore, CM are also used to model data under a signal detection theory framework (e.g., DeCarlo, 2010).
-The Figure \ref{fig:fig-ordinal-models} depict the overall structure of the \emph{cumulative} model.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{img/fig-ordinal-models}
-
-}
-
-\caption{Cumulative model. Adapted from Bürkner and Vuorre (2019).}\label{fig:fig-ordinal-models}
-\end{figure}
-
-\normalsize
+The Figure \ref{fig:fig-explain-cumulative} depict the overall structure of the \emph{cumulative} model.
 
@@
 The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g()\).
-To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:eq:prob-cum-model2}.
+To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:prob-cum-model2}.
 
@@
 For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution.
-Following the notation by Gerhard Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).
+Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).
 
 In the basic version of the model, the thresholds \(\alpha_k\) are considered as fixed and being part of the measurement procedure (Liddell \& Kruschke, 2018) and do not vary as a function of the predictors.
-In a more sophisticated version of the model called location-shift (Gerhard Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha_k\) can vary as a function of the predictors.
-Gerhard Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Gerhard Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.
+In a more sophisticated version of the model called location-shift (Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha_k\) can vary as a function of the predictors.
+Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.
 
@@
 
-\begin{verbatim}
-##   group mean median   sd                  Y1                  Y2
-## a     a 2.46      2 1.20 **15** (*p = 0.30*) **11** (*p = 0.22*)
-## b     b 3.26      4 0.94  **2** (*p = 0.04*) **11** (*p = 0.22*)
-##                    Y3                  Y4
-## a **10** (*p = 0.20*) **14** (*p = 0.28*)
-## b  **9** (*p = 0.18*) **28** (*p = 0.56*)
-\end{verbatim}
-
-\normalsize
-
-\scriptsize
-
-\normalsize
-
-\scriptsize
-
-\begin{verbatim}
-## formula: y ~ x
-## data:    dat
-## 
-##  link  threshold nobs logLik  AIC    niter max.grad cond.H 
-##  logit flexible  100  -125.31 258.63 5(0)  3.87e-12 2.1e+01
-## 
-## Coefficients:
-##    Estimate Std. Error z value Pr(>|z|)    
-## xb   1.3088     0.3828   3.419 0.000629 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Threshold coefficients:
-##     Estimate Std. Error z value
-## 1|2  -1.0373     0.3074  -3.374
-## 2|3   0.1947     0.2791   0.698
-## 3|4   1.0282     0.2947   3.489
-\end{verbatim}
+\begin{table}
+
+\caption{\label{tab:tab-dataset-example}caption here}
+\centering
+\begin{tabular}[t]{llllllll}
+\toprule
+group & mean & median & sd & Y1 & Y2 & Y3 & Y4\\
+\midrule
+a & 2.42 & 2 & 1.13 & 14 (p = 0.28) & 12 (p = 0.24) & 13 (p = 0.26) & 11 (p = 0.22)\\
+b & 3 & 3 & 1.16 & 9 (p = 0.18) & 6 (p = 0.12) & 11 (p = 0.22) & 24 (p = 0.48)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\normalsize
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\FunctionTok{summary}\NormalTok{(fit)}
+\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
+\CommentTok{\#\textgreater{} data:    dat}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
+\CommentTok{\#\textgreater{}  logit flexible  100  {-}132.25 272.51 4(0)  1.08e{-}08 2.0e+01}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Coefficients:}
+\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)   }
+\CommentTok{\#\textgreater{} xb   0.9587     0.3718   2.579  0.00992 **}
+\CommentTok{\#\textgreater{} {-}{-}{-}}
+\CommentTok{\#\textgreater{} Signif.
+codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Threshold coefficients:}
+\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
+\CommentTok{\#\textgreater{} 1|2  {-}0.8103     0.2814  {-}2.879}
+\CommentTok{\#\textgreater{} 2|3   0.0672     0.2636   0.255}
+\CommentTok{\#\textgreater{} 3|4   1.1158     0.2900   3.848}
+\end{Highlighting}
+\end{Shaded}
 
@@
 Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect.
-As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding {[}for an overview about contrasts coding schemes see Schad, Vasishth, Hohenstein, and Kliegl (2020){]}\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
+As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding {[}see Schad, Vasishth, Hohenstein, and Kliegl (2020){]}\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
 By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}
@@
 
-\caption{(ref:fig-odds-example)}\label{fig:fig-odds-example}
+\caption{caption here}\label{fig:fig-odds-example}
 \end{figure}
@@
 \NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\DocumentationTok{\#\# [1] 0.3}
+\CommentTok{\#\textgreater{} [1] 0.28}
 
@@
 \FunctionTok{odds}\NormalTok{(p1a)}
-\DocumentationTok{\#\# [1] 0.4285714}
+\CommentTok{\#\textgreater{} [1] 0.3888889}
 
@@
 \NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\DocumentationTok{\#\# [1] 0.04}
+\CommentTok{\#\textgreater{} [1] 0.18}
 \FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
-\DocumentationTok{\#\# [1] 10.28571}
+\CommentTok{\#\textgreater{} [1] 1.771605}
 \end{Highlighting}
@@
 
-For example the probability of responding \(Y = 1\) in the group ``a'' is 0.30 corresponding to an odds of 0.43.
-When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 10.29 times the odds of responding \(Y = 1\) compared to the group ``b''.
+For example the probability of responding \(Y = 1\) in the group ``a'' is 0.28 corresponding to an odds of 0.39.
+When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 1.77 times the odds of responding \(Y = 1\) compared to the group ``b''.
 
@@
 Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\).
-The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 3.70.
+The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 2.61.
 In fact, the logit model essentially estimate the cumulative odds ratio.
@@
 \NormalTok{(group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
-\DocumentationTok{\#\# y1vs234 y12vs34 y123vs4 }
-\DocumentationTok{\#\#    0.30    0.52    0.72}
+\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
+\CommentTok{\#\textgreater{}    0.28    0.52    0.78}
 \NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
-\DocumentationTok{\#\# y1vs234 y12vs34 y123vs4 }
-\DocumentationTok{\#\#    0.04    0.26    0.44}
+\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
+\CommentTok{\#\textgreater{}    0.18    0.30    0.52}
 
@@
 \FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b)}
-\DocumentationTok{\#\#   y1vs234   y12vs34   y123vs4 }
-\DocumentationTok{\#\# 10.285714  3.083333  3.272727}
+\CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
+\CommentTok{\#\textgreater{} 1.771605 2.527778 3.272727}
 \end{Highlighting}
@@
 This is not a coincidence but an implicit (until now) assumption made by the CM called \emph{proportional odds} (POA).
-Following again the taxonomy by Gerhard Tutz (2022), each of the presented ordinal regression model has a basic version making the (POA).
+Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the (POA).
 There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}).
@@
 Given the relatively strong assumption, there are several methods for testing if data are supporting the POA (see Liu, He, Tu, \& Tang, 2023 for an overview).
-Gerhard Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the POA in terms of implementing \emph{location-shift} or \emph{location-scale} models.
+Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the POA in terms of implementing \emph{location-shift} or \emph{location-scale} models.
 Basically these methods should guarantee more flexibility in modelling the observed probabilities reducing the number of parameters.
@@
 \CommentTok{\# extracting the predicted probabilities for the two groups}
-\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(x)))}\SpecialCharTok{$}\NormalTok{fit}
+\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x)))}\SpecialCharTok{$}\NormalTok{fit}
 
@@
 \FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
-\DocumentationTok{\#\# [1] 3.701591}
+\CommentTok{\#\textgreater{} [1] 2.608304}
 \FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
-\DocumentationTok{\#\# [1] 3.701591}
+\CommentTok{\#\textgreater{} [1] 2.608304}
 \end{Highlighting}
@@
 This can be directly demonstrated in the same way as for the logit model.
-We use the same dataset presented in Table \ref{tab:tab-prop-ord-ex} but fitting a probit model.
+We use the same dataset of the previous simulation.
 The difference between the \(z\) scores comparing the two groups calculated on the cumulative probability is the same regardless the \(k - 1\) level.
 
-\[
+\begin{equation}
 z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
-\]
+\label{eq:probit-prop-odds}
+\end{equation}
 
@@
 \FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
-\DocumentationTok{\#\# [1] 0.8261077}
+\CommentTok{\#\textgreater{} [1] 0.5644823}
 \FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
-\DocumentationTok{\#\# [1] 0.8261077}
+\CommentTok{\#\textgreater{} [1] 0.5644823}
 \end{Highlighting}
@@
 For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in \(x\).
-For the probit model, the \(\beta\) is the increase in terms of \(z\) scores for a unit increase in \(x\).
-This is very convenient especially for categorical variables because parameters can be intepreted as a Cohen's \(d\) like measure.
+
+For the \emph{probit} model, the \(\beta\) is the increase in terms of \(z\) scores for a unit increase in \(x\).
+This is very convenient especially for categorical variables because parameters can be interpreted as a Cohen's \(d\) like measure.
 Thinking about the latent distributions, the \(\beta\) is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric \(x\).
-The intepretation in terms of shifting the latent mean holds also for the logistic model.
+More formally the shift in the latent distribution is \(\frac{\beta}{\sigma}\) (for the \emph{probit} model \(\sigma = 1\)).
+The interpretation in terms of shifting the latent mean holds also for the logistic model.
 However, the standard deviation of the standard logistic regression is \(\frac{\pi}{\sqrt{3}} \approx 1.81\)\footnote{Actually the variance of the logistic distribution is \(\frac{s^2\pi^2}{3}\) and the standard deviation \(\frac{s\pi}{\sqrt{3}}\) where \(s\) is the scale of the distribution.
-For the standard logistic distribution \(s = 1\) (as for the standard normal distribution).
-For the standard normal distribution \(s^2 = \sigma^2 = 1\)}.
-The \(\beta\) for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by \(\frac{\pi}{\sqrt{3}}\) standard deviations.
-
-Another advantage of the probit model is the possibility to directly map parameters from signal detection theory (Green, Swets, \& Others, 1966; Stanislaw \& Todorov, 1999) into the model coefficients.
-In practical terms, the thresholds \(\alpha\) are the criterion cut-offs and the \(\beta_1\) is the d\('\) parameter (DeCarlo, 1998; Knoblauch \& Maloney, 2012).
-
-qualcosa in più qui
+For the standard logistic distribution \(s = 1\) (as for the standard normal distribution).}.
+The \(\beta\) for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by \(\beta/(\frac{\pi}{\sqrt{3}})\) standard deviations.
 
@@
 
-There are mainly two ways to simulate data.
-The first method concerns simulating starting from the latent formulation of the ordinal model.
-Basically we can simulate the underlying latent distribution and then fixing the thresholds converting the latent continuous variable into the observed ordinal variable.
+In this tutorial we present two method for simulating ordinal data.
+The first method concerns calculating the probabilities of each \(k\) of \(Y\) as a function of predictors and then sampling from a multinomial distribution.
+The second method simulate the latent variable \(Y^\star\) as a function of predictors.
 
@@
 
-The first method to simulate ordinal data as a function of predictors is by calculating the true probabilities \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution\footnote{In R we are using the \texttt{sample()} function that simulate values from a \emph{categorical} distribution.
+For the first method we need to calculate \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution\footnote{In R we are using the \texttt{sample()} function that simulate values from a \emph{categorical} distribution.
 The \emph{categorical} distribution is considered a special case of the multinomial distribution}.
-This is similar to the general method to simulate data for a generalized linear model where the linear predictors \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component.
-The following code box simulate a \(n\) binary trials with a continuous predictor \(x\).
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-6-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-6}
-\end{figure}
-
-\normalsize
-
-Essentially the \texttt{lp} (\(\eta_i\)) vector contains true probability of success (\(p_i|x_i = g^{-1}(\eta_i)\)) for the \(x_i\) level of the predictor.
-Then to introduce the random part of the model, we can use \(p_i\) to generate binary values from a Binomial (in this specific case a Bernoulli) distribution.
-This create \texttt{y} that is the vector of 0-1 values generated using the true probabilities of success.
-The supplementary materials contains a more extensive example for simulating data for generalized linear models.
-
-We can apply the same idea to an ordinal outcome but we need \(k - 1\) equations (compared to the previous example) where \(k\) is the number of ordinal levels.
-Let's simulate a similar design with a continuous \(x\) predictor and \(k = 4\) options.
-We fix the baseline probabilities where \(x = 0\) as uniform thus \(p(y_1) = p(y_2) = ...
-p(y_k) = 1/k\).
-The proposed workflow for the simulation is:
+This is similar to the general method to simulate data for a generalized linear model where the linear predictor \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component.
+
+As a simple example, let's simulate two groups or condition with \(n = 100\) participants responding to an item with \(k = 4\) ordered options.
+We simulate that the second group has more probability of responding higher categories of \(Y\).
+As explained in the previous sections, we can summarise the effect size of a CM (assuming POA) using a single \(\beta = \log(OR)\).
+We can assume that the odds ratio is \(OR = 2\).
+For simplicity, the probabilities of the first group \(x = 0\) are uniform thus \(P(Y = 1|x_0) = ...
+P(Y = k|x_0) = 1/k\).
+The following code summarise the first steps of the simulation.
+Basically we define the simulation parameters, we calculate the \(k - 1\) thresholds \(\alpha\) and we apply Equations \eqref{eq:prob-cum-model1} and \eqref{eq:prob-cum-model2}.
+In this way we calculated the true probability of each \(Y\) level for the two groups.
+
+An important step is converting from \(\alpha\) to probability and the opposite.
+This steps are implemented in the \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} functions.
+Basically given the input and the link function, these functions return the thresholds associated with \(k\) probabilities or the \(k - 1\) thresholds.
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{5}
+\NormalTok{(p }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k)) }\CommentTok{\# uniform}
+\CommentTok{\#\textgreater{} [1] 0.2 0.2 0.2 0.2 0.2}
+\FunctionTok{names}\NormalTok{(p) }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"y"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{k)}
+
+\NormalTok{(alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(p, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)) }\CommentTok{\# or prob\_to\_alpha(p, "probit")}
+\CommentTok{\#\textgreater{}         y1         y2         y3         y4 }
+\CommentTok{\#\textgreater{} {-}1.3862944 {-}0.4054651  0.4054651  1.3862944}
+\FunctionTok{alpha\_to\_prob}\NormalTok{(alpha, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\CommentTok{\#\textgreater{} [1] 0.2 0.2 0.2 0.2 0.2}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# SIMULATION PARAMETERS}
+
+\NormalTok{N }\OtherTok{\textless{}{-}} \DecValTok{100} \CommentTok{\# sample size}
+\NormalTok{or }\OtherTok{\textless{}{-}} \DecValTok{2} \CommentTok{\# odds ratio}
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{4} \CommentTok{\# number of ordinal alternatives}
+\NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k) }\CommentTok{\# probabilities for the first group}
+\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs0, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{each =}\NormalTok{ N}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+
+\DocumentationTok{\#\# LINEAR PREDICTOR}
+
+\CommentTok{\# calculate linear predictor using equation }\AlertTok{TODO}\CommentTok{ put equation}
+\CommentTok{\# obtaining k {-} 1 equations}
+
+\NormalTok{lp }\OtherTok{\textless{}{-}} \FunctionTok{lapply}\NormalTok{(alpha, }\ControlFlowTok{function}\NormalTok{(a) a }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(or) }\SpecialCharTok{*}\NormalTok{ dat}\SpecialCharTok{$}\NormalTok{x)}
+\FunctionTok{names}\NormalTok{(lp) }\OtherTok{\textless{}{-}} \FunctionTok{sprintf}\NormalTok{(}\StringTok{"lp\_leq\%s"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{(k }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{)) }\CommentTok{\# giving appropriate names}
+\NormalTok{lp }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(lp)}
+\FunctionTok{head}\NormalTok{(lp)}
+\CommentTok{\#\textgreater{}     lp\_leq1 lp\_leq2  lp\_leq3}
+\CommentTok{\#\textgreater{} 1 {-}1.098612       0 1.098612}
+\CommentTok{\#\textgreater{} 2 {-}1.098612       0 1.098612}
+\CommentTok{\#\textgreater{} 3 {-}1.098612       0 1.098612}
+\CommentTok{\#\textgreater{} 4 {-}1.098612       0 1.098612}
+\CommentTok{\#\textgreater{} 5 {-}1.098612       0 1.098612}
+\CommentTok{\#\textgreater{} 6 {-}1.098612       0 1.098612}
+
+\DocumentationTok{\#\# CUMULATIVE PROBABILITIES}
+
+\CommentTok{\# apply the inverse of the link function (invlogit) to calculate cumulative probabilities}
+\NormalTok{cump }\OtherTok{\textless{}{-}} \FunctionTok{lapply}\NormalTok{(lp, plogis)}
+\NormalTok{cump }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(cump)}
+\FunctionTok{names}\NormalTok{(cump) }\OtherTok{\textless{}{-}} \FunctionTok{sprintf}\NormalTok{(}\StringTok{"cump\_leq\%s"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{(k }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{)) }\CommentTok{\# giving appropriate names}
+\FunctionTok{head}\NormalTok{(cump)}
+\CommentTok{\#\textgreater{}   cump\_leq1 cump\_leq2 cump\_leq3}
+\CommentTok{\#\textgreater{} 1      0.25       0.5      0.75}
+\CommentTok{\#\textgreater{} 2      0.25       0.5      0.75}
+\CommentTok{\#\textgreater{} 3      0.25       0.5      0.75}
+\CommentTok{\#\textgreater{} 4      0.25       0.5      0.75}
+\CommentTok{\#\textgreater{} 5      0.25       0.5      0.75}
+\CommentTok{\#\textgreater{} 6      0.25       0.5      0.75}
+
+\DocumentationTok{\#\# PROBABILITIES OF Y}
+
+\CommentTok{\# for each row, we can calculate P(Y = k) using equation }\AlertTok{TODO}\CommentTok{ put equation}
+\CommentTok{\# P(Y = 1) = P(Y \textless{}= 1)}
+\CommentTok{\# P(Y = 2) = P(Y \textless{}= 2) {-} P(Y \textless{}= 1)}
+\CommentTok{\# P(Y = 3) = P(Y \textless{}= 3) {-} P(Y \textless{}= 2)}
+\CommentTok{\# P(Y = 4) = 1 {-} P(Y \textless{}= 3)}
+
+\CommentTok{\# adding a columns of 0 and 1, then diff() for adjacent differences}
+\NormalTok{cump }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(}\DecValTok{0}\NormalTok{, cump, }\DecValTok{1}\NormalTok{)}
+\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cump, }\DecValTok{1}\NormalTok{, diff, }\AttributeTok{simplify =} \ConstantTok{FALSE}\NormalTok{)}
+
+\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\FunctionTok{do.call}\NormalTok{(rbind, p)) }\CommentTok{\# collapse list of rows into a dataframe}
+\FunctionTok{names}\NormalTok{(p) }\OtherTok{\textless{}{-}} \FunctionTok{sprintf}\NormalTok{(}\StringTok{"p\%s"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{k) }\CommentTok{\# giving appropriate names}
+\FunctionTok{head}\NormalTok{(p)}
+\CommentTok{\#\textgreater{}     p1   p2   p3   p4}
+\CommentTok{\#\textgreater{} 1 0.25 0.25 0.25 0.25}
+\CommentTok{\#\textgreater{} 2 0.25 0.25 0.25 0.25}
+\CommentTok{\#\textgreater{} 3 0.25 0.25 0.25 0.25}
+\CommentTok{\#\textgreater{} 4 0.25 0.25 0.25 0.25}
+\CommentTok{\#\textgreater{} 5 0.25 0.25 0.25 0.25}
+\CommentTok{\#\textgreater{} 6 0.25 0.25 0.25 0.25}
+
+\CommentTok{\# probabilities for id = 1 (x = 0) and id = 51 (x = 1)}
+\NormalTok{p[}\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{51}\NormalTok{), ]}
+\CommentTok{\#\textgreater{}           p1        p2        p3   p4}
+\CommentTok{\#\textgreater{} 1  0.2500000 0.2500000 0.2500000 0.25}
+\CommentTok{\#\textgreater{} 51 0.1428571 0.1904762 0.2666667 0.40}
+
+\DocumentationTok{\#\# CUMULATIVE ODDS RATIO}
+
+\CommentTok{\# calculate the (cumulative) odds ratio}
+
+\NormalTok{x0 }\OtherTok{\textless{}{-}}\NormalTok{ cump[}\DecValTok{1}\NormalTok{, }\DecValTok{2}\SpecialCharTok{:}\NormalTok{k]}
+\NormalTok{x1 }\OtherTok{\textless{}{-}}\NormalTok{ cump[}\DecValTok{51}\NormalTok{, }\DecValTok{2}\SpecialCharTok{:}\NormalTok{k]}
+
+\CommentTok{\# this is the same as the or (the b1) and we are assuming POA}
+\FunctionTok{odds}\NormalTok{(x0) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(x1)}
+\CommentTok{\#\textgreater{}   cump\_leq1 cump\_leq2 cump\_leq3}
+\CommentTok{\#\textgreater{} 1         2         2         2}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+
+\DocumentationTok{\#\# SAMPLING FROM THE MULTINOMIAL DISTRIBUTION}
+
+\CommentTok{\# example of a random Y outcome based on the probabilities}
+\FunctionTok{sample}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ p[}\DecValTok{1}\NormalTok{, ])}
+\CommentTok{\#\textgreater{} [1] 3}
+
+\CommentTok{\# we can apply it to the full dataset, this step lead to different results}
+\CommentTok{\# each time we run the code because we are sampling from a distribution}
+
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(p, }\DecValTok{1}\NormalTok{, }\ControlFlowTok{function}\NormalTok{(ps) }\FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ ps))}
+\FunctionTok{head}\NormalTok{(dat)}
+\CommentTok{\#\textgreater{}   x y}
+\CommentTok{\#\textgreater{} 1 0 2}
+\CommentTok{\#\textgreater{} 2 0 4}
+\CommentTok{\#\textgreater{} 3 0 2}
+\CommentTok{\#\textgreater{} 4 0 4}
+\CommentTok{\#\textgreater{} 5 0 2}
+\CommentTok{\#\textgreater{} 6 0 1}
+
+\CommentTok{\# let\textquotesingle{}s compute the observed probabilities, to be compared to the true}
+\CommentTok{\# probabilities}
+
+\CommentTok{\# observed}
+\NormalTok{(op }\OtherTok{\textless{}{-}} \FunctionTok{prop.table}\NormalTok{(}\FunctionTok{table}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x, dat}\SpecialCharTok{$}\NormalTok{y), }\AttributeTok{margin =} \DecValTok{1}\NormalTok{))}
+\CommentTok{\#\textgreater{}    }
+\CommentTok{\#\textgreater{}        1    2    3    4}
+\CommentTok{\#\textgreater{}   0 0.42 0.22 0.20 0.16}
+\CommentTok{\#\textgreater{}   1 0.14 0.08 0.34 0.44}
+
+\CommentTok{\# true (just selecting a row from x = 0 and x = 1)}
+\NormalTok{p[}\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{51}\NormalTok{), ]}
+\CommentTok{\#\textgreater{}           p1        p2        p3   p4}
+\CommentTok{\#\textgreater{} 1  0.2500000 0.2500000 0.2500000 0.25}
+\CommentTok{\#\textgreater{} 51 0.1428571 0.1904762 0.2666667 0.40}
+
+\CommentTok{\# similarly we can compute the observed cumulative odds ratios}
+
+\NormalTok{(cum\_op }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(op, }\DecValTok{1}\NormalTok{, cumsum))}
+\CommentTok{\#\textgreater{}    }
+\CommentTok{\#\textgreater{}        0    1}
+\CommentTok{\#\textgreater{}   1 0.42 0.14}
+\CommentTok{\#\textgreater{}   2 0.64 0.22}
+\CommentTok{\#\textgreater{}   3 0.84 0.56}
+\CommentTok{\#\textgreater{}   4 1.00 1.00}
+\FunctionTok{odds}\NormalTok{(cum\_op[, }\DecValTok{1}\NormalTok{]) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(cum\_op[, }\DecValTok{2}\NormalTok{])}
+\CommentTok{\#\textgreater{}        1        2        3        4 }
+\CommentTok{\#\textgreater{} 4.448276 6.303030 4.125000      NaN}
+
+\CommentTok{\# the odds ratios are not the same as the parameter. as we increase N}
+\CommentTok{\# the parameter will converge to the true value}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{ordered}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# make an ordered factor in R where 1 \textless{} 2 \textless{} 3 \textless{} 4}
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+\FunctionTok{summary}\NormalTok{(fit)}
+\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
+\CommentTok{\#\textgreater{} data:    dat}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}126.98 261.96 9(0)  1.80e{-}11 2.1e+01}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Coefficients:}
+\CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
+\CommentTok{\#\textgreater{} x   0.9373     0.2275    4.12 3.78e{-}05 ***}
+\CommentTok{\#\textgreater{} {-}{-}{-}}
+\CommentTok{\#\textgreater{} Signif.
+codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Threshold coefficients:}
+\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
+\CommentTok{\#\textgreater{} 1|2  {-}0.1879     0.1683  {-}1.117}
+\CommentTok{\#\textgreater{} 2|3   0.2706     0.1697   1.594}
+\CommentTok{\#\textgreater{} 3|4   1.0595     0.1893   5.596}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+We can generalize the previous workflow as:
 
@@
 \item
-  Define \(n\) (number of observations) and \(k\) (number of ordinal outcomes)
-\item
-  Define the regression coefficients \(\beta\)s.
-\item
-  Define the baseline probabilities when \(x = 0\).
-These probabilities can be converted to the corresponding thresholds \(\alpha\) choosing the appropriate link function (\emph{logit} or \emph{probit}).
+  Define simulation parameters, baseline probabilities, etc. as in the previous workflow
 \item
@@
   Calculate for each observation the probability of the \(k\) outcome as the difference between cumulative probabilities.
-For example, the probability of \(P(y = 2) = P(y \leq 2) - P(y \leq 1)\).
-This is implemented in R by adding a columns of zeros and ones respectively at the beginning and end of the matrix of cumulative probabilities because the probability of the first \(Y = 1\) level is the cumulative probability from \(- \infty\) to the first threshold and the probability of \(Y = k\) is obtained removing \(P(Y \leq k - 1)\) from one.
 \item
   Sample \(n\) outcomes from a multinomial distribution, choosing between \(k\) alternatives with the calculated probabilities.
+\item
+  Fit the appropriate model using \texttt{ordinal::clm()}
 \end{enumerate}
 
-\scriptsize
-
-\normalsize
-
-The linear predictors (\(\eta\)) can be seen in the Figure ref. The \(g\) in this case is the cumulative \emph{probit} function \(\Phi\).
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-8-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-8}
-\end{figure}
-
-\normalsize
-
-Now we can apply the inverse of the link function \(g^{-1} = \Phi^{-1}\) to calculate the corresponding cumulative probabilities.
-
-\scriptsize
-
-\normalsize
-
-\scriptsize
-
-\begin{verbatim}
-##    p_y1  p_y2  p_y3  p_y4
-## 1 0.409 0.263 0.197 0.131
-## 2 0.389 0.264 0.204 0.143
-## 3 0.429 0.261 0.189 0.121
-## 4 0.119 0.188 0.261 0.433
-## 5   ...   ...   ...   ...
-## 6 0.147 0.207 0.264 0.383
-## 7 0.328 0.263 0.226 0.183
-## 8 0.389 0.264 0.204 0.143
-## 9 0.488 0.252 0.166 0.094
-\end{verbatim}
-
-\normalsize
-
-We can plot the expected effect of \(\beta_1\) on the \(k\) probabilities using the \texttt{num\_latent\_plot()} function (see Figure \ref{fig:sim-with-probs-lat-plot}).
-Finally we sample using the \texttt{sample()} function using the calculated probabilities.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/sim-with-probs-lat-plot-1}
-
-}
-
-\caption{sim-with-probs-lat-plot}\label{fig:sim-with-probs-lat-plot}
-\end{figure}
-
-\normalsize
-
-\scriptsize
-
-\normalsize
-
-To check the simulation result we can increase the number of observations, fit the model using the \texttt{ordinal::clm()} function and assess the recovery of simulated parameters.
-
-\scriptsize
-
-\normalsize
-
-\scriptsize
-
-\normalsize
+A general suggestion is to increase the sample size \(n\) to a very large value reducing the random variability to check if simulated parameters are recovered.
 
@@
 
-\normalsize
-
-The Figure (put crossref) depict the simulated \(Y^{*}\) and the corresponding ordinal value.
-As for the previous simulation we can fit the model using \texttt{clm()} and check the estimated parameters.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-14-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-14}
-\end{figure}
-
-\normalsize
-
-\scriptsize
-
-\normalsize
-
-\scriptsize
-
-\normalsize
-
-The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function.
-
-\scriptsize
-
 \begin{Shaded}
 \begin{Highlighting}[]
-
+\DocumentationTok{\#\# SIMULATION PARAMETERS}
+
+\NormalTok{N }\OtherTok{\textless{}{-}} \FloatTok{1e3} \CommentTok{\# sample size}
+\NormalTok{or }\OtherTok{\textless{}{-}} \DecValTok{4} \CommentTok{\# odds ratio, higher here just for a more clear plot}
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{4} \CommentTok{\# number of ordinal alternatives}
+\NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k) }\CommentTok{\# probabilities for the first group}
+\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs0, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{each =}\NormalTok{ N}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+
+\DocumentationTok{\#\# LINEAR PREDICTOR}
+
+\CommentTok{\# calculate the linear predictor using equation }\AlertTok{TODO}\CommentTok{ put equation}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{lp }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(or) }\SpecialCharTok{*}\NormalTok{ dat}\SpecialCharTok{$}\NormalTok{x}
+
+\CommentTok{\# add the random part by sampling errors from a standard logistic (or normal) distribution}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{ystar }\OtherTok{\textless{}{-}}\NormalTok{ dat}\SpecialCharTok{$}\NormalTok{lp }\SpecialCharTok{+} \FunctionTok{rlogis}\NormalTok{(N, }\AttributeTok{location =} \DecValTok{0}\NormalTok{, }\AttributeTok{scale =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{\# cut the latent distribution. The + 1 because the first category is 0 by default.}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{findInterval}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{ystar, alpha) }\SpecialCharTok{+} \DecValTok{1}
+
+\FunctionTok{head\_tail}\NormalTok{(dat, }\AttributeTok{n =} \DecValTok{3}\NormalTok{)}
+\CommentTok{\#\textgreater{}      x       lp       ystar y}
+\CommentTok{\#\textgreater{} 1    0 0.000000 {-}0.63265006 2}
+\CommentTok{\#\textgreater{} 2    0 0.000000  0.96340682 3}
+\CommentTok{\#\textgreater{} 3    0 0.000000  0.95041920 3}
+\CommentTok{\#\textgreater{} 998  1 1.386294  0.07772189 3}
+\CommentTok{\#\textgreater{} 999  1 1.386294  1.09468340 3}
+\CommentTok{\#\textgreater{} 1000 1 1.386294  1.91057532 4}
 \end{Highlighting}
@@
 
+The Figure \ref{fig:fig-sim-from-latent} depict the simulated \(Y^{*}\) and the corresponding ordinal value.
+As for the previous simulation we can fit the model using \texttt{clm()} and check the estimated parameters.
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-sim-from-latent-1}
+
+}
+
+\caption{ }\label{fig:fig-sim-from-latent}
+\end{figure}
+
+\normalsize
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{ordered}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# make an ordered factor in R where 1 \textless{} 2 \textless{} 3 \textless{} 4}
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\FunctionTok{summary}\NormalTok{(fit)}
+\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
+\CommentTok{\#\textgreater{} data:    dat}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{}  link  threshold nobs logLik   AIC     niter max.grad cond.H }
+\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1284.98 2577.96 4(0)  1.51e{-}08 1.8e+01}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Coefficients:}
+\CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
+\CommentTok{\#\textgreater{} x   1.1905     0.1201   9.914   \textless{}2e{-}16 ***}
+\CommentTok{\#\textgreater{} {-}{-}{-}}
+\CommentTok{\#\textgreater{} Signif.
+codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Threshold coefficients:}
+\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
+\CommentTok{\#\textgreater{} 1|2 {-}1.02783    0.09358 {-}10.984}
+\CommentTok{\#\textgreater{} 2|3  0.02280    0.08403   0.271}
+\CommentTok{\#\textgreater{} 3|4  1.06892    0.09184  11.639}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function. Details about the function can be found on the Github repository LINK.
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{N }\OtherTok{\textless{}{-}} \DecValTok{100} \CommentTok{\# sample size}
+\NormalTok{or }\OtherTok{\textless{}{-}} \DecValTok{4} \CommentTok{\# odds ratio, higher here just for a more clear plot}
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{4} \CommentTok{\# number of ordinal alternatives}
+\NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k) }\CommentTok{\# probabilities for the first group}
+\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs0, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{each =}\NormalTok{ N}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+
+\CommentTok{\# same as the previous simulation}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \FunctionTok{log}\NormalTok{(or), }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+
+\FunctionTok{head\_tail}\NormalTok{(dat, }\AttributeTok{n =} \DecValTok{3}\NormalTok{)}
+\CommentTok{\#\textgreater{}     x y         ys}
+\CommentTok{\#\textgreater{} 1   0 1 {-}1.1479983}
+\CommentTok{\#\textgreater{} 2   0 2 {-}0.7163737}
+\CommentTok{\#\textgreater{} 3   0 4  1.3481517}
+\CommentTok{\#\textgreater{} 98  1 3  0.1655840}
+\CommentTok{\#\textgreater{} 99  1 4  1.2106967}
+\CommentTok{\#\textgreater{} 100 1 3  0.6463900}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
 \subsubsection{Choosing parameters values}\label{choosing-parameters-values}
@@
 
-The two proposed simulation workflows can be really flexible for simulating complex models with multiple predictors and interactions.
-The critical part of the simulation is setting appropriate and empirically meaningful parameters values.
-The thresholds are usually not the crucial part of the statistical inference and model interpretations.
-In fact, they can be considered as the intercept in standard linear models.
-In any case to set meaningful \(\alpha\) values the best strategy is to convert them into the corresponding probabilities according to the link function.
-The thresholds are essentially quantiles of the latent distribution (logistic or normal) without a direct empirical interpretation.
-The distance between the thresholds determine the probabililty assigned to a specific \(Y\) value.
-The function \texttt{show\_alpha()} allow to produce a meaningful plot starting from the thresholds or the probabilities.
-For example, the Figure \ref{fig:fig-show-th-example} depict the threshold/probabilities associated with a four-level ordinal variable.
-
-\scriptsize
-
-\normalsize
+The previous simulation can be easily extended by adding predictors and their interactions.
+The crucial part is setting appropriate and empirically meaningful parameters.
+Thresholds \(\alpha\) are usually not of main interest (but see the location-shift models Tutz, 2022) and can be considered as intercepts in standard linear regression.
+The thresholds are quantiles of the latent distribution that produced a certain probability distribution of the ordinal variable.
+To set meaningful \(\alpha\) values we can convert probabilities into thresholds (\texttt{alpha\_to\_prob()}).
+The function \texttt{show\_alpha()} create a meaningful visual representation of using a specific set of thresholds (see Figure \ref{fig:fig-show-th-example}).
+The function can be used as \texttt{show\_alpha(prob\ =\ rep(1/4,\ 4),\ link\ =\ "probit",\ plot\ =\ c("latent",\ "probs"))}.
+In regression terms, the thresholds determine the probabilities when \(x = 0\) (as the intercept).
+Thus with two groups for example, the thresholds are the \(k\) probabilities of the ordinal variable \(Y\) for the reference group.
 
@@
 
-\caption{ }\label{fig:fig-show-th-example}
+\caption{ }\label{fig:fig-show-th-example-1}
 \end{figure}
-
-\normalsize
-
-Thus the thresholds can be fixed according to the desired probabilities distribution.
-For example, when simulating the difference between two groups on an ordinal variable, the threshold can be set thinking about the response distribution in the reference group.
-The \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} functions can be used to convert between probabilities and thresholds.
-
-\scriptsize
-
-\begin{verbatim}
-## [1] -0.2533471  0.0000000  0.8416212
-## [1] 0.02275013 0.47724987 0.34134475 0.15865525
-\end{verbatim}
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-show-th-example-2}
+
+}
+
+\caption{ }\label{fig:fig-show-th-example-2}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-show-th-example-3}
+
+}
+
+\caption{ }\label{fig:fig-show-th-example-3}
+\end{figure}
 
@@
 
-As reported in the sections above, the regression coefficients have a different interpretation according to the link function.
-Fortunately, both parameters have an intuitive interpretation given that they are directly expressed in a interpretable scale.
-For probit models we can think about standardized mean differences (e.g., Cohen's \(d\)) between categorical variables or standardized regression coefficient for a numerical predictor.
-For logit models we can fix the parameters thinking about the odds ratios.
+Regression coefficients are the effect size parameters.
+For \emph{probit} models we can set the \(\beta_j\) to be in standardized (i.e., Cohen's \(d\)-like) units.
+For a categorical variable as the previous example with the group, \(\beta\) is the degree of separation in standard deviation unit between the two latent distributions.
+For \emph{logit} models we can set the odds ratio.
 Meaningful odds ratios can be derived from previous literature, meta-analyses or converting to other effect sizes.
@@
 Using their approach, a Cohen's \(d\) of 0.5 usually considered a plausible medium effect size corresponds to and odds ratio of \(\approx 2.47\).
-Another more informative strategy is about plotting the predicted probabilities \(g^{-1}(\eta)\) associated with certain regression coefficients.
-The \texttt{cat\_latent\_plot()} and \texttt{num\_latent\_plot()} functions can be used to see the impact of choosing a specific regression coefficients for respectively a categorical (Figures \ref{fig:fig-example-cat-latent}) and numerical predictor (Figures \ref{fig:fig-example-num-latent}).
-
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{cat\_latent\_plot}\NormalTok{(}\AttributeTok{m =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\AttributeTok{s =} \DecValTok{1}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{4}\NormalTok{, }\DecValTok{4}\NormalTok{), }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{, }\AttributeTok{plot =} \StringTok{"both"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+
+We can also calculate and plot the predicted probabilities (i.e., \(g^{-1}(\eta)\), without actually sampling) given the predictors and the chosen regression coefficients.
+In this way we can try different values and see if predicted probabilities are plausible or not.
+The \texttt{cat\_latent\_plot()} and \texttt{num\_latent\_plot()} functions can be for respectively a categorical (Figures \ref{fig:fig-example-cat-latent}) and numerical predictor (Figures \ref{fig:fig-example-num-latent}).
+
+\scriptsize
 
@@
 
-\caption{caption here}\label{fig:fig-example-cat-latent}
+\caption{\texttt{cat\_latent\_plot(m\ =\ c(0,\ 0.5),\ s\ =\ 1,\ probs\ =\ rep(1/4,\ 4),\ link\ =\ "logit",\ plot\ =\ "both")}}\label{fig:fig-example-cat-latent}
 \end{figure}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{num\_latent\_plot}\NormalTok{(}\AttributeTok{x =} \FunctionTok{runif}\NormalTok{(}\DecValTok{100}\NormalTok{), }\AttributeTok{b1 =} \DecValTok{2}\NormalTok{, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.6}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.1}\NormalTok{, }\FloatTok{0.1}\NormalTok{), }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}
@@
 
-\caption{caption here}\label{fig:fig-example-num-latent}
-\end{figure}
-
-\normalsize
-
-\subsection{Scale Effects}\label{scale-effects}
-
-The default ordinal regression model assume that the variance of the underlying latent distribution is the same across condition.
-This is similar to a standard linear regression assuming the homogeneity of variance.
-For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test (see Delacre, Lakens, \& Leys, 2017).
-In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution.
-This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
-The Equation (number here) expand the previous model including the linear predictor on the scale of the latent distribution.
-
-\begin{equation}
-g(P(Y \leq k)) = \frac{\alpha_k - X\beta}{e^{X\zeta}}
-\end{equation}
-
-\begin{equation} 
-Y^\star_i = \eta + \epsilon_i
-\epsilon_i \sim \mathcal{N}(0, e^{X\zeta})
-\end{equation}
-
-Where \(X\zeta\) is the linear predictor \(\eta\) for the scale of the distribution.
-By dafult for both the logit and probit model the scale is fixed to 1.
-On scale-location models we put predictors on both parameters.
-Given that the scale cannot be negative we use a log link function \(\eta = \text{log}(X\zeta)\).
-Figure \ref{fig:example-unequal-variance} depict an example of a comparison between two groups where the two underlying latent distributions have unequal variance.
-Ignoring the heterogeneity of variances could lead to underestimation of the effect size, reduced power and inflated type-1 error rate (qualche ref qui).
-Furtermore, as suggested by G.
-Tutz and Berger (2017) location-scale models can be considered as a more parsimonius approach compared to partially or completely relaxing the proportional odds assumption.
-Allowing the scale to be different as a function of the predictor create more modelling flexbility.
-Furthemore, two groups could be theoretically different only in the scale of the latent distribution with a similar location.
-In this example, the only way to capture group differences is by including a scale effect.
-The Figure \ref{fig:fig-scale-effect-example} depict the impact of having different scales between two groups on the ordinal probabilities.
-
-Gerhard Tutz (2022) provide a very clear and intuitive explanation of what happen when including a scale effect and how to interpret the result.
-As suggested before, the scale-location model allow to independently predict changes in the location and the scale.
-While location shifts are simply interpreted as increasing/decreasing the latent \(\mu\) or the odds of responding a certain category scale effects are not straightforward.
-As the scale increase (e.g., the variance increase) there is an higher probability mass on extreme categories.
-On the other side as the scale decrease, responses are more concentrated on single categories.
-The categories are determined by the location parameter.
-For example, if one group have a certain latent mean \(\mu_1\) and a small scale \(\sigma^2 = 1/3\) (thus one third compared to the standard version of the distribution), all responses will be focused on categories around the latent mean.
-On the other side, increasing the scale will increase the cumulative probabilities for all categories and for values that tends to infinity extreme categories are preferred.
-Clearly, the scale parameter can somehow be interpreted as the response style (concentrated or variable).
-This is conceptually similar but implemented and interpreted differently to shift-location models {[}Gerhard Tutz (2022); Tutz2020-xq{]} where thresholds are allowed to vary as a function of predictors.
-Both models tries to increase the flexibility of modelling the probability structure of the responses by adding extra paprameters to predict the probabilities of the ordinal responses.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-19-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-19}
-\end{figure}
-
-\normalsize
-
-\subsubsection{Simulating scale effects}\label{simulating-scale-effects}
-
-The location-scale model can be simulated using the \texttt{sim\_ord\_latent()} function and providing the predictors for the scale parameters.
-Given the \texttt{log} link function, predictors are provided on the \texttt{log} scale.
-For example, we simulate the effect of a binary variable \(x\) representing two independent groups predicting the \(k = 5\) response.
-We simulate a \emph{location} effect of \(\beta_1 = 0.5\) (in \emph{probit} scale) and \(\zeta_1 = \text{log}(2) = 0.70\).
-The first group has a \(\sigma = 1\) and the second group has \(\sigma = 2\).
-Again we simulate that the baseline probabilities are uniform for the first group.
-
-\scriptsize
-
-\normalsize
-
-The table (put ref) reports the simulation results.
-
-\scriptsize
-
-\normalsize
-
-To better understand the impact of assuming (or simulating) a different latent scale we fit \(k - 1\) binomial regressions and check the estimated coefficients.
-We are not simulating a specific beta for each outcome but simulating a scale effect is actually impacting the regression coefficients.
-When generating data for a binary outcome the linear predictor is composed by \(\eta = \beta_0 + \beta_1x\).
-The threshold \(\alpha\) and slope of the function can be estimated using \(\alpha = -\frac{\beta_0}{\beta_1}\) and the slope is \(\frac{1}{\beta_1}\) (Faraggi, Izikson, \& Reiser, 2003; Knoblauch \& Maloney, 2012).
-Under the proportional odds assumption, there is only a change in thresholds \(\alpha\) this a shift in the sigmoid along the \(x\) axis.
-When including a scale effect a change in the sigmoid is combined with a change in the slope.
-
-\scriptsize
-
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-22-1}
-
-}
-
-\caption{ }\label{fig:unnamed-chunk-22}
+\caption{\texttt{num\_latent\_plot(x\ =\ runif(100),\ b1\ =\ 2,\ probs\ =\ c(0.6,\ 0.2,\ 0.1,\ 0.1),\ link\ =\ "probit")}}\label{fig:fig-example-num-latent}
 \end{figure}
@@
 In this way \(\beta_1\) will be the main effect of \(X_1\), \(\beta_2\) the main effect of \(X_2\) and \(\beta_3\) the interaction (thus the difference of differences).
-
-\[
-g(P(Y \leq k))^{-1} = \alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i}
-\]
-
-\scriptsize
-
-\begin{verbatim}
-## formula: y ~ x1 * x2
-## data:    dat
-## 
-##  link   threshold nobs  logLik     AIC        niter max.grad cond.H 
-##  probit flexible  4e+05 -597389.03 1194792.06 5(0)  1.42e-07 4.0e+01
-## 
-## Coefficients:
-##         Estimate Std. Error z value Pr(>|z|)    
-## x11     0.004798   0.003410   1.407    0.159    
-## x21     1.004223   0.003519 285.383   <2e-16 ***
-## x11:x21 0.505333   0.006833  73.954   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Threshold coefficients:
-##      Estimate Std. Error z value
-## 1|2 -0.841241   0.002332  -360.8
-## 2|3 -0.254088   0.002092  -121.4
-## 3|4  0.250954   0.002092   120.0
-## 4|5  0.839533   0.002331   360.2
-\end{verbatim}
-
-\normalsize
-
-The parameters intepretation is the same as introduced in the general case.
-The thresholds \(\alpha\) are fixed the determines the probabilities \(P(Y = k)\) when all predictors are zero.
-In this case by doing \texttt{alpha\_to\_prob(fit\$alpha,\ link\ =\ "probit")} we should recover the \texttt{bprobs} vector.
+Equation \eqref{eq:interaction-equation} depict the model formula.
+
+\begin{equation}
+P(Y \leq k) = g^{-1}(\alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i})
+\label{eq:interaction-equation}
+\end{equation}
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{n }\OtherTok{\textless{}{-}} \DecValTok{100}
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{3}
+\NormalTok{betas }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =} \DecValTok{0}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \FloatTok{0.5}\NormalTok{) }\CommentTok{\# b1 = main effect X1, b2 = main effect X2, b3 = interaction}
+
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{expand.grid}\NormalTok{(}\AttributeTok{x1 =} \FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{x2 =} \FunctionTok{c}\NormalTok{(}\StringTok{"c"}\NormalTok{, }\StringTok{"d"}\NormalTok{), }\AttributeTok{n =} \DecValTok{1}\SpecialCharTok{:}\NormalTok{n)}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{x1 }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x1)}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{x2 }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x2)}
+
+\CommentTok{\# sum to 0 coding}
+\FunctionTok{contrasts}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x1) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.5}\NormalTok{)}
+\FunctionTok{contrasts}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.5}\NormalTok{)}
+\NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k)}
+
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ x1 }\SpecialCharTok{*}\NormalTok{ x2, }\AttributeTok{By =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x1 }\SpecialCharTok{*}\NormalTok{ x2, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+\FunctionTok{summary}\NormalTok{(fit)}
+\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x1 * x2}
+\CommentTok{\#\textgreater{} data:    dat}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
+\CommentTok{\#\textgreater{}  probit flexible  400  {-}401.15 812.29 4(0)  2.53e{-}07 2.1e+01}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Coefficients:}
+\CommentTok{\#\textgreater{}         Estimate Std. Error z value Pr(\textgreater{}|z|)    }
+\CommentTok{\#\textgreater{} x11      0.11896    0.11514   1.033    0.302    }
+\CommentTok{\#\textgreater{} x21      1.00637    0.11746   8.568   \textless{}2e{-}16 ***}
+\CommentTok{\#\textgreater{} x11:x21  0.09606    0.23022   0.417    0.676    }
+\CommentTok{\#\textgreater{} {-}{-}{-}}
+\CommentTok{\#\textgreater{} Signif.
+codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Threshold coefficients:}
+\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
+\CommentTok{\#\textgreater{} 1|2 {-}0.53497    0.06870  {-}7.787}
+\CommentTok{\#\textgreater{} 2|3  0.43367    0.06783   6.393}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+The parameters interpretation is the same as introduced in the general case.
+The thresholds \(\alpha\) are fixed representing the probabilities \(P(Y = k)\) when all predictors are zero.
+In this case by doing \texttt{alpha\_to\_prob(fit\$alpha,\ link\ =\ "probit")} we should recover the \texttt{probs0} vector.
 \texttt{x1} is the main effect thus the difference in \(z\) scores between \emph{a} and \emph{b} averaging over \texttt{x2}.
 The same holds for \texttt{x2}.
-\texttt{x1:x2} is the difference of differences in \(z\) scores thus \((ad - ac) - (bd - bc)\).
-
-To visualize the effects there is the \texttt{ggeffect::ggpredict()} function that compute the predictions according to the specified predictors combination.
-In this case we can plot the interaction using the estimated probability for each group and \(Y\) response.
+\texttt{x1:x2} is the difference of differences in \(z\) scores (i.e., the interaction).
+Figure \ref{fig:fig-effects-2-by-2-interaction} depicts the model results.
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-24-1}
+{\centering \includegraphics{paper-new_files/figure-latex/fig-effects-2-by-2-interaction-1}
 
@@
 
-\caption{ }\label{fig:unnamed-chunk-24}
+\caption{ }\label{fig:fig-effects-2-by-2-interaction}
 \end{figure}
@@
 
-\subsection{Numerical by factor interaction}\label{numerical-by-factor-interaction}
-
-Another common scenario is to compute and interaction between a numerical variable \(x\) and a factor \(g\).
-For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution \(x \sim \mathcal{U}(0, 1)\).
-Again, the thresholds \(\alpha\) represents the baseline probabilities when \(x = 0\) and \(g = 0\).
-
-\scriptsize
-
-\begin{verbatim}
-## formula: y ~ x * g
-## data:    dat
-## 
-##  link   threshold nobs  logLik     AIC       niter max.grad cond.H 
-##  probit flexible  1e+05 -153256.51 306527.02 5(0)  7.97e-08 1.6e+02
-## 
-## Coefficients:
-##      Estimate Std. Error z value Pr(>|z|)    
-## x     0.29419    0.01172   25.11   <2e-16 ***
-## g1    0.49679    0.01351   36.78   <2e-16 ***
-## x:g1  0.51690    0.02343   22.06   <2e-16 ***
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Threshold coefficients:
-##      Estimate Std. Error z value
-## 1|2 -0.841442   0.007474 -112.59
-## 2|3 -0.258062   0.007133  -36.18
-## 3|4  0.250592   0.007124   35.18
-## 4|5  0.838293   0.007404  113.23
-\end{verbatim}
+\subsection{Numerical by categorical interaction}\label{numerical-by-categorical-interaction}
+
+Another common scenario is to compute and interaction between a numerical variable \(x\) and a categorical variable \(g\).
+For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution between 0 and 1.
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{n }\OtherTok{\textless{}{-}} \DecValTok{100}
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{3}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{runif}\NormalTok{(n),}
+                  \AttributeTok{g =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{dat}\SpecialCharTok{$}\NormalTok{g }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{g)}
+\FunctionTok{contrasts}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{g) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.5}\NormalTok{)}
+\NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k)}
+
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ x }\SpecialCharTok{*}\NormalTok{ g, }\AttributeTok{By =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.3}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x }\SpecialCharTok{*}\NormalTok{ g, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+\FunctionTok{summary}\NormalTok{(fit)}
+\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x * g}
+\CommentTok{\#\textgreater{} data:    dat}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{}  link   threshold nobs logLik AIC    niter max.grad cond.H }
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}95.71 201.41 5(0)  2.95e{-}08 1.0e+02}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Coefficients:}
+\CommentTok{\#\textgreater{}      Estimate Std. Error z value Pr(\textgreater{}|z|)}
+\CommentTok{\#\textgreater{} x      0.4754     0.4198   1.132    0.257}
+\CommentTok{\#\textgreater{} g1     0.3572     0.4483   0.797    0.426}
+\CommentTok{\#\textgreater{} x:g1   1.2149     0.8401   1.446    0.148}
+\CommentTok{\#\textgreater{} }
+\CommentTok{\#\textgreater{} Threshold coefficients:}
+\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
+\CommentTok{\#\textgreater{} 1|2  {-}0.5090     0.2356  {-}2.160}
+\CommentTok{\#\textgreater{} 2|3   0.2431     0.2321   1.047}
+\end{Highlighting}
+\end{Shaded}
 
@@
 The \texttt{x:g1} is the difference in slopes between the two groups.
-
-Regardless of the model complexity, the \texttt{sim\_ord\_latent()} function can be expanded by including more predictors.
-Clearly, as the model complexity increase, choosing meaningful and empirically plausible parameters became more challenging.
+Figure \ref{fig:fig-effects-num-by-cat-interaction} depicts the model results.
+
+\scriptsize
+
+\begin{figure}
+
+{\centering \includegraphics{paper-new_files/figure-latex/fig-effects-num-by-cat-interaction-1}
+
+}
+
+\caption{ }\label{fig:fig-effects-num-by-cat-interaction}
+\end{figure}
+
+\normalsize
 
@@
 
-In the previous sections we introduced how to simulate standard and scale-location cumulative models and how to fit the corresponding models using the \texttt{clm()} function from the \texttt{ordinal} package.
-In this section we introduce how to compute the power for some common scenarios.
-We present an example in a very general form that can be expanded for specific use cases.
-For example, by simulating a model fixing all effects to zero it is possibile to estimate the type-1 error rate or fixing the sample size and varying the effect size it is possible to do a sensitivity analysis focusing on the smallest effect size of interest (ref qui).
+In this section we introduce how to estimate the statistical power.
+Formally the power is defined as the probability of correctly rejecting the null hypothesis \(H_0\).
+In some simple case, the power can be computed analytically however using data simulation we can handle every scenario.
+We present very general example that, as for the previous simulations, can be extended and modified.
+
+For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting \(H_0\)) or assessing an eventual bias in model estimates.
 The general workflow for a power simulation can be summarized as:
@@
 \item
-  Choose the research design, e.g., 2x2 factorial design
-\item
-  Indentifify the specific effect to focus on. For example the interaction effect, a main effect or a specific contrast
-\item
-  Write the model equation similarly to what presented in the previous sections and extract the model coefficients
-\item
-  Choose one of the proposed strategy for simulating data and follows the corresponding steps
-\item
-  Fit the model and extract the relevant statistics e.g., the p-value of the interaction
+  Specify the research design, e.g., 2x2 factorial design
+\item
+  Specify the effect/parameter of interest, e.g., the interaction effect
+\item
+  Define the simulation conditions, e.g., a range of sample size, effect size, etc.
+\item
+  Implement one of the simulation workflow described in the previous sections
 \item
@@
 \item
-  Summarise the simulation results in the appropriate way.
-For example, the power of the interaction is the number of times the p-value for the interaction test is lower than the critical level divided by the number of simulations.
+  Summarise the simulation results
 \end{enumerate}
 
-To make a practical example, we simulate the power of a cumulative link model in detecting an effect of \(d = 0.4\) (probit model) with \(k = 6\) and \(n = 40\) participants per group.
-We fix the baseline probabilities in the first group as uniform.
-
-descrizione migliore
-
-\scriptsize
-
-\normalsize
-
-Then to compute the power we can simply compute the proportion of p values lower than the critical level.
-This can be done in R as \texttt{mean(p\ \textless{}=\ alpha)} resulting in an estimated power of 0.39.
-Clearly using only a single condition is not really informative thus we can vary the sample size across plausible values and check the corresponding power curve.
-
-\scriptsize
+To make a practical example, we simulate the power of detecting a group difference of \(d = 0.4\) (assuming a \emph{probit} model).
+Participants respond to an ordinal variable with \(k = 5\).
+The iteration is performed using a \texttt{for} loop that repeat 1000 times the data simulation, model fitting and extract the p-value for the \(\beta_1\).
+The power is then estimated as the number of p-values lower than the critical level over the number of simulations.
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{n }\OtherTok{\textless{}{-}} \DecValTok{40} \CommentTok{\# sample size}
+\NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{5}  \CommentTok{\# number of ordinal variables}
+\NormalTok{d }\OtherTok{\textless{}{-}} \FloatTok{0.4} \CommentTok{\# effect size (i.e., our regression coefficients)}
+\NormalTok{nsim }\OtherTok{\textless{}{-}} \FloatTok{1e3} \CommentTok{\# higher is better, here using 1000 for an example}
+\NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k)}
+\NormalTok{alpha }\OtherTok{\textless{}{-}} \FloatTok{0.05} \CommentTok{\# critical alpha}
+
+\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\ConstantTok{NA}\NormalTok{, nsim) }\CommentTok{\# preallocation to improve loop computational efficiency}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{group =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n)) }\CommentTok{\# data frame}
+
+\FunctionTok{head\_tail}\NormalTok{(dat, }\AttributeTok{n =} \DecValTok{3}\NormalTok{)}
+\CommentTok{\#\textgreater{}    group}
+\CommentTok{\#\textgreater{} 1      a}
+\CommentTok{\#\textgreater{} 2      a}
+\CommentTok{\#\textgreater{} 3      a}
+\CommentTok{\#\textgreater{} 78     b}
+\CommentTok{\#\textgreater{} 79     b}
+\CommentTok{\#\textgreater{} 80     b}
+
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{nsim)\{}
+\NormalTok{  sim }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{group, }\AttributeTok{By =}\NormalTok{ d, }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{  fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ group, }\AttributeTok{data =}\NormalTok{ sim, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+\NormalTok{  p[i] }\OtherTok{\textless{}{-}} \FunctionTok{summary}\NormalTok{(fit)}\SpecialCharTok{$}\NormalTok{coefficients[}\StringTok{"groupb"}\NormalTok{, }\StringTok{"Pr(\textgreater{}|z|)"}\NormalTok{] }\CommentTok{\# extract the pvalue}
+\NormalTok{\}}
+
+\CommentTok{\# estimate the power}
+\FunctionTok{mean}\NormalTok{(p }\SpecialCharTok{\textless{}=}\NormalTok{ alpha, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
+\CommentTok{\#\textgreater{} [1] 0.386}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+Despite useful, calculating power curves instead a single value is more informative.
+For this reason we repeat the previous simulation but for different sample sizes.
+Figure \ref{fig:fig-power-curve} depict the power curve resulting from the simulation.
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{n }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{100}\NormalTok{, }\DecValTok{200}\NormalTok{)}
+\NormalTok{power }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\FunctionTok{length}\NormalTok{(n))}
+
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(n))\{}
+\NormalTok{  p }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\ConstantTok{NA}\NormalTok{, nsim) }\CommentTok{\# preallocation for speed}
+\NormalTok{  dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{group =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n[i]))}
+  \ControlFlowTok{for}\NormalTok{(j }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{nsim)\{}
+\NormalTok{    sim }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{group, }\AttributeTok{By =}\NormalTok{ d, }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{    fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ group, }\AttributeTok{data =}\NormalTok{ sim, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+\NormalTok{    p[j] }\OtherTok{\textless{}{-}} \FunctionTok{summary}\NormalTok{(fit)}\SpecialCharTok{$}\NormalTok{coefficients[}\StringTok{"groupb"}\NormalTok{, }\StringTok{"Pr(\textgreater{}|z|)"}\NormalTok{]}
+\NormalTok{  \}}
+\NormalTok{  power[i] }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(p }\SpecialCharTok{\textless{}=}\NormalTok{ alpha)}
+\NormalTok{\}}
+
+\NormalTok{power}
+\CommentTok{\#\textgreater{} [1] 0.219 0.409 0.531 0.749 0.970}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/unnamed-chunk-28-1}
+{\centering \includegraphics{paper-new_files/figure-latex/fig-power-curve-1}
 
@@
 
-\caption{ }\label{fig:unnamed-chunk-28}
+\caption{caption}\label{fig:fig-power-curve}
 \end{figure}
@@
 
-\bibitem[\citeproctext]{ref-DeCarlo1998-ay}
-DeCarlo, L. T. (1998). Signal detection theory and generalized linear models. \emph{Psychological Methods}, \emph{3}, 186--205. \url{https://doi.org/10.1037/1082-989X.3.2.186}
-
 \bibitem[\citeproctext]{ref-DeCarlo2010-lj}
@@
 
-\bibitem[\citeproctext]{ref-Delacre2017-qy}
-Delacre, M., Lakens, D., \& Leys, C.
-(2017).
-Why psychologists should by default use welch's \emph{t}-test instead of student's \emph{t}-test.
-\emph{International Review of Social Psychology}, \emph{30}, 92.
-\url{https://doi.org/10.5334/irsp.82}
-
-\bibitem[\citeproctext]{ref-Faraggi2003-hm}
-Faraggi, D., Izikson, P., \& Reiser, B. (2003). Confidence intervals for the 50 per cent response dose. \emph{Statistics in Medicine}, \emph{22}, 1977--1988. \url{https://doi.org/10.1002/sim.1368}
-
 \bibitem[\citeproctext]{ref-Gelman2020-tg}
@@
 
-\bibitem[\citeproctext]{ref-Green1966-gy}
-Green, D. M., Swets, J. A., \& Others. (1966). \emph{Signal detection theory and psychophysics} (Vol. 1, pp. xi, 455--xi, 455). Oxford, England: Wiley New York.
-
 \bibitem[\citeproctext]{ref-Jamieson2004-xf}
@@
 Kemp, S., \& Grace, R. C. (2021). Using ordinal scales in psychology. \emph{Methods in Psychology (Online)}, \emph{5}, 100054. \url{https://doi.org/10.1016/j.metip.2021.100054}
-
-\bibitem[\citeproctext]{ref-Knoblauch2012-to}
-Knoblauch, K., \& Maloney, L. T. (2012). \emph{Modeling psychophysical data in r}. New York, NY: Springer New York. \url{https://doi.org/10.1007/978-1-4614-4475-6}
 
@@
 
-\bibitem[\citeproctext]{ref-Stanislaw1999-jr}
-Stanislaw, H., \& Todorov, N.
-(1999).
-Calculation of signal detection theory measures.
-\emph{Behavior Research Methods, Instruments, \& Computers: A Journal of the Psychonomic Society, Inc}, \emph{31}, 137--149.
-\url{https://doi.org/10.3758/bf03207704}
-
 \bibitem[\citeproctext]{ref-Stevens1946-te}
@@
 \bibitem[\citeproctext]{ref-Tutz1990-fe}
-Tutz, Gerhard.
+Tutz, G.
 (1990).
@@
 \bibitem[\citeproctext]{ref-Tutz2022-dg}
-Tutz, Gerhard. (2022). Ordinal regression: A review and a taxonomy of models. \emph{Wiley Interdisciplinary Reviews. Computational Statistics}, \emph{14}. \url{https://doi.org/10.1002/wics.1545}
-
-\bibitem[\citeproctext]{ref-Tutz2017-de}
+Tutz, G. (2022). Ordinal regression: A review and a taxonomy of models. \emph{Wiley Interdisciplinary Reviews. Computational Statistics}, \emph{14}. \url{https://doi.org/10.1002/wics.1545}
+
+\bibitem[\citeproctext]{ref-Tutz2020-xq}
 Tutz, G., \& Berger, M.
-(2017).
-Separating location and dispersion in ordinal regression models.
-\emph{Econometrics and Statistics}, \emph{2}, 131--148.
-\url{https://doi.org/10.1016/j.ecosta.2016.10.002}
-
-\bibitem[\citeproctext]{ref-Tutz2020-xq}
-Tutz, Gerhard, \& Berger, M.
 (2020).

### 2024-01-04 supplementary/supplementary.Rmd
update

---
title: "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
subtitle: "Supplementary Materials"
author:
  - Filippo Gambarota
  - Gianmarco Altoè
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 1
    latex_engine: xelatex
header-includes:
  - \setcounter{table}{0} 
  - \renewcommand*{\thetable}{S\arabic{table}}
  - \setcounter{figure}{0} 
  - \renewcommand*{\thefigure}{S\arabic{figure}}
  - \setcounter{equation}{0}
  - \renewcommand{\theequation}{S\arabic{equation}}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

```{r}
xl <- seq(0, 1, 0.1)
x <- rep(xl, 10)
n <- length(x)
b0 <- qlogis(0.01)
b1 <- 8
lp <- b0 + b1 * x # linear predictor
y <- rbinom(n, 1, plogis(lp)) # sampling from a binomial distribution using the vector of probabilities
p <- tapply(y, x, mean)
plot(unique(x), p, pch = 19)
curve(plogis(x, -(b0/b1), 1/b1), add = TRUE, lwd = 2, col = "firebrick")
```

Essentially the `lp` ($\eta_i$) vector contains true probability of success ($p_i|x_i = g^{-1}(\eta_i)$) for the $x_i$ level of the predictor. Then to introduce the random part of the model, we can use $p_i$ to generate binary values from a Binomial (in this specific case a Bernoulli) distribution. This create `y` that is the vector of 0-1 values generated using the true probabilities of success. The supplementary materials contains a more extensive example for simulating data for generalized linear models.

We can apply the same idea to an ordinal outcome but we need $k - 1$ equations (compared to the previous example) where $k$ is the number of ordinal levels. Let's simulate a similar design with a continuous $x$ predictor and $k = 4$ options. We fix the baseline probabilities where $x = 0$ as uniform thus $p(y_1) = p(y_2) = ... p(y_k) = 1/k$.

```{r}
k <- 4  # number of options (1)
n <- 1e5 # number of observations (1)
b1 <- 0.5 # beta1, the shift in the latent distribution (2)
probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
dat <- data.frame(x = rnorm(n)) # create dataframe
lp <- lapply(alphas, function(a) a - b1 * dat$x) # k - 1 linear predictors (4)
names(lp) <- sprintf("g[p(y <= %s)]", 1:(k - 1)) # giving appropriate names
```

The linear predictors ($\eta$) can be seen in the Figure ref. The $g$ in this case is the cumulative *probit* function $\Phi$.

```{r}
#| echo: false
lpt <- lp
names(lpt) <- latex2exp::TeX(sprintf("$g[p(y \\leq %s)]$", 1:(k - 1)))
datl <- cbind(dat, lpt)
datl <- tidyr::pivot_longer(datl, 2:ncol(datl), names_to = "cumy", values_to = "lp")

ggplot(datl, aes(x = x, y = lp, color = cumy)) +
  facet_wrap(~cumy, labeller = label_parsed) +
  geom_hline(yintercept = alphas, lty = "dashed") +
  geom_line(show.legend = FALSE) +
  theme_minimal(20) +
  ylab("Linear Predictor")
```

Now we can apply the inverse of the link function $g^{-1} = \Phi^{-1}$ to calculate the corresponding cumulative probabilities.

```{r}
cump <- lapply(lp, pnorm) # inverse of the link function (5)
cump <- data.frame(cump)
p <- apply(cbind(0, cump, 1), 1, diff, simplify = FALSE) # probability of each k outcome (6)
p <- do.call(rbind, p)
```

```{r}
p <- data.frame(p)
names(p) <- sprintf("p_y%s", 1:k)
filor::trim_df(data.frame(p))
```

We can plot the expected effect of $\beta_1$ on the $k$ probabilities using the `num_latent_plot()` function (see Figure \@ref(fig:sim-with-probs-lat-plot)). Finally we sample using the `sample()` function using the calculated probabilities.

```{r sim-with-probs-lat-plot, fig.cap="sim-with-probs-lat-plot"}
num_latent_plot(x = dat$x, b1 = b1, probs = probs0, link = "probit")
```

```{r}
dat$y <- apply(p, 1, function(ps) sample(1:k, 1, prob = ps))
```

## Scale Effects

The default ordinal regression model assume that the variance of the underlying latent distribution is the same across condition. This is similar to a standard linear regression assuming the homogeneity of variance. For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy]. In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution. This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included. The Equation (number here) expand the previous model including the linear predictor on the scale of the latent distribution.

```{=tex}
\begin{equation}
g(P(Y \leq k)) = \frac{\alpha_k - X\beta}{e^{X\zeta}}
\end{equation}
```

```{=tex}
\begin{equation} 
Y^\star_i = \eta + \epsilon_i
\epsilon_i \sim \mathcal{N}(0, e^{X\zeta})
\end{equation}
```

Where $X\zeta$ is the linear predictor $\eta$ for the scale of the distribution. By dafult for both the logit and probit model the scale is fixed to 1. On scale-location models we put predictors on both parameters. Given that the scale cannot be negative we use a log link function $\eta = \text{log}(X\zeta)$. Figure \@ref(fig:example-unequal-variance) depict an example of a comparison between two groups where the two underlying latent distributions have unequal variance. Ignoring the heterogeneity of variances could lead to underestimation of the effect size, reduced power and inflated type-1 error rate (qualche ref qui). Furtermore, as suggested by @Tutz2017-de location-scale models can be considered as a more parsimonius approach compared to partially or completely relaxing the proportional odds assumption. Allowing the scale to be different as a function of the predictor create more modelling flexbility. Furthemore, two groups could be theoretically different only in the scale of the latent distribution with a similar location. In this example, the only way to capture group differences is by including a scale effect. The Figure \@ref(fig:fig-scale-effect-example) depict the impact of having different scales between two groups on the ordinal probabilities.

@Tutz2022-dg provide a very clear and intuitive explanation of what happen when including a scale effect and how to interpret the result. As suggested before, the scale-location model allow to independently predict changes in the location and the scale. While location shifts are simply interpreted as increasing/decreasing the latent $\mu$ or the odds of responding a certain category scale effects are not straightforward. As the scale increase (e.g., the variance increase) there is an higher probability mass on extreme categories. On the other side as the scale decrease, responses are more concentrated on single categories. The categories are determined by the location parameter. For example, if one group have a certain latent mean $\mu_1$ and a small scale $\sigma^2 = 1/3$ (thus one third compared to the standard version of the distribution), all responses will be focused on categories around the latent mean. On the other side, increasing the scale will increase the cumulative probabilities for all categories and for values that tends to infinity extreme categories are preferred. Clearly, the scale parameter can somehow be interpreted as the response style (concentrated or variable). This is conceptually similar but implemented and interpreted differently to shift-location models [@Tutz2022-dg; Tutz2020-xq] where thresholds are allowed to vary as a function of predictors. Both models tries to increase the flexibility of modelling the probability structure of the responses by adding extra paprameters to predict the probabilities of the ordinal responses.

```{r}
sigma_high <- cat_latent_plot(m = c(0, 0), 
                s = c(1, 2), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$"))
sigma_low <- cat_latent_plot(m = c(0, 0), 
                s = c(1, 0.5), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$"))
sigma_high_mu1 <- cat_latent_plot(m = c(0, 1), 
                s = c(1, 2), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$, $\\mu_2 = 1$"))
sigma_low_mu1 <- cat_latent_plot(m = c(0, 1), 
                s = c(1, 0.5), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$, $\\mu_2 = 1$"))
plot_grid(sigma_high, sigma_low, sigma_high_mu1, sigma_low_mu1)
```

### Simulating scale effects

The location-scale model can be simulated using the `sim_ord_latent()` function and providing the predictors for the scale parameters. Given the `log` link function, predictors are provided on the `log` scale. For example, we simulate the effect of a binary variable $x$ representing two independent groups predicting the $k = 5$ response. We simulate a *location* effect of $\beta_1 = 0.5$ (in *probit* scale) and $\zeta_1 = \text{log}(2) = 0.70$. The first group has a $\sigma = 1$ and the second group has $\sigma = 2$. Again we simulate that the baseline probabilities are uniform for the first group.

```{r}
k <- 5  # number of options (1)
n <- 1e5 # number of observations (1)
b1 <- 0.5 # beta1, the shift in the latent distribution (2)
z1 <- log(2) # zeta1, the change in the scale
probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
dat <- data.frame(x = rep(c(0, 1), each = n/2))
dat <- sim_ord_latent(~x, scale = ~x, By = b1, Bscale = z1, prob = probs0, data = dat, link = "probit")
fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
```

The table (put ref) reports the simulation results.

```{r}
#| eval: false
truth <- c(alphas, b1, z1)
names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
#clm_table(fit, truth)
```

To better understand the impact of assuming (or simulating) a different latent scale we fit $k - 1$ binomial regressions and check the estimated coefficients. We are not simulating a specific beta for each outcome but simulating a scale effect is actually impacting the regression coefficients. When generating data for a binary outcome the linear predictor is composed by $\eta = \beta_0 + \beta_1x$. The threshold $\alpha$ and slope of the function can be estimated using $\alpha = -\frac{\beta_0}{\beta_1}$ and the slope is $\frac{1}{\beta_1}$ [@Knoblauch2012-to; @Faraggi2003-hm]. Under the proportional odds assumption, there is only a change in thresholds $\alpha$ this a shift in the sigmoid along the $x$ axis. When including a scale effect a change in the sigmoid is combined with a change in the slope.

```{r}
set.seed(2023)
k <- 4
n <- 1e5
b1 <- 3
d1 <- log(2)
dat <- data.frame(x = runif(n))
dat <- sim_ord_latent(~x, ~x, By = b1, Bscale = d1, prob = rep(1/k, k), data = dat, link = "probit")

dat$y1vs234 <- ifelse(dat$y <= 1, 1, 0)
dat$y12vs34 <- ifelse(dat$y <= 2, 1, 0)
dat$y123vs4 <- ifelse(dat$y <= 3, 1, 0)

dat$y <- ordered(dat$y)
fit <- clm(y ~ x, scale = ~x, data = dat, link = "probit")

fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "probit"))
fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "probit"))
fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "probit"))

fits <- list(y1vs234 = fit1vs234, fit12vs34 = fit12vs34, fit123vs4 = fit123vs4)

lapply(fits, function(x) coef(x)) |> 
  bind_rows(.id = "model") |> 
  mutate(xp = list(seq(0, 5, 0.01))) |> 
  unnest(xp) |> 
  ggplot(aes(x = xp, y = `(Intercept)` + x * xp)) +
  geom_line(aes(color = model))
```

The Figure \@ref(fig:fig-odds-example) shows the relationship between probabilities and odds.

```{r fig-odds-example, fig.cap=filor::get_caption()}
p <- seq(0, 1, 0.001)

dd <- data.frame(
  p,
  odds = filor::odds(p)
)

dd$lodds <- log(dd$odds)

dd |> 
  tidyr::pivot_longer(c(odds, lodds)) |> 
  dplyr::mutate(name = ifelse(name == "odds", "Odds", "Log(Odds)")) |> 
  ggplot(aes(x = p, y = value)) +
  geom_line() +
  facet_wrap(~name, scales = "free") +
  theme(axis.title.y = element_blank()) +
  xlab("Probability")
```

### 2024-01-04 R/utils-plotting.R

@@ -2,12 +2,18 @@ show_alpha <- function(prob = NULL,
                        alpha = NULL, 
                        link = c("logit", "probit"),
                        plot = c("latent", "cumulative", "probs")){
-  require(ggplot2)
-  require(cowplot)
-  require(ggtext)
   
   if(is.null(prob) & is.null(alpha)){
     stop("alpha or prob must be specified!")
+  }
+  
+  if(is.null(prob)){
+    prob <- alpha_to_prob(alpha, link)
+  }else{
+    if(sum(prob) != 1){
+      stop("the prob vector must sum to 1!")
+    }
+    alpha <- prob_to_alpha(prob, link)
   }
   
   lf <- get_link(link)
@@ -18,15 +24,6 @@ show_alpha <- function(prob = NULL,
   }else{
     xlim <- c(-5, 5)
     dist <- distributional::dist_normal
-  }
-  
-  if(is.null(prob)){
-    prob <- alpha_to_prob(alpha, link)
-  }else{
-    if(sum(prob) != 1){
-      stop("the prob vector must sum to 1!")
-    }
-    alpha <- prob_to_alpha(prob, link)
   }
   
   k <- length(prob)
@@ -53,7 +50,7 @@ show_alpha <- function(prob = NULL,
                               y = 1, 
                               label = alpha_n), 
                           angle = 90,
-                          hjust = 0.5)
+                          hjust = 1)
   
   y_n <- sprintf("P(Y = %s)", 1:k)
   y_cp <- lf$pfun(c(-Inf, alpha, Inf))
@@ -76,7 +73,7 @@ show_alpha <- function(prob = NULL,
     p = diff(y_cp)
   )
   
-  hplot <- ggplot(pp, aes(x = y, fill = factor(y), y = p)) +
+  hplot <- ggplot(pp, aes(x = factor(y), fill = factor(y), y = p)) +
     geom_col(width = 0.6) +
     ylab("Probability") +
     theme_minimal(15) +
@@ -86,11 +83,8 @@ show_alpha <- function(prob = NULL,
     ylim(c(0, 1))
   
   plist <- list(latent = dplot, cumulative = pplot, probs = hplot)
-  plist <- plist[plot]
-  bl <- get_legend(dplot)
-  
-  pgrid <- plot_grid(plotlist = lapply(plist, function(x) x + theme(legend.position = "none")))
[…]

### 2024-01-04 R/utils.R

@@ -108,8 +108,7 @@ alpha_to_prob <- function(alpha, link = c("logit", "probit"), ...){
   lf <- get_link(link)
   ths <- c(-Inf, unname(alpha), Inf)
   cprobs <- lf$pfun(ths, ...)
-  probs <- diff(cprobs)
-  return(probs)
+  diff(cprobs)
 }
 
 #' Transform probabilities into thresholds

### 2024-01-04 scripts/tables.R

@@ -1,4 +1,7 @@
 library(flextable)
+devtools::load_all()
+
+# MODEL TABLE
 
 models_table <- data.frame(
   stringsAsFactors = FALSE,
@@ -27,8 +30,48 @@ models_table[-1, ] |>
   ftExtra::colformat_md(part = "all") |> 
   save_as_docx(path = sprintf(here::here("paper", "tables", "%s.docx"), "notation-table"))
 
+
+# DATASET EXAMPLE TABLE
+
+b1 <- log(3) # log odds ratio
+k <- 4
+n <- 100
+x <- rep(c("a", "b"), each = n/2)
+dat <- data.frame(x = x)
+probs <- rep(1/k, k) # for the group "a", uniform probabilities
+dat <- sim_ord_latent(~x, By = b1, prob = probs, data = dat, link = "logit")
+
+funs <- list(mean = mean, median = median, sd = sd)
+ss <- sapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
+ff <- t(table(dat$y, dat$x))
+
+ss <- data.frame(cbind(ss, ff))
+names(ss) <- c("mean", "median", "sd", paste0("Y", 1:4))
+
+ss$tot <- apply(ss[, 4:ncol(ss)], 1, sum)
+pp <- ss[, 4:7]/50
+ss <- round(ss, 2)
+ss$Y1 <- sprintf("**%s** (*p = %.2f*)", ss$Y1, pp$Y1)
+ss$Y2 <- sprintf("**%s** (*p = %.2f*)", ss$Y2, pp$Y2)
+ss$Y3 <- sprintf("**%s** (*p = %.2f*)", ss$Y3, pp$Y3)
+ss$Y4 <- sprintf("**%s** (*p = %.2f*)", ss$Y4, pp$Y4)
+
+ss$group <- rownames(ss)
+ss <- select(ss, group, everything(), -tot)
+
+flextable(ss) |> 
+  autofit() |> 
+  set_caption(caption = filor::get_caption()) |> 
+  align(part = "all", align = "center") |> 
+  ftExtra::colformat_md(part = "all") |> 
+  save_as_docx(path = sprintf(here::here("paper", "tables", "%s.docx"), "dataset-example-table"))
+
+
+# Saving
+
 tables <- filor::named_list(
-  models_table
+  models_table,
+  data_example_table = list(dat = dat, table = ss)
 )
 
 saveRDS(tables, here::here("objects", "r-tabs.rds"))

### 2024-01-06 paper/paper-new.Rmd
updating

@@
 footnotelist      : no
+params:
+  floatplaceh: true
 
@@
 }
+```
+
+```{r}
+#| label: fighook
+#| include: false
+#| eval: !expr params$floatplaceh
+
+fig_counter <- filor::scount()
+tbl_counter <- filor::scount()
+
+knitr::knit_hooks$set(fig.place = function(before = FALSE, options) {
+  if(isTRUE(options$fig.place)){
+      if(before){
+          NULL
+      }else{
+          sprintf("\n \\begin{center}\\textbf{[Figure %s about here]} \\end{center}\n", fig_counter())
+      }
+  }
+})
+knitr::knit_hooks$set(tbl.place = function(before = FALSE, options) {
+  if(isTRUE(options$tbl.place)){
+      if(before){
+          NULL
+      }else{
+          sprintf("\n \\begin{center}\\textbf{[Table %s about here]} \\end{center}\n", tbl_counter())
+      }
+  }
+})
 ```
@@
 
-```{r fig-explain-cumulative, fig.cap=filor::get_caption()}
+```{r fig-explain-cumulative, fig.cap=filor::get_caption(), fig.place = TRUE}
 probs <- c(0.4, 0.2, 0.3, 0.1)
@@
 
-```{r fig-logit-vs-probit, fig.cap=filor::get_caption()}
+```{r fig-logit-vs-probit, fig.cap=filor::get_caption(), fig.place = TRUE}
 dplot <- ggplot(data = NULL) +
@@
 
-```{r fig-odds-example, fig.cap=filor::get_caption()}
+```{r fig-odds-example, fig.cap=filor::get_caption(), fig.place = TRUE}
 p <- seq(0, 1, 0.001)
@@
 
-```{r fig-prop-odds, fig.cap=filor::get_caption()}
+```{r fig-prop-odds, fig.cap=filor::get_caption(), fig.place = TRUE}
 rfigs$fig_prop_odds
@@
 \begin{equation}
-z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
+z_{x_1 - x_0} = \Phi(P \leq 1|x_0) - \Phi(P \leq 1|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
 (\#eq:probit-prop-odds)
@@
 
-```{r fig-sim-from-latent}
+```{r fig-sim-from-latent, fig.cap = filor::get_caption(), fig.place = TRUE}
 alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))
@@
 
-```{r}
+```{r simulating-from-latent3}
 #| echo: true
@@
 
-```{r fig-show-th-example}
-p1 <- show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))
-p2 <- show_alpha(prob = c(0.1, 0.2, 0.5, 0.2), link = "probit", plot = c("latent", "probs"))
-plot_grid(p1, p2, ncol = 1)
+```{r fig-show-th-example, fig.cap=filor::get_caption(), fig.place = TRUE}
+p1 <- show_alpha(prob = rep(1/4, 4), link = "probit")
+p2 <- show_alpha(prob = c(0.1, 0.2, 0.5, 0.2), link = "probit")
+
+p1 <- plot_grid(p1$latent, p1$probs)
+p2 <- plot_grid(p2$latent, p2$probs)
+
+plot_grid(p1, 
+          p2, 
+          ncol = 1)
 ```
@@
 
-```{r fig-example-cat-latent, fig.cap=filor::get_caption()}
+```{r fig-example-cat-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
 cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")
@@
 
-```{r fig-example-num-latent, fig.cap=filor::get_caption()}
+```{r fig-example-num-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
 num_latent_plot(x = runif(100), b1 = 2, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit") +
@@
 
-```{r fig-effects-2-by-2-interaction}
+```{r fig-effects-2-by-2-interaction, fig.place = TRUE}
 # plot(ggpredict(fit, terms = c("x1", "x2")))
@@
 
-```{r fig-effects-num-by-cat-interaction}
+```{r fig-effects-num-by-cat-interaction, fig.place = TRUE}
 # plot(ggpredict(fit, terms = c("x[all]", "g")))
@@
 
-```{r fig-power-curve, fig.cap=filor::get_caption()}
+```{r fig-power-curve, fig.cap=filor::get_caption(), fig.place = TRUE}
 power_df <- data.frame(power = power, n = n)
@@
 
+(ref:fig-sim-from-latent) caption
+
+(ref:fig-show-th-example) caption
+
 <!-- TABLES -->

### 2024-01-06 R/deprecated.R

@@ -94,3 +94,28 @@ get_slope <- function(b1){
 get_b0_from_th <- function(th, b1){
   -b1 * th
 }
+get_probs <- function(formula, 
+                      B, 
+                      probs0, 
+                      data, 
+                      link = c("logit", "probit"),
+                      ynames = NULL,
+                      append = FALSE){
+  if(is.null(ynames)){
+    ynames <- paste0("y", 1:length(probs0))
+  }
+  lf <- get_link(link)
+  ths <- c(-Inf, prob_to_alpha(probs0, link = link), Inf)
+  X <- model.matrix(formula, data = data)[, -1]
+  X <- matrix(X)
+  # notice the minus sign t - X %*% B thus an increase in x
+  # is an increase in p
+  P <- lapply(ths, function(t) c(lf$pfun(t - X %*% B)))
+  P <- data.frame(P)
+  P <- data.frame(t(apply(P, 1, diff)))
+  names(P) <- ynames
+  if(append){
+    P <- cbind(data, P)
+  }
+  return(P)
+}

### 2024-01-06 R/utils-plotting.R

@@ -1,7 +1,6 @@
 show_alpha <- function(prob = NULL, 
                        alpha = NULL, 
-                       link = c("logit", "probit"),
-                       plot = c("latent", "cumulative", "probs")){
+                       link = c("logit", "probit")){
   
   if(is.null(prob) & is.null(alpha)){
     stop("alpha or prob must be specified!")
@@ -82,9 +81,7 @@ show_alpha <- function(prob = NULL,
           axis.title.y = element_blank()) +
     ylim(c(0, 1))
   
-  plist <- list(latent = dplot, cumulative = pplot, probs = hplot)
-  print(cowplot::plot_grid(plotlist = (plist[plot])))
-  invisible(plist)
+  list(latent = dplot, cumulative = pplot, probs = hplot)
 }
 
 cat_latent_plot <- function(m = 0, 
@@ -181,7 +178,8 @@ num_latent_plot <- function(x,
     th <- prob_to_alpha(probs, link)
   }
   data <- data.frame(x = seq(min(x), max(x), length.out = nsample))
-  data <- get_probs(~x, b1, probs, data, link, append = TRUE)
+  #data <- get_probs(~x, b1, probs, data, link, append = TRUE)
+  data <- sim_ord_latent(~x, By = b1, prob = probs, data = data, link = link, simulate = FALSE)
   datal <- tidyr::pivot_longer(data, starts_with("y"), names_to = "y", values_to = "value")
   ggplot(datal, aes(x = x, y = value, color = y)) +
     geom_line(linewidth = linewidth) +

### 2024-01-06 R/utils.R

@@ -1,33 +1,3 @@
-get_probs <- function(formula, 
-                      B, 
-                      probs0, 
-                      data, 
-                      link = c("logit", "probit"),
-                      ynames = NULL,
-                      append = FALSE){
-  if(is.null(ynames)){
-    ynames <- paste0("y", 1:length(probs0))
-  }
-  lf <- get_link(link)
-  ths <- c(-Inf, prob_to_alpha(probs0, link = link), Inf)
-  X <- model.matrix(formula, data = data)[, -1]
-  X <- matrix(X)
-  # notice the minus sign t - X %*% B thus an increase in x
-  # is an increase in p
-  P <- lapply(ths, function(t) c(lf$pfun(t - X %*% B)))
-  P <- data.frame(P)
-  P <- data.frame(t(apply(P, 1, diff)))
-  names(P) <- ynames
-  if(append){
-    P <- cbind(data, P)
-  }
-  return(P)
-}
-
-vlogit <- function(scale = 1){
-  (scale^2 * pi^2)/3
-}
-
 #' Simulating Ordinal Data
 #'
 #' @param location a formula with predictors for the location parameter. The components of the formula need to be columns of the \code{data} object.
@@ -38,6 +8,7 @@ vlogit <- function(scale = 1){
 #' @param alpha vector of baseline thresholds. Either \code{probs} or \code{alpha} must be specified. The function calculated thresholds from \code{probs} and viceversa
 #' @param data a dataframe with predictors. both \code{location} and \code{scale} formulas refers to columns of the dataframe.
 #' @param link character vector indicating the link function. Can be \code{logit} or \code{probit}.
+#' @param simulate logical indicating if sampling values (`TRUE`) or returning the true probabilities 
 #'
 #' @return a dataframe with the ordinal variable \code{y} and the latent variable \code{ys}
 #' @export
@@ -57,7 +28,8 @@ sim_ord_latent <- function(location,
                            prob = NULL, 
                            alpha = NULL, 
                            data, 
-                           link = c("logit", "probit")){
+                           link = c("logit", "probit"),
+                           simulate = TRUE){
   
   # get all the correct functions according to the link
   lf <- get_link(link)
[…]

### 2024-01-06 tests/testthat.R

@@ -1,0 +1,12 @@
+# This file is part of the standard setup for testthat.
+# It is recommended that you do not modify it.
+#
+# Where should you do additional test configuration?
+# Learn more about the roles of various files in:
+# * https://r-pkgs.org/testing-design.html#sec-tests-files-overview
+# * https://testthat.r-lib.org/articles/special-files.html
+
+library(testthat)
+library(ordinalsim)
+
+test_check("ordinalsim")

### 2024-01-06 tests/testthat/settings.R

@@ -1,0 +1,1 @@
+.tol <- c(clm = 0.1)

### 2024-01-06 tests/testthat/test-utils.R

@@ -1,0 +1,179 @@
+source("settings.R") # loading objects
+
+test_that("alpha_to_prob() and prob_to_alpha() are working", {
+  p1 <- rep(1/6, 6)
+  a1l <- prob_to_alpha(p1, link = "logit")
+  a1p <- prob_to_alpha(p1, link = "probit")
+  expect_equal(p1, alpha_to_prob(a1l, link = "logit"), info = "logit link")
+  expect_equal(p1, alpha_to_prob(a1p, link = "probit"), info = "probit link")
+})
+
+test_that("dummy_ord() is working!", {
+  k <- 5
+  y <- sample(1:k, 100, TRUE)
+  dy <- dummy_ord(y)
+  expect_equal(ncol(dy), k - 1, info = "number of dummy k - 1")
+})
+
+
+test_that("sim_ord_latent(link = 'logit') works with a binary x, k = 4", {
+  set.seed(0)
+  k <- 4
+  n <- 1e5
+  probs0 <- c(0.1, 0.3, 0.5, 0.1)
+  alpha <- prob_to_alpha(probs0, link = "logit")
+  b1 <- log(2) # log odds ratio
+  dat <- data.frame(x = rep(0:1, each = n))
+  dat <- sim_ord_latent(~x, By = b1, prob = probs0, data = dat, link = "logit")
+  fit <- ordinal::clm(y~x, data = dat, link = "logit")
+  
+  expect_equal(b1, fit$beta, tolerance = .tol["clm"], info = "beta", ignore_attr = TRUE)
+  expect_equal(alpha, fit$alpha, tolerance = .tol["clm"], info = "alpha", ignore_attr = TRUE)
+  expect_equal(nrow(dat), n*2, info = "sample size", ignore_attr = TRUE)
+  
+  dat_fit <- dat[, colnames(fit$model)]
+  expect_equal(fit$model, dat_fit, info = "dataset is the same after passing within the function", ignore_attr = TRUE)
+})
+
+test_that("sim_ord_latent(link = 'probit') works with a binary x, k = 4", {
+  set.seed(1)
+  k <- 4
+  n <- 1e5
+  probs0 <- c(0.1, 0.3, 0.5, 0.1)
+  alpha <- prob_to_alpha(probs0, link = "probit")
+  b1 <- log(2) # log odds ratio
+  dat <- data.frame(x = rep(0:1, each = n))
[…]

### 2024-01-08 README.Rmd
update

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ordinalsim

<!-- badges: start -->
<!-- badges: end -->

### 2024-01-08 docs/paper/paper-new.Rmd
update

--- 2024-01-06 paper/paper-new.Rmd
@@
     header_includes:
-      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline, bm}
+      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline, mathbf}
 ---
@@
 
+For the tutorial, we assume the reader is already familiar with basic R programming and introductory theory about linear regression.
+(VEDI SE SCRIVERE MEGLIO O QUALCOSA DI PIU).
+The current tutorial proposed a set of custom R functions (available on Github ) to understand and simulate basic ordinal regression models.
+More details and extended examples are available on the supplementary materials.
+
 ## Ordinal regression models
@@
 As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
-@Burkner2019-aw suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.
+@Burkner2019-aw suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical intepretation as for the cumulative vs sequential model.
 
@@
 
-The $\mathbf{X} \boldsymbol{\beta}$ is the linear predictor $\eta$ that is the cumulative probability $P(Y \leq k)$ transformed using the link function $g()$.
+The $\mathbf{X} \boldsymbol{\beta}$ is the linear predictor $\eta$ that is the cumulative probability $P(Y \leq k)$ transformed using the link function $g(\cdot)$.
 To obtain the probability of a single outcome $P(Y = k)$ we can compute the difference between cumulative probabilities as shown in Equation \@ref(eq:prob-cum-model2).
@@
 \begin{equation} 
-Y^\star_i = \eta + \epsilon_i
+Y^\star_i = \mathbf{X}\boldsymbol{\beta} + \epsilon_i
 (\#eq:latent-model)
@@
 In a more sophisticated version of the model called location-shift [@Tutz2022-dg], both the location $\mu$ and the thresholds $\alpha_k$ can vary as a function of the predictors.
-@Tutz2022-dg described also another version of the model called location-scale [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur] where the location $\mu$ and the scale $\sigma^2$ of the distribution can vary as a function of the predictors.
 
@@
 @Gelman2020-tg proposed three alternative parametrizations focusing on different definition of the thresholds.
+
+@Tutz2022-dg described also another version of the model called location-scale [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur] where the location $\mu$ and the scale $\sigma^2$ of the distribution can vary as a function of the predictors.
 
@@
 We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
-Table contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
-The participants are divided into two groups (predictor $x$).
+Table \@ref(tab:tab-dataset-example) contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
+The participants are divided into two groups ($x$).
 We can fit a cumulative link model with `clm()` function and check the model summary.
@@
 Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect.
-As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [see @Schad2020-ht][^3]
-
-[^3]: In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
+As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [see @Schad2020-ht].
+In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
 By default, the first level of the factor is taken as the reference level and the $j - 1$ coefficients represent the comparison between the other levels and the reference.
@@
 
-From the previous example, we noticed that the $k - 1$ odds ratios are very similar (the similarity increase if $n$ increase).
-This is not a coincidence but an implicit (until now) assumption made by the CM called *proportional odds* (POA).
-Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the (POA).
+From the previous example, we noticed that the $k - 1$ odds ratios are different.
+However, increasing $n$ will result in odds ratios that are very similar.
+This does not happen by chance but from an implicit (until now) assumption made by the CM called *proportional odds* (PO).
+Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the PO assumption.
 There are more advanced versions of the model relaxing this assumption completely (*non proportional odds*) and partially (*partial proportional odds*).
-The proportional odds assumption is a crucial point when interpreting model parameters.
-Can be formalized as in Equation ().
-
-Basically if we use the *logit* link function $g()$ the $\beta$s are interpreted as standard odds ratio in the logistic regression.
-Given that we have $k > 2$ alternatives there is the need of $k - 1$ equations.
-The POA is formalized in Equation \@ref(eq:prop-odds).
-Basically the cumulative log odds ratio comparing $P(Y_k|x_0)$ with $P(Y_k|x_1)$ is the same regardless the specific $k$ or threshold $\alpha$.
-
-The proportional odds is assuming that the regression coefficients $\beta$ are independent from the thresholds $\alpha$ thus the effect is the same regardless of the $Y$ level.
-Thus $\beta_1 = \beta_2 \dots = \beta_{k - 1}$.
+
+Basically if we use the *logit* link function $g(\cdot)$, regression coefficients are interpreted as odds ratios.
+Given that we have $k > 2$ alternatives we need $k - 1$ equations.
+The PO assumption is formalized in Equation \@ref(eq:prop-odds).
+The cumulative log odds ratio comparing $P(Y_k|x_0)$ with $P(Y_k|x_1)$ is the same regardless the specific threshold ($\beta_1 = \beta_2 \dots = \beta_{k - 1}$).
 The Figure (\@ref(fig:fig-prop-odds)) depicts the proportional odds assumption for the $k - 1$ logistic curves both for probabilities and linear predictors $\eta$.
@@
 
-The proportional odds assumption is convenient because regardless of $k$, the $\beta_j$ ($j$ being the number of regression coefficients) effect is assumed to be the same.
-The model is more parsimonious compared to estimating $k - 1$ coefficients for each $\beta_j$ as in the multinomial regression or the non-proportional model.
-
-Given the relatively strong assumption, there are several methods for testing if data are supporting the POA [see @Liu2023-bp for an overview].
-@Tutz2020-xq suggested a trade-off between assuming/relaxing the POA in terms of implementing *location-shift* or *location-scale* models.
+The PO assumption is convenient because regardless of $k$, the $\beta_j$ ($j$ being the number of regression coefficients) effect is assumed to be the same.
+The model is more parsimonious compared to estimating $k - 1$ coefficients for each $\beta_j$ as in the multinomial regression or the non-proportional odds model.
+
+At the same time, the PO assumption could be considered too strict.
+There are several methods for testing if data are supporting the PO assumption [see @Liu2023-bp for an overview].
+@Tutz2020-xq suggested a trade-off between assuming/relaxing the PO assumption by fitting *location-shift* or *location-scale* models.
 Basically these methods should guarantee more flexibility in modelling the observed probabilities reducing the number of parameters.
 However, these methods and models are outside the scope of the tutorial.
-In the next section, we provide more details only about the *location-scale* model in terms of parameters interpretation and data simulation.
-
-This can be shown by fitting the model with `clm()` that by default assume the POA and using the predicted probabilities to compute the odds ratios.
+In the supplementary materials, we provide more details only about the *location-scale* model in terms of parameters interpretation and data simulation.
+
+This can be shown by fitting the model with `clm()` that by default assume the PO and using the predicted probabilities to compute the odds ratios.
 In the following code we computed the odds ratio comparing $x_a$ and $x_b$ when $k \leq 1$ and $k \leq 2$.
+The odds ratios are exactly the same because we are using the predicted probabilities from the model (assuming PO).
 
@@
 
-In this case the odds ratio is exactly the same because the model (thus the predicted probabilities) assume the POA.
-When estimating the $k - 1$ odds ratio from the data results could be different.
-In fact, the discrepancy between the odds ratio is an index of POA violation.
-metti qualche riferimento, magari a supplementary.
-
-### Proportional odds and *probit* model
-
-The proportional odds assumption is relevant only for the *logit* model because of the link function that put $\beta$ as log odds.
-However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model.
-Equation \@ref(eq:probit-prop-odds) depict the parallel slopes assumption for a probit model.
-Basically the difference in $z$ for a unit increase in $x$ (i.e., the slope) is the same regardless the $Y$ level.
-This can be directly demonstrated in the same way as for the logit model.
-We use the same dataset of the previous simulation.
-The difference between the $z$ scores comparing the two groups calculated on the cumulative probability is the same regardless the $k - 1$ level.
-
-\begin{equation}
-z_{x_1 - x_0} = \Phi(P \leq 1|x_0) - \Phi(P \leq 1|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
-(\#eq:probit-prop-odds)
-\end{equation}
-
-```{r poa-probit-example}
-#| echo: true
-#| collapse: true
-
-# fitting the model
-fit <- clm(y ~ x, data = dat, link = "probit")
-
-# extracting the predicted probabilities for the two groups
-pr <- predict(fit, data.frame(x = unique(dat$x)))$fit
-
-# y <= 1
-y1a <- pr[1, 1]
-y1b <- pr[2, 1]
-
-# y <= 2
-y12a <- sum(pr[1, 1:2])
-y12b <- sum(pr[2, 1:2])
-
-# z score difference y <= 1 (a vs b) VS y <= 2 (a vs b)
-qnorm(y1a) - qnorm(y1b)
-qnorm(y12a) - qnorm(y12b)
-```
-
 ## *Probit Model* $z$ scores
 
-The alternative to fitting a cumulative logit model is using a standard normal distribution thus fitting a *probit* model.
-The main difference is that the latent distribution is now a Gaussian distribution with $\mu = 0$ and $\sigma^2 = 1$ and the link function is the cumulative Gaussian distribution function $\Phi$ implemented in R with the `pnorm()` function.
-The inverse of the Gaussian cumulative distribution function is $\Phi^{-1}$ implemented in R with the `qnorm()` function.
-The logistic and normal distribution are very similar, with the logistic having more variance.
-
+When assuming a standard Gaussian distribution we are fitting *probit* model.
 The main difference regards the parameters interpretation.
-In the logit model the $\beta$ are the log odds ratio.
+In the logit model the $\boldsymbol{\beta}$ are the log odds ratio.
 For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in $x$.
@@
 
+### Proportional odds and *probit* model
+
+The PO assumption is relevant only for the *logit* model.
+When using a different link function (e.g., *probit*) the assumption of a common $\beta$ is called *parallel slopes* (see Equation \@ref(eq:probit-prop-odds)).
+Basically the difference in $z$ scores for a unit increase in $x$ (i.e., the slope) is the same regardless the threshold.
+As for the previous example, we can fit the *probit* model and manually calculate the difference in $z$ scores.
+
+\begin{equation}
+z_{x_1 - x_0} = \Phi(P \leq 1|x_0) - \Phi(P \leq 1|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
+(\#eq:probit-prop-odds)
+\end{equation}
+
+```{r poa-probit-example}
+#| echo: true
+#| collapse: true
+
+# fitting the model
+fit <- clm(y ~ x, data = dat, link = "probit")
+
+# extracting the predicted probabilities for the two groups
+pr <- predict(fit, data.frame(x = unique(dat$x)))$fit
+
+# y <= 1
+y1a <- pr[1, 1]
+y1b <- pr[2, 1]
+
+# y <= 2
+y12a <- sum(pr[1, 1:2])
+y12b <- sum(pr[2, 1:2])
+
+# z score difference y <= 1 (a vs b) VS y <= 2 (a vs b)
+qnorm(y1a) - qnorm(y1b)
+qnorm(y12a) - qnorm(y12b)
+```
+
 ## Simulating data
@@
 In this tutorial we present two method for simulating ordinal data.
-The first method concerns calculating the probabilities of each $k$ of $Y$ as a function of predictors and then sampling from a multinomial distribution.
-The second method simulate the latent variable $Y^\star$ as a function of predictors.
+The first method calculate the probabilities of each $Y$ level as a function of predictors and generate data from a multinomial distribution.
+The second method simulate data using the the latent formulation.
 
@@
 
-For the first method we need to calculate $g^{-1}(\eta)$ as a function of predictors and then sample from a multinomial distribution^[In R we are using the `sample()` function that simulate values from a *categorical* distribution.
-The *categorical* distribution is considered a special case of the multinomial distribution].
-This is similar to the general method to simulate data for a generalized linear model where the linear predictor $\eta$ are calculated and data are generated from the corresponding probability distribution of the random component.
-
-As a simple example, let's simulate two groups or condition with $n = 100$ participants responding to an item with $k = 4$ ordered options.
-We simulate that the second group has more probability of responding higher categories of $Y$.
-As explained in the previous sections, we can summarise the effect size of a CM (assuming POA) using a single $\beta = \log(OR)$.
+For the first method we need to calculate $g^{-1}(\eta)$ as a function of predictors and then sample from a *multinomial* (more specifically *categorical*) distribution using the `sample()` function in R.
+This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview)
+
+As a simple example, let's simulate two groups with $n = 100$ participants responding to an item with $k = 4$ ordered options.
+We simulate the second group having higher probability of responding higher categories of $Y$.
+
+As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single $\beta = \log(OR)$.
 We can assume that the odds ratio is $OR = 2$.
@@
 k <- 4 # number of ordinal alternatives
-probs0 <- rep(1/k, k) # probabilities for the first group
-alpha <- prob_to_alpha(probs0, link = "logit")
+prob0 <- rep(1/k, k) # probabilities for the first group
+alpha <- prob_to_alpha(prob0, link = "logit")
 dat <- data.frame(x = rep(c(0, 1), each = N/2))
@@
 
-1.  Define simulation parameters, baseline probabilities, etc. as in the previous workflow
-4.  Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
+1.  Define simulation parameters, baseline probabilities (`prob0`), sample size, etc.
+4.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the $k - 1$ linear predictors ($\eta$) using the equations
 5.  Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
-6.  Calculate for each observation the probability of the $k$ outcome as the difference between cumulative probabilities.
-7.  Sample $n$ outcomes from a multinomial distribution, choosing between $k$ alternatives with the calculated probabilities.
+6.  Calculate the probabilities of $k$ outcomes
+7.  Sample $n$ outcomes from a *multinomial* distribution, choosing between $k$ alternatives using the calculated probabilities
 8.  Fit the appropriate model using `ordinal::clm()`
 
-A general suggestion is to increase the sample size $n$ to a very large value reducing the random variability to check if simulated parameters are recovered.
-
 ### Simulating from the latent distribution
 
-An equivalent but more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
+A more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
 This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and the cutting the latent values according to the thresholds $\alpha$.
@@
 
-1.  Define $n$ (number of observations) and $k$ (number of ordinal outcomes)
-2.
-Define the regression coefficients $\beta$ <!-- controlla che le soglie non fanno riferimento a quando è 0 ma concettualmente si.
-essendo che non cambiano non è rilevante a quale x si faccia riferimento -->
-3.  Define the baseline probabilities when $x = 0$. These probabilities can be converted to the corresponding thresholds $\alpha$ choosing the appropriate link function (*logit* or *probit*).
-4.
-Calculate the linear predictors $\eta$ using $k - 1$ equations combining the predictors $\mathbf{X}$, the regression coefficients $\boldsymbol{\beta}$ and the error $\mathbf{\epsilon}$ sampled from the appropriate probability distribution.
-5.
-Cut the latent variable into $k$ areas using the thresholds $\alpha$ and assign the corresponding ordinal value.
-This step simply checks if the latent value $Y^{\star}_i$ is between two threshold values and assign the corresponding value.
-This can be done using the `cut()` or the `findInterval()` functions.
+1.  Define simulation parameters as in the previous simulation. `prob0` are the probabilities when all predictors $\mathbf{X}$ are zero.
+2.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the linear predictor $\eta$ using the Equation \@ref(eq:latent-model)
+3.  Add the random errors $\epsilon_i$ sampling from the chosen distribution (logistic or normal)
+4.  Cut the latent variable into $k$ areas using the thresholds $\alpha$ and assign the corresponding ordinal value. This can be done using the `cut()` or the `findInterval()` functions.
+
+The Figure \@ref(fig:fig-sim-from-latent) depict the simulated $Y^{*}$ and the corresponding ordinal value.
+As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.
 
@@
 
-# calculate the linear predictor using equation TODO put equation
+# calculate the linear predictor using the model equation
 dat$lp <- log(or) * dat$x
@@
 ```
-
-The Figure \@ref(fig:fig-sim-from-latent) depict the simulated $Y^{*}$ and the corresponding ordinal value.
-As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.
 
@@
 
-The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function. Details about the function can be found on the Github repository LINK.
+The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function.
+Details about the function can be found on the Github repository LINK.
+Basically, we define the dataset `dat` with predictors.
+Then the model formula is specified within the function as `location = ~` along with the vector of regression coefficients (`beta`), baseline probabilities (`prob0`) and the link function.
+The function return a dataset with the simulated $Y$ and the latent variable (this is possible only because we simulate the data, otherwise $Y^{\star}$ is not observed by definition).
 
@@
 # same as the previous simulation
-dat <- sim_ord_latent(~x, By = log(or), prob = probs0, data = dat, link = "logit")
+dat <- sim_ord_latent(location = ~x, beta = log(or), prob0 = probs0, data = dat, link = "logit")
 
@@
 
-The previous simulation can be easily extended by adding predictors and their interactions.
+The previous simulation can be easily extended by adding more predictors and their interactions.
 The crucial part is setting appropriate and empirically meaningful parameters.
 Thresholds $\alpha$ are usually not of main interest [but see the location-shift models @Tutz2022-dg] and can be considered as intercepts in standard linear regression.
-The thresholds are quantiles of the latent distribution that produced a certain probability distribution of the ordinal variable.
+The thresholds are quantiles of the latent distribution that produced certain $k$ probabilities.
 To set meaningful $\alpha$ values we can convert probabilities into thresholds (`alpha_to_prob()`).
-The function `show_alpha()` create a meaningful visual representation of using a specific set of thresholds (see Figure \@ref(fig:fig-show-th-example)).
-The function can be used as `show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))`.
+The function `show_alpha()` produce a meaningful visual representation of using a specific set of thresholds (see Figure \@ref(fig:fig-show-th-example)).
 In regression terms, the thresholds determine the probabilities when $x = 0$ (as the intercept).
@@
 
-#### Regression coefficients $\beta$
-
-Regression coefficients are the effect size parameters.
+#### Regression coefficients
+
 For *probit* models we can set the $\beta_j$ to be in standardized (i.e., Cohen's $d$-like) units.
-For a categorical variable as the previous example with the group, $\beta$ is the degree of separation in standard deviation unit between the two latent distributions.
+For a categorical variable as the previous example with the group, $\beta_j$ is the degree of separation in standard deviation unit between the two latent distributions.
 For *logit* models we can set the odds ratio.
@@
 For example, @Sanchez-Meca2003-ji proposed some equations to convert between odds ratios and Cohen's $d$.
-Using their approach, a Cohen's $d$ of 0.5 usually considered a plausible medium effect size corresponds to and odds ratio of $\approx 2.47$.
-
-We can also calculate and plot the predicted probabilities (i.e., $g^{-1}(\eta)$, without actually sampling) given the predictors and the chosen regression coefficients.
+Using their approach, a Cohen's $d = 0.5$ usually considered a plausible medium effect size corresponds to and odds ratio of $\approx 2.47$.
+
+We can also calculate and plot the predicted probabilities (i.e., $g^{-1}(\eta)$) given the predictors and the chosen regression coefficients.
 In this way we can try different values and see if predicted probabilities are plausible or not.
@@
 ```{r fig-example-cat-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
-cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")
+cat_latent_plot(m = c(0, 1), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")
 ```
@@
 ```{r fig-example-num-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
-num_latent_plot(x = runif(100), b1 = 2, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit") +
+num_latent_plot(x = rnorm(100), b1 = 1, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit") +
   mtheme() +
@@
 In this example we have 2 main effects and the interaction.
-Regardless the link function, we can focus the model equation on the linear predictor.
-Equation depicts the linear predictor for a 2x2 interaction.
-With a 2x2 design, a good strategy is to use sum to zero contrasts for the two factors.
-By default R use dummy coding but here we set the contrasts for $X_1$ and $X_2$ as $0.5$ and $-0.5$.
+By default R use dummy coding but setting sum-to-zero contrasts (e.g., $0.5$ and $-0.5$) for a factorial design is convenient.
 In this way $\beta_1$ will be the main effect of $X_1$, $\beta_2$ the main effect of $X_2$ and $\beta_3$ the interaction (thus the difference of differences).
@@
 
-dat <- sim_ord_latent(~ x1 * x2, By = c(0, 1, 0.5), prob = probs0, link = "probit", data = dat)
+dat <- sim_ord_latent(~ x1 * x2, beta = betas, prob0 = probs0, link = "probit", data = dat)
 fit <- clm(y ~ x1 * x2, data = dat, link = "probit")
@@
 The same holds for `x2`.
-`x1:x2` is the difference of differences in $z$ scores (i.e., the interaction).
-Figure \@ref(fig:fig-effects-2-by-2-interaction) depicts the model results.
+`x1:x2` is the interaction thus difference of differences in $z$ scores (see Figure \@ref(fig:fig-effects-2-by-2-interaction)).
 
@@
 
-Another common scenario is to compute and interaction between a numerical variable $x$ and a categorical variable $g$.
+Another common scenario is the interaction between a numerical variable $x$ and a categorical variable $g$.
 For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution between 0 and 1.
@@
 
-dat <- sim_ord_latent(~ x * g, By = c(0.3, 0.5, 0.5), prob = probs0, link = "probit", data = dat)
+dat <- sim_ord_latent(~ x * g, beta = c(0.3, 0.5, 0.5), prob0 = probs0, link = "probit", data = dat)
 fit <- clm(y ~ x * g, data = dat, link = "probit")
@@
 
-Again, `x` is the slope of the numerical predictor averaging over `g` (given that `g` has been coded with sum to zero contrasts) thus the increase in $z$ scores for a unit increase in `x`.
-`g1` is the main effect of the factor evaluated when $x = 0$.
-To change the interpretation of `g1` a possibility is to center `x` differently (e.g., mean centering).
-The `x:g1` is the difference in slopes between the two groups.
+Again, `x` is the slope of the numerical predictor averaging over `g` (given that `g` has been coded with sum-to-zero contrasts) thus the increase in $z$ scores for a unit increase in `x`.
+`g1` is the main effect of the factor evaluated when $x = 0$ (centering $x$ will change the parameter interpretation).
+The `x:g1` is the slopes difference the two groups.
 Figure \@ref(fig:fig-effects-num-by-cat-interaction) depicts the model results.
@@
 Formally the power is defined as the probability of correctly rejecting the null hypothesis $H_0$.
-In some simple case, the power can be computed analytically however using data simulation we can handle every scenario.
-We present very general example that, as for the previous simulations, can be extended and modified.
-
+For simple cases (e.g., a t-test) the power can be calculated analytically while in complex cases simulating data is the best approach.
+We present very general example that, as for the previous simulations, can be extended easily extended.
 For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting $H_0$) or assessing an eventual bias in model estimates.
+
 The general workflow for a power simulation can be summarized as:
@@
 
-To make a practical example, we simulate the power of detecting a group difference of $d = 0.4$ (assuming a *probit* model).
+We can estimate the power of detecting a group difference of $d = 0.4$ (assuming a *probit* model).
 Participants respond to an ordinal variable with $k = 5$.
-The iteration is performed using a `for` loop that repeat 1000 times the data simulation, model fitting and extract the p-value for the $\beta_1$.
+The simulation is performed using a `for` loop that repeat 1000 times the data simulation, model fitting and extract the p-value for the $\beta_1$.
 The power is then estimated as the number of p-values lower than the critical level over the number of simulations.
@@
 for(i in 1:nsim){
-  sim <- sim_ord_latent(~group, By = d, prob = probs0, link = "probit", data = dat)
+  sim <- sim_ord_latent(~group, beta = d, prob0 = probs0, link = "probit", data = dat)
   fit <- clm(y ~ group, data = sim, link = "probit")
@@
   for(j in 1:nsim){
-    sim <- sim_ord_latent(~group, By = d, prob = probs0, link = "probit", data = dat)
+    sim <- sim_ord_latent(~group, beta = d, prob0 = probs0, link = "probit", data = dat)
     fit <- clm(y ~ group, data = sim, link = "probit")
@@
 
-## Disclaimer about the functions
-
-The current paper proposed a simplified way with some functions to generate ordinal data.
-For more complex simulations such as simulating correlated ordinal data the `simstudy` package <https://kgoldfeld.github.io/simstudy/articles/ordinal.html> proposed a very comprehensive set of data simulation function also for ordinal data.
-
 \newpage
@@
 
-(ref:fig-show-th-example) caption
+(ref:fig-show-th-example) `show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))`
 

### 2024-01-08 R/utils-plotting.R

@@ -178,8 +178,7 @@ num_latent_plot <- function(x,
     th <- prob_to_alpha(probs, link)
   }
   data <- data.frame(x = seq(min(x), max(x), length.out = nsample))
-  #data <- get_probs(~x, b1, probs, data, link, append = TRUE)
-  data <- sim_ord_latent(~x, By = b1, prob = probs, data = data, link = link, simulate = FALSE)
+  data <- sim_ord_latent(~x, beta = b1, prob0 = probs, data = data, link = link, simulate = FALSE)
   datal <- tidyr::pivot_longer(data, starts_with("y"), names_to = "y", values_to = "value")
   ggplot(datal, aes(x = x, y = value, color = y)) +
     geom_line(linewidth = linewidth) +

### 2024-01-08 R/utils.R

@@ -23,9 +23,9 @@
 #' dat <- sim_ord_latent(~x, By = 0.5, prob = bprobs, data = dat, link = "probit")
 sim_ord_latent <- function(location,
                            scale = NULL,
-                           By,
-                           Bscale = NULL,
-                           prob = NULL, 
+                           beta,
+                           zeta = NULL,
+                           prob0 = NULL, 
                            alpha = NULL, 
                            data, 
                            link = c("logit", "probit"),
@@ -35,26 +35,26 @@ sim_ord_latent <- function(location,
   lf <- get_link(link)
   if(is.null(alpha)){
     # calculate thresholds if not provided
-    alpha <- prob_to_alpha(prob, link = link)
+    alpha <- prob_to_alpha(prob0, link = link)
   }
   k <- length(alpha) + 1 # number of ordinal outcomes
   n <- nrow(data) # number of observations
   
   # model matrix for the location effect
-  Xy <- model.matrix(location, data = data)[, -1, drop = FALSE] # remove intercept
-  lpy <- c(Xy %*% By) # linear predictor for the location
-  lps <- 0 # default scale effect 0, exp(0) = 1 (the default scale)
+  Xloc <- model.matrix(location, data = data)[, -1, drop = FALSE] # remove intercept
+  lploc <- c(Xloc %*% beta) # linear predictor for the location
+  lpscale <- 0 # default scale effect 0, exp(0) = 1 (the default scale)
   
   # check predictors on the scale parameter
   if(!is.null(scale)){
     # model matrix for the scale effect
-    Xsigma <- model.matrix(scale, data = data)[, -1, drop = FALSE] # remove intercept
-    lps <- c(Xsigma %*% Bscale) # linear predictor for the scale
+    Xscale <- model.matrix(scale, data = data)[, -1, drop = FALSE] # remove intercept
+    lpscale <- c(Xscale %*% zeta) # linear predictor for the scale
   }
   
   if(simulate){
     # latent variable with appropriate error function
-    ystar <- lf$rfun(n, lpy, exp(lps))
+    ystar <- lf$rfun(n, lploc, exp(lpscale))
     
     # cut according to thresholds
     y <- findInterval(ystar, alpha) + 1 # to start from 1
@@ -63,7 +63,7 @@ sim_ord_latent <- function(location,
     data$ys <- ystar # save also the latent
   } else{
[…]

### 2024-01-08 scripts/tables.R

@@ -1,4 +1,5 @@
 library(flextable)
+library(tidyverse)
 devtools::load_all()
 
 # MODEL TABLE
@@ -8,10 +9,10 @@ models_table <- data.frame(
   Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
   Link.Function = c("Equation",
                     "$\\text{logit}(p) = \\text{log}(p / (1-p))$",
-                    "$z = \\Phi(p)$"),
+                    "$z = \\Phi^{-1}(p)$"),
   Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
   Inverse.Link.Function = c("Equation",
-                            "$e^{\\text{logit}(p)} / (1 + e^{\\text{logit}(p)})$","$\\Phi^{-1}(z)$"),
+                            "$e^{\\text{logit}(p)} / (1 + e^{\\text{logit}(p)})$","$\\Phi^(z)$"),
   Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
 )
 
@@ -39,7 +40,7 @@ n <- 100
 x <- rep(c("a", "b"), each = n/2)
 dat <- data.frame(x = x)
 probs <- rep(1/k, k) # for the group "a", uniform probabilities
-dat <- sim_ord_latent(~x, By = b1, prob = probs, data = dat, link = "logit")
+dat <- sim_ord_latent(~x, beta = b1, prob0 = probs, data = dat, link = "logit")
 
 funs <- list(mean = mean, median = median, sd = sd)
 ss <- sapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))

### 2024-01-08 docs/paper/paper.tex
updating structure

--- 2024-01-04 paper/paper-new.tex
@@
 
+For the tutorial, we assume the reader is already familiar with basic R programming and introductory theory about linear regression.
+(VEDI SE SCRIVERE MEGLIO O QUALCOSA DI PIU).
+The current tutorial proposed a set of custom R functions (available on Github ) to understand and simulate basic ordinal regression models.
+More details and extended examples are available on the supplementary materials.
+
 \subsection{Ordinal regression models}\label{ordinal-regression-models}
@@
 As suggested by Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
-Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical distinction as for the cumulative vs sequential model.
+Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical intepretation as for the cumulative vs sequential model.
 
@@
 
-The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g()\).
+The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g(\cdot)\).
 To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:prob-cum-model2}.
@@
 
+\begin{center}\textbf{[Figure 1 about here]} \end{center}
+
 \normalsize
@@
 \begin{equation} 
-Y^\star_i = \eta + \epsilon_i
+Y^\star_i = \mathbf{X}\boldsymbol{\beta} + \epsilon_i
 \label{eq:latent-model}
@@
 In a more sophisticated version of the model called location-shift (Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha_k\) can vary as a function of the predictors.
-Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.
 
@@
 Gelman, Hill, and Vehtari (2020) proposed three alternative parametrizations focusing on different definition of the thresholds.
+
+Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.
 
@@
 \end{figure}
+
+\begin{center}\textbf{[Figure 2 about here]} \end{center}
 
@@
 Cumulative Logit & $\text{logit}(p) = \text{log}(p / (1-p))$ & `qlogis()` & $e^{\text{logit}(p)} / (1 + e^{\text{logit}(p)})$ & `plogis()`\\
-Cumulative Probit & $z = \Phi(p)$ & `qnorm()` & $\Phi^{-1}(z)$ & `pnorm()`\\
+Cumulative Probit & $z = \Phi^{-1}(p)$ & `qnorm()` & $\Phi^(z)$ & `pnorm()`\\
 \bottomrule
@@
 We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
-Table contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options.
-The participants are divided into two groups (predictor \(x\)).
+Table \ref{tab:tab-dataset-example} contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options.
+The participants are divided into two groups (\(x\)).
 We can fit a cumulative link model with \texttt{clm()} function and check the model summary.
@@
 \midrule
-a & 2.42 & 2 & 1.13 & 14 (p = 0.28) & 12 (p = 0.24) & 13 (p = 0.26) & 11 (p = 0.22)\\
-b & 3 & 3 & 1.16 & 9 (p = 0.18) & 6 (p = 0.12) & 11 (p = 0.22) & 24 (p = 0.48)\\
+a & 2.84 & 3 & 1.11 & 8 (p = 0.16) & 11 (p = 0.22) & 12 (p = 0.24) & 19 (p = 0.38)\\
+b & 3.02 & 3 & 1.08 & 5 (p = 0.10) & 13 (p = 0.26) & 8 (p = 0.16) & 24 (p = 0.48)\\
 \bottomrule
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  100  {-}132.25 272.51 4(0)  1.08e{-}08 2.0e+01}
+\CommentTok{\#\textgreater{}  logit flexible  100  {-}128.88 265.77 5(0)  9.09e{-}13 2.2e+01}
 \CommentTok{\#\textgreater{} }
 \CommentTok{\#\textgreater{} Coefficients:}
-\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)   }
-\CommentTok{\#\textgreater{} xb   0.9587     0.3718   2.579  0.00992 **}
-\CommentTok{\#\textgreater{} {-}{-}{-}}
-\CommentTok{\#\textgreater{} Signif.
-codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
+\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)}
+\CommentTok{\#\textgreater{} xb   0.3153     0.3670   0.859     0.39}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}0.8103     0.2814  {-}2.879}
-\CommentTok{\#\textgreater{} 2|3   0.0672     0.2636   0.255}
-\CommentTok{\#\textgreater{} 3|4   1.1158     0.2900   3.848}
+\CommentTok{\#\textgreater{} 1|2  {-}1.7503     0.3432  {-}5.100}
+\CommentTok{\#\textgreater{} 2|3  {-}0.3791     0.2722  {-}1.393}
+\CommentTok{\#\textgreater{} 3|4   0.4381     0.2717   1.612}
 \end{Highlighting}
@@
 Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect.
-As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding {[}see Schad, Vasishth, Hohenstein, and Kliegl (2020){]}\footnote{In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
-By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.}
+As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding (see Schad, Vasishth, Hohenstein, \& Kliegl, 2020).
+In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
+By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.
 
@@
 
+\begin{center}\textbf{[Figure 3 about here]} \end{center}
+
 \normalsize
@@
 \NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\CommentTok{\#\textgreater{} [1] 0.28}
+\CommentTok{\#\textgreater{} [1] 0.16}
 
@@
 \FunctionTok{odds}\NormalTok{(p1a)}
-\CommentTok{\#\textgreater{} [1] 0.3888889}
+\CommentTok{\#\textgreater{} [1] 0.1904762}
 
@@
 \NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\CommentTok{\#\textgreater{} [1] 0.18}
+\CommentTok{\#\textgreater{} [1] 0.1}
 \FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
-\CommentTok{\#\textgreater{} [1] 1.771605}
+\CommentTok{\#\textgreater{} [1] 1.714286}
 \end{Highlighting}
@@
 
-For example the probability of responding \(Y = 1\) in the group ``a'' is 0.28 corresponding to an odds of 0.39.
-When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 1.77 times the odds of responding \(Y = 1\) compared to the group ``b''.
+For example the probability of responding \(Y = 1\) in the group ``a'' is 0.16 corresponding to an odds of 0.19.
+When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 1.71 times the odds of responding \(Y = 1\) compared to the group ``b''.
 
@@
 Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\).
-The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 2.61.
+The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 1.37.
 In fact, the logit model essentially estimate the cumulative odds ratio.
@@
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.28    0.52    0.78}
+\CommentTok{\#\textgreater{}    0.16    0.38    0.62}
 \NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.18    0.30    0.52}
+\CommentTok{\#\textgreater{}    0.10    0.36    0.52}
 
@@
 \CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
-\CommentTok{\#\textgreater{} 1.771605 2.527778 3.272727}
+\CommentTok{\#\textgreater{} 1.714286 1.089606 1.506073}
 \end{Highlighting}
@@
 
-From the previous example, we noticed that the \(k - 1\) odds ratios are very similar (the similarity increase if \(n\) increase).
-This is not a coincidence but an implicit (until now) assumption made by the CM called \emph{proportional odds} (POA).
-Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the (POA).
+From the previous example, we noticed that the \(k - 1\) odds ratios are different.
+However, increasing \(n\) will result in odds ratios that are very similar.
+This does not happen by chance but from an implicit (until now) assumption made by the CM called \emph{proportional odds} (PO).
+Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the PO assumption.
 There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}).
-The proportional odds assumption is a crucial point when interpreting model parameters.
-Can be formalized as in Equation ().
-
-Basically if we use the \emph{logit} link function \(g()\) the \(\beta\)s are interpreted as standard odds ratio in the logistic regression.
-Given that we have \(k > 2\) alternatives there is the need of \(k - 1\) equations.
-The POA is formalized in Equation \eqref{eq:prop-odds}.
-Basically the cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) is the same regardless the specific \(k\) or threshold \(\alpha\).
-
-The proportional odds is assuming that the regression coefficients \(\beta\) are independent from the thresholds \(\alpha\) thus the effect is the same regardless of the \(Y\) level.
-Thus \(\beta_1 = \beta_2 \dots = \beta_{k - 1}\).
+
+Basically if we use the \emph{logit} link function \(g(\cdot)\), regression coefficients are interpreted as odds ratios.
+Given that we have \(k > 2\) alternatives we need \(k - 1\) equations.
+The PO assumption is formalized in Equation \eqref{eq:prop-odds}.
+The cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) is the same regardless the specific threshold (\(\beta_1 = \beta_2 \dots = \beta_{k - 1}\)).
 The Figure (\ref{fig:fig-prop-odds}) depicts the proportional odds assumption for the \(k - 1\) logistic curves both for probabilities and linear predictors \(\eta\).
@@
 
-\normalsize
-
-The proportional odds assumption is convenient because regardless of \(k\), the \(\beta_j\) (\(j\) being the number of regression coefficients) effect is assumed to be the same.
-The model is more parsimonious compared to estimating \(k - 1\) coefficients for each \(\beta_j\) as in the multinomial regression or the non-proportional model.
-
-Given the relatively strong assumption, there are several methods for testing if data are supporting the POA (see Liu, He, Tu, \& Tang, 2023 for an overview).
-Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the POA in terms of implementing \emph{location-shift} or \emph{location-scale} models.
+\begin{center}\textbf{[Figure 4 about here]} \end{center}
+
+\normalsize
+
+The PO assumption is convenient because regardless of \(k\), the \(\beta_j\) (\(j\) being the number of regression coefficients) effect is assumed to be the same.
+The model is more parsimonious compared to estimating \(k - 1\) coefficients for each \(\beta_j\) as in the multinomial regression or the non-proportional odds model.
+
+At the same time, the PO assumption could be considered too strict.
+There are several methods for testing if data are supporting the PO assumption (see Liu, He, Tu, \& Tang, 2023 for an overview).
+Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the PO assumption by fitting \emph{location-shift} or \emph{location-scale} models.
 Basically these methods should guarantee more flexibility in modelling the observed probabilities reducing the number of parameters.
 However, these methods and models are outside the scope of the tutorial.
-In the next section, we provide more details only about the \emph{location-scale} model in terms of parameters interpretation and data simulation.
-
-This can be shown by fitting the model with \texttt{clm()} that by default assume the POA and using the predicted probabilities to compute the odds ratios.
+In the supplementary materials, we provide more details only about the \emph{location-scale} model in terms of parameters interpretation and data simulation.
+
+This can be shown by fitting the model with \texttt{clm()} that by default assume the PO and using the predicted probabilities to compute the odds ratios.
 In the following code we computed the odds ratio comparing \(x_a\) and \(x_b\) when \(k \leq 1\) and \(k \leq 2\).
+The odds ratios are exactly the same because we are using the predicted probabilities from the model (assuming PO).
 
@@
 \FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 2.608304}
+\CommentTok{\#\textgreater{} [1] 1.370663}
 \FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 2.608304}
+\CommentTok{\#\textgreater{} [1] 1.370663}
 \end{Highlighting}
@@
 
-In this case the odds ratio is exactly the same because the model (thus the predicted probabilities) assume the POA.
-When estimating the \(k - 1\) odds ratio from the data results could be different.
-In fact, the discrepancy between the odds ratio is an index of POA violation.
-metti qualche riferimento, magari a supplementary.
-
-\subsubsection{\texorpdfstring{Proportional odds and \emph{probit} model}{Proportional odds and probit model}}\label{proportional-odds-and-probit-model}
-
-The proportional odds assumption is relevant only for the \emph{logit} model because of the link function that put \(\beta\) as log odds.
-However the concept of the proportional odds that is more generally called parallel slopes assumption is assumed also for the probit model.
-Equation \eqref{eq:probit-prop-odds} depict the parallel slopes assumption for a probit model.
-Basically the difference in \(z\) for a unit increase in \(x\) (i.e., the slope) is the same regardless the \(Y\) level.
-This can be directly demonstrated in the same way as for the logit model.
-We use the same dataset of the previous simulation.
-The difference between the \(z\) scores comparing the two groups calculated on the cumulative probability is the same regardless the \(k - 1\) level.
-
-\begin{equation}
-z_{x_1 - x_0} = \Phi(P \leq k|x_0) - \Phi(P \leq k|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
-\label{eq:probit-prop-odds}
-\end{equation}
-
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# fitting the model}
-\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
-
-\CommentTok{\# extracting the predicted probabilities for the two groups}
-\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x)))}\SpecialCharTok{$}\NormalTok{fit}
-
-\CommentTok{\# y \textless{}= 1}
-\NormalTok{y1a }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]}
-\NormalTok{y1b }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]}
-
-\CommentTok{\# y \textless{}= 2}
-\NormalTok{y12a }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
-\NormalTok{y12b }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
-
-\CommentTok{\# z score difference y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
-\FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 0.5644823}
-\FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 0.5644823}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
 \subsection{\texorpdfstring{\emph{Probit Model} \(z\) scores}{Probit Model z scores}}\label{probit-model-z-scores}
 
-The alternative to fitting a cumulative logit model is using a standard normal distribution thus fitting a \emph{probit} model.
-The main difference is that the latent distribution is now a Gaussian distribution with \(\mu = 0\) and \(\sigma^2 = 1\) and the link function is the cumulative Gaussian distribution function \(\Phi\) implemented in R with the \texttt{pnorm()} function.
-The inverse of the Gaussian cumulative distribution function is \(\Phi^{-1}\) implemented in R with the \texttt{qnorm()} function.
-The logistic and normal distribution are very similar, with the logistic having more variance.
-
+When assuming a standard Gaussian distribution we are fitting \emph{probit} model.
 The main difference regards the parameters interpretation.
-In the logit model the \(\beta\) are the log odds ratio.
+In the logit model the \(\boldsymbol{\beta}\) are the log odds ratio.
 For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in \(x\).
@@
 
+\subsubsection{\texorpdfstring{Proportional odds and \emph{probit} model}{Proportional odds and probit model}}\label{proportional-odds-and-probit-model}
+
+The PO assumption is relevant only for the \emph{logit} model.
+When using a different link function (e.g., \emph{probit}) the assumption of a common \(\beta\) is called \emph{parallel slopes} (see Equation \eqref{eq:probit-prop-odds}).
+Basically the difference in \(z\) scores for a unit increase in \(x\) (i.e., the slope) is the same regardless the threshold.
+As for the previous example, we can fit the \emph{probit} model and manually calculate the difference in \(z\) scores.
+
+\begin{equation}
+z_{x_1 - x_0} = \Phi(P \leq 1|x_0) - \Phi(P \leq 1|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
+\label{eq:probit-prop-odds}
+\end{equation}
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# fitting the model}
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
+
+\CommentTok{\# extracting the predicted probabilities for the two groups}
+\NormalTok{pr }\OtherTok{\textless{}{-}} \FunctionTok{predict}\NormalTok{(fit, }\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{unique}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{x)))}\SpecialCharTok{$}\NormalTok{fit}
+
+\CommentTok{\# y \textless{}= 1}
+\NormalTok{y1a }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{]}
+\NormalTok{y1b }\OtherTok{\textless{}{-}}\NormalTok{ pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{]}
+
+\CommentTok{\# y \textless{}= 2}
+\NormalTok{y12a }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{1}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
+\NormalTok{y12b }\OtherTok{\textless{}{-}} \FunctionTok{sum}\NormalTok{(pr[}\DecValTok{2}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{])}
+
+\CommentTok{\# z score difference y \textless{}= 1 (a vs b) VS y \textless{}= 2 (a vs b)}
+\FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
+\CommentTok{\#\textgreater{} [1] 0.204808}
+\FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
+\CommentTok{\#\textgreater{} [1] 0.204808}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
 \subsection{Simulating data}\label{simulating-data}
@@
 In this tutorial we present two method for simulating ordinal data.
-The first method concerns calculating the probabilities of each \(k\) of \(Y\) as a function of predictors and then sampling from a multinomial distribution.
-The second method simulate the latent variable \(Y^\star\) as a function of predictors.
+The first method calculate the probabilities of each \(Y\) level as a function of predictors and generate data from a multinomial distribution.
+The second method simulate data using the the latent formulation.
 
@@
 
-For the first method we need to calculate \(g^{-1}(\eta)\) as a function of predictors and then sample from a multinomial distribution\footnote{In R we are using the \texttt{sample()} function that simulate values from a \emph{categorical} distribution.
-The \emph{categorical} distribution is considered a special case of the multinomial distribution}.
-This is similar to the general method to simulate data for a generalized linear model where the linear predictor \(\eta\) are calculated and data are generated from the corresponding probability distribution of the random component.
-
-As a simple example, let's simulate two groups or condition with \(n = 100\) participants responding to an item with \(k = 4\) ordered options.
-We simulate that the second group has more probability of responding higher categories of \(Y\).
-As explained in the previous sections, we can summarise the effect size of a CM (assuming POA) using a single \(\beta = \log(OR)\).
+For the first method we need to calculate \(g^{-1}(\eta)\) as a function of predictors and then sample from a \emph{multinomial} (more specifically \emph{categorical}) distribution using the \texttt{sample()} function in R.
+This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview)
+
+As a simple example, let's simulate two groups with \(n = 100\) participants responding to an item with \(k = 4\) ordered options.
+We simulate the second group having higher probability of responding higher categories of \(Y\).
+
+As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single \(\beta = \log(OR)\).
 We can assume that the odds ratio is \(OR = 2\).
@@
 \NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{4} \CommentTok{\# number of ordinal alternatives}
-\NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k) }\CommentTok{\# probabilities for the first group}
-\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs0, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\NormalTok{prob0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k) }\CommentTok{\# probabilities for the first group}
+\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(prob0, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
 \NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{each =}\NormalTok{ N}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
@@
 \CommentTok{\#\textgreater{} 2 0 4}
-\CommentTok{\#\textgreater{} 3 0 2}
-\CommentTok{\#\textgreater{} 4 0 4}
-\CommentTok{\#\textgreater{} 5 0 2}
+\CommentTok{\#\textgreater{} 3 0 4}
+\CommentTok{\#\textgreater{} 4 0 2}
+\CommentTok{\#\textgreater{} 5 0 1}
 \CommentTok{\#\textgreater{} 6 0 1}
@@
 \CommentTok{\#\textgreater{}        1    2    3    4}
-\CommentTok{\#\textgreater{}   0 0.42 0.22 0.20 0.16}
-\CommentTok{\#\textgreater{}   1 0.14 0.08 0.34 0.44}
+\CommentTok{\#\textgreater{}   0 0.24 0.32 0.12 0.32}
+\CommentTok{\#\textgreater{}   1 0.10 0.28 0.26 0.36}
 
@@
 \CommentTok{\#\textgreater{}        0    1}
-\CommentTok{\#\textgreater{}   1 0.42 0.14}
-\CommentTok{\#\textgreater{}   2 0.64 0.22}
-\CommentTok{\#\textgreater{}   3 0.84 0.56}
+\CommentTok{\#\textgreater{}   1 0.24 0.10}
+\CommentTok{\#\textgreater{}   2 0.56 0.38}
+\CommentTok{\#\textgreater{}   3 0.68 0.64}
 \CommentTok{\#\textgreater{}   4 1.00 1.00}
@@
 \CommentTok{\#\textgreater{}        1        2        3        4 }
-\CommentTok{\#\textgreater{} 4.448276 6.303030 4.125000      NaN}
+\CommentTok{\#\textgreater{} 2.842105 2.076555 1.195312      NaN}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}126.98 261.96 9(0)  1.80e{-}11 2.1e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}133.18 274.37 7(0)  9.13e{-}08 2.1e+01}
 \CommentTok{\#\textgreater{} }
 \CommentTok{\#\textgreater{} Coefficients:}
-\CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x   0.9373     0.2275    4.12 3.78e{-}05 ***}
-\CommentTok{\#\textgreater{} {-}{-}{-}}
-\CommentTok{\#\textgreater{} Signif.
-codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
+\CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)}
+\CommentTok{\#\textgreater{} x   0.3518     0.2190   1.606    0.108}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}0.1879     0.1683  {-}1.117}
-\CommentTok{\#\textgreater{} 2|3   0.2706     0.1697   1.594}
-\CommentTok{\#\textgreater{} 3|4   1.0595     0.1893   5.596}
+\CommentTok{\#\textgreater{} 1|2  {-}0.7949     0.1794  {-}4.431}
+\CommentTok{\#\textgreater{} 2|3   0.1063     0.1693   0.628}
+\CommentTok{\#\textgreater{} 3|4   0.5994     0.1740   3.445}
 \end{Highlighting}
@@
 \item
-  Define simulation parameters, baseline probabilities, etc. as in the previous workflow
-\item
-  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
+  Define simulation parameters, baseline probabilities (\texttt{prob0}), sample size, etc.
+\item
+  Define regression coefficients \(\boldsymbol{\beta}\) and calculate the \(k - 1\) linear predictors (\(\eta\)) using the equations
 \item
@@
 \item
-  Calculate for each observation the probability of the \(k\) outcome as the difference between cumulative probabilities.
-\item
-  Sample \(n\) outcomes from a multinomial distribution, choosing between \(k\) alternatives with the calculated probabilities.
+  Calculate the probabilities of \(k\) outcomes
+\item
+  Sample \(n\) outcomes from a \emph{multinomial} distribution, choosing between \(k\) alternatives using the calculated probabilities
 \item
@@
 
-A general suggestion is to increase the sample size \(n\) to a very large value reducing the random variability to check if simulated parameters are recovered.
-
 \subsubsection{Simulating from the latent distribution}\label{simulating-from-the-latent-distribution}
 
-An equivalent but more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
+A more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
 This require simulating a standard linear regression using the appropriate data generation function (\emph{logistic} or \emph{normal}) and the cutting the latent values according to the thresholds \(\alpha\).
@@
 \item
-  Define \(n\) (number of observations) and \(k\) (number of ordinal outcomes)
-\item
-  Define the regression coefficients \(\beta\)
-\item
-  Define the baseline probabilities when \(x = 0\).
-These probabilities can be converted to the corresponding thresholds \(\alpha\) choosing the appropriate link function (\emph{logit} or \emph{probit}).
-\item
-  Calculate the linear predictors \(\eta\) using \(k - 1\) equations combining the predictors \(\mathbf{X}\), the regression coefficients \(\boldsymbol{\beta}\) and the error \(\mathbf{\epsilon}\) sampled from the appropriate probability distribution.
+  Define simulation parameters as in the previous simulation. \texttt{prob0} are the probabilities when all predictors \(\mathbf{X}\) are zero.
+\item
+  Define regression coefficients \(\boldsymbol{\beta}\) and calculate the linear predictor \(\eta\) using the Equation \eqref{eq:latent-model}
+\item
+  Add the random errors \(\epsilon_i\) sampling from the chosen distribution (logistic or normal)
 \item
   Cut the latent variable into \(k\) areas using the thresholds \(\alpha\) and assign the corresponding ordinal value.
-This step simply checks if the latent value \(Y^{\star}_i\) is between two threshold values and assign the corresponding value.
 This can be done using the \texttt{cut()} or the \texttt{findInterval()} functions.
 \end{enumerate}
+
+The Figure \ref{fig:fig-sim-from-latent} depict the simulated \(Y^{*}\) and the corresponding ordinal value.
+As for the previous simulation we can fit the model using \texttt{clm()} and check the estimated parameters.
 
@@
 
-\CommentTok{\# calculate the linear predictor using equation }\AlertTok{TODO}\CommentTok{ put equation}
+\CommentTok{\# calculate the linear predictor using the model equation}
 \NormalTok{dat}\SpecialCharTok{$}\NormalTok{lp }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(or) }\SpecialCharTok{*}\NormalTok{ dat}\SpecialCharTok{$}\NormalTok{x}
@@
 \CommentTok{\#\textgreater{}      x       lp      ystar y}
-\CommentTok{\#\textgreater{} 1    0 0.000000 {-}0.63265006 2}
-\CommentTok{\#\textgreater{} 2    0 0.000000  0.96340682 3}
-\CommentTok{\#\textgreater{} 3    0 0.000000  0.95041920 3}
-\CommentTok{\#\textgreater{} 998  1 1.386294  0.07772189 3}
-\CommentTok{\#\textgreater{} 999  1 1.386294  1.09468340 3}
-\CommentTok{\#\textgreater{} 1000 1 1.386294  1.91057532 4}
+\CommentTok{\#\textgreater{} 1    0 0.000000 {-}1.4897139 1}
+\CommentTok{\#\textgreater{} 2    0 0.000000 {-}0.9811171 2}
+\CommentTok{\#\textgreater{} 3    0 0.000000 {-}0.2015837 2}
+\CommentTok{\#\textgreater{} 998  1 1.386294  1.0619636 3}
+\CommentTok{\#\textgreater{} 999  1 1.386294  0.9690069 3}
+\CommentTok{\#\textgreater{} 1000 1 1.386294 {-}1.8056165 1}
 \end{Highlighting}
@@
 
-The Figure \ref{fig:fig-sim-from-latent} depict the simulated \(Y^{*}\) and the corresponding ordinal value.
-As for the previous simulation we can fit the model using \texttt{clm()} and check the estimated parameters.
-
 \scriptsize
@@
 
-\caption{ }\label{fig:fig-sim-from-latent}
+\caption{caption}\label{fig:fig-sim-from-latent}
 \end{figure}
+
+\begin{center}\textbf{[Figure 5 about here]} \end{center}
 
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik   AIC     niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1284.98 2577.96 4(0)  1.51e{-}08 1.8e+01}
+\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1255.36 2518.73 4(0)  3.26e{-}08 1.6e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x   1.1905     0.1201   9.914   \textless{}2e{-}16 ***}
+\CommentTok{\#\textgreater{} x   1.3403     0.1217   11.02   \textless{}2e{-}16 ***}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}1.02783    0.09358 {-}10.984}
-\CommentTok{\#\textgreater{} 2|3  0.02280    0.08403   0.271}
-\CommentTok{\#\textgreater{} 3|4  1.06892    0.09184  11.639}
+\CommentTok{\#\textgreater{} 1|2 {-}1.11938    0.09713 {-}11.524}
+\CommentTok{\#\textgreater{} 2|3 {-}0.03563    0.08586  {-}0.415}
+\CommentTok{\#\textgreater{} 3|4  1.10228    0.09360  11.777}
 \end{Highlighting}
@@
 
-The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function. Details about the function can be found on the Github repository LINK.
+The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function.
+Details about the function can be found on the Github repository LINK.
+Basically, we define the dataset \texttt{dat} with predictors.
+Then the model formula is specified within the function as \texttt{location\ =\ \textasciitilde{}} along with the vector of regression coefficients (\texttt{beta}), baseline probabilities (\texttt{prob0}) and the link function.
+The function return a dataset with the simulated \(Y\) and the latent variable (this is possible only because we simulate the data, otherwise \(Y^{\star}\) is not observed by definition).
 
@@
 \CommentTok{\# same as the previous simulation}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{By =} \FunctionTok{log}\NormalTok{(or), }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\AttributeTok{location =} \SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{beta =} \FunctionTok{log}\NormalTok{(or), }\AttributeTok{prob0 =}\NormalTok{ probs0, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
 
@@
 \CommentTok{\#\textgreater{}     x y         ys}
-\CommentTok{\#\textgreater{} 1   0 1 {-}1.1479983}
-\CommentTok{\#\textgreater{} 2   0 2 {-}0.7163737}
-\CommentTok{\#\textgreater{} 3   0 4  1.3481517}
-\CommentTok{\#\textgreater{} 98  1 3  0.1655840}
-\CommentTok{\#\textgreater{} 99  1 4  1.2106967}
-\CommentTok{\#\textgreater{} 100 1 3  0.6463900}
+\CommentTok{\#\textgreater{} 1   0 3  0.1445982}
+\CommentTok{\#\textgreater{} 2   0 1 {-}1.5179018}
+\CommentTok{\#\textgreater{} 3   0 4  1.5222026}
+\CommentTok{\#\textgreater{} 98  1 4  3.3201003}
+\CommentTok{\#\textgreater{} 99  1 4  2.6305353}
+\CommentTok{\#\textgreater{} 100 1 2 {-}0.1008892}
 \end{Highlighting}
@@
 
-The previous simulation can be easily extended by adding predictors and their interactions.
+The previous simulation can be easily extended by adding more predictors and their interactions.
 The crucial part is setting appropriate and empirically meaningful parameters.
 Thresholds \(\alpha\) are usually not of main interest (but see the location-shift models Tutz, 2022) and can be considered as intercepts in standard linear regression.
-The thresholds are quantiles of the latent distribution that produced a certain probability distribution of the ordinal variable.
+The thresholds are quantiles of the latent distribution that produced certain \(k\) probabilities.
 To set meaningful \(\alpha\) values we can convert probabilities into thresholds (\texttt{alpha\_to\_prob()}).
-The function \texttt{show\_alpha()} create a meaningful visual representation of using a specific set of thresholds (see Figure \ref{fig:fig-show-th-example}).
-The function can be used as \texttt{show\_alpha(prob\ =\ rep(1/4,\ 4),\ link\ =\ "probit",\ plot\ =\ c("latent",\ "probs"))}.
+The function \texttt{show\_alpha()} produce a meaningful visual representation of using a specific set of thresholds (see Figure \ref{fig:fig-show-th-example}).
 In regression terms, the thresholds determine the probabilities when \(x = 0\) (as the intercept).
@@
 
-\caption{ }\label{fig:fig-show-th-example-1}
+\caption{\texttt{show\_alpha(prob\ =\ rep(1/4,\ 4),\ link\ =\ "probit",\ plot\ =\ c("latent",\ "probs"))}}\label{fig:fig-show-th-example}
 \end{figure}
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/fig-show-th-example-2}
-
-}
-
-\caption{ }\label{fig:fig-show-th-example-2}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics{paper-new_files/figure-latex/fig-show-th-example-3}
-
-}
-
-\caption{ }\label{fig:fig-show-th-example-3}
-\end{figure}
-
-\normalsize
-
-\paragraph{\texorpdfstring{Regression coefficients \(\beta\)}{Regression coefficients \textbackslash beta}}\label{regression-coefficients-beta}
-
-Regression coefficients are the effect size parameters.
+
+\begin{center}\textbf{[Figure 6 about here]} \end{center}
+
+\normalsize
+
+\paragraph{Regression coefficients}\label{regression-coefficients}
+
 For \emph{probit} models we can set the \(\beta_j\) to be in standardized (i.e., Cohen's \(d\)-like) units.
-For a categorical variable as the previous example with the group, \(\beta\) is the degree of separation in standard deviation unit between the two latent distributions.
+For a categorical variable as the previous example with the group, \(\beta_j\) is the degree of separation in standard deviation unit between the two latent distributions.
 For \emph{logit} models we can set the odds ratio.
@@
 For example, Sánchez-Meca, Marín-Martínez, and Chacón-Moscoso (2003) proposed some equations to convert between odds ratios and Cohen's \(d\).
-Using their approach, a Cohen's \(d\) of 0.5 usually considered a plausible medium effect size corresponds to and odds ratio of \(\approx 2.47\).
-
-We can also calculate and plot the predicted probabilities (i.e., \(g^{-1}(\eta)\), without actually sampling) given the predictors and the chosen regression coefficients.
+Using their approach, a Cohen's \(d = 0.5\) usually considered a plausible medium effect size corresponds to and odds ratio of \(\approx 2.47\).
+
+We can also calculate and plot the predicted probabilities (i.e., \(g^{-1}(\eta)\)) given the predictors and the chosen regression coefficients.
 In this way we can try different values and see if predicted probabilities are plausible or not.
@@
 
+\begin{center}\textbf{[Figure 7 about here]} \end{center}
+
 \normalsize
@@
 
+\begin{center}\textbf{[Figure 8 about here]} \end{center}
+
 \normalsize
@@
 In this example we have 2 main effects and the interaction.
-Regardless the link function, we can focus the model equation on the linear predictor.
-Equation depicts the linear predictor for a 2x2 interaction.
-With a 2x2 design, a good strategy is to use sum to zero contrasts for the two factors.
-By default R use dummy coding but here we set the contrasts for \(X_1\) and \(X_2\) as \(0.5\) and \(-0.5\).
+By default R use dummy coding but setting sum-to-zero contrasts (e.g., \(0.5\) and \(-0.5\)) for a factorial design is convenient.
 In this way \(\beta_1\) will be the main effect of \(X_1\), \(\beta_2\) the main effect of \(X_2\) and \(\beta_3\) the interaction (thus the difference of differences).
@@
 
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ x1 }\SpecialCharTok{*}\NormalTok{ x2, }\AttributeTok{By =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ x1 }\SpecialCharTok{*}\NormalTok{ x2, }\AttributeTok{beta =}\NormalTok{ betas, }\AttributeTok{prob0 =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
 \NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x1 }\SpecialCharTok{*}\NormalTok{ x2, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  400  {-}401.15 812.29 4(0)  2.53e{-}07 2.1e+01}
+\CommentTok{\#\textgreater{}  probit flexible  400  {-}421.76 853.51 5(0)  6.60e{-}13 2.3e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}         Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x11      0.11896    0.11514   1.033    0.302    }
-\CommentTok{\#\textgreater{} x21      1.00637    0.11746   8.568   \textless{}2e{-}16 ***}
-\CommentTok{\#\textgreater{} x11:x21  0.09606    0.23022   0.417    0.676    }
+\CommentTok{\#\textgreater{} x11      0.06325    0.11371   0.556   0.5781    }
+\CommentTok{\#\textgreater{} x21      0.61698    0.11444   5.391    7e{-}08 ***}
+\CommentTok{\#\textgreater{} x11:x21  0.57404    0.22768   2.521   0.0117 *  }
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.53497    0.06870  {-}7.787}
-\CommentTok{\#\textgreater{} 2|3  0.43367    0.06783   6.393}
+\CommentTok{\#\textgreater{} 1|2 {-}0.43993    0.06606  {-}6.659}
+\CommentTok{\#\textgreater{} 2|3  0.45458    0.06642   6.844}
 \end{Highlighting}
@@
 The same holds for \texttt{x2}.
-\texttt{x1:x2} is the difference of differences in \(z\) scores (i.e., the interaction).
-Figure \ref{fig:fig-effects-2-by-2-interaction} depicts the model results.
+\texttt{x1:x2} is the interaction thus difference of differences in \(z\) scores (see Figure \ref{fig:fig-effects-2-by-2-interaction}).
 
@@
 
+\begin{center}\textbf{[Figure 9 about here]} \end{center}
+
 \normalsize
@@
 
-Another common scenario is to compute and interaction between a numerical variable \(x\) and a categorical variable \(g\).
+Another common scenario is the interaction between a numerical variable \(x\) and a categorical variable \(g\).
 For simplicity we simulate the factor with two levels and the numerical variable sampled from an uniform distribution between 0 and 1.
@@
 
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ x }\SpecialCharTok{*}\NormalTok{ g, }\AttributeTok{By =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.3}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{ x }\SpecialCharTok{*}\NormalTok{ g, }\AttributeTok{beta =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.3}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\AttributeTok{prob0 =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
 \NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x }\SpecialCharTok{*}\NormalTok{ g, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}95.71 201.41 5(0)  2.95e{-}08 1.0e+02}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}97.98 205.97 6(0)  5.36e{-}13 1.1e+02}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}      Estimate Std. Error z value Pr(\textgreater{}|z|)}
-\CommentTok{\#\textgreater{} x      0.4754     0.4198   1.132    0.257}
-\CommentTok{\#\textgreater{} g1     0.3572     0.4483   0.797    0.426}
-\CommentTok{\#\textgreater{} x:g1   1.2149     0.8401   1.446    0.148}
+\CommentTok{\#\textgreater{} x     0.03305    0.40738   0.081    0.935}
+\CommentTok{\#\textgreater{} g1    0.30569    0.46246   0.661    0.509}
+\CommentTok{\#\textgreater{} x:g1  1.01724    0.81525   1.248    0.212}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}0.5090     0.2356  {-}2.160}
-\CommentTok{\#\textgreater{} 2|3   0.2431     0.2321   1.047}
+\CommentTok{\#\textgreater{} 1|2 {-}0.65333    0.24337  {-}2.685}
+\CommentTok{\#\textgreater{} 2|3  0.04236    0.23666   0.179}
 \end{Highlighting}
@@
 
-Again, \texttt{x} is the slope of the numerical predictor averaging over \texttt{g} (given that \texttt{g} has been coded with sum to zero contrasts) thus the increase in \(z\) scores for a unit increase in \texttt{x}.
-\texttt{g1} is the main effect of the factor evaluated when \(x = 0\).
-To change the interpretation of \texttt{g1} a possibility is to center \texttt{x} differently (e.g., mean centering).
-The \texttt{x:g1} is the difference in slopes between the two groups.
+Again, \texttt{x} is the slope of the numerical predictor averaging over \texttt{g} (given that \texttt{g} has been coded with sum-to-zero contrasts) thus the increase in \(z\) scores for a unit increase in \texttt{x}.
+\texttt{g1} is the main effect of the factor evaluated when \(x = 0\) (centering \(x\) will change the parameter interpretation).
+The \texttt{x:g1} is the slopes difference the two groups.
 Figure \ref{fig:fig-effects-num-by-cat-interaction} depicts the model results.
@@
 
+\begin{center}\textbf{[Figure 10 about here]} \end{center}
+
 \normalsize
@@
 Formally the power is defined as the probability of correctly rejecting the null hypothesis \(H_0\).
-In some simple case, the power can be computed analytically however using data simulation we can handle every scenario.
-We present very general example that, as for the previous simulations, can be extended and modified.
-
+For simple cases (e.g., a t-test) the power can be calculated analytically while in complex cases simulating data is the best approach.
+We present very general example that, as for the previous simulations, can be extended easily extended.
 For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting \(H_0\)) or assessing an eventual bias in model estimates.
+
 The general workflow for a power simulation can be summarized as:
@@
 
-To make a practical example, we simulate the power of detecting a group difference of \(d = 0.4\) (assuming a \emph{probit} model).
+We can estimate the power of detecting a group difference of \(d = 0.4\) (assuming a \emph{probit} model).
 Participants respond to an ordinal variable with \(k = 5\).
-The iteration is performed using a \texttt{for} loop that repeat 1000 times the data simulation, model fitting and extract the p-value for the \(\beta_1\).
+The simulation is performed using a \texttt{for} loop that repeat 1000 times the data simulation, model fitting and extract the p-value for the \(\beta_1\).
 The power is then estimated as the number of p-values lower than the critical level over the number of simulations.
@@
 \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{nsim)\{}
-\NormalTok{  sim }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{group, }\AttributeTok{By =}\NormalTok{ d, }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{  sim }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{group, }\AttributeTok{beta =}\NormalTok{ d, }\AttributeTok{prob0 =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
 \NormalTok{  fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ group, }\AttributeTok{data =}\NormalTok{ sim, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
@@
 \FunctionTok{mean}\NormalTok{(p }\SpecialCharTok{\textless{}=}\NormalTok{ alpha, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
-\CommentTok{\#\textgreater{} [1] 0.386}
+\CommentTok{\#\textgreater{} [1] 0.381}
 \end{Highlighting}
@@
   \ControlFlowTok{for}\NormalTok{(j }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{nsim)\{}
-\NormalTok{    sim }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{group, }\AttributeTok{By =}\NormalTok{ d, }\AttributeTok{prob =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\NormalTok{    sim }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{group, }\AttributeTok{beta =}\NormalTok{ d, }\AttributeTok{prob0 =}\NormalTok{ probs0, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
 \NormalTok{    fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ group, }\AttributeTok{data =}\NormalTok{ sim, }\AttributeTok{link =} \StringTok{"probit"}\NormalTok{)}
@@
 \NormalTok{power}
-\CommentTok{\#\textgreater{} [1] 0.219 0.409 0.531 0.749 0.970}
+\CommentTok{\#\textgreater{} [1] 0.216 0.407 0.561 0.748 0.964}
 \end{Highlighting}
@@
 
-\normalsize
-
-\subsection{Disclaimer about the functions}\label{disclaimer-about-the-functions}
-
-The current paper proposed a simplified way with some functions to generate ordinal data.
-For more complex simulations such as simulating correlated ordinal data the \texttt{simstudy} package \url{https://kgoldfeld.github.io/simstudy/articles/ordinal.html} proposed a very comprehensive set of data simulation function also for ordinal data.
+\begin{center}\textbf{[Figure 11 about here]} \end{center}
+
+\normalsize
 

### 2024-01-14 docs/supplementary/simulating-functions.qmd
updating

---
title: "Simulation Functions"
format: 
  html:
    toc: true
bibliography: "`r filor::fil()$bib`"
csl: "`r filor::fil()$csl`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(cowplot)
library(ordinal)
devtools::load_all()

knitr::opts_chunk$set(echo = TRUE,
                      dev = "svg",
                      warning = FALSE,
                      message = FALSE)

funs <- filor::get_funs(here::here("R", "utils.R"))
```

The core of the tutorial and the set of R functions is the `sim_ord_latent()` function. Basically the function provide an intuitive way to simulate ordinal data given a dataset with predictors and a set of regression parameters on the location and scale of the latent distribution. The @lst-sim-ord-latent provide the source code of the function.^[The function depends also on other functions loaded using `devtools::load_all()`]

```{r}
#| output: asis
#| echo: false
#| lst-label: lst-sim-ord-latent
#| lst-cap: Simulating ordinal data from the latent formulation of the model
filor::print_fun(funs$sim_ord_latent)
```

The arguments are:

- `location`: is an R formula `~ ...` with the predictors on the location parameters. This is expressed as in standard regression modelling `y ~ x1 + x2 + ...` but omitting the left-side of the formula. The parameters need to match columns of the `data = ` dataset.
- `scale`: The same as the `location` but with predictors on the scale parameter of the latent distribution. By default no predictors are considered thus the scale is assumed to be the same across the dataset. By using predictors on the scale we are fitting a so-called location-scale model.
- `beta`: Vector of regression coefficients for the location effect. These values are expressed in the scale of the link function e.g., odds ratios for a logit model. The order of the vector should match how the formula for the location is expressed
- `zeta`: Vector of regression coefficients for the scale effect. These values are expressed in the scale of the link function e.g., odds ratios for a logit model. The order of the vector should match how the formula for the scale is expressed
- `prob0`: vector with baseline probabilities. With baseline we intended that this vector contains the probabilities of the $k$ outcomes when all predictors are fixed to 0. For example, when $x$ is a dummy indicator for a group effect (0 for group "a" and 1 for group "b"), `prob0` is the vector of probabilities for the group "a". Internally this vector is converted into the vector of $k - 1$ thresholds $\alpha_k$.
- `alpha`: vector with $k - 1$ thresholds. This can be specified alternatively to `prob0` if we already know the thresholds.

### 2024-01-14 R/utils.R

@@ -63,11 +63,13 @@ sim_ord_latent <- function(location,
     data$ys <- ystar # save also the latent
   } else{
     alpha <- c(-Inf, alpha, Inf)
-    cp <- lapply(alpha, function(a) plogis(a - (lploc / exp(lpscale))))
+    cp <- lapply(alpha, function(a) plogis((a - lploc) / exp(lpscale)))
     cp <- data.frame(do.call(cbind, cp))
     p <- data.frame(t(apply(cp, 1, diff)))
     names(p) <- paste0("y", 1:ncol(p))
-    data <- cbind(data, p)
+    cp <- cp[, 2:k]
+    names(cp) <- paste0("yp", Reduce(paste0, 1:(k - 1), accumulate = TRUE))
+    data <- cbind(data, cp, p)
   }
   return(data)
 }

### 2024-01-14 tt.R

@@ -1,0 +1,56 @@
+library(ordinal)
+
+odds <- function(p) p / (1 - p)
+
+set.seed(2026)
+
+N <- 1e4 # total sample size
+k <- 4 # number of ordinal outcomes
+x <- rep(c(0, 1), each = N/2) # group
+b1 <- log(3) # log odds ratio
+z1 <- log(2) # log difference in the scale
+probs0 <- rep(1/k, k) # uniform probabilities for the group 0
+alpha <- qlogis(cumsum(probs0)[-k]) # thresholds
+ystar <- rlogis(N, location = (b1 * x), scale = exp(z1 * x))
+y <- ordered(findInterval(ystar, alpha) + 1)
+dat <- data.frame(x, y)
+
+fit <- clm(y ~ x, scale = ~ x, data = dat, link = "logit")
+summary(fit)
+
+# I fitted a proportional odds model or better the b1 is the same
+# for each cutpoint
+
+fit_non_po <- clm(y ~ 1, nominal = ~ x, data = dat, link = "logit")
+
+# the vector of coefficients b1 for each alpha is the same
+# as fitting k - 1 logistic regression on the cumulative probabilities
+
+y1vs234 <- ifelse(dat$y <= 1, 1, 0)
+y12vs34 <- ifelse(dat$y <= 2, 1, 0)
+y123vs4 <- ifelse(dat$y <= 3, 1, 0)
+
+ylist <- list(y1vs234, y12vs34, y123vs4)
+
+fitl <- lapply(ylist, function(y) glm(y ~ x, family = binomial(link = "logit")))
+
+# betas are the same for the non-po model and the k - 1 logistic regressions
+sapply(fitl, coef)[2, ]
+coef(fit_non_po)[4:6]
+
+# here i estimate the cumulative odds from the observed probabilities
+
+(p <- sapply(ylist, function(y) tapply(y, dat$x, mean)))
+
+# cumulative odds ratios
+apply(p, 2, function(c) log(odds(c[1])/odds(c[2])))
+
+# here the true cumulative odds from the linear predictor of the scale-location
+# model
+
+pr <- data.frame(x = c(0, 1))
+lp <- lapply(c(-Inf, alpha, Inf), function(a) plogis((a - b1 * pr$x) / exp(z1 * pr$x)))
+lp <- data.frame(lp)[, 2:k]
+names(lp) <- c("y1vs234", "y12vs34", "y123vs4")
+apply(lp, 2, function(c) log(odds(c[1])/odds(c[2])))
+

### 2024-01-22 docs/paper/paper.tex
updating

@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-explain-cumulative-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-explain-cumulative-1}
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-logit-vs-probit-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-logit-vs-probit-1}
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-odds-example-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-odds-example-1}
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-prop-odds-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-prop-odds-1}
 
@@
 \FunctionTok{sample}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ p[}\DecValTok{1}\NormalTok{, ])}
-\CommentTok{\#\textgreater{} [1] 3}
+\CommentTok{\#\textgreater{} [1] 1}
 
@@
 \CommentTok{\#\textgreater{}   x y}
-\CommentTok{\#\textgreater{} 1 0 2}
+\CommentTok{\#\textgreater{} 1 0 4}
 \CommentTok{\#\textgreater{} 2 0 4}
@@
 \CommentTok{\#\textgreater{} 4 0 2}
-\CommentTok{\#\textgreater{} 5 0 1}
+\CommentTok{\#\textgreater{} 5 0 3}
 \CommentTok{\#\textgreater{} 6 0 1}
@@
 \CommentTok{\#\textgreater{}        1    2    3    4}
-\CommentTok{\#\textgreater{}   0 0.24 0.32 0.12 0.32}
-\CommentTok{\#\textgreater{}   1 0.10 0.28 0.26 0.36}
+\CommentTok{\#\textgreater{}   0 0.28 0.24 0.22 0.26}
+\CommentTok{\#\textgreater{}   1 0.08 0.12 0.24 0.56}
 
@@
 \CommentTok{\#\textgreater{}        0    1}
-\CommentTok{\#\textgreater{}   1 0.24 0.10}
-\CommentTok{\#\textgreater{}   2 0.56 0.38}
-\CommentTok{\#\textgreater{}   3 0.68 0.64}
+\CommentTok{\#\textgreater{}   1 0.28 0.08}
+\CommentTok{\#\textgreater{}   2 0.52 0.20}
+\CommentTok{\#\textgreater{}   3 0.74 0.44}
 \CommentTok{\#\textgreater{}   4 1.00 1.00}
@@
 \CommentTok{\#\textgreater{}        1        2        3        4 }
-\CommentTok{\#\textgreater{} 2.842105 2.076555 1.195312      NaN}
+\CommentTok{\#\textgreater{} 4.472222 4.333333 3.622378      NaN}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}133.18 274.37 7(0)  9.13e{-}08 2.1e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}125.40 258.80 5(0)  3.05e{-}09 1.8e+01}
 \CommentTok{\#\textgreater{} }
 \CommentTok{\#\textgreater{} Coefficients:}
-\CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)}
-\CommentTok{\#\textgreater{} x   0.3518     0.2190   1.606    0.108}
+\CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
+\CommentTok{\#\textgreater{} x   0.8280     0.2275   3.639 0.000274 ***}
+\CommentTok{\#\textgreater{} {-}{-}{-}}
+\CommentTok{\#\textgreater{} Signif.
+codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}0.7949     0.1794  {-}4.431}
-\CommentTok{\#\textgreater{} 2|3   0.1063     0.1693   0.628}
-\CommentTok{\#\textgreater{} 3|4   0.5994     0.1740   3.445}
+\CommentTok{\#\textgreater{} 1|2 {-}0.58309    0.17598  {-}3.313}
+\CommentTok{\#\textgreater{} 2|3  0.02447    0.16790   0.146}
+\CommentTok{\#\textgreater{} 3|4  0.66422    0.17602   3.774}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}      x       lp       ystar y}
-\CommentTok{\#\textgreater{} 1    0 0.000000 {-}1.4897139 1}
-\CommentTok{\#\textgreater{} 2    0 0.000000 {-}0.9811171 2}
-\CommentTok{\#\textgreater{} 3    0 0.000000 {-}0.2015837 2}
-\CommentTok{\#\textgreater{} 998  1 1.386294  1.0619636 3}
-\CommentTok{\#\textgreater{} 999  1 1.386294  0.9690069 3}
-\CommentTok{\#\textgreater{} 1000 1 1.386294 {-}1.8056165 1}
+\CommentTok{\#\textgreater{} 1    0 0.000000  0.40548307 3}
+\CommentTok{\#\textgreater{} 2    0 0.000000  1.89090227 4}
+\CommentTok{\#\textgreater{} 3    0 0.000000 {-}2.26992221 1}
+\CommentTok{\#\textgreater{} 998  1 1.386294 {-}0.98284222 2}
+\CommentTok{\#\textgreater{} 999  1 1.386294 {-}0.06339477 2}
+\CommentTok{\#\textgreater{} 1000 1 1.386294 {-}0.21570993 2}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-sim-from-latent-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-sim-from-latent-1}
 
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik   AIC     niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1255.36 2518.73 4(0)  3.26e{-}08 1.6e+01}
+\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1265.26 2538.51 4(0)  2.10e{-}08 1.7e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x   1.3403     0.1217   11.02   \textless{}2e{-}16 ***}
+\CommentTok{\#\textgreater{} x   1.2901     0.1209   10.67   \textless{}2e{-}16 ***}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}1.11938    0.09713 {-}11.524}
-\CommentTok{\#\textgreater{} 2|3 {-}0.03563    0.08586  {-}0.415}
-\CommentTok{\#\textgreater{} 3|4  1.10228    0.09360  11.777}
+\CommentTok{\#\textgreater{} 1|2 {-}1.10239    0.09662 {-}11.410}
+\CommentTok{\#\textgreater{} 2|3 {-}0.01977    0.08561  {-}0.231}
+\CommentTok{\#\textgreater{} 3|4  1.09680    0.09307  11.784}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}     x y        ys}
-\CommentTok{\#\textgreater{} 1   0 3  0.1445982}
-\CommentTok{\#\textgreater{} 2   0 1 {-}1.5179018}
-\CommentTok{\#\textgreater{} 3   0 4  1.5222026}
-\CommentTok{\#\textgreater{} 98  1 4  3.3201003}
-\CommentTok{\#\textgreater{} 99  1 4  2.6305353}
-\CommentTok{\#\textgreater{} 100 1 2 {-}0.1008892}
+\CommentTok{\#\textgreater{} 1   0 4  1.925855}
+\CommentTok{\#\textgreater{} 2   0 1 {-}1.663871}
+\CommentTok{\#\textgreater{} 3   0 1 {-}1.215561}
+\CommentTok{\#\textgreater{} 98  1 4  1.935991}
+\CommentTok{\#\textgreater{} 99  1 4  2.818956}
+\CommentTok{\#\textgreater{} 100 1 4  3.384374}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-show-th-example-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-show-th-example-1}
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-example-cat-latent-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-example-cat-latent-1}
 
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-example-num-latent-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-example-num-latent-1}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  400  {-}421.76 853.51 5(0)  6.60e{-}13 2.3e+01}
+\CommentTok{\#\textgreater{}  probit flexible  400  {-}398.87 807.73 5(0)  2.70e{-}10 2.4e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}         Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x11      0.06325    0.11371   0.556   0.5781    }
-\CommentTok{\#\textgreater{} x21      0.61698    0.11444   5.391    7e{-}08 ***}
-\CommentTok{\#\textgreater{} x11:x21  0.57404    0.22768   2.521   0.0117 *  }
+\CommentTok{\#\textgreater{} x11       0.1294     0.1165   1.111  0.26659    }
+\CommentTok{\#\textgreater{} x21       0.9499     0.1181   8.046 8.57e{-}16 ***}
+\CommentTok{\#\textgreater{} x11:x21   0.6674     0.2333   2.861  0.00422 ** }
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.43993    0.06606  {-}6.659}
-\CommentTok{\#\textgreater{} 2|3  0.45458    0.06642   6.844}
+\CommentTok{\#\textgreater{} 1|2 {-}0.31805    0.06668  {-}4.770}
+\CommentTok{\#\textgreater{} 2|3  0.50499    0.06839   7.384}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-effects-2-by-2-interaction-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-effects-2-by-2-interaction-1}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}97.98 205.97 6(0)  5.36e{-}13 1.1e+02}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}99.92 209.84 4(0)  1.85e{-}11 7.2e+01}
 \CommentTok{\#\textgreater{} }
 \CommentTok{\#\textgreater{} Coefficients:}
-\CommentTok{\#\textgreater{}      Estimate Std. Error z value Pr(\textgreater{}|z|)}
-\CommentTok{\#\textgreater{} x     0.03305    0.40738   0.081    0.935}
-\CommentTok{\#\textgreater{} g1    0.30569    0.46246   0.661    0.509}
-\CommentTok{\#\textgreater{} x:g1  1.01724    0.81525   1.248    0.212}
+\CommentTok{\#\textgreater{}      Estimate Std. Error z value Pr(\textgreater{}|z|)  }
+\CommentTok{\#\textgreater{} x      0.9464     0.4066   2.328   0.0199 *}
+\CommentTok{\#\textgreater{} g1     1.0557     0.4603   2.293   0.0218 *}
+\CommentTok{\#\textgreater{} x:g1  {-}0.3967     0.8086  {-}0.491   0.6237  }
+\CommentTok{\#\textgreater{} {-}{-}{-}}
+\CommentTok{\#\textgreater{} Signif.
+codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.65333    0.24337  {-}2.685}
-\CommentTok{\#\textgreater{} 2|3  0.04236    0.23666   0.179}
+\CommentTok{\#\textgreater{} 1|2  {-}0.1297     0.2349  {-}0.552}
+\CommentTok{\#\textgreater{} 2|3   0.9918     0.2494   3.977}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-effects-num-by-cat-interaction-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-effects-num-by-cat-interaction-1}
 
@@
 \FunctionTok{mean}\NormalTok{(p }\SpecialCharTok{\textless{}=}\NormalTok{ alpha, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
-\CommentTok{\#\textgreater{} [1] 0.381}
+\CommentTok{\#\textgreater{} [1] 0.414}
 \end{Highlighting}
@@
 \NormalTok{power}
-\CommentTok{\#\textgreater{} [1] 0.216 0.407 0.561 0.748 0.964}
+\CommentTok{\#\textgreater{} [1] 0.229 0.383 0.540 0.745 0.971}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper-new_files/figure-latex/fig-power-curve-1}
+{\centering \includegraphics{paper_files/figure-latex/fig-power-curve-1}
 

### 2024-01-22 index.qmd
update

---
title: "Test"
---

This is a Quarto website.

To learn more about Quarto websites visit <https://quarto.org/docs/websites>.

### 2024-01-22 supplementary/location-scale-models.qmd
update

---
title: "Location-Scale Ordinal Models"
format: 
  html:
    toc: true
bibliography: "`r filor::fil()$bib`"
csl: "`r filor::fil()$csl`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(cowplot)
library(ordinal)
devtools::load_all()

knitr::opts_chunk$set(echo = TRUE,
                      dev = "svg",
                      warning = FALSE,
                      message = FALSE)
```

## Scale Effects

The default ordinal regression model assume that the variance of the underlying latent distribution is the same across condition. This is similar to a standard linear regression assuming the homogeneity of variance. For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy]. In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution. This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included. The Equations [-@eq-scale-location] and [-@eq-latent-scale-location] expand the standard model including the linear predictor on the scale of the latent distribution.

$$
P(Y \leq k) = g^{-1}\left(\frac{\alpha_k - \mathbf{X}\boldsymbol{\beta}}{e^{\boldsymbol{Z}\boldsymbol{\zeta}}}\right)
$$ {#eq-scale-location}

$$
Y^\star_i = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon} \;\;\; \epsilon_i \sim \mathcal{N}(0, e^{Z\zeta})
$$ {#eq-latent-scale-location}

Where $X\zeta$ is the linear predictor for the scale of the distribution. By default for both the *logit* and *probit* model the scale is fixed to 1. On scale-location models we put predictors on both parameters. Given that the scale cannot be negative we use a log link function $\eta = \text{log}(X\zeta)$. As suggested by @Tutz2017-de location-scale models can be considered as a more parsimonious approach compared to partially or completely relaxing the proportional odds assumption. Allowing the scale to be different as a function of the predictor create more modelling flexibility. Furthermore, two groups could be theoretically different only in the scale of the latent distribution with a similar location. In this example, the only way to capture group differences is by including a scale effect. The fig-scale-effect-example) depict the impact of having different scales between two groups on the ordinal probabilities.

```{r}
#| label: fig-scale-effect-example 
cat_latent_plot(m = c(0, 1), s = c(1, 2), probs = rep(1/4, 4), link = "logit", plot = "both")
```

@Tutz2022-dg provide a very clear and intuitive explanation of what happen when including a scale effect and how to interpret the result. As suggested before, the scale-location model allow to independently predict changes in the location and the scale. While location shifts are simply interpreted as increasing/decreasing the latent $\mu$ or the odds of responding a certain category scale effects are not straightforward. As the scale increase (e.g., the variance increase) there is an higher probability mass on extreme categories. On the other side as the scale decrease, responses are more concentrated on specific categories.

The location parameter determine the category and the scale determine the concentration around the category. For example, if one group have a certain latent mean $\mu_1$ and a small scale $\sigma^2 = 1/3$ (thus one third compared to the standard version of the distribution), all responses will be focused on categories around the latent mean. On the other side, increasing the scale will increase the cumulative probabilities for all categories and for values that tends to infinity extreme categories are preferred. The scale parameter can somehow be interpreted as the **response style**.

```{r}
#| label: fig-sigma
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 10
sigma_high <- cat_latent_plot(m = c(0, 0), 
                s = c(1, 2), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$"))
sigma_low <- cat_latent_plot(m = c(0, 0), 
                s = c(1, 0.5), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$"))
sigma_high_mu1 <- cat_latent_plot(m = c(0, 1), 
                s = c(1, 2), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$, $\\mu_2 = 1$"))
sigma_low_mu1 <- cat_latent_plot(m = c(0, 1), 
                s = c(1, 0.5), 
                probs = rep(1/4, 4), 
                link = "probit", 
                plot = "both",
                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$, $\\mu_2 = 1$"))
plot_grid(sigma_high, sigma_low, sigma_high_mu1, sigma_low_mu1)
```

### Simulating scale effects

The location-scale model can be simulated using the `sim_ord_latent()` function and providing the predictors for the scale parameters. Given the `log` link function, predictors are provided on the `log` scale. For example, we simulate the effect of a binary variable $x$ representing two independent groups predicting the $k = 5$ response. We simulate a *location* effect of $\beta_1 = 0.5$ (in *probit* scale) and $\zeta_1 = \text{log}(2) = 0.70$. The first group has a $\sigma = 1$ and the second group has $\sigma = 2$. Again we simulate that the baseline probabilities are uniform for the first group.

```{r}
k <- 5  # number of options (1)
n <- 1e5 # number of observations (1)
b1 <- 0.5 # beta1, the shift in the latent distribution (2)
z1 <- log(2) # zeta1, the change in the scale
probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
dat <- data.frame(x = rep(c(0, 1), each = n/2))
dat <- sim_ord_latent(~x, scale = ~x, beta = b1, zeta = z1, prob0 = probs0, data = dat, link = "probit")
fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
```

The table (put ref) reports the simulation results.

```{r}
#| eval: false
truth <- c(alphas, b1, z1)
names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
#clm_table(fit, truth)
```

To better understand the impact of assuming (or simulating) a different latent scale we fit $k - 1$ binomial regressions and check the estimated coefficients. We are not simulating a specific beta for each outcome but simulating a scale effect is actually impacting the regression coefficients. When generating data for a binary outcome the linear predictor is composed by $\eta = \beta_0 + \beta_1x$. The threshold $\alpha$ and slope of the function can be estimated using $\alpha = -\frac{\beta_0}{\beta_1}$ and the slope is $\frac{1}{\beta_1}$ [@Knoblauch2012-to; @Faraggi2003-hm]. Under the proportional odds assumption, there is only a change in thresholds $\alpha$ this a shift in the sigmoid along the $x$ axis. When including a scale effect a change in the sigmoid is combined with a change in the slope.

```{r}
set.seed(2023)
k <- 4
n <- 1e5
b1 <- 3
d1 <- log(2)
dat <- data.frame(x = runif(n))
dat <- sim_ord_latent(~x, ~x, beta = b1, zeta = d1, prob0 = rep(1/k, k), data = dat, link = "probit")

dat$y1vs234 <- ifelse(dat$y <= 1, 1, 0)
dat$y12vs34 <- ifelse(dat$y <= 2, 1, 0)
dat$y123vs4 <- ifelse(dat$y <= 3, 1, 0)

dat$y <- ordered(dat$y)
fit <- clm(y ~ x, scale = ~x, data = dat, link = "probit")

fit1vs234 <- glm(y1vs234 ~ x, data = dat, family = binomial(link = "probit"))
fit12vs34 <- glm(y12vs34 ~ x, data = dat, family = binomial(link = "probit"))
fit123vs4 <- glm(y123vs4 ~ x, data = dat, family = binomial(link = "probit"))

fits <- list(y1vs234 = fit1vs234, fit12vs34 = fit12vs34, fit123vs4 = fit123vs4)

lapply(fits, function(x) coef(x)) |> 
  bind_rows(.id = "model") |> 
  mutate(xp = list(seq(0, 5, 0.01))) |> 
  unnest(xp) |> 
  ggplot(aes(x = xp, y = `(Intercept)` + x * xp)) +
  geom_line(aes(color = model))
```

### Parallel slopes with scale effects

When fitting a model with `clm()` the way of relaxing the proportional odds or parallel slopes assumption is including what is called a *nominal* effect. Basically when a predictor is included as a nominal effect, there will be $k - 1$ coefficients (where $k$ is the number of ordered levels). In this way, the odds ratio or the difference in $z$ scores will not be the same across $k$ levels.

If a scale effect is included in `clm()` is not immediately clear what happens to the proportional odds assumption. In fact, we are still estimating a single $\beta$ but we have extra parameters on the scale of the distribution. However, changing the scale of the distribution is the same as changing the slope. This is clear from @fig-scale-slope-effect where changing the scale of the latent distribution is the same as changing the slope of the cumulative probability function. The slope ($\beta_1$) is related to the scale of the distribution (logistic or normal) as $\beta_1 = \frac{1}{s}$. In fact, if we include a scale effect we are changing the scale of the underlying latent distribution and thus the slope of of the functions.

```{r}
#| label: fig-scale-slope-effect
#| echo: false
location <- 0
scale <- c(0.2, 1, 2)

par(mfrow = c(3,1))

curve(dnorm(x, location, scale[1]), -4, 4, col = 1, lwd = 1.5, ylab = "Density")
curve(dnorm(x, location, scale[2]), -4, 4, add = TRUE, col = 2, lwd = 1.5)
curve(dnorm(x, location, scale[3]), -4, 4, add = TRUE, col = 3, lwd = 1.5)

curve(pnorm(x, location, scale[1]), -4, 4, col = 1, lwd = 1.5, ylab = "Probability")
curve(pnorm(x, location, scale[2]), -4, 4, add = TRUE, col = 2, lwd = 1.5)
curve(pnorm(x, location, scale[3]), -4, 4, add = TRUE, col = 3, lwd = 1.5)

b0 <- location
b1 <- 1/scale

x <- seq(-4, 4, 0.01)
lp <- lapply(b1, function(slope) b0 + slope * x)

plot(x, lp[[1]], type = "l", lwd = 1.5, col = 1, ylab = latex2exp::TeX("$\\eta$"))
lines(x, lp[[2]], lwd = 1.5, col = 2)
lines(x, lp[[3]], lwd = 1.5, col = 3)
```

In fact, if we simulate a model with a scale effect, odds ratios or $z$ differences are no longer constant.

We will simulate the following situation where a binary predictor $x$ has an effect both on location and scale of the logistic distribution.

```{r}
odds <- function(p) p / (1 - p)

n <- 1e4 # large N
k <- 4
b1 <- log(3)
z1 <- log(2)
dat <- data.frame(x = rep(c(0, 1), each = n))
dat <- sim_ord_latent(~x, ~x, beta = log(3), zeta = log(2), prob0 = rep(1/k, k), data = dat, link = "logit")

# model with scale effects, log(odds ratios) are different

fit <- clm(y ~ x, scale = ~ x, data = dat, link = "logit")

preds <- predict(fit, newdata = data.frame(x = c(0, 1)))$fit
pcum <- t(apply(preds, 1, cumsum))[, 1:(k - 1)]

apply(pcum, 2, function(c) log(odds(c[1])/odds(c[2])))

# model without scale effects, log(odds ratios) are equal

fit0 <- clm(y ~ x, scale = ~ 1, data = dat, link = "logit")
preds <- predict(fit0, newdata = data.frame(x = c(0, 1)))$fit
pcum <- t(apply(preds, 1, cumsum))[, 1:(k - 1)]
apply(pcum, 2, function(c) log(odds(c[1])/odds(c[2])))
```

```{r}
cat_latent_plot(c(0, 0 + b1), s = c(exp(0), (exp(0) + z1)), probs = rep(1/k, k), link = "logit", plot = "both")
```

Actually we can model this dataset using a non-proportional odds model. The model correctly capture the more complex probability structure but we are not able to recover the parameters because we simulated a scale effect.

```{r}
fit_nominal <- clm(y ~ 1, nominal = ~ x, data = dat, link = "logit")
summary(fit_nominal)
```

Essentially we estimated two sets of thresholds one for the group 0 and one for the group 1. The difference with the previous model (beyond the fact that data are generated under a scale-location model) is that we need more parameters in a non-proportional odds model compared to the scale-location model. This is also more relevant when including more predictors. Also the likelihood ratio test correctly suggest that there is no relevant difference in the likelihood of the two models.

```{r}
anova(fit, fit_nominal)
```

## Location Shift Models

The location-shift model is considered an alternative way to include dispersion effects (together with location effects) [@Tutz2022-dg; @Tutz2017-de; @Tutz2020-xq]. The core of the models is including predictors on the thresholds $\alpha$ to decrease/increase the probability of response categories $k$ together with the parameters on the location $\boldsymbol{\beta}$

They are similar to the location-scale models in terms of parsimony but different in terms of parameters interpretation. Given that these models cannot be directly fitted by the `ordinal` package we are not going into details but we included some references to understand the model parametrization. The models are implemented in the `ordDisp` R package (https://cran.r-project.org/web/packages/ordDisp/index.html).

However, as noted by Berger and Tutz^[https://www.researchgate.net/profile/Gerhard-Tutz/publication/305082385_Modelling_of_Varying_Dispersion_in_Cumulative_Regression_Models/links/578115fd08ae9485a43bd0f0/Modelling-of-Varying-Dispersion-in-Cumulative-Regression-Models.pdf] there is a correspondence between the partial proportional odds model and the location-shift model. Let's simulate a model with a scale effect where the true model has a scale effect:

```{r}
library(ordinal)
library(ordDisp)

n <- 1e3
k <- 3 # number of ordinal levels
probs0 <- rep(1/k, k) # baseline probabilities

dat <- data.frame(x = rep(c(0, 1), each = n))

dat <- sim_ord_latent(~x, ~x, 
                      beta = b1, zeta = z1, 
                      prob0 = probs0, 
                      link = "logit", 
                      data = dat, 
                      simulate = TRUE)

head(dat)
```

Now let's fit the location-shift model and the non-proportional odds model:

```{r}
#| collapse: true
#| echo: true
fit_shift <- ordDisp(y ~ x|x, data = dat)
fit_nopo <- clm(y ~ 1, nominal = ~x, data = dat, link = "logit")

summary(fit_shift)
summary(fit_nopo)
```

Firstly, there is a correspondence in the baseline thresholds:

```{r}
#| collapse: true
#| echo: true
# location-shift model
coef(fit_shift)[1:(k - 1)]

# non proportional odds model
coef(fit_nopo)[1:(k - 1)]
```

Then also linear predictors $\eta$ are the same:

```{r}
#| collapse: true
#| echo: true
# let's see the linear predictor for x = 1

# location-shift model
predict(fit_shift)[n + 1, ]

# non proportional odds model
predict(fit_nopo, newdata = data.frame(x = 1), type = "linear.predictor")$eta1[, -k]
```

### 2024-01-22 supplementary/simulating-glm.qmd
update

---
title: "Simulating Generalized Linear Models"
format: 
  html:
    toc: true
bibliography: "`r filor::fil()$bib`"
csl: "`r filor::fil()$csl`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

# Introduction

This document present a general workflow to simulate data for a binomial generalized linear model. This simulation is similar to what is presented in the paper for the special case of an ordinal regression.

# Simulation

The general formula for a binomial GLM is presented in @eq-glm. Where $g^{-1}(\cdot)$ is the inverse of the link function. In the case of a binomial regression with a *logit* link function $g(p) = \text{logit}(p) = log(\frac{p}{1 - p})$ and $g^{-1}[logit(p)] = \text{invlogit}(p) = \frac{e^p}{1 + e^p}$.

$$
P(Y = 1|\mathbf{X}) = g^{-1}(\mathbf{X} \boldsymbol{\beta})
$$ {#eq-glm}

The linear combination of parameters $\eta = \mathbf{X} \boldsymbol{\beta}$ define the true probabilities. Then the random part of the model (in this case a *logistic* distribution) define the random variability around the true values. To make an example, let's assume to predict $P(Y = 1)$ with a numerical variable $x$ coming from an uniform distribution $x \sim \mathcal{U}(0, 1)$. As in standard regression we have two parameters:

-   $\beta_0$: the intercept. The probability of having 1 when all predictors are fixed to zero ($P(Y = 1|x = 0)$).
-   $\beta_1$: the slope. Is the increase in the log-odds of $Y = 1$ for a unit increase in $x$.

Thus we can choose two values for the parameters and plot the true relationship between $x$ and $P(Y = 1)$. The link function is `qlogis()` and the inverse of the link function is `plogis()`. Parameters are expressed in the scale of the of the link function.

```{r}
b0 <- qlogis(0.01) # probability of success when x = 0, in the link function space
b1 <- 10 # increase in the log odds of success for a unit increase in x
x <- seq(0, 1, 0.01) # x values
p <- plogis(b0 + b1 * x)

plot(x, p, type = "l")
```

The curve depict the true probability of success for each value of $x$. To include the random error we need to sample the observed values from a Binomial distribution using the vector of probabilities `p`. We sample a vector of 0 and 1.

```{r}
ns <- length(x) # number of subjects
y <- rbinom(ns, 1, p)
head(y)

par(mfrow = c(1,2))

# binary values
plot(x, jitter(y, amount = 0.05))
lines(x, p)

# grouped values
xc <- cut(x, seq(0, 1, 0.1), include.lowest = TRUE)
yc <- tapply(y, xc, mean)

plot(as.integer(unique(xc))/10, yc)
lines(x, p)
```

In practical terms, the $\eta$ define the true probability of success for each observation (i.e., combination of predictors $\mathbf{X}$) and then the random part is introduced by sampling from the assumed probability distribution.

Then we can fit the logistic regression using `glm()` and we should recover the simulation parameters. Increasing the number of trials/observations will reduce the distance between the simulated the true values.

```{r}
dat <- data.frame(x, y)
fit <- glm(y ~ x, data = dat, family = binomial(link = "logit"))
summary(fit)

# model
coef(fit)

# truth
c(b0 = b0, b1 = b1)
```

Changing $\beta_0$ will affect the lower bound of the sigmoid curve while $\beta_1$ determine the slope of the function. @fig-ex-sigmoid depicts logistic curves with different parameters.

```{r}
#| label: fig-ex-sigmoid
#| echo: false
#| code-fold: true

b0 <- qlogis(c(0.01, 0.3, 0.5))
b1 <- c(5, 10, 20)

pars <- expand.grid(b0 = b0, b1 = b1)
pars$x <- list(seq(0, 1, 0.01))
pars <- tibble(pars)
pars$b0t <- factor(sprintf("$\\beta_0 = g(%s)$", plogis(pars$b0)))
pars$b0t <- factor(pars$b0t, labels = latex2exp::TeX(levels(pars$b0t)))

pars$b1t <- factor(sprintf("$\\beta_1 = %s$", pars$b1))
pars$b1t <- factor(pars$b1t, labels = latex2exp::TeX(levels(pars$b1t)))

pars <- unnest(pars, x)
pars$p <- with(pars, plogis(b0 + b1 * x))

ggplot(pars, aes(x = x, y = p)) +
  geom_line() +
  facet_grid(b0t ~ b1t, labeller = label_parsed)
```

This method is implemented in the paper when sampling from a categorical distribution. In the case of ordinal data we need $k - 1$ equations where $k$ is the number of ordered categories.

### 2024-01-22 R/utils-project.R

@@ -1,0 +1,41 @@
+qrender <- function(){
+  rmarkdown::render("paper/paper.Rmd")
+  system("quarto render")
+}
+
+mkdirif <- function(dir){
+  if(!fs::is_dir(dir)){
+    fs::dir_create(dir)
+  }
+}
+
+render <- function(file){
+  db <- getdb()
+  parent <- dirname(file)
+  current_md5 <- tools::md5sum(file)
+  update <- !(db$md5[db$files == names(current_md5)] == current_md5)
+  if(update){
+    rmarkdown::render(file)
+    db$md5[db$files == names(current_md5)] <- current_md5
+    saveRDS(db, ".dbcache/dbcache.rds")
+  }else{
+    cat("file not changed!")
+  }
+}
+
+getdb <- function(){
+  mkdirif(file.path(".dbcache"))
+  dbpath <- here::here(".dbcache", "dbcache.rds")
+  if(!fs::file_exists(dbpath)){
+    files <- list.files(pattern = "Rmd", full.names = TRUE, recursive = TRUE)
+    md5 <- tools::md5sum(files)
+    db <- data.frame(
+      file = gsub("\\./", "", names(md5)),
+      md5 = md5
+    )
+    names(db) <- c("files", "md5")
+    rownames(db) <- NULL
+    saveRDS(db, dbpath)
+  }
+  db <- readRDS(dbpath)
+}

### 2024-01-22 paper/paper.Rmd
update
revision 1st part gianmarco

--- 2024-01-08 docs/paper/paper-new.Rmd
@@
 
-bibliography      : "`r filor::fil()$bib`"
+bibliography:
+  - "`r filor::fil()$bib`"
 
@@
 One of the main reason is probably the usage of Likert scales [@Likert1932-xw].
-Ordinal data refers to a specific type of measurement scale [@Stevens1946-te] where ordered numbers are assigned to a variable.
-Compared to nominal scale and as the name suggest the labels are ordered.
-Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
+Ordinal data, as defined by @Stevens1946-te, belongs to a specific type of measurement scale where ordered numbers are assigned to a variable.
+In contrast to nominal scales, the labels in ordinal scales are ordered.
+Unlike interval or ratio scales, there is no explicit assumption about the distance between labels.
 An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
@@
 For the tutorial, we assume the reader is already familiar with basic R programming and introductory theory about linear regression.
-(VEDI SE SCRIVERE MEGLIO O QUALCOSA DI PIU).
+(VEDI SE SCRIVERE MEGLIO O QULCOSA DI PIU).
 The current tutorial proposed a set of custom R functions (available on Github ) to understand and simulate basic ordinal regression models.
@@
 
-Despite the actual modeling proposal by @Liddell2018-wu, there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions.
+While Liddell and Kruschke [-@Liddell2018-wu] proposed a specific modeling approach, there exists a class of regression models that consider the ordinal nature of the response variable without relying on metric assumptions.
 We can name this general class of models as *ordinal regression*.
@@
 @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
-We can identify three main classes: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe] and *adjacent category models*.
-The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.
-
-The *sequential model* as suggested by the name is appropriate when modelling sequential processes.
+
+We can identify three main classes of models: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe], and *adjacent category models*.
+Among these, the cumulative model is the most widely used, assuming the existence of a latent variable.
+This latent variable, categorized using a set of thresholds, produces the observed ordinal variable.
+The *sequential model* as suggested by the name is appropriate when modeling sequential processes.
 Assuming to have five response options, the sequential model assume that responding "3" assume a sequential process where steps "1" and "2" are already reached.
@@
 When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached [@Tutz2020-xq].
+
 The adjacent category model compare the category $k$ with $k + 1$ still assuming a latent distribution for each $k$.
 As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
-@Burkner2019-aw suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical intepretation as for the cumulative vs sequential model.
+@Burkner2019-aw suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical interpretation as for the cumulative vs sequential model.
 
@@
 Furthermore, CM are also used to model data under a signal detection theory framework [e.g., @DeCarlo2010-lj].
-The Figure \@ref(fig:fig-explain-cumulative) depict the overall structure of the *cumulative* model.
+Figure \@ref(fig:fig-explain-cumulative) depicts the overall structure of the *cumulative* model.
 
@@
 
-In this section we introduce some notation for the CM that is used through the paper and in the R code.
-We proposed a notation as consistent as possible with the literature and the `ordinal` R package used in the tutorial.
+In this section we introduce some notation for the CM that is used through the paper and in the R code [@r-lang].
+We proposed a notation as consistent as possible with the literature and the `ordinal` [@Christensen2022-jk] R package used in the tutorial.
 We define $Y_k$ as the observed ordinal response with $1, \dots, k$ levels and $Y^\star$ is the underlying latent variable.
 The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \dots, \alpha_{k - 1}$.
-Similarly to the generalized linear models framework, we define $g(x) = \eta$ as the link function that maps probabilities into the linear predictor $\eta$.
+Similarly to the generalized linear models framework [e.g., @Fox2015-ps], we define $g(x) = \eta$ as the link function that maps probabilities into the linear predictor $\eta$.
 To transform back $\eta$ into probabilities we use the inverse of the link function $x = g^{-1}(\eta)$.
 The specific link function define the type of model and require a different R function.
-For example, when assuming a Gaussian distribution we are fitting a model with a `probit` link function is the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
+For example, when assuming a Gaussian distribution, fitting a model with a *probit* link function entails using the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
 When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$[^1].
-Equation \@ref(eq:prob-cum-model1) depict the general cumulative model including predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
+Equation \@ref(eq:prob-cum-model1) shows the general cumulative model including predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
 The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to interpret the $\beta$ as in the standard regression model [@Agresti2010-rz] where higher $\beta$ values corresponds to increased probability of responding higher $k$ categories.
@@
 Despite the presence of other possibilities the `ordinal` package provide the most complete and intuitive way to implement the ordinal models.
-The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.[^2].
+The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.[^2]
 
@@
 -   `link`: is the link function. In this tutorial we consider only the *logit* and *probit* link but other link functions are available.
--   `data`: is the dataset where with the variables included in the `formula`
--   `nominal`: formula with predictors where the proportional odds assumption (See Section ) is relaxed (i.e., partial or non proportional odds)
+-   `data`: is the dataset with variables included in the `formula`
+-   `nominal`: formula with predictors where the proportional odds assumption (See Section \@ref(proportional-odds-assumption)) is relaxed (i.e., partial or non proportional odds)
 -   `scale`: formula with predictors for the scale (standard deviation) parameter.
-This argument allow to fit a scale-location model (see Section ).
+This argument allow to fit a scale-location model (see supplementary materials).
 The main `formula` argument refers to predictors on the location parameter (i.e., the mean $\mu$).
@@
 The odds takes value ranging from 0 to $\infty$.
-For example with a probability of $p = 0.8$ we have an odds of $4$, thus we have a 4 successes for each failure.
+For instance, with a probability of $p = 0.8$, we have odds of $4$, indicating that there are four successes for each failure.
 The same as having $p = 0.2$ and an odds of $0.25$ means that for each $0.25$ successes we have a failure or that we have $4$ failures for each success.
@@
 The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from $-\infty$ to $\infty$ with $p = 0.5$ as the midpoint because $\text{log}(\frac{0.5}{1 - 0.5}) = 0$.
-The standard logistic regression with two outcome model the logit transformed probability.
 
@@
 
+<!-- TODO vedi perchè forse fare un'altra simulazione con N grande per fare vedere che anche se non super diversi sono comunque simili -->
+
 ```{r cumulative-odds-example}
@@
 Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the PO assumption.
-There are more advanced versions of the model relaxing this assumption completely (*non proportional odds*) and partially (*partial proportional odds*).
+There are more advanced versions of the model relaxing this assumption completely (*non proportional odds*) and partially (*partial proportional odds*) [@Peterson1990-cz].
 
@@
 
-## *Probit Model* $z$ scores
+## *Probit Model* $\mathbf{z}$ scores
 

### 2024-01-22 paper/paper.tex
update
revision 1st part gianmarco

--- 2024-01-22 docs/paper/paper.tex
@@
 One of the main reason is probably the usage of Likert scales (Likert, 1932).
-Ordinal data refers to a specific type of measurement scale (Stevens, 1946) where ordered numbers are assigned to a variable.
-Compared to nominal scale and as the name suggest the labels are ordered.
-Compared to interval or ratio scales there is no explicit assumption about the distance between labels.
+Ordinal data, as defined by Stevens (1946), belongs to a specific type of measurement scale where ordered numbers are assigned to a variable.
+In contrast to nominal scales, the labels in ordinal scales are ordered.
+Unlike interval or ratio scales, there is no explicit assumption about the distance between labels.
 An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement).
@@
 
-Despite the actual modeling proposal by Liddell and Kruschke (2018), there is a class of regression models taking into account the ordinal nature of the response variable without metric assumptions.
+While Liddell and Kruschke (2018) proposed a specific modeling approach, there exists a class of regression models that consider the ordinal nature of the response variable without relying on metric assumptions.
 We can name this general class of models as \emph{ordinal regression}.
@@
 Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
-We can identify three main classes: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990) and \emph{adjacent category models}.
-The cumulative is the mostly used model assuming the existence of a latent variable that categorized using a set of thresholds produces the observed ordinal variable.
-
-The \emph{sequential model} as suggested by the name is appropriate when modelling sequential processes.
+
+We can identify three main classes of models: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}.
+Among these, the cumulative model is the most widely used, assuming the existence of a latent variable.
+This latent variable, categorized using a set of thresholds, produces the observed ordinal variable.
+The \emph{sequential model} as suggested by the name is appropriate when modeling sequential processes.
 Assuming to have five response options, the sequential model assume that responding ``3'' assume a sequential process where steps ``1'' and ``2'' are already reached.
@@
 When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Tutz \& Berger, 2020).
+
 The adjacent category model compare the category \(k\) with \(k + 1\) still assuming a latent distribution for each \(k\).
 As suggested by Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
-Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical intepretation as for the cumulative vs sequential model.
+Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical interpretation as for the cumulative vs sequential model.
 
@@
 Furthermore, CM are also used to model data under a signal detection theory framework (e.g., DeCarlo, 2010).
-The Figure \ref{fig:fig-explain-cumulative} depict the overall structure of the \emph{cumulative} model.
+Figure \ref{fig:fig-explain-cumulative} depicts the overall structure of the \emph{cumulative} model.
 
@@
 
-In this section we introduce some notation for the CM that is used through the paper and in the R code.
-We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} R package used in the tutorial.
+In this section we introduce some notation for the CM that is used through the paper and in the R code (R Core Team, 2023).
+We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} (Christensen, 2022) R package used in the tutorial.
 We define \(Y_k\) as the observed ordinal response with \(1, \dots, k\) levels and \(Y^\star\) is the underlying latent variable.
 The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\).
-Similarly to the generalized linear models framework, we define \(g(x) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\).
+Similarly to the generalized linear models framework (e.g., Fox, 2015), we define \(g(x) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\).
 To transform back \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\).
 The specific link function define the type of model and require a different R function.
-For example, when assuming a Gaussian distribution we are fitting a model with a \texttt{probit} link function is the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
+For example, when assuming a Gaussian distribution, fitting a model with a \emph{probit} link function entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
 When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(x\) value \(P(Y \leq k | x_i)\)}.
-Equation \eqref{eq:prob-cum-model1} depict the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
+Equation \eqref{eq:prob-cum-model1} shows the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
 The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta\) as in the standard regression model (Agresti, 2010) where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories.
@@
 Despite the presence of other possibilities the \texttt{ordinal} package provide the most complete and intuitive way to implement the ordinal models.
-The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}.
+The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}
 
@@
 \item
-  \texttt{data}: is the dataset where with the variables included in the \texttt{formula}
-\item
-  \texttt{nominal}: formula with predictors where the proportional odds assumption (See Section ) is relaxed (i.e., partial or non proportional odds)
+  \texttt{data}: is the dataset with variables included in the \texttt{formula}
+\item
+  \texttt{nominal}: formula with predictors where the proportional odds assumption (See Section \ref{proportional-odds-assumption}) is relaxed (i.e., partial or non proportional odds)
 \item
   \texttt{scale}: formula with predictors for the scale (standard deviation) parameter.
-This argument allow to fit a scale-location model (see Section ).
+This argument allow to fit a scale-location model (see supplementary materials).
 The main \texttt{formula} argument refers to predictors on the location parameter (i.e., the mean \(\mu\)).
@@
 The odds takes value ranging from 0 to \(\infty\).
-For example with a probability of \(p = 0.8\) we have an odds of \(4\), thus we have a 4 successes for each failure.
+For instance, with a probability of \(p = 0.8\), we have odds of \(4\), indicating that there are four successes for each failure.
 The same as having \(p = 0.2\) and an odds of \(0.25\) means that for each \(0.25\) successes we have a failure or that we have \(4\) failures for each success.
@@
 The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from \(-\infty\) to \(\infty\) with \(p = 0.5\) as the midpoint because \(\text{log}(\frac{0.5}{1 - 0.5}) = 0\).
-The standard logistic regression with two outcome model the logit transformed probability.
 
@@
 Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the PO assumption.
-There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}).
+There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}) (Peterson \& Harrell, 1990).
 
@@
 
-\subsection{\texorpdfstring{\emph{Probit Model} \(z\) scores}{Probit Model z scores}}\label{probit-model-z-scores}
+\subsection{\texorpdfstring{\emph{Probit Model} \(\mathbf{z}\) scores}{Probit Model \textbackslash mathbf\{z\} scores}}\label{probit-model-mathbfz-scores}
 
@@
 \FunctionTok{sample}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ p[}\DecValTok{1}\NormalTok{, ])}
-\CommentTok{\#\textgreater{} [1] 1}
+\CommentTok{\#\textgreater{} [1] 3}
 
@@
 \CommentTok{\#\textgreater{} 1 0 4}
-\CommentTok{\#\textgreater{} 2 0 4}
-\CommentTok{\#\textgreater{} 3 0 4}
-\CommentTok{\#\textgreater{} 4 0 2}
-\CommentTok{\#\textgreater{} 5 0 3}
-\CommentTok{\#\textgreater{} 6 0 1}
+\CommentTok{\#\textgreater{} 2 0 1}
+\CommentTok{\#\textgreater{} 3 0 2}
+\CommentTok{\#\textgreater{} 4 0 3}
+\CommentTok{\#\textgreater{} 5 0 1}
+\CommentTok{\#\textgreater{} 6 0 4}
 
@@
 \CommentTok{\#\textgreater{}        1    2    3    4}
-\CommentTok{\#\textgreater{}   0 0.28 0.24 0.22 0.26}
-\CommentTok{\#\textgreater{}   1 0.08 0.12 0.24 0.56}
+\CommentTok{\#\textgreater{}   0 0.26 0.24 0.26 0.24}
+\CommentTok{\#\textgreater{}   1 0.18 0.12 0.28 0.42}
 
@@
 \CommentTok{\#\textgreater{}        0    1}
-\CommentTok{\#\textgreater{}   1 0.28 0.08}
-\CommentTok{\#\textgreater{}   2 0.52 0.20}
-\CommentTok{\#\textgreater{}   3 0.74 0.44}
+\CommentTok{\#\textgreater{}   1 0.26 0.18}
+\CommentTok{\#\textgreater{}   2 0.50 0.30}
+\CommentTok{\#\textgreater{}   3 0.76 0.58}
 \CommentTok{\#\textgreater{}   4 1.00 1.00}
@@
 \CommentTok{\#\textgreater{}        1        2        3        4 }
-\CommentTok{\#\textgreater{} 4.472222 4.333333 3.622378      NaN}
+\CommentTok{\#\textgreater{} 1.600601 2.333333 2.293103      NaN}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}125.40 258.80 5(0)  3.05e{-}09 1.8e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}134.09 276.18 6(0)  2.78e{-}08 1.9e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)  }
-\CommentTok{\#\textgreater{} x   0.8280     0.2275   3.639 0.000274 ***}
+\CommentTok{\#\textgreater{} x   0.4420     0.2198   2.011   0.0443 *}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.58309    0.17598  {-}3.313}
-\CommentTok{\#\textgreater{} 2|3  0.02447    0.16790   0.146}
-\CommentTok{\#\textgreater{} 3|4  0.66422    0.17602   3.774}
+\CommentTok{\#\textgreater{} 1|2 {-}0.57303    0.17219  {-}3.328}
+\CommentTok{\#\textgreater{} 2|3 {-}0.04306    0.16493  {-}0.261}
+\CommentTok{\#\textgreater{} 3|4  0.67182    0.17437   3.853}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}      x       lp      ystar y}
-\CommentTok{\#\textgreater{} 1    0 0.000000  0.40548307 3}
-\CommentTok{\#\textgreater{} 2    0 0.000000  1.89090227 4}
-\CommentTok{\#\textgreater{} 3    0 0.000000 {-}2.26992221 1}
-\CommentTok{\#\textgreater{} 998  1 1.386294 {-}0.98284222 2}
-\CommentTok{\#\textgreater{} 999  1 1.386294 {-}0.06339477 2}
-\CommentTok{\#\textgreater{} 1000 1 1.386294 {-}0.21570993 2}
+\CommentTok{\#\textgreater{} 1    0 0.000000  0.9531361 3}
+\CommentTok{\#\textgreater{} 2    0 0.000000  1.1862471 4}
+\CommentTok{\#\textgreater{} 3    0 0.000000 {-}0.7356740 2}
+\CommentTok{\#\textgreater{} 998  1 1.386294 {-}5.0039165 1}
+\CommentTok{\#\textgreater{} 999  1 1.386294  1.8689381 4}
+\CommentTok{\#\textgreater{} 1000 1 1.386294  4.0016754 4}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik   AIC     niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1265.26 2538.51 4(0)  2.10e{-}08 1.7e+01}
+\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1268.84 2545.68 4(0)  2.84e{-}08 1.7e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x   1.2901     0.1209   10.67   \textless{}2e{-}16 ***}
+\CommentTok{\#\textgreater{} x    1.429      0.122   11.71   \textless{}2e{-}16 ***}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}1.10239    0.09662 {-}11.410}
-\CommentTok{\#\textgreater{} 2|3 {-}0.01977    0.08561  {-}0.231}
-\CommentTok{\#\textgreater{} 3|4  1.09680    0.09307  11.784}
+\CommentTok{\#\textgreater{} 1|2 {-}1.01156    0.09476 {-}10.675}
+\CommentTok{\#\textgreater{} 2|3  0.16670    0.08565   1.946}
+\CommentTok{\#\textgreater{} 3|4  1.26980    0.09575  13.262}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}     x y         ys}
-\CommentTok{\#\textgreater{} 1   0 4  1.925855}
-\CommentTok{\#\textgreater{} 2   0 1 {-}1.663871}
-\CommentTok{\#\textgreater{} 3   0 1 {-}1.215561}
-\CommentTok{\#\textgreater{} 98  1 4  1.935991}
-\CommentTok{\#\textgreater{} 99  1 4  2.818956}
-\CommentTok{\#\textgreater{} 100 1 4  3.384374}
+\CommentTok{\#\textgreater{} 1   0 3  0.0555103}
+\CommentTok{\#\textgreater{} 2   0 4  1.2882548}
+\CommentTok{\#\textgreater{} 3   0 4  2.0276663}
+\CommentTok{\#\textgreater{} 98  1 2 {-}1.0176796}
+\CommentTok{\#\textgreater{} 99  1 4  1.4546417}
+\CommentTok{\#\textgreater{} 100 1 2 {-}0.5004862}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  400  {-}398.87 807.73 5(0)  2.70e{-}10 2.4e+01}
+\CommentTok{\#\textgreater{}  probit flexible  400  {-}400.35 810.70 6(0)  2.95e{-}13 2.8e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}         Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x11       0.1294     0.1165   1.111  0.26659    }
-\CommentTok{\#\textgreater{} x21       0.9499     0.1181   8.046 8.57e{-}16 ***}
-\CommentTok{\#\textgreater{} x11:x21   0.6674     0.2333   2.861  0.00422 ** }
+\CommentTok{\#\textgreater{} x11       0.2514     0.1171   2.147   0.0318 *  }
+\CommentTok{\#\textgreater{} x21       0.8774     0.1181   7.427 1.11e{-}13 ***}
+\CommentTok{\#\textgreater{} x11:x21   0.5215     0.2343   2.226   0.0260 *  }
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.31805    0.06668  {-}4.770}
-\CommentTok{\#\textgreater{} 2|3  0.50499    0.06839   7.384}
+\CommentTok{\#\textgreater{} 1|2 {-}0.30679    0.06605  {-}4.645}
+\CommentTok{\#\textgreater{} 2|3  0.40471    0.06714   6.028}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}99.92 209.84 4(0)  1.85e{-}11 7.2e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}95.21 200.41 5(0)  3.33e{-}14 9.2e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}      Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x      0.9464     0.4066   2.328   0.0199 *}
-\CommentTok{\#\textgreater{} g1     1.0557     0.4603   2.293   0.0218 *}
-\CommentTok{\#\textgreater{} x:g1  {-}0.3967     0.8086  {-}0.491   0.6237  }
+\CommentTok{\#\textgreater{} x      1.4377     0.4303   3.341 0.000835 ***}
+\CommentTok{\#\textgreater{} g1     0.4508     0.5041   0.894 0.371205    }
+\CommentTok{\#\textgreater{} x:g1   1.1330     0.8536   1.327 0.184431    }
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}0.1297     0.2349  {-}0.552}
-\CommentTok{\#\textgreater{} 2|3   0.9918     0.2494   3.977}
+\CommentTok{\#\textgreater{} 1|2   0.2913     0.2553   1.141}
+\CommentTok{\#\textgreater{} 2|3   1.1938     0.2731   4.371}
 \end{Highlighting}
@@
 \FunctionTok{mean}\NormalTok{(p }\SpecialCharTok{\textless{}=}\NormalTok{ alpha, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
-\CommentTok{\#\textgreater{} [1] 0.414}
+\CommentTok{\#\textgreater{} [1] 0.382}
 \end{Highlighting}
@@
 
+\bibitem[\citeproctext]{ref-Christensen2022-jk}
+Christensen, R. H. B. (2022). \emph{Ordinal---regression models for ordinal data}.
+
 \bibitem[\citeproctext]{ref-Cliff1996-qo}
@@
 \url{https://doi.org/10.1016/j.jmp.2010.01.001}
+
+\bibitem[\citeproctext]{ref-Fox2015-ps}
+Fox, J. (2015). \emph{Applied regression analysis and generalized linear models}. SAGE Publications. Retrieved from \url{https://play.google.com/store/books/details?id=BlsKogEACAAJ}
 
@@
 McCullagh, P. (1980). Regression models for ordinal data. \emph{Journal of the Royal Statistical Society}, \emph{42}, 109--127. \url{https://doi.org/10.1111/j.2517-6161.1980.tb01109.x}
+
+\bibitem[\citeproctext]{ref-Peterson1990-cz}
+Peterson, B., \& Harrell, F.
+E.
+(1990).
+Partial proportional odds models for ordinal response variables.
+\emph{Journal of the Royal Statistical Society.
+Series C, Applied Statistics}, \emph{39}, 205.
+\url{https://doi.org/10.2307/2347760}
+
+\bibitem[\citeproctext]{ref-r-lang}
+R Core Team.
+(2023).
+\emph{R: A language and environment for statistical computing} {[}Computer software{]}.
+Vienna, Austria: R Foundation for Statistical Computing.
+Retrieved from \url{https://www.R-project.org/}
 

### 2024-01-25 paper/paper.Rmd
updating after giamma revision

@@
 shorttitle        : "Ordinal regression models made easy"
-
 author: 
@@
     corresponding : yes    # Define only one corresponding author
-    address       : "Postal address"
+    address       : "Via Venezia 8, 35131, Padova (Italy)"
     email         : "filippo.gambarota@unipd.it"
@@
       - "Writing - Review & Editing"
+      - "Methodology"
+      - "Software"
   - name          : "Gianmarco Altoè"
@@
     role:
+      - "Conceptualization"
       - "Writing - Review & Editing"
       - "Supervision"
-
 affiliation:
@@
     institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"
-
-authornote: |
-  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-
-  Enter author note here.
-
 abstract: |
-  Abstract here
-
-keywords          : "ordinal, likert, simulations, power"
+  Ordinal data are extremely used in psychology.
+Common examples include Likert items or variables where categories are ordered without assuming a common distance between labels.
+Despite the prevalence of ordinal data, a majority of published psychological research relies on applying metric models (e.g., standard linear regression) even when the dependent variable is ordinal.
+This approach leads to several statistical drawbacks.
+Similar to what happens with binary data (e.g., accuracies in a cognitive experiment), using metric models instead of e.g.
+logistic regression is probably caused by several reasons.
+Firstly, parameter interpretation can be challenging.
+Ordinal models commonly used a link function (e.g., logit) for model fitting thus transforming the parameters scale.
+This is even more critical when estimating models with multiple parameters or interactions.
+Secondly, nowadays journals require power or sensitivity analysis when planning or evaluating a research work.
+For common statistical models such as t-tests or the analysis of variance, there are tools to easily estimate the statistical power.
+For ordinal regressions, estimating the statistical power often requires simulating data thus clearly understanding the statistical background and implementing a Monte Carlo simulation.
+The objective of this tutorial is to present the essential statistical background of ordinal regression, highlighting the research areas that would benefit more by introducing this model.
+In particular, we present how to interpret the model parameters with categorical and numerical variables also linking them to common standardized effect size measures.
+Then we propose methods to report the results, especially focusing on graphical representation.
+Finally, we present how to estimate the statistical power using Monte Carlo simulations with suggestions on setting parameters in a meaningful way.
+Through the tutorial, we use the R programming language using specific packages for ordinal models and sharing also ad-hoc developed functions supporting each step.
+For the sake of reproducibility and to empower researchers in analyzing their studies autonomously, all data and scripts used in the article have been made publicly available.
+
+keywords          : "ordinal, simulations, power"
 wordcount         : "X"
-
 bibliography:
+  - "packages.bib"
   - "`r filor::fil()$bib`"
-
 floatsintext      : yes
@@
 mask              : no
-
 figurelist        : no
@@
   floatplaceh: true
-
-classoption       : "man"
+  preprint: true
+classoption       : "man, mask"
 output: 
@@
       - \usepackage{booktabs, caption, longtable, colortbl, array, hhline, mathbf}
+  papaja::apa6_docx: default
 ---
@@
 ```{r setup, include = FALSE}
+pdf2png <- function(path) {
+  # only do the conversion for non-LaTeX output
+  if (knitr::is_latex_output())
+    return(path)
+  path2 <- xfun::with_ext(path, "png")
+  img <- magick::image_read(path)
+  magick::image_write(img, path2, format = "png")
+  path2
+}
+
 knitr::opts_chunk$set(echo = FALSE,
                       dev = "pdf",
+                      fig.width = 8,
+                      fig.asp = 0.618,
+                      fig.process = pdf2png,
+                      out.width = "100%",
                       fig.align = "center",
@@
                       comment = "#>")
+
+which_output <- function(format){
+  format %in% knitr::opts_knit$get("rmarkdown.pandoc.to")
+}
 ```
@@
 ```{r packages, include = FALSE}
-library(tidyverse)
+library(dplyr)
+library(tidyr)
 library(ordinal)
 library(here)
-library(kableExtra)
 library(cowplot)
@@
 library(ggeffects)
+library(papaja)
+library(ggplot2)
+library(purrr)
 devtools::load_all()
+```
+
+```{r packages-latex, include = FALSE, eval = which_output("latex")}
+library(kableExtra)
 ```
@@
 mtheme <- function(){
-  ggthemes::theme_pander(15) +
+  theme_minimal(15) +
     theme(legend.position = "bottom",
@@
 scale_fill_discrete <- scale_fill_viridis_d
-```
-
-```{r scripts, include = FALSE}
+options(tikzDefaultEngine = "xetex")
+```
+
+```{r scripts, include = FALSE, eval = TRUE}
 temp <- new.env()
@@
 
+```{r refs-r}
+knitr::write_bib("base", file = "packages.bib")
+```
+
 # Introduction
@@
 Ordinal data, as defined by @Stevens1946-te, belongs to a specific type of measurement scale where ordered numbers are assigned to a variable.
+Beyond Likert-like items, in Psychology there are several applications of ordinal scales.
+For instance, sociodemographic variables like educational levels or socioeconomic status, as well as general ratings such as pain severity, agreement with a statement, or the evaluation of sensory experiences.
+
 In contrast to nominal scales, the labels in ordinal scales are ordered.
@@
 
-For the tutorial, we assume the reader is already familiar with basic R programming and introductory theory about linear regression.
-(VEDI SE SCRIVERE MEGLIO O QULCOSA DI PIU).
-The current tutorial proposed a set of custom R functions (available on Github ) to understand and simulate basic ordinal regression models.
-More details and extended examples are available on the supplementary materials.
+For the tutorial, we assume that the reader is already familiar with basic R programming and introductory theory about linear regression.
+Despite the practical and applicative spirit of the tutorial, we still introduced some statistical theory and equations supported by translation into R code.
+Similar to other tutorials [@DeBruine2021-id; @Gambarota2023-on], we believe this is the best educational approach especially with the involvement of Monte Carlo simulations.
+
+In the first part we introduce the ordinal models explaining the general structure and assumptions.
+Then we move with model fitting and parameters interpretation.
+Finally we introduce the simulation approach for common research scenarios and an application to power analysis.
+We used `r cite_r()` for the code, figures and tables.
+Details about R packages an The tutorial is supported by a set of custom R functions available on Github.
+Other examples and model extensions along with details about used R packages are available on the Github repository (supplementary materials).
+For the tutorial we used the following packages.
+
+```{r tutorial-pkgs, eval = FALSE}
+library(dplyr)
+library(tidyr)
+library(ordinal)
+library(ggplot2)
+```
 
@@
 
-In this section we introduce some notation for the CM that is used through the paper and in the R code [@r-lang].
-We proposed a notation as consistent as possible with the literature and the `ordinal` [@Christensen2022-jk] R package used in the tutorial.
-We define $Y_k$ as the observed ordinal response with $1, \dots, k$ levels and $Y^\star$ is the underlying latent variable.
+In this section we introduce some notation for the CM that is used through the paper.
+We proposed a notation as consistent as possible with the literature and the `ordinal` package [@Christensen2022-jk].
+We define $Y_k$ as the observed ordinal variable with $k$ levels and $Y^\star$ is the underlying latent variable.
 The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \dots, \alpha_{k - 1}$.
@@
 For example, when assuming a Gaussian distribution, fitting a model with a *probit* link function entails using the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
-When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$[^1].
+When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$^[As done by @Agresti2010-rz, when referring to $P(Y \leq k)$ we are implicitly conditioning on a particular $\mathbf{X}$ value $P(Y \leq k | \mathbf{X}_i)$].
 Equation \@ref(eq:prob-cum-model1) shows the general cumulative model including predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
-The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to interpret the $\beta$ as in the standard regression model [@Agresti2010-rz] where higher $\beta$ values corresponds to increased probability of responding higher $k$ categories.
-
-[^1]: As done by @Agresti2010-rz, when referring to $P(Y \leq k)$ we are implicitly conditioning on a particular $x$ value $P(Y \leq k | x_i)$
+The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to interpret the $\beta_j$ as in the standard regression models where higher $\beta$ values corresponds to increased probability of responding higher $k$ categories [@Agresti2010-rz].
 
 \begin{equation}
-P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}), \;\;k = 1, \dots, k - 1
+P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}) \;\;k = 1, \dots, k - 1
 (\#eq:prob-cum-model1)
@@
   geom_vline(xintercept = c(-4, th, 4)) +
-  theme_minimal(15) +
   theme(legend.position = "bottom",
@@
 \begin{equation}
-Y^\star_i = \mathbf{X}\boldsymbol{\beta} + \epsilon_i
+\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \mathbf{\epsilon}
 (\#eq:latent-model)
@@
 
-The crucial part is $\epsilon_i$ that is the random component of the model coming from a certain probability distribution.
+The crucial part is $\epsilon$ that is the random component of the model coming from a certain probability distribution.
 For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution.
@@
 Among several available functions the *logit* and *probit* models are the most common.
-The *logit* model assume a *logit* link function and a logistic distribution as latent variable.
-On the other side, the *probit* model assume a Gaussian distribution.
+The *logit* model use a *logit* link function and a logistic distribution as latent variable.
+On the other side, the *probit* model assume a latent Gaussian distribution using the inverse of the cumulative distribution ($\Phi^{-1}$) function as link function.
 
@@
 In the next sections we will illustrate the differences and simulation strategies.
-Figure \@ref(fig:fig-logit-vs-probit) depicts the two distributions while Table \@ref(tab:tab-model-summary) summarise the presented cumulative models with the proposed link function and the corresponding R code.
+Figure \@ref(fig:fig-logit-vs-probit) depicts the two distributions while Table L'output è `r if(which_output("latex")) "\\@ref(tab:tab-model-summary-latex)" else "\\@ref(tab:tab-model-summary-word)"` summarise the presented cumulative models with the proposed link function and the corresponding R code.
 
@@
   theme(legend.position = "bottom") +
-  viridis::scale_color_viridis(discrete = TRUE)
+  viridis::scale_color_viridis(discrete = TRUE) +
+  ylab("Density")
 
@@
   xlim(c(-7, 7)) +
-  viridis::scale_color_viridis(discrete = TRUE)
+  viridis::scale_color_viridis(discrete = TRUE) +
+  ylab("Probability")
 
@@
 
-```{r tab-model-summary}
-rtabs$models_table[-1, ] |> 
+```{r prep-model-tab-summary}
+models_table <- data.frame(
+  stringsAsFactors = FALSE,
+  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
+  Link.Function = c("Equation",
+                    "$\\text{logit}(p) = \\text{log}(p / (1-p))$",
+                    "$z = \\Phi^{-1}(p)$"),
+  Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
+  Inverse.Link.Function = c("Equation",
+                            "$e^{\\text{logit}(p)} / (1 + e^{\\text{logit}(p)})$","$p = \\Phi(z)$"),
+  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
+)
+models_table <- models_table[-1, ]
+```
+
+```{r tab-model-summary-latex, eval = which_output("latex")}
+models_table_latex <- models_table
+models_table_latex$Link.Function.1 <- replace_before_after(models_table_latex$Link.Function.1,
+                                                          before = "`",
+                                                          new = "\\\\texttt{%s}")
+models_table_latex$Inverse.Link.Function.1 <- replace_before_after(models_table_latex$Inverse.Link.Function.1,
+                                                           before = "`",
+                                                           new = "\\\\texttt{%s}")
+models_table_latex[-1, ] |> 
   kable(col.names = c("Model", "Equation", "R Code", "Equation", "R Code"), row.names = FALSE,
-        booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE) |> 
+        booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE, align = "c") |> 
   kable_styling(full_width = FALSE) |> 
-  add_header_above(header = c("", "Link Function" = 2, "Inverse Link Function" = 2))
+  add_header_above(header = c("", "Link Function $g(.)$" = 2, "Inverse Link Function $g^{-1}(.)$" = 2), escape = FALSE) |> 
+  row_spec(0, bold = TRUE)
+```
+
+```{r tab-model-summary-docx, eval=which_output("docx")}
+models_table_docx <- models_table
+names(models_table) <- c("Model", "Link Function $g(\\cdot)$", "", "Inverse Link Function $g^{-1}(\\cdot)$", "")
+papaja::apa_table(models_table, escape = FALSE, align = "c", row.names = FALSE, caption = filor::get_caption())
 ```
@@
 We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
-Table \@ref(tab:tab-dataset-example) contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
-The participants are divided into two groups ($x$).
+Table `r if(which_output("latex")) "\\@ref(tab:tab-dataset-example-latex)" else "\\@ref(tab:tab-dataset-example-docx)"` contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
+The participants are divided into two groups ($x_a$ and $x_b$).
 We can fit a cumulative link model with `clm()` function and check the model summary.
 
-```{r tab-dataset-example}
+```{r prep-tab-dataset-example}
+set.seed(2024)
+b1 <- log(3) # log odds ratio
+k <- 4
+n <- 100
+x <- rep(c("a", "b"), each = n/2)
+dat <- data.frame(x = x)
+probs <- rep(1/k, k) # for the group "a", uniform probabilities
+dat <- sim_ord_latent(~x, beta = b1, prob0 = probs, data = dat, link = "logit")
+
+funs <- list(mean = mean, median = median, sd = sd)
+ss <- sapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
+ff <- t(table(dat$y, dat$x))
+
+ss <- data.frame(cbind(ss, ff))
+names(ss) <- c("mean", "median", "sd", paste0("Y", 1:4))
+
+ss$tot <- apply(ss[, 4:ncol(ss)], 1, sum)
+pp <- ss[, 4:7]/50
+ss <- round(ss, 2)
+ss$Y1 <- sprintf("%s (p = %s)", fmt(ss$Y1, "bold", "latex"), fmt(pp$Y1, "bold", "latex"))
+ss$Y2 <- sprintf("%s (p = %s)", fmt(ss$Y2, "bold", "latex"), fmt(pp$Y2, "bold", "latex"))
+ss$Y3 <- sprintf("%s (p = %s)", fmt(ss$Y3, "bold", "latex"), fmt(pp$Y3, "bold", "latex"))
+ss$Y4 <- sprintf("%s (p = %s)", fmt(ss$Y4, "bold", "latex"), fmt(pp$Y4, "bold", "latex"))
+
+ss$group <- rownames(ss)
+ss <- select(ss, group, everything(), -tot)
 dat <- rtabs$data_example_table$dat
-map(rtabs$data_example_table$table, ~gsub("\\**", "", .x)) |> 
-  bind_rows() |> 
+
+ss <- map(rtabs$data_example_table$table, ~gsub("\\**", "", .x)) |> 
+  bind_rows()
+
+names(ss) <- stringr::str_to_title(names(ss))
+names(ss)[names(ss) == "Sd"] <- "SD"
+```
+
+```{r tab-dataset-example-latex, eval = which_output("latex")}
+ss |> 
   kable(row.names = FALSE,
-        booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE) |> 
-  kable_styling(full_width = FALSE)
+        booktabs = TRUE, 
+        format = "latex", 
+        caption = filor::get_caption(), 
+        escape = FALSE, 
+        align = "c") |> 
+  kable_styling(full_width = FALSE) |> 
+  row_spec(0,bold=TRUE)
+```
+
+```{r tab-dataset-example-docx, eval = which_output("docx")}
+papaja::apa_table(ss, caption = filor::get_caption())
 ```
@@
 The two main sections of the model summary are the *Coefficients* section reporting the regression coefficients $\beta$ and the *Threshold* section reporting the $\alpha$ estimation.
-Given that $k = 4$ we have $k - 1 = 3$ thresholds and a single $\beta$ associated with the $x$ effect.
-As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients where the interpretation depends on the contrast coding [see @Schad2020-ht].
+Given that $k = 4$ we have $k - 1 = 3$ thresholds and one $\beta$ associated with the $x$ effect.
+As in standard regression models, when $x$ is a categorical predictor with $j$ level, we will estimate $j - 1$ regression coefficients (plus the intercept term) where the interpretation depends on the contrast coding [see @Schad2020-ht].
 In R the default is the dummy coding where a factor of $j$ levels is converted into $j - 1$ dummy variables.
@@
 To understand the logit model we need to introduce odds and odds ratio.
-The odds of a probability $p$ is defined as $\frac{p}{1 - p}$ thus the probability of success divided by the probability of failure.
+The odds of a probability $p$ is defined as $p/(1 - p)$ thus the probability of success divided by the probability of failure.
 The odds takes value ranging from 0 to $\infty$.
@@
 The Figure \@ref(fig:fig-odds-example) shows the relationship between probabilities and odds.
-The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from $-\infty$ to $\infty$ with $p = 0.5$ as the midpoint because $\text{log}(\frac{0.5}{1 - 0.5}) = 0$.
+The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from $-\infty$ to $\infty$ with $p = 0.5$ as the midpoint because $\text{log}(0.5/(1 - 0.5)) = 0$.
 
@@
 
-For an ordinal variable we have $k$ levels and we can calculate $k - 1$ cumulative probabilities (excluding the last associated with $p = 1$).
+For an ordinal variable we have $k$ levels and we can calculate $k - 1$ cumulative probabilities (excluding the last associated with $P = 1$).
 Thus we have $k - 1$ odds ratio or what is called a cumulative odds ratio.
@@
 
-<!-- TODO vedi perchè forse fare un'altra simulazione con N grande per fare vedere che anche se non super diversi sono comunque simili -->
-
 ```{r cumulative-odds-example}
@@
 
-From the previous example, we noticed that the $k - 1$ odds ratios are different.
+From the previous example, we noticed that the $k - 1$ estimated odds ratios are different.
 However, increasing $n$ will result in odds ratios that are very similar.
@@
 Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the PO assumption.
-There are more advanced versions of the model relaxing this assumption completely (*non proportional odds*) and partially (*partial proportional odds*) [@Peterson1990-cz].
+There are more advanced versions of the model relaxing this assumption completely (*non proportional odds*) and partially [*partial proportional odds*, @Peterson1990-cz].
+
+```{r}
+#| echo: true
+#| collapse: true
+set.seed(2024)
+# using same parameters as the previous simulation increasing n
+n <- 1e5 # very large n
+x <- rep(c("a", "b"), each = n)
+dat <- data.frame(x = x)
+dat <- sim_ord_latent(~x, beta = b1, prob0 = probs, data = dat, link = "logit")
+
+# same steps as before
+cum_p <- cbind(x = dat$x, dummy_ord(dat$y))
+group_a <- apply(cum_p[cum_p$x == "a", -1], 2, mean)
+group_b <- apply(cum_p[cum_p$x == "b", -1], 2, mean)
+# now odds are almost the same and equal to the simulated b1 = ~1.09
+odds(group_a) / odds(group_b)
+log(odds(group_a) / odds(group_b))
+```
 
@@
 
-<!-- - https://hbiostat.org/ordinal/impactpo.pdf blog su proportional odds, mi pare di capire che non sia così problematica come assunzione -->
-
 \begin{equation}
@@
 ```{r fig-prop-odds, fig.cap=filor::get_caption(), fig.place = TRUE}
-rfigs$fig_prop_odds
+k <- 4
+dd <- tibble(
+  l = c(0, 1, 3),
+  s = c(1, 1, 1),
+  y = sprintf("$Y \\leq %s", 1:(k - 1))
+)
+
+dd <- rbind(dd, dd)
+dd$type <- rep(c("po", "npo"), each = 3)
+dd$s <- ifelse(dd$type == "npo", c(0.5, 1.5, 2), dd$s)
+
+dd$x <- list(runif(1e3, -10, 10))
+dd$p <- map2(dd$l, dd$s, ~plogis(dd$x[[1]], .x, .y))
+
+dd <- unnest(dd, c(x, p))
+dd$type <- ifelse(dd$type == "po", "PO", "NPO")
+dd$lp <- qlogis(dd$p)
+dd <- pivot_longer(dd, c(p, lp))
+dd$name <- factor(dd$name, labels = c(latex2exp::TeX("$\\eta$"), latex2exp::TeX("Probability")))
+
+dd |> 
+  ggplot(aes(x = x, y = value, color = y)) +
+  facet_grid(name~type, scales = "free", labeller = label_parsed) +
+  geom_line(lwd = 1) +
+  scale_color_discrete(labels = latex2exp::TeX(unique(dd$y))) +
+  theme(legend.title = element_blank(),
+        axis.title.y = element_blank()) +
+  xlab("x")
 ```
@@
 
-## *Probit Model* $\mathbf{z}$ scores
+## *Probit Model* and $\mathbf{z}$ scores
 
@@
 Thinking about the latent distributions, the $\beta$ is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric $x$.
-More formally the shift in the latent distribution is $\frac{\beta}{\sigma}$ (for the *probit* model $\sigma = 1$).
+More formally the shift in the latent distribution is $\beta/\sigma$ (for the *probit* model $\sigma = 1$).
 The interpretation in terms of shifting the latent mean holds also for the logistic model.
 However, the standard deviation of the standard logistic regression is $\frac{\pi}{\sqrt{3}} \approx 1.81$[^4].
-The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\beta/(\frac{\pi}{\sqrt{3}})$ standard deviations.
+The $\beta$ for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by $\beta/(\frac{\pi}{\sqrt{3}})$ standard deviations [@Agresti2010-rz].
 
@@
 \begin{equation}
-z_{x_1 - x_0} = \Phi(P \leq 1|x_0) - \Phi(P \leq 1|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
+z_{x_1 - x_0} = \Phi(P \leq 1|x_1) - \Phi(P \leq 1|x_0) = \dots = \Phi(P \leq k - 1|x_1) - \Phi(P \leq k - 1|x_0)
 (\#eq:probit-prop-odds)
@@
 In this tutorial we present two method for simulating ordinal data.
-The first method calculate the probabilities of each $Y$ level as a function of predictors and generate data from a multinomial distribution.
+Simulating data is a powerful strategy to understand the model [@DeBruine2021-id] and estimate statistical properties (e.g., power or type-1 error).
+The first simulation method calculate the probabilities of each $Y$ level as a function of predictors and generate data from a *multinomial* distribution.
 The second method simulate data using the the latent formulation.
+Whenever random number generation occurs, it is appropriate to set a `seed` (`set.seed()` function, in this case we use `set.seed(2024)`).
+Running again the same code will produce the same result.
+A general simulation approach concerns generating a dataset from the assumed data generation process, fitting the statistical model and assessing the recovery of simulated parameters.
+To check the simulation approach, using a large sample size produce estimations with small sampling error.
+A limited sample size even when fitting the true model will produce variable estimations.
 
@@
 For the first method we need to calculate $g^{-1}(\eta)$ as a function of predictors and then sample from a *multinomial* (more specifically *categorical*) distribution using the `sample()` function in R.
-This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview)
-
-As a simple example, let's simulate two groups with $n = 100$ participants responding to an item with $k = 4$ ordered options.
+This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview).
+
+As a simple example, let's simulate two groups with $n = 100$ participants responding to an item ($Y$) with $k = 4$ ordered options.
 We simulate the second group having higher probability of responding higher categories of $Y$.
@@
 We can assume that the odds ratio is $OR = 2$.
-For simplicity, the probabilities of the first group $x = 0$ are uniform thus $P(Y = 1|x_0) = ...
-P(Y = k|x_0) = 1/k$.
+For simplicity, the probabilities of the first group $x = 0$ are uniform thus $P(Y = 1|x_0) = \dots P(Y = k|x_0) = 1/k$.
 The following code summarise the first steps of the simulation.
@@
 
-An important step is converting from $\alpha$ to probability and the opposite.
-This steps are implemented in the `alpha_to_prob()` and `prob_to_alpha()` functions.
-Basically given the input and the link function, these functions return the thresholds associated with $k$ probabilities or the $k - 1$ thresholds.
+An important step is converting from $\alpha$ to probabilities or the opposite.
+This steps are implemented in the `alpha_to_prob()` and `prob_to_alpha()` functions that given the input return thresholds or probabilities.
 
@@
 
-# calculate linear predictor using equation TODO put equation
-# obtaining k - 1 equations
+# calculate linear predictor using equation 1 obtaining k - 1 equations
 
@@
 
-# for each row, we can calculate P(Y = k) using equation TODO put equation
+# for each row, we can calculate P(Y = k) using equation 2
 # P(Y = 1) = P(Y <= 1)
@@
 
+set.seed(2024)
+
 # example of a random Y outcome based on the probabilities
@@
 (cum_op <- apply(op, 1, cumsum))
-odds(cum_op[, 1]) / odds(cum_op[, 2])
+odds(cum_op[-k, 1]) / odds(cum_op[-k, 2]) # -k remove the P(y = k) = 1
 
@@
 
-The Figure \@ref(fig:fig-sim-from-latent) depict the simulated $Y^{*}$ and the corresponding ordinal value.
+The Figure \@ref(fig:fig-sim-from-latent) depicts the simulated $Y^{*}$ and the corresponding ordinal value.
 As for the previous simulation we can fit the model using `clm()` and check the estimated parameters.
@@
 
+set.seed(2024)
 N <- 1e3 # sample size
@@
 
-```{r fig-sim-from-latent, fig.cap = filor::get_caption(), fig.place = TRUE}
+```{r fig-sim-from-latent, fig.cap = filor::get_caption(), fig.place = TRUE, fig.asp=0.8}
 alpha_n <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", th_names(k)))
@@
 The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function.
-Details about the function can be found on the Github repository LINK.
 Basically, we define the dataset `dat` with predictors.
@@
 #| collapse: true
+
+set.seed(2024)
 N <- 100 # sample size
@@
 A common research design could be a 2x2 factorial design.
-In this example we have 2 main effects and the interaction.
+In this example we have two main effects and the interaction.
 By default R use dummy coding but setting sum-to-zero contrasts (e.g., $0.5$ and $-0.5$) for a factorial design is convenient.
@@
 \begin{equation}
-P(Y \leq k) = g^{-1}(\alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i})
+P(Y_i \leq k) = g^{-1}[\alpha_k - (\beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i})]
 (\#eq:interaction-equation)
@@
 #| collapse: true
+
+set.seed(2024)
 n <- 100
@@
 
-```{r fig-effects-2-by-2-interaction, fig.place = TRUE}
+```{r fig-effects-2-by-2-interaction, fig.place = TRUE, fig.cap=filor::get_caption()}
 # plot(ggpredict(fit, terms = c("x1", "x2")))
@@
   ylab("Probability") +
-  mtheme() +
   viridis::scale_color_viridis(discrete = TRUE)
@@
 
-```{r fig-effects-num-by-cat-interaction, fig.place = TRUE}
+```{r fig-effects-num-by-cat-interaction, fig.place = TRUE, fig.width=9, fig.height=9, fig.cap=filor::get_caption()}
 # plot(ggpredict(fit, terms = c("x[all]", "g")))
@@
   geom_line(aes(group = group), linewidth = 1) +
-  facet_grid(~response.level) +
+  facet_grid(response.level~1) +
   ylab("Probability") +
@@
 In this section we introduce how to estimate the statistical power.
+Power analysis is a crucial and often necessary step when planning a confirmatory study [@Lakens2022-jp] or preparing a pre-registration or registered report [@Chambers2022-tz].
+Furthermore, statistical power for complex models can only be estimated using Monte Carlo simulations.
+
 Formally the power is defined as the probability of correctly rejecting the null hypothesis $H_0$.
 For simple cases (e.g., a t-test) the power can be calculated analytically while in complex cases simulating data is the best approach.
-We present very general example that, as for the previous simulations, can be extended easily extended.
+We present very general example that, as for the previous simulations, can be easily extended.
 For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting $H_0$) or assessing an eventual bias in model estimates.
@@
 #| collapse: true
+
+set.seed(2024)
 n <- 40 # sample size
@@
 #| collapse: true
+
+set.seed(2024)
 
@@
 
-(ref:fig-ordinal-models) Cumulative model. Adapted from @Burkner2019-aw.
-
-(ref:fig-prop-odds) caption here
-
-(ref:fig-logit-vs-probit) caption here
-
-(ref:fig-example-cat-latent) `cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")`
-
-(ref:fig-example-num-latent) `num_latent_plot(x = runif(100), b1 = 2, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit")`
-
-(ref:fig-explain-cumulative) caption here
-
-(ref:fig-odds-example) caption here
-
-(ref:fig-effects-2-by-2-interaction) `plot(ggpredict(fit, terms = c("x1", "x2")))`
-
-(ref:fig-effects-num-by-cat-interaction) `plot(ggpredict(fit, terms = c("x[all]", "g")))`
-
-(ref:fig-power-curve) caption
-
-(ref:fig-sim-from-latent) caption
-
-(ref:fig-show-th-example) `show_alpha(prob = rep(1/4, 4), link = "probit", plot = c("latent", "probs"))`
+(ref:fig-explain-cumulative) Relationship between cumulative probabilities and ordinal outcomes.
+The $Y^\star$ latent variable is segmented into $k$ levels using $k - 1$ thresholds ($\alpha$).
+The distance between thresholds determine the probability of each outcome.
+The grey thresholds are only used to compute the $Y = 1$ and $Y = k$ probabilities.
+
+(ref:fig-logit-vs-probit) Difference between *logit* and *probit* models. On the left the probability density function (PDF). On the right the cumulative distribution function (CDF).
+
+(ref:fig-odds-example) Relationship between probability (x axis) odds and the log odds. Using log odds ensure a symmetric relationship with a zero midpoint ($p = 0.5$).
+
+(ref:fig-prop-odds) Example of proportional odds (PO, right) and non-proportional odds (NPO, left) for cumulative probabilities and the linear predictor $\eta$ assuming a continuous predictor $x$.
+Data supporting the PO assumption shows an horizontal shift in the cumulative probability with the same slope while for NPO model slopes can be heterogeneous.
+
+(ref:fig-sim-from-latent) Example of cutting a latent variable for $k = 4$ ordinal outcomes with the effect of a binary predictor $x$.
+The thresholds are fixed and the difference between the latent means of the two groups increase the number of points (thus probability) in higher outcomes.
+
+(ref:fig-show-th-example) Example of the `show_alpha()` function. On the left the assumed latent variable and $k - 1$ thresholds with the corresponding probabilities on the right.
+
+(ref:fig-example-cat-latent) Example of the `cat_latent_plot()` function depicting the effect of a categorical predictor.
+On the left the shift in the latent mean and on the right the impact on the expected probabilities.
+The plot can be created using `cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")`.
+
+(ref:fig-example-num-latent) Example of the `num_latent_plot()` function depicting the effect of a continuous predictor on the expected probabilities.
+$\beta_1$ is the regression coefficient.
+The plot can be created using `num_latent_plot(x = runif(100), b1 = 1, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit")`
+
+(ref:fig-effects-2-by-2-interaction) Results from the simulated model with a 2x2 interaction between categorical predictors.
+The plot shows the predicted probabilities for each ordinal outcome as a function of the two predictors.
+A similar plot can be produced using `plot(ggeffects::ggpredict(fit, terms = c("x1", "x2")))`.
+
+(ref:fig-effects-num-by-cat-interaction) Results from the simulated model with the interaction between a continuous and categorical variables.
+The plot shows the predicted probabilities with 95% confidence intervals as a function of predictors.
+A similar plot can be produced using `plot(ggeffects::ggpredict(fit, terms = c("x[all]", "g")))`
+
+(ref:fig-power-curve) Results from the power analysis. The x-axis represents the sample size (per group) and the y axis the estimated power. The red dotted line highlight the 80% power level.
 
@@
 
-(ref:tab-model-summary) caption here
-
-(ref:tab-dataset-example) caption here
+(ref:tab-model-summary-docx) Summary of the probit and logit models in terms of link ($g(\cdot)$) and inverse link function ($g^{-1}(\cdot)$) and the corresponding R code.
+
+(ref:tab-model-summary-latex) (ref:tab-model-summary-docx)
+
+(ref:tab-dataset-example-docx) Summary of the simulated dataset.
+For each group (a and b) we reported mean, median and standard deviation of the ordinal outcome (metric descriptive statistics).
+For each ordinal outcome ($Y$) we reported the frequency and (within parentheses) the probability.
+
+(ref:tab-dataset-example-latex) (ref:tab-dataset-example-docx)

### 2024-01-25 paper/paper.tex
updating after giamma revision

@@
 \documentclass[
-  man,floatsintext]{apa6}
+  man, mask,floatsintext]{apa6}
 \usepackage{amsmath,amssymb}
@@
 {}{\PatchFailed}
-\keywords{ordinal, likert, simulations, power\newline\indent Word count: X}
+\keywords{ordinal, simulations, power\newline\indent Word count: X}
 \usepackage{csquotes}
@@
   pdflang={en-EN},
-  pdfkeywords={ordinal, likert, simulations, power},
+  pdfkeywords={ordinal, simulations, power},
   hidelinks,
@@
 
-Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
-
-Enter author note here.
-
 The authors made the following contributions.
-Filippo Gambarota: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Gianmarco Altoè: Writing - Review \& Editing, Supervision.
-
-Correspondence concerning this article should be addressed to Filippo Gambarota, Postal address. E-mail: \href{mailto:filippo.gambarota@unipd.it}{\nolinkurl{filippo.gambarota@unipd.it}}
+Filippo Gambarota: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing, Methodology, Software; Gianmarco Altoè: Conceptualization, Writing - Review \& Editing, Supervision.
+
+Correspondence concerning this article should be addressed to Filippo Gambarota, Via Venezia 8, 35131, Padova (Italy).
+E-mail: \href{mailto:filippo.gambarota@unipd.it}{\nolinkurl{filippo.gambarota@unipd.it}}
 
@@
 \abstract{%
-Abstract here
+Ordinal data are extremely used in psychology.
+Common examples include Likert items or variables where categories are ordered without assuming a common distance between labels.
+Despite the prevalence of ordinal data, a majority of published psychological research relies on applying metric models (e.g., standard linear regression) even when the dependent variable is ordinal.
+This approach leads to several statistical drawbacks.
+Similar to what happens with binary data (e.g., accuracies in a cognitive experiment), using metric models instead of e.g.~logistic regression is probably caused by several reasons.
+Firstly, parameter interpretation can be challenging.
+Ordinal models commonly used a link function (e.g., logit) for model fitting thus transforming the parameters scale.
+This is even more critical when estimating models with multiple parameters or interactions.
+Secondly, nowadays journals require power or sensitivity analysis when planning or evaluating a research work.
+For common statistical models such as t-tests or the analysis of variance, there are tools to easily estimate the statistical power.
+For ordinal regressions, estimating the statistical power often requires simulating data thus clearly understanding the statistical background and implementing a Monte Carlo simulation.
+The objective of this tutorial is to present the essential statistical background of ordinal regression, highlighting the research areas that would benefit more by introducing this model.
+In particular, we present how to interpret the model parameters with categorical and numerical variables also linking them to common standardized effect size measures.
+Then we propose methods to report the results, especially focusing on graphical representation.
+Finally, we present how to estimate the statistical power using Monte Carlo simulations with suggestions on setting parameters in a meaningful way.
+Through the tutorial, we use the R programming language using specific packages for ordinal models and sharing also ad-hoc developed functions supporting each step.
+For the sake of reproducibility and to empower researchers in analyzing their studies autonomously, all data and scripts used in the article have been made publicly available.
 }
@@
 \maketitle
+
+\scriptsize
+
+\normalsize
 
@@
 Ordinal data, as defined by Stevens (1946), belongs to a specific type of measurement scale where ordered numbers are assigned to a variable.
+Beyond Likert-like items, in Psychology there are several applications of ordinal scales.
+For instance, sociodemographic variables like educational levels or socioeconomic status, as well as general ratings such as pain severity, agreement with a statement, or the evaluation of sensory experiences.
+
 In contrast to nominal scales, the labels in ordinal scales are ordered.
@@
 
-For the tutorial, we assume the reader is already familiar with basic R programming and introductory theory about linear regression.
-(VEDI SE SCRIVERE MEGLIO O QULCOSA DI PIU).
-The current tutorial proposed a set of custom R functions (available on Github ) to understand and simulate basic ordinal regression models.
-More details and extended examples are available on the supplementary materials.
+For the tutorial, we assume that the reader is already familiar with basic R programming and introductory theory about linear regression.
+Despite the practical and applicative spirit of the tutorial, we still introduced some statistical theory and equations supported by translation into R code.
+Similar to other tutorials (DeBruine \& Barr, 2021; Gambarota \& Altoè, 2023), we believe this is the best educational approach especially with the involvement of Monte Carlo simulations.
+
+In the first part we introduce the ordinal models explaining the general structure and assumptions.
+Then we move with model fitting and parameters interpretation.
+Finally we introduce the simulation approach for common research scenarios and an application to power analysis.
+We used R {[}Version 4.3.1; R Core Team (2023){]} for the code, figures and tables.
+Details about R packages an The tutorial is supported by a set of custom R functions available on Github.
+Other examples and model extensions along with details about used R packages are available on the Github repository (supplementary materials).
+For the tutorial we used the following packages.
+
+\scriptsize
+
+\normalsize
 
@@
 
-In this section we introduce some notation for the CM that is used through the paper and in the R code (R Core Team, 2023).
-We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} (Christensen, 2022) R package used in the tutorial.
-We define \(Y_k\) as the observed ordinal response with \(1, \dots, k\) levels and \(Y^\star\) is the underlying latent variable.
+In this section we introduce some notation for the CM that is used through the paper.
+We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} package (Christensen, 2022).
+We define \(Y_k\) as the observed ordinal variable with \(k\) levels and \(Y^\star\) is the underlying latent variable.
 The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\).
@@
 For example, when assuming a Gaussian distribution, fitting a model with a \emph{probit} link function entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
-When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(x\) value \(P(Y \leq k | x_i)\)}.
+When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(\mathbf{X}\) value \(P(Y \leq k | \mathbf{X}_i)\)}.
 Equation \eqref{eq:prob-cum-model1} shows the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
-The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta\) as in the standard regression model (Agresti, 2010) where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories.
+The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta_j\) as in the standard regression models where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories (Agresti, 2010).
 
 \begin{equation}
-P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}), \;\;k = 1, \dots, k - 1
+P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}) \;\;k = 1, \dots, k - 1
 \label{eq:prob-cum-model1}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-explain-cumulative-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-explain-cumulative-1}
 
@@
 
-\caption{caption here}\label{fig:fig-explain-cumulative}
+\caption{Relationship between cumulative probabilities and ordinal outcomes.
+The \(Y^\star\) latent variable is segmented into \(k\) levels using \(k - 1\) thresholds (\(\alpha\)).
+The distance between thresholds determine the probability of each outcome.
+The grey thresholds are only used to compute the \(Y = 1\) and \(Y = k\) probabilities.}\label{fig:fig-explain-cumulative}
 \end{figure}
@@
 \begin{equation}
-Y^\star_i = \mathbf{X}\boldsymbol{\beta} + \epsilon_i
+\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \mathbf{\epsilon}
 \label{eq:latent-model}
@@
 
-The crucial part is \(\epsilon_i\) that is the random component of the model coming from a certain probability distribution.
+The crucial part is \(\epsilon\) that is the random component of the model coming from a certain probability distribution.
 For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution.
@@
 Among several available functions the \emph{logit} and \emph{probit} models are the most common.
-The \emph{logit} model assume a \emph{logit} link function and a logistic distribution as latent variable.
-On the other side, the \emph{probit} model assume a Gaussian distribution.
+The \emph{logit} model use a \emph{logit} link function and a logistic distribution as latent variable.
+On the other side, the \emph{probit} model assume a latent Gaussian distribution using the inverse of the cumulative distribution (\(\Phi^{-1}\)) function as link function.
 
@@
 In the next sections we will illustrate the differences and simulation strategies.
-Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table \ref{tab:tab-model-summary} summarise the presented cumulative models with the proposed link function and the corresponding R code.
+Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table L'output è \ref{tab:tab-model-summary-latex} summarise the presented cumulative models with the proposed link function and the corresponding R code.
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-logit-vs-probit-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-logit-vs-probit-1}
 
@@
 
-\caption{caption here}\label{fig:fig-logit-vs-probit}
+\caption{Difference between \emph{logit} and \emph{probit} models.
+On the left the probability density function (PDF).
+On the right the cumulative distribution function (CDF).}\label{fig:fig-logit-vs-probit}
 \end{figure}
@@
 
+\normalsize
+
+\scriptsize
+
 \begin{table}
 
-\caption{\label{tab:tab-model-summary}caption here}
+\caption{\label{tab:tab-model-summary-latex}(ref:tab-model-summary-docx)}
 \centering
-\begin{tabular}[t]{lllll}
+\begin{tabular}[t]{ccccc}
 \toprule
-\multicolumn{1}{c}{} & \multicolumn{2}{c}{Link Function} & \multicolumn{2}{c}{Inverse Link Function} \\
+\multicolumn{1}{c}{} & \multicolumn{2}{c}{Link Function $g(.)$} & \multicolumn{2}{c}{Inverse Link Function $g^{-1}(.)$} \\
 \cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Model & Equation & R Code & Equation & R Code\\
+\textbf{Model} & \textbf{Equation} & \textbf{R Code} & \textbf{Equation} & \textbf{R Code}\\
 \midrule
-Cumulative Logit & $\text{logit}(p) = \text{log}(p / (1-p))$ & `qlogis()` & $e^{\text{logit}(p)} / (1 + e^{\text{logit}(p)})$ & `plogis()`\\
-Cumulative Probit & $z = \Phi^{-1}(p)$ & `qnorm()` & $\Phi^(z)$ & `pnorm()`\\
+Cumulative Probit & $z = \Phi^{-1}(p)$ & \texttt{qnorm()} & $p = \Phi(z)$ & \texttt{pnorm()}\\
 \bottomrule
@@
 
+\scriptsize
+
+\normalsize
+
 \section{Model fitting}\label{model-fitting}
@@
 We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
-Table \ref{tab:tab-dataset-example} contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options.
-The participants are divided into two groups (\(x\)).
+Table \ref{tab:tab-dataset-example-latex} contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options.
+The participants are divided into two groups (\(x_a\) and \(x_b\)).
 We can fit a cumulative link model with \texttt{clm()} function and check the model summary.
@@
 
+\normalsize
+
+\scriptsize
+
 \begin{table}
 
-\caption{\label{tab:tab-dataset-example}caption here}
+\caption{\label{tab:tab-dataset-example-latex}(ref:tab-dataset-example-docx)}
 \centering
-\begin{tabular}[t]{llllllll}
+\begin{tabular}[t]{cccccccc}
 \toprule
-group & mean & median & sd & Y1 & Y2 & Y3 & Y4\\
+\textbf{Group} & \textbf{Mean} & \textbf{Median} & \textbf{SD} & \textbf{Y1} & \textbf{Y2} & \textbf{Y3} & \textbf{Y4}\\
 \midrule
-a & 2.84 & 3 & 1.11 & 8 (p = 0.16) & 11 (p = 0.22) & 12 (p = 0.24) & 19 (p = 0.38)\\
-b & 3.02 & 3 & 1.08 & 5 (p = 0.10) & 13 (p = 0.26) & 8 (p = 0.16) & 24 (p = 0.48)\\
+a & 2.16 & 2 & 1.09 & 19 (p = 0.38) & 11 (p = 0.22) & 13 (p = 0.26) & 7 (p = 0.14)\\
+b & 3.12 & 4 & 1.1 & 7 (p = 0.14) & 6 (p = 0.12) & 11 (p = 0.22) & 26 (p = 0.52)\\
 \bottomrule
@@
 \end{table}
+
+\normalsize
+
+\scriptsize
 
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  100  {-}128.88 265.77 5(0)  9.09e{-}13 2.2e+01}
+\CommentTok{\#\textgreater{}  logit flexible  100  {-}127.04 262.08 4(0)  5.42e{-}07 2.0e+01}
 \CommentTok{\#\textgreater{} }
 \CommentTok{\#\textgreater{} Coefficients:}
-\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)}
-\CommentTok{\#\textgreater{} xb   0.3153     0.3670   0.859     0.39}
+\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)    }
+\CommentTok{\#\textgreater{} xb   1.5929     0.3895    4.09 4.32e{-}05 ***}
+\CommentTok{\#\textgreater{} {-}{-}{-}}
+\CommentTok{\#\textgreater{} Signif.
+codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}1.7503     0.3432  {-}5.100}
-\CommentTok{\#\textgreater{} 2|3  {-}0.3791     0.2722  {-}1.393}
-\CommentTok{\#\textgreater{} 3|4   0.4381     0.2717   1.612}
+\CommentTok{\#\textgreater{} 1|2  {-}0.4243     0.2745  {-}1.546}
+\CommentTok{\#\textgreater{} 2|3   0.4293     0.2711   1.584}
+\CommentTok{\#\textgreater{} 3|4   1.5851     0.3165   5.008}
 \end{Highlighting}
@@
 The two main sections of the model summary are the \emph{Coefficients} section reporting the regression coefficients \(\beta\) and the \emph{Threshold} section reporting the \(\alpha\) estimation.
-Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect.
-As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding (see Schad, Vasishth, Hohenstein, \& Kliegl, 2020).
+Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and one \(\beta\) associated with the \(x\) effect.
+As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients (plus the intercept term) where the interpretation depends on the contrast coding (see Schad, Vasishth, Hohenstein, \& Kliegl, 2020).
 In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables.
@@
 To understand the logit model we need to introduce odds and odds ratio.
-The odds of a probability \(p\) is defined as \(\frac{p}{1 - p}\) thus the probability of success divided by the probability of failure.
+The odds of a probability \(p\) is defined as \(p/(1 - p)\) thus the probability of success divided by the probability of failure.
 The odds takes value ranging from 0 to \(\infty\).
@@
 The Figure \ref{fig:fig-odds-example} shows the relationship between probabilities and odds.
-The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from \(-\infty\) to \(\infty\) with \(p = 0.5\) as the midpoint because \(\text{log}(\frac{0.5}{1 - 0.5}) = 0\).
+The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from \(-\infty\) to \(\infty\) with \(p = 0.5\) as the midpoint because \(\text{log}(0.5/(1 - 0.5)) = 0\).
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-odds-example-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-odds-example-1}
 
@@
 
-\caption{caption here}\label{fig:fig-odds-example}
+\caption{Relationship between probability (x axis) odds and the log odds. Using log odds ensure a symmetric relationship with a zero midpoint (\(p = 0.5\)).}\label{fig:fig-odds-example}
 \end{figure}
@@
 \NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\CommentTok{\#\textgreater{} [1] 0.16}
+\CommentTok{\#\textgreater{} [1] 0.38}
 
@@
 \FunctionTok{odds}\NormalTok{(p1a)}
-\CommentTok{\#\textgreater{} [1] 0.1904762}
+\CommentTok{\#\textgreater{} [1] 0.6129032}
 
@@
 \NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\CommentTok{\#\textgreater{} [1] 0.1}
+\CommentTok{\#\textgreater{} [1] 0.14}
 \FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
-\CommentTok{\#\textgreater{} [1] 1.714286}
+\CommentTok{\#\textgreater{} [1] 3.764977}
 \end{Highlighting}
@@
 
-For example the probability of responding \(Y = 1\) in the group ``a'' is 0.16 corresponding to an odds of 0.19.
-When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 1.71 times the odds of responding \(Y = 1\) compared to the group ``b''.
-
-For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(p = 1\)).
+For example the probability of responding \(Y = 1\) in the group ``a'' is 0.38 corresponding to an odds of 0.61.
+When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 3.76 times the odds of responding \(Y = 1\) compared to the group ``b''.
+
+For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(P = 1\)).
 Thus we have \(k - 1\) odds ratio or what is called a cumulative odds ratio.
 Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\).
-The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 1.37.
+The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 4.92.
 In fact, the logit model essentially estimate the cumulative odds ratio.
@@
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.16    0.38    0.62}
+\CommentTok{\#\textgreater{}    0.38    0.60    0.86}
 \NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.10    0.36    0.52}
+\CommentTok{\#\textgreater{}    0.14    0.26    0.48}
 
@@
 \CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
-\CommentTok{\#\textgreater{} 1.714286 1.089606 1.506073}
+\CommentTok{\#\textgreater{} 3.764977 4.269231 6.654762}
 \end{Highlighting}
@@
 
-From the previous example, we noticed that the \(k - 1\) odds ratios are different.
+From the previous example, we noticed that the \(k - 1\) estimated odds ratios are different.
 However, increasing \(n\) will result in odds ratios that are very similar.
@@
 Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the PO assumption.
-There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}) (Peterson \& Harrell, 1990).
+There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}, Peterson \& Harrell, 1990).
+
+\scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
+\CommentTok{\# using same parameters as the previous simulation increasing n}
+\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e5} \CommentTok{\# very large n}
+\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{beta =}\NormalTok{ b1, }\AttributeTok{prob0 =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+
+\CommentTok{\# same steps as before}
+\NormalTok{cum\_p }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(}\AttributeTok{x =}\NormalTok{ dat}\SpecialCharTok{$}\NormalTok{x, }\FunctionTok{dummy\_ord}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y))}
+\NormalTok{group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean)}
+\NormalTok{group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean)}
+\CommentTok{\# now odds are almost the same and equal to the simulated b1 = \textasciitilde{}1.09}
+\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b)}
+\CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
+\CommentTok{\#\textgreater{} 2.914579 2.985242 2.991333}
+\FunctionTok{log}\NormalTok{(}\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b))}
+\CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
+\CommentTok{\#\textgreater{} 1.069725 1.093681 1.095719}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-prop-odds-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-prop-odds-1}
 
@@
 
-\caption{caption here}\label{fig:fig-prop-odds}
+\caption{Example of proportional odds (PO, right) and non-proportional odds (NPO, left) for cumulative probabilities and the linear predictor \(\eta\) assuming a continuous predictor \(x\).
+Data supporting the PO assumption shows an horizontal shift in the cumulative probability with the same slope while for NPO model slopes can be heterogeneous.}\label{fig:fig-prop-odds}
 \end{figure}
@@
 \FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 1.370663}
+\CommentTok{\#\textgreater{} [1] 2.976954}
 \FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 1.370663}
+\CommentTok{\#\textgreater{} [1] 2.976954}
 \end{Highlighting}
@@
 
-\subsection{\texorpdfstring{\emph{Probit Model} \(\mathbf{z}\) scores}{Probit Model \textbackslash mathbf\{z\} scores}}\label{probit-model-mathbfz-scores}
+\subsection{\texorpdfstring{\emph{Probit Model} and \(\mathbf{z}\) scores}{Probit Model and \textbackslash mathbf\{z\} scores}}\label{probit-model-and-mathbfz-scores}
 
@@
 Thinking about the latent distributions, the \(\beta\) is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric \(x\).
-More formally the shift in the latent distribution is \(\frac{\beta}{\sigma}\) (for the \emph{probit} model \(\sigma = 1\)).
+More formally the shift in the latent distribution is \(\beta/\sigma\) (for the \emph{probit} model \(\sigma = 1\)).
 The interpretation in terms of shifting the latent mean holds also for the logistic model.
@@
 For the standard logistic distribution \(s = 1\) (as for the standard normal distribution).}.
-The \(\beta\) for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by \(\beta/(\frac{\pi}{\sqrt{3}})\) standard deviations.
+The \(\beta\) for the logistic distribution can be interpreted as the location shift of the latent logistic distribution by \(\beta/(\frac{\pi}{\sqrt{3}})\) standard deviations (Agresti, 2010).
 
@@
 \begin{equation}
-z_{x_1 - x_0} = \Phi(P \leq 1|x_0) - \Phi(P \leq 1|x_1) = \dots = \Phi(P \leq k - 1|x_0) - \Phi(P \leq k - 1|x_1)
+z_{x_1 - x_0} = \Phi(P \leq 1|x_1) - \Phi(P \leq 1|x_0) = \dots = \Phi(P \leq k - 1|x_1) - \Phi(P \leq k - 1|x_0)
 \label{eq:probit-prop-odds}
@@
 \FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 0.204808}
+\CommentTok{\#\textgreater{} [1] 0.6523531}
 \FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 0.204808}
+\CommentTok{\#\textgreater{} [1] 0.6523531}
 \end{Highlighting}
@@
 In this tutorial we present two method for simulating ordinal data.
-The first method calculate the probabilities of each \(Y\) level as a function of predictors and generate data from a multinomial distribution.
+Simulating data is a powerful strategy to understand the model (DeBruine \& Barr, 2021) and estimate statistical properties (e.g., power or type-1 error).
+The first simulation method calculate the probabilities of each \(Y\) level as a function of predictors and generate data from a \emph{multinomial} distribution.
 The second method simulate data using the the latent formulation.
+Whenever random number generation occurs, it is appropriate to set a \texttt{seed} (\texttt{set.seed()} function, in this case we use \texttt{set.seed(2024)}).
+Running again the same code will produce the same result.
+A general simulation approach concerns generating a dataset from the assumed data generation process, fitting the statistical model and assessing the recovery of simulated parameters.
+To check the simulation approach, using a large sample size produce estimations with small sampling error.
+A limited sample size even when fitting the true model will produce variable estimations.
 
@@
 For the first method we need to calculate \(g^{-1}(\eta)\) as a function of predictors and then sample from a \emph{multinomial} (more specifically \emph{categorical}) distribution using the \texttt{sample()} function in R.
-This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview)
-
-As a simple example, let's simulate two groups with \(n = 100\) participants responding to an item with \(k = 4\) ordered options.
+This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview).
+
+As a simple example, let's simulate two groups with \(n = 100\) participants responding to an item (\(Y\)) with \(k = 4\) ordered options.
 We simulate the second group having higher probability of responding higher categories of \(Y\).
@@
 We can assume that the odds ratio is \(OR = 2\).
-For simplicity, the probabilities of the first group \(x = 0\) are uniform thus \(P(Y = 1|x_0) = ...
-P(Y = k|x_0) = 1/k\).
+For simplicity, the probabilities of the first group \(x = 0\) are uniform thus \(P(Y = 1|x_0) = \dots P(Y = k|x_0) = 1/k\).
 The following code summarise the first steps of the simulation.
@@
 
-An important step is converting from \(\alpha\) to probability and the opposite.
-This steps are implemented in the \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} functions.
-Basically given the input and the link function, these functions return the thresholds associated with \(k\) probabilities or the \(k - 1\) thresholds.
+An important step is converting from \(\alpha\) to probabilities or the opposite.
+This steps are implemented in the \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} functions that given the input return thresholds or probabilities.
 
@@
 
-\CommentTok{\# calculate linear predictor using equation }\AlertTok{TODO}\CommentTok{ put equation}
-\CommentTok{\# obtaining k {-} 1 equations}
+\CommentTok{\# calculate linear predictor using equation 1 obtaining k {-} 1 equations}
 
@@
 
-\CommentTok{\# for each row, we can calculate P(Y = k) using equation }\AlertTok{TODO}\CommentTok{ put equation}
+\CommentTok{\# for each row, we can calculate P(Y = k) using equation 2}
 \CommentTok{\# P(Y = 1) = P(Y \textless{}= 1)}
@@
 
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
+
 \CommentTok{\# example of a random Y outcome based on the probabilities}
 \FunctionTok{sample}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ p[}\DecValTok{1}\NormalTok{, ])}
-\CommentTok{\#\textgreater{} [1] 3}
+\CommentTok{\#\textgreater{} [1] 1}
 
@@
 \CommentTok{\#\textgreater{}   x y}
-\CommentTok{\#\textgreater{} 1 0 4}
-\CommentTok{\#\textgreater{} 2 0 1}
-\CommentTok{\#\textgreater{} 3 0 2}
+\CommentTok{\#\textgreater{} 1 0 3}
+\CommentTok{\#\textgreater{} 2 0 4}
+\CommentTok{\#\textgreater{} 3 0 4}
 \CommentTok{\#\textgreater{} 4 0 3}
-\CommentTok{\#\textgreater{} 5 0 1}
-\CommentTok{\#\textgreater{} 6 0 4}
+\CommentTok{\#\textgreater{} 5 0 4}
+\CommentTok{\#\textgreater{} 6 0 3}
 
@@
 \CommentTok{\#\textgreater{}        1    2    3    4}
-\CommentTok{\#\textgreater{}   0 0.26 0.24 0.26 0.24}
-\CommentTok{\#\textgreater{}   1 0.18 0.12 0.28 0.42}
+\CommentTok{\#\textgreater{}   0 0.26 0.22 0.26 0.26}
+\CommentTok{\#\textgreater{}   1 0.10 0.18 0.30 0.42}
 
@@
 \CommentTok{\#\textgreater{}        0    1}
-\CommentTok{\#\textgreater{}   1 0.26 0.18}
-\CommentTok{\#\textgreater{}   2 0.50 0.30}
-\CommentTok{\#\textgreater{}   3 0.76 0.58}
+\CommentTok{\#\textgreater{}   1 0.26 0.10}
+\CommentTok{\#\textgreater{}   2 0.48 0.28}
+\CommentTok{\#\textgreater{}   3 0.74 0.58}
 \CommentTok{\#\textgreater{}   4 1.00 1.00}
-\FunctionTok{odds}\NormalTok{(cum\_op[, }\DecValTok{1}\NormalTok{]) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(cum\_op[, }\DecValTok{2}\NormalTok{])}
-\CommentTok{\#\textgreater{}        1        2        3        4 }
-\CommentTok{\#\textgreater{} 1.600601 2.333333 2.293103      NaN}
+\FunctionTok{odds}\NormalTok{(cum\_op[}\SpecialCharTok{{-}}\NormalTok{k, }\DecValTok{1}\NormalTok{]) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(cum\_op[}\SpecialCharTok{{-}}\NormalTok{k, }\DecValTok{2}\NormalTok{]) }\CommentTok{\# {-}k remove the P(y = k) = 1}
+\CommentTok{\#\textgreater{}        1        2        3 }
+\CommentTok{\#\textgreater{} 3.162162 2.373626 2.061008}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}134.09 276.18 6(0)  2.78e{-}08 1.9e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}132.59 273.17 5(0)  2.86e{-}09 1.7e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)  }
-\CommentTok{\#\textgreater{} x   0.4420     0.2198   2.011   0.0443 *}
+\CommentTok{\#\textgreater{} x   0.5183     0.2197   2.359   0.0183 *}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.57303    0.17219  {-}3.328}
-\CommentTok{\#\textgreater{} 2|3 {-}0.04306    0.16493  {-}0.261}
-\CommentTok{\#\textgreater{} 3|4  0.67182    0.17437   3.853}
+\CommentTok{\#\textgreater{} 1|2 {-}0.68708    0.17711  {-}3.879}
+\CommentTok{\#\textgreater{} 2|3 {-}0.05305    0.16717  {-}0.317}
+\CommentTok{\#\textgreater{} 3|4  0.68706    0.17462   3.935}
 \end{Highlighting}
@@
 
-The Figure \ref{fig:fig-sim-from-latent} depict the simulated \(Y^{*}\) and the corresponding ordinal value.
+The Figure \ref{fig:fig-sim-from-latent} depicts the simulated \(Y^{*}\) and the corresponding ordinal value.
 As for the previous simulation we can fit the model using \texttt{clm()} and check the estimated parameters.
@@
 
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
 \NormalTok{N }\OtherTok{\textless{}{-}} \FloatTok{1e3} \CommentTok{\# sample size}
@@
 \CommentTok{\#\textgreater{}      x       lp      ystar y}
-\CommentTok{\#\textgreater{} 1    0 0.000000  0.9531361 3}
-\CommentTok{\#\textgreater{} 2    0 0.000000  1.1862471 4}
-\CommentTok{\#\textgreater{} 3    0 0.000000 {-}0.7356740 2}
-\CommentTok{\#\textgreater{} 998  1 1.386294 {-}5.0039165 1}
-\CommentTok{\#\textgreater{} 999  1 1.386294  1.8689381 4}
-\CommentTok{\#\textgreater{} 1000 1 1.386294  4.0016754 4}
+\CommentTok{\#\textgreater{} 1    0 0.000000  1.6356528 4}
+\CommentTok{\#\textgreater{} 2    0 0.000000 {-}0.7497878 2}
+\CommentTok{\#\textgreater{} 3    0 0.000000  0.7554420 3}
+\CommentTok{\#\textgreater{} 998  1 1.386294  2.2813322 4}
+\CommentTok{\#\textgreater{} 999  1 1.386294  2.2839112 4}
+\CommentTok{\#\textgreater{} 1000 1 1.386294  2.0325030 4}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-sim-from-latent-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-sim-from-latent-1}
 
@@
 
-\caption{caption}\label{fig:fig-sim-from-latent}
+\caption{Example of cutting a latent variable for \(k = 4\) ordinal outcomes with the effect of a binary predictor \(x\).
+The thresholds are fixed and the difference between the latent means of the two groups increase the number of points (thus probability) in higher outcomes.}\label{fig:fig-sim-from-latent}
 \end{figure}
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik   AIC     niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1268.84 2545.68 4(0)  2.84e{-}08 1.7e+01}
+\CommentTok{\#\textgreater{}  logit flexible  1000 {-}1226.66 2461.32 4(0)  2.25e{-}07 1.7e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x    1.429      0.122   11.71   \textless{}2e{-}16 ***}
+\CommentTok{\#\textgreater{} x   1.4078     0.1233   11.41   \textless{}2e{-}16 ***}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}1.01156    0.09476 {-}10.675}
-\CommentTok{\#\textgreater{} 2|3  0.16670    0.08565   1.946}
-\CommentTok{\#\textgreater{} 3|4  1.26980    0.09575  13.262}
+\CommentTok{\#\textgreater{} 1|2 {-}1.13646    0.09757 {-}11.648}
+\CommentTok{\#\textgreater{} 2|3 {-}0.07934    0.08603  {-}0.922}
+\CommentTok{\#\textgreater{} 3|4  0.98793    0.09269  10.658}
 \end{Highlighting}
@@
 The simulation using the latent formulation of the model is implemented in the \texttt{sim\_ord\_latent()} function.
-Details about the function can be found on the Github repository LINK.
 Basically, we define the dataset \texttt{dat} with predictors.
@@
 \begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
 \NormalTok{N }\OtherTok{\textless{}{-}} \DecValTok{100} \CommentTok{\# sample size}
@@
 \CommentTok{\#\textgreater{}     x y         ys}
-\CommentTok{\#\textgreater{} 1   0 3  0.0555103}
-\CommentTok{\#\textgreater{} 2   0 4  1.2882548}
-\CommentTok{\#\textgreater{} 3   0 4  2.0276663}
-\CommentTok{\#\textgreater{} 98  1 2 {-}1.0176796}
-\CommentTok{\#\textgreater{} 99  1 4  1.4546417}
-\CommentTok{\#\textgreater{} 100 1 2 {-}0.5004862}
+\CommentTok{\#\textgreater{} 1   0 4  1.6356528}
+\CommentTok{\#\textgreater{} 2   0 2 {-}0.7497878}
+\CommentTok{\#\textgreater{} 3   0 3  0.7554420}
+\CommentTok{\#\textgreater{} 98  1 4  1.4134548}
+\CommentTok{\#\textgreater{} 99  1 4  5.0032143}
+\CommentTok{\#\textgreater{} 100 1 4  1.6703959}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-show-th-example-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-show-th-example-1}
 
@@
 
-\caption{\texttt{show\_alpha(prob\ =\ rep(1/4,\ 4),\ link\ =\ "probit",\ plot\ =\ c("latent",\ "probs"))}}\label{fig:fig-show-th-example}
+\caption{Example of the \texttt{show\_alpha()} function.
+On the left the assumed latent variable and \(k - 1\) thresholds with the corresponding probabilities on the right.}\label{fig:fig-show-th-example}
 \end{figure}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-example-cat-latent-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-example-cat-latent-1}
 
@@
 
-\caption{\texttt{cat\_latent\_plot(m\ =\ c(0,\ 0.5),\ s\ =\ 1,\ probs\ =\ rep(1/4,\ 4),\ link\ =\ "logit",\ plot\ =\ "both")}}\label{fig:fig-example-cat-latent}
+\caption{Example of the \texttt{cat\_latent\_plot()} function depicting the effect of a categorical predictor.
+On the left the shift in the latent mean and on the right the impact on the expected probabilities.
+The plot can be created using \texttt{cat\_latent\_plot(m\ =\ c(0,\ 0.5),\ s\ =\ 1,\ probs\ =\ rep(1/4,\ 4),\ link\ =\ "logit",\ plot\ =\ "both")}.}\label{fig:fig-example-cat-latent}
 \end{figure}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-example-num-latent-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-example-num-latent-1}
 
@@
 
-\caption{\texttt{num\_latent\_plot(x\ =\ runif(100),\ b1\ =\ 2,\ probs\ =\ c(0.6,\ 0.2,\ 0.1,\ 0.1),\ link\ =\ "probit")}}\label{fig:fig-example-num-latent}
+\caption{Example of the \texttt{num\_latent\_plot()} function depicting the effect of a continuous predictor on the expected probabilities.
+\(\beta_1\) is the regression coefficient.
+The plot can be created using \texttt{num\_latent\_plot(x\ =\ runif(100),\ b1\ =\ 1,\ probs\ =\ c(0.6,\ 0.2,\ 0.1,\ 0.1),\ link\ =\ "probit")}}\label{fig:fig-example-num-latent}
 \end{figure}
@@
 A common research design could be a 2x2 factorial design.
-In this example we have 2 main effects and the interaction.
+In this example we have two main effects and the interaction.
 By default R use dummy coding but setting sum-to-zero contrasts (e.g., \(0.5\) and \(-0.5\)) for a factorial design is convenient.
@@
 \begin{equation}
-P(Y \leq k) = g^{-1}(\alpha_k - \beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i})
+P(Y_i \leq k) = g^{-1}[\alpha_k - (\beta_1X_{1_i} + \beta_2X_{2_i} + \beta_3X_{1_i}X_{2_i})]
 \label{eq:interaction-equation}
@@
 \begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
 \NormalTok{n }\OtherTok{\textless{}{-}} \DecValTok{100}
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  400  {-}400.35 810.70 6(0)  2.95e{-}13 2.8e+01}
+\CommentTok{\#\textgreater{}  probit flexible  400  {-}390.29 790.57 4(0)  2.20e{-}07 2.1e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}         Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x11       0.2514     0.1171   2.147   0.0318 *  }
-\CommentTok{\#\textgreater{} x21       0.8774     0.1181   7.427 1.11e{-}13 ***}
-\CommentTok{\#\textgreater{} x11:x21   0.5215     0.2343   2.226   0.0260 *  }
+\CommentTok{\#\textgreater{} x11     {-}0.02969    0.11625  {-}0.255    0.798    }
+\CommentTok{\#\textgreater{} x21      1.15703    0.11935   9.695   \textless{}2e{-}16 ***}
+\CommentTok{\#\textgreater{} x11:x21  0.29058    0.23254   1.250    0.211    }
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.30679    0.06605  {-}4.645}
-\CommentTok{\#\textgreater{} 2|3  0.40471    0.06714   6.028}
+\CommentTok{\#\textgreater{} 1|2 {-}0.50503    0.06932  {-}7.286}
+\CommentTok{\#\textgreater{} 2|3  0.46102    0.06900   6.681}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-effects-2-by-2-interaction-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-effects-2-by-2-interaction-1}
 
@@
 
-\caption{ }\label{fig:fig-effects-2-by-2-interaction}
+\caption{Results from the simulated model with a 2x2 interaction between categorical predictors.
+The plot shows the predicted probabilities for each ordinal outcome as a function of the two predictors.
+A similar plot can be produced using \texttt{plot(ggeffects::ggpredict(fit,\ terms\ =\ c("x1",\ "x2")))}.}\label{fig:fig-effects-2-by-2-interaction}
 \end{figure}
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}95.21 200.41 5(0)  3.33e{-}14 9.2e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}101.63 213.26 5(0)  3.01e{-}08 1.0e+02}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}      Estimate Std. Error z value Pr(\textgreater{}|z|)  }
-\CommentTok{\#\textgreater{} x      1.4377     0.4303   3.341 0.000835 ***}
-\CommentTok{\#\textgreater{} g1     0.4508     0.5041   0.894 0.371205    }
-\CommentTok{\#\textgreater{} x:g1   1.1330     0.8536   1.327 0.184431    }
+\CommentTok{\#\textgreater{} x     0.68616    0.40959   1.675   0.0939 .}
+\CommentTok{\#\textgreater{} g1    0.07207    0.45768   0.157   0.8749  }
+\CommentTok{\#\textgreater{} x:g1  1.21471    0.81881   1.484   0.1379  }
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2   0.2913     0.2553   1.141}
-\CommentTok{\#\textgreater{} 2|3   1.1938     0.2731   4.371}
+\CommentTok{\#\textgreater{} 1|2  {-}0.1059     0.2352  {-}0.450}
+\CommentTok{\#\textgreater{} 2|3   0.6529     0.2408   2.711}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-effects-num-by-cat-interaction-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-effects-num-by-cat-interaction-1}
 
@@
 
-\caption{ }\label{fig:fig-effects-num-by-cat-interaction}
+\caption{Results from the simulated model with the interaction between a continuous and categorical variables.
+The plot shows the predicted probabilities with 95\% confidence intervals as a function of predictors.
+A similar plot can be produced using \texttt{plot(ggeffects::ggpredict(fit,\ terms\ =\ c("x{[}all{]}",\ "g")))}}\label{fig:fig-effects-num-by-cat-interaction}
 \end{figure}
@@
 In this section we introduce how to estimate the statistical power.
+Power analysis is a crucial and often necessary step when planning a confirmatory study (Lakens, 2022) or preparing a pre-registration or registered report (Chambers \& Tzavella, 2022).
+Furthermore, statistical power for complex models can only be estimated using Monte Carlo simulations.
+
 Formally the power is defined as the probability of correctly rejecting the null hypothesis \(H_0\).
 For simple cases (e.g., a t-test) the power can be calculated analytically while in complex cases simulating data is the best approach.
-We present very general example that, as for the previous simulations, can be extended easily extended.
+We present very general example that, as for the previous simulations, can be easily extended.
 For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting \(H_0\)) or assessing an eventual bias in model estimates.
@@
 \begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
 \NormalTok{n }\OtherTok{\textless{}{-}} \DecValTok{40} \CommentTok{\# sample size}
@@
 \FunctionTok{mean}\NormalTok{(p }\SpecialCharTok{\textless{}=}\NormalTok{ alpha, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)}
-\CommentTok{\#\textgreater{} [1] 0.382}
+\CommentTok{\#\textgreater{} [1] 0.418}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
+
 \NormalTok{n }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{40}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{100}\NormalTok{, }\DecValTok{200}\NormalTok{)}
@@
 \NormalTok{power}
-\CommentTok{\#\textgreater{} [1] 0.229 0.383 0.540 0.745 0.971}
+\CommentTok{\#\textgreater{} [1] 0.225 0.412 0.522 0.750 0.964}
 \end{Highlighting}
@@
 
-{\centering \includegraphics{paper_files/figure-latex/fig-power-curve-1}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-power-curve-1}
 
@@
 
-\caption{caption}\label{fig:fig-power-curve}
+\caption{Results from the power analysis.
+The x-axis represents the sample size (per group) and the y axis the estimated power.
+The red dotted line highlight the 80\% power level.}\label{fig:fig-power-curve}
 \end{figure}
@@
 
+\bibitem[\citeproctext]{ref-Chambers2022-tz}
+Chambers, C. D., \& Tzavella, L. (2022). The past, present and future of registered reports. \emph{Nature Human Behaviour}, \emph{6}, 29--42. \url{https://doi.org/10.1038/s41562-021-01193-7}
+
 \bibitem[\citeproctext]{ref-Christensen2019-cz}
@@
 \url{https://doi.org/10.1002/sim.4780141105}
+
+\bibitem[\citeproctext]{ref-DeBruine2021-id}
+DeBruine, L.
+M., \& Barr, D.
+J.
+(2021).
+Understanding mixed-effects models through data simulation.
+\emph{Advances in Methods and Practices in Psychological Science}, \emph{4}, 2515245920965119.
+\url{https://doi.org/10.1177/2515245920965119}
 
@@
 
+\bibitem[\citeproctext]{ref-Gambarota2023-on}
+Gambarota, F., \& Altoè, G.
+(2023).
+Understanding meta-analysis through data simulation with applications to power analysis.
+\emph{Advances in Methods and Practices in Psychological Science}.
+\url{https://doi.org/10.1177/25152459231209330}
+
 \bibitem[\citeproctext]{ref-Gelman2020-tg}
@@
 Kruschke, J. K. (2015). \emph{Doing bayesian data analysis}. \url{https://doi.org/10.1016/c2012-0-00477-2}
+
+\bibitem[\citeproctext]{ref-Lakens2022-jp}
+Lakens, D. (2022). Sample size justification. \emph{Collabra. Psychology}, \emph{8}. \url{https://doi.org/10.1525/collabra.33267}
 
@@
 
-\bibitem[\citeproctext]{ref-r-lang}
-R Core Team.
-(2023).
-\emph{R: A language and environment for statistical computing} {[}Computer software{]}.
-Vienna, Austria: R Foundation for Statistical Computing.
-Retrieved from \url{https://www.R-project.org/}
+\bibitem[\citeproctext]{ref-R-base}
+R Core Team. (2023). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}
 

### 2024-01-25 R/deprecated/deprecated.R

@@ -119,3 +119,47 @@ get_probs <- function(formula,
   }
   return(P)
 }
+
+
+qrender <- function(){
+  rmarkdown::render("paper/paper.Rmd")
+  system("quarto render")
+}
+
+mkdirif <- function(dir){
+  if(!fs::is_dir(dir)){
+    fs::dir_create(dir)
+  }
+}
+
+render <- function(file){
+  db <- getdb()
+  parent <- dirname(file)
+  current_md5 <- tools::md5sum(file)
+  update <- !(db$md5[db$files == names(current_md5)] == current_md5)
+  if(update){
+    out <- rmarkdown::render(file)
+    db$md5[db$files == names(current_md5)] <- current_md5
+    saveRDS(db, ".dbcache/dbcache.rds")
+    return(out)
+  }else{
+    cat("file not changed!")
+  }
+}
+
+getdb <- function(){
+  mkdirif(file.path(".dbcache"))
+  dbpath <- here::here(".dbcache", "dbcache.rds")
+  if(!fs::file_exists(dbpath)){
+    files <- list.files(pattern = "Rmd", full.names = TRUE, recursive = TRUE)
+    md5 <- tools::md5sum(files)
+    db <- data.frame(
+      file = gsub("\\./", "", names(md5)),
+      md5 = md5
+    )
+    names(db) <- c("files", "md5")
+    rownames(db) <- NULL
+    saveRDS(db, dbpath)
+  }
+  db <- readRDS(dbpath)
+}

### 2024-01-25 R/utils-paper.R

@@ -116,3 +116,38 @@ qapa <- function(data, caption){
 rm_legend <- function(){
   theme(legend.position = "none")
 }
+
+replace_before_after <- function(x, before, after = NULL, new){
+  if(is.null(after)) after <- before
+  whole <- sprintf("%s.*%s", before, after)
+  within <- sprintf("(?<=%s).*(?=%s)", before, after)
+  within_string <- regmatches(x, regexpr(within, x, perl = TRUE))
+  new_string <- sprintf(new, within_string)
+  x[grepl(whole, x)] <- stringr::str_replace_all(x[grepl(whole, x)], whole, new_string)
+  return(x)
+}
+
+fmt <- function(x, what, type = "markdown", digits = 2, ...){
+  type <- match.arg(type, c("latex", "markdown"))
+  what <- match.arg(what, c("bold", "italic"))
+  if(is.numeric(x)) x <- round(x, digits)
+  fun <- switch(what,
+    "bold" = bold,
+    "italic" = italic
+  )
+  fun(x, type, ...)
+}
+
+bold <- function(x, type = "markdown"){
+  switch(type,
+    "markdown" = sprintf("**%s**", x),
+    "latex" = sprintf("\\textbf{%s}", x)
+  )
+}
+
+italic <- function(x, type = "markdown"){
+  switch(type,
+         "markdown" = sprintf("*%s*", x),
+         "latex" = sprintf("\\textit{%s}", x)
+  )
+}

### 2024-01-25 R/utils-plotting.R

@@ -179,6 +179,7 @@ num_latent_plot <- function(x,
   }
   data <- data.frame(x = seq(min(x), max(x), length.out = nsample))
   data <- sim_ord_latent(~x, beta = b1, prob0 = probs, data = data, link = link, simulate = FALSE)
+  data <- data[, !grepl("^yp", names(data))]
   datal <- tidyr::pivot_longer(data, starts_with("y"), names_to = "y", values_to = "value")
   ggplot(datal, aes(x = x, y = value, color = y)) +
     geom_line(linewidth = linewidth) +

### 2024-01-25 R/utils-project.R

@@ -1,41 +1,12 @@
-qrender <- function(){
-  rmarkdown::render("paper/paper.Rmd")
-  system("quarto render")
+anonymize_paper <- function(file){
+  # remove title page
+  qpdf::pdf_subset(file, pages = -1, output = sprintf("%s_anonymized.pdf", xfun::sans_ext(file)))
+  
+  # create title page
+  qpdf::pdf_subset(file, pages = 1, output = sprintf("%s_title_page.pdf", xfun::sans_ext(file)))
 }
 
-mkdirif <- function(dir){
-  if(!fs::is_dir(dir)){
-    fs::dir_create(dir)
-  }
-}
-
-render <- function(file){
-  db <- getdb()
-  parent <- dirname(file)
-  current_md5 <- tools::md5sum(file)
-  update <- !(db$md5[db$files == names(current_md5)] == current_md5)
-  if(update){
-    rmarkdown::render(file)
-    db$md5[db$files == names(current_md5)] <- current_md5
-    saveRDS(db, ".dbcache/dbcache.rds")
-  }else{
-    cat("file not changed!")
-  }
-}
-
-getdb <- function(){
-  mkdirif(file.path(".dbcache"))
-  dbpath <- here::here(".dbcache", "dbcache.rds")
-  if(!fs::file_exists(dbpath)){
-    files <- list.files(pattern = "Rmd", full.names = TRUE, recursive = TRUE)
-    md5 <- tools::md5sum(files)
-    db <- data.frame(
-      file = gsub("\\./", "", names(md5)),
-      md5 = md5
-    )
-    names(db) <- c("files", "md5")
-    rownames(db) <- NULL
-    saveRDS(db, dbpath)
-  }
-  db <- readRDS(dbpath)
+render_paper <- function(){
+  rmarkdown::render(here::here("paper", "paper.Rmd"), output_format = "all")
+  anonymize_paper("paper/paper.pdf")
 }

### 2024-01-25 scripts/figures.R

@@ -1,100 +1,0 @@
-# Packages
-
-library(tidyverse)
-library(cowplot)
-library(distributional)
-library(ggdist)
-library(here)
-devtools::load_all()
-
-# Figure Ordinal Models from @Burkner2019-aw ------------------------------
-
-# k <- 4
-# dd <- data.frame(m = 0, s = 1)
-# dd$dist <- distributional::dist_normal(dd$m, dd$s)
-# alpha <- prob_to_alpha(c(0.1, 0.4, 0.4, 0.1), link = "probit")
-# 
-# 
-# cumulative <- ggplot(dd, aes(dist = dist, fill = after_stat(factor(findInterval(x, alpha) + 1)))) +
-#   stat_slab(orientation = "horizontal") +
-#   xlim(c(-4, 4)) +
-#   xlab(latex2exp::TeX("$Y^{*}$")) +
-#   theme(legend.position = "none") +
-#   geom_vline(xintercept = alpha) +
-#   annotate("label", x = alpha, y = 0, label = latex2exp::TeX(sprintf("$\\alpha_{%s}$", 1:(k - 1)))) +
-#   annotate("label", x = seq(-2, 2, length.out = 4), y = 1, label = sprintf("Y = %s", 1:k))
-# 
-# seq1 <- ggplot(dd, aes(dist = dist, fill = after_stat(factor(findInterval(x, alpha[1]) + 1)))) +
-#   stat_slab(orientation = "horizontal") +
-#   xlim(c(-4, 4)) +
-#   xlab(latex2exp::TeX("$Y^{*}_1$")) +
-#   cowplot::theme_nothing() +
-#   theme(axis.title.x = element_text(),
-#         legend.position = "none") +
-#   geom_vline(xintercept = alpha[1]) +
-#   annotate("label", x = alpha[1], y = 0, label = latex2exp::TeX(sprintf("$\\alpha_{%s}$", 1))) +
-#   annotate("label", x = c(-2, -0.5), y = 1, label = c("Y = 1", "Y > 1"))
-# 
-# seq2 <- ggplot(dd, aes(dist = dist, fill = after_stat(factor(findInterval(x, alpha[2]) + 1)))) +
-#   stat_slab(orientation = "horizontal") +
-#   xlim(c(-4, 4)) +
-#   xlab(latex2exp::TeX("$Y^{*}_2$")) +
-#   cowplot::theme_nothing() +
-#   theme(axis.title.x = element_text(),
[…]

### 2024-01-25 scripts/tables.R

@@ -1,78 +1,0 @@
-library(flextable)
-library(tidyverse)
-devtools::load_all()
-
-# MODEL TABLE
-
-models_table <- data.frame(
-  stringsAsFactors = FALSE,
-  Model = c("Model", "Cumulative Logit", "Cumulative Probit"),
-  Link.Function = c("Equation",
-                    "$\\text{logit}(p) = \\text{log}(p / (1-p))$",
-                    "$z = \\Phi^{-1}(p)$"),
-  Link.Function = c("R Code", "`qlogis()`", "`qnorm()`"),
-  Inverse.Link.Function = c("Equation",
-                            "$e^{\\text{logit}(p)} / (1 + e^{\\text{logit}(p)})$","$\\Phi^(z)$"),
-  Inverse.Link.Function = c("R Code", "`plogis()`", "`pnorm()`")
-)
-
-models_table[-1, ] |> 
-  flextable() |> 
-  autofit() |> 
-  set_header_labels(values = list(Model = "Model",
-                                  Link.Function = "Equation",
-                                  Link.Function.1 = "R Code",
-                                  Inverse.Link.Function = "Equation",
-                                  Inverse.Link.Function.1 = "R Code")) |>
-  add_header_row(values = c("", "Link Function", "Link Function", "Inverse Link Function", "Inverse Link Function")) |>
-  merge_h(1, part = "header") |>
-  theme_booktabs(bold_header = TRUE) |>
-  align(part = "all", align = "center") |> 
-  ftExtra::colformat_md(part = "all") |> 
-  save_as_docx(path = sprintf(here::here("paper", "tables", "%s.docx"), "notation-table"))
-
-
-# DATASET EXAMPLE TABLE
-
-b1 <- log(3) # log odds ratio
-k <- 4
-n <- 100
-x <- rep(c("a", "b"), each = n/2)
-dat <- data.frame(x = x)
-probs <- rep(1/k, k) # for the group "a", uniform probabilities
-dat <- sim_ord_latent(~x, beta = b1, prob0 = probs, data = dat, link = "logit")
-
-funs <- list(mean = mean, median = median, sd = sd)
-ss <- sapply(funs, function(f) tapply(as.numeric(dat$y), dat$x, f))
[…]

### 2024-01-26 README.Rmd
major update

@@
 )
+
+devtools::load_all()
+pkgs <- .get_packages()
 ```
@@
 <!-- badges: start -->
+[<img alt="alt_text" src="https://img.shields.io/badge/OSF-https://osf.io/93h5j/-337AB7"/>](https://osf.io/93h5j/)
 <!-- badges: end -->
+
+This repository contains the code and functions to reproduce the analysis and simulations presented in the paper *Ordinal regression models made easy.
+A tutorial on parameter interpretation, data simulation, and power analysis.* by Filippo Gambarota and Gianmarco Altoè.
+This repository is organized as an R package:
+
+- `R/`: folder with all custom functions used in the paper and the supplementary materials
+  - `utils.R`: main functions for simulations
+  - `utils-plotting.R`: functions for plots
+  - `utils-project.R`: functions for the project and folders organization, mainly for internal usage
+  - `utils-paper.R`: functions for the paper formatting, mainly for internal usage
+- `paper/`: folder with the source of the paper (`.Rmd`) and the compiled pdf document
+- `supplementary/`: folder with some extra documents (`.qmd`) with examples
+- `tests/`: folder with unit tests for the main functions (`R/utils.R`)
+
+## Using the functions
+
+Once cloned or downloaded the repository, the functions can be used running `devtools::load_all()` (the `devtools` package is required).
+This is very similar to `library(package)` thus all functions are available but not as objects in the global environment (compared to using `source(file)`).
+
+## Packages
+
+These packages are used through the project:
+
+```{r, echo = FALSE, message=FALSE, warning=FALSE, results='asis'}
+sprintf("- `%s` (%s)", pkgs$pkg, pkgs$version) |> 
+  cat(sep = "\n")
+```
+
+## Session info
+
+```{r, echo=FALSE}
+sinfo <- sessioninfo::session_info()$platform
+sinfo <- sinfo[names(sinfo) != "date"]
+paste0(
+  cli::style_bold(cli::col_blue(names(sinfo))), ": ", sinfo
+) |> 
+  cat(sep = "\n")
+```

### 2024-01-26 index.qmd
major update

@@
 ---
-title: "Test"
+title: "Simulating ordinal data"
 ---
 
-This is a Quarto website.
+This is the online supplementary material for the tutorial-paper *Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis*.
 
-To learn more about Quarto websites visit <https://quarto.org/docs/websites>.
+In the next links there are useful materials and examples that extended what presented in the tutorial-paper.
+
+- [Useful resources](supplementary/useful-resources.qmd): a collection of books, blog posts, R packages and more about ordinal data
+- [Ordinal notes](supplementary/ordinal-notes.qmd): a miscellanea of code snippets and quick examples about ordinal regression models
+- [Location-scale models](supplementary/location-scale-models.qmd): a tutorial to fit, interpret and simulate location-scale cumulative link models.
+- [Simulating a GLM](supplementary/simulating-glm.qmd): a quick example on how to simulate data for a Generalized Linear Model (GLM).

### 2024-01-26 paper/paper.Rmd
major update

@@
 abstract: |
-  Ordinal data are extremely used in psychology.
-Common examples include Likert items or variables where categories are ordered without assuming a common distance between labels.
-Despite the prevalence of ordinal data, a majority of published psychological research relies on applying metric models (e.g., standard linear regression) even when the dependent variable is ordinal.
-This approach leads to several statistical drawbacks.
-Similar to what happens with binary data (e.g., accuracies in a cognitive experiment), using metric models instead of e.g.
-logistic regression is probably caused by several reasons.
-Firstly, parameter interpretation can be challenging.
-Ordinal models commonly used a link function (e.g., logit) for model fitting thus transforming the parameters scale.
-This is even more critical when estimating models with multiple parameters or interactions.
-Secondly, nowadays journals require power or sensitivity analysis when planning or evaluating a research work.
-For common statistical models such as t-tests or the analysis of variance, there are tools to easily estimate the statistical power.
-For ordinal regressions, estimating the statistical power often requires simulating data thus clearly understanding the statistical background and implementing a Monte Carlo simulation.
-The objective of this tutorial is to present the essential statistical background of ordinal regression, highlighting the research areas that would benefit more by introducing this model.
-In particular, we present how to interpret the model parameters with categorical and numerical variables also linking them to common standardized effect size measures.
-Then we propose methods to report the results, especially focusing on graphical representation.
-Finally, we present how to estimate the statistical power using Monte Carlo simulations with suggestions on setting parameters in a meaningful way.
-Through the tutorial, we use the R programming language using specific packages for ordinal models and sharing also ad-hoc developed functions supporting each step.
-For the sake of reproducibility and to empower researchers in analyzing their studies autonomously, all data and scripts used in the article have been made publicly available.
+  Ordinal data are very common in psychology such as Likert items, ratings or generic ordinal variables.
+Despite the widespread these variables are usually analyzed using metric models (e.g., standard linear regression).
+This approach can have important drawbacks in terms of statistical inference (e.g., reduced power) and prediction.
+One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis.
+The aim of the tutorial is presenting ordinal regression models using a simulation-based approach.
+In the first section we introduced the general model highlighting the crucial components and assumptions.
+Then we explained how to intepret the parameters for a logit and probit model.
+Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor.
+Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors.
+The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models.
+The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Github repository.
 
@@
   preprint: true
-classoption       : "man, mask"
+classoption       : "man"
 output: 
@@
 library(purrr)
+library(kableExtra)
 devtools::load_all()
-```
-
-```{r packages-latex, include = FALSE, eval = which_output("latex")}
-library(kableExtra)
 ```
@@
 
-theme_set(mtheme())
-
 # palettes
@@
 scale_fill_discrete <- scale_fill_viridis_d
-options(tikzDefaultEngine = "xetex")
-```
-
-```{r scripts, include = FALSE, eval = TRUE}
-temp <- new.env()
-# figures
-# source(here("scripts", "figures.R"), local = temp)
-rfigs <- readRDS(here("objects", "r-figs.rds"))
-
-# tables
-# source(here("scripts", "tables.R"), local = temp)
-rtabs <- readRDS(here("objects", "r-tabs.rds"))
+theme_set(mtheme())
 ```
@@
 Beyond Likert-like items, in Psychology there are several applications of ordinal scales.
-For instance, sociodemographic variables like educational levels or socioeconomic status, as well as general ratings such as pain severity, agreement with a statement, or the evaluation of sensory experiences.
+For instance, sociodemographic variables like educational levels or socioeconomic status, as well as general ratings such as pain severity, agreement with a statement, or the evaluation of sensory experiences [e.g., @Overgaard2021-hs].
 
@@
 Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels [@Liddell2018-wu].
-More generally, @Cliff2016-ck suggest that most of the research questions in behavioral sciences can be considered as ordinal (*is the score* $x$ higher than the score $y$?) concerning variables where the most appropriate measurement scale is probably ordinal.
-
-In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable.
-@Liddell2018-wu reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models.
-In the same work, @Liddell2018-wu showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric [but see @Robitzsch2020-la for an alternative perspective].
-They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and distorted effect size estimates.
-Some authors suggested that individual ordinal items and a collection of ordinal (averaged or summed) items can safely be considered and analyzed as metric [@Carifio2008-qd; @Carifio2007-kk; but see @Jamieson2004-xf].
-Despite @Liddell2018-wu provide some example that averaging ordinal items and applying metric models is not appropriate, in the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.).
-
-For the tutorial, we assume that the reader is already familiar with basic R programming and introductory theory about linear regression.
-Despite the practical and applicative spirit of the tutorial, we still introduced some statistical theory and equations supported by translation into R code.
-Similar to other tutorials [@DeBruine2021-id; @Gambarota2023-on], we believe this is the best educational approach especially with the involvement of Monte Carlo simulations.
+More generally, @Cliff2016-ck suggests that several research questions in behavioral sciences can be considered as ordinal (*is the score* $x$ higher than the score $y$?) concerning variables where the most appropriate measurement scale is probably ordinal.
+
+In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a linear regression that makes an explicit assumption about metric features of the response variable.
+@Liddell2018-wu reviewed the psychological literature using Likert-based measures and reported how the majority of papers used metric-based statistical models.
+In the same work, @Liddell2018-wu showed examples about the potential pitfalls of metric models applied to ordinal variables [but see @Robitzsch2020-la for an alternative perspective].
+They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and biased estimates.
+While sums or averages of ordinal items could be treated as metric [@Carifio2008-qd; @Carifio2007-kk; but see @Jamieson2004-xf] @Liddell2018-wu provided some examples of potential pitfalls even in this case.
+In the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.) without considering the case of aggregating ordinal responses.
+
+For the tutorial, we assume that the reader is familiar with basic R programming and introductory theory about linear regression.
+Similar to other tutorials [@DeBruine2021-id; @Gambarota2023-on] we used few equations that are necessary to understand the model parameters and implement the Monte Carlo simulation.
 
@@
 library(ggplot2)
+
+# custom functions
+
+# show first and last rows of a dataframe
+head_tail <- function(x, n = 5){
+  rbind(
+    head(x, n = n),
+    tail(x, n = n)
+  )
+}
 ```
@@
 
-While Liddell and Kruschke [-@Liddell2018-wu] proposed a specific modeling approach, there exists a class of regression models that consider the ordinal nature of the response variable without relying on metric assumptions.
-We can name this general class of models as *ordinal regression*.
-The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures [@Tutz2022-dg].
+Despite the specific model proposed by @Liddell2018-wu proposed a specific modeling approach, there is a class of regression models that consider the ordinal nature of the response variable.
+The actual statistical nomenclature of *ordinal regression* models can be confusing [@Tutz2022-dg].
 @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
 
-We can identify three main classes of models: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe], and *adjacent category models*.
-Among these, the cumulative model is the most widely used, assuming the existence of a latent variable.
-This latent variable, categorized using a set of thresholds, produces the observed ordinal variable.
-The *sequential model* as suggested by the name is appropriate when modeling sequential processes.
+We can identify three main types: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe], and *adjacent category models*.
+Among these, the cumulative model is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
+The *sequential model* is appropriate when modeling sequential processes.
 Assuming to have five response options, the sequential model assume that responding "3" assume a sequential process where steps "1" and "2" are already reached.
@@
 In the current paper we put the focus on the CM for several reasons.
-The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation.
+Firstly, the latent formulation of the model is particularly convenient for parameter interpretation and data simulation.
 The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item.
-Furthermore, CM are also used to model data under a signal detection theory framework [e.g., @DeCarlo2010-lj].
-Figure \@ref(fig:fig-explain-cumulative) depicts the overall structure of the *cumulative* model.
+Furthermore, the signal detection theory framework that is very common in experimental psychology can be implemented using a CM [e.g., @DeCarlo2010-lj].
+Figure \@ref(fig:fig-explain-cumulative) depicts the overall structure of the CM.
 
@@
 
-In this section we introduce some notation for the CM that is used through the paper.
-We proposed a notation as consistent as possible with the literature and the `ordinal` package [@Christensen2022-jk].
+In this section we introduce some notation for the CM to be consistent with the existing literature and the `ordinal` R package [@Christensen2022-jk].
 We define $Y_k$ as the observed ordinal variable with $k$ levels and $Y^\star$ is the underlying latent variable.
@@
 The specific link function define the type of model and require a different R function.
-For example, when assuming a Gaussian distribution, fitting a model with a *probit* link function entails using the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
+For example, when assuming a Gaussian distribution (*probit* link function) entails using the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
+
 When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$^[As done by @Agresti2010-rz, when referring to $P(Y \leq k)$ we are implicitly conditioning on a particular $\mathbf{X}$ value $P(Y \leq k | \mathbf{X}_i)$].
@@
 
-<!-- notazione https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf e @Burkner2019-aw per la parte di notazione e quella più formale anche @DeCarlo2010-lj -->
-
-The same model can be written in the so-called latent formulation as reported in Equation \@ref(eq:latent-model).
-The model is no longer directly about the cumulative probabilities but focused on the continuous latent variable $Y^\star$ as a function of the linear predictor $\eta = \mathbf{X} \boldsymbol{\beta}$ similar to a standard linear regression.
+<!-- see https://people.vcu.edu/~dbandyop/BIOS625/CLM_R.pdf for consistent notation -->
+
+Equation \@ref(eq:latent-model) shows the latent formulation of the previous model.
+The latent variable $Y^\star$ is a function of the linear predictor $\eta = \mathbf{X} \boldsymbol{\beta}$ similar to a standard regression.
 
@@
 
-The crucial part is $\epsilon$ that is the random component of the model coming from a certain probability distribution.
-For a *probit* model, errors are sampled from a standard Gaussian distribution while for a *logit* model from a standard logistic distribution.
+The crucial part is $\epsilon$ that is the random part (errors) of the model.
+For a *probit* model, errors are sampled from a standard Normal distribution while for a *logit* model from a standard logistic distribution.
 Following the notation by @Tutz2022-dg, the observed ordinal value $Y_i = k$ comes from $Y^\star_i$ belonging to the interval defined by the thresholds $Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}$ where $- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty$.
 
-In the basic version of the model, the thresholds $\alpha_k$ are considered as fixed and being part of the measurement procedure [@Liddell2018-wu] and do not vary as a function of the predictors.
-In a more sophisticated version of the model called location-shift [@Tutz2022-dg], both the location $\mu$ and the thresholds $\alpha_k$ can vary as a function of the predictors.
-
+The thresholds $\alpha_k$ are considered fixed and part of the measurement procedure [@Liddell2018-wu; see the *location-shift* models by @Tutz2022-dg, where thresholds vary as a function of predictors].
 The model can be also formalized in alternative ways.
-@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian version of the model with a different threshold parametrization.
-@Gelman2020-tg proposed three alternative parametrizations focusing on different definition of the thresholds.
-
-@Tutz2022-dg described also another version of the model called location-scale [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur] where the location $\mu$ and the scale $\sigma^2$ of the distribution can vary as a function of the predictors.
+@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian version of the model or @Gelman2020-tg for other thresholds parametrizations.
+
+The scale (spread) of the latent variable is fixed to one for both *probit* and *logit* models.
+In the supplementary materials we showed how to implement a location-scale models including predictors on the scale parameter [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur].
 
@@
 
-The cumulative link model implemented in Equations \@ref(eq:prob-cum-model1) and \@ref(eq:latent-model) can be considered the general formulation that requires specifying the link function $g(\cdot)$ or the errors distribution $\epsilon_i \sim D(\mu, \sigma^2)$.
+The CM implemented in Equations \@ref(eq:prob-cum-model1) and \@ref(eq:latent-model) requires specifying the link function $g(\cdot)$ or the errors distribution $\epsilon_i \sim D(\mu, \sigma^2)$ ($D$ being a certain probability distribution).
 Among several available functions the *logit* and *probit* models are the most common.
@@
 In the next sections we will illustrate the differences and simulation strategies.
-Figure \@ref(fig:fig-logit-vs-probit) depicts the two distributions while Table L'output è `r if(which_output("latex")) "\\@ref(tab:tab-model-summary-latex)" else "\\@ref(tab:tab-model-summary-word)"` summarise the presented cumulative models with the proposed link function and the corresponding R code.
-
-In terms of parameters, both distributions can be defined with a location $\mu$ and a scale $s$ parameter.
+Figure \@ref(fig:fig-logit-vs-probit) depicts the two distributions while Table `r if(which_output("latex")) "\\@ref(tab:tab-model-summary-latex)" else "\\@ref(tab:tab-model-summary-word)"` summarise the presented CM with the proposed link function and the corresponding R code.
+
+In terms of parameters, both distributions can be defined with in terms of location $\mu$ and scale $s$.
 The standard normal distribution has $\mu = 0$ and $s = 1$.
@@
 ```{r tab-model-summary-latex, eval = which_output("latex")}
+
+replace_before_after <- function(x, before, after = NULL, new){
+  if(is.null(after)) after <- before
+  whole <- sprintf("%s.*%s", before, after)
+  within <- sprintf("(?<=%s).*(?=%s)", before, after)
+  within_string <- regmatches(x, regexpr(within, x, perl = TRUE))
+  new_string <- sprintf(new, within_string)
+  x[grepl(whole, x)] <- stringr::str_replace_all(x[grepl(whole, x)], whole, new_string)
+  return(x)
+}
+
 models_table_latex <- models_table
@@
                                                            new = "\\\\texttt{%s}")
-models_table_latex[-1, ] |> 
+models_table_latex |>
   kable(col.names = c("Model", "Equation", "R Code", "Equation", "R Code"), row.names = FALSE,
@@
 
-The function to fit the model is `clm()` and the model equation is specified using the R formula syntax `y ~ x` where `y` is the ordinal dependent variable and `x` is one or more predictors eventually including also the interactions.
-The package also implements mixed-effects models (see the `clmm()` function) including random intercepts and slopes but the topic is beyond the scope of the current tutorial.
+The main function is `clm()` and the formula is specified using the syntax `y ~ ...` where `y` is the ordinal response and `...` is a combination of predictors.
+The package also implements mixed-effects models including random intercepts and slopes.
 
@@
 ss <- round(ss, 2)
-ss$Y1 <- sprintf("%s (p = %s)", fmt(ss$Y1, "bold", "latex"), fmt(pp$Y1, "bold", "latex"))
-ss$Y2 <- sprintf("%s (p = %s)", fmt(ss$Y2, "bold", "latex"), fmt(pp$Y2, "bold", "latex"))
-ss$Y3 <- sprintf("%s (p = %s)", fmt(ss$Y3, "bold", "latex"), fmt(pp$Y3, "bold", "latex"))
-ss$Y4 <- sprintf("%s (p = %s)", fmt(ss$Y4, "bold", "latex"), fmt(pp$Y4, "bold", "latex"))
+ss$Y1 <- sprintf("%s (p = %s)", fmt(ss$Y1, "bold", "latex"), fmt(pp$Y1, "italic", "latex"))
+ss$Y2 <- sprintf("%s (p = %s)", fmt(ss$Y2, "bold", "latex"), fmt(pp$Y2, "italic", "latex"))
+ss$Y3 <- sprintf("%s (p = %s)", fmt(ss$Y3, "bold", "latex"), fmt(pp$Y3, "italic", "latex"))
+ss$Y4 <- sprintf("%s (p = %s)", fmt(ss$Y4, "bold", "latex"), fmt(pp$Y4, "italic", "latex"))
 
@@
 ss <- select(ss, group, everything(), -tot)
-dat <- rtabs$data_example_table$dat
-
-ss <- map(rtabs$data_example_table$table, ~gsub("\\**", "", .x)) |> 
+
+ss <- map(ss, ~gsub("\\**", "", .x)) |> 
   bind_rows()
@@
 ss |> 
-  kable(row.names = FALSE,
-        booktabs = TRUE, 
-        format = "latex", 
-        caption = filor::get_caption(), 
-        escape = FALSE, 
-        align = "c") |> 
+  kable(booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE, align = "c", row.names = FALSE) |> 
   kable_styling(full_width = FALSE) |> 
-  row_spec(0,bold=TRUE)
+  row_spec(0, bold = TRUE) 
 ```
@@
 To understand the logit model we need to introduce odds and odds ratio.
-The odds of a probability $p$ is defined as $p/(1 - p)$ thus the probability of success divided by the probability of failure.
-The odds takes value ranging from 0 to $\infty$.
-For instance, with a probability of $p = 0.8$, we have odds of $4$, indicating that there are four successes for each failure.
-The same as having $p = 0.2$ and an odds of $0.25$ means that for each $0.25$ successes we have a failure or that we have $4$ failures for each success.
-When comparing two groups or conditions we can compare the two odds calculating an odds ratio.
-The odds ratio is the mostly used statistics to compare groups or conditions on a binary outcome.
-An odds ratio of $4$ means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator.
+The odds of a probability $p$ is defined as $p/(1 - p)$ thus the success probability over the failure probability.
+The odds takes value ranging from $0$ to $\infty$.
+With a probability of $p = 0.8$, we have odds of $4$, indicating that there are four successes for each failure.
+The same as having $p = 0.2$ and an odds of $0.25$ means that for each $0.25$ successes we have a failure (or $4$ failures for each success).
+When comparing two groups or conditions we can take the ratio of two odds.
+An odds ratio (OR) of $4$ means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator.
 The Figure \@ref(fig:fig-odds-example) shows the relationship between probabilities and odds.
@@
 With e.g.
-$k = 4$ outcomes we have $k - 1$ models determined by the cumulative probability in terms of $P(Y \leq 1), \dots, P(Y \leq k - 1)$.
-We can manually calculate the odds ratio starting from the previous dataset.
+$k = 4$ outcomes we have $k - 1$ odds ratios determined by the cumulative probability in terms of $P(Y \leq 1), \dots, P(Y \leq k - 1)$ comparing the two groups.
 
@@
 #| collapse: true
+#| include: false
 
@@
 
-For example the probability of responding $Y = 1$ in the group "a" is `r p1a` corresponding to an odds of `r odds(p1a)`.
-When calculating the odds ratio comparing "a" vs "b" for $Y = 1$ we obtain that the group "a" has `r odds(p1a) / odds(p1b)` times the odds of responding $Y = 1$ compared to the group "b".
-
-For an ordinal variable we have $k$ levels and we can calculate $k - 1$ cumulative probabilities (excluding the last associated with $P = 1$).
-Thus we have $k - 1$ odds ratio or what is called a cumulative odds ratio.
-Basically we repeat the previous steps but on the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$.
-The three odds ratios are very similar correspond to the model estimate $e^\beta_1$, in R `exp(fit$beta)` `r exp(fit$beta)`.
-In fact, the logit model essentially estimate the cumulative odds ratio.
+For example the probability of responding $Y \leq 1$ in the group "a" is `r p1a` corresponding to an odds of `r odds(p1a)`.
+When calculating the odds ratio comparing "a" vs "b" for $Y \leq 1$ we obtain that the group "a" has `r odds(p1a) / odds(p1b)` times the odds of responding $Y \leq 1$ compared to the group "b".
 
@@
 #| collapse: true
+
+# function to calculate odds
+odds <- function(p) p / (1 - p)
+
 # the dummy_ord() is a custom function creating k - 1 dummy variable for the cumulative probabilities
@@
 
-From the previous example, we noticed that the $k - 1$ estimated odds ratios are different.
-However, increasing $n$ will result in odds ratios that are very similar.
-This does not happen by chance but from an implicit (until now) assumption made by the CM called *proportional odds* (PO).
 Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the PO assumption.
@@
 
-```{r}
-#| echo: true
-#| collapse: true
-set.seed(2024)
-# using same parameters as the previous simulation increasing n
-n <- 1e5 # very large n
-x <- rep(c("a", "b"), each = n)
-dat <- data.frame(x = x)
-dat <- sim_ord_latent(~x, beta = b1, prob0 = probs, data = dat, link = "logit")
-
-# same steps as before
-cum_p <- cbind(x = dat$x, dummy_ord(dat$y))
-group_a <- apply(cum_p[cum_p$x == "a", -1], 2, mean)
-group_b <- apply(cum_p[cum_p$x == "b", -1], 2, mean)
-# now odds are almost the same and equal to the simulated b1 = ~1.09
-odds(group_a) / odds(group_b)
-log(odds(group_a) / odds(group_b))
-```
-
-Basically if we use the *logit* link function $g(\cdot)$, regression coefficients are interpreted as odds ratios.
-Given that we have $k > 2$ alternatives we need $k - 1$ equations.
+When fitting a logit model, the coefficients $\beta_j$ are the (log) odds ratios for a unit increase in $x$.
+However, the previous model estimate a single $\beta$ even if we have four ordinal outcomes.
+The CM model (in the basic version) made an assumption called *proportional odds* (PO).
+Under this assumption we assume that the $k - 1$ cumulative odds ratios are the same.
 The PO assumption is formalized in Equation \@ref(eq:prop-odds).
@@
 
-The PO assumption is convenient because regardless of $k$, the $\beta_j$ ($j$ being the number of regression coefficients) effect is assumed to be the same.
-The model is more parsimonious compared to estimating $k - 1$ coefficients for each $\beta_j$ as in the multinomial regression or the non-proportional odds model.
-
-At the same time, the PO assumption could be considered too strict.
+There are CM models relaxing this assumption completely [*non proportional odds*, @Tutz2022-dg] or partially [*partial proportional odds*, @Peterson1990-cz].
+The PO model is more parsimonious but the PO assumption could be considered too strict.
 There are several methods for testing if data are supporting the PO assumption [see @Liu2023-bp for an overview].
@@
 However, these methods and models are outside the scope of the tutorial.
-In the supplementary materials, we provide more details only about the *location-scale* model in terms of parameters interpretation and data simulation.
-
-This can be shown by fitting the model with `clm()` that by default assume the PO and using the predicted probabilities to compute the odds ratios.
-In the following code we computed the odds ratio comparing $x_a$ and $x_b$ when $k \leq 1$ and $k \leq 2$.
-The odds ratios are exactly the same because we are using the predicted probabilities from the model (assuming PO).
+
+In the previous code block, we estimated the $k - 1$ odds ratios and they were different even if data have been simulated under the PO assumption (see the Simulating data section).
+Calculating the odds ratios using predicted probabilities from the model fitted with `clm()` (that by default assume the PO) the odds ratios are the same.
+In the following code block we computed the odds ratio comparing $x_a$ and $x_b$ when $k \leq 1$ and $k \leq 2$.
 
@@
 The PO assumption is relevant only for the *logit* model.
-When using a different link function (e.g., *probit*) the assumption of a common $\beta$ is called *parallel slopes* (see Equation \@ref(eq:probit-prop-odds)).
-Basically the difference in $z$ scores for a unit increase in $x$ (i.e., the slope) is the same regardless the threshold.
+When using the *probit* link function we can use the terms *parallel slopes* for the assumption of a common $\beta$ (see Equation \@ref(eq:probit-prop-odds)).
+In practical terms, difference in $z$ scores for a unit increase in $x$ (the actual interpretation of $\beta$ in a *probit* model) is the same regardless of the threshold.
 As for the previous example, we can fit the *probit* model and manually calculate the difference in $z$ scores.
@@
 Then the model formula is specified within the function as `location = ~` along with the vector of regression coefficients (`beta`), baseline probabilities (`prob0`) and the link function.
-The function return a dataset with the simulated $Y$ and the latent variable (this is possible only because we simulate the data, otherwise $Y^{\star}$ is not observed by definition).
+The function return a dataset with the simulated $Y$ and the latent variable (this is possible only because we simulate the data, otherwise $Y^{\star}$ cannot be observed by definition).
 
@@
 The crucial part is setting appropriate and empirically meaningful parameters.
-Thresholds $\alpha$ are usually not of main interest [but see the location-shift models @Tutz2022-dg] and can be considered as intercepts in standard linear regression.
-The thresholds are quantiles of the latent distribution that produced certain $k$ probabilities.
+In the PO model thresholds are usually not of main interest as intercepts in standard linear regression.
+Thresholds are essentially quantiles of the latent distribution that produced certain $k$ probabilities.
 To set meaningful $\alpha$ values we can convert probabilities into thresholds (`alpha_to_prob()`).
 The function `show_alpha()` produce a meaningful visual representation of using a specific set of thresholds (see Figure \@ref(fig:fig-show-th-example)).
-In regression terms, the thresholds determine the probabilities when $x = 0$ (as the intercept).
-Thus with two groups for example, the thresholds are the $k$ probabilities of the ordinal variable $Y$ for the reference group.
+Thus with two groups for example, the thresholds are the $k$ probabilities of the ordinal variable $Y$ for the reference group (when $x = 0$).
 
 ```{r fig-show-th-example, fig.cap=filor::get_caption(), fig.place = TRUE}
-p1 <- show_alpha(prob = rep(1/4, 4), link = "probit")
-p2 <- show_alpha(prob = c(0.1, 0.2, 0.5, 0.2), link = "probit")
+p1 <- show_alpha(prob = rep(1/4, 4), link = "probit", plot = FALSE)
+p2 <- show_alpha(prob = c(0.1, 0.2, 0.5, 0.2), link = "probit", plot = FALSE)
 
@@
 The `cat_latent_plot()` and `num_latent_plot()` functions can be for respectively a categorical (Figures \@ref(fig:fig-example-cat-latent)) and numerical predictor (Figures \@ref(fig:fig-example-num-latent)).
+In the supplementary materials we show how to use the `sim_ord_latent()` function for checking the impact of certain $\beta$ for more complex models.
 
 ```{r fig-example-cat-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
-cat_latent_plot(m = c(0, 1), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")
-```
-
-```{r fig-example-num-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
-num_latent_plot(x = rnorm(100), b1 = 1, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit") +
-  mtheme() +
-  viridis::scale_color_viridis(discrete = TRUE)
+cp <- cat_latent_plot(location = c(0, 1), scale = 1, prob0 = rep(1/4, 4), link = "logit", plot = FALSE) 
+legend <- cowplot::get_legend(cp$probs_plot)
+plt <- cowplot::plot_grid(
+  cp$lat_plot + rm_legend(),
+  cp$probs_plot + rm_legend()
+)
+cowplot::plot_grid(plt, legend, nrow = 2, rel_heights = c(0.95, 0.05))
+```
+
+```{r fig-example-num-latent, fig.cap=filor::get_caption(), fig.place = TRUE, fig.height=12}
+np <- num_latent_plot(x = rnorm(100), b1 = 1, prob0 = c(0.6, 0.2, 0.1, 0.1), link = "probit", plot = FALSE)
+
+legend <- cowplot::get_legend(np$qplot)
+#title <- get_title(np$pplot)
+
+np$pplot <- np$pplot + scale_color_discrete()
+
+plt <- cowplot::plot_grid(
+  np$pplot + rm_legend() + theme(plot.title = element_blank()),
+  np$qplot + rm_legend(),
+  nrow = 2
+)
+
+cowplot::plot_grid(plt, legend, nrow = 2, rel_heights = c(0.95, 0.05))
 ```
@@
 We present very general example that, as for the previous simulations, can be easily extended.
-For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting $H_0$) or assessing an eventual bias in model estimates.
 
@@
 On the left the shift in the latent mean and on the right the impact on the expected probabilities.
-The plot can be created using `cat_latent_plot(m = c(0, 0.5), s = 1, probs = rep(1/4, 4), link = "logit", plot = "both")`.
+The plot can be created using `cat_latent_plot(m = c(0, 1), s = 1, prob0 = rep(1/4, 4), link = "logit")`.
 
 (ref:fig-example-num-latent) Example of the `num_latent_plot()` function depicting the effect of a continuous predictor on the expected probabilities.
-$\beta_1$ is the regression coefficient.
-The plot can be created using `num_latent_plot(x = runif(100), b1 = 1, probs = c(0.6, 0.2, 0.1, 0.1), link = "probit")`
+Dotted lines are the 25%, 50% and 75% quantiles.
+$\beta_1 = 1$ is the regression coefficient.
+The plot can be created using `num_latent_plot(x = runif(100), b1 = 1, prob0 = c(0.6, 0.2, 0.1, 0.1), link = "probit")`
 
@@
 
-(ref:tab-model-summary-latex) (ref:tab-model-summary-docx)
+(ref:tab-model-summary-latex) Summary of the probit and logit models in terms of link ($g(\cdot)$) and inverse link function ($g^{-1}(\cdot)$) and the corresponding R code.
 
@@
 
-(ref:tab-dataset-example-latex) (ref:tab-dataset-example-docx)
+(ref:tab-dataset-example-latex) Summary of the simulated dataset.
+For each group (a and b) we reported mean, median and standard deviation of the ordinal outcome (metric descriptive statistics).
+For each ordinal outcome ($Y$) we reported the frequency and (within parentheses) the probability.

### 2024-01-26 paper/paper.tex
major update

@@
 \documentclass[
-  man, mask,floatsintext]{apa6}
+  man,floatsintext]{apa6}
 \usepackage{amsmath,amssymb}
@@
 \abstract{%
-Ordinal data are extremely used in psychology.
-Common examples include Likert items or variables where categories are ordered without assuming a common distance between labels.
-Despite the prevalence of ordinal data, a majority of published psychological research relies on applying metric models (e.g., standard linear regression) even when the dependent variable is ordinal.
-This approach leads to several statistical drawbacks.
-Similar to what happens with binary data (e.g., accuracies in a cognitive experiment), using metric models instead of e.g.~logistic regression is probably caused by several reasons.
-Firstly, parameter interpretation can be challenging.
-Ordinal models commonly used a link function (e.g., logit) for model fitting thus transforming the parameters scale.
-This is even more critical when estimating models with multiple parameters or interactions.
-Secondly, nowadays journals require power or sensitivity analysis when planning or evaluating a research work.
-For common statistical models such as t-tests or the analysis of variance, there are tools to easily estimate the statistical power.
-For ordinal regressions, estimating the statistical power often requires simulating data thus clearly understanding the statistical background and implementing a Monte Carlo simulation.
-The objective of this tutorial is to present the essential statistical background of ordinal regression, highlighting the research areas that would benefit more by introducing this model.
-In particular, we present how to interpret the model parameters with categorical and numerical variables also linking them to common standardized effect size measures.
-Then we propose methods to report the results, especially focusing on graphical representation.
-Finally, we present how to estimate the statistical power using Monte Carlo simulations with suggestions on setting parameters in a meaningful way.
-Through the tutorial, we use the R programming language using specific packages for ordinal models and sharing also ad-hoc developed functions supporting each step.
-For the sake of reproducibility and to empower researchers in analyzing their studies autonomously, all data and scripts used in the article have been made publicly available.
+Ordinal data are very common in psychology such as Likert items, ratings or generic ordinal variables.
+Despite the widespread these variables are usually analyzed using metric models (e.g., standard linear regression).
+This approach can have important drawbacks in terms of statistical inference (e.g., reduced power) and prediction.
+One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis.
+The aim of the tutorial is presenting ordinal regression models using a simulation-based approach.
+In the first section we introduced the general model highlighting the crucial components and assumptions.
+Then we explained how to intepret the parameters for a logit and probit model.
+Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor.
+Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors.
+The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models.
+The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Github repository.
 }
@@
 Beyond Likert-like items, in Psychology there are several applications of ordinal scales.
-For instance, sociodemographic variables like educational levels or socioeconomic status, as well as general ratings such as pain severity, agreement with a statement, or the evaluation of sensory experiences.
+For instance, sociodemographic variables like educational levels or socioeconomic status, as well as general ratings such as pain severity, agreement with a statement, or the evaluation of sensory experiences (e.g., Overgaard \& Sandberg, 2021).
 
@@
 Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018).
-More generally, Norman Cliff (2016) suggest that most of the research questions in behavioral sciences can be considered as ordinal (\emph{is the score} \(x\) higher than the score \(y\)?) concerning variables where the most appropriate measurement scale is probably ordinal.
-
-In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a normal linear regression that makes an explicit assumption about metric features of the response variable.
-Liddell and Kruschke (2018) reviewed the psychological literature using likert-based measures and reported how the majority of papers used metric-based statistical models.
-In the same work, Liddell and Kruschke (2018) showed extensive examples and simulations about the potential pitfalls of treating an ordinal variable as metric (but see Robitzsch, 2020 for an alternative perspective).
-They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and distorted effect size estimates.
-Some authors suggested that individual ordinal items and a collection of ordinal (averaged or summed) items can safely be considered and analyzed as metric (Carifio \& Perla, 2008; Carifio \& Perla, 2007; but see Jamieson, 2004).
-Despite Liddell and Kruschke (2018) provide some example that averaging ordinal items and applying metric models is not appropriate, in the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.).
-
-For the tutorial, we assume that the reader is already familiar with basic R programming and introductory theory about linear regression.
-Despite the practical and applicative spirit of the tutorial, we still introduced some statistical theory and equations supported by translation into R code.
-Similar to other tutorials (DeBruine \& Barr, 2021; Gambarota \& Altoè, 2023), we believe this is the best educational approach especially with the involvement of Monte Carlo simulations.
+More generally, Norman Cliff (2016) suggests that several research questions in behavioral sciences can be considered as ordinal (\emph{is the score} \(x\) higher than the score \(y\)?) concerning variables where the most appropriate measurement scale is probably ordinal.
+
+In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a linear regression that makes an explicit assumption about metric features of the response variable.
+Liddell and Kruschke (2018) reviewed the psychological literature using Likert-based measures and reported how the majority of papers used metric-based statistical models.
+In the same work, Liddell and Kruschke (2018) showed examples about the potential pitfalls of metric models applied to ordinal variables (but see Robitzsch, 2020 for an alternative perspective).
+They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and biased estimates.
+While sums or averages of ordinal items could be treated as metric (Carifio \& Perla, 2008; Carifio \& Perla, 2007; but see Jamieson, 2004) Liddell and Kruschke (2018) provided some examples of potential pitfalls even in this case.
+In the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.) without considering the case of aggregating ordinal responses.
+
+For the tutorial, we assume that the reader is familiar with basic R programming and introductory theory about linear regression.
+Similar to other tutorials (DeBruine \& Barr, 2021; Gambarota \& Altoè, 2023) we used few equations that are necessary to understand the model parameters and implement the Monte Carlo simulation.
 
@@
 
-While Liddell and Kruschke (2018) proposed a specific modeling approach, there exists a class of regression models that consider the ordinal nature of the response variable without relying on metric assumptions.
-We can name this general class of models as \emph{ordinal regression}.
-The actual statistical nomenclature can be confusing mainly because there are several types of models with different assumptions and structures (Tutz, 2022).
+Despite the specific model proposed by Liddell and Kruschke (2018) proposed a specific modeling approach, there is a class of regression models that consider the ordinal nature of the response variable.
+The actual statistical nomenclature of \emph{ordinal regression} models can be confusing (Tutz, 2022).
 Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
 
-We can identify three main classes of models: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}.
-Among these, the cumulative model is the most widely used, assuming the existence of a latent variable.
-This latent variable, categorized using a set of thresholds, produces the observed ordinal variable.
-The \emph{sequential model} as suggested by the name is appropriate when modeling sequential processes.
+We can identify three main types: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}.
+Among these, the cumulative model is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
+The \emph{sequential model} is appropriate when modeling sequential processes.
 Assuming to have five response options, the sequential model assume that responding ``3'' assume a sequential process where steps ``1'' and ``2'' are already reached.
@@
 In the current paper we put the focus on the CM for several reasons.
-The first reason is that the latent formulation of the model is particularly convenient both for parameter interpretation and data simulation.
+Firstly, the latent formulation of the model is particularly convenient for parameter interpretation and data simulation.
 The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item.
-Furthermore, CM are also used to model data under a signal detection theory framework (e.g., DeCarlo, 2010).
-Figure \ref{fig:fig-explain-cumulative} depicts the overall structure of the \emph{cumulative} model.
+Furthermore, the signal detection theory framework that is very common in experimental psychology can be implemented using a CM (e.g., DeCarlo, 2010).
+Figure \ref{fig:fig-explain-cumulative} depicts the overall structure of the CM.
 
@@
 
-In this section we introduce some notation for the CM that is used through the paper.
-We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} package (Christensen, 2022).
+In this section we introduce some notation for the CM to be consistent with the existing literature and the \texttt{ordinal} R package (Christensen, 2022).
 We define \(Y_k\) as the observed ordinal variable with \(k\) levels and \(Y^\star\) is the underlying latent variable.
@@
 The specific link function define the type of model and require a different R function.
-For example, when assuming a Gaussian distribution, fitting a model with a \emph{probit} link function entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
+For example, when assuming a Gaussian distribution (\emph{probit} link function) entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
+
 When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(\mathbf{X}\) value \(P(Y \leq k | \mathbf{X}_i)\)}.
@@
 
-The same model can be written in the so-called latent formulation as reported in Equation \eqref{eq:latent-model}.
-The model is no longer directly about the cumulative probabilities but focused on the continuous latent variable \(Y^\star\) as a function of the linear predictor \(\eta = \mathbf{X} \boldsymbol{\beta}\) similar to a standard linear regression.
+Equation \eqref{eq:latent-model} shows the latent formulation of the previous model.
+The latent variable \(Y^\star\) is a function of the linear predictor \(\eta = \mathbf{X} \boldsymbol{\beta}\) similar to a standard regression.
 
@@
 
-The crucial part is \(\epsilon\) that is the random component of the model coming from a certain probability distribution.
-For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution.
+The crucial part is \(\epsilon\) that is the random part (errors) of the model.
+For a \emph{probit} model, errors are sampled from a standard Normal distribution while for a \emph{logit} model from a standard logistic distribution.
 Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).
 
-In the basic version of the model, the thresholds \(\alpha_k\) are considered as fixed and being part of the measurement procedure (Liddell \& Kruschke, 2018) and do not vary as a function of the predictors.
-In a more sophisticated version of the model called location-shift (Tutz, 2022), both the location \(\mu\) and the thresholds \(\alpha_k\) can vary as a function of the predictors.
-
+The thresholds \(\alpha_k\) are considered fixed and part of the measurement procedure (Liddell \& Kruschke, 2018; see the \emph{location-shift} models by Tutz, 2022, where thresholds vary as a function of predictors).
 The model can be also formalized in alternative ways.
-Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model with a different threshold parametrization.
-Gelman, Hill, and Vehtari (2020) proposed three alternative parametrizations focusing on different definition of the thresholds.
-
-Tutz (2022) described also another version of the model called location-scale (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022) where the location \(\mu\) and the scale \(\sigma^2\) of the distribution can vary as a function of the predictors.
+Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model or Gelman, Hill, and Vehtari (2020) for other thresholds parametrizations.
+
+The scale (spread) of the latent variable is fixed to one for both \emph{probit} and \emph{logit} models.
+In the supplementary materials we showed how to implement a location-scale models including predictors on the scale parameter (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022).
 
@@
 
-The cumulative link model implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} can be considered the general formulation that requires specifying the link function \(g(\cdot)\) or the errors distribution \(\epsilon_i \sim D(\mu, \sigma^2)\).
+The CM implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} requires specifying the link function \(g(\cdot)\) or the errors distribution \(\epsilon_i \sim D(\mu, \sigma^2)\) (\(D\) being a certain probability distribution).
 Among several available functions the \emph{logit} and \emph{probit} models are the most common.
@@
 In the next sections we will illustrate the differences and simulation strategies.
-Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table L'output è \ref{tab:tab-model-summary-latex} summarise the presented cumulative models with the proposed link function and the corresponding R code.
-
-In terms of parameters, both distributions can be defined with a location \(\mu\) and a scale \(s\) parameter.
+Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table \ref{tab:tab-model-summary-latex} summarise the presented CM with the proposed link function and the corresponding R code.
+
+In terms of parameters, both distributions can be defined with in terms of location \(\mu\) and scale \(s\).
 The standard normal distribution has \(\mu = 0\) and \(s = 1\).
@@
 \begin{table}
-
-\caption{\label{tab:tab-model-summary-latex}(ref:tab-model-summary-docx)}
+\centering
+\caption{\label{tab:tab-model-summary-latex}Summary of the probit and logit models in terms of link (\(g(\cdot)\)) and inverse link function (\(g^{-1}(\cdot)\)) and the corresponding R code.}
 \centering
@@
 \midrule
+Cumulative Logit & $\text{logit}(p) = \text{log}(p / (1-p))$ & \texttt{qlogis()} & $e^{\text{logit}(p)} / (1 + e^{\text{logit}(p)})$ & \texttt{plogis()}\\
 Cumulative Probit & $z = \Phi^{-1}(p)$ & \texttt{qnorm()} & $p = \Phi(z)$ & \texttt{pnorm()}\\
@@
 
-The function to fit the model is \texttt{clm()} and the model equation is specified using the R formula syntax \texttt{y\ \textasciitilde{}\ x} where \texttt{y} is the ordinal dependent variable and \texttt{x} is one or more predictors eventually including also the interactions.
-The package also implements mixed-effects models (see the \texttt{clmm()} function) including random intercepts and slopes but the topic is beyond the scope of the current tutorial.
+The main function is \texttt{clm()} and the formula is specified using the syntax \texttt{y\ \textasciitilde{}\ ...} where \texttt{y} is the ordinal response and \texttt{...} is a combination of predictors.
+The package also implements mixed-effects models including random intercepts and slopes.
 
@@
 \begin{table}
-
-\caption{\label{tab:tab-dataset-example-latex}(ref:tab-dataset-example-docx)}
+\centering
+\caption{\label{tab:tab-dataset-example-latex}Summary of the simulated dataset.
+For each group (a and b) we reported mean, median and standard deviation of the ordinal outcome (metric descriptive statistics).
+For each ordinal outcome (\(Y\)) we reported the frequency and (within parentheses) the probability.}
 \centering
@@
 \midrule
-a & 2.16 & 2 & 1.09 & 19 (p = 0.38) & 11 (p = 0.22) & 13 (p = 0.26) & 7 (p = 0.14)\\
-b & 3.12 & 4 & 1.1 & 7 (p = 0.14) & 6 (p = 0.12) & 11 (p = 0.22) & 26 (p = 0.52)\\
+a & 2.62 & 3 & 1.1 & \textbf{10} (p = \textit{0.2}) & \textbf{13} (p = \textit{0.26}) & \textbf{13} (p = \textit{0.26}) & \textbf{14} (p = \textit{0.28})\\
+b & 3.06 & 3 & 1.04 & \textbf{6} (p = \textit{0.12}) & \textbf{7} (p = \textit{0.14}) & \textbf{15} (p = \textit{0.3}) & \textbf{22} (p = \textit{0.44})\\
 \bottomrule
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  100  {-}127.04 262.08 4(0)  5.42e{-}07 2.0e+01}
+\CommentTok{\#\textgreater{}  logit flexible  100  {-}131.80 271.60 4(0)  2.13e{-}11 1.7e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)  }
-\CommentTok{\#\textgreater{} xb   1.5929     0.3895    4.09 4.32e{-}05 ***}
+\CommentTok{\#\textgreater{} xb   0.7556     0.3685    2.05   0.0403 *}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}0.4243     0.2745  {-}1.546}
-\CommentTok{\#\textgreater{} 2|3   0.4293     0.2711   1.584}
-\CommentTok{\#\textgreater{} 3|4   1.5851     0.3165   5.008}
+\CommentTok{\#\textgreater{} 1|2  {-}1.3328     0.3126  {-}4.264}
+\CommentTok{\#\textgreater{} 2|3  {-}0.2185     0.2703  {-}0.808}
+\CommentTok{\#\textgreater{} 3|4   0.9767     0.2893   3.377}
 \end{Highlighting}
@@
 To understand the logit model we need to introduce odds and odds ratio.
-The odds of a probability \(p\) is defined as \(p/(1 - p)\) thus the probability of success divided by the probability of failure.
-The odds takes value ranging from 0 to \(\infty\).
-For instance, with a probability of \(p = 0.8\), we have odds of \(4\), indicating that there are four successes for each failure.
-The same as having \(p = 0.2\) and an odds of \(0.25\) means that for each \(0.25\) successes we have a failure or that we have \(4\) failures for each success.
-When comparing two groups or conditions we can compare the two odds calculating an odds ratio.
-The odds ratio is the mostly used statistics to compare groups or conditions on a binary outcome.
-An odds ratio of \(4\) means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator.
+The odds of a probability \(p\) is defined as \(p/(1 - p)\) thus the success probability over the failure probability.
+The odds takes value ranging from \(0\) to \(\infty\).
+With a probability of \(p = 0.8\), we have odds of \(4\), indicating that there are four successes for each failure.
+The same as having \(p = 0.2\) and an odds of \(0.25\) means that for each \(0.25\) successes we have a failure (or \(4\) failures for each success).
+When comparing two groups or conditions we can take the ratio of two odds.
+An odds ratio (OR) of \(4\) means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator.
 The Figure \ref{fig:fig-odds-example} shows the relationship between probabilities and odds.
@@
 With an ordinal variable and a cumulative model we can use the cumulative odds ratio.
-With e.g.~\(k = 4\) outcomes we have \(k - 1\) models determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\).
-We can manually calculate the odds ratio starting from the previous dataset.
+With e.g.~\(k = 4\) outcomes we have \(k - 1\) odds ratios determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\) comparing the two groups.
+
+For example the probability of responding \(Y \leq 1\) in the group ``a'' is 0.20 corresponding to an odds of 0.25.
+When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y \leq 1\) we obtain that the group ``a'' has 1.83 times the odds of responding \(Y \leq 1\) compared to the group ``b''.
 
@@
 
-\CommentTok{\# probability of Y = 1 for the group a}
-\NormalTok{dat}\SpecialCharTok{$}\NormalTok{yn }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# better as number here}
-\NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\CommentTok{\#\textgreater{} [1] 0.38}
-
-\CommentTok{\# thus P(Y \textgreater{} 1) = 1 {-} p1a}
-\FunctionTok{odds}\NormalTok{(p1a)}
-\CommentTok{\#\textgreater{} [1] 0.6129032}
-
-\CommentTok{\# odds ratio Y = 1 a vs b}
-\NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
-\CommentTok{\#\textgreater{} [1] 0.14}
-\FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
-\CommentTok{\#\textgreater{} [1] 3.764977}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-For example the probability of responding \(Y = 1\) in the group ``a'' is 0.38 corresponding to an odds of 0.61.
-When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 3.76 times the odds of responding \(Y = 1\) compared to the group ``b''.
-
-For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(P = 1\)).
-Thus we have \(k - 1\) odds ratio or what is called a cumulative odds ratio.
-Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\).
-The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 4.92.
-In fact, the logit model essentially estimate the cumulative odds ratio.
-
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \CommentTok{\# the dummy\_ord() is a custom function creating k {-} 1 dummy variable for the cumulative probabilities}
@@
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.38    0.60    0.86}
+\CommentTok{\#\textgreater{}    0.20    0.46    0.72}
 \NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.14    0.26    0.48}
+\CommentTok{\#\textgreater{}    0.12    0.26    0.56}
 
@@
 \CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
-\CommentTok{\#\textgreater{} 3.764977 4.269231 6.654762}
+\CommentTok{\#\textgreater{} 1.833333 2.424501 2.020408}
 \end{Highlighting}
@@
 
-From the previous example, we noticed that the \(k - 1\) estimated odds ratios are different.
-However, increasing \(n\) will result in odds ratios that are very similar.
-This does not happen by chance but from an implicit (until now) assumption made by the CM called \emph{proportional odds} (PO).
 Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the PO assumption.
@@
 
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
-\CommentTok{\# using same parameters as the previous simulation increasing n}
-\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e5} \CommentTok{\# very large n}
-\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n)}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x)}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{beta =}\NormalTok{ b1, }\AttributeTok{prob0 =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
-
-\CommentTok{\# same steps as before}
-\NormalTok{cum\_p }\OtherTok{\textless{}{-}} \FunctionTok{cbind}\NormalTok{(}\AttributeTok{x =}\NormalTok{ dat}\SpecialCharTok{$}\NormalTok{x, }\FunctionTok{dummy\_ord}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y))}
-\NormalTok{group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean)}
-\NormalTok{group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean)}
-\CommentTok{\# now odds are almost the same and equal to the simulated b1 = \textasciitilde{}1.09}
-\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b)}
-\CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
-\CommentTok{\#\textgreater{} 2.914579 2.985242 2.991333}
-\FunctionTok{log}\NormalTok{(}\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b))}
-\CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
-\CommentTok{\#\textgreater{} 1.069725 1.093681 1.095719}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-Basically if we use the \emph{logit} link function \(g(\cdot)\), regression coefficients are interpreted as odds ratios.
-Given that we have \(k > 2\) alternatives we need \(k - 1\) equations.
+When fitting a logit model, the coefficients \(\beta_j\) are the (log) odds ratios for a unit increase in \(x\).
+However, the previous model estimate a single \(\beta\) even if we have four ordinal outcomes.
+The CM model (in the basic version) made an assumption called \emph{proportional odds} (PO).
+Under this assumption we assume that the \(k - 1\) cumulative odds ratios are the same.
 The PO assumption is formalized in Equation \eqref{eq:prop-odds}.
@@
 
-The PO assumption is convenient because regardless of \(k\), the \(\beta_j\) (\(j\) being the number of regression coefficients) effect is assumed to be the same.
-The model is more parsimonious compared to estimating \(k - 1\) coefficients for each \(\beta_j\) as in the multinomial regression or the non-proportional odds model.
-
-At the same time, the PO assumption could be considered too strict.
+There are CM models relaxing this assumption completely (\emph{non proportional odds}, Tutz, 2022) or partially (\emph{partial proportional odds}, Peterson \& Harrell, 1990).
+The PO model is more parsimonious but the PO assumption could be considered too strict.
 There are several methods for testing if data are supporting the PO assumption (see Liu, He, Tu, \& Tang, 2023 for an overview).
@@
 However, these methods and models are outside the scope of the tutorial.
-In the supplementary materials, we provide more details only about the \emph{location-scale} model in terms of parameters interpretation and data simulation.
-
-This can be shown by fitting the model with \texttt{clm()} that by default assume the PO and using the predicted probabilities to compute the odds ratios.
-In the following code we computed the odds ratio comparing \(x_a\) and \(x_b\) when \(k \leq 1\) and \(k \leq 2\).
-The odds ratios are exactly the same because we are using the predicted probabilities from the model (assuming PO).
+
+In the previous code block, we estimated the \(k - 1\) odds ratios and they were different even if data have been simulated under the PO assumption (see the Simulating data section).
+Calculating the odds ratios using predicted probabilities from the model fitted with \texttt{clm()} (that by default assume the PO) the odds ratios are the same.
+In the following code block we computed the odds ratio comparing \(x_a\) and \(x_b\) when \(k \leq 1\) and \(k \leq 2\).
 
@@
 \FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 2.976954}
+\CommentTok{\#\textgreater{} [1] 2.128857}
 \FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 2.976954}
+\CommentTok{\#\textgreater{} [1] 2.128857}
 \end{Highlighting}
@@
 The PO assumption is relevant only for the \emph{logit} model.
-When using a different link function (e.g., \emph{probit}) the assumption of a common \(\beta\) is called \emph{parallel slopes} (see Equation \eqref{eq:probit-prop-odds}).
-Basically the difference in \(z\) scores for a unit increase in \(x\) (i.e., the slope) is the same regardless the threshold.
+When using the \emph{probit} link function we can use the terms \emph{parallel slopes} for the assumption of a common \(\beta\) (see Equation \eqref{eq:probit-prop-odds}).
+In practical terms, difference in \(z\) scores for a unit increase in \(x\) (the actual interpretation of \(\beta\) in a \emph{probit} model) is the same regardless of the threshold.
 As for the previous example, we can fit the \emph{probit} model and manually calculate the difference in \(z\) scores.
@@
 \FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 0.6523531}
+\CommentTok{\#\textgreater{} [1] 0.4408949}
 \FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 0.6523531}
+\CommentTok{\#\textgreater{} [1] 0.4408949}
 \end{Highlighting}
@@
 \NormalTok{(alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(p, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)) }\CommentTok{\# or prob\_to\_alpha(p, "probit")}
-\CommentTok{\#\textgreater{}         y1         y2         y3         y4 }
+\CommentTok{\#\textgreater{}        1|2        2|3        3|4        4|5 }
 \CommentTok{\#\textgreater{} {-}1.3862944 {-}0.4054651  0.4054651  1.3862944}
 \FunctionTok{alpha\_to\_prob}\NormalTok{(alpha, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
-\CommentTok{\#\textgreater{} [1] 0.2 0.2 0.2 0.2 0.2}
+\CommentTok{\#\textgreater{}  p1  p2  p3  p4  p5 }
+\CommentTok{\#\textgreater{} 0.2 0.2 0.2 0.2 0.2}
 \end{Highlighting}
@@
 Then the model formula is specified within the function as \texttt{location\ =\ \textasciitilde{}} along with the vector of regression coefficients (\texttt{beta}), baseline probabilities (\texttt{prob0}) and the link function.
-The function return a dataset with the simulated \(Y\) and the latent variable (this is possible only because we simulate the data, otherwise \(Y^{\star}\) is not observed by definition).
+The function return a dataset with the simulated \(Y\) and the latent variable (this is possible only because we simulate the data, otherwise \(Y^{\star}\) cannot be observed by definition).
 
@@
 The crucial part is setting appropriate and empirically meaningful parameters.
-Thresholds \(\alpha\) are usually not of main interest (but see the location-shift models Tutz, 2022) and can be considered as intercepts in standard linear regression.
-The thresholds are quantiles of the latent distribution that produced certain \(k\) probabilities.
+In the PO model thresholds are usually not of main interest as intercepts in standard linear regression.
+Thresholds are essentially quantiles of the latent distribution that produced certain \(k\) probabilities.
 To set meaningful \(\alpha\) values we can convert probabilities into thresholds (\texttt{alpha\_to\_prob()}).
 The function \texttt{show\_alpha()} produce a meaningful visual representation of using a specific set of thresholds (see Figure \ref{fig:fig-show-th-example}).
-In regression terms, the thresholds determine the probabilities when \(x = 0\) (as the intercept).
-Thus with two groups for example, the thresholds are the \(k\) probabilities of the ordinal variable \(Y\) for the reference group.
+Thus with two groups for example, the thresholds are the \(k\) probabilities of the ordinal variable \(Y\) for the reference group (when \(x = 0\)).
 
@@
 The \texttt{cat\_latent\_plot()} and \texttt{num\_latent\_plot()} functions can be for respectively a categorical (Figures \ref{fig:fig-example-cat-latent}) and numerical predictor (Figures \ref{fig:fig-example-num-latent}).
+In the supplementary materials we show how to use the \texttt{sim\_ord\_latent()} function for checking the impact of certain \(\beta\) for more complex models.
 
@@
 On the left the shift in the latent mean and on the right the impact on the expected probabilities.
-The plot can be created using \texttt{cat\_latent\_plot(m\ =\ c(0,\ 0.5),\ s\ =\ 1,\ probs\ =\ rep(1/4,\ 4),\ link\ =\ "logit",\ plot\ =\ "both")}.}\label{fig:fig-example-cat-latent}
+The plot can be created using \texttt{cat\_latent\_plot(m\ =\ c(0,\ 1),\ s\ =\ 1,\ prob0\ =\ rep(1/4,\ 4),\ link\ =\ "logit")}.}\label{fig:fig-example-cat-latent}
 \end{figure}
@@
 \caption{Example of the \texttt{num\_latent\_plot()} function depicting the effect of a continuous predictor on the expected probabilities.
-\(\beta_1\) is the regression coefficient.
-The plot can be created using \texttt{num\_latent\_plot(x\ =\ runif(100),\ b1\ =\ 1,\ probs\ =\ c(0.6,\ 0.2,\ 0.1,\ 0.1),\ link\ =\ "probit")}}\label{fig:fig-example-num-latent}
+Dotted lines are the 25\%, 50\% and 75\% quantiles.
+\(\beta_1 = 1\) is the regression coefficient.
+The plot can be created using \texttt{num\_latent\_plot(x\ =\ runif(100),\ b1\ =\ 1,\ prob0\ =\ c(0.6,\ 0.2,\ 0.1,\ 0.1),\ link\ =\ "probit")}}\label{fig:fig-example-num-latent}
 \end{figure}
@@
 We present very general example that, as for the previous simulations, can be easily extended.
-For example, by simulating a model fixing a specific parameter to zero, we can estimate the type-1 error rate (wrongly rejecting \(H_0\)) or assessing an eventual bias in model estimates.
 
@@
 
+\bibitem[\citeproctext]{ref-Overgaard2021-hs}
+Overgaard, M., \& Sandberg, K.
+(2021).
+The perceptual awareness scale---recent controversies and debates.
+\emph{Neuroscience of Consciousness}.
+Retrieved from \url{https://academic.oup.com/nc/article-pdf/doi/10.1093/nc/niab044/41765024/niab044.pdf}
+
 \bibitem[\citeproctext]{ref-Peterson1990-cz}

### 2024-01-26 supplementary/location-scale-models.qmd
major update

@@
                       message = FALSE)
+mtheme <- function(){
+  theme_minimal(15)
+}
+theme_set(mtheme())
 ```
@@
 
-The default ordinal regression model assume that the variance of the underlying latent distribution is the same across condition.
-This is similar to a standard linear regression assuming the homogeneity of variance.
-For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy].
+The default ordinal regression model assume that the variance of the underlying latent distribution is the same across conditions.
+This is similar to a standard linear regression assuming the homogeneity of variances.
+
+For example, when comparing two groups or conditions we can run a standard linear model (i.e., a t-test) assuming homogeneity of variances or using the Welch t-test [see @Delacre2017-qy] that relaxes the assumption.
 In addition, there are the so-called location-scale models that allows to include predictors also for the scale (e.g., the variance) of the distribution.
+
 This can be done also in ordinal regression where instead of assuming the same variance between conditions, the linear predictors can be included.
@@
 $$
-Y^\star_i = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon} \;\;\; \epsilon_i \sim \mathcal{N}(0, e^{Z\zeta})
+Y^\star = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon} \;\;\; \epsilon_i \sim \mathcal{N}(0, e^{Z\zeta})
 $$ {#eq-latent-scale-location}
@@
 In this example, the only way to capture group differences is by including a scale effect.
-The fig-scale-effect-example) depict the impact of having different scales between two groups on the ordinal probabilities.
+The @fig-scale-effect-example depict the impact of having different scales between two groups on the ordinal probabilities.
 
@@
 #| label: fig-scale-effect-example
-cat_latent_plot(m = c(0, 1), s = c(1, 2), probs = rep(1/4, 4), link = "logit", plot = "both")
+cat_latent_plot(location = c(0, 1), scale = c(1, 2), prob0 = rep(1/4, 4), link = "logit", plot = FALSE) |> 
+  cowplot::plot_grid(plotlist = _, nrow = 1)
 ```
@@
 #| fig-height: 10
-sigma_high <- cat_latent_plot(m = c(0, 0), 
-                s = c(1, 2), 
-                probs = rep(1/4, 4), 
+
+ggrid <- function(plt, title){
+  plt[[1]] <- plt[[1]] + ggtitle(title)
+  cowplot::plot_grid(plotlist = plt, nrow = 1)
+}
+
+sigma_high <- cat_latent_plot(location = c(0, 0), 
+                scale = c(1, 2), 
+                prob0 = rep(1/4, 4), 
                 link = "probit",
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$"))
-sigma_low <- cat_latent_plot(m = c(0, 0), 
-                s = c(1, 0.5), 
-                probs = rep(1/4, 4), 
+                plot = FALSE)
+
+sigma_low <- cat_latent_plot(location = c(0, 0), 
+                scale = c(1, 0.5), 
+                prob0 = rep(1/4, 4), 
                 link = "probit",
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$"))
-sigma_high_mu1 <- cat_latent_plot(m = c(0, 1), 
-                s = c(1, 2), 
-                probs = rep(1/4, 4), 
+                plot = FALSE)
+
+sigma_high_mu1 <- cat_latent_plot(location = c(0, 1), 
+                scale = c(1, 2), 
+                prob0 = rep(1/4, 4), 
                 link = "probit",
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 2$, $\\mu_2 = 1$"))
-sigma_low_mu1 <- cat_latent_plot(m = c(0, 1), 
-                s = c(1, 0.5), 
-                probs = rep(1/4, 4), 
+                plot = FALSE)
+
+sigma_low_mu1 <- cat_latent_plot(location = c(0, 1), 
+                scale = c(1, 0.5), 
+                prob0 = rep(1/4, 4), 
                 link = "probit",
-                plot = "both",
-                title = latex2exp::TeX("$\\sigma^2_{g_2} = 0.5$, $\\mu_2 = 1$"))
-plot_grid(sigma_high, sigma_low, sigma_high_mu1, sigma_low_mu1)
+                plot = FALSE)
+
+plot_grid(
+  ggrid(sigma_high, latex2exp::TeX("$\\mu = 0, 0$, $\\sigma = 1, 2$")), 
+  ggrid(sigma_low, latex2exp::TeX("$\\mu = 0, 0$, $\\sigma = 1, 0.5$")), 
+  ggrid(sigma_high_mu1, latex2exp::TeX("$\\mu = 0, 1$, $\\sigma = 1, 2$")), 
+  ggrid(sigma_low_mu1, latex2exp::TeX("$\\mu = 0, 1$, $\\sigma = 1, 0.5$"))
+)
 ```
@@
   ggplot(aes(x = xp, y = `(Intercept)` + x * xp)) +
-  geom_line(aes(color = model))
+  geom_line(aes(color = model)) +
+  theme_minimal(15)
 ```
@@
 #| echo: false
+#| fig-height: 8
+#| fig-width: 8
 location <- 0
@@
 ```{r}
-cat_latent_plot(c(0, 0 + b1), s = c(exp(0), (exp(0) + z1)), probs = rep(1/k, k), link = "logit", plot = "both")
+cat_latent_plot(c(0, 0 + b1), s = c(exp(0), (exp(0) + z1)), prob0 = rep(1/k, k), link = "logit")
 ```

### 2024-01-26 supplementary/ordinal-notes.qmd
major update

---
title: "Ordinal Notes"
format: 
  html:
    toc: true
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(cowplot)
library(ordinal)
devtools::load_all()

knitr::opts_chunk$set(echo = TRUE,
                      dev = "svg",
                      warning = FALSE,
                      message = FALSE)

mtheme <- function(){
  theme_minimal(15)
}
theme_set(mtheme())
```

## $-\mathbf{X}\boldsymbol{\beta}$ vs $\mathbf{X}\boldsymbol{\beta}$ parametrization

In the tutorial we used the $\alpha_k -\mathbf{X}\boldsymbol{\beta}$ parametrization (thus with the minus sign) because this force the $\beta$ to have the interpretation as in standard regression models. Usually, $\beta$ is the increase in $y$ for a unit increase in $x$. A negative $\beta$ means that the expected value of $y$ decrease for an increase in $x$. Let's see an example using the positive sign for $\beta$.

```{r}
k <- 4 # number of ordinal outcomes
x <- c(0, 1) # a binary predictor
b1 <- log(3) # log odds ratio comparing x1 and x0
probs0 <- rep(1/k, k)
alpha <- prob_to_alpha(probs0, link = "logit")
X <- matrix(x, nrow = 2)

# positive sign
(lp <- lapply(alpha, function(a) c(a + X %*% b1)))
cump <- lapply(lp, plogis)
cump <- cbind(0, data.frame(cump), 1)
p <- data.frame(t(apply(cump, 1, diff)))
names(p) <- paste0("y", 1:k)
p$x <- x
p

p |> 
  pivot_longer(starts_with("y")) |> 
  ggplot(aes(x = x, y = value, fill = name)) +
  geom_col(position = position_dodge()) +
  ylab("Probability") +
  ylim(c(0, 1)) +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  ggtitle(latex2exp::TeX("$P(Y \\leq k) = \\alpha_k + X\\beta$"))
```

Clearly a positive $\beta$ create higher probability for lower $Y$ categories. This can be somehow not intuitive thus we can use the negative sign.

```{r}
# negative sign

(lp <- lapply(alpha, function(a) c(a - X %*% b1)))
cump <- lapply(lp, plogis)
cump <- cbind(0, data.frame(cump), 1)
p <- data.frame(t(apply(cump, 1, diff)))
names(p) <- paste0("y", 1:k)
p$x <- x

p |> 
  pivot_longer(starts_with("y")) |> 
  ggplot(aes(x = x, y = value, fill = name)) +
  geom_col(position = position_dodge()) +
  ylab("Probability") +
  ylim(c(0, 1)) +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  ggtitle(latex2exp::TeX("$P(Y \\leq k) = \\alpha_k - X\\beta$"))
```

Using the second parametrization, with positive $\beta$ we have higher probability for higher $Y$ categories and the opposite.

The negative-sign parametrization is implicit when simulating from the latent distribution. Let's see an example. We use a continuous $x$ predictor because it is easier to see the results.

```{r}
#| fig-cap: Simulated data using the latent variable approach. The dotted lines are the thresholds $\\alpha$
n <- 1e3
x <- runif(1e3)
B <- 3
ys_p <- x * B + rnorm(n)
ys_n <- x * -B + rnorm(n)
y_p <- findInterval(ys_p, alpha) + 1
y_n <- findInterval(ys_n, alpha) + 1

par(mfrow = c(1,2))
plot(x, ys_p, col = y_p, pch = 19, ylim = c(-7, 7), main = latex2exp::TeX("$\\Y^{*} = X \\beta + \\epsilon$, $\\beta = 3$"), ylab = latex2exp::TeX("$Y^{*}$"))
abline(h = alpha, lty = "dashed")
legend("bottomleft", fill = 1:k, legend = paste0("Y", 1:k))

plot(x, ys_n, col = y_n, pch = 19, ylim = c(-7, 7), main = latex2exp::TeX("$\\Y^{*} = X \\beta + \\epsilon$, $\\beta = -3$"), ylab = latex2exp::TeX("$Y^{*}$"))
abline(h = alpha, lty = "dashed")
```

## Checking the impact of $\mathbf{\beta}$

Choosing one or more plausible $\beta_j$ values can be challenging. For a single $\beta$ we can easily think about the odds ratio (for a *logit* model) or the Cohen's $d$ (for a *probit*) model. With multiple predictors and their interactions is not easy to fix plausible values. A good strategy is to try different values and check the impact on the predicted probabilities. In practice, we need to compute the predicted probabilities using the model equation for the $k$ ordinal outcomes. This can be easily done with the `sim_ord_latent()` function, fixing the `simulate = FALSE` parameter. In this way, only the predicted probabilities are computed. Let's see an example for a single $x$ sampled for a standard normal distribution.

```{r}
k <- 4
dat <- data.frame(x = seq(-4, 4, 0.1))
b1 <- 0.5
probs0 <- rep(1/k, k)
dat <- sim_ord_latent(~x, beta = b1, prob0 = probs0, data = dat, simulate = FALSE)
head(dat)
```

Then we can plot the results:

```{r}
dat |> 
  pivot_longer(matches("^y[1-9]")) |> 
  ggplot(aes(x = x, y = value, color = name)) +
  geom_line()
```

In this case the $\beta_1 = 0.5$ can be considered a plausible value. Let's see what happens increasing it:

```{r}
data.frame(x = seq(-4, 4, 0.1)) |> 
  sim_ord_latent(~x, beta = 4, prob0 = probs0, data = _, simulate = FALSE) |> 
  pivot_longer(matches("^y[1-9]")) |> 
  ggplot(aes(x = x, y = value, color = name)) +
  geom_line()
```

We can clearly see the difference and probably $\beta = 4$ can be considered too large. To note, the same result can be achieved using the `num_latent_plot()` (that under the hood uses the `sim_ord_latent()` function).

Let's make now an example, with an interaction between a continous and categorical predictor.

```{r}
dat <- expand_grid(x = seq(-4, 4, 0.1), g = c("a", "b"))
dat$g <- factor(dat$g)
contrasts(dat$g) <- c(-0.5, 0.5)
beta <- c(b1 = 0.5, b2 = 1, b3 = 0.1)
dat <- sim_ord_latent(~ x * g, beta = beta, prob0 = probs0, data = dat, simulate = FALSE)

dat |> 
  pivot_longer(matches("^y[1-9]")) |> 
  ggplot(aes(x = x, y = value, color = name, lty = g)) +
  geom_line()
```

We can see the impact of $\beta_3 = 0.1$ that is the difference in slopes between the two groups. We can also use another plot to better see the group effect on each $Y$.

```{r}
dat |> 
  pivot_longer(matches("^y[1-9]")) |> 
  ggplot(aes(x = x, y = value, color = g)) +
  geom_line() +
  facet_wrap(~name)
```

### 2024-01-26 R/deprecated/deprecated.R

@@ -1,165 +1,0 @@
-get_y <- function(data = NULL, P, ordered = TRUE){
-  y <- apply(P, 1, function(x) sample(names(P), 1, replace = TRUE, prob = x))
-  y <- factor(y, levels = names(P), ordered = TRUE)
-  if(!is.null(data)){
-    cbind(data, y)
-  }else{
-    y
-  }
-}
-
-codds <- function(p){
-  co <- cumsum(p)/(1 - cumsum(p))
-  co <- co[-length(co)]
-  nn <- Reduce(paste0, 1:(length(p) - 1), accumulate = TRUE)
-  names(co) <- paste0("o", nn)
-  co
-}
-
-
-show_th <- function(th = NULL, probs = NULL, link = c("logit", "probit"),
-                    cnames = NULL){
-  link <- match.arg(link)
-  if(!is.null(th) & !is.null(probs)){
-    stop("Only one between th and probs need to be supplied")
-  }
-  
-  fn <- get_link(link)
-  
-  if(is.null(th)){
-    th <- probs_to_th(probs, link = link)
-  }
-  if(is.null(probs)){
-    probs <- th_to_probs(th, link = link)
-  }
-  
-  if(is.null(cnames)){
-    cnames <- 1:length(probs)
-  }
-  
-  if(sum(probs) != 1){
-    stop("The vector of probabilities must sum to 1")
-  }
-  
-  par(mfrow = c(1,3))
-  curve(fn$dfun(x), -5, 5, main = "Latent Distribution",
-        ylab = fn$dfun_n)
-  abline(v = th)
-  
-  curve(fn$pfun(x), -4, 4, main = "Cumulative Probabilities",
-        ylab = fn$pfun_n)
-  segments(th, 0, th, fn$pfun(th))
-  
-  barplot(probs, ylim = c(0, 1), main = "Expected Probabilities",
-          col = RColorBrewer::brewer.pal(n = length(probs), name = "RdBu"),
-          names = cnames)
-  
-  par(mfrow = c(1,1))
-}
-
-dummy_ord <- function(y){
-  if(!is.numeric(y)) y <- as.integer(y)
-  yc <- sort(unique(y))
-  dummy <- lapply(yc, function(t) ifelse(y <= t, 1, 0))
-  nn1 <- Reduce(paste0, yc, accumulate = TRUE)
-  nn2 <- Reduce(paste0, yc, accumulate = TRUE, right = TRUE)
[…]

### 2024-01-26 R/utils-paper.R

@@ -1,130 +1,5 @@
-tidy_clm <- function(fit, ...){
-  fits <- broom::tidy(fit, ...)
-  return(fits)
-}
-
-clm_table <- function(fit, truth = NULL, caption = NULL){
-  require(flextable)
-  require(ftExtra)
-  fits <- tidy_clm(fit)
-  nalpha <- length(fit$alpha)
-  if(!is.null(truth)){
-    fits$estimate <- sprintf("%.3f (*%.3f*)", fits$estimate, truth)
-  }
-  fits <- dplyr::select(fits, coef.type, dplyr::everything())
-  fits$coef.type <- dplyr::case_when(
-    fits$coef.type == "intercept" ~ "$\\alpha$",
-    fits$coef.type == "location" ~ "$\\mu$",
-    fits$coef.type == "scale" ~ "$\\sigma$",
-    TRUE ~ NA
-  )
-  
-  flextable(fits) |> 
-    flextable::merge_v(1) |> 
-    colformat_double(digits = 3) |> 
-    autofit() |> 
-    theme_vanilla() |> 
-    set_header_labels(coef.type = "", 
-                      term = "Parameter", 
-                      estimate = "Value", 
-                      std.error = "SE",
-                      statistic = "z",
-                      p.value = "p") |> 
-    align(part = "all", align = "center") |> 
-    colformat_md(part = "all") |> 
-    set_caption(caption = caption)
-}
-
-z_to_x <- function(z, mean, sd){
-  (z * sd) + mean
-}
-
-clm_table_old <- function(fit, truth = NULL, caption = NULL){
-  require(gt)
-  fits <- tidy_clm(fit)
-  nalpha <- length(fit$alpha)
-  if(!is.null(truth)){
-    fits$estimate <- sprintf("%.3f (*%.3f*)", fits$estimate, truth)
-  }
-  fits <- dplyr::select(fits, coef.type, dplyr::everything())
-  fits$coef.type <- dplyr::case_when(
-    fits$coef.type == "intercept" ~ "$$\\alpha$$",
-    fits$coef.type == "location" ~ "$$\\mu$$",
-    fits$coef.type == "scale" ~ "$$\\sigma$$",
-    TRUE ~ NA
-  )
-  fits |> 
-    mutate(row = row_number(), .by = coef.type) |> 
-    mutate(across(coef.type, ~ifelse(row == 1, ., ""))) |> 
-    gt::gt() |> 
[…]

### 2024-01-26 R/utils-plotting.R

@@ -1,6 +1,18 @@
+#' Show the thresholds and associated probabilities
+#'
+#' @param prob vector of probabilities
+#' @param alpha vector of thresholds
+#' @param link link function (logit or probit)
+#'
+#' @return a list of plots
+#' @export
 show_alpha <- function(prob = NULL, 
                        alpha = NULL, 
-                       link = c("logit", "probit")){
+                       link = "logit",
+                       plot = TRUE){
+  
+  # not elegant, but effective
+  suppressMessages(suppressWarnings(library(ggplot2)))
   
   if(is.null(prob) & is.null(alpha)){
     stop("alpha or prob must be specified!")
@@ -9,9 +21,7 @@ show_alpha <- function(prob = NULL,
   if(is.null(prob)){
     prob <- alpha_to_prob(alpha, link)
   }else{
-    if(sum(prob) != 1){
-      stop("the prob vector must sum to 1!")
-    }
+    sum_to_1(prob)
     alpha <- prob_to_alpha(prob, link)
   }
   
@@ -35,7 +45,7 @@ show_alpha <- function(prob = NULL,
     ggdist::stat_slab(aes(dist = dist,
                           fill = factor(after_stat(findInterval(x, alpha) + 1))),
                       orientation = "horizontal",
-                      alpha = 0.8) +
+                      alpha = 1) +
     xlim(xlim) +
     theme_minimal(15) +
     theme(legend.title = element_blank(),
@@ -55,16 +65,20 @@ show_alpha <- function(prob = NULL,
   y_cp <- lf$pfun(c(-Inf, alpha, Inf))
   y_p <- 0.5 * (y_cp[1:(length(y_cp) - 1)] + y_cp[2:length(y_cp)])
   
-  pplot <- ggplot() +
-    stat_function(geom = "line", fun = lf$pfun) +
-    xlim(xlim) +
-    geom_segment(aes(x = alpha, y = 0, xend = alpha, yend = lf$pfun(alpha))) +
-    geom_segment(aes(x = -Inf, y = lf$pfun(alpha), xend = alpha, yend = lf$pfun(alpha))) +
+  pplot <- ggplot(data = NULL, aes(x = xlim)) +
+    geom_line(stat = "function", 
+              fun = lf$pfun, 
+              #aes(color = factor(after_stat(findInterval(x, alpha)))),
+              linewidth = 1.5) +
+    geom_segment(data = NULL, aes(x = alpha, y = 0, xend = alpha, yend = lf$pfun(alpha))) +
[…]

### 2024-01-26 R/utils-project.R

@@ -10,3 +10,17 @@ render_paper <- function(){
   rmarkdown::render(here::here("paper", "paper.Rmd"), output_format = "all")
   anonymize_paper("paper/paper.pdf")
 }
+
+.get_packages <- function(){
+  options(renv.verbose = FALSE)
+  pkgs <- renv::dependencies()
+  pkgs <- unique(pkgs$Package)
+  pkgs <- pkgs[!pkgs %in% c("R", "shiny")]
+  pkgs_data <- data.frame(
+    pkg = pkgs
+  )
+  
+  pkgs_data$version <- sapply(pkgs_data$pkg, 
+                              function(x) paste(packageVersion(x), collapse = "."))
+  pkgs_data
+}

### 2024-01-26 R/utils.R

@@ -18,9 +18,9 @@
 #' set.seed(123)
 #' n <- 100
 #' k <- 5 # 5 ordinal values
-#' bprobs <- rep(1/5, 5) # baseline probabilities for the group 0
+#' probs0 <- rep(1/k, k) # baseline probabilities for the group 0
 #' dat <- data.frame(x = rep(c(0, 1), each = n))
-#' dat <- sim_ord_latent(~x, By = 0.5, prob = bprobs, data = dat, link = "probit")
+#' dat <- sim_ord_latent(~x, By = 0.5, prob = probs0, data = dat, link = "probit")
 sim_ord_latent <- function(location,
                            scale = NULL,
                            beta,
@@ -28,7 +28,7 @@ sim_ord_latent <- function(location,
                            prob0 = NULL, 
                            alpha = NULL, 
                            data, 
-                           link = c("logit", "probit"),
+                           link = "logit",
                            simulate = TRUE){
   
   # get all the correct functions according to the link
@@ -63,7 +63,7 @@ sim_ord_latent <- function(location,
     data$ys <- ystar # save also the latent
   } else{
     alpha <- c(-Inf, alpha, Inf)
-    cp <- lapply(alpha, function(a) plogis((a - lploc) / exp(lpscale)))
+    cp <- lapply(alpha, function(a) lf$pfun((a - lploc) / exp(lpscale)))
     cp <- data.frame(do.call(cbind, cp))
     p <- data.frame(t(apply(cp, 1, diff)))
     names(p) <- paste0("y", 1:ncol(p))
@@ -87,11 +87,13 @@ sim_ord_latent <- function(location,
 #' alpha <- c(-2, -0.5, 1, 1.5)
 #' alpha_to_prob(alpha, link = "logit")
 #' alpha_to_prob(alpha, link = "probit")
-alpha_to_prob <- function(alpha, link = c("logit", "probit"), ...){
+alpha_to_prob <- function(alpha, link = "logit", ...){
   lf <- get_link(link)
   ths <- c(-Inf, unname(alpha), Inf)
   cprobs <- lf$pfun(ths, ...)
-  diff(cprobs)
+  prob <- diff(cprobs)
+  names(prob) <- paste0("p", 1:length(prob))
+  return(prob)
 }
 
 #' Transform probabilities into thresholds
[…]

### 2024-01-26 deprecated/deprecated.R

@@ -1,0 +1,309 @@
+tidy_clm <- function(fit, ...){
+  fits <- broom::tidy(fit, ...)
+  return(fits)
+}
+
+plot_grid_common_legend <- function(..., nrow = NULL, ncol = NULL){
+  plts <- list(...)
+  legend <- cowplot::get_legend(plts[[1]] + theme(legend.position = "bottom", legend.title = element_blank()))
+  plts <- lapply(plts, function(p) p + theme(legend.position = "none"))
+  grid <- cowplot::plot_grid(plotlist = plts)
+  cowplot::plot_grid(grid, legend, nrow = nrow, ncol = ncol, rel_heights = c(0.9, 0.1))
+}
+
+
+z_to_x <- function(z, mean, sd){
+  (z * sd) + mean
+}
+
+clm_table_old <- function(fit, truth = NULL, caption = NULL){
+  require(gt)
+  fits <- tidy_clm(fit)
+  nalpha <- length(fit$alpha)
+  if(!is.null(truth)){
+    fits$estimate <- sprintf("%.3f (*%.3f*)", fits$estimate, truth)
+  }
+  fits <- dplyr::select(fits, coef.type, dplyr::everything())
+  fits$coef.type <- dplyr::case_when(
+    fits$coef.type == "intercept" ~ "$$\\alpha$$",
+    fits$coef.type == "location" ~ "$$\\mu$$",
+    fits$coef.type == "scale" ~ "$$\\sigma$$",
+    TRUE ~ NA
+  )
+  fits |> 
+    mutate(row = row_number(), .by = coef.type) |> 
+    mutate(across(coef.type, ~ifelse(row == 1, ., ""))) |> 
+    gt::gt() |> 
+    gt::fmt_number(decimals = 3) |> 
+    tab_caption(caption)
+}
+
+clm_table <- function(fit, truth = NULL, caption = NULL){
+  require(flextable)
+  require(ftExtra)
+  fits <- tidy_clm(fit)
+  nalpha <- length(fit$alpha)
+  if(!is.null(truth)){
+    fits$estimate <- sprintf("%.3f (*%.3f*)", fits$estimate, truth)
+  }
+  fits <- dplyr::select(fits, coef.type, dplyr::everything())
+  fits$coef.type <- dplyr::case_when(
+    fits$coef.type == "intercept" ~ "$\\alpha$",
+    fits$coef.type == "location" ~ "$\\mu$",
[…]

### 2024-01-29 paper/paper.Rmd
before manual revisione

@@
 abstract: |
-  Ordinal data are very common in psychology such as Likert items, ratings or generic ordinal variables.
-Despite the widespread these variables are usually analyzed using metric models (e.g., standard linear regression).
-This approach can have important drawbacks in terms of statistical inference (e.g., reduced power) and prediction.
+  Ordinal data are very common in psychology such as Likert items, ratings or generic ordered variables.
+Despite the widespread, these variables are usually analyzed using metric models (e.g., standard linear regression).
+This approach can have important drawbacks in terms of statistical inference (e.g., reduced power and increased type-1 error) and prediction.
 One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis.

### 2024-01-29 paper/paper.Rmd
reducing words

@@
 Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels [@Liddell2018-wu].
-More generally, @Cliff2016-ck suggests that several research questions in behavioral sciences can be considered as ordinal (*is the score* $x$ higher than the score $y$?) concerning variables where the most appropriate measurement scale is probably ordinal.
-
-In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a linear regression that makes an explicit assumption about metric features of the response variable.
+
+In Psychology especially with item-based measures (questionnaires, surveys, etc.) the common practice is running a linear regression that makes an explicit assumption about metric features of the response variable.
 @Liddell2018-wu reviewed the psychological literature using Likert-based measures and reported how the majority of papers used metric-based statistical models.
@@
 ```{r tutorial-pkgs, eval = FALSE}
+#| echo: true
 library(dplyr)
@@
 
-Despite the specific model proposed by @Liddell2018-wu proposed a specific modeling approach, there is a class of regression models that consider the ordinal nature of the response variable.
-The actual statistical nomenclature of *ordinal regression* models can be confusing [@Tutz2022-dg].
-@Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
-
-We can identify three main types: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe], and *adjacent category models*.
-Among these, the cumulative model is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
-The *sequential model* is appropriate when modeling sequential processes.
-Assuming to have five response options, the sequential model assume that responding "3" assume a sequential process where steps "1" and "2" are already reached.
-A clear example is proposed by @Burkner2019-aw where the marriage duration in years is predicted as a function of some explanatory variables.
-For each level of the response variable there is a latent distribution where the step between a marriage year $k = 1$ and the next years $k > 1$ is modeled by the sequential model.
-When comparing $k$ with $k > 1$, everything lower than $k$ is assumed to be already reached [@Tutz2020-xq].
-
-The adjacent category model compare the category $k$ with $k + 1$ still assuming a latent distribution for each $k$.
-As suggested by @Tutz2022-dg the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
-@Burkner2019-aw suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical interpretation as for the cumulative vs sequential model.
-
-In the current paper we put the focus on the CM for several reasons.
+We can generally define as *ordinal regression*, a statistical model explicitly considering the ordinal (and not metric) nature of the response variable.
+The nomenclature of *ordinal regression* models can be confusing but @Tutz2022-dg and @Burkner2019-aw provide a clear and updated taxonomy of ordinal regression models.
+
+We can identify three main models: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe], and *adjacent category models*.
+Among these, the CM is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
+The *sequential model* is appropriate when modeling sequential processes where responding to a certain category implies have already reached lower categories.
+The *adjacent category model* compares a category with the successive one and can be seen  the adjacent-category model can be seen as a series of binomial regressions taking into account the ordering [@Tutz2022-dg].
+
+In the current paper we focus on the CM for several reasons.
 Firstly, the latent formulation of the model is particularly convenient for parameter interpretation and data simulation.
-The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item.
+The second reason is that several psychological construct can be formalized as a latent continuous variable observed as an ordinal item(s).
 Furthermore, the signal detection theory framework that is very common in experimental psychology can be implemented using a CM [e.g., @DeCarlo2010-lj].
@@
 The latent variable is segmented using $k - 1$ thresholds $\alpha_1, \dots, \alpha_{k - 1}$.
-Similarly to the generalized linear models framework [e.g., @Fox2015-ps], we define $g(x) = \eta$ as the link function that maps probabilities into the linear predictor $\eta$.
-To transform back $\eta$ into probabilities we use the inverse of the link function $x = g^{-1}(\eta)$.
-The specific link function define the type of model and require a different R function.
-For example, when assuming a Gaussian distribution (*probit* link function) entails using the cumulative distribution function $g(x) = \Phi^{-1}(x) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $x = g^{-1}(\eta) = \Phi(\eta)$.
-
-When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$^[As done by @Agresti2010-rz, when referring to $P(Y \leq k)$ we are implicitly conditioning on a particular $\mathbf{X}$ value $P(Y \leq k | \mathbf{X}_i)$].
-Equation \@ref(eq:prob-cum-model1) shows the general cumulative model including predictors $\mathbf{X}$ and regression coefficients $\boldsymbol{\beta}$.
-The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to interpret the $\beta_j$ as in the standard regression models where higher $\beta$ values corresponds to increased probability of responding higher $k$ categories [@Agresti2010-rz].
+Similarly to the generalized linear models framework [e.g., @Fox2015-ps], we define $g(p) = \eta$ as the link function that maps probabilities into the linear predictor $\eta$.
+To transform back $\eta$ into probabilities we use the inverse of the link function $p = g^{-1}(\eta)$.
+The specific link function define the type of model and require different R functions.
+For example, using a Normal distribution (*probit* link function) requires using the cumulative distribution function $g(p) = \Phi^{-1}(p) = \eta$ and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined $p = g^{-1}(\eta) = \Phi(\eta)$.
+
+When modelling an ordinal variable in a CM we include predictors on the cumulative probability $P(Y \leq k), k = 1, \dots, k - 1$^[As done by @Agresti2010-rz, when referring to $P(Y \leq k)$ we are implicitly conditioning on a particular $\mathbf{X}$ value $P(Y \leq k | \mathbf{X}_i)$].
+Equation \@ref(eq:prob-cum-model1) shows the general cumulative model with predictors $\mathbf{X}$ and coefficients $\boldsymbol{\beta}$.
+The minus sign in $\mathbf{X} \boldsymbol{\beta}$ is used to interpret the $\beta_j$ as in the standard regression models where higher positive values corresponds to increased probability of responding higher $k$ categories [@Agresti2010-rz].
 
@@
 The $\mathbf{X} \boldsymbol{\beta}$ is the linear predictor $\eta$ that is the cumulative probability $P(Y \leq k)$ transformed using the link function $g(\cdot)$.
-To obtain the probability of a single outcome $P(Y = k)$ we can compute the difference between cumulative probabilities as shown in Equation \@ref(eq:prob-cum-model2).
+The probability of a single outcome $P(Y = k)$ can be calculated as the difference between cumulative probabilities (see Equation \@ref(eq:prob-cum-model2)).
 
@@
 There are always two special cases when computing the probability of a single outcome $Y$ that is when $Y = 1$ and $Y = k$.
-In the first case the cumulative probability is calculated from $-\infty$ that is the same as temporary assuming an $\alpha_0 = -\infty$.
-In the second case ($Y = 1$) the probability is calculated as $P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)$ that is the same as assuming a temporary threshold $\alpha_k = +\infty$.
+In the first case the cumulative probability is calculated from $-\infty$ (assuming that $\alpha_0 = -\infty$).
+In the second case ($Y = 1$) the probability is calculated as $P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)$ (a $\alpha_k = +\infty$.
 The Figure \@ref(fig:fig-explain-cumulative) shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.
@@
 \begin{equation}
-\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \mathbf{\epsilon}
+\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
 (\#eq:latent-model)
@@
 
-The crucial part is $\epsilon$ that is the random part (errors) of the model.
-For a *probit* model, errors are sampled from a standard Normal distribution while for a *logit* model from a standard logistic distribution.
+The $\epsilon$ term is the random part (errors) of the model.
+For a *probit* model, errors are distributed as a standard Normal distribution and for a *logit* model as a standard logistic distribution.
 Following the notation by @Tutz2022-dg, the observed ordinal value $Y_i = k$ comes from $Y^\star_i$ belonging to the interval defined by the thresholds $Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}$ where $- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty$.
 
-The thresholds $\alpha_k$ are considered fixed and part of the measurement procedure [@Liddell2018-wu; see the *location-shift* models by @Tutz2022-dg, where thresholds vary as a function of predictors].
-The model can be also formalized in alternative ways.
-@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian version of the model or @Gelman2020-tg for other thresholds parametrizations.
-
-The scale (spread) of the latent variable is fixed to one for both *probit* and *logit* models.
-In the supplementary materials we showed how to implement a location-scale models including predictors on the scale parameter [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur].
+In the CM, thresholds $\alpha_k$ are considered fixed and part of the measurement procedure [@Liddell2018-wu; but see the *location-shift* models by @Tutz2022-dg, where thresholds vary as a function of predictors] and can be considered as intercepts.
+The proposed CM can be parametrized in different ways.
+@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian version of the model and @Gelman2020-tg proposed other thresholds parametrizations.
 
@@
 Among several available functions the *logit* and *probit* models are the most common.
-The *logit* model use a *logit* link function and a logistic distribution as latent variable.
-On the other side, the *probit* model assume a latent Gaussian distribution using the inverse of the cumulative distribution ($\Phi^{-1}$) function as link function.
+The *logit* model use a *logit* link function and a standard logistic distribution as latent variable.
+On the other side, the *probit* model assume a standard Normal distribution using the inverse of the cumulative distribution ($\Phi^{-1}(\cdot)$) function as link function.
 
@@
 In practical terms, fixing $\mu$ and $s$ lead to an higher standard deviation for the logistic distribution.
+In the supplementary materials we showed how to implement a location-scale models for including predictors on the scale parameter [@Tutz2022-dg; @Rigby2005-ko; @Cox1995-ur].
 
@@
 
-For fitting the cumulative models we used the `ordinal` package [@Christensen2019-cz].
-Despite the presence of other possibilities the `ordinal` package provide the most complete and intuitive way to implement the ordinal models.
-The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.[^2]
+For fitting the CM we used the `ordinal` package [@Christensen2019-cz].
+The `ordinal` package provide a complete and intuitive way to implement the ordinal models.
+The syntax is very similar to standard linear models in R with functions to calculate predictions, perform model comparison and extract relevant model information.[^2]
 
@@
 
-When fitting the model the crucial arguments are the `formula`, the `link` function and the `data`. More advanced arguments are the `nominal`, `scale` and `threshold`.
+When fitting the model the crucial arguments are the `formula`, the `link` function and the `data`. More advanced arguments are `nominal`, `scale` and `threshold`.
 
 -   `formula`: the formula `y ~ x` with the dependent variable and predictors.
--   `link`: is the link function. In this tutorial we consider only the *logit* and *probit* link but other link functions are available.
+-   `link`: is the link function (in this tutorial only *logit* and *probit*)
 -   `data`: is the dataset with variables included in the `formula`
 -   `nominal`: formula with predictors where the proportional odds assumption (See Section \@ref(proportional-odds-assumption)) is relaxed (i.e., partial or non proportional odds)
--   `scale`: formula with predictors for the scale (standard deviation) parameter.
-This argument allow to fit a scale-location model (see supplementary materials).
-The main `formula` argument refers to predictors on the location parameter (i.e., the mean $\mu$).
+-   `scale`: formula with predictors for the scale parameter for fitting a scale-location model
 -   `threshold`: different structures for estimating the thresholds. The default is `threshold = "flexible"` where $k - 1$ threshold (where $k$ is the number of ordinal levels for $Y$) are estimated.
@@
 We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
-Table `r if(which_output("latex")) "\\@ref(tab:tab-dataset-example-latex)" else "\\@ref(tab:tab-dataset-example-docx)"` contains simulated data from $n = 100$ participants rating the agreement about a certain item with $k = 4$ ordered options.
-The participants are divided into two groups ($x_a$ and $x_b$).
-We can fit a cumulative link model with `clm()` function and check the model summary.
+Table `r if(which_output("latex")) "\\@ref(tab:tab-dataset-example-latex)" else "\\@ref(tab:tab-dataset-example-docx)"` contains simulated data from $n = 5000$ large $n$ for having stable estimations in the examples) participants rating the agreement about an item with $k = 4$ ordered options.
+The participants are divided into two groups ($X_a$ and $X_b$).
+We can fit a cumulative link model (`clm()` function)  and check the model summary.
 
@@
 k <- 4
-n <- 100
+n <- 5e3
 x <- rep(c("a", "b"), each = n/2)
@@
 Odds and odds ratios are clearly defined with $k = 2$ outcomes.
-With an ordinal variable and a cumulative model we can use the cumulative odds ratio.
-With e.g.
-$k = 4$ outcomes we have $k - 1$ odds ratios determined by the cumulative probability in terms of $P(Y \leq 1), \dots, P(Y \leq k - 1)$ comparing the two groups.
+With an ordinal variable we have $k - 1$ odds ratios determined by the cumulative probability in terms of $P(Y \leq 1), \dots, P(Y \leq k - 1)$ comparing the two groups.
 
 ```{r odds-example}
-#| echo: true
-#| collapse: true
 #| include: false
@@
 
-# function to calculate odds
+# the dummy_ord() is a custom function creating k - 1 dummy variable for the cumulative probabilities
+
 odds <- function(p) p / (1 - p)
-
-# the dummy_ord() is a custom function creating k - 1 dummy variable for the cumulative probabilities
 
@@
 # calculating k - 1 odds ratios on the cumulative probabilities as a/b
-odds(group_a) / odds(group_b)
+(ors <- odds(group_a) / odds(group_b))
+log(ors) # logarithm of the odds
 ```
@@
 
-Following again the taxonomy by @Tutz2022-dg, each of the presented ordinal regression model has a basic version making the PO assumption.
-There are more advanced versions of the model relaxing this assumption completely (*non proportional odds*) and partially [*partial proportional odds*, @Peterson1990-cz].
-
-When fitting a logit model, the coefficients $\beta_j$ are the (log) odds ratios for a unit increase in $x$.
-However, the previous model estimate a single $\beta$ even if we have four ordinal outcomes.
-The CM model (in the basic version) made an assumption called *proportional odds* (PO).
-Under this assumption we assume that the $k - 1$ cumulative odds ratios are the same.
+The ORs calculated above are very similar to the parameter estimated from the model (thanks to the large $n$).
+The CM estimate essentially the (cumulative) OR as a function of predictors.
+Furthermore, the $k - 1$ ORs are very similar because data are generated under the so-called proportional odds (PO) assumption.
+The basic version of the CM model fitted above assume the PO estimating a single $\beta$ (instead of $k - 1$).
 The PO assumption is formalized in Equation \@ref(eq:prop-odds).
-The cumulative log odds ratio comparing $P(Y_k|x_0)$ with $P(Y_k|x_1)$ is the same regardless the specific threshold ($\beta_1 = \beta_2 \dots = \beta_{k - 1}$).
+The cumulative log odds ratio comparing $P(Y_k|x_0)$ with $P(Y_k|x_1)$ ($x_0$ and $x_1$ being two levels of a predictor) is the same regardless the specific threshold ($\beta_1 = \beta_2 \dots = \beta_{k - 1}$).
 The Figure (\@ref(fig:fig-prop-odds)) depicts the proportional odds assumption for the $k - 1$ logistic curves both for probabilities and linear predictors $\eta$.
@@
 There are CM models relaxing this assumption completely [*non proportional odds*, @Tutz2022-dg] or partially [*partial proportional odds*, @Peterson1990-cz].
-The PO model is more parsimonious but the PO assumption could be considered too strict.
-There are several methods for testing if data are supporting the PO assumption [see @Liu2023-bp for an overview].
+The advantage of the PO model is the parsimony.
+In some scenarios, the PO assumption can be considered too strict.
+There are several methods for testing if data are supporting the PO [see @Liu2023-bp for an overview].
 @Tutz2020-xq suggested a trade-off between assuming/relaxing the PO assumption by fitting *location-shift* or *location-scale* models.
@@
 
-In the previous code block, we estimated the $k - 1$ odds ratios and they were different even if data have been simulated under the PO assumption (see the Simulating data section).
-Calculating the odds ratios using predicted probabilities from the model fitted with `clm()` (that by default assume the PO) the odds ratios are the same.
+The PO can also be clearly seen calculating the ORs using predicted probabilities.
+By default, the model fitted with `clm()` assume the PO.
 In the following code block we computed the odds ratio comparing $x_a$ and $x_b$ when $k \leq 1$ and $k \leq 2$.
@@
 
-When assuming a standard Gaussian distribution we are fitting *probit* model.
+The previous model was fitted using a *logit* link function.
+When assuming a standard normal distribution we are fitting *probit* model.
 The main difference regards the parameters interpretation.
 In the logit model the $\boldsymbol{\beta}$ are the log odds ratio.
-For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in $x$.
-
-For the *probit* model, the $\beta$ is the increase in terms of $z$ scores for a unit increase in $x$.
+For categorical variables they represents the increase in the log odds of moving from one category to another while for numerical variables is the increase in the log odds for a unit increase in $x$.
+
+In the *probit* model, the $\beta$ is the increase in terms of $z$ scores for a unit increase in $x$.
 This is very convenient especially for categorical variables because parameters can be interpreted as a Cohen's $d$ like measure.
-Thinking about the latent distributions, the $\beta$ is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric $x$.
+In terms of latent distributions, the $\beta$ is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric $x$.
 More formally the shift in the latent distribution is $\beta/\sigma$ (for the *probit* model $\sigma = 1$).
@@
 
-The PO assumption is relevant only for the *logit* model.
-When using the *probit* link function we can use the terms *parallel slopes* for the assumption of a common $\beta$ (see Equation \@ref(eq:probit-prop-odds)).
-In practical terms, difference in $z$ scores for a unit increase in $x$ (the actual interpretation of $\beta$ in a *probit* model) is the same regardless of the threshold.
-As for the previous example, we can fit the *probit* model and manually calculate the difference in $z$ scores.
+More generally, the PO assumption can be called *parallel slopes*.
+When applying the link function, cumulative probabilities have no longer a sigmoid shape but linear (see Figure \@ref(fig:fig-prop-odds)).
+If the steepness of the sigmoid (i.e., $\beta_j$) is not constant the $k - 1$ lines have different slopes.
+The PO assumption is relevant only for the *logit* model because parameters are ORs.
+
+For the *probit* model, the difference in $z$ scores for a unit increase in $x$ is the same regardless of the threshold (see Equation \@ref(eq:probit-prop-odds).
+As for the previous example, we can fit the CM and manually calculate the difference in $z$ scores (from predicted probabilities).
 
@@
 Whenever random number generation occurs, it is appropriate to set a `seed` (`set.seed()` function, in this case we use `set.seed(2024)`).
-Running again the same code will produce the same result.
+Running again the same code with the same seed will produce the same result.
 A general simulation approach concerns generating a dataset from the assumed data generation process, fitting the statistical model and assessing the recovery of simulated parameters.
@@
 
-For the first method we need to calculate $g^{-1}(\eta)$ as a function of predictors and then sample from a *multinomial* (more specifically *categorical*) distribution using the `sample()` function in R.
-This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview).
-
-As a simple example, let's simulate two groups with $n = 100$ participants responding to an item ($Y$) with $k = 4$ ordered options.
-We simulate the second group having higher probability of responding higher categories of $Y$.
-
-As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single $\beta = \log(OR)$.
-We can assume that the odds ratio is $OR = 2$.
+For the first method we need to calculate $P(Y \leq k) = g^{-1}(\eta)$ as a function of predictors and then sample from a *multinomial* (more specifically *categorical*) distribution using the `sample()` function in R.
+This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials).
+
+As a simple example, we simulate two groups with $n = 100$ participants responding to an item ($Y$) with $k = 4$ ordered options.
+As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single $\beta = \log(OR) = \log(3)$.
 For simplicity, the probabilities of the first group $x = 0$ are uniform thus $P(Y = 1|x_0) = \dots P(Y = k|x_0) = 1/k$.
 The following code summarise the first steps of the simulation.
-Basically we define the simulation parameters, we calculate the $k - 1$ thresholds $\alpha$ and we apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2).
-In this way we calculated the true probability of each $Y$ level for the two groups.
-
-An important step is converting from $\alpha$ to probabilities or the opposite.
-This steps are implemented in the `alpha_to_prob()` and `prob_to_alpha()` functions that given the input return thresholds or probabilities.
+Basically we define the simulation parameters, we calculate the $k - 1$ thresholds $\alpha$ and we apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2) calculating the probability of each response.
+
+The other parameters to define are the thresholds.
+We can consider them as intercepts thus the transformed probabilities of each ordinal response when $x = 0$ (in this case the first group).
+The `alpha_to_prob()` and `prob_to_alpha()` can be used to go back and forth from thresholds and probabilities.
 
@@
 N <- 100 # sample size
-or <- 2 # odds ratio
+or <- 3 # odds ratio
 k <- 4 # number of ordinal alternatives
@@
 A more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
-This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and the cutting the latent values according to the thresholds $\alpha$.
+This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and cutting the latent variable using the thresholds.
 The workflow is slightly different compared to the previous approach.
@@
 The crucial part is setting appropriate and empirically meaningful parameters.
-In the PO model thresholds are usually not of main interest as intercepts in standard linear regression.
-Thresholds are essentially quantiles of the latent distribution that produced certain $k$ probabilities.
+Thresholds are the quantiles of the latent distribution that produced certain $k$ probabilities when $\mathbf{X} = 0$.
 To set meaningful $\alpha$ values we can convert probabilities into thresholds (`alpha_to_prob()`).

### 2024-01-29 paper/paper.tex
reducing words

@@
 \abstract{%
-Ordinal data are very common in psychology such as Likert items, ratings or generic ordinal variables.
-Despite the widespread these variables are usually analyzed using metric models (e.g., standard linear regression).
-This approach can have important drawbacks in terms of statistical inference (e.g., reduced power) and prediction.
+Ordinal data are very common in psychology such as Likert items, ratings or generic ordered variables.
+Despite the widespread, these variables are usually analyzed using metric models (e.g., standard linear regression).
+This approach can have important drawbacks in terms of statistical inference (e.g., reduced power and increased type-1 error) and prediction.
 One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis.
@@
 Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer.
-Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools (Agresti, 2010; e.g., N.
-Cliff, 1996).
+Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools (Agresti, 2010; e.g., Cliff, 1996).
 This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data.
 Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018).
-More generally, Norman Cliff (2016) suggests that several research questions in behavioral sciences can be considered as ordinal (\emph{is the score} \(x\) higher than the score \(y\)?) concerning variables where the most appropriate measurement scale is probably ordinal.
-
-In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a linear regression that makes an explicit assumption about metric features of the response variable.
+
+In Psychology especially with item-based measures (questionnaires, surveys, etc.) the common practice is running a linear regression that makes an explicit assumption about metric features of the response variable.
 Liddell and Kruschke (2018) reviewed the psychological literature using Likert-based measures and reported how the majority of papers used metric-based statistical models.
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(dplyr)}
+\FunctionTok{library}\NormalTok{(tidyr)}
+\FunctionTok{library}\NormalTok{(ordinal)}
+\FunctionTok{library}\NormalTok{(ggplot2)}
+
+\CommentTok{\# custom functions}
+
+\CommentTok{\# show first and last rows of a dataframe}
+\NormalTok{head\_tail }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, }\AttributeTok{n =} \DecValTok{5}\NormalTok{)\{}
+  \FunctionTok{rbind}\NormalTok{(}
+    \FunctionTok{head}\NormalTok{(x, }\AttributeTok{n =}\NormalTok{ n),}
+    \FunctionTok{tail}\NormalTok{(x, }\AttributeTok{n =}\NormalTok{ n)}
+\NormalTok{  )}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
 \normalsize
@@
 
-Despite the specific model proposed by Liddell and Kruschke (2018) proposed a specific modeling approach, there is a class of regression models that consider the ordinal nature of the response variable.
-The actual statistical nomenclature of \emph{ordinal regression} models can be confusing (Tutz, 2022).
-Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
-
-We can identify three main types: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}.
-Among these, the cumulative model is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
-The \emph{sequential model} is appropriate when modeling sequential processes.
-Assuming to have five response options, the sequential model assume that responding ``3'' assume a sequential process where steps ``1'' and ``2'' are already reached.
-A clear example is proposed by Bürkner and Vuorre (2019) where the marriage duration in years is predicted as a function of some explanatory variables.
-For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential model.
-When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Tutz \& Berger, 2020).
-
-The adjacent category model compare the category \(k\) with \(k + 1\) still assuming a latent distribution for each \(k\).
-As suggested by Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories.
-Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical interpretation as for the cumulative vs sequential model.
-
-In the current paper we put the focus on the CM for several reasons.
+We can generally define as \emph{ordinal regression}, a statistical model explicitly considering the ordinal (and not metric) nature of the response variable.
+The nomenclature of \emph{ordinal regression} models can be confusing but Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.
+
+We can identify three main models: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}.
+Among these, the CM is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
+The \emph{sequential model} is appropriate when modeling sequential processes where responding to a certain category implies have already reached lower categories.
+The \emph{adjacent category model} compares a category with the successive one and can be seen the adjacent-category model can be seen as a series of binomial regressions taking into account the ordering (Tutz, 2022).
+
+In the current paper we focus on the CM for several reasons.
 Firstly, the latent formulation of the model is particularly convenient for parameter interpretation and data simulation.
-The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item.
+The second reason is that several psychological construct can be formalized as a latent continuous variable observed as an ordinal item(s).
 Furthermore, the signal detection theory framework that is very common in experimental psychology can be implemented using a CM (e.g., DeCarlo, 2010).
@@
 The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\).
-Similarly to the generalized linear models framework (e.g., Fox, 2015), we define \(g(x) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\).
-To transform back \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\).
-The specific link function define the type of model and require a different R function.
-For example, when assuming a Gaussian distribution (\emph{probit} link function) entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).
-
-When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(\mathbf{X}\) value \(P(Y \leq k | \mathbf{X}_i)\)}.
-Equation \eqref{eq:prob-cum-model1} shows the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\).
-The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta_j\) as in the standard regression models where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories (Agresti, 2010).
+Similarly to the generalized linear models framework (e.g., Fox, 2015), we define \(g(p) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\).
+To transform back \(\eta\) into probabilities we use the inverse of the link function \(p = g^{-1}(\eta)\).
+The specific link function define the type of model and require different R functions.
+For example, using a Normal distribution (\emph{probit} link function) requires using the cumulative distribution function \(g(p) = \Phi^{-1}(p) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(p = g^{-1}(\eta) = \Phi(\eta)\).
+
+When modelling an ordinal variable in a CM we include predictors on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(\mathbf{X}\) value \(P(Y \leq k | \mathbf{X}_i)\)}.
+Equation \eqref{eq:prob-cum-model1} shows the general cumulative model with predictors \(\mathbf{X}\) and coefficients \(\boldsymbol{\beta}\).
+The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta_j\) as in the standard regression models where higher positive values corresponds to increased probability of responding higher \(k\) categories (Agresti, 2010).
 
@@
 The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g(\cdot)\).
-To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:prob-cum-model2}.
+The probability of a single outcome \(P(Y = k)\) can be calculated as the difference between cumulative probabilities (see Equation \eqref{eq:prob-cum-model2}).
 
@@
 There are always two special cases when computing the probability of a single outcome \(Y\) that is when \(Y = 1\) and \(Y = k\).
-In the first case the cumulative probability is calculated from \(-\infty\) that is the same as temporary assuming an \(\alpha_0 = -\infty\).
-In the second case (\(Y = 1\)) the probability is calculated as \(P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)\) that is the same as assuming a temporary threshold \(\alpha_k = +\infty\).
+In the first case the cumulative probability is calculated from \(-\infty\) (assuming that \(\alpha_0 = -\infty\)).
+In the second case (\(Y = 1\)) the probability is calculated as \(P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)\) (a \(\alpha_k = +\infty\).
 The Figure \ref{fig:fig-explain-cumulative} shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.
@@
 \begin{equation}
-\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \mathbf{\epsilon}
+\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
 \label{eq:latent-model}
@@
 
-The crucial part is \(\epsilon\) that is the random part (errors) of the model.
-For a \emph{probit} model, errors are sampled from a standard Normal distribution while for a \emph{logit} model from a standard logistic distribution.
+The \(\epsilon\) term is the random part (errors) of the model.
+For a \emph{probit} model, errors are distributed as a standard Normal distribution and for a \emph{logit} model as a standard logistic distribution.
 Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).
 
-The thresholds \(\alpha_k\) are considered fixed and part of the measurement procedure (Liddell \& Kruschke, 2018; see the \emph{location-shift} models by Tutz, 2022, where thresholds vary as a function of predictors).
-The model can be also formalized in alternative ways.
-Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model or Gelman, Hill, and Vehtari (2020) for other thresholds parametrizations.
-
-The scale (spread) of the latent variable is fixed to one for both \emph{probit} and \emph{logit} models.
-In the supplementary materials we showed how to implement a location-scale models including predictors on the scale parameter (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022).
+In the CM, thresholds \(\alpha_k\) are considered fixed and part of the measurement procedure (Liddell \& Kruschke, 2018; but see the \emph{location-shift} models by Tutz, 2022, where thresholds vary as a function of predictors) and can be considered as intercepts.
+The proposed CM can be parametrized in different ways.
+Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model and Gelman, Hill, and Vehtari (2020) proposed other thresholds parametrizations.
 
@@
 Among several available functions the \emph{logit} and \emph{probit} models are the most common.
-The \emph{logit} model use a \emph{logit} link function and a logistic distribution as latent variable.
-On the other side, the \emph{probit} model assume a latent Gaussian distribution using the inverse of the cumulative distribution (\(\Phi^{-1}\)) function as link function.
+The \emph{logit} model use a \emph{logit} link function and a standard logistic distribution as latent variable.
+On the other side, the \emph{probit} model assume a standard Normal distribution using the inverse of the cumulative distribution (\(\Phi^{-1}(\cdot)\)) function as link function.
 
@@
 In practical terms, fixing \(\mu\) and \(s\) lead to an higher standard deviation for the logistic distribution.
+In the supplementary materials we showed how to implement a location-scale models for including predictors on the scale parameter (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022).
 
@@
 
-For fitting the cumulative models we used the \texttt{ordinal} package (Christensen, 2019).
-Despite the presence of other possibilities the \texttt{ordinal} package provide the most complete and intuitive way to implement the ordinal models.
-The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}
+For fitting the CM we used the \texttt{ordinal} package (Christensen, 2019).
+The \texttt{ordinal} package provide a complete and intuitive way to implement the ordinal models.
+The syntax is very similar to standard linear models in R with functions to calculate predictions, perform model comparison and extract relevant model information.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}
 
@@
 When fitting the model the crucial arguments are the \texttt{formula}, the \texttt{link} function and the \texttt{data}.
-More advanced arguments are the \texttt{nominal}, \texttt{scale} and \texttt{threshold}.
+More advanced arguments are \texttt{nominal}, \texttt{scale} and \texttt{threshold}.
 
@@
 \item
-  \texttt{link}: is the link function. In this tutorial we consider only the \emph{logit} and \emph{probit} link but other link functions are available.
+  \texttt{link}: is the link function (in this tutorial only \emph{logit} and \emph{probit})
 \item
@@
 \item
-  \texttt{scale}: formula with predictors for the scale (standard deviation) parameter.
-This argument allow to fit a scale-location model (see supplementary materials).
-The main \texttt{formula} argument refers to predictors on the location parameter (i.e., the mean \(\mu\)).
+  \texttt{scale}: formula with predictors for the scale parameter for fitting a scale-location model
 \item
@@
 We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections.
-Table \ref{tab:tab-dataset-example-latex} contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options.
-The participants are divided into two groups (\(x_a\) and \(x_b\)).
-We can fit a cumulative link model with \texttt{clm()} function and check the model summary.
+Table \ref{tab:tab-dataset-example-latex} contains simulated data from \(n = 5000\) large \(n\) for having stable estimations in the examples) participants rating the agreement about an item with \(k = 4\) ordered options.
+The participants are divided into two groups (\(X_a\) and \(X_b\)).
+We can fit a cumulative link model (\texttt{clm()} function) and check the model summary.
 
@@
 \midrule
-a & 2.62 & 3 & 1.1 & \textbf{10} (p = \textit{0.2}) & \textbf{13} (p = \textit{0.26}) & \textbf{13} (p = \textit{0.26}) & \textbf{14} (p = \textit{0.28})\\
-b & 3.06 & 3 & 1.04 & \textbf{6} (p = \textit{0.12}) & \textbf{7} (p = \textit{0.14}) & \textbf{15} (p = \textit{0.3}) & \textbf{22} (p = \textit{0.44})\\
+a & 2.5 & 2 & 1.1 & \textbf{599} (p = \textit{11.98}) & \textbf{662} (p = \textit{13.24}) & \textbf{627} (p = \textit{12.54}) & \textbf{612} (p = \textit{12.24})\\
+b & 3.17 & 3.5 & 0.99 & \textbf{221} (p = \textit{4.42}) & \textbf{381} (p = \textit{7.62}) & \textbf{648} (p = \textit{12.96}) & \textbf{1250} (p = \textit{25})\\
 \bottomrule
@@
 \CommentTok{\#\textgreater{}  link  threshold nobs logLik   AIC      niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  logit flexible  100  {-}131.80 271.60 4(0)  2.13e{-}11 1.7e+01}
+\CommentTok{\#\textgreater{}  logit flexible  5000 {-}6458.45 12924.89 4(0)  1.59e{-}08 1.7e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} xb   0.7556     0.3685    2.05   0.0403 *}
+\CommentTok{\#\textgreater{} xb  1.14864    0.05333   21.54   \textless{}2e{-}16 ***}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}1.3328     0.3126  {-}4.264}
-\CommentTok{\#\textgreater{} 2|3  {-}0.2185     0.2703  {-}0.808}
-\CommentTok{\#\textgreater{} 3|4   0.9767     0.2893   3.377}
+\CommentTok{\#\textgreater{} 1|2 {-}1.16300    0.04345 {-}26.765}
+\CommentTok{\#\textgreater{} 2|3  0.01284    0.03835   0.335}
+\CommentTok{\#\textgreater{} 3|4  1.14143    0.04194  27.214}
 \end{Highlighting}
@@
 Odds and odds ratios are clearly defined with \(k = 2\) outcomes.
-With an ordinal variable and a cumulative model we can use the cumulative odds ratio.
-With e.g.~\(k = 4\) outcomes we have \(k - 1\) odds ratios determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\) comparing the two groups.
-
-For example the probability of responding \(Y \leq 1\) in the group ``a'' is 0.20 corresponding to an odds of 0.25.
-When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y \leq 1\) we obtain that the group ``a'' has 1.83 times the odds of responding \(Y \leq 1\) compared to the group ``b''.
+With an ordinal variable we have \(k - 1\) odds ratios determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\) comparing the two groups.
+
+For example the probability of responding \(Y \leq 1\) in the group ``a'' is 0.24 corresponding to an odds of 0.32.
+When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y \leq 1\) we obtain that the group ``a'' has 3.25 times the odds of responding \(Y \leq 1\) compared to the group ``b''.
 
@@
 \begin{Highlighting}[]
-\CommentTok{\# function to calculate odds}
+\CommentTok{\# the dummy\_ord() is a custom function creating k {-} 1 dummy variable for the cumulative probabilities}
+
 \NormalTok{odds }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(p) p }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ p)}
-
-\CommentTok{\# the dummy\_ord() is a custom function creating k {-} 1 dummy variable for the cumulative probabilities}
 
@@
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.20    0.46    0.72}
+\CommentTok{\#\textgreater{}  0.2396  0.5044  0.7552}
 \NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
 \CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
-\CommentTok{\#\textgreater{}    0.12    0.26    0.56}
+\CommentTok{\#\textgreater{}  0.0884  0.2408  0.5000}
 
 \CommentTok{\# calculating k {-} 1 odds ratios on the cumulative probabilities as a/b}
-\FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b)}
+\NormalTok{(ors }\OtherTok{\textless{}{-}} \FunctionTok{odds}\NormalTok{(group\_a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(group\_b))}
 \CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
-\CommentTok{\#\textgreater{} 1.833333 2.424501 2.020408}
+\CommentTok{\#\textgreater{} 3.249352 3.208806 3.084967}
+\FunctionTok{log}\NormalTok{(ors) }\CommentTok{\# logarithm of the odds}
+\CommentTok{\#\textgreater{}  y1vs234  y12vs34  y123vs4 }
+\CommentTok{\#\textgreater{} 1.178456 1.165899 1.126541}
 \end{Highlighting}
@@
 
-Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the PO assumption.
-There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}, Peterson \& Harrell, 1990).
-
-When fitting a logit model, the coefficients \(\beta_j\) are the (log) odds ratios for a unit increase in \(x\).
-However, the previous model estimate a single \(\beta\) even if we have four ordinal outcomes.
-The CM model (in the basic version) made an assumption called \emph{proportional odds} (PO).
-Under this assumption we assume that the \(k - 1\) cumulative odds ratios are the same.
+The ORs calculated above are very similar to the parameter estimated from the model (thanks to the large \(n\)).
+The CM estimate essentially the (cumulative) OR as a function of predictors.
+Furthermore, the \(k - 1\) ORs are very similar because data are generated under the so-called proportional odds (PO) assumption.
+The basic version of the CM model fitted above assume the PO estimating a single \(\beta\) (instead of \(k - 1\)).
 The PO assumption is formalized in Equation \eqref{eq:prop-odds}.
-The cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) is the same regardless the specific threshold (\(\beta_1 = \beta_2 \dots = \beta_{k - 1}\)).
+The cumulative log odds ratio comparing \(P(Y_k|x_0)\) with \(P(Y_k|x_1)\) (\(x_0\) and \(x_1\) being two levels of a predictor) is the same regardless the specific threshold (\(\beta_1 = \beta_2 \dots = \beta_{k - 1}\)).
 The Figure (\ref{fig:fig-prop-odds}) depicts the proportional odds assumption for the \(k - 1\) logistic curves both for probabilities and linear predictors \(\eta\).
@@
 There are CM models relaxing this assumption completely (\emph{non proportional odds}, Tutz, 2022) or partially (\emph{partial proportional odds}, Peterson \& Harrell, 1990).
-The PO model is more parsimonious but the PO assumption could be considered too strict.
-There are several methods for testing if data are supporting the PO assumption (see Liu, He, Tu, \& Tang, 2023 for an overview).
+The advantage of the PO model is the parsimony.
+In some scenarios, the PO assumption can be considered too strict.
+There are several methods for testing if data are supporting the PO (see Liu, He, Tu, \& Tang, 2023 for an overview).
 Tutz and Berger (2020) suggested a trade-off between assuming/relaxing the PO assumption by fitting \emph{location-shift} or \emph{location-scale} models.
@@
 
-In the previous code block, we estimated the \(k - 1\) odds ratios and they were different even if data have been simulated under the PO assumption (see the Simulating data section).
-Calculating the odds ratios using predicted probabilities from the model fitted with \texttt{clm()} (that by default assume the PO) the odds ratios are the same.
+The PO can also be clearly seen calculating the ORs using predicted probabilities.
+By default, the model fitted with \texttt{clm()} assume the PO.
 In the following code block we computed the odds ratio comparing \(x_a\) and \(x_b\) when \(k \leq 1\) and \(k \leq 2\).
@@
 \FunctionTok{odds}\NormalTok{(y1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 2.128857}
+\CommentTok{\#\textgreater{} [1] 3.153894}
 \FunctionTok{odds}\NormalTok{(y12a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 2.128857}
+\CommentTok{\#\textgreater{} [1] 3.153894}
 \end{Highlighting}
@@
 
-When assuming a standard Gaussian distribution we are fitting \emph{probit} model.
+The previous model was fitted using a \emph{logit} link function.
+When assuming a standard normal distribution we are fitting \emph{probit} model.
 The main difference regards the parameters interpretation.
 In the logit model the \(\boldsymbol{\beta}\) are the log odds ratio.
-For categorical variables they represents the increase in the log odds of moving from one level to the other while for numerical variables is the increase in the log odds for a unit increase in \(x\).
-
-For the \emph{probit} model, the \(\beta\) is the increase in terms of \(z\) scores for a unit increase in \(x\).
+For categorical variables they represents the increase in the log odds of moving from one category to another while for numerical variables is the increase in the log odds for a unit increase in \(x\).
+
+In the \emph{probit} model, the \(\beta\) is the increase in terms of \(z\) scores for a unit increase in \(x\).
 This is very convenient especially for categorical variables because parameters can be interpreted as a Cohen's \(d\) like measure.
-Thinking about the latent distributions, the \(\beta\) is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric \(x\).
+In terms of latent distributions, the \(\beta\) is the shift in the latent mean comparing two or more groups or the slope of latent scores as a function of a numeric \(x\).
 More formally the shift in the latent distribution is \(\beta/\sigma\) (for the \emph{probit} model \(\sigma = 1\)).
@@
 
-The PO assumption is relevant only for the \emph{logit} model.
-When using the \emph{probit} link function we can use the terms \emph{parallel slopes} for the assumption of a common \(\beta\) (see Equation \eqref{eq:probit-prop-odds}).
-In practical terms, difference in \(z\) scores for a unit increase in \(x\) (the actual interpretation of \(\beta\) in a \emph{probit} model) is the same regardless of the threshold.
-As for the previous example, we can fit the \emph{probit} model and manually calculate the difference in \(z\) scores.
+More generally, the PO assumption can be called \emph{parallel slopes}.
+When applying the link function, cumulative probabilities have no longer a sigmoid shape but linear (see Figure \ref{fig:fig-prop-odds}).
+If the steepness of the sigmoid (i.e., \(\beta_j\)) is not constant the \(k - 1\) lines have different slopes.
+The PO assumption is relevant only for the \emph{logit} model because parameters are ORs.
+
+For the \emph{probit} model, the difference in \(z\) scores for a unit increase in \(x\) is the same regardless of the threshold (see Equation \eqref{eq:probit-prop-odds}.
+As for the previous example, we can fit the CM and manually calculate the difference in \(z\) scores (from predicted probabilities).
 
@@
 \FunctionTok{qnorm}\NormalTok{(y1a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y1b)}
-\CommentTok{\#\textgreater{} [1] 0.4408949}
+\CommentTok{\#\textgreater{} [1] 0.6869165}
 \FunctionTok{qnorm}\NormalTok{(y12a) }\SpecialCharTok{{-}} \FunctionTok{qnorm}\NormalTok{(y12b)}
-\CommentTok{\#\textgreater{} [1] 0.4408949}
+\CommentTok{\#\textgreater{} [1] 0.6869165}
 \end{Highlighting}
@@
 Whenever random number generation occurs, it is appropriate to set a \texttt{seed} (\texttt{set.seed()} function, in this case we use \texttt{set.seed(2024)}).
-Running again the same code will produce the same result.
+Running again the same code with the same seed will produce the same result.
 A general simulation approach concerns generating a dataset from the assumed data generation process, fitting the statistical model and assessing the recovery of simulated parameters.
@@
 
-For the first method we need to calculate \(g^{-1}(\eta)\) as a function of predictors and then sample from a \emph{multinomial} (more specifically \emph{categorical}) distribution using the \texttt{sample()} function in R.
-This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials for a general overview).
-
-As a simple example, let's simulate two groups with \(n = 100\) participants responding to an item (\(Y\)) with \(k = 4\) ordered options.
-We simulate the second group having higher probability of responding higher categories of \(Y\).
-
-As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single \(\beta = \log(OR)\).
-We can assume that the odds ratio is \(OR = 2\).
+For the first method we need to calculate \(P(Y \leq k) = g^{-1}(\eta)\) as a function of predictors and then sample from a \emph{multinomial} (more specifically \emph{categorical}) distribution using the \texttt{sample()} function in R.
+This method is similar to the general way of simulating data for a generalized linear model (see the supplementary materials).
+
+As a simple example, we simulate two groups with \(n = 100\) participants responding to an item (\(Y\)) with \(k = 4\) ordered options.
+As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single \(\beta = \log(OR) = \log(3)\).
 For simplicity, the probabilities of the first group \(x = 0\) are uniform thus \(P(Y = 1|x_0) = \dots P(Y = k|x_0) = 1/k\).
 The following code summarise the first steps of the simulation.
-Basically we define the simulation parameters, we calculate the \(k - 1\) thresholds \(\alpha\) and we apply Equations \eqref{eq:prob-cum-model1} and \eqref{eq:prob-cum-model2}.
-In this way we calculated the true probability of each \(Y\) level for the two groups.
-
-An important step is converting from \(\alpha\) to probabilities or the opposite.
-This steps are implemented in the \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} functions that given the input return thresholds or probabilities.
+Basically we define the simulation parameters, we calculate the \(k - 1\) thresholds \(\alpha\) and we apply Equations \eqref{eq:prob-cum-model1} and \eqref{eq:prob-cum-model2} calculating the probability of each response.
+
+The other parameters to define are the thresholds.
+We can consider them as intercepts thus the transformed probabilities of each ordinal response when \(x = 0\) (in this case the first group).
+The \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} can be used to go back and forth from thresholds and probabilities.
 
@@
 \NormalTok{N }\OtherTok{\textless{}{-}} \DecValTok{100} \CommentTok{\# sample size}
-\NormalTok{or }\OtherTok{\textless{}{-}} \DecValTok{2} \CommentTok{\# odds ratio}
+\NormalTok{or }\OtherTok{\textless{}{-}} \DecValTok{3} \CommentTok{\# odds ratio}
 \NormalTok{k }\OtherTok{\textless{}{-}} \DecValTok{4} \CommentTok{\# number of ordinal alternatives}
@@
 \CommentTok{\#\textgreater{}      p1   p2   p3   p4}
-\CommentTok{\#\textgreater{} 1  0.2500000 0.2500000 0.2500000 0.25}
-\CommentTok{\#\textgreater{} 51 0.1428571 0.1904762 0.2666667 0.40}
+\CommentTok{\#\textgreater{} 1  0.25 0.25 0.25 0.25}
+\CommentTok{\#\textgreater{} 51 0.10 0.15 0.25 0.50}
 
@@
 \CommentTok{\#\textgreater{}   cump\_leq1 cump\_leq2 cump\_leq3}
-\CommentTok{\#\textgreater{} 1         2         2         2}
+\CommentTok{\#\textgreater{} 1         3         3         3}
 \end{Highlighting}
@@
 \CommentTok{\#\textgreater{}   0 0.26 0.22 0.26 0.26}
-\CommentTok{\#\textgreater{}   1 0.10 0.18 0.30 0.42}
+\CommentTok{\#\textgreater{}   1 0.06 0.08 0.30 0.56}
 
@@
 \CommentTok{\#\textgreater{}        0    1}
-\CommentTok{\#\textgreater{}   1 0.26 0.10}
-\CommentTok{\#\textgreater{}   2 0.48 0.28}
-\CommentTok{\#\textgreater{}   3 0.74 0.58}
+\CommentTok{\#\textgreater{}   1 0.26 0.06}
+\CommentTok{\#\textgreater{}   2 0.48 0.14}
+\CommentTok{\#\textgreater{}   3 0.74 0.44}
 \CommentTok{\#\textgreater{}   4 1.00 1.00}
@@
 \CommentTok{\#\textgreater{}        1        2        3 }
-\CommentTok{\#\textgreater{} 3.162162 2.373626 2.061008}
+\CommentTok{\#\textgreater{} 5.504505 5.670330 3.622378}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}132.59 273.17 5(0)  2.86e{-}09 1.7e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}122.46 252.91 6(0)  2.05e{-}10 1.6e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x   0.5183     0.2197   2.359   0.0183 *}
+\CommentTok{\#\textgreater{} x   0.8823     0.2290   3.854 0.000116 ***}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2 {-}0.68708    0.17711  {-}3.879}
-\CommentTok{\#\textgreater{} 2|3 {-}0.05305    0.16717  {-}0.317}
-\CommentTok{\#\textgreater{} 3|4  0.68706    0.17462   3.935}
+\CommentTok{\#\textgreater{} 1|2  {-}0.6564     0.1788  {-}3.672}
+\CommentTok{\#\textgreater{} 2|3  {-}0.1026     0.1686  {-}0.608}
+\CommentTok{\#\textgreater{} 3|4   0.6974     0.1778   3.923}
 \end{Highlighting}
@@
 A more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
-This require simulating a standard linear regression using the appropriate data generation function (\emph{logistic} or \emph{normal}) and the cutting the latent values according to the thresholds \(\alpha\).
+This require simulating a standard linear regression using the appropriate data generation function (\emph{logistic} or \emph{normal}) and cutting the latent variable using the thresholds.
 The workflow is slightly different compared to the previous approach.
@@
 The crucial part is setting appropriate and empirically meaningful parameters.
-In the PO model thresholds are usually not of main interest as intercepts in standard linear regression.
-Thresholds are essentially quantiles of the latent distribution that produced certain \(k\) probabilities.
+Thresholds are the quantiles of the latent distribution that produced certain \(k\) probabilities when \(\mathbf{X} = 0\).
 To set meaningful \(\alpha\) values we can convert probabilities into thresholds (\texttt{alpha\_to\_prob()}).
@@
 
-\bibitem[\citeproctext]{ref-Cliff2016-ck}
-Cliff, Norman. (2016). \emph{Ordinal methods for behavioral data analysis}. London, England: Psychology Press.
-
 \bibitem[\citeproctext]{ref-Cox1995-ur}

### 2024-01-29 paper/paper.Rmd
testing website

@@
 The following code summarise the first steps of the simulation.
-Basically we define the simulation parameters, we calculate the $k - 1$ thresholds $\alpha$ and we apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2) calculating the probability of each response.
-
-The other parameters to define are the thresholds.
-We can consider them as intercepts thus the transformed probabilities of each ordinal response when $x = 0$ (in this case the first group).
+We define the simulation parameters, calculate the $k - 1$ thresholds $\alpha$ and apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2) calculating the probability of each response.
+
+To set the thresholds, we can consider them as intercepts thus the transformed probabilities of each ordinal response when $x = 0$ (in this case the first group).
 The `alpha_to_prob()` and `prob_to_alpha()` can be used to go back and forth from thresholds and probabilities.
@@
 lp <- data.frame(lp)
-head(lp)
+head_tail(lp, n = 3)
 
@@
 cump <- data.frame(cump)
-names(cump) <- sprintf("cump_leq%s", 1:(k - 1)) # giving appropriate names
-head(cump)
+
+# giving appropriate names, cump = cumulative probability, leq = less or equal
+names(cump) <- sprintf("cump_leq%s", 1:(k - 1))
+head_tail(cump, n = 3)
 
@@
 names(p) <- sprintf("p%s", 1:k) # giving appropriate names
-head(p)
 
@@
 
-# example of a random Y outcome based on the probabilities
-sample(x = 1:k, size = 1, prob = p[1, ])
+# example of n random Y outcome based on the probabilities
+sample(x = 1:k, size = 10, prob = p[1, ], replace = TRUE)
 
@@
 dat$y <- apply(p, 1, function(ps) sample(1:k, size = 1, prob = ps))
-head(dat)
+head_tail(dat, n = 3)
 
@@
 
-1.  Define simulation parameters, baseline probabilities (`prob0`), sample size, etc.
+1.  Define simulation parameters, intercept probabilities (`prob0`), sample size, etc.
 4.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the $k - 1$ linear predictors ($\eta$) using the equations
@@
 6.  Calculate the probabilities of $k$ outcomes
-7.  Sample $n$ outcomes from a *multinomial* distribution, choosing between $k$ alternatives using the calculated probabilities
+7.  Sample $n$ outcomes from a *multinomial* distribution using the calculated probabilities
 8.  Fit the appropriate model using `ordinal::clm()`
@@
 
-A more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
+More efficiently, we can use the latent formulation of the CM.
 This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and cutting the latent variable using the thresholds.
@@
 2.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the linear predictor $\eta$ using the Equation \@ref(eq:latent-model)
-3.  Add the random errors $\epsilon_i$ sampling from the chosen distribution (logistic or normal)
+3.  Add the random errors $\epsilon_i$ sampling from the logistic or normal distributions
 4.  Cut the latent variable into $k$ areas using the thresholds $\alpha$ and assign the corresponding ordinal value. This can be done using the `cut()` or the `findInterval()` functions.
@@
 probs0 <- rep(1/k, k) # probabilities for the first group
-alpha <- prob_to_alpha(probs0, link = "logit")
+alpha <- prob_to_alpha(probs0, link = "logit") # thresholds
 dat <- data.frame(x = rep(c(0, 1), each = N/2))
@@
 Then the model formula is specified within the function as `location = ~` along with the vector of regression coefficients (`beta`), baseline probabilities (`prob0`) and the link function.
-The function return a dataset with the simulated $Y$ and the latent variable (this is possible only because we simulate the data, otherwise $Y^{\star}$ cannot be observed by definition).
+The function return a dataset with the simulated $Y$ and the latent variable (this is possible only because we simulateds the data, otherwise $Y^{\star}$ cannot be observed by definition).
 
@@
 For *logit* models we can set the odds ratio.
-Meaningful odds ratios can be derived from previous literature, meta-analyses or converting to other effect sizes.
+Meaningful odds ratios can be derived from previous literature, meta-analyses or converting from other effect sizes.
 For example, @Sanchez-Meca2003-ji proposed some equations to convert between odds ratios and Cohen's $d$.
@@
 ```{r fig-example-cat-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
-cp <- cat_latent_plot(location = c(0, 1), scale = 1, prob0 = rep(1/4, 4), link = "logit", plot = FALSE) 
+cp <- cat_latent_plot(location = c(0, 1), scale = c(1, 1), prob0 = rep(1/4, 4), link = "logit", plot = FALSE)
 legend <- cowplot::get_legend(cp$probs_plot)
@@
 
-The parameters interpretation is the same as introduced in the general case.
-The thresholds $\alpha$ are fixed representing the probabilities $P(Y = k)$ when all predictors are zero.
+The thresholds are fixed, representing the probabilities $P(Y = k)$ when all predictors are zero.
 In this case by doing `alpha_to_prob(fit$alpha, link = "probit")` we should recover the `probs0` vector.
-`x1` is the main effect thus the difference in $z$ scores between *a* and *b* averaging over `x2`.
-The same holds for `x2`.
-`x1:x2` is the interaction thus difference of differences in $z$ scores (see Figure \@ref(fig:fig-effects-2-by-2-interaction)).
+`x11` is the main effect thus the difference in $z$ scores between *a* and *b* averaging over `x2`.
+The same holds for `x21`.
+`x11:x21` is the interaction thus difference of differences in $z$ scores (see Figure \@ref(fig:fig-effects-2-by-2-interaction)).
 
 ```{r fig-effects-2-by-2-interaction, fig.place = TRUE, fig.cap=filor::get_caption()}
-# plot(ggpredict(fit, terms = c("x1", "x2")))
+# plot(ggpredict(fit, terms = c ("x1", "x2")))
 ggpredict(fit, terms = c("x1", "x2")) |> 
@@
 
-Again, `x` is the slope of the numerical predictor averaging over `g` (given that `g` has been coded with sum-to-zero contrasts) thus the increase in $z$ scores for a unit increase in `x`.
+Again, `x` is the slope of the numerical predictor averaging over `g` (given that `g` has been coded with sum-to-zero contrast) thus the increase in $z$ scores for a unit increase in `x`.
 `g1` is the main effect of the factor evaluated when $x = 0$ (centering $x$ will change the parameter interpretation).
-The `x:g1` is the slopes difference the two groups.
+The `x:g1` is the difference between the slopes of the two groups.
 Figure \@ref(fig:fig-effects-num-by-cat-interaction) depicts the model results.
@@
 
+# Conclusions
+
+In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation.
+The basic model can be expanded in several ways.
+When assuming PO or considering only a location effect is limiting, including scale (*location-scale*), thresholds (*location-shift*) effects or relaxing the PO assumption can help in modeling complex phenomenon.
+The supplementary materials contains some examples of these complex models.
+Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.
+
 \newpage
@@
 
-(ref:fig-show-th-example) Example of the `show_alpha()` function. On the left the assumed latent variable and $k - 1$ thresholds with the corresponding probabilities on the right.
+(ref:fig-show-th-example) Example of the `show_alpha()` function.
+On the left the assumed latent variable and $k - 1$ thresholds with the corresponding probabilities on the right.
+The distance between thresholds (quantiles) determine the probability of the $k$ outcomes.
 
@@
 
-(ref:tab-model-summary-docx) Summary of the probit and logit models in terms of link ($g(\cdot)$) and inverse link function ($g^{-1}(\cdot)$) and the corresponding R code.
-
-(ref:tab-model-summary-latex) Summary of the probit and logit models in terms of link ($g(\cdot)$) and inverse link function ($g^{-1}(\cdot)$) and the corresponding R code.
+(ref:tab-model-summary-docx) Summary of the probit and logit models in terms of link $g(\cdot)$ and inverse link function $g^{-1}(\cdot)$ and the corresponding R code.
+
+(ref:tab-model-summary-latex) Summary of the probit and logit models in terms of link $g(\cdot)$ and inverse link function $g^{-1}(\cdot)$ and the corresponding R code.
 

### 2024-01-29 supplementary/mixed-effects-ordinal.qmd
testing website

---
title: "Random-effects Models"
format: 
  html:
    toc: true
bibliography: "`r filor::fil()$bib`"
csl: "`r filor::fil()$csl`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(cowplot)
library(ordinal)
devtools::load_all()

knitr::opts_chunk$set(echo = TRUE,
                      dev = "svg",
                      warning = FALSE,
                      message = FALSE)
mtheme <- function(){
  theme_minimal(15)
}
theme_set(mtheme())
```

We can extended the base CM model by including random-effects. For example, if the same participant responds to the same item/trial multiple times or responds to multiple items we need to take into account the nested data-structure. @Agresti2010-rz formalized the random-intercept CM model in @eq-cm-random-intercept.

$$
P(Y \leq k) = g^{-1}[(\alpha_k + u_i) - \mathbf{X}\boldsymbol{\beta}]
$$ {#eq-cm-random-intercept}

Where $u_i$ is the by-subject adjustment to the overall intercept-threshold. As in standard mixed-effects models, the random-effect of the intercept is sampled from a normal distribution with $\mu = 0$ and standard deviation $\sigma_{u}$, thus $u_i \sim \mathcal{N}(0, \sigma_u)$. We can use the same simulation strategies (sampling from multinomial distribution or latent approach) of the paper but introducing the random-effect. We simulate $N = 100$ subjects divided into two groups. Each subject performs $100$ trial responding on a ordinal item.

```{r}
# Simulation parameters
set.seed(2024)
k <- 4
probs0 <- rep(1/k, k)
N <- 100 # sample size
n <- N/2
nt <- 100 # number of trials
b1 <- log(3)
sb0 <- 0.5 # intercept standard deviation
group <- c(0, 1) # group
alpha <- prob_to_alpha(probs0, link = "logit")

# Data structure
dat0 <- dat1 <- expand.grid(id = 1:n, trial = 1:nt)
dat0$group <- 0
dat1$group <- 1
dat0$id <- rep(1:n, nt)
dat1$id <- rep((n + 1):N, nt)
dat <- rbind(dat0, dat1)

alphai <- rnorm(N, 0, sb0)
```

## Sampling from multinomial distribution

```{r}
set.seed(2024)
cump <- lapply(alpha, function(a) plogis(with(dat, (a + alphai[id]) - (b1 * group))))
p <- t(apply(cbind(0, data.frame(cump), 1), 1, diff))
names(p) <- paste0("py", 1:k)
dat <- cbind(dat, p)

dat$y <- apply(dat[, 4:ncol(dat)], 1, function(p) sample(1:k, 1, replace = TRUE, prob = p))
dat$y <- ordered(dat$y)

fit <- clmm(y ~ group + (1|id), data = dat, link = "logit")
summary(fit)
```

## Sampling from the latent variable

```{r}
set.seed(2024)
ystar <- with(dat, alphai[id] + (b1*group)) + rlogis(nrow(dat), 0, 1)
dat$y2 <- ordered(findInterval(ystar, alpha) + 1)
fit2 <- clmm(y2 ~ group + (1|id), data = dat, link = "logit")
summary(fit2)
```

In both simulations we are able to recover the parameters. For fitting the model we use the `clmm()` function that allows specifying the random-effects structure. The syntax is the same as the `lme4` package for standard mixed-effects models.

We can extend the simulation including also random slopes. The `group` effect need to be a within effect now, thus we can imagine to have two conditions in the experiment with 100 trials each. We can extend the @eq-cm-random-intercept with @eq-cm-random-slope. $\beta_{1_i}$ are the by-subjects adjustments to the overall $\beta_1$ effect, still sampled from a normal distribution.

$$
P(Y \leq k) = g^{-1}[(\alpha_k + u_i) - (\beta_1 + \beta_{1_i})X_1]
$${#eq-cm-random-slope}

Let's use the latent formulation directly:

```{r}
#| cache: true

set.seed(2024)
dat <- expand.grid(id = 1:N, trials = 1:nt, cond = c(0, 1))
sb1 <- 0.2 # slope standard deviation
alphai <- rnorm(N, 0, sb0)
b1i <- rnorm(N, 0, sb1)

ystar <- with(dat, alphai[id] + ((b1 + b1i[id]) * cond)) + rlogis(nrow(dat), 0, 1)
dat$y <- ordered(findInterval(ystar, alpha) + 1)
fit <- clmm(y ~ cond + (cond|id), data = dat, link = "logit", )
summary(fit)
```

### 2024-01-29 R/utils-plotting.R

@@ -129,7 +129,7 @@ cat_latent_plot <- function(location = 0,
     sum_to_1(prob0)
     alpha <- prob_to_alpha(prob0, link = link)
   }
-  
+
   data <- data.frame(x = factor(1:length(location)))
   k <- length(alpha) + 1
   if(length(location) > 1){

### 2024-01-29 paper/paper.Rmd
updating

@@
 The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models.
-The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Github repository.
+The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Open Science Framework repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).
 
@@
 We used `r cite_r()` for the code, figures and tables.
-Details about R packages an The tutorial is supported by a set of custom R functions available on Github.
-Other examples and model extensions along with details about used R packages are available on the Github repository (supplementary materials).
+Details about R packages an The tutorial is supported by a set of custom R functions that can be found on the online OSF repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).
 For the tutorial we used the following packages.
@@
 When assuming PO or considering only a location effect is limiting, including scale (*location-scale*), thresholds (*location-shift*) effects or relaxing the PO assumption can help in modeling complex phenomenon.
-The supplementary materials contains some examples of these complex models.
+The supplementary materials ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)) contains some examples of these complex models.
 Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.

### 2024-01-29 paper/paper.tex
updating

@@
 The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models.
-The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Github repository.
+The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Open Science Framework repository (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}).
 }
@@
 We used R {[}Version 4.3.1; R Core Team (2023){]} for the code, figures and tables.
-Details about R packages an The tutorial is supported by a set of custom R functions available on Github.
-Other examples and model extensions along with details about used R packages are available on the Github repository (supplementary materials).
+Details about R packages an The tutorial is supported by a set of custom R functions that can be found on the online OSF repository (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}).
 For the tutorial we used the following packages.
@@
 \centering
-\caption{\label{tab:tab-model-summary-latex}Summary of the probit and logit models in terms of link (\(g(\cdot)\)) and inverse link function (\(g^{-1}(\cdot)\)) and the corresponding R code.}
+\caption{\label{tab:tab-model-summary-latex}Summary of the probit and logit models in terms of link \(g(\cdot)\) and inverse link function \(g^{-1}(\cdot)\) and the corresponding R code.}
 \centering
@@
 The following code summarise the first steps of the simulation.
-Basically we define the simulation parameters, we calculate the \(k - 1\) thresholds \(\alpha\) and we apply Equations \eqref{eq:prob-cum-model1} and \eqref{eq:prob-cum-model2} calculating the probability of each response.
-
-The other parameters to define are the thresholds.
-We can consider them as intercepts thus the transformed probabilities of each ordinal response when \(x = 0\) (in this case the first group).
+We define the simulation parameters, calculate the \(k - 1\) thresholds \(\alpha\) and apply Equations \eqref{eq:prob-cum-model1} and \eqref{eq:prob-cum-model2} calculating the probability of each response.
+
+To set the thresholds, we can consider them as intercepts thus the transformed probabilities of each ordinal response when \(x = 0\) (in this case the first group).
 The \texttt{alpha\_to\_prob()} and \texttt{prob\_to\_alpha()} can be used to go back and forth from thresholds and probabilities.
@@
 \NormalTok{lp }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(lp)}
-\FunctionTok{head}\NormalTok{(lp)}
+\FunctionTok{head\_tail}\NormalTok{(lp, }\AttributeTok{n =} \DecValTok{3}\NormalTok{)}
 \CommentTok{\#\textgreater{}       lp\_leq1   lp\_leq2  lp\_leq3}
-\CommentTok{\#\textgreater{} 1 {-}1.098612       0 1.098612}
-\CommentTok{\#\textgreater{} 2 {-}1.098612       0 1.098612}
-\CommentTok{\#\textgreater{} 3 {-}1.098612       0 1.098612}
-\CommentTok{\#\textgreater{} 4 {-}1.098612       0 1.098612}
-\CommentTok{\#\textgreater{} 5 {-}1.098612       0 1.098612}
-\CommentTok{\#\textgreater{} 6 {-}1.098612       0 1.098612}
+\CommentTok{\#\textgreater{} 1   {-}1.098612  0.000000 1.098612}
+\CommentTok{\#\textgreater{} 2   {-}1.098612  0.000000 1.098612}
+\CommentTok{\#\textgreater{} 3   {-}1.098612  0.000000 1.098612}
+\CommentTok{\#\textgreater{} 98  {-}2.197225 {-}1.098612 0.000000}
+\CommentTok{\#\textgreater{} 99  {-}2.197225 {-}1.098612 0.000000}
+\CommentTok{\#\textgreater{} 100 {-}2.197225 {-}1.098612 0.000000}
 
@@
 \NormalTok{cump }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(cump)}
-\FunctionTok{names}\NormalTok{(cump) }\OtherTok{\textless{}{-}} \FunctionTok{sprintf}\NormalTok{(}\StringTok{"cump\_leq\%s"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{(k }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{)) }\CommentTok{\# giving appropriate names}
-\FunctionTok{head}\NormalTok{(cump)}
+
+\CommentTok{\# giving appropriate names, cump = cumulative probability, leq = less or equal}
+\FunctionTok{names}\NormalTok{(cump) }\OtherTok{\textless{}{-}} \FunctionTok{sprintf}\NormalTok{(}\StringTok{"cump\_leq\%s"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{(k }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{))}
+\FunctionTok{head\_tail}\NormalTok{(cump, }\AttributeTok{n =} \DecValTok{3}\NormalTok{)}
 \CommentTok{\#\textgreater{}     cump\_leq1 cump\_leq2 cump\_leq3}
-\CommentTok{\#\textgreater{} 1      0.25       0.5      0.75}
-\CommentTok{\#\textgreater{} 2      0.25       0.5      0.75}
-\CommentTok{\#\textgreater{} 3      0.25       0.5      0.75}
-\CommentTok{\#\textgreater{} 4      0.25       0.5      0.75}
-\CommentTok{\#\textgreater{} 5      0.25       0.5      0.75}
-\CommentTok{\#\textgreater{} 6      0.25       0.5      0.75}
+\CommentTok{\#\textgreater{} 1        0.25      0.50      0.75}
+\CommentTok{\#\textgreater{} 2        0.25      0.50      0.75}
+\CommentTok{\#\textgreater{} 3        0.25      0.50      0.75}
+\CommentTok{\#\textgreater{} 98       0.10      0.25      0.50}
+\CommentTok{\#\textgreater{} 99       0.10      0.25      0.50}
+\CommentTok{\#\textgreater{} 100      0.10      0.25      0.50}
 
@@
 \FunctionTok{names}\NormalTok{(p) }\OtherTok{\textless{}{-}} \FunctionTok{sprintf}\NormalTok{(}\StringTok{"p\%s"}\NormalTok{, }\DecValTok{1}\SpecialCharTok{:}\NormalTok{k) }\CommentTok{\# giving appropriate names}
-\FunctionTok{head}\NormalTok{(p)}
-\CommentTok{\#\textgreater{}     p1   p2   p3   p4}
-\CommentTok{\#\textgreater{} 1 0.25 0.25 0.25 0.25}
-\CommentTok{\#\textgreater{} 2 0.25 0.25 0.25 0.25}
-\CommentTok{\#\textgreater{} 3 0.25 0.25 0.25 0.25}
-\CommentTok{\#\textgreater{} 4 0.25 0.25 0.25 0.25}
-\CommentTok{\#\textgreater{} 5 0.25 0.25 0.25 0.25}
-\CommentTok{\#\textgreater{} 6 0.25 0.25 0.25 0.25}
 
@@
 
-\CommentTok{\# example of a random Y outcome based on the probabilities}
-\FunctionTok{sample}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ p[}\DecValTok{1}\NormalTok{, ])}
-\CommentTok{\#\textgreater{} [1] 1}
+\CommentTok{\# example of n random Y outcome based on the probabilities}
+\FunctionTok{sample}\NormalTok{(}\AttributeTok{x =} \DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{10}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ p[}\DecValTok{1}\NormalTok{, ], }\AttributeTok{replace =} \ConstantTok{TRUE}\NormalTok{)}
+\CommentTok{\#\textgreater{}  [1] 1 3 4 4 3 4 3 3 1 2}
 
@@
 \NormalTok{dat}\SpecialCharTok{$}\NormalTok{y }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(p, }\DecValTok{1}\NormalTok{, }\ControlFlowTok{function}\NormalTok{(ps) }\FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{k, }\AttributeTok{size =} \DecValTok{1}\NormalTok{, }\AttributeTok{prob =}\NormalTok{ ps))}
-\FunctionTok{head}\NormalTok{(dat)}
+\FunctionTok{head\_tail}\NormalTok{(dat, }\AttributeTok{n =} \DecValTok{3}\NormalTok{)}
 \CommentTok{\#\textgreater{}     x y}
-\CommentTok{\#\textgreater{} 1 0 3}
-\CommentTok{\#\textgreater{} 2 0 4}
+\CommentTok{\#\textgreater{} 1   0 1}
+\CommentTok{\#\textgreater{} 2   0 1}
 \CommentTok{\#\textgreater{} 3   0 4}
-\CommentTok{\#\textgreater{} 4 0 3}
-\CommentTok{\#\textgreater{} 5 0 4}
-\CommentTok{\#\textgreater{} 6 0 3}
+\CommentTok{\#\textgreater{} 98  1 4}
+\CommentTok{\#\textgreater{} 99  1 4}
+\CommentTok{\#\textgreater{} 100 1 2}
 
@@
 \CommentTok{\#\textgreater{}        1    2    3    4}
-\CommentTok{\#\textgreater{}   0 0.26 0.22 0.26 0.26}
-\CommentTok{\#\textgreater{}   1 0.06 0.08 0.30 0.56}
+\CommentTok{\#\textgreater{}   0 0.26 0.24 0.26 0.24}
+\CommentTok{\#\textgreater{}   1 0.06 0.10 0.30 0.54}
 
@@
 \CommentTok{\#\textgreater{}   1 0.26 0.06}
-\CommentTok{\#\textgreater{}   2 0.48 0.14}
-\CommentTok{\#\textgreater{}   3 0.74 0.44}
+\CommentTok{\#\textgreater{}   2 0.50 0.16}
+\CommentTok{\#\textgreater{}   3 0.76 0.46}
 \CommentTok{\#\textgreater{}   4 1.00 1.00}
@@
 \CommentTok{\#\textgreater{}        1        2        3 }
-\CommentTok{\#\textgreater{} 5.504505 5.670330 3.622378}
+\CommentTok{\#\textgreater{} 5.504505 5.250000 3.717391}
 
@@
 \CommentTok{\#\textgreater{}  link   threshold nobs logLik  AIC    niter max.grad cond.H }
-\CommentTok{\#\textgreater{}  probit flexible  100  {-}122.46 252.91 6(0)  2.05e{-}10 1.6e+01}
+\CommentTok{\#\textgreater{}  probit flexible  100  {-}124.19 256.38 5(0)  5.62e{-}08 1.5e+01}
 \CommentTok{\#\textgreater{} }
@@
 \CommentTok{\#\textgreater{}   Estimate Std. Error z value Pr(\textgreater{}|z|)    }
-\CommentTok{\#\textgreater{} x   0.8823     0.2290   3.854 0.000116 ***}
+\CommentTok{\#\textgreater{} x   0.8831     0.2276    3.88 0.000104 ***}
 \CommentTok{\#\textgreater{} {-}{-}{-}}
@@
 \CommentTok{\#\textgreater{}     Estimate Std. Error z value}
-\CommentTok{\#\textgreater{} 1|2  {-}0.6564     0.1788  {-}3.672}
-\CommentTok{\#\textgreater{} 2|3  {-}0.1026     0.1686  {-}0.608}
-\CommentTok{\#\textgreater{} 3|4   0.6974     0.1778   3.923}
+\CommentTok{\#\textgreater{} 1|2 {-}0.65508    0.17861  {-}3.668}
+\CommentTok{\#\textgreater{} 2|3 {-}0.04056    0.16814  {-}0.241}
+\CommentTok{\#\textgreater{} 3|4  0.75376    0.17871   4.218}
 \end{Highlighting}
@@
 \item
-  Define simulation parameters, baseline probabilities (\texttt{prob0}), sample size, etc.
+  Define simulation parameters, intercept probabilities (\texttt{prob0}), sample size, etc.
 \item
@@
 \item
-  Sample \(n\) outcomes from a \emph{multinomial} distribution, choosing between \(k\) alternatives using the calculated probabilities
+  Sample \(n\) outcomes from a \emph{multinomial} distribution using the calculated probabilities
 \item
@@
 
-A more efficient way to simulate an ordinal outcome is using the latent formulation of the model.
+More efficiently, we can use the latent formulation of the CM.
 This require simulating a standard linear regression using the appropriate data generation function (\emph{logistic} or \emph{normal}) and cutting the latent variable using the thresholds.
@@
 \item
-  Add the random errors \(\epsilon_i\) sampling from the chosen distribution (logistic or normal)
+  Add the random errors \(\epsilon_i\) sampling from the logistic or normal distributions
 \item
@@
 \NormalTok{probs0 }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{k, k) }\CommentTok{\# probabilities for the first group}
-\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs0, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
+\NormalTok{alpha }\OtherTok{\textless{}{-}} \FunctionTok{prob\_to\_alpha}\NormalTok{(probs0, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{) }\CommentTok{\# thresholds}
 \NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{each =}\NormalTok{ N}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
@@
 Then the model formula is specified within the function as \texttt{location\ =\ \textasciitilde{}} along with the vector of regression coefficients (\texttt{beta}), baseline probabilities (\texttt{prob0}) and the link function.
-The function return a dataset with the simulated \(Y\) and the latent variable (this is possible only because we simulate the data, otherwise \(Y^{\star}\) cannot be observed by definition).
+The function return a dataset with the simulated \(Y\) and the latent variable (this is possible only because we simulateds the data, otherwise \(Y^{\star}\) cannot be observed by definition).
 
@@
 \caption{Example of the \texttt{show\_alpha()} function.
-On the left the assumed latent variable and \(k - 1\) thresholds with the corresponding probabilities on the right.}\label{fig:fig-show-th-example}
+On the left the assumed latent variable and \(k - 1\) thresholds with the corresponding probabilities on the right.
+The distance between thresholds (quantiles) determine the probability of the \(k\) outcomes.}\label{fig:fig-show-th-example}
 \end{figure}
@@
 For \emph{logit} models we can set the odds ratio.
-Meaningful odds ratios can be derived from previous literature, meta-analyses or converting to other effect sizes.
+Meaningful odds ratios can be derived from previous literature, meta-analyses or converting from other effect sizes.
 For example, Sánchez-Meca, Marín-Martínez, and Chacón-Moscoso (2003) proposed some equations to convert between odds ratios and Cohen's \(d\).
@@
 
-The parameters interpretation is the same as introduced in the general case.
-The thresholds \(\alpha\) are fixed representing the probabilities \(P(Y = k)\) when all predictors are zero.
+The thresholds are fixed, representing the probabilities \(P(Y = k)\) when all predictors are zero.
 In this case by doing \texttt{alpha\_to\_prob(fit\$alpha,\ link\ =\ "probit")} we should recover the \texttt{probs0} vector.
-\texttt{x1} is the main effect thus the difference in \(z\) scores between \emph{a} and \emph{b} averaging over \texttt{x2}.
-The same holds for \texttt{x2}.
-\texttt{x1:x2} is the interaction thus difference of differences in \(z\) scores (see Figure \ref{fig:fig-effects-2-by-2-interaction}).
+\texttt{x11} is the main effect thus the difference in \(z\) scores between \emph{a} and \emph{b} averaging over \texttt{x2}.
+The same holds for \texttt{x21}.
+\texttt{x11:x21} is the interaction thus difference of differences in \(z\) scores (see Figure \ref{fig:fig-effects-2-by-2-interaction}).
 
@@
 
-Again, \texttt{x} is the slope of the numerical predictor averaging over \texttt{g} (given that \texttt{g} has been coded with sum-to-zero contrasts) thus the increase in \(z\) scores for a unit increase in \texttt{x}.
+Again, \texttt{x} is the slope of the numerical predictor averaging over \texttt{g} (given that \texttt{g} has been coded with sum-to-zero contrast) thus the increase in \(z\) scores for a unit increase in \texttt{x}.
 \texttt{g1} is the main effect of the factor evaluated when \(x = 0\) (centering \(x\) will change the parameter interpretation).
-The \texttt{x:g1} is the slopes difference the two groups.
+The \texttt{x:g1} is the difference between the slopes of the two groups.
 Figure \ref{fig:fig-effects-num-by-cat-interaction} depicts the model results.
@@
 
+\section{Conclusions}\label{conclusions}
+
+In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation.
+The basic model can be expanded in several ways.
+When assuming PO or considering only a location effect is limiting, including scale (\emph{location-scale}), thresholds (\emph{location-shift}) effects or relaxing the PO assumption can help in modeling complex phenomenon.
+The supplementary materials (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}) contains some examples of these complex models.
+Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.
+
 \newpage

### 2024-01-30 paper/paper.Rmd
update

@@
 abstract: |
-  Ordinal data are very common in psychology such as Likert items, ratings or generic ordered variables.
-Despite the widespread, these variables are usually analyzed using metric models (e.g., standard linear regression).
-This approach can have important drawbacks in terms of statistical inference (e.g., reduced power and increased type-1 error) and prediction.
-One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis.
-The aim of the tutorial is presenting ordinal regression models using a simulation-based approach.
-In the first section we introduced the general model highlighting the crucial components and assumptions.
-Then we explained how to intepret the parameters for a logit and probit model.
-Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor.
+    Ordinal data are widespread in psychology such as Likert items, ratings, or generic ordered variables.
+These variables are usually analyzed using metric models (e.g., standard linear regression) with important drawbacks in terms of statistical inference (reduced power and increased type-1 error) and prediction.
+One possible reason for not using ordinal regression models could be difficulty in understanding parameters or conducting a power analysis.
+The tutorial aims to present ordinal regression models using a simulation-based approach.
+Firstly, we introduced the general model highlighting crucial components and assumptions.
+Then we explained how to interpret parameters for a logit and probit model.
+Then we proposed two ways for simulating data as a function of predictors showing a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor.
 Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors.
 The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models.
-The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Open Science Framework repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).
-
-keywords          : "ordinal, simulations, power"
-wordcount         : "X"
+The code to reproduce the proposed simulation, the custom R functions, and additional examples of ordinal regression models can be found on the online Open Science Framework repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).
+
+keywords          : "ordinal regression, monte carlo simulations, power analysis"
+wordcount         : "6987"
 bibliography:
@@
 
-We can identify three main models: *cumulative models* [CM, @McCullagh1980-cw; @Agresti2010-rz], *sequential models* [@Tutz1990-fe], and *adjacent category models*.
+We can identify three main models: *cumulative models* [CM, @Agresti2010-rz;  @McCullagh1980-cw], *sequential models* [@Tutz1990-fe], and *adjacent category models*.
 Among these, the CM is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
@@
 
-In this section we introduce some notation for the CM to be consistent with the existing literature and the `ordinal` R package [@Christensen2022-jk].
+In this section we introduce some notation for the CM to be consistent with the existing literature and the `ordinal` R package.
 We define $Y_k$ as the observed ordinal variable with $k$ levels and $Y^\star$ is the underlying latent variable.
@@
 The proposed CM can be parametrized in different ways.
-@Liddell2018-wu and @Kruschke2015-re proposed a Bayesian version of the model and @Gelman2020-tg proposed other thresholds parametrizations.
+@Liddell2018-wu proposed a Bayesian version of the model and @Gelman2020-tg proposed other thresholds parametrizations.
 
@@
 
-For fitting the CM we used the `ordinal` package [@Christensen2019-cz].
+For fitting the CM we used the `ordinal` package [@Christensen2023-yi].
 The `ordinal` package provide a complete and intuitive way to implement the ordinal models.
@@
 
-[^2]: For a very complete overview of the ordinal package see @Christensen2019-cz and the Package documentation <https://cran.r-project.org/web/packages/ordinal/ordinal.pdf>
+[^2]: For a very complete overview of the ordinal package see <https://cran.r-project.org/web/packages/ordinal/ordinal.pdf> and <https://cran.uni-muenster.de/web/packages/ordinal/vignettes/clm_article.pdf>
 
@@
   kable(booktabs = TRUE, format = "latex", caption = filor::get_caption(), escape = FALSE, align = "c", row.names = FALSE) |> 
-  kable_styling(full_width = FALSE) |> 
+  kable_styling(full_width = FALSE, latex_options = c("scale_down")) |> 
   row_spec(0, bold = TRUE)
@@
 
-```{r fig-effects-num-by-cat-interaction, fig.place = TRUE, fig.width=9, fig.height=9, fig.cap=filor::get_caption()}
+```{r fig-effects-num-by-cat-interaction, fig.place = TRUE, fig.width=7, fig.height=12, fig.cap=filor::get_caption()}
 # plot(ggpredict(fit, terms = c("x[all]", "g")))
@@
   geom_line(aes(group = group), linewidth = 1) +
-  facet_grid(response.level~1) +
+  facet_grid(response.level~.) +
   ylab("Probability") +

### 2024-01-30 paper/paper.tex
update

@@
 {}{\PatchFailed}
-\keywords{ordinal, simulations, power\newline\indent Word count: X}
+\keywords{ordinal regression, monte carlo simulations, power analysis\newline\indent Word count: 6987}
 \usepackage{csquotes}
@@
   pdflang={en-EN},
-  pdfkeywords={ordinal, simulations, power},
+  pdfkeywords={ordinal regression, monte carlo simulations, power analysis},
   hidelinks,
@@
 \abstract{%
-Ordinal data are very common in psychology such as Likert items, ratings or generic ordered variables.
-Despite the widespread, these variables are usually analyzed using metric models (e.g., standard linear regression).
-This approach can have important drawbacks in terms of statistical inference (e.g., reduced power and increased type-1 error) and prediction.
-One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis.
-The aim of the tutorial is presenting ordinal regression models using a simulation-based approach.
-In the first section we introduced the general model highlighting the crucial components and assumptions.
-Then we explained how to intepret the parameters for a logit and probit model.
-Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor.
+Ordinal data are widespread in psychology such as Likert items, ratings, or generic ordered variables.
+These variables are usually analyzed using metric models (e.g., standard linear regression) with important drawbacks in terms of statistical inference (reduced power and increased type-1 error) and prediction.
+One possible reason for not using ordinal regression models could be difficulty in understanding parameters or conducting a power analysis.
+The tutorial aims to present ordinal regression models using a simulation-based approach.
+Firstly, we introduced the general model highlighting crucial components and assumptions.
+Then we explained how to interpret parameters for a logit and probit model.
+Then we proposed two ways for simulating data as a function of predictors showing a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor.
 Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors.
 The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models.
-The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Open Science Framework repository (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}).
+The code to reproduce the proposed simulation, the custom R functions, and additional examples of ordinal regression models can be found on the online Open Science Framework repository (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}).
 }
@@
 
-We can identify three main models: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}.
+We can identify three main models: \emph{cumulative models} (CM, Agresti, 2010; McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}.
 Among these, the CM is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds.
@@
 
-In this section we introduce some notation for the CM to be consistent with the existing literature and the \texttt{ordinal} R package (Christensen, 2022).
+In this section we introduce some notation for the CM to be consistent with the existing literature and the \texttt{ordinal} R package.
 We define \(Y_k\) as the observed ordinal variable with \(k\) levels and \(Y^\star\) is the underlying latent variable.
@@
 The proposed CM can be parametrized in different ways.
-Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model and Gelman, Hill, and Vehtari (2020) proposed other thresholds parametrizations.
+Liddell and Kruschke (2018) proposed a Bayesian version of the model and Gelman, Hill, and Vehtari (2020) proposed other thresholds parametrizations.
 
@@
 
-For fitting the CM we used the \texttt{ordinal} package (Christensen, 2019).
+For fitting the CM we used the \texttt{ordinal} package (Christensen, 2023).
 The \texttt{ordinal} package provide a complete and intuitive way to implement the ordinal models.
-The syntax is very similar to standard linear models in R with functions to calculate predictions, perform model comparison and extract relevant model information.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}
+The syntax is very similar to standard linear models in R with functions to calculate predictions, perform model comparison and extract relevant model information.\footnote{For a very complete overview of the ordinal package see \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf} and \url{https://cran.uni-muenster.de/web/packages/ordinal/vignettes/clm_article.pdf}}
 
@@
 \centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
 \begin{tabular}[t]{cccccccc}
@@
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
@@
 
-\bibitem[\citeproctext]{ref-Christensen2019-cz}
-Christensen, R. H. B. (2019). \emph{Ordinal---regression models for ordinal data}.
-
-\bibitem[\citeproctext]{ref-Christensen2022-jk}
-Christensen, R. H. B. (2022). \emph{Ordinal---regression models for ordinal data}.
+\bibitem[\citeproctext]{ref-Christensen2023-yi}
+Christensen, R. H. B. (2023). \emph{Ordinal---regression models for ordinal data}. Retrieved from \url{https://CRAN.R-project.org/package=ordinal}
 
@@
 \bibitem[\citeproctext]{ref-Fox2015-ps}
-Fox, J. (2015). \emph{Applied regression analysis and generalized linear models}. SAGE Publications. Retrieved from \url{https://play.google.com/store/books/details?id=BlsKogEACAAJ}
+Fox, J. (2015). \emph{Applied regression analysis and generalized linear models}. SAGE Publications.
 
@@
 Kemp, S., \& Grace, R. C. (2021). Using ordinal scales in psychology. \emph{Methods in Psychology (Online)}, \emph{5}, 100054. \url{https://doi.org/10.1016/j.metip.2021.100054}
-
-\bibitem[\citeproctext]{ref-Kruschke2015-re}
-Kruschke, J. K. (2015). \emph{Doing bayesian data analysis}. \url{https://doi.org/10.1016/c2012-0-00477-2}
 
@@
 \bibitem[\citeproctext]{ref-Likert1932-xw}
-Likert, R. (1932). A technique for the measurement of attitudes. \emph{Archives of Psychology}, \emph{22}, 55. Retrieved from \url{https://psycnet.apa.org/record/1933-01885-001}
+Likert, R. (1932). A technique for the measurement of attitudes. \emph{Archives of Psychology}, \emph{22}, 55.
 
@@
 \emph{Neuroscience of Consciousness}.
-Retrieved from \url{https://academic.oup.com/nc/article-pdf/doi/10.1093/nc/niab044/41765024/niab044.pdf}
+\url{https://doi.org/10.1093/nc/niab044/41765024/niab044}
 

### 2024-01-30 submission/cover-letter/cover-letter.Rmd
update

---
author: "Filippo Gambarota"
role: "Postdoctoral Researcher"
journal: "International Journal of Psychology"
affiliation: "Department of Developmental Psychology and Socialization"
paper: "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis"
university: "University of Padova"
editor: "Prof. Antonio Zuffianò and Prof. David Giofrè"
date: "31\\textsuperscript{th} January 2024"
output: filor::letter
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We wish to submit our work "*`r rmarkdown::metadata$paper`*" for consideration by `r rmarkdown::metadata$journal` (special issue "Tutorials on Quantitative Methods"). Ordinal data are very common in any area of Psychology from social psychology, to clinical assessment and also experimental psychology. Several authors suggested that ordinal data requires appropriate descriptive statistics and modeling methods but very often researchers ignore the ordinal nature of the variable using standard statistical techniques e.g., linear regressions. This approach can introduce inferential errors, biased estimates, and wrong predictions.

Ordinal regression models are not new statistical approaches but one of the reasons why researchers are not using them could be the challenging statistical implementation, the not intuitive parameters interpretation, and the difficulty in implementing research-related activities such as power or design analysis. All these aspects are surely more challenging for ordinal models compared to the standard linear regression.

This tutorial aims to introduce a specific but very applicable class of ordinal regression models called cumulative models. We adopted a specific approach for the tutorial as a mixture of concise but necessary statistical background (model assumptions, minimal equations, and model components). Beyond the theoretical introduction, we used a simulation-based approach where we showed how to generate realistic data under different common research scenarios. The simulation-based approach has several advantages in terms of forcing people to understand the model foundations still not requiring them to completely understand the statistical details. Furthermore, introducing Monte Carlo simulations has the advantage of giving a powerful tool also for daily research work. By using simulations, researchers can not only understand but plan their experiments/research in terms of power and design analysis. We presented a simple but extendable framework to simulate data and implement a power analysis. We believe that when presenting or revisiting a statistical approach is always important to suggest or propose methods to estimate the statistical power. Given the widespread of registered reports and preregistration, power analysis is often a mandatory step.

In addition to the topics and examples proposed in the tutorial, we created a set of custom R functions supporting the simulation and visualization of ordinal models. Furthermore, we added other examples and model extensions such as mixed-effects models and scale-location models. All the materials to reproduce the paper and the additional materials are available on the online OSF repository ([https://osf.io/93h5j/](https://osf.io/93h5j/)) linked to a Github repository ([github.com/shared-research/ordinalsim](github.com/shared-research/ordinalsim)).

Thank you for your consideration of this manuscript.

Best regards,

Filippo Gambarota and Gianmarco Altoè

### 2024-01-30 submission/cover-letter/cover-letter.Rmd
updating

@@
 
-We wish to submit our work "*`r rmarkdown::metadata$paper`*" for consideration by `r rmarkdown::metadata$journal` (special issue "Tutorials on Quantitative Methods").
+We wish to submit our full manuscript "*`r rmarkdown::metadata$paper`*" for consideration by `r rmarkdown::metadata$journal` (special issue "Tutorials on Quantitative Methods").
 Ordinal data are very common in any area of Psychology from social psychology, to clinical assessment and also experimental psychology.

### 2024-01-31 paper/paper.Rmd
updating

@@
 The code to reproduce the proposed simulation, the custom R functions, and additional examples of ordinal regression models can be found on the online Open Science Framework repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).
-
 keywords          : "ordinal regression, monte carlo simulations, power analysis"
@@
   preprint: true
+  submission: true
 classoption       : "man"
 output: 
+  papaja::apa6_docx: default
   papaja::apa6_pdf:
@@
       - \usepackage{booktabs, caption, longtable, colortbl, array, hhline, mathbf}
-  papaja::apa6_docx: default
 ---
@@
 ```{r tab-dataset-example-docx, eval = which_output("docx")}
-papaja::apa_table(ss, caption = filor::get_caption())
+papaja::apa_table(ss, caption = filor::get_caption(), escape = FALSE)
 ```
@@
 The supplementary materials ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)) contains some examples of these complex models.
-Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.
+Furthermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.
+
+```{r, child = 'ethical-section.Rmd'}
+```
 

### 2024-01-31 paper/paper.tex
updating

@@
 The supplementary materials (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}) contains some examples of these complex models.
-Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.
+Furthermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.
+
+\section{Ethical Compliance Section}\label{ethical-compliance-section}
+
+\subsubsection{Funding}\label{funding}
+
+The authors have no funding to disclose.
+
+\subsubsection{Compliance with Ethical Standards}\label{compliance-with-ethical-standards}
+
+Compliance with Ethical Standards: Not applicable
+
+\subsubsection{Conflicts of Interest}\label{conflicts-of-interest}
+
+The authors declare they have no conflict of interest
 

### 2024-01-31 supplementary/location-scale-models.qmd
updating

@@
 In this example, the only way to capture group differences is by including a scale effect.
-The @fig-scale-effect-example depict the impact of having different scales between two groups on the ordinal probabilities.
+The @fig-scale-effect-example depict the impact of having different scales between two groups on the ordinal probabilities but the same location.
+Clearly, there is no location effect and the two groups are predicted to be the same considering only the location effect.
 
@@
 #| label: fig-scale-effect-example
-cat_latent_plot(location = c(0, 1), scale = c(1, 2), prob0 = rep(1/4, 4), link = "logit", plot = FALSE) |> 
+cat_latent_plot(location = c(0, 0), scale = c(1, 2), prob0 = rep(1/4, 4), link = "logit", plot = FALSE) |> 
   cowplot::plot_grid(plotlist = _, nrow = 1)
@@
 Again we simulate that the baseline probabilities are uniform for the first group.
+The $\log$ link function (and $e$ as inverse link function) is used to make sure that the variance is always positive.
+If the first group has a scale of 1 and the second group a scale of 2, the $\zeta_1 = \log(\frac{s_2}{s_1}) = \log(\frac{2}{1}) \approx 0.7$.
+Thus the two groups have a log difference in terms of scale of 0.7 (or a ratio of 2).
 
 ```{r}
-k <- 5  # number of options (1)
-n <- 1e5 # number of observations (1)
-b1 <- 0.5 # beta1, the shift in the latent distribution (2)
+#| echo: true
+k <- 5  # number of options
+n <- 1e5 # number of observations
+b1 <- 0.5 # beta1, the shift in the latent distribution
 z1 <- log(2) # zeta1, the change in the scale
-probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
-alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
+probs0 <- rep(1/k, k) # probabilities when x = 0
+alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities
 dat <- data.frame(x = rep(c(0, 1), each = n/2))
@@
 fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
-```
-
-The table (put ref) reports the simulation results.
-
-```{r}
-#| eval: false
-truth <- c(alphas, b1, z1)
-names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
-#clm_table(fit, truth)
+summary(fit)
 ```

### 2024-01-31 supplementary/ordinal-notes.qmd
updating

@@
 ```{r}
-#| fig-cap: Simulated data using the latent variable approach. The dotted lines are the thresholds $\\alpha$
+#| fig-cap: Simulated data using the latent variable approach. The dotted lines are the thresholds $\alpha$
 n <- 1e3

### 2024-01-31 supplementary/simulating-glm.qmd
major update
updating

@@
 library(tidyverse)
+library(cowplot)
+library(ordinal)
+devtools::load_all()
+
+knitr::opts_chunk$set(echo = TRUE,
+                      dev = "svg",
+                      warning = FALSE,
+                      message = FALSE)
+mtheme <- function(){
+  theme_minimal(15)
+}
+theme_set(mtheme())
 ```
@@
 
-The curve depict the true probability of success for each value of $x$.
+The curve depicts the true probability of success for each value of $x$.
 To include the random error we need to sample the observed values from a Binomial distribution using the vector of probabilities `p`.
@@
 y <- rbinom(ns, 1, p)
-head(y)
+summary(p)
 

### 2024-01-31 supplementary/useful-resources.qmd
updating

---
title: "Useful Resources"
format: html
---

In this document we collect some useful resources for ordinal regression models.

# Books

- [Analysis of Ordinal Categorical Data](https://onlinelibrary.wiley.com/doi/book/10.1002/9780470594001): a very complete book about modeling ordinal data.
- [Categorical Data Analysis](https://onlinelibrary.wiley.com/doi/book/10.1002/0471249688)
- [An Introduction to Categorical Data Analysis](https://onlinelibrary.wiley.com/doi/book/10.1002/0470114754): Introduction about categorical data in R with some examples with ordinal data (Chapter 6)
- [Doing Bayesian Data Analysis](https://www.sciencedirect.com/book/9780124058880/doing-bayesian-data-analysis): Overview of Bayesian data analysis. Chapter 23 is about ordinal regression models.
- [Regression and Other Stories](https://avehtari.github.io/ROS-Examples/): Chapter 15 contains some examples and parametrizations of ordinal regression models

# Posts

- [Violation of Proportional Odds is Not Fatal](https://www.fharrell.com/post/po/)
- [Assessing the Proportional Odds Assumption and Its Impact](https://www.fharrell.com/post/impactpo/)
- [Resources for Ordinal Regression Models](https://www.fharrell.com/post/rpo/)
- [Bayesian Estimation of Signal Detection Models](https://mvuorre.github.io/posts/2017-10-09-bayesian-estimation-of-signal-detection-theory-models/)
- [Causal inference with ordinal regression](https://solomonkurz.netlify.app/blog/2023-05-21-causal-inference-with-ordinal-regression/)

# Packages

- `ordinal`: https://cran.r-project.org/web/packages/ordinal/index.html
- `VGAM`: https://cran.r-project.org/web/packages/VGAM/index.html
- `brms`: https://paul-buerkner.github.io/brms/
- `MASS::polr()`: https://rdrr.io/cran/MASS/man/polr.html
- `ordDisp`: https://cran.r-project.org/web/packages/ordDisp/index.html

### 2024-01-31 R/utils-plotting.R

@@ -29,10 +29,10 @@ show_alpha <- function(prob = NULL,
   
   if(link == "logit"){
     xlim <- c(-7, 7)
-    dist <- distributional::dist_logistic
+    dist <- function(...) distributional::dist_logistic(...)
   }else{
     xlim <- c(-5, 5)
-    dist <- distributional::dist_normal
+    dist <- function(...) distributional::dist_normal(...)
   }
   
   k <- length(prob)
@@ -68,7 +68,6 @@ show_alpha <- function(prob = NULL,
   pplot <- ggplot(data = NULL, aes(x = xlim)) +
     geom_line(stat = "function", 
               fun = lf$pfun, 
-              #aes(color = factor(after_stat(findInterval(x, alpha)))),
               linewidth = 1.5) +
     geom_segment(data = NULL, aes(x = alpha, y = 0, xend = alpha, yend = lf$pfun(alpha))) +
     geom_segment(data = NULL,aes(x = -Inf, y = lf$pfun(alpha), xend = alpha, yend = lf$pfun(alpha))) +
@@ -125,6 +124,10 @@ cat_latent_plot <- function(location = 0,
   require(ggplot2)
   lf <- get_link(link)
   
+  if(is.null(prob0) & is.null(alpha)){
+    stop("alpha or prob0 must be specified!")
+  }
+  
   if(is.null(alpha)){
     sum_to_1(prob0)
     alpha <- prob_to_alpha(prob0, link = link)
@@ -148,11 +151,11 @@ cat_latent_plot <- function(location = 0,
   thl <- latex2exp::TeX(sprintf("$\\alpha_{%s}$", 1:length(alpha)))
   
   if(link == "logit"){
-    dfun <- distributional::dist_logistic
+    dfun <- function(...) distributional::dist_logistic(...)
     sd <- sqrt(vlogit(scale))
     ylab <- latex2exp::TeX("Logistic Distribution $Y^{*}$")
   }else{
-    dfun <- distributional::dist_normal
+    dfun <- function(...) distributional::dist_normal(...)
     sd <- data$scale
     ylab <- latex2exp::TeX("Gaussian Distribution $Y^{*}$")
   }
@@ -223,6 +226,16 @@ num_latent_plot <- function(x,
                             size = 20,
                             linewidth = 1,
                             plot = TRUE){
+  
+  if(is.null(prob0) & is.null(alpha)){
[…]

### 2024-01-31 R/utils.R

@@ -2,9 +2,9 @@
 #'
 #' @param location a formula with predictors for the location parameter. The components of the formula need to be columns of the \code{data} object.
 #' @param scale a formula with predictors for the scale parameter. Default to \code{NULL} thus if not specified the scale is assumed to be 1.
-#' @param By vector of regression coefficients for the location formula. The order is the same as returned by running \code{model.matrix(~formula, data = data)}
-#' @param Bscale vector of regression coefficients for the scale formula. Default to \code{NULL} thus if not specified the scale is assumed to be 1.
-#' @param prob vector of baseline probabilities from which thresholds are calculated. They have a role similar to intercepts in standard regression modelling. The probabilities are fixed assuming all predictors to 0.
+#' @param beta vector of regression coefficients for the location formula. The order is the same as returned by running \code{model.matrix(~formula, data = data)}
+#' @param zeta vector of regression coefficients for the scale formula. Default to \code{NULL} thus if not specified the scale is assumed to be 1.
+#' @param prob0 vector of baseline probabilities from which thresholds are calculated. They have a role similar to intercepts in standard regression modelling. The probabilities are fixed assuming all predictors to 0.
 #' @param alpha vector of baseline thresholds. Either \code{probs} or \code{alpha} must be specified. The function calculated thresholds from \code{probs} and viceversa
 #' @param data a dataframe with predictors. both \code{location} and \code{scale} formulas refers to columns of the dataframe.
 #' @param link character vector indicating the link function. Can be \code{logit} or \code{probit}.
@@ -20,7 +20,7 @@
 #' k <- 5 # 5 ordinal values
 #' probs0 <- rep(1/k, k) # baseline probabilities for the group 0
 #' dat <- data.frame(x = rep(c(0, 1), each = n))
-#' dat <- sim_ord_latent(~x, By = 0.5, prob = probs0, data = dat, link = "probit")
+#' dat <- sim_ord_latent(~x, beta = 0.5, prob0 = probs0, data = dat, link = "probit")
 sim_ord_latent <- function(location,
                            scale = NULL,
                            beta,

### 2024-01-31 paper/render.R

@@ -1,0 +1,42 @@
+mkdirif <- function(dir, ...){
+  if(!dir.exists(dir)){
+    dir.create(dir, ...)
+  }
+}
+
+# render paper
+
+rmarkdown::render("paper/paper.Rmd", output_format = "all") # pdf and word
+
+# anonymize paper
+
+anonymize_paper("paper/paper.pdf")
+
+# figures
+
+figs_names <- c(
+  "fig-explain-cumulative",
+  "fig-logit-vs-probit",
+  "fig-odds-example",
+  "fig-prop-odds",
+  "fig-sim-from-latent",
+  "fig-show-th-example",
+  "fig-example-cat-latent",
+  "fig-example-num-latent",
+  "fig-effects-2-by-2-interaction",
+  "fig-effects-num-by-cat-interaction",
+  "fig-power-curve"
+)
+
+figs <- paste0("figure", 1:length(figs_names))
+names(figs) <- paste0(figs_names, "-1")
+
+mkdirif("paper/figures-submission")
+actual_figures <- list.files("paper/paper_files/figure-latex/", pattern = "*.pdf", full.names = TRUE)
+fs::file_copy(actual_figures, "paper/figures-submission/")
+submission_figures <- list.files("paper/figures-submission/", pattern = "*.pdf")
+
+fs::file_move(
+  file.path("paper", "figures-submission", submission_figures),
+  file.path("paper", "figures-submission", sprintf("%s.pdf", figs[tools::file_path_sans_ext(submission_figures)]))
+)

### 2024-01-31 tests/testthat/test-utils.R

@@ -4,8 +4,8 @@ test_that("alpha_to_prob() and prob_to_alpha() are working", {
   p1 <- rep(1/6, 6)
   a1l <- prob_to_alpha(p1, link = "logit")
   a1p <- prob_to_alpha(p1, link = "probit")
-  expect_equal(p1, alpha_to_prob(a1l, link = "logit"), info = "logit link")
-  expect_equal(p1, alpha_to_prob(a1p, link = "probit"), info = "probit link")
+  expect_equal(p1, alpha_to_prob(a1l, link = "logit"), info = "logit link", ignore_attr = TRUE)
+  expect_equal(p1, alpha_to_prob(a1p, link = "probit"), info = "probit link", ignore_attr = TRUE)
 })
 
 test_that("dummy_ord() is working!", {
@@ -24,7 +24,7 @@ test_that("sim_ord_latent(link = 'logit') works with a binary x, k = 4", {
   alpha <- prob_to_alpha(probs0, link = "logit")
   b1 <- log(2) # log odds ratio
   dat <- data.frame(x = rep(0:1, each = n))
-  dat <- sim_ord_latent(~x, By = b1, prob = probs0, data = dat, link = "logit")
+  dat <- sim_ord_latent(~x, beta = b1, prob0 = probs0, data = dat, link = "logit")
   fit <- ordinal::clm(y~x, data = dat, link = "logit")
   
   expect_equal(b1, fit$beta, tolerance = .tol["clm"], info = "beta", ignore_attr = TRUE)
@@ -43,7 +43,7 @@ test_that("sim_ord_latent(link = 'probit') works with a binary x, k = 4", {
   alpha <- prob_to_alpha(probs0, link = "probit")
   b1 <- log(2) # log odds ratio
   dat <- data.frame(x = rep(0:1, each = n))
-  dat <- sim_ord_latent(~x, By = b1, prob = probs0, data = dat, link = "probit")
+  dat <- sim_ord_latent(~x, beta = b1, prob0 = probs0, data = dat, link = "probit")
   fit <- ordinal::clm(y~x, data = dat, link = "probit")
   
   expect_equal(b1, fit$beta, tolerance = .tol["clm"], info = "beta", ignore_attr = TRUE)
@@ -62,7 +62,7 @@ test_that("sim_ord_latent(link = 'logit') works with a continuous x, k = 4", {
[…]
